---
title: "09 Supplement — Extended Exercises"
subtitle: "WBA-Softmax, Multilevel SBC, and Bidirectional Model Recovery"
output: html_document
date: "2025-03-10"
---

```{r sup_setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "80%",
  dpi        = 300
)

regenerate_simulations <- FALSE

for (d in c("stan", "simdata", "simmodels", "figures")) {
  if (!dir.exists(d)) dir.create(d)
}

pacman::p_load(
  tidyverse,
  cmdstanr,
  posterior,
  bayesplot,
  patchwork,
  loo,
  SBC,
  furrr,
  future
)

theme_set(theme_classic())
plan(sequential)
```

# Extended Exercises for Chapter 9

> **📍 Prerequisites:** These exercises depend on code objects defined in
> the main Chapter 9. Run Chapter 9 first (or at minimum, compile the
> Stan models and generate the simulated datasets used there). Each
> exercise is self-contained once those objects exist.

The main chapter deliberately scoped out three analyses to keep the
narrative focused. Each of those analyses is a genuine validation step
that the cheatsheet mandates, and each teaches something the main chapter
does not. This supplement provides full implementations.

---

## Exercise 1: The Softmax Observation Model

### Motivation

Chapter 9 names the BetaBinomial observation model as a cognitive
hypothesis — probability matching — and presents a table of
alternatives. The most important alternative is the **softmax** (inverse
temperature) model, which replaces the two-step BetaBinomial draw with a
compressed logistic transformation of the posterior mean. This exercise
builds that model, validates it via SBC, and uses LOO to compare it
against the BetaBinomial WBA on the same simulated data.

The distinction matters because:

- The BetaBinomial propagates the agent's **full belief uncertainty**
  into the response. An agent who is genuinely uncertain produces more
  variable choices.
- The softmax collapses the belief to a single number (the posterior
  mean $\hat{\theta}$) and uses an inverse temperature $\phi$ to control
  how deterministically the agent acts on that number. It discards
  distributional information.

If real participants are more deterministic than probability matching
predicts, the WBA-BetaBinomial will show a hump-shaped LOO-PIT
(overestimating variability), and the softmax model will improve LOO.

### The WBA-Softmax Stan Model

The belief-update mechanism is identical to the WBA. Only the
observation model changes: instead of $y_i \sim \text{BetaBin}(1,
\alpha_i, \beta_i)$, we compute $\hat{\theta}_i = \alpha_i / (\alpha_i
+ \beta_i)$ and pass it through a logistic link scaled by $\phi$:

$$y_i \sim \text{Bernoulli}\!\left(\sigma\!\left(\phi \cdot
\text{logit}(\hat{\theta}_i)\right)\right)$$

At $\phi = 1$ the model is a logistic regression on the posterior mean.
At $\phi \to \infty$ it approaches a threshold rule. At $\phi \to 0$ it
approaches random responding.

> **Identifiability caution.** The parameters $\kappa$ and $\phi$ are
> partially confounded: increasing $\kappa$ makes the Beta posterior
> mean more extreme (higher effective sample size), and increasing $\phi$
> amplifies any departure from 0.5. Separating them requires trials
> where the *shape* of the evidence (how ambiguous the posterior is)
> matters independently of its *direction*. The 9 × 4 evidence grid
> helps, but SBC may reveal wider calibration bands for $\kappa$ than
> in the BetaBinomial WBA. This is expected and informative.

```{r ex1_softmax_stan}
WBA_Softmax_stan <- "
// WBA with softmax observation model.
// Same belief update as the BetaBinomial WBA (rho, kappa parameterisation).
// Observation model: Bernoulli with logistic link on the posterior mean.
// phi (inverse temperature) controls response determinism.
data {
  int<lower=1> N;
  array[N] int<lower=0, upper=1> choice;
  array[N] int<lower=0> blue1;
  array[N] int<lower=0> total1;
  array[N] int<lower=0> blue2;
  array[N] int<lower=0> total2;
}

parameters {
  real<lower=0, upper=1> rho;
  real<lower=0>          kappa;
  real<lower=0>          phi;   // inverse temperature
}

transformed parameters {
  real<lower=0> weight_direct = rho * kappa;
  real<lower=0> weight_social = (1.0 - rho) * kappa;
}

model {
  // Priors — rho and kappa identical to the BetaBinomial WBA
  target += beta_lpdf(rho   | 2, 2);
  target += lognormal_lpdf(kappa | log(2), 0.5);
  // phi: lognormal(0, 1) spans roughly [0.1, 7].
  // phi < 0.5 = near-random; phi > 5 = near-deterministic.
  target += lognormal_lpdf(phi   | 0, 1);

  for (i in 1:N) {
    real alpha_post = 0.5
                    + weight_direct * blue1[i]
                    + weight_social * blue2[i];
    real beta_post  = 0.5
                    + weight_direct * (total1[i] - blue1[i])
                    + weight_social * (total2[i] - blue2[i]);
    // Posterior mean of the Beta belief distribution
    real theta_hat = alpha_post / (alpha_post + beta_post);
    // Softmax observation model
    target += bernoulli_logit_lpmf(choice[i] | phi * logit(theta_hat));
  }
}

generated quantities {
  vector[N] log_lik;
  array[N] int posterior_pred;
  array[N] int prior_pred;

  real rho_pr   = beta_rng(2, 2);
  real kappa_pr = lognormal_rng(log(2), 0.5);
  real phi_pr   = lognormal_rng(0, 1);
  real wd_pr    = rho_pr * kappa_pr;
  real ws_pr    = (1.0 - rho_pr) * kappa_pr;

  for (i in 1:N) {
    real alpha_post = 0.5
                    + weight_direct * blue1[i]
                    + weight_social * blue2[i];
    real beta_post  = 0.5
                    + weight_direct * (total1[i] - blue1[i])
                    + weight_social * (total2[i] - blue2[i]);
    real theta_hat = alpha_post / (alpha_post + beta_post);

    log_lik[i]        = bernoulli_logit_lpmf(choice[i] | phi * logit(theta_hat));
    posterior_pred[i] = bernoulli_logit_rng(phi * logit(theta_hat));

    // Prior predictive
    real ap = 0.5 + wd_pr * blue1[i] + ws_pr * blue2[i];
    real bp = 0.5 + wd_pr * (total1[i] - blue1[i])
                  + ws_pr * (total2[i] - blue2[i]);
    real th_pr = ap / (ap + bp);
    prior_pred[i] = bernoulli_logit_rng(phi_pr * logit(th_pr));
  }
}
"

write_stan_file(WBA_Softmax_stan,
                dir      = "stan/",
                basename = "ch9_wba_softmax.stan")

mod_softmax <- cmdstan_model("stan/ch9_wba_softmax.stan")
```

### SBC: WBA-Softmax

```{r ex1_sbc_softmax, message=FALSE}
n_sbc_softmax <- 500

if (regenerate_simulations) {

  sbc_gen_softmax <- SBC_generator_function(
    function(N) {
      rho_t   <- rbeta(1, 2, 2)
      kappa_t <- rlnorm(1, log(2), 0.5)
      phi_t   <- rlnorm(1, 0, 1)
      wd <- rho_t * kappa_t
      ws <- (1 - rho_t) * kappa_t

      ev <- evidence_combinations |>
        slice_sample(n = N, replace = TRUE)

      alpha_p <- 0.5 + wd * ev$blue1 + ws * ev$blue2
      beta_p  <- 0.5 + wd * (ev$total1 - ev$blue1) +
                       ws * (ev$total2 - ev$blue2)
      theta_hat <- alpha_p / (alpha_p + beta_p)
      p_blue    <- plogis(phi_t * qlogis(theta_hat))
      choices   <- rbinom(N, 1, p_blue)

      list(
        variables = list(rho = rho_t, kappa = kappa_t, phi = phi_t),
        generated = list(
          N      = N,
          choice = choices,
          blue1  = ev$blue1,
          total1 = ev$total1,
          blue2  = ev$blue2,
          total2 = ev$total2
        )
      )
    },
    N = 180  # same order as the main chapter's single-agent datasets
  )

  sbc_ds_softmax <- generate_datasets(sbc_gen_softmax, n_sbc_softmax)

  sbc_be_softmax <- SBC::SBC_backend_cmdstan_sample(
    mod_softmax,
    chains          = 2,
    iter_warmup     = 500,
    iter_sampling   = 500,
    parallel_chains = 2,
    refresh         = 0,
    show_messages   = FALSE
  )

  sbc_res_softmax <- compute_SBC(sbc_ds_softmax, sbc_be_softmax,
                                  keep_fits = FALSE)
  saveRDS(sbc_res_softmax, "simmodels/ch9_sbc_softmax.rds")

} else {
  sbc_res_softmax <- readRDS("simmodels/ch9_sbc_softmax.rds")
}

SBC::plot_ecdf_diff(
  sbc_res_softmax, variables = c("rho", "kappa", "phi")
) +
  labs(
    title    = "SBC Calibration: WBA-Softmax",
    subtitle = paste0(
      n_sbc_softmax, " datasets. Compare kappa calibration bands to the ",
      "BetaBinomial WBA — wider bands here indicate the kappa/phi confound."
    )
  )
```

**What to look for:** If $\kappa$ shows wider calibration bands than in
the BetaBinomial WBA's SBC, that is the $\kappa$/$\phi$ confound at
work. Increasing $\kappa$ concentrates the posterior mean toward the
extremes; increasing $\phi$ amplifies that same posterior mean. The two
operations have similar effects on the choice probability, making them
hard to disentangle from binary data alone. This is not a model failure —
it is an inherent limitation of the softmax parameterisation that the
BetaBinomial avoids by propagating full distributional uncertainty.

### Fitting and LOO Comparison

```{r ex1_fit_softmax, message=FALSE}
# Reuse the stan_data_self_focused from Chapter 9
fit_softmax_self <- if (regenerate_simulations) {
  fit <- mod_softmax$sample(
    data            = stan_data_self_focused,
    seed            = 701,
    chains          = 4,
    parallel_chains = 4,
    iter_warmup     = 1000,
    iter_sampling   = 1000,
    refresh         = 0
  )
  fit$save_object("simmodels/ch9_fit_softmax_self.rds")
  fit
} else {
  readRDS("simmodels/ch9_fit_softmax_self.rds")
}

# Diagnostics
fit_softmax_self$cmdstan_diagnose()

fit_softmax_self$summary(variables = c("rho", "kappa", "phi")) |>
  knitr::kable(digits = 3,
               caption = "WBA-Softmax posteriors: Self-Focused agent data")
```

```{r ex1_loo_comparison}
loo_softmax_self <- loo(fit_softmax_self$draws("log_lik", format = "matrix"))

# The BetaBinomial WBA's LOO was computed in Chapter 9 as loo_weighted_self.
# If running this supplement standalone, recompute it:
# loo_weighted_self <- loo(fit_weighted_self_focused$draws("log_lik",
#                          format = "matrix"))

cat("=== Observation Model Comparison: BetaBinomial vs. Softmax ===\n")
cat("(Both use the WBA belief-update with rho/kappa parameterisation)\n\n")

obs_model_compare <- loo_compare(
  list("WBA-BetaBin" = loo_weighted_self,
       "WBA-Softmax" = loo_softmax_self)
)
print(obs_model_compare)

diff_ratio <- abs(obs_model_compare[2, "elpd_diff"] /
                  obs_model_compare[2, "se_diff"])
cat(sprintf("\n|ELPD diff| / SE = %.2f\n", diff_ratio))

obs_weights <- loo_model_weights(
  list("WBA-BetaBin" = loo_weighted_self,
       "WBA-Softmax" = loo_softmax_self),
  method = "stacking"
)
cat("\nBayesian stacking weights:\n")
print(obs_weights)
```

**Interpretation guidance:**

- If the **BetaBinomial wins** (expected, since the simulated agents used
  the BetaBinomial generative process): the probability-matching
  hypothesis is supported by the data that were generated under it. The
  softmax's extra parameter $\phi$ is not justified.

- If the **Softmax wins** on simulated BetaBinomial data: this would be
  surprising and would suggest a numerical issue or SBC calibration
  failure worth investigating.

- On **real human data**, the comparison has genuine empirical content.
  If participants are more deterministic than probability matching
  predicts, the softmax will improve LOO. The LOO-PIT from the
  BetaBinomial model would show a characteristic hump shape (model
  overestimates response variability), confirming the diagnosis.

```{r ex1_phi_interpretation}
# Visualise the posterior for phi — how deterministic is the agent?
draws_sm <- as_draws_df(fit_softmax_self$draws("phi"))

ggplot(draws_sm, aes(x = phi)) +
  geom_histogram(bins = 50, fill = "#756bb1", alpha = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 0.3, y = Inf, label = "Near-random\n(probability matching)",
           vjust = 1.5, hjust = 0, size = 3, colour = "grey40") +
  annotate("text", x = 3.5, y = Inf, label = "Near-deterministic\n(maximising)",
           vjust = 1.5, hjust = 0, size = 3, colour = "grey40") +
  labs(
    title    = "Posterior: Inverse Temperature (phi)",
    subtitle = "phi = 1 approximates logistic probability matching.\nphi > 1 = more deterministic; phi < 1 = noisier.",
    x        = expression(phi),
    y        = "Posterior count"
  )
```

---

## Exercise 2: SBC for the Multilevel WBA and Sequential Model

### Motivation

The main chapter validates the single-level WBA and PBA via SBC but
skips the multilevel WBA and the sequential model. The cheatsheet
mandates that "empirical fitting must never occur before simulation
validation." The multilevel WBA introduces correlated NCP random effects
(Cholesky-factored), which can develop funnel geometry that single-level
SBC cannot detect. The sequential model has a monotonically accumulating
state variable that could produce gradient pathologies at long time
horizons.

### 2a: SBC for the Multilevel WBA

The generator must simulate a full multi-agent dataset: draw population
parameters from the prior, draw individual parameters from the
population distribution, and generate choices for each agent.

```{r ex2a_sbc_ml_wba, message=FALSE}
n_sbc_ml <- 200  # multilevel SBC is expensive; 200 is a reasonable lower bound

if (regenerate_simulations) {

  J_sbc <- 20          # agents per SBC universe
  n_per_comb_sbc <- 2  # repetitions per evidence combination per agent

  sbc_gen_ml <- SBC_generator_function(
    function(N_unused) {
      # Draw population parameters from priors
      mu_log    <- rnorm(2, 0, 1)
      sigma_log <- rexp(2, 2)

      # Draw LKJ(2) correlation via the Cholesky factor
      # For 2x2, the off-diagonal element of L is Beta(2, 2)-distributed
      # after appropriate transformation
      l21       <- 2 * rbeta(1, 2, 2) - 1  # marginal on rho under LKJ(2)
      L_Omega   <- matrix(c(1, l21, 0, sqrt(1 - l21^2)), 2, 2)
      Sigma_chol <- diag(sigma_log) %*% L_Omega

      # Draw individual z-scores and reconstruct
      z <- matrix(rnorm(2 * J_sbc), nrow = 2)
      theta <- Sigma_chol %*% z
      log_wd <- mu_log[1] + theta[1, ]
      log_ws <- mu_log[2] + theta[2, ]
      wd_ind <- exp(log_wd)
      ws_ind <- exp(log_ws)

      # Simulate choices
      ev_grid <- expand_grid(blue1 = 0:8, blue2 = 0:3,
                              total1 = 8, total2 = 3)
      sim_rows <- list()

      for (j in seq_len(J_sbc)) {
        ev_j <- ev_grid |> slice_sample(n = n_per_comb_sbc * nrow(ev_grid),
                                         replace = TRUE)
        alpha_p <- 0.5 + wd_ind[j] * ev_j$blue1 + ws_ind[j] * ev_j$blue2
        beta_p  <- 0.5 + wd_ind[j] * (ev_j$total1 - ev_j$blue1) +
                         ws_ind[j] * (ev_j$total2 - ev_j$blue2)
        theta_draws <- rbeta(nrow(ev_j), alpha_p, beta_p)
        choices     <- rbinom(nrow(ev_j), 1, theta_draws)

        sim_rows[[j]] <- tibble(
          agent_id = j,
          choice   = choices,
          blue1    = ev_j$blue1,
          total1   = ev_j$total1,
          blue2    = ev_j$blue2,
          total2   = ev_j$total2
        )
      }

      sim_df <- bind_rows(sim_rows)

      # SBC tracks population-level parameters
      rho_pop <- l21  # the correlation

      list(
        variables = list(
          `mu_log[1]`    = mu_log[1],
          `mu_log[2]`    = mu_log[2],
          `sigma_log[1]` = sigma_log[1],
          `sigma_log[2]` = sigma_log[2]
          # rho is derived; tracking mu_log and sigma_log is sufficient
        ),
        generated = list(
          N        = nrow(sim_df),
          J        = J_sbc,
          agent_id = sim_df$agent_id,
          choice   = sim_df$choice,
          blue1    = sim_df$blue1,
          total1   = sim_df$total1,
          blue2    = sim_df$blue2,
          total2   = sim_df$total2
        )
      )
    },
    N_unused = 0
  )

  sbc_ds_ml <- generate_datasets(sbc_gen_ml, n_sbc_ml)

  sbc_be_ml <- SBC::SBC_backend_cmdstan_sample(
    mod_ml_wba,  # compiled in Chapter 9
    chains          = 2,
    iter_warmup     = 500,
    iter_sampling   = 500,
    parallel_chains = 2,
    adapt_delta     = 0.9,
    refresh         = 0,
    show_messages   = FALSE
  )

  sbc_res_ml <- compute_SBC(sbc_ds_ml, sbc_be_ml, keep_fits = FALSE)
  saveRDS(sbc_res_ml, "simmodels/ch9_sbc_ml_wba.rds")

} else {
  sbc_res_ml <- readRDS("simmodels/ch9_sbc_ml_wba.rds")
}

# ── ECDF-difference plots: population parameters ────────────────────
SBC::plot_ecdf_diff(
  sbc_res_ml,
  variables = c("mu_log[1]", "mu_log[2]", "sigma_log[1]", "sigma_log[2]")
) +
  labs(
    title    = "SBC Calibration: Multilevel WBA (population parameters)",
    subtitle = paste0(n_sbc_ml, " datasets, ", J_sbc, " agents each.",
                      "\nsigma_log parameters are most prone to funnel issues.")
  )
```

```{r ex2a_divergence_profile}
# Divergence profile: does funnel geometry concentrate at small sigma?
if (!is.null(sbc_res_ml$backend_diagnostics)) {
  diags_ml <- sbc_res_ml$backend_diagnostics
  fail_rate <- mean(diags_ml$n_divergent > 0) * 100

  cat(sprintf("%.0f%% of SBC universes produced at least one divergence.\n",
              fail_rate))

  # If the SBC object stores generated variables, plot sigma vs divergence
  diags_ml |>
    ggplot(aes(x = n_divergent)) +
    geom_histogram(binwidth = 1, fill = "darkred", alpha = 0.8) +
    labs(
      title    = "ML-WBA Divergence Counts Across SBC Universes",
      subtitle = sprintf("%.0f%% of universes had >=1 divergence.", fail_rate),
      x        = "Number of Divergent Transitions",
      y        = "Number of SBC Universes"
    )
}
```

**What to look for:**

- If $\sigma_\text{log}$ parameters show systematic ECDF-difference
  deviations (especially U-shaped or skewed) while $\mu_\text{log}$
  parameters are well calibrated, this indicates Neal's Funnel: the NCP
  is failing to fully resolve the small-$\sigma$ geometry. The
  diagnostic response is to check whether divergences concentrate at
  small $\sigma$ values (plot sigma against divergence count).

- If both $\mu$ and $\sigma$ are well calibrated but divergence rates
  are non-trivial (say > 5%), this indicates mild geometry issues that
  do not bias the marginal posteriors but slow mixing. Raising
  `adapt_delta` to 0.95 is the appropriate treatment in this case (after
  confirming the NCP is correct).

### 2b: SBC for the Sequential Model

The sequential model has three parameters (`weight_direct`,
`weight_social`, `learning_rate`) and a monotonically accumulating
internal state. SBC checks whether the sampler can navigate this state
space without systematic bias.

```{r ex2b_sbc_sequential, message=FALSE}
n_sbc_seq <- 200

if (regenerate_simulations) {

  T_sbc <- 60  # trials per SBC dataset (same as chapter example)

  sbc_gen_seq <- SBC_generator_function(
    function(T_unused) {
      wd_t <- rlnorm(1, 0, 0.5)
      ws_t <- rlnorm(1, 0, 0.5)
      lr_t <- rlnorm(1, -1, 0.5)

      # Simulate sequential data
      alpha_t <- 1.0
      beta_t  <- 1.0
      p_blue_env <- 0.75  # stable environment

      blue1_vec  <- integer(T_sbc)
      total1_vec <- rep(8L, T_sbc)
      blue2_vec  <- integer(T_sbc)
      total2_vec <- rep(3L, T_sbc)
      choice_vec <- integer(T_sbc)

      for (t in seq_len(T_sbc)) {
        belief_t <- alpha_t / (alpha_t + beta_t)
        blue1_vec[t]  <- rbinom(1, 8, p_blue_env)
        blue2_vec[t]  <- rbinom(1, 3, p_blue_env)
        choice_vec[t] <- rbinom(1, 1, belief_t)

        if (t < T_sbc) {
          alpha_t <- alpha_t +
            lr_t * (wd_t * blue1_vec[t] + ws_t * blue2_vec[t])
          beta_t  <- beta_t +
            lr_t * (wd_t * (8 - blue1_vec[t]) + ws_t * (3 - blue2_vec[t]))
        }
      }

      list(
        variables = list(
          weight_direct = wd_t,
          weight_social = ws_t,
          learning_rate = lr_t
        ),
        generated = list(
          T      = T_sbc,
          choice = choice_vec,
          blue1  = blue1_vec,
          total1 = total1_vec,
          blue2  = blue2_vec,
          total2 = total2_vec
        )
      )
    },
    T_unused = 0
  )

  sbc_ds_seq <- generate_datasets(sbc_gen_seq, n_sbc_seq)

  sbc_be_seq <- SBC::SBC_backend_cmdstan_sample(
    mod_seq,  # compiled in Chapter 9
    chains          = 2,
    iter_warmup     = 500,
    iter_sampling   = 500,
    parallel_chains = 2,
    refresh         = 0,
    show_messages   = FALSE
  )

  sbc_res_seq <- compute_SBC(sbc_ds_seq, sbc_be_seq, keep_fits = FALSE)
  saveRDS(sbc_res_seq, "simmodels/ch9_sbc_sequential.rds")

} else {
  sbc_res_seq <- readRDS("simmodels/ch9_sbc_sequential.rds")
}

SBC::plot_ecdf_diff(
  sbc_res_seq,
  variables = c("weight_direct", "weight_social", "learning_rate")
) +
  labs(
    title    = "SBC Calibration: Sequential Agent",
    subtitle = paste0(
      n_sbc_seq, " datasets, ", T_sbc, " trials each.",
      "\nlearning_rate is hardest to identify — expect wider bands."
    )
  )
```

**What to look for:**

- `learning_rate` is the most vulnerable parameter. Because it scales
  the *pace* of accumulation rather than the *direction*, its effect on
  early trials is negligible (the denominator $\alpha_t + \beta_t$
  is small, so the mean $\alpha_t / (\alpha_t + \beta_t)$ is sensitive
  to each addition), but becomes visible only as evidence accumulates
  over many trials. With $T = 60$, the SBC bands for `learning_rate`
  will be wider than for the weight parameters.

- If the ECDF-difference for `learning_rate` shows a systematic hump
  (underconfident posteriors), it means the prior is too wide relative
  to what the data can identify. Tightening the prior from
  `lognormal(-1, 0.5)` to `lognormal(-1, 0.3)` and re-running SBC
  would test this hypothesis.

---

## Exercise 3: Bidirectional Model Recovery — Circular Inference vs. WBA

### Motivation

Chapter 9 fits both the WBA and the Circular Inference (CI) model to
data generated by the WBA, confirming that the LOO comparison does not
spuriously prefer the more complex model. But this is only one direction
of the test. A complete model-recovery validation (following Ch. 7's
established practice) requires the reverse: generate data from the CI
model and verify that the CI model wins.

Without this check, we cannot distinguish two scenarios: (a) the LOO
comparison has genuine discriminative power between WBA and CI, or (b)
the LOO comparison always favours the model with fewer parameters,
regardless of the generating process. Only the bidirectional test
resolves this.

### Generating Data from the Circular Inference Model

```{r ex3_ci_data_generation}
set.seed(42)

# Circular inference agent: moderate overcounting of self-evidence
ci_true_params <- list(
  w_self       = 1.0,
  w_other      = 0.8,
  alpha_self   = 2.5,   # overcounts direct evidence by factor 2.5
  alpha_other  = 1.0    # no overcounting of social evidence
)

alpha0 <- beta0 <- 0.5
lo_prior_val <- log(alpha0 / beta0)  # = 0 for Jeffreys

# Generate choices across the full evidence grid
ci_sim_data <- evidence_combinations |>
  slice(rep(seq_len(n()), each = 5)) |>  # 5 repetitions per combination
  mutate(
    lo_self   = log((alpha0 + blue1) / (beta0 + total1 - blue1)),
    lo_social = log((alpha0 + blue2) / (beta0 + total2 - blue2)),
    lo_combined = ci_true_params$w_self  * ci_true_params$alpha_self *
                    (lo_self - lo_prior_val) +
                  ci_true_params$w_other * ci_true_params$alpha_other *
                    (lo_social - lo_prior_val) +
                  lo_prior_val,
    p_blue = plogis(lo_combined),
    choice = rbinom(n(), 1, p_blue)
  )

stan_data_ci_gen <- list(
  N      = nrow(ci_sim_data),
  choice = ci_sim_data$choice,
  blue1  = ci_sim_data$blue1,
  total1 = ci_sim_data$total1,
  blue2  = ci_sim_data$blue2,
  total2 = ci_sim_data$total2
)

# Visualise the generated response curve
ci_sim_data |>
  mutate(social_evidence = factor(
    blue2, levels = 0:3,
    labels = c("Clear Red", "Maybe Red", "Maybe Blue", "Clear Blue")
  )) |>
  group_by(blue1, social_evidence) |>
  summarise(p_blue_obs = mean(choice), .groups = "drop") |>
  ggplot(aes(x = blue1, y = p_blue_obs,
             colour = social_evidence, group = social_evidence)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  scale_colour_brewer(palette = "Set1") +
  scale_x_continuous(breaks = 0:8) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title    = "Simulated Choices from Circular Inference Agent",
    subtitle = paste0("True params: w_self=", ci_true_params$w_self,
                      ", w_other=", ci_true_params$w_other,
                      ", alpha_self=", ci_true_params$alpha_self,
                      ", alpha_other=", ci_true_params$alpha_other),
    x        = "Blue marbles in direct sample",
    y        = "P(choose blue)",
    colour   = "Social evidence"
  ) +
  theme(legend.position = "bottom")
```

Note the steeper response curves compared to WBA-generated data. This is
the qualitative signature of circular inference: overcounting amplifies
small differences on the log-odds scale, producing sharper transitions
from "definitely red" to "definitely blue" as direct evidence increases.

### Fitting Both Models to CI-Generated Data

```{r ex3_fit_both, message=FALSE}
# Fit the WBA (BetaBinomial, rho/kappa) to CI-generated data
fit_wba_on_ci <- if (regenerate_simulations) {
  fit <- mod_weighted$sample(
    data            = stan_data_ci_gen,
    seed            = 801,
    chains          = 4,
    parallel_chains = 4,
    iter_warmup     = 1000,
    iter_sampling   = 1000,
    refresh         = 0
  )
  fit$save_object("simmodels/ch9_fit_wba_on_ci_data.rds")
  fit
} else {
  readRDS("simmodels/ch9_fit_wba_on_ci_data.rds")
}

# Fit the CI model (Bernoulli, log-odds) to CI-generated data
fit_ci_on_ci <- if (regenerate_simulations) {
  fit <- mod_circular$sample(
    data            = stan_data_ci_gen,
    seed            = 802,
    chains          = 4,
    parallel_chains = 4,
    iter_warmup     = 1000,
    iter_sampling   = 1000,
    refresh         = 0
  )
  fit$save_object("simmodels/ch9_fit_ci_on_ci_data.rds")
  fit
} else {
  readRDS("simmodels/ch9_fit_ci_on_ci_data.rds")
}

# Diagnostics
check_diagnostics(fit_wba_on_ci, "WBA on CI-generated data")
check_diagnostics(fit_ci_on_ci,  "CI on CI-generated data")
```

### LOO Comparison on CI-Generated Data

```{r ex3_loo_comparison}
loo_wba_on_ci <- loo(fit_wba_on_ci$draws("log_lik", format = "matrix"))
loo_ci_on_ci  <- loo(fit_ci_on_ci$draws("log_lik", format = "matrix"))

cat("=== Bidirectional Test: CI-Generated Data ===\n")
cat("If the CI model wins here AND the WBA wins on WBA-generated data,\n")
cat("the comparison pipeline has genuine discriminative power.\n\n")

bidir_compare <- loo_compare(
  list(WBA = loo_wba_on_ci, CI = loo_ci_on_ci)
)
print(bidir_compare)

diff_ratio <- abs(bidir_compare[2, "elpd_diff"] /
                  bidir_compare[2, "se_diff"])
cat(sprintf("\n|ELPD diff| / SE = %.2f\n", diff_ratio))

bidir_weights <- loo_model_weights(
  list(WBA = loo_wba_on_ci, CI = loo_ci_on_ci),
  method = "stacking"
)
cat("\nBayesian stacking weights:\n")
print(bidir_weights)
```

### Parameter Recovery: Does the CI Model Recover the True Loop Parameters?

```{r ex3_ci_recovery}
ci_true_df <- tibble(
  parameter  = c("w_self", "w_other", "alpha_self", "alpha_other"),
  true_value = c(1.0, 0.8, 2.5, 1.0)
)

draws_ci <- as_draws_df(
  fit_ci_on_ci$draws(c("w_self", "w_other", "alpha_self", "alpha_other"))
) |>
  pivot_longer(everything(), names_to = "parameter", values_to = "value") |>
  filter(!parameter %in% c(".chain", ".iteration", ".draw"))

ggplot(draws_ci, aes(x = value)) +
  geom_histogram(bins = 50, fill = "#756bb1", alpha = 0.7) +
  geom_vline(data = ci_true_df, aes(xintercept = true_value),
             colour = "#d73027", linewidth = 1.2, linetype = "dashed") +
  facet_wrap(~ parameter, scales = "free") +
  labs(
    title    = "Parameter Recovery: CI Model on CI-Generated Data",
    subtitle = "Red dashed = true generating values. alpha_self = 2.5 is the key test.",
    x        = "Parameter value",
    y        = "Posterior count"
  )
```

### Summary Table: Bidirectional Model Recovery

```{r ex3_summary_table}
# Combine results from Chapter 9 (WBA-generated data) and this exercise
# Chapter 9 result: WBA should win on WBA-generated data
# This exercise result: CI should win on CI-generated data

recovery_summary <- tibble(
  `Generating model` = c("WBA (w_d=1.5, w_s=0.5)",
                          "CI (w_self=1, alpha_self=2.5)"),
  `Data source`      = c("Chapter 9, Self-Focused agent",
                          "This exercise"),
  `Expected winner`  = c("WBA", "CI"),
  `ELPD diff / SE`   = c(
    "See Ch. 9 output",
    sprintf("%.2f", diff_ratio)
  ),
  `Stacking: WBA`    = c("See Ch. 9 output",
                          sprintf("%.2f", bidir_weights["WBA"])),
  `Stacking: CI`     = c("See Ch. 9 output",
                          sprintf("%.2f", bidir_weights["CI"]))
)

knitr::kable(recovery_summary,
             caption = "Bidirectional Model Recovery: WBA vs. Circular Inference")
```

**Interpreting the bidirectional test:**

- **Both directions correct** (WBA wins on WBA data, CI wins on CI
  data): The LOO comparison pipeline has discriminative power for these
  two models under the simulated conditions. The 9 × 4 evidence grid
  provides sufficient contrast to separate differential trust (WBA) from
  overcounting (CI).

- **CI fails to win on its own data:** This would indicate that the
  WBA's pseudocount scaling can approximate the CI's overcounting
  signature well enough that LOO cannot distinguish them. The
  implication is that the evidence grid is not sufficiently diagnostic —
  you need trials at intermediate evidence levels (3–5 blue of 8), where
  the nonlinear amplification of the log-odds overcounting is most
  distinct from linear pseudocount scaling.

- **WBA fails to win on its own data:** This would indicate the CI
  model's extra parameters absorb variance even when overcounting is
  absent. Check whether the posterior for `alpha_self` concentrates near
  1.0 (the CI model correctly identifies no overcounting) — if so, the
  ELPD penalty for extra parameters is too small to overcome the
  flexibility advantage. A stronger prior on `alpha_self_m1` (e.g.,
  `half_normal(0, 0.5)` instead of `half_normal(0, 1)`) would regularise
  more aggressively and might restore the expected preference.

---

## Summary: What These Exercises Add to Chapter 9

| Exercise | Validation gap filled | Key diagnostic output |
|---|---|---|
| 1. WBA-Softmax | Tests the observation-model dimension; certifies the BetaBinomial is not taken on faith | SBC for $(\rho, \kappa, \phi)$; LOO comparison of observation models |
| 2a. ML-WBA SBC | Certifies the hierarchical NCP + LKJ sampler before interpreting population parameters | ECDF-diff for $\mu_\text{log}$, $\sigma_\text{log}$; divergence profile |
| 2b. Sequential SBC | Certifies the accumulating-state sampler before interpreting learning rates | ECDF-diff for `learning_rate`; sensitivity to prior width |
| 3. Bidirectional recovery | Proves LOO can tell WBA and CI apart (not just that it avoids false positives) | 2 × 2 recovery table; stacking weights in both directions |

Together with the main chapter, these exercises complete the full
validation battery mandated by the cheatsheet for every model that
touches data.
