---
title: "11 — The Generalized Context Model"
output:
  html_document:
    mathjax: default
    toc: true
    toc_float: true
  pdf_document:
    toc: true
date: "2025-04-28"
---

```{r, W11 load packages, include=FALSE}
# Set to TRUE to run computationally intensive simulations/fits.
# Set to FALSE to load pre-computed results.
regenerate_simulations <- TRUE
run_intensive_checks   <- TRUE   # controls recovery / SBC / LFO-CV chunks

knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(
  tidyverse,    # data manipulation and visualization
  future,       # parallel processing setup
  purrr,        # functional programming tools
  furrr,        # parallel functional programming
  patchwork,    # combining plots
  cmdstanr,     # interface to Stan
  posterior,    # tidy posterior arrays
  tidybayes,    # tidy extraction of draws
  bayesplot,    # Bayesian visualization
  loo,          # LOO-CV model comparison
  SBC,          # Simulation-Based Calibration (ECDF-diff plots)
  priorsense,   # prior sensitivity via power-scaling
  MCMCpack      # rdirichlet
)

theme_set(theme_classic())

# Prior hyperparameters for sensitivity c
LOG_C_PRIOR_MEAN <- log(2)   # prior median c = 2
LOG_C_PRIOR_SD   <- 1.0      # covers c ≈ 0.27–14.8 at ±2 SD

# Create output directories
for (d in c("stan", "simdata", "simmodels")) {
  if (!dir.exists(d)) dir.create(d)
}

# ── Shared diagnostic helper used throughout the chapter ─────────────────────
# This function is the canonical "have we passed the mandatory MCMC diagnostic
# battery" check. It returns a tibble that the chapter prints whenever a fit
# is completed, and chapters 12, 13, 15 source the same helper.
diagnostic_summary_table <- function(fit, params = NULL) {
  if (is.null(fit)) return(NULL)
  diag <- fit$diagnostic_summary(quiet = TRUE)
  
  s <- fit$summary(
    variables = params, 
    "mean", "sd", "rhat", "ess_bulk", "ess_tail", "mcse_mean"
  )

  tibble(
    metric = c(
      "Divergences (zero tolerance)",
      "Max rank-normalised R-hat",
      "Min bulk ESS",
      "Min tail ESS",
      "Min E-BFMI",
      "Max MCSE / posterior SD"
    ),
    value = c(
      sum(diag$num_divergent),
      max(s$rhat,    na.rm = TRUE),
      min(s$ess_bulk, na.rm = TRUE),
      min(s$ess_tail, na.rm = TRUE),
      min(diag$ebfmi, na.rm = TRUE),
      max(s$mcse_mean / s$sd, na.rm = TRUE)
    ),
    threshold = c("== 0", "< 1.01", "> 400", "> 400", "> 0.2", "< 0.05"),
    pass = c(
      sum(diag$num_divergent)               == 0,
      max(s$rhat, na.rm = TRUE)             <  1.01,
      min(s$ess_bulk, na.rm = TRUE)         >  400,
      min(s$ess_tail, na.rm = TRUE)         >  400,
      min(diag$ebfmi, na.rm = TRUE)         >  0.2,
      max(s$mcse_mean / s$sd, na.rm = TRUE) <  0.05
    )
  )
}
```

> **📍 Where we are in the Bayesian modeling workflow:**
> Chapters 1–10 built the complete inference toolkit: parameter estimation
> (Ch. 4), the six-phase validation pipeline (Ch. 5), hierarchical models
> (Ch. 6), model comparison via LOO (Ch. 7), mixture models with discrete
> marginalization (Ch. 8), and hierarchical Bayesian cognition (Ch. 9–10).
> This chapter applies the full toolkit to *categorization* — asking not only
> what parameters describe a participant's behavior, but which *cognitive
> mechanism* best accounts for how people learn to assign objects to kinds.
> We begin with the exemplar approach: the Generalized Context Model (GCM). 
> The following chapters will introduce prototype and rule-based models, 
> and then compare all three approaches head-to-head using the full validation 
> battery.
>
> Two important caveats that previous did not face. 
> First, the categorization models learn, that is, are **path-dependent** —
> the likelihood at trial $t$ depends on the full sequence of stimuli on
> trials $1, \ldots, t-1$ through the exemplar memory store. Standard LOO-CV
> is therefore technically invalid here. Rather than wave at this and hope it
> doesn't matter, we introduce leave-future-out cross-validation (LFO-CV; 
> Bürkner, Gabry, & Vehtari, 2020).
> Second, the rule-based models we will discuss cannot really be fit using
> MCMC (we'll see why), so we need to introduce a new slower method: Approximate
> Bayesian Computation (ABC).

# Introduction to Categorization Models

## The Fundamental Problem of Categorization

Categorization is one of the most fundamental cognitive abilities that humans possess. From early childhood, we learn to organize the world around us into meaningful categories: distinguishing food from non-food, safe from dangerous, or one letter from another. This process of assigning objects to categories is complex, involving perception, memory, attention, and decision-making.

How do humans learn categories and make categorization decisions? This seemingly simple question has generated decades of research and theoretical debate in cognitive science.

## Why Model Categorization?

As we have seen in previous chapters, computational models offer powerful ways to formalize theories about categorization. By implementing these theories as computer algorithms, we can:

* Explore the consequences of different theoretical assumptions

* Make precise predictions about human behavior in categorization tasks

* Test competing theories against empirical data

In this chapter, we will explore three major approaches to modeling categorization:

* **Exemplar models** — which propose that categories are represented by storing individual examples

* **Prototype models** — which suggest categories are represented by an abstract "average" or central tendency

* **Rule-based models** — which posit that categories are represented by explicit rules or decision boundaries

Each approach captures different aspects of human categorization and has generated substantial empirical research. We'll implement computational versions of each model type, allowing us to compare their behavior and predictions.
Note that there are many hybrid models and more complex approaches, but covering these three will give usa solid foundation also for understanding the more complex models.

## Historical Development of Categorization Models

### Early Views: The Classical Approach

Early theories of categorization followed what is now called the classical view, where categories were defined by necessary and sufficient features. In this view, category membership is an all-or-nothing affair: an object either satisfies the logical criteria or it doesn't. While intuitive, this approach struggled to explain many aspects of human categorization, such as:

* Graded category membership (some members seem "more typical" than others)

* Unclear boundaries between categories

* Context-dependent categorization

* Family resemblance structures (where no single feature is necessary)

### The Prototype Revolution

In the 1970s, Eleanor Rosch's pioneering work on prototypes challenged the classical view. She demonstrated that categories appear to be organized around central tendencies or "prototypes," with membership determined by similarity to these prototypes. Objects closer to the prototype are categorized more quickly and consistently.

Prototype models formalize this idea by representing categories as their central tendency in a psychological feature space. New items are classified based on their similarity to these prototypes.

### The Exemplar Alternative

In the late 1970s and 1980s, researchers like Douglas Medin and Robert Nosofsky proposed that rather than abstracting prototypes, people might store individual exemplars of categories and make judgments based on similarity to these stored examples.

The Generalized Context Model (GCM), developed by Nosofsky, became the standard exemplar model. It proposes that categorization decisions are based on the **summed** similarity of a new stimulus to all stored exemplars of each category, weighted by attention to different stimulus dimensions.

### Rule-Based Models

While similarity-based approaches (both prototype and exemplar) gained prominence, other researchers argued that people sometimes use explicit rules for categorization. Rule-based models propose that categorization involves applying decision rules that partition the stimulus space. To overcome at least some of the critiques against a more classical approach, many contemporary rule-based approaches are probabilistic, reflecting the complexity of real-world categories.

Models like Bayesian particle filters, multinomial trees, COVIS (COmpetition between Verbal and Implicit Systems) and RULEX (RULe-plus-EXception) formalize these ideas, suggesting that rule learning and application form a core part of human categorization, particularly for well-defined categories.

### Hybrid Approaches

More recent work has focused on hybrid models that incorporate elements of multiple approaches. Models like SUSTAIN (Supervised and Unsupervised STratified Adaptive Incremental Network) and ATRIUM (Attention to Rules and Instances in a Unified Model) propose that humans can flexibly switch between strategies or that different systems operate in parallel.

## The Three Model Classes: Core Assumptions and Predictions

Let's examine the core assumptions of the three model classes we'll be implementing:

| Feature / Dimension | Exemplar Models (e.g., GCM) | Prototype Models | Rule-Based Models |
| :--- | :--- | :--- | :--- |
| **Core Representation** | Individual instances. | An abstract central tendency ("average"). | Explicit rules (necessary/sufficient conditions). |
| **Level of Abstraction** | **None:** All encountered examples are stored. | **High:** Individual details are integrated into a summary. | **High:** Continuous space partitioned into discrete regions. |
| **Similarity / Distance** | distance to stored exemplars (often exponential decay). | distance to the central prototype (often exponential decay). | Probability that the exemplar is generated by the rule. |
| **Attention Allocation** | continuous attention weights to dimensions. | continuous attention weights to dimensions. | **Selective:**  continuous attention "weights" to rules. |
| **Key Predictions** | High sensitivity to specific training examples; categorization depends on specific memories. | "Typicality effect": The most mathematically average items are categorized most easily. | Fast decisions once rules are learned; selective attention to specific features. |

## Our Implementation Approach

We'll implement three models representing each major theoretical approach:

* **Generalized Context Model (GCM)** — A canonical exemplar model where categorization is based on summed similarity to all stored examples

* **Kalman Filter Prototype Model** — A dynamic prototype model that builds on the GCM infrastructure, but incrementally updates category representations instead of storing new exemplars.  (Ch. 12)

* **Bayesian Particle Filter for Rules** — A rule-based model that maintains a distribution over potential categorization rules (Ch. 13)

Each implementation endeavors to follow a similar structure: core model function as a forward simulator, Stan implementation for parameter estimation, the mandatory MCMC diagnostic battery, prior/posterior predictive checks, parameter recovery, prior sensitivity, and SBC. However, as we will see Bayesian Particle Filters are not so amenable to this style of inference, so we will introduce Approximate Bayesian Computation (ABC) for that model in Ch. 13.

---

# The Generalized Context Model (GCM)

## Mathematical Foundations and Cognitive Principles

The Generalized Context Model (GCM), developed by Robert Nosofsky in the 1980s, represents one of the most influential exemplar-based approaches to categorization. Unlike models that abstract category information into prototypes or rules, the GCM proposes that people store individual exemplars in memory and make categorization decisions based on similarity to these stored examples.

### Core Assumptions of the GCM

* **Memory for Instances**: Every encountered example is stored in memory with its category label.

* **Similarity Computation**: Distannce is calculated using a city-block metric (Manhattan distance, steps in each direction, without allowing diagonal movements). Similarity between a new stimulus and each stored exemplar is an exponentially decreasing function of their city-metric distance, with the decay rate governed by parameter $c$.

* **Selective Attention**: Per-dimension weights $w$ rescale the psychological space — stretching or shrinking each axis — before distance is computed.

* **Bias**: A parameter $\beta$ weights the category-A summed similarity against category-B's in the choice rule, independent of the stimuli.

* **Choice Probability**: The probability of assigning a stimulus to a category is proportional to its summed similarity to exemplars of that category. No parameter is defined here.

The point of the previous paragraph is to define the specific commitments we are making in this chapter, which are: "the Nosofsky 1986 GCM with a Manhattan distance, no forgetting, $\beta$ as both a static bias and the cold-start default, and a simple probabilistic decision rule (no softmax)." However, any of these core assumptions can be challenged and/or implemented in different ways. Just because Nosofsky did it one way doesn't mean we have to. E.g. we could implement the GCM with a geometric distance, with memory decay, with a different cold-start strategy, or with a softmax temperature for decisions. 

### Parameters: Cognitive Commitment, Algebraic Role, Stan Parameterization

The GCM has three primary parameters in this formulation. For each, we make three commitments simultaneously — a cognitive one (what the parameter *means*), an algebraic one (how it enters the equations below), and a Stan-level one (how it is declared so the sampler has a well-behaved geometry to move on). Keeping these three layers aligned from the very start help us creating consistent code for both forward and inferential modeling.

**Attention weights $w$.** Cognitively, $w$ is a simplex of dimension-specific attention weights: simplex means that there is finite attention (1) and higher weight means more attention, and therefore more diagnostic contribution, for that feature. Algebraically, $w$ enters the distance formula as the per-dimension coefficient in the attention-weighted city-block metric, so changing $w$ literally stretches or compresses each axis of the psychological space before similarity is computed. At the Stan level $w$ is declared as `simplex[nfeatures]`, which enforces $w_m \ge 0$ and $\sum_m w_m = 1$ directly in the parameterization and gives the Dirichlet prior its natural support.

**Sensitivity $c$.** Cognitively, $c$ represents perceptual discriminability — the width of the generalization gradient and, in this formulation, the model's decisional precision as well (there is no separate softmax temperature). Higher $c$ = faster similarity decay = less generalization = higher specificity. Algebraically, $c$ is the exponential decay rate in $\eta_{ij} = e^{-c\, d_{ij}}$, so it controls the slope of the similarity curve seen in the visualization below. At the Stan level, we need to ensure that $c$ stays a positive value, so we cannot sample it directly. We sample an unconstrained `real log_c` and set `c = exp(log_c)` in `transformed parameters`. This log-normal reparameterization is constraint-free and geometrically well-behaved — the sampler moves on an unbounded real line and the positivity constraint is preserved through exponentiation.

**Response bias $\beta$.** Cognitively, $\beta$ is a baseline preference for category A independent of stimulus similarity; $\beta = 0.5$ is unbiased, and deviations reflect asymmetric response tendencies or unequal base rates. In this chapter $\beta$ also serves as the cold-start fallback: before *both* categories have been observed, the agent responds at probability $\beta$. Algebraically, $\beta$ weights the category-A summed similarity against category-B's in $P(A\mid i)$. At the Stan level $\beta$ is declared as `real<lower=0, upper=1>` with a Beta prior, bounding the parameter to the open interval on which it is defined.

## Mathematical Formulation

**1. Distance Calculation**

The psychological distance between two stimuli in a multidimensional space is:

$$d_{ij} = \left[ \sum_{m=1}^{M} w_m |x_{im} - x_{jm}|^r \right]^{\frac{1}{r}}$$

Where $x_{im}$ is the value of stimulus $i$ on dimension $m$, $w_m$ is the attention weight for dimension $m$ (with $\sum_m w_m = 1$), and $r$ determines the distance metric. We use the city-block metric ($r = 1$), appropriate for separable stimulus dimensions:

```{r, 11 distance formula}
# City-block distance (r = 1), attention-weighted
distance <- function(vect1, vect2, w) {
  sum(w * abs(vect1 - vect2))
}
```

**2. Similarity Computation**

Similarity is an exponentially decreasing function of distance:

$$\eta_{ij} = e^{-c \cdot d_{ij}}$$

where $c > 0$ is the sensitivity parameter. Higher $c$ means similarity drops more steeply with distance (less generalization); lower $c$ means more gradual decay (more generalization).

```{r, 11 similarity formula}
# Exponential similarity kernel
similarity <- function(dist_val, c) {
  exp(-c * dist_val)
}
```

Let's visualize how similarity changes with distance for different values of $c$:

```{r, 11 visualizing effects of sensitivity}
dd <- expand_grid(
  distance  = seq(0, 6, by = 0.1),
  c         = c(0.3, 0.7, 1.5, 3.0, 6.0, 10.0)
) |>
  mutate(
    sim     = similarity(distance, c),
    c_label = factor(c)
  )

ggplot(dd, aes(x = distance, y = sim, color = c_label, group = c_label)) +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(option = "plasma", name = "Sensitivity (c)") +
  labs(
    title    = "Effect of Sensitivity Parameter (c) on Similarity",
    subtitle = "c is on the natural (original) scale; prior places most mass on c ∈ (0.3, 15)",
    x        = "Psychological Distance",
    y        = "Similarity"
  ) +
  theme(legend.position = "bottom")
```

From this visualization: low $c$ (≈ 0.3) means similarity remains high even for distant stimuli — the model generalizes broadly. High $c$ (≈ 10) means similarity collapses rapidly — the model treats only near-identical stimuli as similar. This visual is the geometric content of the Q3 commitment above: the slope of these curves *is* the model's decisional precision.

**3. Category Response Probability**

The probability of assigning stimulus $i$ to category A is based on summed similarity (Nosofsky 1986):

$$P(A|i) = \frac{\beta \sum_{j \in A} \eta_{ij}}{\beta \sum_{j \in A} \eta_{ij} + (1-\beta) \sum_{k \in B} \eta_{ik}}$$

Where $\sum_{j \in A} \eta_{ij}$ is the summed similarity to all exemplars of category A, $\sum_{k \in B} \eta_{ik}$ is the summed similarity to all exemplars of category B, and $\beta \in (0,1)$ is a response bias for category A. Note that we use the **sum** of similarities throughout — in both the R simulation and the Stan likelihood — consistent with the original GCM formulation. The cold-start fallback applies whenever one of the summed similarities is undefined: if either $\sum_{j \in A}$ or $\sum_{k \in B}$ has no terms yet, the response probability is $\beta$.

### The GCM Agent Implementation (R Simulation)

With distance and similarity functions in place, we implement an agent that observes stimuli, stores them into categories according to feedback, and compares new stimuli to stored exemplars.

```{r, 11 generative GCM model}
# Generative GCM model function.
# Takes the sensitivity parameter on its natural (cognitive) scale.
gcm_simulate <- function(w,            # attention weights (sums to 1)
                         c,            # sensitivity (natural scale, strictly positive)
                         bias,         # response bias towards category 1
                         obs,          # matrix of stimulus features (trials x features)
                         cat_feedback, # true category labels (0 or 1)
                         quiet = TRUE) {
  ntrials  <- nrow(obs)
  nfeatures <- ncol(obs)

  prob_cat1    <- numeric(ntrials)
  sim_response <- numeric(ntrials)

  memory_obs <- matrix(NA_real_, nrow = 0, ncol = nfeatures)
  memory_cat <- numeric(0)

  for (i in seq_len(ntrials)) {
    if (!quiet && i %% 20 == 0) cat("Simulating trial:", i, "\n")

    current_stim <- as.numeric(obs[i, ])
    n_mem        <- nrow(memory_obs)
    has_cat0     <- any(memory_cat == 0)
    has_cat1     <- any(memory_cat == 1)

    if (n_mem == 0 || !has_cat0 || !has_cat1) {
      prob_cat1[i] <- bias
    } else {
      # Compute similarity from current stimulus to every stored exemplar
      sims <- vapply(seq_len(n_mem), function(e) {
        similarity(distance(current_stim, memory_obs[e, ], w), c)
      }, numeric(1))

      # Sum (not mean) similarity per category — matches original GCM formulation
      s1  <- sum(sims[memory_cat == 1])
      s0  <- sum(sims[memory_cat == 0])
      num <- bias * s1
      den <- num + (1 - bias) * s0

      prob_cat1[i] <- if (den > 1e-9) num / den else bias
      prob_cat1[i] <- pmax(1e-9, pmin(1 - 1e-9, prob_cat1[i]))
    }

    sim_response[i] <- rbinom(1, 1, prob_cat1[i])

    # Update memory
    memory_obs <- rbind(memory_obs, current_stim)
    memory_cat <- c(memory_cat, cat_feedback[i])
  }

  tibble(prob_cat1 = prob_cat1, sim_response = sim_response)
}
```

This implementation captures the core cognitive processes proposed by the GCM: storing exemplars, computing similarity, weighting by attention, and making probabilistic choices based on relative summed similarity. Critically, the *exact* same control flow has to appear in the Stan code that we will produce.

---

## Experimental Setup: Kruschke (1993)

To understand how the GCM behaves, we simulate performance on a categorization task from Kruschke (1993): 8 stimuli varying on two continuous dimensions (height and position), assigned to two categories. Participants complete 8 blocks of 8 trials each (64 trials total), receiving feedback after each choice.

```{r, 11 GCM experimental setup}
stimulus_info <- tibble(
  stimulus      = c(5, 3, 7, 1, 8, 2, 6, 4),
  height        = c(1, 1, 2, 2, 3, 3, 4, 4),
  position      = c(2, 3, 1, 4, 1, 4, 2, 3),
  category_true = c(0, 0, 1, 0, 1, 0, 1, 1)
)

ggplot(stimulus_info, aes(x = position, y = height,
                          color = factor(category_true), label = stimulus)) +
  geom_point(size = 5) +
  geom_text(nudge_y = 0.18) +
  scale_color_manual(values = c("0" = "#0072B2", "1" = "#D55E00"), name = "Category") +
  labs(title = "Kruschke (1993) Stimulus Structure", x = "Position", y = "Height") +
  coord_fixed()

n_blocks          <- 8
n_stim_per_block  <- nrow(stimulus_info)
total_trials      <- n_blocks * n_stim_per_block
```

Notice that the categories are not linearly separable: no single threshold on height or position perfectly partitions the stimuli. This is what makes the design useful for testing whether the model can recover both attention weights $w$ and sensitivity $c$.

---

## Per-Subject Schedule Generation

In a real experiment each participant sees the stimuli in an **independently randomized** order. Because the GCM is path-dependent — the memory state at trial $t$ depends on the exact ordering of trials $1, \ldots, t-1$ — using the same ordering for every simulated agent would be a modeling error: it would calibrate the inference engine on a single memory trajectory rather than the distribution of trajectories the model will encounter in real data. We therefore generate a distinct randomized schedule for each subject using a deterministic, reproducible seed.

```{r, 11 per-subject schedule}
# Generate one randomized experiment schedule for one subject.
# Returns a tibble with trial_within_subject, block, stimulus_id, height,
# position, and category_feedback. The trial_within_subject column is the
# critical within-subject temporal index used for sorting data before Stan.
make_subject_schedule <- function(stimulus_info, n_blocks, seed) {
  set.seed(seed)
  n_stim <- nrow(stimulus_info)

  sequence <- unlist(lapply(seq_len(n_blocks), function(b) {
    sample(stimulus_info$stimulus, n_stim, replace = FALSE)
  }))

  tibble(
    trial_within_subject = seq_along(sequence),
    block                = rep(seq_len(n_blocks), each = n_stim),
    stimulus_id          = sequence
  ) |>
    left_join(stimulus_info, by = c("stimulus_id" = "stimulus")) |>
    rename(category_feedback = category_true) |>
    dplyr::select(trial_within_subject, block, stimulus_id,
                  height, position, category_feedback)
}
```

We will be using this very same randomization function throughout the chapter.

---

## Simulating Behavior with the GCM

We simulate agents with different parameter settings. `simulate_gcm_agent` wraps `gcm_simulate` by generating the agent's own randomized schedule from a subject-specific seed:

```{r, 11 GCM response simulations}
simulate_gcm_agent <- function(agent_id, w_true, c_true, bias_true,
                               stimulus_info, n_blocks, subject_seed) {
  schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  obs      <- schedule |> dplyr::select(height, position)
  feedback <- schedule$category_feedback

  sim_results <- gcm_simulate(
    w = w_true, c = c_true, bias = bias_true,
    obs = obs, cat_feedback = feedback, quiet = TRUE
  )

  schedule |>
    mutate(
      agent_id   = agent_id,
      w1_true    = w_true[1],
      w2_true    = w_true[2],
      c_true     = c_true,
      log_c_true = log(c_true),
      bias_true  = bias_true,
      prob_cat1  = sim_results$prob_cat1,
      sim_response = sim_results$sim_response,
      correct    = as.integer(category_feedback == sim_response)
    ) |>
    group_by(agent_id) |>
    mutate(performance = cumsum(correct) / row_number()) |>
    ungroup()
}

# Parameter grid on the natural c scale
# (c = 1 corresponds to moderate generalization;
#  c = 4 to high specificity)
param_grid <- expand_grid(
  w_setting    = list(c(0.5, 0.5), c(0.7, 0.3), c(0.9, 0.1)),
  c_setting    = c(0.5, 1.0, 2.0, 4.0, 5.0, 15.0),
  bias_setting  = 0.5
) |>
  mutate(
    setting_id  = row_number(),
    w_label     = map_chr(w_setting, ~paste(round(.x, 1), collapse = ",")),
    c_label     = paste0("c≈", round(c_setting, 1)),
    label       = paste0("w=(", w_label, "), ", c_label)
  )

n_agents_per_setting <- 5

simulation_plan <- param_grid |>
  uncount(n_agents_per_setting) |>
  mutate(agent_id = row_number())

plan(multisession, workers = max(1, availableCores() - 1))

if (regenerate_simulations) {
  simulated_responses <- future_pmap_dfr(
    list(
      agent_id     = simulation_plan$agent_id,
      w_true       = simulation_plan$w_setting,
      c_true       = simulation_plan$c_setting,
      bias_true    = simulation_plan$bias_setting
    ),
    simulate_gcm_agent,
    stimulus_info = stimulus_info,
    n_blocks      = n_blocks,
    subject_seed  = simulation_plan$agent_id,  # unique seed per agent
    .options = furrr_options(seed = TRUE)
  )
  write_csv(simulated_responses, "simdata/W11_gcm_simulated_responses.csv")
  cat("Simulation complete. Results saved.\n")
} else {
  if (file.exists("simdata/W11_gcm_simulated_responses.csv")) {
    simulated_responses <- read_csv("simdata/W11_gcm_simulated_responses.csv",
                                    show_col_types = FALSE)
    cat("Loaded existing simulation results.\n")
  } else {
    stop("Simulation file not found. Set regenerate_simulations = TRUE.")
  }
}

plan(sequential)
```

Let's visualize how different parameter settings affect learning:

```{r, 11 visualize parameter effects}
simulated_responses_labeled <- simulated_responses |>
  mutate(
    w_label = paste0("w=(", round(w1_true, 1), ",", round(w2_true, 1), ")"),
    c_label = paste0("c≈", round(c_true, 1))
  )

# Effect of sensitivity (c) on learning curves
p_c_effect <- ggplot(
  simulated_responses_labeled,
  aes(x = trial_within_subject, y = performance,
      color = factor(round(c_true, 1)),
      group = interaction(agent_id, round(c_true, 1)))
) +
  stat_summary(fun = mean, geom = "line",
               aes(group = factor(round(c_true, 1))), linewidth = 1.2) +
  facet_wrap(~w_label) +
  scale_color_viridis_d(option = "plasma", name = "Sensitivity (c)") +
  labs(
    title    = "Effect of Sensitivity (c) on GCM Learning",
    subtitle = "Mean across simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme(legend.position = "bottom")

# Effect of attention weights on learning curves
p_w_effect <- ggplot(
  simulated_responses_labeled,
  aes(x = trial_within_subject, y = performance,
      color = w_label, group = interaction(agent_id, w_label))
) +
  stat_summary(fun = mean, geom = "line", aes(group = w_label), linewidth = 1.2) +
  facet_wrap(~c_label) +
  scale_color_brewer(palette = "Set1", name = "Attention Weights") +
  labs(
    title    = "Effect of Attention Weights (w) on GCM Learning",
    subtitle = "Mean across simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme(legend.position = "bottom")

p_c_effect
p_w_effect
```

From these plots we observe:

* **Attention weights matter**: For this category structure, equal weights `w=(0.5,0.5)` generally lead to the best performance, especially at higher sensitivity values — both dimensions are informative.

* **Sensitivity affects the learning trajectory**: Too low a $c$ value produces overgeneralization and slow learning. The literature reports that too high a value produces rapid forgetting-by-distance that could hurt accuracy when weights are skewed, but we don't see that here.

* **Parameters interact**: The effect of sensitivity $c$ depends on attention weights. Unless we have very unequal weights, very high $c$ is indistinguishable from lower values. *This interaction is exactly the source of the c–w identifiability concern*: if performance is similar for several different $(c, w)$ combinations, the data will not pin them down separately.

**Demonstrating path-dependence:** Because each agent receives a unique stimulus order, agents with identical parameters can follow divergent learning curves in early trials — the cold-start trajectory and the specific sequence of exemplars stored in memory differs. This is exactly the diversity of trajectories that real data presents, and it is why per-subject randomization in the simulation matters for realistic model validation.

---

## Prior Predictive Check

Before fitting a single data point, we ask: do the priors generate only plausible categorization curves? If the prior predictive distribution includes trajectories with performance well below chance throughout the entire experiment, the prior is implying behaviour that no cognitive model would predict. We pick performance as a visualization, because it's easy to think through, but other targets would be possible.

We need to define the priors for Stan model already here:

- $w \sim \text{Dirichlet}(1, 1)$ — uniform over the simplex: all dimensions are a priori equally important.
- $\log c \sim \text{Normal}(\log(2), 1)$ — implies $c \in (0.27, 14.8)$ at $\pm 2$ SD
- $\text{bias} \sim \text{Beta}(1, 1)$ — uniform over $(0, 1)$

We draw $S = 500$ parameter vectors from those priors and generate the simulations.

```{r, 11 prior predictive check}
n_prior_samples <- 500

prior_draws <- tibble(
  sample_id = seq_len(n_prior_samples),
  log_c     = rnorm(n_prior_samples, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD),
  w1        = runif(n_prior_samples),   # Dirichlet(1,1) marginal = Uniform(0,1)
  bias      = rbeta(n_prior_samples, 1, 1)
) |>
  mutate(c_true = exp(log_c), w2 = 1 - w1)

# Use a single fixed schedule for the prior predictive illustration
# (we are checking the parameter space, not order sensitivity)
ppc_schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = 999)

prior_pred_curves <- pmap_dfr(
  list(
    sample_id = prior_draws$sample_id,
    log_c     = prior_draws$log_c,
    w1        = prior_draws$w1,
    bias      = prior_draws$bias
  ),
  function(sample_id, log_c, w1, bias) {
    c <- exp(log_c)
    sim <- gcm_simulate(
      w = c(w1, 1 - w1), c = c, bias = bias,
      obs         = ppc_schedule |> dplyr::select(height, position),
      cat_feedback = ppc_schedule$category_feedback
    )
    tibble(
      sample_id   = sample_id,
      trial       = seq_len(nrow(ppc_schedule)),
      correct     = as.integer(sim$sim_response == ppc_schedule$category_feedback)
    ) |>
      mutate(cum_acc = cumsum(correct) / row_number())
  }
)

prior_pred_summary <- prior_pred_curves |>
  group_by(trial) |>
  summarise(
    q05 = quantile(cum_acc, 0.05),
    q25 = quantile(cum_acc, 0.25),
    q50 = median(cum_acc),
    q75 = quantile(cum_acc, 0.75),
    q95 = quantile(cum_acc, 0.95),
    .groups = "drop"
  )

ggplot(prior_pred_summary, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#0072B2", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Prior Predictive Check: GCM Learning Curves",
    subtitle = "Ribbons: 50% and 90% prior predictive intervals; line: median\nPriors: log(c) ~ Normal(log 2, 1),  w ~ Dirichlet(1,1),  bias ~ Beta(1,1)",
    x = "Trial", y = "Cumulative Accuracy"
  )
```

The prior predictive envelope covers the full plausible range: chance performance (≈ 0.5) for uninformative parameter draws, and accuracy approaching 1.0 for high-sensitivity agents with well-directed attention. No draws push accuracy below chance across the experiment — the prior does not imply absurd behavior. 

---

## Implementing the GCM in Stan for Parameter Estimation

While simulating the GCM helps understand its behavior, we ultimately want to estimate its parameters ($w$, $c$, $\beta$) from observed categorization data. We implement the GCM in Stan for Bayesian parameter estimation.

### Architectural choice: `prob_cat1[i]` in `transformed parameters`

A naive Stan implementation of a sequential cognitive model places the entire trial-by-trial loop inside the `model{}` block as a local scope, accumulates `target +=` along the way, and then *duplicates the entire loop* in `generated quantities{}` to compute `log_lik[i]`. This works, but it doubles the per-iteration cost (the loop runs twice on every saved draw) and it creates two copies of the same code that must be kept in sync — and the second copy is the one that quietly rots when the first is updated.

There is a better pattern. Because the GCM's per-trial choice probability $p_i$ is a deterministic function of the parameters $(w, c, \beta)$ and the data sequence, it belongs in `transformed parameters` as a single source of truth. The `model{}` block then becomes a one-line vectorised Bernoulli likelihood, and `generated quantities{}` becomes a lookup against the already-computed `prob_cat1` vector. Anything that depends only on observed data is evaluated once in `transformed data`; anything that depends on parameters is cached once in `transformed parameters`; the `model{}` and `generated quantities{}` blocks consume those caches. This is the same pattern the Kalman prototype model in Ch. 12 will use, and it should be the default for any cognitive model whose per-trial likelihood is deterministic given the parameters.

```{r, 11 gcm stan model}
gcm_single_stan <- "
// Generalized Context Model — Single Subject (refactored architecture)
//
// Key design:
//   prob_cat1[i] is the per-trial choice probability. It is deterministic
//   given (w, log_c, bias) and the data sequence, so it lives in
//   transformed parameters. model{} reads it once via bernoulli_lpmf.
//   generated quantities{} reads the same vector for log_lik[i].
//
// Parameterization:
//   w      : attention weights (simplex)
//   log_c  : log sensitivity (unconstrained); c = exp(log_c)
//   bias   : response bias towards category 1 (0–1)

data {
  int<lower=1> ntrials;
  int<lower=1> nfeatures;
  array[ntrials] int<lower=0, upper=1> y;
  array[ntrials, nfeatures] real obs;
  array[ntrials] int<lower=0, upper=1> cat_feedback;

  // Prior hyperparameters (passed as data for flexibility)
  vector[nfeatures] w_prior_alpha;
  real log_c_prior_mean;
  real<lower=0> log_c_prior_sd;
  real<lower=0> bias_prior_alpha;
  real<lower=0> bias_prior_beta;
}

transformed data {
  // Precompute absolute differences between all trials for all features
  // abs_diff[i, j, f] holds the absolute difference between trial i and trial j on feature f
  array[ntrials, ntrials, nfeatures] real abs_diff;
  
  for (i in 1:ntrials) {
    for (j in 1:ntrials) {
      for (f in 1:nfeatures) {
        abs_diff[i, j, f] = abs(obs[i, f] - obs[j, f]);
      }
    }
  }
}

parameters {
  simplex[nfeatures] w;
  real log_c;
  real<lower=0, upper=1> bias;
}

transformed parameters {
  real<lower=0> c = exp(log_c);

  // Per-trial choice probability — computed once per leapfrog step.
  vector<lower=1e-9, upper=1-1e-9>[ntrials] prob_cat1;

  {
    array[ntrials] int memory_trial_idx;
    array[ntrials] int memory_cat;
    int n_mem = 0;

    for (i in 1:ntrials) {
      real p_i;
      int has_cat0 = 0;
      int has_cat1 = 0;

      for (k in 1:n_mem) {
        if (memory_cat[k] == 0) has_cat0 = 1;
        if (memory_cat[k] == 1) has_cat1 = 1;
      }

      if (n_mem == 0 || has_cat0 == 0 || has_cat1 == 0) {
        // Cold-start branch — answers Q1 of Phase 0.
        p_i = bias;
      } else {
        vector[n_mem] sims;
        for (e in 1:n_mem) {
          real d = 0;
          int past_i = memory_trial_idx[e];
          for (f in 1:nfeatures)
            d += w[f] * abs_diff[i, past_i, f];
          sims[e] = exp(-c * d);
        }
        real s1 = 0;
        real s0 = 0;
        for (e in 1:n_mem) {
          if (memory_cat[e] == 1) s1 += sims[e];
          else                    s0 += sims[e];
        }
        real num = bias * s1;
        real den = num + (1 - bias) * s0;
        p_i = (den > 1e-9) ? num / den : bias;
      }

      prob_cat1[i] = fmax(1e-9, fmin(1 - 1e-9, p_i));

      // Memory update
      n_mem += 1;
      memory_trial_idx[n_mem] = i;
      memory_cat[n_mem] = cat_feedback[i];
    }
  }
}

model {
  // Priors
  target += dirichlet_lpdf(w    | w_prior_alpha);
  target += normal_lpdf(log_c   | log_c_prior_mean, log_c_prior_sd);
  target += beta_lpdf(bias      | bias_prior_alpha, bias_prior_beta);

  // Vectorised Bernoulli likelihood — one line, no duplicated loop.
  target += bernoulli_lpmf(y | prob_cat1);
}

generated quantities {
  // log_lik reads from prob_cat1 — the loop is not re-run.
  // This is the architectural payoff of the transformed-parameters layout.
  vector[ntrials] log_lik;
  real lprior;
  for (i in 1:ntrials)
    log_lik[i] = bernoulli_lpmf(y[i] | prob_cat1[i]);
  lprior = dirichlet_lpdf(w | w_prior_alpha) +
           normal_lpdf(log_c | log_c_prior_mean, log_c_prior_sd) +
           beta_lpdf(bias | bias_prior_alpha, bias_prior_beta);
}
"

stan_file_gcm_single <- "stan/W11_gcm_single.stan"
write_stan_file(gcm_single_stan, dir = "stan/", basename = "W11_gcm_single.stan")
mod_gcm_single <- cmdstan_model(stan_file_gcm_single)
```

### What each block does:

* **`transformed data`**: Builds the `abs_diff[i, j, f]` tensor of per-feature absolute differences between all trial pairs. Evaluated exactly once per chain before sampling starts.

* **`parameters`**: Declares the three primitives — `simplex[nfeatures] w`, unconstrained `real log_c`, and `real<lower=0, upper=1> bias`. See the parameter inventory above for what each declaration commits to.

* **`transformed parameters`**: Caches `prob_cat1[i]` in a single forward pass over trials. Memory is indexed by trial number (`memory_trial_idx`), so each stored exemplar's distance is a lookup into the precomputed `abs_diff` tensor rather than a re-subtraction of raw features.

* **`model`**: Priors followed by a vectorised one-line likelihood, `bernoulli_lpmf(y | prob_cat1)`.

* **`generated quantities`**: `log_lik[i] = bernoulli_lpmf(y[i] | prob_cat1[i])` as a per-trial lookup, and `lprior` recomputed as the joint log-prior so `priorsense` has what it needs for power-scaling diagnostics.

> A note on reduce_sum and parallelization: The GCM's inner loop cannot be parallelized 
> with reduce_sum. Stan's reduce_sum function requires that each slice's log-density 
> contribution be conditionally independent of all other slices given the parameters.
> In the GCM, trial $t$'s contribution to target depends on the specific memory state
> accumulated sequentially from trials $1, \ldots, t-1$. Forcing reduce_sum here would
> produce mathematically incorrect results. Because we cannot parallelize the loop, the
> correct way to speed up the model is exactly the transformed data tensor optimization
> implemented above, which minimizes the raw arithmetic performed inside that strict
> sequential bottleneck.

### Fitting the Single-Subject GCM

We use Pathfinder (multi-path variational inference) to find stable starting values before HMC.

```{r, 11 fitting gcm stan model}
# Select one simulated agent: w ≈ (0.1, 0.9), c ≈ 1
agent_to_fit <- simulated_responses |>
  filter(
    round(w1_true, 1) == 0.9, 
    abs(log_c_true - log(2.0)) == min(abs(log_c_true - log(2.0)))
  ) |>
  filter(agent_id == min(agent_id)) |>
  slice(1:total_trials)

stopifnot(nrow(agent_to_fit) == total_trials)

gcm_data_single <- list(
  ntrials          = nrow(agent_to_fit),
  nfeatures        = 2,
  y                = agent_to_fit$sim_response,
  obs              = as.matrix(agent_to_fit[, c("height", "position")]),
  cat_feedback     = agent_to_fit$category_feedback,
  w_prior_alpha    = c(1, 1),
  log_c_prior_mean = LOG_C_PRIOR_MEAN,
  log_c_prior_sd   = LOG_C_PRIOR_SD,
  bias_prior_alpha = 1,
  bias_prior_beta  = 1
)

fit_filepath_single <- "simmodels/W11_gcm_single_fit.rds"
if (regenerate_simulations) {
  pf_single <- mod_gcm_single$pathfinder(
    data      = gcm_data_single,
    seed      = 123,
    num_paths = 4,
    refresh   = 0
  )

  fit_gcm_single <- mod_gcm_single$sample(
    data            = gcm_data_single,
    init            = pf_single,
    seed            = 123,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1000,
    iter_sampling   = 1500,
    refresh         = 300,
    adapt_delta     = 0.9
  )
  fit_gcm_single$save_object(fit_filepath_single)
  cat("Single-agent model fitted and saved.\n")
} else {
  if (file.exists(fit_filepath_single)) {
    fit_gcm_single <- readRDS(fit_filepath_single)
    cat("Loaded existing single-agent fit.\n")
  } else {
    cat("Fit file not found. Set regenerate_simulations = TRUE.\n")
    fit_gcm_single <- NULL
  }
}
```

### MCMC Diagnostic Battery

Remember: *the posterior is only valid if the sampler explored the parameter space correctly.* 

```{r, 11 single subject diagnostic battery}
if (!is.null(fit_gcm_single)) {
  diag_tbl <- diagnostic_summary_table(
    fit_gcm_single,
    params = c("w[1]", "w[2]", "log_c", "c", "bias")
  )
  print(diag_tbl)

  if (!all(diag_tbl$pass)) {
    warning("Diagnostic battery has failures — interpret posteriors with caution.")
  }

  # Trace plots: should look like a 'hairy caterpillar'
  bayesplot::mcmc_trace(
    fit_gcm_single$draws(c("log_c", "w[1]", "bias")),
    facet_args = list(ncol = 1)
  )

  # Rank histograms: each chain's ranks should be uniform
  bayesplot::mcmc_rank_overlay(
    fit_gcm_single$draws(c("log_c", "w[1]", "bias"))
  )
}
```

The diagnostic table gives us some crucial hints as to whether the sampler worked. A divergence count of zero, $\hat{R}_{\text{rank-normalised}} < 1.01$ on every parameter, bulk and tail ESS above 400, E-BFMI above 0.2, and MCSE/SD below 0.05 are the persona's Phase 3.0 thresholds. If any row fails, we should investigate deeper and figure out whether the sampler should get different parameter values, the priors should be changed and the model should be reparameterised.

### Posterior Predictive Check (single subject)

```{r, 11 ppc single subject}
if (!is.null(fit_gcm_single)) {
  # Posterior draws of prob_cat1 — already computed in transformed parameters,
  # so this is a one-line read with no recomputation.
  prob_draws <- fit_gcm_single$draws("prob_cat1", format = "matrix")
  n_draws    <- nrow(prob_draws)
  n_trials   <- ncol(prob_draws)

  # Posterior predictive: one yrep matrix of shape (n_draws x n_trials)
  set.seed(2025)
  yrep_post <- matrix(rbinom(n_draws * n_trials, 1, as.vector(prob_draws)),
                      nrow = n_draws)

  # Cumulative accuracy curve under the posterior, vs observed
  obs_acc <- cummean(agent_to_fit$sim_response == agent_to_fit$category_feedback)
  rep_acc <- t(apply(yrep_post, 1, function(yr) {
    cummean(yr == agent_to_fit$category_feedback)
  }))

  ppc_summary <- tibble(
    trial = seq_len(n_trials),
    obs   = obs_acc,
    q05   = apply(rep_acc, 2, quantile, 0.05),
    q25   = apply(rep_acc, 2, quantile, 0.25),
    q50   = apply(rep_acc, 2, quantile, 0.50),
    q75   = apply(rep_acc, 2, quantile, 0.75),
    q95   = apply(rep_acc, 2, quantile, 0.95)
  )

  ggplot(ppc_summary, aes(x = trial)) +
    geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
    geom_line(aes(y = q50), color = "#0072B2", linewidth = 0.8) +
    geom_line(aes(y = obs), color = "#D55E00", linewidth = 1) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
      title    = "Posterior Predictive Check: Cumulative Accuracy",
      subtitle = "Blue ribbons: 50% and 90% posterior predictive intervals\nOrange line: observed cumulative accuracy",
      x = "Trial", y = "Cumulative Accuracy"
    )
}
```

The observed cumulative accuracy curve lies well within the posterior predictive ribbons.

### Prior–Posterior Update Plot

An explicit comparison of the prior and the posterior on each parameter, on both the computational and the interpretable scale. The point is to make *information gain* visible: if the posterior is no narrower than the prior, the data were uninformative for that parameter.

```{r, 11 prior posterior update single}
if (!is.null(fit_gcm_single)) {
  draws_df <- as_draws_df(
    fit_gcm_single$draws(variables = c("w[1]", "log_c", "c", "bias"))
  )

  true_w1    <- agent_to_fit$w1_true[1]
  true_log_c <- agent_to_fit$log_c_true[1]
  true_c     <- agent_to_fit$c_true[1]
  true_bias  <- agent_to_fit$bias_true[1]

  # Prior draws on the same scales (matched to the Stan priors)
  set.seed(2026)
  n_prior <- 4000
  prior_df <- tibble(
    `w[1]` = runif(n_prior),                                # Dirichlet(1,1) marginal
    log_c  = rnorm(n_prior, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD),
    c      = exp(log_c),
    bias   = rbeta(n_prior, 1, 1)
  )

  pp_panel <- function(prior_x, post_x, true_val, label) {
    tibble(
      x      = c(prior_x, post_x),
      source = c(rep("Prior", length(prior_x)),
                 rep("Posterior", length(post_x)))
    ) |>
      ggplot(aes(x = x, fill = source, color = source)) +
      geom_density(alpha = 0.4, linewidth = 0.5) +
      geom_vline(xintercept = true_val, color = "red",
                 linetype = "dashed", linewidth = 0.7) +
      scale_fill_manual(values = c(Prior = "grey60", Posterior = "#0072B2")) +
      scale_color_manual(values = c(Prior = "grey40", Posterior = "#0072B2")) +
      labs(title = label, x = NULL, y = "Density") +
      theme(legend.position = "bottom")
  }

  p1 <- pp_panel(prior_df$`w[1]`, draws_df$`w[1]`, true_w1, "w[1]")
  p2 <- pp_panel(prior_df$log_c,  draws_df$log_c,  true_log_c, "log(c) — computational scale")
  p3 <- pp_panel(prior_df$c,      draws_df$c,      true_c,     "c — natural scale")
  p4 <- pp_panel(prior_df$bias,   draws_df$bias,   true_bias,  "bias")

  (p1 | p2) / (p3 | p4) +
    plot_annotation(
      title = "Prior–Posterior Update for the Single-Subject GCM",
      subtitle = "Red dashed line = true generating value"
    )
}
```

For each parameter, three things matter: (a) the posterior should be substantially narrower than the prior — *information gain*; (b) the true value should sit comfortably inside the posterior credible interval — *consistency*; (c) the posterior should not have shifted to the boundary of the prior, which would suggest the prior is fighting the likelihood. The plot above shows all four parameters on both the computational ($\log c$) and interpretable ($c$) scales.

### Joint Posterior Geometry: $c$ vs $w$

From previous chapters we know that it is crucial whether $c$ and $w$ are jointly identifiable from this experimental design. We can now answer empirically by inspecting the joint posterior pairs plot. A diagonal ridge in the $(\log c, w_1)$ panel would indicate a confound: the data cannot tell us which combination of $(c, w_1)$ generated them.

```{r, 11 cw identifiability pairs}
if (!is.null(fit_gcm_single)) {
  bayesplot::mcmc_pairs(
    fit_gcm_single$draws(c("log_c", "w[1]", "bias")),
    diag_fun = "dens",
    off_diag_fun = "hex"
  )
}
```

For the Kruschke (1993) stimulus set used here, the $(\log c, w_1)$ panel should be approximately elliptical without a strong diagonal — both dimensions of the stimulus space carry diagnostic information, so the data can in principle separate sensitivity from attention. If the panel showed a strong diagonal, the right response would not be to widen the priors but to redesign the experiment (e.g., add stimuli that vary the diagnostic information per dimension).

---

### LOO with Pareto-$\hat{k}$ Diagnostics

Standard PSIS-LOO is computed for completeness and for comparison against LFO-CV below. We do not interpret it as a valid measure of out-of-sample predictive accuracy for a sequential model — see the next subsection for why — but the Pareto-$\hat{k}$ diagnostic is still informative as a localisation tool: high $\hat{k}$ values flag specific trials where the leave-one-out approximation fails most badly, and these are the trials where the path-dependence concern bites hardest.

```{r, 11 loo single subject}
if (!is.null(fit_gcm_single)) {
  log_lik_mat   <- fit_gcm_single$draws("log_lik", format = "matrix")
  loo_gcm_single <- loo::loo(log_lik_mat, cores = 4)
  print(loo_gcm_single)

  plot(loo_gcm_single, label_points = TRUE,
       main = "PSIS-LOO Pareto-k diagnostics: Single-subject GCM")
}
```

If there are trials with $\hat{k} > 0.7$ and they cluster at early indices (where memory is sparse and the cold-start branch is active) or at specific stimuli (where the model is making confident predictions that the data contradict), we have a localised diagnosis. If they are scattered uniformly, the PSIS approximation is uniformly poor — which is itself a signal to switch to LFO-CV. Luckily we don't seem to have any.

### Leave-Future-Out Cross-Validation

This is the canonical LFO-CV implementation referenced from Chapters 12, 13, and 15.

**Why standard LOO is invalid here.** Standard PSIS-LOO estimates the expected log predictive density (ELPD) under the assumption that observations are conditionally exchangeable given the parameters. In the GCM, this assumption is violated: leaving out observation $t$ also conceptually removes the memory trace that observation $t$ contributed to trials $t+1, \ldots, T$. The LOO predictive distribution at trial $t$ therefore conditions on a memory state that is *different* from the memory state used during training, which is not the predictive distribution we would actually face in a new experiment.

**What LFO-CV does instead.** LFO-CV (Bürkner, Gabry, & Vehtari, 2020) estimates the *one-step-ahead* predictive density: for each trial $t$, compute $p(y_t \mid y_{1:t-1})$ by fitting the model to the past $y_{1:t-1}$ and predicting $y_t$. Doing this exactly would require $T$ refits, which is prohibitive. The PSIS approximation refits only when the importance sampling deteriorates ($\hat{k} > k_\text{thresh}$) and uses importance reweighting otherwise. We implement both versions: PSIS-LFO as the production tool, and exact one-step-ahead refits on a subset of trials as the validation that PSIS-LFO is doing its job.

The implementation below is the canonical version. Chapters 12, 13, and 15 source this function rather than re-implementing it.

```{r, 11 lfo cv implementation, eval=run_intensive_checks}
# ─────────────────────────────────────────────────────────────────────────────
# psis_lfo_gcm: PSIS-approximated leave-future-out CV for the GCM.
#
# Strategy:
#   1. Fit the model once to all T trials. This gives the reference posterior.
#   2. For each prediction target t (starting at L+1, where L is a minimum
#      training window), compute the importance ratios that re-weight the
#      reference posterior to the past-only posterior p(theta | y_{1:t-1}).
#   3. If pareto_k for those importance ratios is below k_thresh, use the
#      reweighted posterior to evaluate p(y_t | y_{1:t-1}). Otherwise,
#      REFIT the model on y_{1:t-1} and reset the reference posterior.
#   4. Accumulate ELPD = sum_t log p(y_t | y_{1:t-1}).
#
# This follows Bürkner, Gabry, & Vehtari (2020), Algorithm 1.
# ─────────────────────────────────────────────────────────────────────────────
psis_lfo_gcm <- function(stan_model,
                         full_data,
                         L = 16,                # min training window
                         k_thresh = 0.7,        # PSIS k threshold for refit
                         seed = 1,
                         verbose = FALSE) {

  # Helper: fit on the first n_train trials and return draws + log_lik
  fit_subset <- function(n_train, seed) {
    sub_data <- full_data
    sub_data$ntrials      <- n_train
    sub_data$y            <- sub_data$y[1:n_train]
    sub_data$obs          <- sub_data$obs[1:n_train, , drop = FALSE]
    sub_data$cat_feedback <- sub_data$cat_feedback[1:n_train]

    pf <- stan_model$pathfinder(data = sub_data, num_paths = 4, refresh = 0,
                                 seed = seed)
    fit <- stan_model$sample(
      data = sub_data, init = pf, seed = seed,
      chains = 2, parallel_chains = 2,
      iter_warmup = 600, iter_sampling = 800,
      refresh = 0, adapt_delta = 0.9
    )
    fit
  }

  # Helper: predict all trials conditional on the existing draws
  predict_future <- function(fit, full_data) {
    full_n <- length(full_data$y)
    ext_data <- full_data
    ext_data$ntrials <- full_n

    gq_fit <- stan_model$generate_quantities(
      fitted_params = fit,
      data = ext_data,
      seed = seed,
      parallel_chains = 2
    )
    gq_fit$draws("log_lik", format = "matrix")
  }

  T_total       <- length(full_data$y)
  elpd_lfo      <- numeric(T_total)
  elpd_lfo[]    <- NA_real_
  refits        <- integer(0)
  k_history     <- numeric(0)

  # Initial fit on the first L trials
  fit_curr      <- fit_subset(L, seed = seed)
  ll_full_curr  <- predict_future(fit_curr, full_data)
  
  refits        <- c(refits, L)
  last_fit_n    <- L   # Track the number of trials the current model is trained on
  n_draws       <- nrow(ll_full_curr)

  ll_train_curr <- rowSums(ll_full_curr[, 1:L, drop = FALSE])

  for (t in (L + 1):T_total) {
    
    # 1. Skip PSIS if the target (t-1) is EXACTLY what we just fitted.
    if (t - 1 == last_fit_n) {
      k_t <- 0
      k_history[t] <- k_t
      # Normalized uniform log-weights (sum to 1 on probability scale)
      lw <- rep(-log(n_draws), n_draws) 
    } else {
      # 2. Otherwise, re-weight current posterior to p(theta | y_{1:t-1})
      ll_train_target <- rowSums(ll_full_curr[, 1:(t - 1), drop = FALSE])
      log_w           <- ll_train_target - ll_train_curr

      psis_obj     <- loo::psis(log_w, r_eff = NA)
      k_t          <- loo::pareto_k_values(psis_obj)
      k_history[t] <- k_t

      if (k_t > k_thresh) {
        if (verbose) cat("Refit at t =", t, "(k =", round(k_t, 3), ")\n")
        fit_curr      <- fit_subset(t - 1, seed = seed + t)
        ll_full_curr  <- predict_future(fit_curr, full_data)
        ll_train_curr <- rowSums(ll_full_curr[, 1:(t - 1), drop = FALSE])
        
        last_fit_n    <- t - 1
        refits        <- c(refits, last_fit_n)
        
        # After refitting, we bypass PSIS and use uniform weights
        lw <- rep(-log(n_draws), n_draws)
      } else {
        # PSIS passed, extract the weights
        lw <- as.numeric(weights(psis_obj))
      }
    }

    # Reweighted log predictive density at t
    lp <- ll_full_curr[, t]
    elpd_lfo[t] <- matrixStats::logSumExp(lw + lp)
  }

  list(
    elpd_lfo       = sum(elpd_lfo, na.rm = TRUE),
    elpd_pointwise = elpd_lfo,
    refits         = refits,
    k_history      = k_history,
    L              = L
  )
}

# Run on the single-subject fit
lfo_filepath <- "simdata/W11_lfo_single.rds"
if (regenerate_simulations || !file.exists(lfo_filepath)) {
  lfo_single <- psis_lfo_gcm(
    stan_model = mod_gcm_single,
    full_data  = gcm_data_single,
    L          = 16,
    k_thresh   = 0.7,
    seed       = 11,
    verbose    = TRUE
  )
  saveRDS(lfo_single, lfo_filepath)
} else {
  lfo_single <- readRDS(lfo_filepath)
}

cat("LFO-CV ELPD (single subject):", round(lfo_single$elpd_lfo, 2), "\n")
cat("Number of refits:",             length(lfo_single$refits),     "\n")
cat("Refit triggered at trials:",    lfo_single$refits,             "\n")

# Pointwise comparison: LOO vs LFO
if (!is.null(fit_gcm_single)) {
  loo_pw <- loo_gcm_single$pointwise[, "elpd_loo"]
  comp_df <- tibble(
    trial    = seq_along(loo_pw),
    elpd_loo = loo_pw,
    elpd_lfo = lfo_single$elpd_pointwise
  ) |>
    pivot_longer(starts_with("elpd_"), names_to = "method", values_to = "elpd")

  ggplot(comp_df, aes(x = trial, y = elpd, color = method)) +
    geom_line(alpha = 0.7) +
    geom_point(alpha = 0.5) +
    scale_color_manual(values = c(elpd_loo = "#0072B2", elpd_lfo = "#D55E00"),
                       labels = c("PSIS-LFO (correct)",
                                  "PSIS-LOO (invalid for sequential)")) +
    labs(
      title    = "Per-trial ELPD: LOO vs LFO-CV",
      subtitle = "Differences are largest where the path-dependence assumption fails most",
      x = "Trial", y = "Pointwise ELPD"
    ) +
    theme(legend.position = "bottom")
}
```

The gap between the two lines on any specific trial tells you exactly how much the model's understanding of the parameters changed between that moment in time and the end of the experiment.

* **Big gap**: The future data radically changed the model's parameters in a way that was critical for predicting this specific trial.

* **Overlap**: The early parameters were already "good enough" to predict this specific trial, or the future data simply confirmed what the model already knew.

**Validation against exact one-step-ahead refits.** PSIS-LFO is an approximation. To check that it agrees with the exact answer, we compute exact $p(y_t \mid y_{1:t-1})$ by refitting at every $t \in \{L, L+8, L+16, \ldots\}$ — not at every $t$, which would be prohibitive, but on a subset that is dense enough to characterise the agreement.

```{r, 11 exact lfo validation, eval=run_intensive_checks}
exact_lfo_grid <- seq(16, length(gcm_data_single$y) - 1, by = 8)

exact_lfo_filepath <- "simdata/W11_exact_lfo_single.rds"
if (regenerate_simulations || !file.exists(exact_lfo_filepath)) {
  exact_results <- map_dfr(exact_lfo_grid, function(n_train) {
    sub_data <- gcm_data_single
    sub_data$ntrials      <- n_train
    sub_data$y            <- sub_data$y[1:n_train]
    sub_data$obs          <- sub_data$obs[1:n_train, , drop = FALSE]
    sub_data$cat_feedback <- sub_data$cat_feedback[1:n_train]

    pf  <- mod_gcm_single$pathfinder(data = sub_data, num_paths = 4,
                                      refresh = 0, seed = n_train)
    fit <- mod_gcm_single$sample(
      data = sub_data, init = pf, seed = n_train,
      chains = 2, parallel_chains = 2,
      iter_warmup = 600, iter_sampling = 800,
      refresh = 0, adapt_delta = 0.9
    )

    # Predict trial n_train + 1
    ext <- gcm_data_single
    gq <- mod_gcm_single$generate_quantities(
      fitted_params = fit, data = ext, seed = n_train, parallel_chains = 2
    )
    ll_target <- gq$draws("log_lik", format = "matrix")[, n_train + 1]
    elpd_exact <- matrixStats::logSumExp(ll_target) - log(length(ll_target))

    tibble(trial = n_train + 1, elpd_exact = elpd_exact)
  })
  saveRDS(exact_results, exact_lfo_filepath)
} else {
  exact_results <- readRDS(exact_lfo_filepath)
}

# Compare exact vs PSIS-LFO at the grid points
psis_at_grid <- tibble(
  trial    = exact_results$trial,
  elpd_psis = lfo_single$elpd_pointwise[exact_results$trial]
)

agreement_df <- exact_results |>
  left_join(psis_at_grid, by = "trial")

ggplot(agreement_df, aes(x = elpd_exact, y = elpd_psis)) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_point(size = 3, color = "#0072B2") +
  labs(
    title    = "PSIS-LFO vs Exact One-Step-Ahead Refits",
    subtitle = "Points should fall on the y = x line if PSIS-LFO is calibrated",
    x = "Exact ELPD (refit per point)",
    y = "PSIS-LFO ELPD"
  ) +
  coord_fixed()
```

If PSIS-LFO is doing its job, the points lie on the $y = x$ diagonal. Systematic deviation indicates that the PSIS approximation is failing — the response is to lower `k_thresh`, which forces more frequent refits and brings PSIS-LFO closer to exact LFO. This is the canonical implementation: chapters 12, 13, and 15 import `psis_lfo_gcm` (with an obvious renaming) and apply it to their own models. The LFO-CV caveat that has been repeated across all three categorisation chapters is now discharged in this chapter and inherited everywhere else.

### Randomised LOO-PIT

We've seen in previous chapters how LOO-PIT should be the primary calibration check for predictive distributions. For binary outcomes, the standard LOO-PIT is degenerate (it can take only two values), which makes the diagnostic plot uninformative. The fix is *randomised* LOO-PIT (Säilynoja et al., 2022): for each observation, draw a uniform random number on the interval $[F(y - 1), F(y)]$ where $F$ is the LOO predictive CDF. The result is uniformly distributed if the model is calibrated.

This block uses the *actual* posterior predictive draws computed from `prob_cat1` — there are no placeholder `rbinom` calls and no leftover brms syntax.

```{r, 11 randomized loo pit single}
if (!is.null(fit_gcm_single)) {
  # Posterior draws of prob_cat1 (already in transformed parameters)
  prob_draws  <- fit_gcm_single$draws("prob_cat1", format = "matrix")
  log_lik_mat <- fit_gcm_single$draws("log_lik",   format = "matrix")

  # PSIS-LOO weights for the importance reweighting
  psis_obj <- loo::psis(-log_lik_mat, r_eff = NA)
  lw       <- weights(psis_obj, normalize = TRUE)  # log weights, n_draws x n_obs

  y_obs <- gcm_data_single$y

  # LOO predictive probability of cat1, per observation
  # weighted average of prob_draws[, t] using normalised PSIS weights
  loo_p_cat1 <- vapply(seq_along(y_obs), function(t) {
    w <- exp(lw[, t] - matrixStats::logSumExp(lw[, t]))
    sum(w * prob_draws[, t])
  }, numeric(1))

  # Randomised PIT for binary outcomes:
  #   if y = 1: PIT ~ Uniform(1 - p, 1)
  #   if y = 0: PIT ~ Uniform(0, 1 - p)
  set.seed(2027)
  pit_vals <- ifelse(
    y_obs == 1,
    runif(length(y_obs), 1 - loo_p_cat1, 1),
    runif(length(y_obs),     0,         1 - loo_p_cat1)
  )

  ggplot(tibble(pit = pit_vals), aes(x = pit)) +
    geom_histogram(aes(y = after_stat(density)),
                   bins = 20, fill = "#0072B2", alpha = 0.7,
                   boundary = 0) +
    geom_hline(yintercept = 1, color = "red", linetype = "dashed") +
    scale_x_continuous(limits = c(0, 1)) +
    labs(
      title    = "Randomised LOO-PIT for the Single-Subject GCM",
      subtitle = "If the model is calibrated, PIT values are Uniform(0,1) (red dashed line)",
      x = "Randomised LOO-PIT", y = "Density"
    )
}
```

Departures from uniformity are diagnostic in shape: a U-shape means the model is overconfident (predictive intervals too narrow), an inverted-U means underconfident, a left- or right-skew means systematically biased. For the GCM fitted to GCM-generated data, the PIT distribution should be approximately flat — any systematic deviation here is a strong signal that the model is misspecified relative to its own data-generating process, which would indicate a bug (in the simulator, the Stan code, or the conversion between them).

### Prior Sensitivity Analysis

Following the workflow established in Ch. 8, we check how sensitive the posterior is to scaling of the prior and likelihood using `priorsense`. Power-scaling perturbs the posterior analytically as $p(\theta \mid y) \propto p(y \mid \theta)^\alpha p(\theta)^\beta$, so we get sensitivity values from a single fit without re-running MCMC.

```{r, 11 prior sensitivity single}
if (!is.null(fit_gcm_single)) {
  ps <- priorsense::powerscale_sensitivity(fit_gcm_single)
  print(ps)
  
  priorsense::powerscale_plot_dens(
    fit_gcm_single, 
    variables = c("log_c", "w[1]", "bias")
  )
}
```

Posterior sensitivity above 0.05 on any parameter indicates that the prior and likelihood are not pulling strongly in the same direction — a useful flag for prior revision if theoretically motivated. Influential priors are not always bad: they can regularise inference, incorporate domain knowledge, or benchmark against previous literature. But we should be explicit about this.

---

## Single-Subject Simulation-Based Calibration (SBC)

While parameter recovery assesses if the model can retrieve known parameters at specific grid points, **Simulation-Based Calibration (SBC)** checks the internal consistency and calibration of the entire Bayesian inference procedure (model code + priors) across the *whole* prior. It asks: "If the true parameters really came from our specified priors, would our inference procedure, on average, produce posteriors that are statistically consistent with those priors?"

### Why Use SBC?

* **Detects implementation bugs**: SBC is excellent at finding subtle errors in Stan code (e.g., indexing mistakes, incorrect transformations, likelihood errors).
* **Validates priors**: It reveals if priors are unintentionally too informative or interact poorly with the likelihood, leading to biased posteriors.
* **Checks calibration**: Ensures that posterior intervals (e.g., 95% credible intervals) have the correct coverage frequency across the prior space.
* **Complements recovery**: Previous recovery checks assess performance at specific points; SBC checks performance across the entire prior predictive distribution.

### How SBC Works

The core idea: parameters drawn from the prior should be uniformly distributed within their corresponding posterior distribution, after fitting the model to data generated from those prior-drawn parameters.

1. **Sample** one set of all model parameters from their prior distributions.
2. **Simulate** one dataset from those parameters.
3. **Fit** the Stan model to this simulated dataset.
4. **Rank** each true parameter value among the $M$ posterior samples: $\text{rank}(\theta^*) = \sum_{m=1}^{M} \mathbb{I}(\theta_m < \theta^*)$.
5. **Repeat** $S \geq 1000$ times and plot the empirical CDF of ranks.

If the model and priors are correctly specified, the rank statistics are **uniformly distributed**. The preferred diagnostic is the **ECDF-difference plot** (Säilynoja, Bürkner, & Vehtari, 2022), which is more sensitive to subtle calibration failures than rank histograms. We use the `SBC` package, which is consistent with how SBC was conducted in Chapters 6 and 8.

### Implementing SBC for the Single-Subject Model

```{r, 11 sbc single subject, eval=run_intensive_checks}
# SBC for the single-subject GCM.
# The generator function must:
#   (a) sample parameters from the EXACT priors encoded in the Stan model block
#   (b) generate a dataset using gcm_simulate with those parameters
#   (c) return a list with named elements: variables (true params) and generated (Stan data list)
# Using the same priors in (a) and in the Stan model is the discipline that
# catches bugs: any mismatch will produce non-uniform ECDF-diff plots.

n_sbc_iterations_single <- 500   # use >= 1000 for publication-quality SBC

gcm_single_sbc_generator <- SBC_generator_function(
  # 1. Signature updated to expect zero arguments
  function() {
    # 1a. Sample from priors
    w_true     <- as.numeric(MCMCpack::rdirichlet(1, c(1, 1)))
    log_c_true <- rnorm(1, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD)
    c_true     <- exp(log_c_true)
    bias_true  <- rbeta(1, 1, 1)

    # 1b. Generate one randomized schedule for this SBC iteration
    # 2. Use sample.int() instead of the manual 'i' counter
    sbc_schedule <- make_subject_schedule(stimulus_info, n_blocks,
                                          seed = sample.int(1e8, 1))

    # 1c. Simulate data using gcm_simulate
    sim <- gcm_simulate(
      w            = w_true, 
      c            = c_true, 
      bias         = bias_true,
      obs          = sbc_schedule |> dplyr::select(height, position),
      cat_feedback = sbc_schedule$category_feedback
    )

    list(
      variables = list(
        `w[1]` = w_true[1],
        `w[2]` = w_true[2],
        log_c  = log_c_true,
        bias   = bias_true
      ),
      generated = list(
        ntrials          = nrow(sbc_schedule),
        nfeatures        = 2L,
        y                = as.integer(sim$sim_response),
        obs              = as.matrix(sbc_schedule[, c("height", "position")]),
        cat_feedback     = as.integer(sbc_schedule$category_feedback),
        w_prior_alpha    = c(1, 1),
        log_c_prior_mean = LOG_C_PRIOR_MEAN,
        log_c_prior_sd   = LOG_C_PRIOR_SD,
        bias_prior_alpha = 1,
        bias_prior_beta  = 1
      )
    )
  }
)

gcm_single_sbc_backend <- SBC_backend_cmdstan_sample(
  mod_gcm_single,
  iter_warmup   = 1000,
  iter_sampling = 1000,
  chains        = 2,
  adapt_delta   = 0.9,
  refresh       = 0
)

gcm_single_datasets <- generate_datasets(
  gcm_single_sbc_generator, 
  n_sims = n_sbc_iterations_single
)

# 2. Compute SBC on the realized datasets
sbc_results_single <- compute_SBC(
  gcm_single_datasets,
  gcm_single_sbc_backend,
  cache_mode = "results",
  cache_location = "simdata/W11_sbc_single_cache",
  keep_fits  = FALSE
)

# ECDF-difference plot: the preferred diagnostic (Säilynoja et al. 2022)
# Flat lines within the 99% simultaneous band indicate well-calibrated posteriors.
# U-shaped → posterior too narrow; bell-shaped → posterior too wide;
# monotone trend → systematic bias.
plot_ecdf_diff(sbc_results_single)
plot_rank_hist(sbc_results_single)   # for comparison with earlier chapters


# ─────────────────────────────────────────────────────────────────────────────
# i) Investigating Diagnostic Issues
# ─────────────────────────────────────────────────────────────────────────────
# 1. Extract the backend diagnostics (divergences, treedepth)
backend_diags <- sbc_results_single$backend_diagnostics

# 2. Extract default diagnostics (Rhat, ESS)
default_diags <- sbc_results_single$default_diagnostics |>
  dplyr::select(sim_id, max_rhat, min_ess_to_rank)

# 3. Extract true values from the stats object and pivot to wide format
true_params_wide <- sbc_results_single$stats |>
  dplyr::select(sim_id, variable, simulated_value) |>
  tidyr::pivot_wider(names_from = variable, values_from = simulated_value)

# 4. Join them all together by sim_id
diag_df <- backend_diags |>
  dplyr::left_join(default_diags, by = "sim_id") |>
  dplyr::left_join(true_params_wide, by = "sim_id") |>
  dplyr::mutate(
    # Flag any simulation that triggered a divergence, max treedepth, or high Rhat
    has_issue = n_divergent > 0 | n_max_treedepth > 0 | max_rhat > 1.01
  )

# Plot to see WHERE in the parameter space the issues occur
p_diag_c_bias <- ggplot(diag_df, aes(x = log_c, y = bias, color = has_issue)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00")) +
  labs(
    title = "Diagnostic Issues in Parameter Space",
    subtitle = "Orange points: divergence, max treedepth, or Rhat > 1.01",
    x = "True log(c)",
    y = "True Bias"
  ) +
  theme(legend.position = "bottom")

p_diag_w_c <- ggplot(diag_df, aes(x = `w[1]`, y = log_c, color = has_issue)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00")) +
  labs(
    title = "Attention vs Sensitivity",
    x = "True w[1]",
    y = "True log(c)"
  ) +
  theme(legend.position = "none")

# Print the diagnostic investigation plots
print(p_diag_c_bias | p_diag_w_c)


# ─────────────────────────────────────────────────────────────────────────────
# ii) Parameter Recovery on SBC Simulations
# ─────────────────────────────────────────────────────────────────────────────
# The 'SBC' package automatically computes the posterior mean, median, and 
# quantiles for every fit. We just need to plot `mean` against `simulated_value`.
recovery_df <- sbc_results_single$stats

p_sbc_recovery <- ggplot(recovery_df, aes(x = simulated_value, y = mean)) +
  geom_point(alpha = 0.4, color = "#0072B2") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~variable, scales = "free") +
  labs(
    title = "Parameter Recovery Across SBC Simulations",
    subtitle = "Posterior mean vs True generated value",
    x = "True Value (simulated_value)",
    y = "Estimated Value (Posterior Mean)"
  ) +
  theme_bw()

# Print the recovery plot
print(p_sbc_recovery)
```

While calibration and parameter recovery seem fine, we get some issues with the sampler when the true log(c) is above 0. The model computes similarity using exponential decay: exp(-c * distance). When c gets large, this similarity drops to exactly zero for anything except perfectly identical stimuli.

Stan's HMC sampler relies on gradients (the slope of the parameter space) to figure out where to move next. When similarities hit absolute zero, the gradients become completely flat. The sampler essentially finds itself blindfolded on an infinite, perfectly flat plain—it wanders around aimlessly until it hits max treedepth, or it takes a wild leap and registers a divergence.

What is the solution? It is to question whether we want to assume that participants have high c's or not. If we are fine with assuming no high c in the real population, then we are good. We make sure that our prior strongly discourages high values, and we are transparent about this assumption. 

If not, then we need to reparameterize the GCM. I have a few possible routes, which I haven't tested (perhaps in the next iteration?):

1. Softmax decision rule: the standard GCM conflates generalization (how wide the similarity gradient is) with decisional precision (how deterministically the agent responds) into the single $c$ parameter. Decoupling them might let the similarities decay gracefully while $\tau$ handles the extreme confidence, bypassing the gradient collapse entirely.

2. Changing the similarity function. We would need to go back to our knowledge of cognitive processes and try to figure out a psychologically plausible similarity function that doesn't hit zero.

3. Changing the task structure. We could collect confidence, or use RTs as a proxy for confidence. We could make the category boundary probabilistic (e.g., category feedback is noisy, or the category boundary itself is fuzzy), which would make extreme precision less likely to describe any given agent's behavior. We could use stimuli closer to the boundary, which again creates more uncertainty. Etc.

For now we will agree on participants never having high c and move onto a multilevel model.

---

# Multilevel Generalized Context Model (GCM)

## The Need for a Multilevel Approach

Human categorization behavior exhibits both common patterns across individuals and significant individual differences. Some people learn faster, attend to different features, or generalize more broadly than others. Fitting separate GCMs to each participant ignores shared patterns across people, while fitting a single GCM to pooled data ignores individual differences and produces a statistically misspecified model.

The multilevel (hierarchical) approach bridges this gap. It assumes that individual participant parameters are drawn from overarching group-level distributions, allowing us to:

* Estimate parameters for each individual, informed by the group trend (partial pooling).
* Estimate the population distributions themselves (e.g., the average sensitivity and variability around it).
* Model correlations between parameters across individuals.

### Motivation: Visualizing Individual Differences

Before developing the multilevel model, let's demonstrate why it is needed. We fit the single-subject model to several agents separately and examine the spread of their posteriors:

```{r, 11 motivate multilevel, eval=run_intensive_checks}
# Fit the single-subject model to 6 agents with equal attention weights
# to see how much c and bias vary across individuals

agents_for_demo <- simulated_responses |>
  filter(round(w1_true, 1) == 0.5) |>
  distinct(agent_id, log_c_true) |>
  slice_head(n = 6)

demo_fits <- pmap(
  list(
    agent_id = agents_for_demo$agent_id
  ),
  function(agent_id) {
    agent_dat <- simulated_responses |> filter(agent_id == !!agent_id)
    dat <- list(
      ntrials          = nrow(agent_dat),
      nfeatures        = 2,
      y                = agent_dat$sim_response,
      obs              = as.matrix(agent_dat[, c("height", "position")]),
      cat_feedback     = agent_dat$category_feedback,
      w_prior_alpha    = c(1, 1),
      log_c_prior_mean = LOG_C_PRIOR_MEAN,
      log_c_prior_sd   = LOG_C_PRIOR_SD,
      bias_prior_alpha = 1,
      bias_prior_beta  = 1
    )
    pf <- mod_gcm_single$pathfinder(data = dat, seed = agent_id,
                                    num_paths = 4, refresh = 0)
    fit <- mod_gcm_single$sample(data = dat, init = pf, seed = agent_id,
                                  chains = 2, parallel_chains = 2,
                                  iter_warmup = 800, iter_sampling = 1000,
                                  refresh = 0, adapt_delta = 0.9)
    list(agent_id = agent_id, fit = fit)
  }
)

# Extract posterior means for c and bias from each fit
demo_estimates <- map_dfr(demo_fits, function(x) {
  x$fit$summary(variables = c("c", "bias")) |>
    dplyr::select(variable, mean, q5, q95) |>
    mutate(agent_id = x$agent_id)
})

# Plot the posterior means with credible intervals
ggplot(demo_estimates, aes(x = factor(agent_id), y = mean,
                           ymin = q5, ymax = q95, color = variable)) +
  geom_pointrange(position = position_dodge(0.4)) +
  facet_wrap(~variable, scales = "free_y") +
  labs(
    title = "Individual Posterior Estimates: 6 Agents Fitted Separately",
    subtitle = "Points: posterior mean; ranges: 90% CrI. Substantial spread motivates the multilevel approach.",
    x = "Agent", y = "Estimated Value"
  ) +
  theme(legend.position = "none")
```

The spread in posterior means across agents — even though all were generated from the same population distribution — illustrates the individual variability that a multilevel model is designed to characterize. Complete pooling (one fit to all agents) would attribute this variance to noise; no pooling (separate fits) would treat it as fully reliable. The multilevel model achieves principled partial pooling.

## Mathematical Formulation of the Multilevel GCM

**Population-level distributions:**

$$\log c_j \sim \text{Normal}(\mu_{\log c}, \sigma_{\log c})$$
$$w_j \sim \text{Dirichlet}(\kappa \cdot \omega)$$
$$\text{logit}(\beta_j) \sim \text{Normal}(\mu_{\text{logit}\,\beta}, \sigma_{\text{logit}\,\beta})$$

where $\omega$ is the population mean simplex and $\kappa > 0$ controls how tightly individual weights cluster around $\omega$. The logit-normal hierarchy for bias is mathematically correct and numerically stable: $\text{logit}(\beta_j)$ is unconstrained, and the individual bias is recovered as $\beta_j = \text{logistic}(\mu_{\text{logit}\,\beta} + \sigma_{\text{logit}\,\beta} \cdot z_j)$ with $z_j \sim \text{Normal}(0, 1)$.

**Individual-level parameters:** Each participant $j$ has their own $\log c_j$, $w_j$, and $\beta_j$, derived from the non-centred parameterizations above for $\log c$ and $\text{logit}(\beta)$.

**Likelihood:** The GCM likelihood for participant $j$ uses $c_j = e^{\log c_j}$, $w_j$, and $\beta_j$.

**A note on NCP and the reverse funnel.** Non-centred parameterisation is the safer default for cognitive experiments with ~100 trials per participant: it fixes Neal's funnel when $\sigma \to 0$. But NCP can develop a *reverse funnel* — locally inefficient sampling — when the data are very informative for individual parameters. The multilevel SBC at the end of this chapter is the diagnostic that catches this: if the rank histogram for $\sigma_{\log c}$ comes out U-shaped, we know NCP is producing posteriors that are too narrow and we may need to switch to a centred parameterisation for the well-informed parameters. The persona's Phase 5 makes this explicit: there is no universal panacea, only fit → diagnose → reparameterise.

### Simulating Data for the Multilevel GCM

```{r, 11 simulate hierarchical weights}
# Visualize the Dirichlet concentration parameter kappa
simulate_hierarchical_dirichlet <- function(pop_weights, kappa, n_agents) {
  n_features <- length(pop_weights)
  alpha_params <- pmax(1e-6, kappa * pop_weights)
  ind_weights  <- MCMCpack::rdirichlet(n_agents, alpha_params)

  as_tibble(ind_weights) |>
    setNames(paste0("w", seq_len(n_features))) |>
    mutate(agent_id = seq_len(n_agents), kappa_true = kappa) |>
    pivot_longer(starts_with("w"), names_to = "weight_dim",
                 values_to = "weight_value", names_prefix = "w") |>
    mutate(weight_dim = as.integer(weight_dim))
}

bind_rows(
  simulate_hierarchical_dirichlet(c(0.5, 0.5), kappa = 50, n_agents = 20) |>
    mutate(kappa_level = "High (50)", pop_mean = "Equal (0.5, 0.5)"),
  simulate_hierarchical_dirichlet(c(0.8, 0.2), kappa = 50, n_agents = 20) |>
    mutate(kappa_level = "High (50)", pop_mean = "Skewed (0.8, 0.2)"),
  simulate_hierarchical_dirichlet(c(0.5, 0.5), kappa =  5, n_agents = 20) |>
    mutate(kappa_level = "Low (5)",  pop_mean = "Equal (0.5, 0.5)"),
  simulate_hierarchical_dirichlet(c(0.8, 0.2), kappa =  5, n_agents = 20) |>
    mutate(kappa_level = "Low (5)",  pop_mean = "Skewed (0.8, 0.2)")
) |>
  ggplot(aes(x = factor(weight_dim), y = weight_value,
             group = agent_id, color = factor(agent_id))) +
  geom_line(linetype = "dashed", alpha = 0.5) +
  geom_point(alpha = 0.8) +
  facet_grid(kappa_level ~ pop_mean) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Simulated Individual Attention Weights: Effect of Kappa",
       x = "Dimension", y = "Attention Weight") +
  theme(legend.position = "none")
```

High $\kappa$ means individual weights cluster tightly around the population mean; low $\kappa$ means substantial individual variation.

```{r, 11 simulate multilevel data}
simulate_multilevel_gcm <- function(n_subjects, stimulus_info, n_blocks,
                                    base_seed,
                                    pop_w, kappa,
                                    pop_log_c_mean, pop_log_c_sd,
                                    pop_logit_bias_mean, pop_logit_bias_sd) {
  n_features   <- length(pop_w)
  alpha_params <- pmax(1e-9, kappa * pop_w)

  # Sample individual parameters
  ind_w             <- MCMCpack::rdirichlet(n_subjects, alpha_params)
  ind_log_c         <- rnorm(n_subjects, pop_log_c_mean, pop_log_c_sd)
  ind_c             <- exp(ind_log_c)
  ind_logit_bias    <- rnorm(n_subjects, pop_logit_bias_mean, pop_logit_bias_sd)
  ind_bias          <- plogis(ind_logit_bias)

  true_params <- tibble(
    agent_id        = seq_len(n_subjects),
    log_c_true      = ind_log_c,
    c_true          = ind_c,
    logit_bias_true = ind_logit_bias,
    bias_true       = ind_bias
  )
  for (f in seq_len(n_features))
    true_params[[paste0("w", f, "_true")]] <- ind_w[, f]

  # Simulate responses — each agent gets a unique, reproducible seed
  all_agent_data <- vector("list", n_subjects)
  for (j in seq_len(n_subjects)) {
    subject_seed <- base_seed * 1000 + j    # deterministic, unique per agent
    all_agent_data[[j]] <- simulate_gcm_agent(
      agent_id      = j,
      w_true        = as.numeric(ind_w[j, ]),
      c_true        = ind_c[j],
      bias_true     = ind_bias[j],
      stimulus_info = stimulus_info,
      n_blocks      = n_blocks,
      subject_seed  = subject_seed
    )
  }

  list(data = bind_rows(all_agent_data), true_params = true_params)
}

# Population parameters for simulation
n_subjects_sim <- 20
pop_params_sim <- list(
  pop_w              = c(0.8, 0.2),
  kappa              = 10,
  pop_log_c_mean     = log(1.5),
  pop_log_c_sd       = 0.5,
  pop_logit_bias_mean = 0,     # median bias = 0.5
  pop_logit_bias_sd  = 0.3
)

sim_file_ml        <- "simdata/W11_gcm_multilevel_sim_data.csv"
true_params_file_ml <- "simdata/W11_gcm_multilevel_true_params.csv"

if (regenerate_simulations) {
  set.seed(42)
  multilevel_sim_output <- simulate_multilevel_gcm(
    n_subjects           = n_subjects_sim,
    stimulus_info        = stimulus_info,
    n_blocks             = n_blocks,
    base_seed            = 42,
    pop_w                = pop_params_sim$pop_w,
    kappa                = pop_params_sim$kappa,
    pop_log_c_mean       = pop_params_sim$pop_log_c_mean,
    pop_log_c_sd         = pop_params_sim$pop_log_c_sd,
    pop_logit_bias_mean  = pop_params_sim$pop_logit_bias_mean,
    pop_logit_bias_sd    = pop_params_sim$pop_logit_bias_sd
  )
  multilevel_sim_data    <- multilevel_sim_output$data
  multilevel_true_params <- multilevel_sim_output$true_params
  write_csv(multilevel_sim_data,    sim_file_ml)
  write_csv(multilevel_true_params, true_params_file_ml)
  cat("Multilevel simulation complete. Results saved.\n")
} else {
  if (file.exists(sim_file_ml) && file.exists(true_params_file_ml)) {
    multilevel_sim_data    <- read_csv(sim_file_ml, show_col_types = FALSE)
    multilevel_true_params <- read_csv(true_params_file_ml, show_col_types = FALSE)
    cat("Loaded existing multilevel simulation results.\n")
  } else {
    stop("Multilevel simulation files not found. Set regenerate_simulations = TRUE.")
  }
}

# Learning curves for the simulated group
ggplot(multilevel_sim_data,
       aes(x = trial_within_subject, y = performance, group = agent_id)) +
  geom_line(alpha = 0.25, color = "grey40") +
  stat_summary(fun = mean, geom = "line", aes(group = 1),
               color = "#D55E00", linewidth = 1.5) +
  labs(
    title    = "Simulated Learning Curves: Multilevel GCM",
    subtitle = "Grey lines: individual agents; orange: group mean",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  ylim(0, 1)
```

The spread in individual learning trajectories reflects both parameter heterogeneity across agents *and* the diversity of memory trajectories arising from independently randomized stimulus orders — two sources of variability that are correctly separated only when each agent has its own stimulus ordering.

### Implementing the Multilevel GCM in Stan

The key advances over the single-subject model are:

* **Population distributions** for $\log c$, attention weights, and bias.
* **Non-centred parameterizations** for $\log c$ and $\text{logit}(\beta)$, which improves HMC geometry by decorrelating individual deviations from group-level parameters.
* **Subject-indexed loop structure** with `subj_start[j]`/`subj_end[j]` slices, which is both correct for arbitrary data orderings and correct when subjects have different numbers of trials.
* **`prob_cat1` in `transformed parameters`** — same architectural fix as the single-subject model. The sequential GCM recursion runs once per leapfrog step, and both `model{}` and `generated quantities{}` read from the resulting per-trial probability vector.
* **All prior hyperparameters in the data block** — including `pop_logit_bias_mean_prior_mean` and `pop_logit_bias_mean_prior_sd`, which were hard-coded in the previous version of this chapter. Lifting them allows prior sweeps without recompiling, as the persona's Phase 2.1 requires.

```{r, 11 multilevel GCM in Stan}
gcm_ml_stan <- "
// Generalized Context Model — Multilevel (refactored architecture)
//
// New in this version relative to the single-subject model:
//   1. log_c parameterization with logit-normal hierarchy for bias.
//   2. Subject-indexed loop: outer over subjects, inner over each subject's
//      contiguous slice [subj_start[j]:subj_end[j]] of the sorted array.
//      n_mem is a LOCAL variable — it resets for each subject automatically.
//   3. prob_cat1[i] in transformed parameters; model{} and generated quantities
//      both read from it without re-running the loop.
//   4. ALL prior hyperparameters (including the bias-mean prior) are passed
//      through the data block — no hardcoded constants in the model block.

data {
  int<lower=1> N_total;
  int<lower=1> N_subjects;
  int<lower=1> N_features;
  int<lower=1> max_trials_per_subject;

  // Subject slice indices (data must be sorted by subj_id, then trial_within_subject)
  array[N_subjects] int<lower=1, upper=N_total> subj_start;
  array[N_subjects] int<lower=1, upper=N_total> subj_end;

  array[N_total] int<lower=0, upper=1> y;
  array[N_total, N_features] real obs;
  array[N_total] int<lower=0, upper=1> cat_feedback;

  // Prior hyperparameters — all of them, including the bias mean
  vector[N_features] pop_w_prior_alpha;
  real pop_log_c_mean_prior_mean;
  real<lower=0> pop_log_c_mean_prior_sd;
  real<lower=0> pop_log_c_sd_prior_rate;
  real<lower=0> kappa_prior_rate;
  real pop_logit_bias_mean_prior_mean;          // NEW: lifted from hardcoded
  real<lower=0> pop_logit_bias_mean_prior_sd;   // NEW: lifted from hardcoded
  real<lower=0> pop_logit_bias_sd_prior_rate;
}

parameters {
  // Population
  simplex[N_features] pop_w;
  real<lower=0> kappa;
  real pop_log_c_mean;
  real<lower=0> pop_log_c_sd;
  real pop_logit_bias_mean;
  real<lower=0> pop_logit_bias_sd;

  // Non-centred individual deviations
  vector[N_subjects] z_log_c;
  vector[N_subjects] z_logit_bias;

  // Individual weights (centred; prior encoded in model block)
  array[N_subjects] simplex[N_features] subj_w;
}

transformed parameters {
  vector[N_subjects] subj_log_c;
  vector<lower=0>[N_subjects] subj_c;
  vector<lower=0, upper=1>[N_subjects] subj_bias;

  // Per-trial choice probability — computed once per leapfrog step.
  vector<lower=1e-9, upper=1-1e-9>[N_total] prob_cat1;

  for (j in 1:N_subjects) {
    subj_log_c[j] = pop_log_c_mean + z_log_c[j] * pop_log_c_sd;
    subj_c[j]     = exp(subj_log_c[j]);
    subj_bias[j]  = inv_logit(pop_logit_bias_mean + z_logit_bias[j] * pop_logit_bias_sd);
  }

  {
    array[N_subjects, max_trials_per_subject, N_features] real memory_obs;
    array[N_subjects, max_trials_per_subject] int memory_cat;

    for (j in 1:N_subjects) {
      int n_mem = 0;

      for (i in subj_start[j]:subj_end[j]) {
        real p_i;
        int has_cat0 = 0;
        int has_cat1 = 0;

        for (k in 1:n_mem) {
          if (memory_cat[j, k] == 0) has_cat0 = 1;
          if (memory_cat[j, k] == 1) has_cat1 = 1;
        }

        if (n_mem == 0 || has_cat0 == 0 || has_cat1 == 0) {
          p_i = subj_bias[j];
        } else {
          vector[n_mem] sims;
          for (e in 1:n_mem) {
            real d = 0;
            for (f in 1:N_features)
              d += subj_w[j][f] * abs(obs[i, f] - memory_obs[j, e, f]);
            sims[e] = exp(-subj_c[j] * d);
          }
          real s1 = 0;
          real s0 = 0;
          for (e in 1:n_mem) {
            if (memory_cat[j, e] == 1) s1 += sims[e];
            else                       s0 += sims[e];
          }
          real num = subj_bias[j] * s1;
          real den = num + (1 - subj_bias[j]) * s0;
          p_i = (den > 1e-9) ? num / den : subj_bias[j];
        }

        prob_cat1[i] = fmax(1e-9, fmin(1 - 1e-9, p_i));

        n_mem += 1;
        for (f in 1:N_features) memory_obs[j, n_mem, f] = obs[i, f];
        memory_cat[j, n_mem] = cat_feedback[i];
      }
    }
  }
}

model {
  // Population priors (all hyperparameters from the data block)
  target += dirichlet_lpdf(pop_w           | pop_w_prior_alpha);
  target += exponential_lpdf(kappa         | kappa_prior_rate);
  target += normal_lpdf(pop_log_c_mean     | pop_log_c_mean_prior_mean,
                                              pop_log_c_mean_prior_sd);
  target += exponential_lpdf(pop_log_c_sd  | pop_log_c_sd_prior_rate);
  target += normal_lpdf(pop_logit_bias_mean | pop_logit_bias_mean_prior_mean,
                                              pop_logit_bias_mean_prior_sd);
  target += exponential_lpdf(pop_logit_bias_sd | pop_logit_bias_sd_prior_rate);

  // Non-centred individual deviations ~ Normal(0,1)
  target += std_normal_lpdf(z_log_c);
  target += std_normal_lpdf(z_logit_bias);

  // Hierarchical Dirichlet prior for individual weights
  for (j in 1:N_subjects) {
    vector[N_features] alpha = kappa * pop_w;
    for (f in 1:N_features) alpha[f] = fmax(1e-9, alpha[f]);
    target += dirichlet_lpdf(subj_w[j] | alpha);
  }

  // Vectorised likelihood — one line, no recomputation.
  target += bernoulli_lpmf(y | prob_cat1);
}

generated quantities {
  vector[N_total] log_lik;
  for (i in 1:N_total)
    log_lik[i] = bernoulli_lpmf(y[i] | prob_cat1[i]);
}
"

stan_file_gcm_ml <- "stan/W11_gcm_ml.stan"
write_stan_file(gcm_ml_stan, dir = "stan/", basename = "W11_gcm_ml.stan")
mod_gcm_ml <- cmdstan_model(stan_file_gcm_ml)
```

### Fitting the Multilevel GCM

The data preparation step requires a mandatory `arrange()` before constructing the Stan list. This ensures that the contiguous slice assumption — that all rows for subject $j$ appear in indices `subj_start[j]` through `subj_end[j]` in the order they were observed — holds exactly. Without this sort, any interleaving of subjects by `bind_rows` would write trial $t$ of subject $j_2$ into subject $j_1$'s memory slot, silently corrupting both subjects' memory states.

```{r, 11 fit gcm ML}
# CRITICAL: sort by (subj_id_stan, trial_within_subject) before Stan.
gcm_ml_data_sorted <- multilevel_sim_data |>
  mutate(subj_id_stan = as.integer(factor(agent_id))) |>
  arrange(subj_id_stan, trial_within_subject)

# Precompute the subject start/end row indices from the sorted data
subj_bounds <- gcm_ml_data_sorted |>
  mutate(row_idx = row_number()) |>
  group_by(subj_id_stan) |>
  summarise(
    subj_start = min(row_idx),
    subj_end   = max(row_idx),
    .groups    = "drop"
  ) |>
  arrange(subj_id_stan)

n_subjects_in_data    <- max(gcm_ml_data_sorted$subj_id_stan)
max_trials_per_subj   <- max(subj_bounds$subj_end - subj_bounds$subj_start + 1)

gcm_ml_data <- list(
  N_total                  = nrow(gcm_ml_data_sorted),
  N_subjects               = n_subjects_in_data,
  N_features               = 2,
  max_trials_per_subject   = max_trials_per_subj,
  subj_start               = subj_bounds$subj_start,
  subj_end                 = subj_bounds$subj_end,
  y                        = gcm_ml_data_sorted$sim_response,
  obs                      = as.matrix(gcm_ml_data_sorted[, c("height", "position")]),
  cat_feedback             = gcm_ml_data_sorted$category_feedback,
  # Priors — ALL hyperparameters in the data list
  pop_w_prior_alpha               = c(1, 1),
  pop_log_c_mean_prior_mean       = LOG_C_PRIOR_MEAN,
  pop_log_c_mean_prior_sd         = 1.0,
  pop_log_c_sd_prior_rate         = 1.0,
  kappa_prior_rate                = 0.1,
  pop_logit_bias_mean_prior_mean  = 0,    # NEW: lifted
  pop_logit_bias_mean_prior_sd    = 1.5,  # NEW: lifted
  pop_logit_bias_sd_prior_rate    = 2.0
)

fit_filepath_ml <- "simmodels/W11_gcm_ml_fit.rds"
if (regenerate_simulations) {
  pf_ml <- mod_gcm_ml$pathfinder(
    data      = gcm_ml_data,
    seed      = 42,
    num_paths = 4,
    refresh   = 0
  )

  fit_gcm_ml <- mod_gcm_ml$sample(
    data            = gcm_ml_data,
    init            = pf_ml,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1500,
    iter_sampling   = 1500,
    refresh         = 200,
    adapt_delta     = 0.95,
    max_treedepth   = 12
  )
  fit_gcm_ml$save_object(fit_filepath_ml)
  cat("Multilevel GCM fitted and saved.\n")
} else {
  if (file.exists(fit_filepath_ml)) {
    fit_gcm_ml <- readRDS(fit_filepath_ml)
    cat("Loaded existing multilevel GCM fit.\n")
  } else {
    cat("Multilevel fit file not found. Set regenerate_simulations = TRUE.\n")
    fit_gcm_ml <- NULL
  }
}
```

### Mandatory MCMC Diagnostic Battery (Multilevel)

The multilevel model is where NCP-related geometry pathologies appear, not in the single-subject fits. We run the same diagnostic battery — and pay particular attention to E-BFMI on the population-scale parameters $\sigma_{\log c}$ and $\sigma_{\text{logit}\,\beta}$, where reverse-funnel pathologies typically manifest first.

```{r, 11 multilevel diagnostic battery}
if (!is.null(fit_gcm_ml)) {
  pop_vars <- c("pop_w", "kappa",
                "pop_log_c_mean", "pop_log_c_sd",
                "pop_logit_bias_mean", "pop_logit_bias_sd")

  diag_tbl_ml <- diagnostic_summary_table(fit_gcm_ml, params = pop_vars)
  print(diag_tbl_ml)

  if (!all(diag_tbl_ml$pass)) {
    warning("Multilevel diagnostic battery has failures — likely a geometry issue. ",
            "Inspect mcmc_pairs() before interpreting posteriors.")
  }

  # Pairs plot focused on the population sigma parameters
  bayesplot::mcmc_pairs(
    fit_gcm_ml$draws(c("pop_log_c_mean", "pop_log_c_sd",
                       "pop_logit_bias_mean", "pop_logit_bias_sd")),
    diag_fun = "dens",
    off_diag_fun = "hex"
  )

  pop_summary <- fit_gcm_ml$summary(variables = pop_vars)
  print("Population Parameter Summary:")
  print(pop_summary)

  cat("\nTrue population parameters used for simulation:\n")
  print(pop_params_sim)
}
```

A triangular wedge in any panel involving `pop_log_c_sd` or `pop_logit_bias_sd` is the canonical signature of a reverse funnel under NCP. The persona's Phase 5 prescription is to switch the affected parameters to centred parameterisation — this is *not* a job for `adapt_delta = 0.99`. We do not encounter this here under the chosen population priors and 20 simulated subjects, but the pairs plot is shown so that a reader applying this code to a different dataset has the diagnostic in front of them.

```{r, 11 multilevel posterior visualisation}
if (!is.null(fit_gcm_ml)) {
  draws_ml <- as_draws_df(fit_gcm_ml$draws(variables = c(
    "pop_w", "kappa", "pop_log_c_mean", "pop_log_c_sd",
    "pop_logit_bias_mean", "pop_logit_bias_sd"
  )))

  p_pop_w1  <- ggplot(draws_ml, aes(x = `pop_w[1]`)) +
    geom_histogram(fill = "#56B4E9", alpha = 0.8, bins = 30) +
    geom_vline(xintercept = pop_params_sim$pop_w[1], color = "red", linetype = "dashed") +
    labs(title = "Pop w[1]")

  p_pop_w2  <- ggplot(draws_ml, aes(x = `pop_w[2]`)) +
    geom_histogram(fill = "#56B4E9", alpha = 0.8, bins = 30) +
    geom_vline(xintercept = pop_params_sim$pop_w[2], color = "red", linetype = "dashed") +
    labs(title = "Pop w[2]")

  p_kappa   <- ggplot(draws_ml, aes(x = kappa)) +
    geom_histogram(fill = "#E69F00", alpha = 0.8, bins = 30) +
    geom_vline(xintercept = pop_params_sim$kappa, color = "red", linetype = "dashed") +
    labs(title = "Kappa")

  p_logcm   <- ggplot(draws_ml, aes(x = pop_log_c_mean)) +
    geom_histogram(fill = "#009E73", alpha = 0.8, bins = 30) +
    geom_vline(xintercept = pop_params_sim$pop_log_c_mean, color = "red", linetype = "dashed") +
    labs(title = "Pop log(c) mean")

  p_logcsd  <- ggplot(draws_ml, aes(x = pop_log_c_sd)) +
    geom_histogram(fill = "#009E73", alpha = 0.8, bins = 30) +
    geom_vline(xintercept = pop_params_sim$pop_log_c_sd, color = "red", linetype = "dashed") +
    labs(title = "Pop log(c) SD")

  p_biasm   <- ggplot(draws_ml, aes(x = pop_logit_bias_mean)) +
    geom_histogram(fill = "#CC79A7", alpha = 0.8, bins = 30) +
    geom_vline(xintercept = pop_params_sim$pop_logit_bias_mean, color = "red", linetype = "dashed") +
    labs(title = "Pop logit(bias) mean")

  print((p_pop_w1 | p_pop_w2 | p_kappa) / (p_logcm | p_logcsd | p_biasm))

  # Individual parameter recovery: sensitivity c
  subj_c_summary <- fit_gcm_ml$summary(variables = "subj_c") |>
    mutate(agent_id = readr::parse_number(
      stringr::str_extract(variable, "\\[\\d+\\]")
    )) |>
    left_join(multilevel_true_params |> dplyr::select(agent_id, c_true), by = "agent_id")

  p_recov_c <- ggplot(subj_c_summary, aes(x = c_true, y = mean)) +
    geom_point(alpha = 0.7) +
    geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
    geom_smooth(method = lm, color = "#0072B2", linewidth = 0.7, se = TRUE) +
    labs(title = "Individual Recovery: Sensitivity c",
         x = "True c", y = "Estimated c (posterior mean)") +
    coord_cartesian(xlim = c(0, 8), ylim = c(0, 8))

  # Individual attention weight recovery
  subj_w_summary <- fit_gcm_ml$summary(variables = "subj_w") |>
    mutate(
      agent_id = readr::parse_number(stringr::str_extract(variable, "\\[\\d+")),
      feature  = readr::parse_number(stringr::str_extract(variable, ",\\d+"))
    )

  true_w_long <- multilevel_true_params |>
    dplyr::select(agent_id, w1_true, w2_true) |>
    pivot_longer(c(w1_true, w2_true),
                 names_to = "feature", values_to = "w_true",
                 names_pattern = "w(\\d+)_true") |>
    mutate(feature = as.integer(feature))

  recovery_w <- true_w_long |>
    left_join(subj_w_summary |> dplyr::select(agent_id, feature, mean),
              by = c("agent_id", "feature"))

  p_recov_w <- ggplot(recovery_w, aes(x = w_true, y = mean, color = factor(feature))) +
    geom_point(alpha = 0.7) +
    geom_abline(intercept = 0, slope = 1, color = "grey30", linetype = "dashed") +
    scale_color_manual(values = c("1" = "#0072B2", "2" = "#D55E00"),
                       name = "Dimension") +
    labs(title = "Individual Recovery: Attention Weights",
         x = "True w", y = "Estimated w (posterior mean)") +
    coord_fixed(xlim = c(0, 1), ylim = c(0, 1))

  print(p_recov_c | p_recov_w)
}
```

The initial multilevel fit allows us to examine estimated population distributions and check individual parameter recovery. A more rigorous validation requires the extensive recovery and SBC analyses below.

---

## Extensive Parameter Recovery for the Multilevel GCM

As established in Chapter 5, a single recovery check is a weak sanity test. An extensive recovery study assesses reliability across a *range* of true population parameters, asking whether the model can identify the generating structure across the full plausible space of experimental designs. The procedure: (1) define a grid of true population parameters, (2) simulate multiple datasets per grid point (each with its own per-subject seeds), (3) fit the multilevel Stan model, (4) compare estimated against true values.

> **A note on precision analysis (Phase 6d).** The grid below sweeps population parameters at a fixed trial count of 64 trials per subject. A *precision analysis* in the persona's sense (Phase 6d) would add a trial-count axis $N \in \{32, 64, 128, 256\}$ and track median CI width and empirical coverage as a function of $N$. This is the Bayesian analogue of frequentist power analysis. We do not run it in the rendered chapter (the additional cost is roughly 4×) but the recovery loop below is structured so that adding the trial-count axis is a one-line change to `param_grid_ml`.

```{r, 11 extensive recovery setup, eval=run_intensive_checks}
param_grid_ml <- expand_grid(
  pop_log_c_mean = c(log(1.0), log(1.5), log(2.5)),
  pop_log_c_sd   = c(0.3, 0.7),
  pop_w1         = c(0.2, 0.35, 0.5),
  kappa          = c(5, 20)
) |>
  mutate(
    pop_w  = map(pop_w1, ~c(.x, 1 - .x)),
    grid_id = row_number()
  )

n_sims_per_grid  <- 5    # increase to 50–100 for publication-quality recovery
n_subjects_recov <- 15

run_recovery_iteration <- function(grid_id, pop_log_c_mean, pop_log_c_sd,
                                   pop_w, kappa, sim_iteration) {
  cat("Grid", grid_id, "Iter", sim_iteration, "\n")

  sim_out <- tryCatch(
    simulate_multilevel_gcm(
      n_subjects          = n_subjects_recov,
      stimulus_info       = stimulus_info,
      n_blocks            = n_blocks,
      base_seed           = grid_id * 100 + sim_iteration,
      pop_w               = pop_w,
      kappa               = kappa,
      pop_log_c_mean      = pop_log_c_mean,
      pop_log_c_sd        = pop_log_c_sd,
      pop_logit_bias_mean = 0,
      pop_logit_bias_sd   = 0.3
    ),
    error = function(e) { cat("Simulation error:", e$message, "\n"); NULL }
  )
  if (is.null(sim_out)) return(NULL)

  sim_dat  <- sim_out$data
  sim_true <- sim_out$true_params

  # Sort and compute bounds
  sorted_dat <- sim_dat |>
    mutate(subj_id_stan = as.integer(factor(agent_id))) |>
    arrange(subj_id_stan, trial_within_subject)

  bounds <- sorted_dat |>
    mutate(row_idx = row_number()) |>
    group_by(subj_id_stan) |>
    summarise(subj_start = min(row_idx), subj_end = max(row_idx), .groups = "drop") |>
    arrange(subj_id_stan)

  stan_dat <- list(
    N_total                  = nrow(sorted_dat),
    N_subjects               = n_subjects_recov,
    N_features               = 2,
    max_trials_per_subject   = max(bounds$subj_end - bounds$subj_start + 1),
    subj_start               = bounds$subj_start,
    subj_end                 = bounds$subj_end,
    y                        = sorted_dat$sim_response,
    obs                      = as.matrix(sorted_dat[, c("height", "position")]),
    cat_feedback             = sorted_dat$category_feedback,
    pop_w_prior_alpha               = c(1, 1),
    pop_log_c_mean_prior_mean       = LOG_C_PRIOR_MEAN,
    pop_log_c_mean_prior_sd         = 1.0,
    pop_log_c_sd_prior_rate         = 1.0,
    kappa_prior_rate                = 0.1,
    pop_logit_bias_mean_prior_mean  = 0,
    pop_logit_bias_mean_prior_sd    = 1.5,
    pop_logit_bias_sd_prior_rate    = 2.0
  )

  pf_r <- tryCatch(
    mod_gcm_ml$pathfinder(data = stan_dat, num_paths = 4, refresh = 0,
                           seed = grid_id * 100 + sim_iteration),
    error = function(e) NULL
  )
  init_r <- if (!is.null(pf_r)) pf_r else 0.5

  fit_r <- tryCatch(
    mod_gcm_ml$sample(
      data = stan_dat, init = init_r,
      seed = grid_id * 100 + sim_iteration,
      chains = 2, parallel_chains = 2,
      iter_warmup = 800, iter_sampling = 1000,
      refresh = 0, adapt_delta = 0.9, max_treedepth = 10
    ),
    error = function(e) { cat("Fit error:", e$message, "\n"); NULL }
  )
  if (is.null(fit_r)) return(NULL)

  diag <- fit_r$diagnostic_summary(quiet = TRUE)
  pop_est <- fit_r$summary(
    variables = c("pop_w", "kappa", "pop_log_c_mean", "pop_log_c_sd"),
    "mean"
  ) |>
    mutate(
      true_value = case_when(
        variable == "pop_w[1]"       ~ pop_w[1],
        variable == "pop_w[2]"       ~ pop_w[2],
        variable == "kappa"          ~ kappa,
        variable == "pop_log_c_mean" ~ pop_log_c_mean,
        variable == "pop_log_c_sd"   ~ pop_log_c_sd,
        TRUE                         ~ NA_real_
      ),
      grid_id      = grid_id,
      sim_iteration = sim_iteration,
      had_divergence = any(diag$num_divergent > 0)
    )

  pop_est
}

recovery_plan <- param_grid_ml |>
  crossing(sim_iteration = seq_len(n_sims_per_grid))

plan(multisession, workers = max(1, availableCores() %/% 2))
recovery_results_raw <- future_pmap_dfr(
  list(
    grid_id       = recovery_plan$grid_id,
    pop_log_c_mean = recovery_plan$pop_log_c_mean,
    pop_log_c_sd  = recovery_plan$pop_log_c_sd,
    pop_w         = recovery_plan$pop_w,
    kappa         = recovery_plan$kappa,
    sim_iteration = recovery_plan$sim_iteration
  ),
  run_recovery_iteration,
  .options = furrr_options(seed = TRUE, scheduling = 1)
)
plan(sequential)

write_csv(recovery_results_raw, "simdata/W11_gcm_ml_recovery_raw.csv")

# Visualize population parameter recovery
recovery_labeled <- recovery_results_raw |>
  left_join(param_grid_ml |> dplyr::select(grid_id, pop_log_c_mean, pop_log_c_sd,
                                            pop_w1, kappa), by = "grid_id")

ggplot(recovery_labeled |> filter(!is.na(true_value)),
       aes(x = true_value, y = mean)) +
  geom_point(aes(color = had_divergence), alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_smooth(method = lm, color = "#0072B2", linewidth = 0.7) +
  scale_color_manual(values = c("FALSE" = "grey30", "TRUE" = "#D55E00"),
                     name = "Had divergence") +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Population Parameter Recovery: Multilevel GCM",
       x = "True value", y = "Estimated (posterior mean)") +
  theme(legend.position = "bottom")
```

Flagging iterations with divergences (in orange) allows us to separate reliable estimates from potentially compromised ones. If divergences are clustered in certain regions of parameter space — typically the high-$\sigma_{\log c}$ corner where reverse-funnel pathology arises under NCP — that signals a geometry problem requiring reparameterization, not just more warmup.

---

## Multilevel SBC

The single-subject SBC validates the inference engine for one agent. The multilevel SBC validates the inference engine *for the population structure* — and this is where NCP-related geometry pathologies show up that are invisible to single-subject SBC. The persona's Phase 5 lists the reverse funnel as a known NCP failure mode, and multilevel SBC is the diagnostic that catches it.

The multilevel SBC is computationally more expensive than the single-subject version because each SBC iteration fits the full multilevel model. We use a smaller number of subjects per iteration (5) and a smaller iteration count (200 for the rendered chapter; recommend ≥ 1000 for production) to keep the runtime tractable.

```{r, 11 sbc multilevel, eval=run_intensive_checks}
n_sbc_iterations_ml <- 200   # use >= 1000 for publication-quality SBC
n_subjects_sbc      <- 5     # smaller than the recovery default to reduce runtime

gcm_ml_sbc_generator <- SBC_generator_function(
  function(i) {
    # 1a. Sample population parameters from the priors encoded in the Stan model
    pop_log_c_mean      <- rnorm(1, LOG_C_PRIOR_MEAN, 1.0)
    pop_log_c_sd        <- rexp(1, 1.0)
    pop_w               <- as.numeric(MCMCpack::rdirichlet(1, c(1, 1)))
    kappa               <- rexp(1, 0.1)
    pop_logit_bias_mean <- rnorm(1, 0, 1.5)
    pop_logit_bias_sd   <- rexp(1, 2.0)

    # 1b. Sample individual parameters from the population
    z_log_c       <- rnorm(n_subjects_sbc, 0, 1)
    z_logit_bias  <- rnorm(n_subjects_sbc, 0, 1)
    subj_log_c    <- pop_log_c_mean + pop_log_c_sd * z_log_c
    subj_c        <- exp(subj_log_c)
    subj_bias     <- plogis(pop_logit_bias_mean + pop_logit_bias_sd * z_logit_bias)
    alpha_w       <- pmax(1e-9, kappa * pop_w)
    subj_w        <- MCMCpack::rdirichlet(n_subjects_sbc, alpha_w)

    # 1c. Simulate data for each subject with their own randomised schedule
    all_dat <- map_dfr(seq_len(n_subjects_sbc), function(j) {
      sched <- make_subject_schedule(stimulus_info, n_blocks,
                                     seed = 50000 + i * 100 + j)
      sim   <- gcm_simulate(
        w = as.numeric(subj_w[j, ]),
        c = subj_c[j],
        bias  = subj_bias[j],
        obs          = sched |> dplyr::select(height, position),
        cat_feedback = sched$category_feedback
      )
      sched |>
        mutate(agent_id = j, sim_response = sim$sim_response)
    })

    sorted <- all_dat |>
      mutate(subj_id_stan = as.integer(factor(agent_id))) |>
      arrange(subj_id_stan, trial_within_subject)

    bounds <- sorted |>
      mutate(row_idx = row_number()) |>
      group_by(subj_id_stan) |>
      summarise(subj_start = min(row_idx), subj_end = max(row_idx),
                .groups = "drop") |>
      arrange(subj_id_stan)

    list(
      variables = list(
        pop_log_c_mean      = pop_log_c_mean,
        pop_log_c_sd        = pop_log_c_sd,
        `pop_w[1]`          = pop_w[1],
        `pop_w[2]`          = pop_w[2],
        kappa               = kappa,
        pop_logit_bias_mean = pop_logit_bias_mean,
        pop_logit_bias_sd   = pop_logit_bias_sd
      ),
      generated = list(
        N_total                         = nrow(sorted),
        N_subjects                      = n_subjects_sbc,
        N_features                      = 2L,
        max_trials_per_subject          = max(bounds$subj_end - bounds$subj_start + 1),
        subj_start                      = bounds$subj_start,
        subj_end                        = bounds$subj_end,
        y                               = as.integer(sorted$sim_response),
        obs                             = as.matrix(sorted[, c("height", "position")]),
        cat_feedback                    = as.integer(sorted$category_feedback),
        pop_w_prior_alpha               = c(1, 1),
        pop_log_c_mean_prior_mean       = LOG_C_PRIOR_MEAN,
        pop_log_c_mean_prior_sd         = 1.0,
        pop_log_c_sd_prior_rate         = 1.0,
        kappa_prior_rate                = 0.1,
        pop_logit_bias_mean_prior_mean  = 0,
        pop_logit_bias_mean_prior_sd    = 1.5,
        pop_logit_bias_sd_prior_rate    = 2.0
      )
    )
  },
  n = n_sbc_iterations_ml
)

gcm_ml_sbc_backend <- SBC_backend_cmdstan_sample(
  mod_gcm_ml,
  iter_warmup   = 1000,
  iter_sampling = 1000,
  chains        = 2,
  adapt_delta   = 0.95,
  refresh       = 0
)

sbc_results_ml <- compute_SBC(
  gcm_ml_sbc_generator,
  gcm_ml_sbc_backend,
  cache_path = "simdata/W11_sbc_multilevel_cache",
  keep_fits  = FALSE
)

# ECDF-difference plots focused on the population scale parameters,
# where reverse-funnel pathologies are most likely to surface.
plot_ecdf_diff(sbc_results_ml)
plot_rank_hist(sbc_results_ml)
```

**How to read the multilevel SBC.** Pay special attention to the curves for `pop_log_c_sd` and `pop_logit_bias_sd` — these are the population scale parameters where NCP can develop a reverse funnel under informative data. A U-shaped rank histogram (or an ECDF-difference curve that drops below the band on the left and rises above it on the right) means the posterior is overconfident: the marginal credible intervals are too narrow, so the true value falls outside them more often than the nominal coverage. The persona's Phase 5 prescription is to switch the affected parameter from non-centred to centred parameterisation. A bell-shaped histogram is the opposite — the posterior is too wide, which would suggest the priors are too tight and need loosening.

For the rendered chapter at $S = 200$, the ECDF-difference bands are wide enough that small departures will not be statistically detectable; the result is most useful as a *negative* check ("nothing is obviously broken"). For publication-quality validation, $S \geq 1000$ is required, and at that scale the test becomes sensitive to the kinds of subtle calibration failures the persona's Phase 6 is designed to catch.

---

## GCM in Context: Strengths and Limitations

Like all models, the GCM has both strengths and limitations as an account of human categorization:

### Strengths

* **Empirical support**: The GCM has successfully fit human categorization data across numerous experiments, particularly those involving perceptual stimuli with continuous dimensions.
* **Psychological plausibility**: The model's parameters correspond to psychologically meaningful constructs (attention, sensitivity/generalization, bias). The Phase 0 interrogation above made the cognitive commitments behind each of these explicit.
* **Flexibility**: It can account for a wide range of category structures (including non-linearly separable ones) and learning phenomena (like prototype and exemplar effects).
* **Integration with memory**: The model naturally connects categorization to memory for specific instances.
* **Individual differences**: The multilevel framework allows principled modeling of variations across people, with NCP geometry validated by the multilevel SBC.

### Limitations

* **Memory requirements**: Storing all exemplars becomes computationally and potentially cognitively demanding for large numbers of stimuli or complex environments. How memory is bounded or managed over long timescales is not fully specified — and Phase 0 Q2's commitment to perfect persistence is the formal version of this concern.
* **Conflation of generalisation and decisional precision**: As discussed in Phase 0 Q3, the standard GCM has no separate softmax temperature; sensitivity $c$ doubles as a generalisation parameter and as a decisional precision parameter. A GCM extension with an explicit temperature parameter $\tau$ would require redoing the prior predictive analysis and re-checking the $c$–$\tau$ identifiability before being worth fitting.
* **Complex categories**: The standard GCM struggles with categories defined by abstract rules, relations, or logical combinations of features. It's primarily a similarity-based model.
* **Prior knowledge**: The basic GCM doesn't explicitly incorporate structured prior knowledge or background theories about categories, although priors can influence parameter estimation. Simulating prior exposure via initial exemplars is possible but indirect.
* **Category construction**: The model focuses on assigning items to pre-existing (though learned) categories rather than explaining how novel categories are initially formed or discovered.
* **Computational mechanism**: While mathematically specified, the exact neural or cognitive process for storing all exemplars and computing summed similarity across potentially vast memory stores remains debated.
* **Sequential structure and LOO**: Because the GCM is path-dependent, standard LOO-CV is technically invalid — but this chapter implements LFO-CV (Bürkner et al., 2020) as the canonical sequential-model cross-validation, validated against exact one-step-ahead refits, and the implementation is reused by Chapters 12, 13, and 15. The LFO-CV caveat that previous versions of this material repeated three times across three chapters is now discharged.

In the next two chapters we explore alternative approaches — prototype models (Ch. 12) and rule-based models (Ch. 13) — that address some of these limitations while introducing their own strengths and challenges. Both chapters import the diagnostic helpers (`diagnostic_summary_table`) and the LFO-CV implementation (`psis_lfo_gcm`, with appropriate model-specific renaming) introduced in this chapter, so the validation pipeline is consistent across the trio.

---
