---
title: "12 — Prototype Models: Learning Category Averages"
output:
  html_document:
    mathjax: default
    toc: true
    toc_float: true
  pdf_document:
    toc: true
date: "2025-04-28"
---

```{r 12_setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)

# Set to TRUE to rerun simulations and Stan fits.
# FALSE loads pre-computed results from 'simdata/'.
regenerate_simulations <- FALSE

# Set to TRUE to run the long-running validation chunks
# (LFO-CV, precision analysis, SBC). FALSE loads cached results
# or skips the chunk if no cache exists.
run_intensive_checks <- FALSE

pacman::p_load(
  tidyverse,    # data manipulation and visualization
  mvtnorm,      # multivariate normal densities
  patchwork,    # combining plots
  cmdstanr,     # Stan interface
  posterior,    # tidy posterior arrays
  tidybayes,    # tidy extraction of draws
  bayesplot,    # MCMC diagnostic plots
  loo,          # LOO-CV and PSIS
  SBC,          # Simulation-Based Calibration
  priorsense,   # prior sensitivity
  future,       # parallel processing
  furrr,        # parallel functional programming
  ellipse,      # uncertainty ellipses
  matrixStats,  # logSumExp for the LFO-CV implementation
  MASS          # ginv (pseudo-inverse fallback)
)

theme_set(theme_classic())

# Prior hyperparameter for log(r)
LOG_R_PRIOR_MEAN <- 0          # prior median r = 1
LOG_R_PRIOR_SD   <- 1.0        # covers r ≈ 0.13–7.4 at ±2 SD

for (d in c("stan", "simdata", "simmodels")) {
  if (!dir.exists(d)) dir.create(d)
}

# ── Shared diagnostic helper (canonical version lives in Ch 11) ──────────────
# We re-define it here for standalone use. The function returns a one-row
# tibble summarising the mandatory MCMC diagnostic battery for a cmdstanr
# fit, and prints a warning if any threshold is breached. Chapters 11, 12,
# 13, and 15 all call this so the diagnostic table looks the same everywhere.
diagnostic_summary_table <- function(fit, pars = NULL) {
  diag <- fit$diagnostic_summary(quiet = TRUE)
  draws_summary <- fit$summary(variables = pars)

  out <- tibble(
    n_divergent      = sum(diag$num_divergent),
    n_max_treedepth  = sum(diag$num_max_treedepth),
    min_ebfmi        = round(min(diag$ebfmi), 3),
    max_rhat         = round(max(draws_summary$rhat,    na.rm = TRUE), 4),
    min_ess_bulk     = round(min(draws_summary$ess_bulk, na.rm = TRUE), 0),
    min_ess_tail     = round(min(draws_summary$ess_tail, na.rm = TRUE), 0)
  ) |>
    mutate(
      max_mcse_ratio = {
        # MCSE_mean / posterior SD across all monitored params
        ratios <- draws_summary$mcse_mean / draws_summary$sd
        round(max(ratios, na.rm = TRUE), 4)
      }
    )

  # Threshold checks
  ok_div  <- out$n_divergent     == 0
  ok_rhat <- out$max_rhat        <  1.01
  ok_bulk <- out$min_ess_bulk    >  400
  ok_tail <- out$min_ess_tail    >  400
  ok_ebfmi<- out$min_ebfmi       >  0.2
  ok_mcse <- out$max_mcse_ratio  <  0.05

  if (!all(c(ok_div, ok_rhat, ok_bulk, ok_tail, ok_ebfmi, ok_mcse))) {
    warning(
      "MCMC diagnostic battery: at least one threshold breached. ",
      "Inspect divergences and reparameterize before interpreting the posterior.",
      call. = FALSE
    )
  }

  out
}

# ── Kruschke (1993) stimulus set ─────────────────────────────────────────────
# Re-defined here for standalone use; identical to Ch. 11.
stimulus_info <- tibble(
  stimulus      = c(5, 3, 7, 1, 8, 2, 6, 4),
  height        = c(1, 1, 2, 2, 3, 3, 4, 4),
  position      = c(2, 3, 1, 4, 1, 4, 2, 3),
  category_true = c(0, 0, 1, 0, 1, 0, 1, 1)
)

n_blocks         <- 8
n_stim_per_block <- nrow(stimulus_info)
total_trials     <- n_blocks * n_stim_per_block

# ── Per-subject schedule helper ───────────────────────────────────────────────
# Identical to Ch. 11. Redefined here for standalone use.
make_subject_schedule <- function(stimulus_info, n_blocks, seed) {
  set.seed(seed)
  n_stim <- nrow(stimulus_info)
  sequence <- unlist(lapply(seq_len(n_blocks), function(b) {
    sample(stimulus_info$stimulus, n_stim, replace = FALSE)
  }))
  tibble(
    trial_within_subject = seq_along(sequence),
    block                = rep(seq_len(n_blocks), each = n_stim),
    stimulus_id          = sequence
  ) |>
    left_join(stimulus_info, by = c("stimulus_id" = "stimulus")) |>
    rename(category_feedback = category_true) |>
    dplyr::select(trial_within_subject, block, stimulus_id,
                  height, position, category_feedback)
}

plan(multisession, workers = max(1, availableCores() - 1))
```

> **📍 Where we are in the Bayesian modeling workflow:**
> Chapter 11 built the GCM — an exemplar model that stores every encountered
> stimulus and decides by summed similarity to memory. This chapter introduces
> a structurally different account: the **prototype model**, which abstracts
> each category to a single running estimate (its mean and uncertainty) rather
> than storing individual instances. We implement prototype learning as a
> multivariate Kalman filter and then add a **Stan parameter-estimation model**
> for the prototype architecture.
>
> The full validation pipeline from Ch. 5 — Phase 0 interrogation, prior
> predictive check, Pathfinder initialization, mandatory MCMC diagnostic
> battery, posterior predictive check, prior–posterior update plot, randomised
> LOO-PIT, LFO-CV, parameter recovery, precision analysis, prior sensitivity,
> and SBC — is applied to this new model. Crucially, the **LFO-CV machinery
> implemented in Chapter 11** (`psis_lfo_gcm`) is generalised here as
> `psis_lfo_kalman` and applied to the prototype model. The repeated LFO-CV
> caveat that earlier drafts of this material carried across three chapters
> is now discharged: this chapter actually computes leave-future-out ELPD
> against exact one-step-ahead refits and reports both.

# Prototype-Based Models of Categorization

> **📋 Validation battery roadmap for this chapter**
>
> | Persona phase | Where it lives in this chapter |
> |---|---|
> | 0. Conceptual interrogation | §"Phase 0: Interrogating the Prototype Model" |
> | 1. Forward simulator | `prototype_kalman()` and `simulate_prototype_agent()` |
> | 2. Stan implementation | §"The Prototype Model in Stan" — `prob` in `transformed parameters` |
> | 3.0 MCMC diagnostic battery | §"Mandatory MCMC Diagnostic Battery" — uses `diagnostic_summary_table()` |
> | 3.2 Prior predictive | §"Prior Predictive Check" |
> | 3.3 Posterior predictive | §"Posterior Predictive Check" |
> | 3.4 Prior–posterior update | §"Prior–Posterior Update Plot" |
> | 3.5 Prior sensitivity | §"Prior Sensitivity Analysis" — `priorsense::powerscale_sensitivity()` |
> | 4.5 LOO-PIT (randomised, binary) | §"Randomised LOO-PIT Calibration" |
> | 4.6 LFO-CV | §"Leave-Future-Out Cross-Validation" — inherited from Ch 11 |
> | 6a. Parameter recovery | §"Parameter Recovery" |
> | 6b. SBC (single-subject) | §"Simulation-Based Calibration" |
> | 6d. Precision analysis | §"Precision Analysis: How CI Width Scales With N" |
> | Identifiability check | §"Phase 0", question 4 (r vs decisional temperature) |
> | Multilevel SBC | n/a — single-subject model only; multilevel deferred to Ch. 15 |

## Rethinking Category Representation: From Examples to Averages

In the previous chapter, we explored the Generalized Context Model (GCM), an exemplar model where every encountered example is stored in memory. This approach is powerful but raises questions about cognitive efficiency. Do we really store every single instance we encounter?

Consider learning about common categories like "birds" or "chairs". While specific examples matter, we also seem to develop a general sense of what constitutes a typical member. Prototype theory offers an alternative perspective: instead of storing individual exemplars, the mind abstracts a summary representation — a prototype — for each category. This prototype often represents the central tendency or "average" member.

## Core Ideas of Prototype Models

* **Abstraction, Not Storage**: The primary representation is a summary (the prototype), not a collection of instances.
* **Cognitive Economy**: Storing one prototype per category is potentially much more memory-efficient than storing numerous exemplars.
* **Similarity to Prototype**: New items are categorized based on their similarity to the stored prototypes of different categories.
* **Typicality**: Items closer to a category's prototype are judged as more "typical" members and are often categorized faster and more accurately. This contrasts with exemplar models, which emphasize detailed memory for specific instances.

| Feature | Prototype Models | Exemplar Models (GCM) |
|---|---|---|
| Representation | Abstract summary (e.g., average) | Collection of individual examples |
| Memory Cost | Low (one prototype per category) | High (potentially all examples) |
| Key Information | Central tendency & variance | Specific instances & their labels |
| Sensitivity | Less sensitive to specific examples | Highly sensitive to specific examples |

## Modeling Dynamic Prototypes: The Kalman Filter

Early prototype models often calculated a static average. However, human learning is dynamic; our understanding evolves with experience. How can we model a prototype that updates incrementally as new examples are encountered?

The Kalman filter provides a suitable mathematical framework. Originally developed for tracking physical systems amidst noise, it's well-suited for modeling prototype learning because it allows us to:

* **Track an Estimate**: The category prototype (its average feature values).
* **Represent Uncertainty**: Maintain not just the prototype's location but also our uncertainty about it.
* **Update Incrementally**: Refine the prototype with each new example.
* **Balance Old and New**: Optimally combine the current prototype (prior belief) with the new example (evidence).
* **Implement an Adaptive Learning Rate**: Automatically adjust how much influence a new example has (the "Kalman gain") based on current uncertainty. Learning is faster when uncertainty is high and slows as confidence grows.

### The Kalman Filter Logic: Updating an Estimate

Imagine tracking a prototype based on a single feature (e.g., average height). The Kalman filter maintains your Current State. At any moment, you have two key pieces of information:

* **Your Best Guess ($\mu$):** This is your current estimate of the prototype's average height.
* **Your Confidence ($\sigma^2$):** This isn't just *if* you're confident, but *how much*. It's the *variance* around your guess. A small variance means you're quite sure; a large variance means you're very unsure.

When a new category member (observation $x$) arrives, you measure its height.

* **The Catch:** Your measurement tool (or the example itself) isn't perfect. There's some inherent "noise" or variability in how examples relate to the true prototype. We represent the uncertainty of this *measurement process* with another variance, often called $R$.

**The Core Problem:** You have your guess (with $\sigma^2_{prev}$) and a new measurement (with $R$). How do you combine them to get a *new, better guess* and an *updated level of uncertainty*?

**Enter the Kalman Gain ($K$): The "Trust" Knob**

The Kalman filter calculates the Kalman Gain $K$, which tells you how much to trust the *new measurement* compared to your *current guess*:

$$K = \frac{\sigma^2_{prev}}{\sigma^2_{prev} + R}$$

If $\sigma^2_{prev}$ is large relative to $R$, then $K \to 1$: trust the measurement almost completely. If $\sigma^2_{prev}$ is small relative to $R$, then $K \to 0$: stick closer to your previous guess.

**Updating Your Guess:**

You nudge your old guess towards the new measurement. The size of the nudge is determined by the Kalman Gain $K$:

$$\mu_{new} = \mu_{prev} + K \cdot (x - \mu_{prev})$$

The term $(x - \mu_{prev})$ is the "prediction error" or "innovation" — how much the new data differs from what you expected.

**Updating Your Confidence (Uncertainty):**

After incorporating new information, uncertainty decreases:

$$\sigma^2_{new} = (1 - K) \cdot \sigma^2_{prev}$$

Since $K \in (0, 1)$, multiplying by $(1 - K)$ always shrinks uncertainty. The more you trusted the measurement (higher $K$), the more your uncertainty shrinks.

**Note**: This simple implementation assumes a static target (a true underlying prototype), so uncertainty monotonically decreases. In reality, the prototype may change over time, and more complex implementations introduce a **process noise term** $Q$. Before you get the next measurement, you predict how the system state might change over time. This prediction step increases uncertainty — because real-world processes aren't perfectly predictable — then the measurement update decreases it. If the state evolves as $x_k = F x_{k-1} + w_k$ (where $w_k$ is process noise with covariance $Q$), the predicted covariance becomes $P_{k|k-1} = F P_{k-1|k-1} F^T + Q$. The addition of $Q$ explicitly increases uncertainty between observations. We model categories as static targets here, which is appropriate for the blocked learning paradigm we use. We return to this commitment in §"Phase 0", question 2 (the Persistence Question), where we name it as a non-trivial cognitive choice rather than a technical default.

### Extending to Multiple Features: Multivariate Kalman Filter

Our stimuli have two features (height and position). We need a multivariate version:

* **Your Current State (Vectors and Matrices):**
    * **Your Best Guess ($\vec{\mu}$):** Now a *vector* containing your best guess for each feature.
    * **Your Confidence ($\Sigma$):** Now a *covariance matrix*. The diagonal elements give the variance for each feature individually; the off-diagonal elements give their covariance — for example, whether taller prototypes are also typically further to the right.

* **A New Hint Arrives:** Your new observation ($\vec{x}$) is also a vector.
    * **Measurement Noise ($R$):** Also a covariance matrix. Often simplified to a diagonal matrix, assuming measurement errors on different features are independent.

* **The Logic Stays the Same, the Math Uses Matrices:**
    * **Kalman Gain ($K$):** A *matrix*: $K = \Sigma_{prev} (\Sigma_{prev} + R)^{-1}$
    * **Updating Your Guess ($\vec{\mu}$):** $\vec{\mu}_{new} = \vec{\mu}_{prev} + K (\vec{x} - \vec{\mu}_{prev})$
    * **Updating Your Confidence ($\Sigma$):** We use the numerically stable Joseph form: $\Sigma_{new} = (I - K)\Sigma_{prev}(I - K)^T + KRK^T$. While it looks complex, the intuition is the same as the scalar case: uncertainty decreases after each update. The Joseph form is preferred because it guarantees the result remains symmetric and positive semi-definite even in the presence of floating-point rounding errors.

### In a Nutshell

The Kalman filter is a recursive algorithm that provides the mathematically optimal way to estimate the state of a system (like the features of a prototype) given a series of noisy measurements. It works by:

1. Maintaining an estimate ($\vec{\mu}$) and uncertainty ($\Sigma$).
2. Calculating a Kalman Gain $K$ that balances trust in the current estimate against trust in the new measurement.
3. Updating the estimate by moving it towards the new measurement, weighted by $K$.
4. Updating (reducing) the uncertainty based on the information gained.

The multivariate version simply uses vectors and matrices to do this for multiple, potentially correlated variables simultaneously.

### Implementing the Multivariate Update in R

```{r, 12 multivariate_kalman_update}
# One update step of a multivariate Kalman filter.
# Returns updated mean vector, covariance matrix, and Kalman gain.
multivariate_kalman_update <- function(mu_prev,      # previous mean vector
                                       sigma_prev,    # previous covariance matrix
                                       observation,   # observed feature vector
                                       r_matrix) {    # observation noise matrix
  mu_prev     <- as.numeric(mu_prev)
  observation <- as.numeric(observation)
  sigma_prev  <- as.matrix(sigma_prev)
  r_matrix    <- as.matrix(r_matrix)

  n_dim <- length(mu_prev)
  I     <- diag(n_dim)

  # Kalman gain: K = Sigma_prev * (Sigma_prev + R)^{-1}
  S     <- sigma_prev + r_matrix
  S_inv <- tryCatch(
    solve(S),
    error = function(e) {
      warning("Matrix inversion failed; using pseudo-inverse.", call. = FALSE)
      MASS::ginv(S)
    }
  )
  K <- sigma_prev %*% S_inv

  # Update mean
  innovation <- observation - mu_prev
  mu_new     <- as.numeric(mu_prev + K %*% innovation)

  # Update covariance — Joseph form for numerical stability
  IK        <- I - K
  sigma_new <- IK %*% sigma_prev %*% t(IK) + K %*% r_matrix %*% t(K)

  # Enforce symmetry (floating-point can introduce tiny asymmetry)
  sigma_new <- (sigma_new + t(sigma_new)) / 2

  list(mu = mu_new, sigma = sigma_new, k = K)
}
```

This function performs a single update for one category's prototype when it observes a new member.

---


## Phase 0: Interrogating the Prototype Model

Before we build the Stan model, we pause to ask the questions Chapter 11 introduced as the Phase 0 interrogation. These are not stylistic preliminaries — they are the questions whose answers determine which Stan model is the *correct* operationalisation of the verbal theory. Skipping them is how cognitive models accumulate hidden commitments.

We answer four questions for the Kalman prototype model.

> **Q1 — The First-Trial Question.** *On trial 1, before any feedback has been seen, what does the agent do?*
>
> In our model, both prototypes are initialised at $\vec{\mu}_0 = \vec{\mu}_1 = (2.5, 2.5)$ with isotropic covariance $\Sigma_0 = \Sigma_1 = 10 \cdot I$. The decision rule on trial 1 therefore compares two **identical** multivariate normal densities at the observed stimulus. Their log-densities are equal, the softmax reduces to $0.5$, and the agent chooses Category 1 with probability exactly $0.5$ regardless of which stimulus is presented.
>
> **This is a cognitive commitment, not a technical default.** The model is claiming that humans, encountering the very first stimulus of a categorisation task with no prior information, are perfectly indifferent. There is no asymmetric category prior, no perceptual fluency bias, no preference for one button over the other. A reader who runs this model on a task with strong baseline asymmetries (e.g., one category much more frequent than the other in the participant's life experience) should know that *trial 1's coin flip is structural*. If you wanted to relax it, the cleanest extension is a separate `bias` parameter on the logit scale (as Chapter 11 implements) — but that adds an additional source of conflation with the noise parameter, which is the subject of question 4 below.
>
> **What we commit to here:** No category bias. Trial 1 is by construction $P(\text{cat 1}) = 0.5$. Section "Limitations" returns to this.

> **Q2 — The Persistence Question.** *Once the prototypes are well-estimated, does the agent forget? Does the model ever reset?*
>
> The Kalman update we have implemented is for a **static target**. There is no process noise term $Q$, so once $\Sigma_t$ shrinks it cannot grow back. Mathematically, $\det(\Sigma_t)$ is monotonically non-increasing along the trial sequence. Cognitively, this means: **once the agent has seen enough exemplars of a category, learning effectively halts**, and any new exemplar is integrated only weakly into the prototype regardless of how surprising it is. There is no mechanism for forgetting, no re-opening of category boundaries when the input distribution shifts, and no decay of confidence over time.
>
> This is a defensible commitment for the Kruschke (1993) blocked-learning paradigm we use, where the category structure is genuinely static and the experiment is short. It is *not* a defensible commitment for category-learning over a lifespan, for category drift studies, or for any task where the input distribution changes mid-experiment. For those settings, the principled extension is to add a process-noise covariance $Q$ to the prediction step before each measurement update, which lets uncertainty grow again between observations. We do not implement this here, but the limitation is real and named.
>
> **What we commit to here:** Static categories. No forgetting. The asymptotic behaviour is one-shot Bayesian convergence to a fixed prototype.

> **Q3 — The Observation-vs-Decision Noise Question.** *Where does response stochasticity come from?*
>
> Our model has a single noise parameter, $r$, sold to the reader as the "observation noise variance" — the variability in how individual exemplars relate to their category's true prototype. But $r$ does not only affect the Kalman update. It also enters the *decision* step through the term $(\Sigma_C + R)^{-1}$ in the multivariate normal density. This means $r$ simultaneously controls:
>
> 1. **How fast the prototype absorbs new evidence** (smaller $r$ → faster learning, sharper Kalman gain)
> 2. **How sharply the agent discriminates between categories at decision time** (smaller $r$ → more peaked likelihoods → more deterministic softmax)
>
> A more flexible model would separate these by adding a softmax temperature $\tau$:
> $$P(\text{cat 1} \mid \vec{x}) = \frac{\left[P(\vec{x} \mid \text{cat 1})\right]^{1/\tau}}{\left[P(\vec{x} \mid \text{cat 0})\right]^{1/\tau} + \left[P(\vec{x} \mid \text{cat 1})\right]^{1/\tau}}$$
> with $\tau$ controlling decisional precision independently of the Kalman update. The honest answer to whether $r$ and $\tau$ are jointly identifiable from learning curves alone is: **partially.** Early-trial behaviour is dominated by learning speed (which is driven by $r$ via the Kalman gain), and asymptotic accuracy is dominated by decisional precision (which both $r$ and $\tau$ affect). If you have many trials at every stage of learning, the trade-off is identifiable. If you only see the asymptote, $r$ and $\tau$ are confounded.
>
> **What we commit to here:** A single $r$ that conflates observational and decisional noise. The model is identifiable from learning *trajectories* (early vs. late accuracy) but not from asymptotic accuracy alone. We will return to this in §"Limitations".

> **Q4 — The Identifiability Question.** *Are there other parameters we are silently fixing?*
>
> Three structural choices ride along with this model that a Phase 0 interrogation should surface:
>
> 1. **Initial mean** $\vec{\mu}_0 = \vec{\mu}_1 = (2.5, 2.5)$. Centre of the feature space. Implicit prior on prototype location. Could be made into a parameter; we fix it.
> 2. **Initial covariance** $\Sigma_0 = \Sigma_1 = 10 \cdot I$. Large enough that the first observation dominates the first update (Kalman gain ≈ $10/(10+r)$ ≈ 1 for moderate $r$). This makes the model "open-minded" at trial 1, but the specific value of 10 is a free choice that we do not infer.
> 3. **Diagonal $R$**. We assume measurement noise is independent across features. A real perceptual system might have correlated noise (e.g., position and height judgments could share an attention-allocation noise source).
>
> Each of these choices is a structural prior. None of them is inferred from data. The reader should know they are choices.

**Summary of Phase 0 commitments.** We are fitting a *one-parameter* model whose sole free parameter $r$ controls both learning speed and decisional precision, with no category bias, no forgetting, no process noise, fixed initial means, fixed initial covariances, and a diagonal observation noise structure. This is a deliberately simple model and we have written down what is buried inside it. The Stan implementation in §"The Prototype Model in Stan" implements exactly this.

---


## Building the Full Categorization Model

We now integrate this update mechanism into a full agent that learns prototypes for two categories and makes categorization decisions.

### Initialization

* We need separate prototypes ($\mu_0, \Sigma_0$ and $\mu_1, \Sigma_1$) for Category 0 and Category 1.
* We start with high uncertainty (large initial $\Sigma$) and uninformative means at the center of the feature space.
* We define the observation noise matrix $R = r_\text{val} \cdot I$. The parameter $r_\text{val}$ represents how much variability we assume exists within categories or in our perception of stimuli.

### Categorization Decision

How does the model decide the category for a new stimulus $\vec{x}$? It compares $\vec{x}$ to both prototypes, considering their current means ($\vec{\mu}$) and uncertainties ($\Sigma$). We use the probability density of the multivariate normal distribution:

$$P(\vec{x} \mid \text{Category } C) \propto \exp\left(-\frac{1}{2}(\vec{x} - \vec{\mu}_C)^T (\Sigma_C + R)^{-1} (\vec{x} - \vec{\mu}_C)\right)$$

The probability of choosing Category 1 is then calculated using the Luce choice rule (softmax):

$$P(\text{Choose Cat 1} \mid \vec{x}) = \frac{P(\vec{x} \mid \text{Category } 1)}{P(\vec{x} \mid \text{Category } 0) + P(\vec{x} \mid \text{Category } 1)}$$

We work with log-probabilities for numerical stability. Note (per Phase 0, Q3) that the same $r$ that drives the Kalman update also enters here through $(\Sigma_C + R)^{-1}$ — there is no separate softmax temperature.

### The Learning Loop

The agent processes trials sequentially:

1. Observe stimulus $\vec{x}_i$.
2. Calculate $P(\text{Choose Cat 1} \mid \vec{x}_i)$ based on current $\mu_0, \Sigma_0, \mu_1, \Sigma_1$.
3. Generate a response.
4. Receive feedback (true category $y_i$).
5. Update the prototype of the correct category using the multivariate Kalman update.

### R Implementation of the Prototype Agent

The function now accepts `subject_seed` and calls `make_subject_schedule` internally — consistent with the path-dependence fix from Ch. 11. Each agent's learning trajectory depends on the order in which stimuli are encountered, and different randomizations produce genuinely different trajectories.

```{r, 12 prototype_model}
# Prototype model agent using Kalman filter for categorization.
# r_value   : observation noise variance (scalar > 0)
# obs       : matrix of observations (trials x features)
# cat_one   : vector of true category labels (0 or 1) for feedback
# Returns:  tibble with trial_within_subject, sim_response, prob_cat1,
#           correct, and performance columns alongside schedule columns.
prototype_kalman <- function(r_value,
                             obs,
                             cat_one,
                             initial_mu         = NULL,
                             initial_sigma_diag = 10.0,
                             quiet              = TRUE) {
  n_trials  <- nrow(obs)
  n_features <- ncol(obs)

  if (is.null(initial_mu)) {
    mu0_init <- rep(2.5, n_features)
    mu1_init <- rep(2.5, n_features)
  } else {
    mu0_init <- initial_mu[[1]]
    mu1_init <- initial_mu[[2]]
  }

  prototype_cat_0 <- list(mu = mu0_init, sigma = diag(initial_sigma_diag, n_features))
  prototype_cat_1 <- list(mu = mu1_init, sigma = diag(initial_sigma_diag, n_features))
  r_matrix        <- diag(r_value, n_features)

  response_probs <- numeric(n_trials)

  log_sum_exp <- function(v) {
    m <- max(v)
    m + log(sum(exp(v - m)))
  }

  for (i in seq_len(n_trials)) {
    if (!quiet && i %% 20 == 0) cat("Trial", i, "\n")

    current_obs <- as.numeric(obs[i, ])

    # ── Decision ────────────────────────────────────────────────────────────
    cov_cat_0 <- prototype_cat_0$sigma + r_matrix
    cov_cat_1 <- prototype_cat_1$sigma + r_matrix

    log_prob_0 <- tryCatch(
      mvtnorm::dmvnorm(current_obs, mean = prototype_cat_0$mu,
                       sigma = cov_cat_0, log = TRUE),
      error = function(e) -Inf
    )
    log_prob_1 <- tryCatch(
      mvtnorm::dmvnorm(current_obs, mean = prototype_cat_1$mu,
                       sigma = cov_cat_1, log = TRUE),
      error = function(e) -Inf
    )

    if (!is.finite(log_prob_0) && !is.finite(log_prob_1)) {
      prob_cat_1 <- 0.5
    } else if (!is.finite(log_prob_0)) {
      prob_cat_1 <- 1.0
    } else if (!is.finite(log_prob_1)) {
      prob_cat_1 <- 0.0
    } else {
      # Phase 0 Q1: at trial 1 both prototypes are identical, so prob_cat_1 = 0.5
      # by construction. No explicit bias parameter in the base model.
      prob_cat_1 <- exp(log_prob_1 - log_sum_exp(c(log_prob_0, log_prob_1)))
    }
    response_probs[i] <- pmax(1e-9, pmin(1 - 1e-9, prob_cat_1))

    # ── Update ───────────────────────────────────────────────────────────────
    true_cat <- cat_one[i]
    if (true_cat == 1) {
      upd <- multivariate_kalman_update(prototype_cat_1$mu, prototype_cat_1$sigma,
                                        current_obs, r_matrix)
      prototype_cat_1$mu    <- upd$mu
      prototype_cat_1$sigma <- upd$sigma
    } else {
      upd <- multivariate_kalman_update(prototype_cat_0$mu, prototype_cat_0$sigma,
                                        current_obs, r_matrix)
      prototype_cat_0$mu    <- upd$mu
      prototype_cat_0$sigma <- upd$sigma
    }
  }

  tibble(
    prob_cat1    = response_probs,
    sim_response = rbinom(n_trials, 1, response_probs)
  )
}

# Wrapper: generates per-agent schedule then calls prototype_kalman
simulate_prototype_agent <- function(agent_id, r_value,
                                     stimulus_info, n_blocks, subject_seed) {
  schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  obs      <- as.matrix(schedule[, c("height", "position")])
  cat_one  <- schedule$category_feedback

  result <- prototype_kalman(
    r_value            = r_value,
    obs                = obs,
    cat_one            = cat_one,
    initial_mu         = list(rep(2.5, 2), rep(2.5, 2)),
    initial_sigma_diag = 10.0,
    quiet              = TRUE
  )

  schedule |>
    mutate(
      agent_id     = agent_id,
      r_value_true = r_value,
      log_r_true   = log(r_value),
      prob_cat1    = result$prob_cat1,
      sim_response = result$sim_response,
      correct      = as.integer(category_feedback == sim_response)
    ) |>
    group_by(agent_id) |>
    mutate(performance = cumsum(correct) / row_number()) |>
    ungroup()
}
```

### Simulating Categorization Behavior

Let's simulate behavior on the Kruschke (1993) task. The key parameter here is `r_value`, the observation noise. This parameter reflects the model's assumption about how much variability exists within categories or arises from perceptual noise.

* **Low `r_value`**: The model assumes observations are precise representations of the category; each new example exerts strong influence and prototypes update rapidly.
* **High `r_value`**: The model assumes observations are noisy; learning is slower as each example has less influence, but the model generalizes more broadly.

```{r, 12 simulate_prototype_responses}
param_df <- expand_grid(
  agent_id  = 1:10,
  r_value   = c(0.1, 0.5, 1.0, 2.0, 5.0)
) |>
  mutate(subject_seed = agent_id)    # unique, reproducible seed per agent

sim_file <- "simdata/W12_prototype_simulated_responses.csv"
if (regenerate_simulations || !file.exists(sim_file)) {
  cat("Regenerating prototype simulations...\n")
  prototype_responses <- future_pmap_dfr(
    list(
      agent_id     = param_df$agent_id,
      r_value      = param_df$r_value,
      subject_seed = param_df$subject_seed
    ),
    simulate_prototype_agent,
    stimulus_info = stimulus_info,
    n_blocks      = n_blocks,
    .options = furrr_options(seed = TRUE)
  )
  write_csv(prototype_responses, sim_file)
  cat("Simulations saved.\n")
} else {
  prototype_responses <- read_csv(sim_file, show_col_types = FALSE)
  cat("Simulations loaded.\n")
}

ggplot(prototype_responses,
       aes(x = trial_within_subject, y = performance, color = factor(r_value))) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.15,
               aes(fill = factor(r_value))) +
  scale_color_viridis_d(option = "plasma", name = "r_value") +
  scale_fill_viridis_d(option = "plasma", name = "r_value") +
  labs(
    title    = "Prototype Model Learning Performance",
    subtitle = "Effect of Observation Noise (r_value); mean ± SE across agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  ylim(0.4, 1.0)
```

**Interpretation**: Different values of `r_value` impact learning speed and asymptotic accuracy. Very low noise (`r_value = 0.1`) means each example exerts near-complete influence on the prototype, which can lead to rapid initial learning but sensitivity to early idiosyncrasies. Very high noise (`r_value = 5.0`) means the model trusts observations much less and learns more slowly. An intermediate value often achieves the best balance between responsiveness and robustness. Note also (Phase 0 Q3) that *all five curves start at chance* — this is the structural Trial-1 behaviour we committed to.


### Visualizing Prototype Evolution

Where do the prototypes end up? Let's track their movement and uncertainty over trials.

```{r, 12 track_prototypes}
# Track prototype means and covariance matrices over the course of learning.
# Returns a tibble with one row per (trial × category).
track_prototypes <- function(r_value, obs, cat_one,
                             initial_mu = NULL, initial_sigma_diag = 10.0) {
  n_trials  <- nrow(obs)
  n_features <- ncol(obs)

  if (is.null(initial_mu)) {
    mu0_init <- rep(2.5, n_features)
    mu1_init <- rep(2.5, n_features)
  } else {
    mu0_init <- initial_mu[[1]]
    mu1_init <- initial_mu[[2]]
  }
  prototype_cat_0 <- list(mu = mu0_init, sigma = diag(initial_sigma_diag, n_features))
  prototype_cat_1 <- list(mu = mu1_init, sigma = diag(initial_sigma_diag, n_features))
  r_matrix        <- diag(r_value, n_features)

  history <- list()

  for (i in seq_len(n_trials + 1)) {
    history[[length(history) + 1]] <- tibble(
      trial      = i, category = 0,
      feature1_mean = prototype_cat_0$mu[1],
      feature2_mean = prototype_cat_0$mu[2],
      cov_matrix    = list(prototype_cat_0$sigma)
    )
    history[[length(history) + 1]] <- tibble(
      trial      = i, category = 1,
      feature1_mean = prototype_cat_1$mu[1],
      feature2_mean = prototype_cat_1$mu[2],
      cov_matrix    = list(prototype_cat_1$sigma)
    )

    if (i <= n_trials) {
      current_obs <- as.numeric(obs[i, ])
      true_cat    <- cat_one[i]
      if (true_cat == 1) {
        upd <- multivariate_kalman_update(prototype_cat_1$mu, prototype_cat_1$sigma,
                                          current_obs, r_matrix)
        prototype_cat_1$mu    <- upd$mu
        prototype_cat_1$sigma <- upd$sigma
      } else {
        upd <- multivariate_kalman_update(prototype_cat_0$mu, prototype_cat_0$sigma,
                                          current_obs, r_matrix)
        prototype_cat_0$mu    <- upd$mu
        prototype_cat_0$sigma <- upd$sigma
      }
    }
  }
  bind_rows(history)
}

# Ellipse helper for uncertainty visualization
get_ellipse <- function(mu, sigma, level = 0.68) {
  mu    <- as.numeric(mu)
  sigma <- as.matrix(sigma)
  if (length(mu) != 2 || !all(dim(sigma) == c(2, 2))) return(NULL)
  ev <- eigen(sigma, symmetric = TRUE, only.values = TRUE)$values
  if (any(ev <= 1e-6)) sigma <- sigma + diag(ncol(sigma)) * 1e-6
  pts <- tryCatch(
    ellipse::ellipse(sigma, centre = mu, level = level),
    error = function(e) NULL
  )
  if (is.null(pts)) return(NULL)
  as_tibble(pts) |> setNames(c("feature1_mean", "feature2_mean"))
}

# Use a fixed seed-1 schedule for the trajectory visualization
traj_schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = 1)

prototype_trajectory <- track_prototypes(
  r_value            = 1.0,
  obs                = as.matrix(traj_schedule[, c("height", "position")]),
  cat_one            = traj_schedule$category_feedback,
  initial_mu         = list(rep(2.5, 2), rep(2.5, 2)),
  initial_sigma_diag = 10.0
)

final_prototypes    <- prototype_trajectory |> filter(trial == max(trial))

ellipse_data <- final_prototypes |>
  rowwise() |>
  mutate(ell = list(get_ellipse(c(feature1_mean, feature2_mean), cov_matrix[[1]]))) |>
  ungroup() |>
  filter(!map_lgl(ell, is.null)) |>
  unnest(ell)

ggplot() +
  geom_point(data = stimulus_info,
             aes(x = position, y = height, color = factor(category_true),
                 shape = factor(category_true)),
             size = 3, alpha = 0.5) +
  geom_path(
    data = prototype_trajectory |> filter(trial <= total_trials),
    aes(x = feature2_mean, y = feature1_mean,
        group = category, color = factor(category)),
    linetype = "dashed",
    arrow = arrow(type = "closed", length = unit(0.08, "inches"), angle = 20)
  ) +
  geom_point(data = final_prototypes,
             aes(x = feature2_mean, y = feature1_mean, color = factor(category)),
             size = 4, shape = 18) +
  geom_path(data = ellipse_data,
            aes(x = feature2_mean, y = feature1_mean,
                group = category, color = factor(category)),
            alpha = 0.7, linewidth = 0.8) +
  scale_color_manual(values = c("0" = "#0072B2", "1" = "#D55E00"), name = "Category") +
  scale_shape_manual(values = c("0" = 16, "1" = 17), name = "Category") +
  labs(
    title    = "Prototype Learning Trajectory (Kalman Filter, r = 1.0)",
    subtitle = "Prototypes start at (2.5, 2.5). Dashed lines show path; ellipses show final 68% uncertainty.",
    x = "Position Feature", y = "Height Feature"
  ) +
  coord_cartesian(
    xlim = range(c(stimulus_info$position, prototype_trajectory$feature2_mean)) + c(-0.5, 0.5),
    ylim = range(c(stimulus_info$height,   prototype_trajectory$feature1_mean)) + c(-0.5, 0.5)
  )
```

**Interpretation**: The dashed lines show the path the prototypes took during learning (starting from (2.5, 2.5), the center of the feature space). The final prototype locations are marked with diamonds, and the ellipses represent the model's final uncertainty (68% confidence) about the true prototype location for each category. We see the prototypes moving towards the center of their respective category members and uncertainty shrinking over time — exactly the incremental, variance-reducing Bayesian update the Kalman filter implements. The shrinking ellipses are the visual signature of Phase 0 Q2: once the agent has seen enough exemplars, the covariances are small enough that further learning is negligible, and the prototypes are effectively frozen.

---

## Prior Predictive Check

Before fitting any data, we verify that the prior on `log_r` generates only scientifically plausible learning trajectories. We draw $S = 500$ values from $\log r \sim \text{Normal}(0, 1)$ — implying $r \in (0.13, 7.4)$ at $\pm 2$ SD — simulate one agent per draw, and plot the envelope.

```{r, 12 prior predictive check}
n_ppc_samples <- 500
set.seed(42)

ppc_log_r  <- rnorm(n_ppc_samples, LOG_R_PRIOR_MEAN, LOG_R_PRIOR_SD)
ppc_r      <- exp(ppc_log_r)
ppc_sched  <- make_subject_schedule(stimulus_info, n_blocks, seed = 999)

prior_pred_curves <- map_dfr(seq_len(n_ppc_samples), function(s) {
  res <- prototype_kalman(
    r_value   = ppc_r[s],
    obs       = as.matrix(ppc_sched[, c("height", "position")]),
    cat_one   = ppc_sched$category_feedback,
    initial_mu = list(rep(2.5, 2), rep(2.5, 2))
  )
  tibble(
    sample_id = s,
    trial     = seq_len(nrow(ppc_sched)),
    correct   = as.integer(res$sim_response == ppc_sched$category_feedback)
  ) |>
    mutate(cum_acc = cumsum(correct) / row_number())
})

ppc_summary <- prior_pred_curves |>
  group_by(trial) |>
  summarise(
    q05 = quantile(cum_acc, 0.05), q25 = quantile(cum_acc, 0.25),
    q50 = median(cum_acc),
    q75 = quantile(cum_acc, 0.75), q95 = quantile(cum_acc, 0.95),
    .groups = "drop"
  )

ggplot(ppc_summary, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#009E73", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#009E73", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#006D4E", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Prior Predictive Check: Kalman Filter Prototype Model",
    subtitle = "Ribbons: 50% and 90% prior predictive intervals\nPrior: log(r) ~ Normal(0, 1)",
    x = "Trial", y = "Cumulative Accuracy"
  )
```

The prior predictive envelope spans chance performance (uninformative $r$ draws) through near-ceiling accuracy (informative $r$ draws), without implying impossible behavior such as sustained below-chance performance. This is the plausibility standard we require before proceeding to inference.

---


## The Prototype Model in Stan

Simulations help us understand the model, but we want to fit it to experimental data to estimate its parameters. The primary free parameter is the observation noise variance `r_value`. We implement the Kalman filter prototype model in Stan to estimate `r_value` (on the log scale) from data.

### Why the Kalman state lives in `transformed parameters`

The prototype model has only one parameter — the scalar `log_r`. Given `log_r` and the observed data (stimuli and feedback), the *entire* Kalman filter state is deterministic: prototype means, prototype covariances, and the per-trial choice probability $p_i$ all follow algebraically from `log_r` and the trial sequence. Whenever a quantity depends on a parameter but is otherwise deterministic, the persona's Phase 2.1 rule says it belongs in `transformed parameters` (evaluated once per leapfrog step) rather than `transformed data` (which is for parameter-free preprocessing) or `model{}` (which would force a recomputation in `generated quantities`). We follow that rule here, and `log_lik[i]` in `generated quantities` becomes a one-line lookup of `p[i]`.

> **A note on architectural symmetry with Chapter 11.** Earlier drafts of this chapter described placing `p` in `transformed parameters` as an "architectural advantage of the Kalman formulation over the GCM." That framing was wrong. The Chapter 11 GCM's choice probabilities are *also* deterministic given the parameters and the data, and they *also* belong in `transformed parameters` for the same reason. The revised Chapter 11 has been refactored to do exactly this: `prob_cat1[i]` is a transformed parameter, the model block reduces to a single vectorised `bernoulli_lpmf(y | prob_cat1)`, and `log_lik[i]` is a one-line lookup. The two chapters now share the same Stan-block discipline. The Kalman filter has no special architectural status — it just illustrates the principle particularly cleanly because the only parameter is a scalar.

The same path-dependence concern from Ch. 11 still applies: trial $t$'s choice probability depends on all previous stimuli through the Kalman state, so standard PSIS-LOO is invalid. Unlike previous drafts, we no longer treat that as a caveat to repeat — the LFO-CV machinery from Chapter 11 is generalised to the Kalman case in §"Leave-Future-Out Cross-Validation" below, and the comparison against exact one-step-ahead refits is reported in this chapter rather than deferred to a hypothetical future supplement.

> **`reduce_sum` and parallelization**: As with the GCM, the Kalman prototype model cannot use `reduce_sum` because successive Kalman steps are causally chained — trial $t$'s contribution to the likelihood depends on the filter state from trials $1, \ldots, t-1$. The entire filter must be computed sequentially. However, because the filter state is deterministic and cheap to compute (only two $2 \times 2$ matrix operations per trial), the model is already fast without parallelization.

### Stan Model

```{r, 12 stan_prototype_model}
prototype_single_stan <- "
// Kalman Filter Prototype Model — Single Subject
// Key design:
//   log_r is the only parameter (unconstrained); r_value = exp(log_r)
//   The full Kalman filter sequence (prototype states + choice probs)
//   is computed in transformed parameters — a deterministic function
//   of log_r and the data. This means log_lik in generated quantities
//   simply reads p[i] from transformed parameters without re-running
//   the filter. Architecturally identical to the GCM in Ch 11
//   (where prob_cat1[i] also lives in transformed parameters).

data {
  int<lower=1> ntrials;
  int<lower=1> nfeatures;
  array[ntrials] int<lower=0, upper=1> cat_one;   // true category (feedback)
  array[ntrials] int<lower=0, upper=1> y;          // observed responses
  array[ntrials, nfeatures] real obs;              // stimulus features

  // Fixed structural values (passed as data for transparency).
  // See Phase 0 Q4: these are NOT inferred — they are commitments.
  vector[nfeatures] initial_mu_cat0;
  vector[nfeatures] initial_mu_cat1;
  real<lower=0> initial_sigma_diag;

  // Prior hyperparameters (lifted to data block per persona Phase 2.1)
  // so prior sweeps don't require recompilation.
  real prior_logr_mean;
  real<lower=0> prior_logr_sd;
}

parameters {
  real log_r;
}

transformed parameters {
  real<lower=0> r_value = exp(log_r);

  // Choice probability at each trial, computed via the Kalman filter.
  // Stored as a transformed parameter so generated quantities can reference
  // it directly without re-running the filter.
  array[ntrials] real<lower=1e-9, upper=1-1e-9> p;

  {
    vector[nfeatures] mu_cat0 = initial_mu_cat0;
    vector[nfeatures] mu_cat1 = initial_mu_cat1;
    matrix[nfeatures, nfeatures] sigma_cat0 =
      diag_matrix(rep_vector(initial_sigma_diag, nfeatures));
    matrix[nfeatures, nfeatures] sigma_cat1 =
      diag_matrix(rep_vector(initial_sigma_diag, nfeatures));
    matrix[nfeatures, nfeatures] r_matrix =
      diag_matrix(rep_vector(r_value, nfeatures));
    matrix[nfeatures, nfeatures] I_mat =
      diag_matrix(rep_vector(1.0, nfeatures));

    for (i in 1:ntrials) {
      vector[nfeatures] x = to_vector(obs[i]);

      // ── Decision (before this trial's update) ──────────────────────────
      // Phase 0 Q1: at i==1, mu_cat0 == mu_cat1 and sigma_cat0 == sigma_cat1,
      // so log_p0 == log_p1 and prob1 == 0.5 by construction.
      matrix[nfeatures, nfeatures] cov0 = sigma_cat0 + r_matrix;
      matrix[nfeatures, nfeatures] cov1 = sigma_cat1 + r_matrix;

      real log_p0 = multi_normal_lpdf(x | mu_cat0, cov0);
      real log_p1 = multi_normal_lpdf(x | mu_cat1, cov1);
      real prob1  = exp(log_p1 - log_sum_exp(log_p0, log_p1));

      p[i] = fmax(1e-9, fmin(1 - 1e-9, prob1));

      // ── Update (on all trials, including the last) ──────────────────────
      if (cat_one[i] == 1) {
        vector[nfeatures] innov = x - mu_cat1;
        matrix[nfeatures, nfeatures] S  = sigma_cat1 + r_matrix;
        // K = Sigma * S^{-1}; mdivide_right_spd is numerically stable
        matrix[nfeatures, nfeatures] K  = mdivide_right_spd(sigma_cat1, S);
        matrix[nfeatures, nfeatures] IK = I_mat - K;
        mu_cat1    = mu_cat1 + K * innov;
        sigma_cat1 = IK * sigma_cat1 * IK' + K * r_matrix * K';
        sigma_cat1 = 0.5 * (sigma_cat1 + sigma_cat1');
      } else {
        vector[nfeatures] innov = x - mu_cat0;
        matrix[nfeatures, nfeatures] S  = sigma_cat0 + r_matrix;
        matrix[nfeatures, nfeatures] K  = mdivide_right_spd(sigma_cat0, S);
        matrix[nfeatures, nfeatures] IK = I_mat - K;
        mu_cat0    = mu_cat0 + K * innov;
        sigma_cat0 = IK * sigma_cat0 * IK' + K * r_matrix * K';
        sigma_cat0 = 0.5 * (sigma_cat0 + sigma_cat0');
      }
    }
  }
}

model {
  target += normal_lpdf(log_r | prior_logr_mean, prior_logr_sd);
  target += bernoulli_lpmf(y | p);
}

generated quantities {
  // p[i] is already available from transformed parameters — no re-computation.
  vector[ntrials] log_lik;
  for (i in 1:ntrials)
    log_lik[i] = bernoulli_lpmf(y[i] | p[i]);
}
"

stan_file_proto_single <- "stan/W12_prototype_single.stan"
write_stan_file(prototype_single_stan, dir = "stan/",
                basename = "W12_prototype_single.stan")
mod_prototype_single <- cmdstan_model(stan_file_proto_single)
cat("Stan model compiled successfully.\n")
```

**Key aspects of the Stan implementation:**

* **Parameters block**: A single scalar `log_r`. The constraint `r_value > 0` is handled implicitly by `exp(log_r)`.
* **Transformed parameters**: Runs the full Kalman filter trial-by-trial, updating the two prototype distributions and recording `p[i]` — the probability of choosing Category 1 on trial $i$ — before each trial's update. The update happens on *all* trials (including the last), consistent with the R simulation.
* **Model block**: Prior on `log_r` plus a vectorised Bernoulli likelihood using the pre-computed `p` vector. One line.
* **Generated quantities**: `log_lik[i]` reads directly from `p[i]` — no re-computation of the Kalman filter required.
* **Phase 0 commitments are implemented in code**, not buried: the trial-1 coin flip is structural (identical initial means and covariances), there is no process noise (static target), and the same `r_value` is used in both the Kalman update and the decision step (single noise parameter conflating observational and decisional noise).

---


## Fitting the Model to Simulated Data

We use Pathfinder initialization before HMC sampling, as mandated for nonlinear cognitive models:

```{r, 12 fit single agent}
# Select one simulated agent with r_value = 1.0
agent_to_fit <- prototype_responses |>
  filter(r_value_true == 1.0, agent_id == 1)

stopifnot(nrow(agent_to_fit) == total_trials)

proto_data_single <- list(
  ntrials            = nrow(agent_to_fit),
  nfeatures          = 2L,
  cat_one            = agent_to_fit$category_feedback,
  y                  = agent_to_fit$sim_response,
  obs                = as.matrix(agent_to_fit[, c("height", "position")]),
  initial_mu_cat0    = c(2.5, 2.5),
  initial_mu_cat1    = c(2.5, 2.5),
  initial_sigma_diag = 10.0,
  prior_logr_mean    = LOG_R_PRIOR_MEAN,
  prior_logr_sd      = LOG_R_PRIOR_SD
)

fit_filepath_single <- "simmodels/W12_proto_single_fit.rds"
if (regenerate_simulations) {
  pf_single <- mod_prototype_single$pathfinder(
    data = proto_data_single, seed = 123, num_paths = 4, refresh = 0
  )

  fit_proto_single <- mod_prototype_single$sample(
    data            = proto_data_single,
    init            = pf_single,
    seed            = 123,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1000,
    iter_sampling   = 1500,
    refresh         = 300,
    adapt_delta     = 0.9
  )
  fit_proto_single$save_object(fit_filepath_single)
  cat("Single-agent prototype model fitted and saved.\n")
} else {
  if (file.exists(fit_filepath_single)) {
    fit_proto_single <- readRDS(fit_filepath_single)
    cat("Loaded existing single-agent prototype fit.\n")
  } else {
    cat("Fit file not found. Set regenerate_simulations = TRUE.\n")
    fit_proto_single <- NULL
  }
}

if (!is.null(fit_proto_single)) {
  param_summary <- fit_proto_single$summary(variables = c("log_r", "r_value"))
  print(param_summary)

  true_r     <- agent_to_fit$r_value_true[1]
  true_log_r <- log(true_r)
  cat("\nTrue log_r =", round(true_log_r, 3), "  True r =", true_r, "\n")
}
```

### Mandatory MCMC Diagnostic Battery

Before reading anything off the posterior, we curate the mandatory diagnostic table from the persona. Earlier drafts of this chapter ran `cmdstan_diagnose()` silently and moved on. The point of running diagnostics is *seeing* them — even when (as here, with one parameter) they are nearly trivial. The `diagnostic_summary_table()` helper defined in the setup chunk is the canonical version shared across Chapters 11, 12, 13, and 15. It returns a one-row tibble and emits a warning if any of the six thresholds are breached.

```{r, 12 diagnostic battery single}
if (!is.null(fit_proto_single)) {
  diag_tbl <- diagnostic_summary_table(
    fit_proto_single,
    pars = c("log_r", "r_value")
  )
  print(diag_tbl)

  cat("\nThreshold reference (persona Phase 3.0):\n",
      "  divergences        = 0\n",
      "  rank-normalized Rhat < 1.01\n",
      "  bulk ESS           > 400\n",
      "  tail ESS           > 400\n",
      "  E-BFMI             > 0.2\n",
      "  MCSE_mean / SD     < 0.05\n", sep = "")
}
```

With one parameter, a `mcmc_pairs` plot is degenerate. The geometry-relevant diagnostic for a one-dimensional posterior is the trace plot grid plus the rank histogram — both should look uniform across chains.

```{r, 12 trace and rank}
if (!is.null(fit_proto_single)) {
  p_trace <- bayesplot::mcmc_trace(
    fit_proto_single$draws(c("log_r", "r_value")),
    facet_args = list(ncol = 2)
  ) +
    ggtitle("Trace plots — should look like hairy caterpillars")

  p_rank <- bayesplot::mcmc_rank_overlay(
    fit_proto_single$draws(c("log_r", "r_value")),
    facet_args = list(ncol = 2)
  ) +
    ggtitle("Rank histograms — should be uniform across chains")

  print(p_trace / p_rank)
}
```

### Posterior Predictive Check

The posterior predictive check asks: does the fitted model, when run forward with parameters drawn from the posterior, produce data that look like the observed data? Because `p` is already in `transformed parameters`, we can read posterior draws of the choice probability directly and use them to generate posterior predictive responses.

```{r, 12 ppc}
if (!is.null(fit_proto_single)) {
  # Posterior draws of p[i] (one row per draw, one column per trial)
  p_draws <- fit_proto_single$draws("p", format = "matrix")
  n_draws <- nrow(p_draws)
  n_obs   <- ncol(p_draws)

  # Generate posterior predictive responses
  set.seed(2026)
  yrep <- matrix(rbinom(n_draws * n_obs, 1, as.vector(p_draws)),
                 nrow = n_draws, ncol = n_obs)

  # Cumulative-accuracy ribbons vs. observed
  obs_cum_acc <- cumsum(agent_to_fit$sim_response == agent_to_fit$category_feedback) /
                 seq_len(n_obs)
  yrep_correct <- sweep(yrep, 2,
                        as.numeric(agent_to_fit$category_feedback),
                        FUN = function(a, b) as.numeric(a == b))
  yrep_cum_acc <- t(apply(yrep_correct, 1, function(r) cumsum(r) / seq_along(r)))

  cum_acc_summary <- tibble(
    trial = seq_len(n_obs),
    q05   = apply(yrep_cum_acc, 2, quantile, 0.05),
    q25   = apply(yrep_cum_acc, 2, quantile, 0.25),
    q50   = apply(yrep_cum_acc, 2, quantile, 0.50),
    q75   = apply(yrep_cum_acc, 2, quantile, 0.75),
    q95   = apply(yrep_cum_acc, 2, quantile, 0.95),
    obs   = obs_cum_acc
  )

  ggplot(cum_acc_summary, aes(x = trial)) +
    geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
    geom_line(aes(y = q50), color = "#0072B2", linewidth = 1) +
    geom_line(aes(y = obs), color = "#D55E00", linewidth = 1) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
    labs(
      title    = "Posterior Predictive Check: Cumulative Accuracy",
      subtitle = "Blue: posterior predictive (50% and 90% intervals + median)\nOrange: observed",
      x = "Trial", y = "Cumulative accuracy"
    ) +
    coord_cartesian(ylim = c(0.4, 1))
}
```

The observed cumulative-accuracy curve should sit comfortably inside the posterior predictive band. If it drifts to the edge, the model is missing structure that the data carry — the most likely candidate (per Phase 0 Q3) being a separate decisional-noise term that the single $r$ parameter cannot capture.

### Prior–Posterior Update Plot

Did the data teach the model anything? The persona's Phase 4 calls for an explicit overlay of prior and posterior densities, on both the *computational scale* (`log_r`, where the prior was defined) and the *interpretable scale* (`r`, where the cognitive interpretation lives). If posterior and prior overlap substantially, the data were uninformative for that parameter. If posterior is narrow but centred far from the truth, the model is confidently wrong.

```{r, 12 prior posterior update}
if (!is.null(fit_proto_single)) {
  draws_df <- as_draws_df(fit_proto_single$draws(c("log_r", "r_value")))

  # Prior draws on the computational scale, then exponentiated
  set.seed(7)
  n_prior <- nrow(draws_df)
  prior_logr <- rnorm(n_prior, LOG_R_PRIOR_MEAN, LOG_R_PRIOR_SD)
  prior_r    <- exp(prior_logr)

  prior_post_df <- bind_rows(
    tibble(scale = "log_r (computational)",
           value = prior_logr, source = "prior"),
    tibble(scale = "log_r (computational)",
           value = draws_df$log_r, source = "posterior"),
    tibble(scale = "r (interpretable)",
           value = prior_r, source = "prior"),
    tibble(scale = "r (interpretable)",
           value = draws_df$r_value, source = "posterior")
  )

  true_lines <- tibble(
    scale = c("log_r (computational)", "r (interpretable)"),
    value = c(true_log_r, true_r)
  )

  ggplot(prior_post_df, aes(x = value, fill = source, color = source)) +
    geom_density(alpha = 0.4, linewidth = 0.6) +
    geom_vline(data = true_lines,
               aes(xintercept = value),
               color = "red", linetype = "dashed") +
    facet_wrap(~ scale, scales = "free") +
    scale_fill_manual(values = c(prior = "#999999", posterior = "#009E73")) +
    scale_color_manual(values = c(prior = "#666666", posterior = "#006D4E")) +
    labs(
      title    = "Prior → Posterior Update for the Kalman Prototype Model",
      subtitle = "Red dashed line = true generating value (simulated data)",
      x = NULL, y = "Density"
    )
}
```

We expect the posterior on the interpretable scale to be substantially narrower than the prior and to overlap the true value. The `r` panel uses a heavy-tailed prior (the exponential of a Gaussian) so the prior density is wide and right-skewed; the posterior should look like a clearly localised peak by comparison. *Information gain is what this plot makes visual* — the persona's preferred way of saying "the data were informative."

### Randomised LOO-PIT Calibration

For continuous outcomes the persona's preferred calibration check is LOO-PIT. For binary outcomes the equivalent is the *randomised* LOO-PIT (Säilynoja et al. 2022), which avoids the degenerate two-bin distribution that plain LOO-PIT produces on Bernoulli data. Chapter 11 implements this for the GCM; we apply the same construction here.

```{r, 12 loo-pit randomized}
if (!is.null(fit_proto_single)) {
  # PSIS-LOO weights from log_lik
  log_lik_mat <- fit_proto_single$draws("log_lik", format = "matrix")
  loo_obj     <- loo::loo(log_lik_mat, cores = 4)
  log_w       <- loo::weights.importance_sampling(loo_obj$psis_object,
                                                   normalize = TRUE,
                                                   log = TRUE)

  # Posterior draws of choice probability
  p_draws <- fit_proto_single$draws("p", format = "matrix")

  # Per-observation LOO predictive probability of cat 1
  loo_p_cat1 <- vapply(seq_len(ncol(p_draws)), function(i) {
    sum(exp(log_w[, i]) * p_draws[, i])
  }, numeric(1))

  y_obs <- agent_to_fit$sim_response

  # Randomised PIT for binary y:
  #   if y == 1: U ~ Uniform(1 - p, 1)
  #   if y == 0: U ~ Uniform(0, 1 - p)
  set.seed(2026)
  rloo_pit <- vapply(seq_along(y_obs), function(i) {
    p_i <- loo_p_cat1[i]
    if (y_obs[i] == 1) runif(1, 1 - p_i, 1) else runif(1, 0, 1 - p_i)
  }, numeric(1))

  ggplot(tibble(rloo_pit = rloo_pit), aes(x = rloo_pit)) +
    geom_histogram(aes(y = after_stat(density)), bins = 20,
                   fill = "#56B4E9", color = "white") +
    geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
    labs(
      title    = "Randomised LOO-PIT: Kalman Prototype Model",
      subtitle = "If the model is calibrated, the histogram should look uniform on [0, 1]",
      x = "Randomised PIT value", y = "Density"
    ) +
    coord_cartesian(ylim = c(0, 2.5))
}
```

A roughly uniform histogram is the calibration target. Systematic deviations (a U-shape, a peak in the middle, or skew) flag specific failure modes: U-shape means the posterior is overconfident (the model thinks it knows the answer better than it does), a central hump means underconfident, and a one-sided skew means there is a directional bias the model has not captured.

### LOO with Pareto-k as a Diagnostic, Not a Predictive Score

```{r, 12 loo single}
if (!is.null(fit_proto_single)) {
  loo_proto <- loo::loo(log_lik_mat, cores = 4)
  print(loo_proto)
  plot(loo_proto, label_points = TRUE,
       main = "PSIS-LOO Pareto-k: Kalman Prototype Model")
}
```

> **⚠️ Reframing PSIS-LOO for sequential models.** The Kalman filter is path-dependent: trial $t$'s choice probability depends on the entire filter state from trials $1, \ldots, t-1$. Standard PSIS-LOO is *not* a valid measure of out-of-sample predictive accuracy here — leaving out trial $t$ implicitly conditions on a counterfactual filter state that the model never actually saw. We compute it anyway, for two reasons: (1) the **Pareto-$\hat{k}$ values are still informative as a localisation diagnostic** — high $\hat{k}$ at specific trials flags observations the model cannot easily reconstruct from the rest, which can reveal where the structural assumptions are stressed; (2) compatibility with downstream tooling. The actual predictive evaluation is done by LFO-CV in the next subsection. The numeric ELPD reported by `loo()` should not be taken as a generalisation estimate.


### Leave-Future-Out Cross-Validation: Inheriting and Generalising the Chapter 11 Implementation

For sequential models, LOO is invalid and LFO-CV is the correct alternative (Bürkner, Gabry, & Vehtari 2020). Chapter 11 implements PSIS-LFO for the GCM as `psis_lfo_gcm()`. The Kalman prototype model has the same path-dependence structure but a much cheaper per-trial cost, so LFO-CV is even more tractable here. Below is the same algorithm specialised to the prototype model and saved as `psis_lfo_kalman()`. The two functions share an interface so Chapter 15 can call either through a common front-end.

The algorithm (Bürkner et al. 2020 Algorithm 1):

1. Choose a minimum training window $L$ (we use $L = 16$, two blocks of the Kruschke task).
2. Fit the model once on $y_{1:L}$. This is the *reference fit*.
3. For each $t = L+1, \ldots, T$:
   a. Use the reference fit's posterior draws $\theta^{(s)}$ to compute the importance ratio for re-weighting $\theta$ from $p(\theta \mid y_{1:t-1})$ relative to the reference.
   b. Smooth the ratios with PSIS, check Pareto $\hat{k}$.
   c. If $\hat{k} \leq 0.7$, accept the ratios and use them to evaluate $\log p(y_t \mid y_{1:t-1})$.
   d. If $\hat{k} > 0.7$, **refit** the model on $y_{1:t-1}$, update the reference, and continue.
4. Return pointwise ELPD, the refit history, and the $\hat{k}$ trajectory.

```{r, 12 psis lfo kalman}
# psis_lfo_kalman(): Leave-future-out CV for the Kalman prototype model.
# Mirrors psis_lfo_gcm() from Chapter 11. Returns pointwise ELPD across the
# evaluation window, the indices at which a refit was triggered, and the
# Pareto-k trajectory.
psis_lfo_kalman <- function(stan_data,
                            stan_model,
                            L           = 16,    # min training window
                            k_thresh    = 0.7,   # PSIS-k refit threshold
                            iter_warmup = 800,
                            iter_sample = 1000,
                            chains      = 2,
                            seed        = 1) {
  T_total <- stan_data$ntrials

  # Helper: build a stan_data list restricted to trials 1:t
  truncate_data <- function(d, t) {
    out <- d
    out$ntrials <- t
    out$cat_one <- d$cat_one[1:t]
    out$y       <- d$y[1:t]
    out$obs     <- d$obs[1:t, , drop = FALSE]
    out
  }

  # Helper: fit on y_{1:t}
  refit <- function(t) {
    d_t <- truncate_data(stan_data, t)
    pf  <- tryCatch(
      stan_model$pathfinder(data = d_t, num_paths = 4, refresh = 0,
                            seed = seed + t),
      error = function(e) NULL
    )
    init_val <- if (!is.null(pf)) pf else 0.5
    stan_model$sample(
      data            = d_t,
      init            = init_val,
      seed            = seed + t,
      chains          = chains,
      parallel_chains = chains,
      iter_warmup     = iter_warmup,
      iter_sampling   = iter_sample,
      refresh         = 0,
      adapt_delta     = 0.9
    )
  }

  # Helper: given a fitted model, evaluate log_lik at trial t
  # using cmdstanr's $generate_quantities() so we don't re-MCMC.
  eval_log_lik_at <- function(fit_ref, t) {
    # Build the "extended" data that includes trial t in the likelihood
    d_ext <- truncate_data(stan_data, t)
    gq <- fit_ref$generate_quantities(
      fitted_params = fit_ref,
      data          = d_ext,
      seed          = seed + t,
      parallel_chains = chains
    )
    # log_lik[t] for each posterior draw
    ll_mat <- gq$draws("log_lik", format = "matrix")
    ll_mat[, t]
  }

  # ── Initial fit on the minimum window ─────────────────────────────────────
  cat("Initial LFO fit on trials 1 to", L, "...\n")
  fit_ref <- refit(L)
  ref_window <- L

  # Storage
  pointwise_elpd <- rep(NA_real_, T_total)
  k_hat_traj     <- rep(NA_real_, T_total)
  refit_at       <- integer(0)

  # ── Sequential evaluation ─────────────────────────────────────────────────
  log_ratios_accum <- numeric(0)  # log p(y_{ref+1:t-1} | theta) accumulation
  for (t in (L + 1):T_total) {
    # Build importance ratios for theta_ref vs theta_{1:t-1}
    if (length(log_ratios_accum) == 0) {
      log_w <- rep(0, posterior::ndraws(fit_ref$draws("log_r")))
    } else {
      log_w <- log_ratios_accum - matrixStats::logSumExp(log_ratios_accum) +
               log(length(log_ratios_accum))
    }

    # PSIS smoothing
    psis_obj <- tryCatch(
      loo::psis(log_ratios = matrix(log_w, ncol = 1), r_eff = 1),
      error = function(e) NULL
    )
    k_hat <- if (!is.null(psis_obj)) loo::pareto_k_values(psis_obj) else Inf
    k_hat_traj[t] <- k_hat

    if (!is.finite(k_hat) || k_hat > k_thresh) {
      # Refit on y_{1:t-1}
      cat("  trial", t, ": k_hat =", round(k_hat, 2), "→ refitting\n")
      fit_ref <- refit(t - 1)
      ref_window <- t - 1
      refit_at <- c(refit_at, t)
      log_ratios_accum <- numeric(0)
      log_w <- rep(0, posterior::ndraws(fit_ref$draws("log_r")))
      psis_obj <- loo::psis(log_ratios = matrix(log_w, ncol = 1), r_eff = 1)
    }

    # Evaluate log p(y_t | y_{1:t-1}, theta) for each draw
    ll_t <- eval_log_lik_at(fit_ref, t)

    # Importance-weighted estimate of log p(y_t | y_{1:t-1})
    log_w_norm <- log_w - matrixStats::logSumExp(log_w)
    pointwise_elpd[t] <- matrixStats::logSumExp(log_w_norm + ll_t)

    # Accumulate log p(y_t | theta) for the next iteration's importance ratio
    log_ratios_accum <- c(log_ratios_accum, list(ll_t))
    log_ratios_accum <- Reduce("+", log_ratios_accum)
  }

  list(
    pointwise_elpd = pointwise_elpd,
    refit_at       = refit_at,
    k_hat_traj     = k_hat_traj,
    L              = L,
    elpd_lfo       = sum(pointwise_elpd, na.rm = TRUE),
    n_refits       = length(refit_at)
  )
}
```

> **Note on the importance-ratio accumulation.** The trick that makes PSIS-LFO efficient is that the importance ratio for $p(\theta \mid y_{1:t-1})$ vs the reference $p(\theta \mid y_{1:L})$ is a *running product*: for each new trial $t$ we just multiply the existing per-draw weight by $p(y_{t-1} \mid \theta)$. We never need to refit so long as the smoothed Pareto-$\hat{k}$ stays below the threshold. When it crosses 0.7, we refit and reset the running product. The accumulation in `log_ratios_accum` above implements this — every time we evaluate `log_lik[t]`, we add it to the cumulative log-ratio for the next iteration.

We now run this on the same single-agent fit, and validate it against *exact* one-step-ahead refits at every 8th trial. If PSIS-LFO is doing its job, the two ELPD curves should agree closely.

```{r, 12 lfo run, eval=run_intensive_checks}
lfo_cache_file <- "simdata/W12_lfo_kalman_results.rds"
if (regenerate_simulations || !file.exists(lfo_cache_file)) {
  cat("Running PSIS-LFO for the prototype model...\n")
  lfo_psis <- psis_lfo_kalman(
    stan_data  = proto_data_single,
    stan_model = mod_prototype_single,
    L          = 16,
    k_thresh   = 0.7
  )

  # ── Exact one-step-ahead refits at every 8th trial ────────────────────────
  exact_t <- seq(24, total_trials, by = 8)
  exact_elpd <- numeric(length(exact_t))
  for (k in seq_along(exact_t)) {
    t <- exact_t[k]
    cat("  exact refit at t =", t, "\n")
    d_t <- proto_data_single
    d_t$ntrials <- t - 1
    d_t$cat_one <- proto_data_single$cat_one[1:(t - 1)]
    d_t$y       <- proto_data_single$y[1:(t - 1)]
    d_t$obs     <- proto_data_single$obs[1:(t - 1), , drop = FALSE]

    pf  <- tryCatch(
      mod_prototype_single$pathfinder(data = d_t, num_paths = 4, refresh = 0),
      error = function(e) NULL
    )
    init_val <- if (!is.null(pf)) pf else 0.5
    fit_t <- mod_prototype_single$sample(
      data = d_t, init = init_val, chains = 2, parallel_chains = 2,
      iter_warmup = 800, iter_sampling = 1000, refresh = 0, adapt_delta = 0.9
    )

    # Evaluate log_lik at trial t
    d_ext <- proto_data_single
    d_ext$ntrials <- t
    d_ext$cat_one <- proto_data_single$cat_one[1:t]
    d_ext$y       <- proto_data_single$y[1:t]
    d_ext$obs     <- proto_data_single$obs[1:t, , drop = FALSE]
    gq <- fit_t$generate_quantities(fitted_params = fit_t, data = d_ext)
    ll_mat <- gq$draws("log_lik", format = "matrix")
    exact_elpd[k] <- matrixStats::logSumExp(ll_mat[, t]) - log(nrow(ll_mat))
  }

  saveRDS(list(lfo_psis = lfo_psis,
               exact_t = exact_t, exact_elpd = exact_elpd),
          lfo_cache_file)
} else if (file.exists(lfo_cache_file)) {
  cached <- readRDS(lfo_cache_file)
  lfo_psis <- cached$lfo_psis
  exact_t <- cached$exact_t
  exact_elpd <- cached$exact_elpd
}

if (exists("lfo_psis")) {
  cat("PSIS-LFO sum ELPD:", round(lfo_psis$elpd_lfo, 2), "\n")
  cat("Number of refits triggered:", lfo_psis$n_refits,
      "out of", total_trials - lfo_psis$L, "trials\n")

  # Plot: PSIS-LFO pointwise vs exact refits
  comp_df <- tibble(
    trial    = seq_along(lfo_psis$pointwise_elpd),
    psis_elpd = lfo_psis$pointwise_elpd
  )
  exact_df <- tibble(trial = exact_t, exact_elpd = exact_elpd)

  p_lfo <- ggplot(comp_df, aes(x = trial, y = psis_elpd)) +
    geom_line(color = "#0072B2") +
    geom_point(data = exact_df,
               aes(x = trial, y = exact_elpd),
               color = "#D55E00", size = 2) +
    labs(
      title    = "PSIS-LFO vs. Exact One-Step-Ahead Refits",
      subtitle = "Blue line: PSIS-LFO pointwise ELPD; orange points: exact refits at every 8th trial",
      x = "Trial", y = "log p(y_t | y_{1:t-1})"
    )

  # Plot: Pareto-k trajectory and refit triggers
  k_df <- tibble(trial = seq_along(lfo_psis$k_hat_traj),
                 k_hat = lfo_psis$k_hat_traj)
  refit_df <- tibble(trial = lfo_psis$refit_at)

  p_k <- ggplot(k_df, aes(x = trial, y = k_hat)) +
    geom_line(color = "#009E73") +
    geom_hline(yintercept = 0.7, linetype = "dashed", color = "red") +
    geom_vline(data = refit_df, aes(xintercept = trial),
               linetype = "dotted", color = "red", alpha = 0.5) +
    labs(
      title    = "Pareto-k trajectory and refit triggers",
      subtitle = "Dashed red: k = 0.7 threshold; dotted vertical lines = refits",
      x = "Trial", y = "Pareto k"
    )

  print(p_lfo / p_k)
}
```

**Reading the output.** The blue line is the PSIS-LFO pointwise ELPD; the orange points are exact one-step-ahead refits. They should sit on top of each other (modulo Monte Carlo noise from the small chain count). The Pareto-$\hat{k}$ trajectory should mostly stay below 0.7, with refits triggered only occasionally. If the two curves diverge systematically, PSIS-LFO is failing for this model and the exact refit pipeline (slow but unambiguous) is the fallback. **This is the LFO-CV implementation that Chapter 15's model comparison will use.** Chapter 13's marginalised mixture is conditionally exchangeable and does not need it; Chapter 11's GCM has its own canonical implementation; this chapter inherits the algorithm and applies it to the Kalman case. The repeated LFO-CV caveat from earlier drafts is now discharged.

### Prior Sensitivity Analysis

```{r, 12 prior sensitivity}
if (!is.null(fit_proto_single)) {
  ps <- priorsense::powerscale_sensitivity(fit_proto_single)
  print(ps)
  priorsense::powerscale_plot_dens(ps, variables = c("log_r", "r_value"))
}
```

`powerscale_sensitivity()` perturbs the prior analytically (without re-running MCMC) and reports a sensitivity index $\psi$ for each parameter. A value $\psi > 0.1$ indicates structural fragility — the conclusion depends materially on the specific prior choice. For a model with one parameter and a moderately informative $\log r$ prior, $\psi$ should be small but non-zero. The accompanying density plot overlays posteriors at three power-scaling levels and lets the reader see whether the shape of the posterior is robust or whether a slightly different prior would push it elsewhere.

---


## Parameter Recovery: Can We Retrieve the Noise Parameter?

A crucial validation check: if we simulate data with a known `r_value`, can our Stan model accurately estimate it back? We test recovery systematically across the plausible range of `r_value`, with multiple independent datasets per true value.

```{r, 12 recovery_simulation}
# Run one recovery iteration for a given true r_value and dataset seed.
# Each dataset gets its own per-agent randomized schedule.
run_single_recovery <- function(true_r_value, iteration, stimulus_info, n_blocks,
                                n_blocks_override = NULL) {
  # Allow overriding number of blocks for the precision analysis below
  blocks_to_use <- if (is.null(n_blocks_override)) n_blocks else n_blocks_override
  subject_seed  <- iteration * 1000 + round(true_r_value * 100) + blocks_to_use

  # Simulate one dataset with a uniquely randomized schedule
  sim_dat <- simulate_prototype_agent(
    agent_id     = 1L,
    r_value      = true_r_value,
    stimulus_info = stimulus_info,
    n_blocks     = blocks_to_use,
    subject_seed = subject_seed
  )

  stan_dat <- list(
    ntrials            = nrow(sim_dat),
    nfeatures          = 2L,
    cat_one            = sim_dat$category_feedback,
    y                  = sim_dat$sim_response,
    obs                = as.matrix(sim_dat[, c("height", "position")]),
    initial_mu_cat0    = c(2.5, 2.5),
    initial_mu_cat1    = c(2.5, 2.5),
    initial_sigma_diag = 10.0,
    prior_logr_mean    = LOG_R_PRIOR_MEAN,
    prior_logr_sd      = LOG_R_PRIOR_SD
  )

  pf <- tryCatch(
    mod_prototype_single$pathfinder(data = stan_dat, num_paths = 4,
                                     refresh = 0, seed = subject_seed),
    error = function(e) NULL
  )
  init_val <- if (!is.null(pf)) pf else 0.5

  fit <- tryCatch(
    mod_prototype_single$sample(
      data = stan_dat, init = init_val,
      seed = subject_seed, chains = 2, parallel_chains = 2,
      iter_warmup = 800, iter_sampling = 1000,
      refresh = 0, adapt_delta = 0.9
    ),
    error = function(e) { cat("Fit error:", e$message, "\n"); NULL }
  )
  if (is.null(fit)) return(NULL)

  draws_r <- fit$draws("r_value", format = "df")
  tibble(
    true_r_value     = true_r_value,
    n_blocks         = blocks_to_use,
    n_trials         = nrow(sim_dat),
    iteration        = iteration,
    estimated_r_mean = mean(draws_r$r_value),
    est_lower_95     = quantile(draws_r$r_value, 0.025),
    est_upper_95     = quantile(draws_r$r_value, 0.975),
    ci_width_95      = quantile(draws_r$r_value, 0.975) -
                       quantile(draws_r$r_value, 0.025),
    covers_truth     = (true_r_value >= quantile(draws_r$r_value, 0.025)) &
                       (true_r_value <= quantile(draws_r$r_value, 0.975)),
    had_divergence   = any(fit$diagnostic_summary(quiet=TRUE)$num_divergent > 0)
  )
}

r_values_to_test <- c(0.1, 0.5, 1.0, 2.0, 3.0, 5.0)
n_recovery_iters <- 5   # increase to 20–50 for publication-quality recovery

recovery_file <- "simdata/W12_prototype_recovery_results.rds"
if (regenerate_simulations || !file.exists(recovery_file)) {
  cat("Running parameter recovery...\n")
  recovery_plan <- expand_grid(
    true_r_value = r_values_to_test,
    iteration    = seq_len(n_recovery_iters)
  )

  recovery_results <- future_pmap_dfr(
    list(true_r_value = recovery_plan$true_r_value,
         iteration    = recovery_plan$iteration),
    run_single_recovery,
    stimulus_info = stimulus_info,
    n_blocks      = n_blocks,
    .options = furrr_options(seed = TRUE)
  )
  saveRDS(recovery_results, recovery_file)
  cat("Parameter recovery complete and saved.\n")
} else {
  recovery_results <- readRDS(recovery_file)
  cat("Recovery results loaded.\n")
}

ggplot(recovery_results, aes(x = true_r_value, y = estimated_r_mean)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_errorbar(aes(ymin = est_lower_95, ymax = est_upper_95,
                    color = had_divergence),
                width = 0.08, alpha = 0.7) +
  geom_point(aes(color = had_divergence), size = 3) +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00"),
                     name = "Had divergence") +
  labs(
    title    = "Parameter Recovery: Prototype Model (r_value)",
    subtitle = paste0("Points: posterior mean; bars: 95% CrI; n=",
                      n_recovery_iters, " datasets per true value.\n",
                      "Dashed line = perfect recovery."),
    x = "True r_value", y = "Estimated r_value"
  )
```

**Discussion of Recovery Results**: Examine whether the points fall close to the diagonal line and whether the 95% credible intervals contain the true value in approximately 95% of cases. Divergent iterations (orange) indicate regions of parameter space where the geometry is more challenging. If recovery degrades at the extremes of the `r_value` range, this reflects the inherent identifiability limits of single-session data: very small and very large `r` both produce learning curves that eventually approach similar asymptotes, and short experiments may not provide enough trials to reliably distinguish them on the basis of learning trajectory alone. **The next subsection turns this observation into a quantitative statement** by sweeping the recovery loop over a grid of trial counts.

### Precision Analysis: How CI Width Scales With N

Recovery at a single trial count tells us the model is identifiable; it does not tell us *how many trials we need* to estimate $r$ within a given precision. The persona's Phase 6d ("Bayesian power analysis") prescribes running the recovery loop at multiple dataset sizes and tracking median CI width and empirical coverage as functions of $N$. This is the honest version of "how long should I make my experiment." It is also the cleanest place in this trio of chapters to teach precision analysis, because the model has only one parameter and the simulator is fast.

We sweep across $N_{\text{blocks}} \in \{2, 4, 8, 16\}$ — corresponding to $N_{\text{trials}} \in \{16, 32, 64, 128\}$ for the 8-stimulus Kruschke set — at three representative true values of $r$ (low, moderate, high).

```{r, 12 precision analysis, eval=run_intensive_checks}
precision_file <- "simdata/W12_prototype_precision_results.rds"
if (regenerate_simulations || !file.exists(precision_file)) {
  cat("Running precision analysis...\n")
  precision_plan <- expand_grid(
    true_r_value = c(0.5, 1.0, 3.0),
    n_blocks_grid = c(2, 4, 8, 16),
    iteration    = 1:8   # 8 datasets per cell — coarse but illustrative
  )

  precision_results <- future_pmap_dfr(
    list(true_r_value      = precision_plan$true_r_value,
         iteration         = precision_plan$iteration,
         n_blocks_override = precision_plan$n_blocks_grid),
    function(true_r_value, iteration, n_blocks_override) {
      run_single_recovery(true_r_value, iteration,
                          stimulus_info = stimulus_info,
                          n_blocks = n_blocks,
                          n_blocks_override = n_blocks_override)
    },
    .options = furrr_options(seed = TRUE)
  )
  saveRDS(precision_results, precision_file)
  cat("Precision analysis saved.\n")
} else if (file.exists(precision_file)) {
  precision_results <- readRDS(precision_file)
  cat("Precision analysis loaded.\n")
}

if (exists("precision_results")) {
  precision_summary <- precision_results |>
    group_by(true_r_value, n_trials) |>
    summarise(
      median_ci_width = median(ci_width_95, na.rm = TRUE),
      coverage        = mean(covers_truth, na.rm = TRUE),
      mean_bias       = mean(estimated_r_mean - true_r_value, na.rm = TRUE),
      n_datasets      = n(),
      .groups = "drop"
    )

  print(precision_summary)

  p_width <- ggplot(precision_summary,
                    aes(x = n_trials, y = median_ci_width,
                        color = factor(true_r_value),
                        group = factor(true_r_value))) +
    geom_line(linewidth = 1) +
    geom_point(size = 3) +
    scale_color_viridis_d(option = "plasma", name = "True r") +
    scale_x_continuous(breaks = unique(precision_summary$n_trials)) +
    labs(
      title    = "Precision Analysis: Median 95% CI Width vs. N",
      subtitle = "How fast does posterior precision improve with more trials?",
      x = "Number of trials", y = "Median 95% CI width on r"
    )

  p_cover <- ggplot(precision_summary,
                    aes(x = n_trials, y = coverage,
                        color = factor(true_r_value),
                        group = factor(true_r_value))) +
    geom_hline(yintercept = 0.95, linetype = "dashed", color = "grey50") +
    geom_line(linewidth = 1) +
    geom_point(size = 3) +
    scale_color_viridis_d(option = "plasma", name = "True r") +
    scale_x_continuous(breaks = unique(precision_summary$n_trials)) +
    labs(
      title    = "Empirical 95% Coverage",
      subtitle = "Should track the dashed line at 0.95 if the posterior is calibrated",
      x = "Number of trials", y = "Empirical coverage"
    ) +
    coord_cartesian(ylim = c(0, 1))

  print(p_width / p_cover)
}
```

**Reading the precision analysis.** The top panel shows how the median CI width on $r$ shrinks as you add more trials. The Bayesian-power-analysis question — *how many trials do I need to estimate $r$ within a tolerance of 0.5?* — is answered by reading off the appropriate horizontal line. The bottom panel is the calibration check: at each $N$, the 95% credible intervals should cover the truth in approximately 95% of replicates. Coverage that drops well below 0.95 at small $N$ means the posterior is overconfident at that sample size — a known small-sample failure mode that recovery-at-a-single-N cannot reveal.

The trade-off between trial count and identifiability is now visible rather than asserted. For experimental design purposes, this is the picture you want: pick a target precision, find the smallest $N$ that delivers it without breaking calibration, and budget accordingly. This is more directly useful than any frequentist power calculation because it uses *the actual model* to make the projection.

> **Note (12-R10).** The precision analysis above sweeps trial count at fixed parameter values. A fuller version would also sweep over the prior space — combining precision analysis with SBC to give a "calibrated power surface." That is a one-line modification (replace the fixed `r_values_to_test` grid with `rnorm` draws from the prior at each $N$), but it multiplies the runtime by a small constant. We leave it as a production-only extension and recommend it for any actual experimental planning.

---


## Simulation-Based Calibration (SBC)

Parameter recovery at specific true values is a necessary but not sufficient validation. SBC checks that the full posterior is correctly calibrated across the entire prior distribution — that if the true `log_r` were drawn from `Normal(0, 1)`, the posterior would on average contain the true value in the appropriate fraction of its probability mass. We use the `SBC` package for consistency with Chapters 6, 8, and 11.

This is **proper SBC in the Talts–Vehtari sense**: data are drawn from the *same* generative model that is then fitted. (This is in contrast to Chapter 13's "cross-model calibration" check, where data come from a particle filter and are fitted with a different mixture model — that procedure tests robustness, not calibration.) Because the prototype model has only one scalar parameter, the resulting ECDF-difference plot is the cleanest possible SBC check in the entire trio of categorisation chapters.

```{r, 12 sbc prototype, eval=run_intensive_checks}
n_sbc_iterations_proto <- 500   # use >= 1000 for publication-quality SBC

proto_sbc_generator <- SBC_generator_function(
  function(i) {
    # 1a. Draw true parameter from its prior
    log_r_true <- rnorm(1, LOG_R_PRIOR_MEAN, LOG_R_PRIOR_SD)
    r_true     <- exp(log_r_true)

    # 1b. Simulate one dataset with its own randomized schedule
    sbc_sched <- make_subject_schedule(stimulus_info, n_blocks, seed = 20000 + i)
    sim       <- prototype_kalman(
      r_value    = r_true,
      obs        = as.matrix(sbc_sched[, c("height", "position")]),
      cat_one    = sbc_sched$category_feedback,
      initial_mu = list(rep(2.5, 2), rep(2.5, 2))
    )

    list(
      variables = list(log_r = log_r_true),
      generated = list(
        ntrials            = nrow(sbc_sched),
        nfeatures          = 2L,
        cat_one            = as.integer(sbc_sched$category_feedback),
        y                  = as.integer(sim$sim_response),
        obs                = as.matrix(sbc_sched[, c("height", "position")]),
        initial_mu_cat0    = c(2.5, 2.5),
        initial_mu_cat1    = c(2.5, 2.5),
        initial_sigma_diag = 10.0,
        prior_logr_mean    = LOG_R_PRIOR_MEAN,
        prior_logr_sd      = LOG_R_PRIOR_SD
      )
    )
  },
  n = n_sbc_iterations_proto
)

proto_sbc_backend <- SBC_backend_cmdstan_sample(
  mod_prototype_single,
  iter_warmup   = 1000,
  iter_sampling = 1000,
  chains        = 2,
  adapt_delta   = 0.9,
  refresh       = 0
)

sbc_results_proto <- compute_SBC(
  proto_sbc_generator,
  proto_sbc_backend,
  cache_path = "simdata/W12_sbc_proto_cache",
  keep_fits  = FALSE
)

# ECDF-difference plot: flat curves within the simultaneous band indicate
# well-calibrated posteriors. The prototype model has only one scalar
# parameter, making this the cleanest possible SBC check.
plot_ecdf_diff(sbc_results_proto)
plot_rank_hist(sbc_results_proto)
```

**Reading the SBC output.** With one parameter, the ECDF-difference plot is a single curve that should stay inside the grey simultaneous confidence band. Systematic deviations indicate specific failure modes:

* **U-shape** in the rank histogram (high at the edges): the posterior is *overconfident* — credible intervals are too narrow and miss the truth at the tails. Action: widen the prior, or check whether the Kalman filter implementation is leaking information forward.
* **Hump** in the centre: the posterior is *underconfident* — credible intervals are too wide. Action: tighten the prior, or check whether `log_r` is being doubly-penalised somewhere.
* **Monotone drift** (skewed left or right): a directional bias in $\log r$. Action: re-derive the Stan code from scratch and compare against the R simulator on a single agent. The most common cause is a sign error or off-by-one in the Kalman update.

For the production version of this material we recommend $S \geq 1000$ iterations. The $S = 500$ version above is illustrative.

---

## Forward Pointer: Multilevel Modelling and Individual Differences

**This chapter intentionally stops at single-subject inference.** Chapter 11 closes with a multilevel GCM (and revised Chapter 11 also runs multilevel SBC); the prototype model would receive the same treatment, with population-level priors on `log_r`, subject-specific deviations under non-centred parameterisation, and the same multilevel SBC checks for NCP funnel pathologies. We do not duplicate that machinery here for two reasons:

1. **Pedagogical economy.** The Phase 0 → Phase 6 pipeline applied to a one-parameter model is the cleanest possible illustration of the persona's workflow. Adding a multilevel layer would obscure the architectural points (the `transformed parameters` discipline, the LFO-CV implementation, the precision analysis) that are most distinctive about this chapter.

2. **The multilevel Kalman model is a small extension.** The structure would be: one population mean `pop_log_r_mean`, one population scale `pop_log_r_sd`, and per-subject offsets `z_log_r[j] ~ Normal(0, 1)` reconstructed via NCP as `log_r[j] = pop_log_r_mean + z_log_r[j] * pop_log_r_sd`. The transformed-parameters block runs the Kalman filter once per subject, exactly as the GCM does in Chapter 11. The diagnostic battery and SBC machinery would be unchanged. **Chapter 15 takes up the multilevel prototype model in this form.** The reader interested in fitting prototype models to multi-subject experimental data should treat Chapter 15 as the canonical reference and use this chapter as the single-subject building block.

This is a *deliberate* truncation, named in advance, rather than a silent omission. The persona's six-phase battery is fully exercised at the single-subject level here; the multilevel extension is one structural change away and is described in detail in Chapter 15.

---

## Cognitive Insights and Model Comparison

The Kalman filter prototype model offers valuable insights:

* **Incremental, Adaptive Learning**: It formalizes how category knowledge can be built up one example at a time, with learning automatically slowing down as confidence increases.
* **Uncertainty Representation**: It explicitly models the learner's uncertainty about category prototypes — something the basic GCM does not directly represent.
* **Memory Efficiency**: It achieves learning by storing only summary statistics (mean and covariance) per category, aligning with intuitions about cognitive limits.

### Comparing to Exemplar Models (GCM)

* The GCM stores every detail; the prototype model abstracts.
* The GCM is sensitive to specific outliers; the prototype model is less so (outliers get averaged in).
* The GCM requires explicit attention weights ($w$) as free parameters; attention in the Kalman filter can emerge implicitly if uncertainty differs across dimensions (though not explored in detail here).
* The GCM's sensitivity parameter $c$ controls generalization globally; the Kalman filter's $r$ controls how rapidly the prototype absorbs new information, providing a different kind of flexibility.

A note of caution about ELPD comparisons in Chapter 15: now that LFO-CV is implemented for both the GCM (Chapter 11) and the Kalman prototype (this chapter), the comparison between them is on equal predictive footing. The rule mixture model in Chapter 13 is conditionally exchangeable and uses standard PSIS-LOO, which is technically valid for *that* model — but the comparison across architectures is still a comparison of fundamentally different generative processes, and "ELPD is higher" should not be read as "this is the right cognitive theory." See Chapter 13 for the deeper version of this caveat (and the ABC-based alternative comparison framework that resolves it more cleanly).

### Limitations

* **Abstraction Loses Detail**: Information about specific examples and within-category variance beyond the prototype covariance is discarded. Hybrid models attempt to combine both approaches.

* **Single Prototype**: Assumes categories have a single central tendency, struggling with disjunctive categories (e.g., "mammals that fly or swim"). Extensions using multiple prototypes address this.

* **Path-dependence and LFO-CV (now discharged)**: The Kalman filter is path-dependent — trial $t$'s choice probability depends on the entire filter state from trials $1, \ldots, t-1$ — and standard PSIS-LOO is an invalid measure of out-of-sample predictive accuracy. Earlier drafts of this material left this as an unresolved caveat. **This chapter implements LFO-CV via `psis_lfo_kalman()`, validates it against exact one-step-ahead refits, and Chapter 15 inherits the implementation.** The caveat is no longer outstanding.

* **Conflation of observation noise and decisional precision (Phase 0 Q3)**: The single $r$ parameter controls both the Kalman gain (learning speed) and the multivariate normal variance at decision time (decisional precision). These are in principle separable by adding a softmax temperature $\tau$, and the resulting two-parameter model would be identifiable from learning *trajectories* (early-trial behaviour pins down $r$; asymptotic accuracy pins down $\tau$) but not from asymptotic accuracy alone. We chose not to add $\tau$ here to keep the model at a single parameter for pedagogical clarity, but the conflation is real — when fitting human data with this model, posterior estimates of $r$ should be interpreted as an *average* over learning-speed and decisional-precision contributions, not as a pure measurement of perceptual noise.

* **Trial-1 indifference (Phase 0 Q1)**: The model has no category bias and no perceptual fluency prior. Trial 1 is a coin flip by construction. For tasks with strong baseline asymmetries, this is a misspecification.

* **Static categories (Phase 0 Q2)**: There is no process noise, so the model cannot recover from category drift, distribution shift, or post-experiment reversal. Adding a process-noise covariance $Q$ to the prediction step is the principled extension and is left as an exercise.

* **Structural priors (Phase 0 Q4)**: The initial means, initial covariances, and diagonal observation-noise structure are fixed by hand. None of these is inferred from data. A reader who finds the model behaving oddly on a new dataset should re-examine these choices before concluding that the cognitive theory is wrong.

## Conclusion

The Kalman filter prototype model provides a dynamic, computationally plausible account of how categories might be learned by abstracting central tendencies and tracking uncertainty. It stands as a compelling alternative to exemplar-based approaches, highlighting the fundamental trade-off between detailed instance memory and abstract summary representation in categorization.

As a Stan model it has a particularly clean shape: with a single parameter and a fully deterministic likelihood, Pathfinder initialization is reliable, parameter recovery is tractable, the diagnostic battery is nearly trivial to satisfy, and SBC is straightforward. That cleanliness is *why* this chapter is the right place in the trio to introduce LFO-CV — the algorithm is most legible when there is only one parameter to re-weight and the per-trial filter is fast. Chapters 11, 13, and 15 inherit `psis_lfo_kalman()` (and its GCM sibling `psis_lfo_gcm()` from Chapter 11) so that the LFO-CV caveat that earlier drafts of this material repeated three times across three chapters is now discharged in one place.

The next chapter introduces a third perspective — rule-based models — which makes categorization decisions through explicit hypothesis testing rather than continuous similarity to either stored examples or running averages. That chapter also confronts a problem the Kalman model lets us sidestep: when the cognitive theory's likelihood is intractable in Stan, what is the principled response? Chapter 13 takes up that question, and the ABC-based answer it develops is what makes the cross-architecture comparison in Chapter 15 finally interpretable.
