---
title: "Rule-Based Categorization Models"
format:
  html:
    code-fold: true
    code-tools: true
---

```{r ch14_setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "80%",
  dpi        = 300
)

# Set to TRUE to rerun simulations and Stan fits.
# FALSE loads pre-computed results from 'simdata/'.
regenerate_simulations <- FALSE

# Set to TRUE to run the long-running validation chunks
# (rule-set sensitivity, real SBC for the Stan proxy, NPE pipeline,
# coverage validation, NPE-vs-Stan comparison).
run_intensive_checks   <- regenerate_simulations

library(here)
pacman::p_load(
  tidyverse,    # data manipulation and visualization
  patchwork,    # combining plots
  cmdstanr,     # Stan interface
  posterior,    # tidy posterior arrays
  tidybayes,    # tidy extraction of draws
  bayesplot,    # MCMC diagnostic plots
  loo,          # LOO-CV and PSIS
  SBC,          # Simulation-Based Calibration Checks
  priorsense,   # prior sensitivity
  future,       # parallel processing
  furrr,        # parallel functional programming
  matrixStats,  # logSumExp for PSIS calculations
  reticulate    # Python interop for sbi / BayesFlow
)

theme_set(theme_classic())

# ── Prior hyperparameter for logit(error_prob) ────────────────────────────────
# Normal(0, 1.5) is approximately uniform on the probability scale (persona
# Phase 0 Q4 default). Prior median ε ≈ 0.50; 95% mass on (0.05, 0.95).
# This is deliberately wider than the Normal(-2, 1) used in earlier drafts,
# which concentrated mass below ε = 0.27 and cut off the posterior for
# near-chance subjects. priorsense::powerscale_sensitivity() should be
# checked; if ψ > 0.1 for any quantity, the prior matters and this choice
# must be defended.
LOGIT_ERROR_PRIOR_MEAN <- 0.0
LOGIT_ERROR_PRIOR_SD   <- 1.5

# ── Structural commitments for the particle filter (Model A) ─────────────────
# These are NOT inferred from data — they are cognitive choices the modeller
# makes. Phase 0, question 4 surfaces them.
N_PARTICLES_FIXED  <- 100    # working-memory bound on hypothesis count
MAX_RULE_DIMS      <- 2      # maximum number of conditions in a rule
RULE_OPERATORS     <- c(">", "<")
RULE_LOGIC         <- c("AND", "OR")
RESAMPLE_THRESHOLD <- 0.5    # resample when ESS < N_particles * this factor

for (d in c("stan", "simdata", "simmodels")) {
  if (!dir.exists(d)) dir.create(d)
}

# ── Shared diagnostic helper (canonical version lives in Ch 11) ──────────────
# Re-defined here for standalone use. Same signature as Ch 11 / Ch 12.
diagnostic_summary_table <- function(fit, params = NULL) {
  diag <- fit$diagnostic_summary(quiet = TRUE)
  draws_summary <- fit$summary(variables = params)

  n_div    <- sum(diag$num_divergent)
  max_rhat <- round(max(draws_summary$rhat,     na.rm = TRUE), 2)
  min_bulk <- round(min(draws_summary$ess_bulk,  na.rm = TRUE), 0)
  min_tail <- round(min(draws_summary$ess_tail,  na.rm = TRUE), 0)
  min_ebfmi <- round(min(diag$ebfmi), 3)

  mcse_col <- intersect(names(draws_summary), c("mcse_mean", "mcse", "mcse_cp"))[1]
  if (!is.na(mcse_col)) {
    max_mcse <- round(max(draws_summary[[mcse_col]] / draws_summary$sd, na.rm = TRUE), 4)
  } else {
    max_mcse <- round(max(1 / sqrt(draws_summary$ess_bulk), na.rm = TRUE), 4)
  }

  out <- tibble::tibble(
    metric = c(
      "Divergences (zero tolerance)",
      "Max rank-normalised R-hat",
      "Min bulk ESS",
      "Min tail ESS",
      "Min E-BFMI",
      "Max MCSE / posterior SD"
    ),
    value = c(n_div, max_rhat, min_bulk, min_tail, min_ebfmi, max_mcse),
    threshold = c("== 0", "< 1.01", "> 400", "> 400", "> 0.2", "< 0.05"),
    pass = c(
      n_div == 0,
      max_rhat < 1.01,
      min_bulk > 400,
      min_tail > 400,
      min_ebfmi > 0.2,
      max_mcse < 0.05
    )
  )

  if (!all(out$pass)) {
    warning(
      "MCMC diagnostic battery: at least one threshold breached. ",
      "Inspect divergences and reparameterize before interpreting the posterior.",
      call. = FALSE
    )
  }

  out
}

# ── Kruschke (1993) stimulus set ─────────────────────────────────────────────
# Identical to Ch. 11 and Ch. 12; redefined here for standalone use.
stimulus_info <- tibble(
  stimulus      = c(5, 3, 7, 1, 8, 2, 6, 4),
  height        = c(1, 1, 2, 2, 3, 3, 4, 4),
  position      = c(2, 3, 1, 4, 1, 4, 2, 3),
  category_true = c(0, 0, 1, 0, 1, 0, 1, 1)
)

n_blocks         <- 8
n_stim_per_block <- nrow(stimulus_info)
total_trials     <- n_blocks * n_stim_per_block

# Feature ranges for threshold sampling (min and max of each feature)
feature_range <- list(
  height   = range(stimulus_info$height),
  position = range(stimulus_info$position)
)

# ── Per-subject schedule helper ───────────────────────────────────────────────
make_subject_schedule <- function(stimulus_info, n_blocks, seed) {
  set.seed(seed)
  n_stim <- nrow(stimulus_info)
  sequence <- unlist(lapply(seq_len(n_blocks), function(b) {
    sample(stimulus_info$stimulus, n_stim, replace = FALSE)
  }))
  tibble(
    trial_within_subject = seq_along(sequence),
    block                = rep(seq_len(n_blocks), each = n_stim),
    stimulus_id          = sequence
  ) |>
    left_join(stimulus_info, by = c("stimulus_id" = "stimulus")) |>
    rename(category_feedback = category_true) |>
    dplyr::select(trial_within_subject, block, stimulus_id,
                  height, position, category_feedback)
}

plan(multisession, workers = max(1, availableCores() - 1))
```

> **📍 Where we are in the Bayesian modeling workflow:**
> Chapters 11 and 12 implemented similarity-based accounts of categorization:
> the GCM stores individual exemplars and decides by summed similarity, while
> the Kalman filter prototype model tracks running category averages. Both
> update continuously and live as well-formed Stan likelihoods. This chapter
> takes a third route: **rule-based categorization**, where learners maintain
> and update discrete hypotheses about logical criteria for category
> membership. We implement this using a Bayesian particle filter — a sequential
> Monte Carlo method that maintains a weighted distribution over candidate
> rules.
>
> **Two distinct models will appear in this chapter, and the distinction is
> load-bearing**:
>
> * **Model A (the cognitive theory)** is the particle filter itself. Its
>   parameters are $\varepsilon$ (error rate), $N_{\text{particles}}$ (a
>   working-memory bound), `max_dims`, the rule grammar, and the resampling
>   threshold. Its likelihood has no closed form because the rule space is
>   discrete and combinatorial, the resampling step is non-differentiable, and
>   the particle trajectories cannot be marginalised analytically. **Stan
>   cannot fit Model A.** It can only be fitted via simulator-based inference,
>   for which we use Neural Posterior Estimation (NPE) via the `sbi` Python
>   package, accessed from R through `reticulate`.
> * **Model B (a tractable Stan proxy)** is a static, conditionally
>   exchangeable mixture over a small fixed set of hand-picked candidate
>   rules. Its likelihood is closed-form Bernoulli marginalised with
>   `log_sum_exp`. It admits the full HMC validation pipeline. **Validation of
>   Model B is not validation of Model A.** A reader who skips this distinction
>   will end up using LOO scores from Model B as if they reported on the
>   particle filter's predictive accuracy. They do not.
>
> The pedagogical move in this chapter is to fit *both* models, using HMC for
> Model B and Neural Posterior Estimation (NPE) for Model A, and to read the
> gap between their posteriors as the chapter's main empirical finding: the
> magnitude of the structural bias you accept when you replace an intractable
> cognitive theory with a tractable proxy.
>
> Standard PSIS-LOO is **technically valid for Model B** (it is conditionally
> exchangeable — leaving out trial $t$ does not affect the likelihood of any
> other trial), so the LFO-CV machinery introduced in Chapters 11 and 12 is
> not needed here. We say this once, in §"LFO-CV does not apply here," and
> reference Chapter 11's `psis_lfo_gcm()` and Chapter 12's `psis_lfo_kalman()`
> for completeness.

# Rule-Based Models of Categorization

> **📋 Validation battery roadmap for this chapter**
>
> Two columns because two models. Validation of Model B does not transfer to
> Model A.
>
> | Persona phase | Model A (particle filter) | Model B (Stan mixture proxy) |
> |---|---|---|
> | 0. Conceptual interrogation | §"Phase 0: Two Models, Not One" | §"Phase 0: Two Models, Not One" |
> | 1. Forward simulator | `rule_particle_filter()` | none — no separate simulator needed |
> | 2. Implementation | R only — no closed-form likelihood | §"Stan Model: Marginalised Mixture" |
> | 3.0 MCMC diagnostics | n/a (no MCMC) | §"Mandatory MCMC Diagnostic Battery" |
> | 3.2 Prior predictive | §"Prior Predictive Check (Model A)" — and the cross-model warning that goes with it | uses Model A's simulator → cross-model warning applies |
> | 3.3 Posterior predictive | via NPE posterior | §"Posterior Predictive Check (Model B)" |
> | 3.4 Prior–posterior update | §"NPE: Prior–Posterior Update for $\varepsilon$" | §"Prior–Posterior Update Plot (Model B)" |
> | 3.5 Prior sensitivity | embedding network learns summaries; prior-width sweep | `priorsense::powerscale_sensitivity()` |
> | 4.5 LOO-PIT (randomised, binary) | n/a | §"Randomised LOO-PIT (Model B)" |
> | 4.6 LFO-CV | n/a (Model A is fitted via NPE, not likelihood) | §"LFO-CV Does Not Apply Here" — Model B is exchangeable |
> | 6a. Parameter recovery | §"NPE: Parameter Recovery via SBC + Expected Coverage" | §"Parameter Recovery (Model B)" |
> | 6b. SBC | §"NPE: SBC + Expected Coverage + TARP" | §"Real SBC for the Stan Mixture (Model B)" |
> | 6b.cross. Cross-model calibration check | §"Cross-Model Calibration Check" — generates from Model A, fits Model B; **labelled** as a robustness check, not SBC |
> | 6c. Posterior SBC | n/a | §"Posterior SBC for Model B" — local calibration at the fitted posterior |
> | Identifiability check | §"Phase 0", Q3 ($\varepsilon$ vs rule set) | §"Phase 0", Q3 |

## Theoretical Foundations

In the previous chapters, we explored models based on similarity. Exemplar models compute similarity to stored instances, while prototype models compute similarity to category averages. However, humans sometimes categorize using explicit criteria or rules. Think about definitions like "a square is a shape with four equal sides and four right angles" or simple diagnostic rules like "if the animal has feathers, it's a bird."

Rule-based models propose that categorization often involves learning and applying such explicit rules that define category membership by partitioning the feature space.

### Core Assumptions of Rule-Based Models

1. **Explicit Rules**: Categories are represented by rules specifying conditions for membership (e.g., "Height > 2.5 AND Position < 3.0").

2. **Rule Testing**: Learning involves generating, testing, weighting and refining candidate rules based on feedback.

3. **Selective Attention**: Rules typically focus attention on specific, relevant features mentioned in the rule, ignoring irrelevant ones.

4. **Deterministic Boundaries**: Simple rules often create sharp, axis-aligned boundaries in the feature space.

5. **Verbal Representation**: Rules are often easily verbalizable (e.g., "If height > 2.5, then Category A"), and might actually become easier to learn and represent through verbalization.

### Rules vs. Similarity-Based Approaches

The distinction between rule-based and similarity-based approaches lies not just in the computational details, but in fundamental assumptions about how the mind represents categories:

| Aspect | Rule-Based Models | Similarity-Based Models |
|--------|------------------|------------------------|
| **Representation** | Explicit decision criteria | Stored examples or prototypes |
| **Decision Process** | Rule application | Similarity computation |
| **Feature Use** | Selective (focus on rule-relevant features) | Holistic (all features might contribute) |
| **Category Boundaries** | Sharp and often axis-aligned | Fuzzy and sensitive to overall similarity |
| **Individual Differences** | Different rules | Different attention weights |
| **Generalization Pattern** | Rectangle-like regions | Similarity-gradient regions |

### Rule-Based Categorization in the Six-Step Pipeline

Chapters 11 and 12 used a six-stage framework to locate GCM and the Kalman filter prototype in the space of possible categorization models. The rule-based model occupies a distinct position at almost every stage:

| Stage | GCM (Ch. 11) | Kalman prototype (Ch. 12) | Rule particle filter (Ch. 13) |
|---|---|---|---|
| **1. Input representation** | Raw feature vector | Raw feature vector | Raw feature vector |
| **2. Attention** | Continuous weights $w_d$ across all features | None (equal implicit weight) | Selective — only rule-relevant dimensions attend; irrelevant dimensions are structurally ignored |
| **3. Intermediate representation** | Per-category similarity sums | Per-category Gaussian running mean + covariance | Weighted distribution over discrete rule hypotheses $\{r_1, \ldots, r_N\}$ |
| **4. Evidential mechanism** | Luce-ratio over similarities | Ratio of Gaussian likelihoods | Weighted vote over rule predictions, softened by $\varepsilon$ |
| **5. Decision rule** | Softmax threshold | Softmax threshold | Same |
| **6. Learning** | Exemplar accumulation (decay at rate $\lambda$) | Kalman update (process noise $q$) | Sequential Monte Carlo: weight update → resample → (mutate at rate $\mu$) |

The key contrasts are at stages 2, 3, and 6. At stage 2, the rule model is the only one that uses *structural* selective attention — the rule itself encodes which dimensions are relevant, rather than learned continuous weights. At stage 3, the model maintains a *hypothesis distribution* over discrete logical objects, not a similarity sum or a running mean. At stage 6, learning is a particle filter rather than a closed-form update, which is why the likelihood is intractable and NPE is required to fit Model A.

::: {.callout-note title="Why the six-step pipeline does not produce a single clean Stan model here"}
In Chapters 11 and 12, each stage of the pipeline mapped onto a differentiable Stan operation: gradient-based sampling could reach every parameter. In Chapter 13, stage 3 (the discrete hypothesis distribution) and stage 6 (the stochastic resample step) break differentiability. The particle filter is a valid Bayesian cognitive model, but it is not a Stan model. This is why the chapter introduces two models: Model A is the cognitive theory, fitted via NPE; Model B is a tractable differentiable proxy, fitted via HMC.
:::

### Representing and Evaluating Rules

How can we represent a rule computationally? For our 2D stimuli (height, position), a simple rule might involve one or both dimensions and a threshold.

Example Rule: "If Height > 3.5, then predict Category 1, otherwise Category 0."

We can represent this rule in R as a list:

```{r ch14_rule_representation}
# Example: a simple 1D rule
rule_example_1d <- list(
  dimensions        = 1,
  thresholds        = 3.5,
  operations        = ">",
  prediction_if_true = 1
)

# Example: a 2D rule (Height > 2 AND Position < 3 => Cat 0)
rule_example_2d <- list(
  dimensions        = c(1, 2),
  thresholds        = c(2.0, 3.0),
  operations        = c(">", "<"),
  logic             = "AND",
  prediction_if_true = 0
)

# Example: a nested rule — (Height > 2.5) AND ((Height > 3.5) OR (Position < 2.0))
# Cognitive reading: "tall-enough AND (very tall OR low on screen)"
# This is a RULEX-style conjunctive rule with a disjunctive exception clause.
# The 'nested_logic' field marks this as a three-condition rule; the first
# condition is in clause A and clauses B and C are combined by the inner connective.
# The outer connective links clause A with (B OR C).
rule_example_nested <- list(
  dimensions         = c(1, 1, 2),      # features for each of the 3 conditions
  thresholds         = c(2.5, 3.5, 2.0),
  operations         = c(">", ">", "<"),
  outer_logic        = "AND",           # connects clause A with (B inner_logic C)
  inner_logic        = "OR",            # connects clauses B and C
  prediction_if_true = 1
)
```

Now, a function to evaluate a given rule against a stimulus:

```{r ch14_evaluate_rule}
# Evaluate a rule for a given stimulus.
# Returns the category prediction (0 or 1).
evaluate_rule <- function(rule, stimulus) {

  eval_cond <- function(dim_idx, thresh, op) {
    switch(op,
           ">" = stimulus[dim_idx] > thresh,
           "<" = stimulus[dim_idx] < thresh,
           stop("Invalid operation in rule"))
  }

  if (length(rule$dimensions) == 1) {
    condition_met <- eval_cond(rule$dimensions[1], rule$thresholds[1], rule$operations[1])

  } else if (length(rule$dimensions) == 2) {
    c1 <- eval_cond(rule$dimensions[1], rule$thresholds[1], rule$operations[1])
    c2 <- eval_cond(rule$dimensions[2], rule$thresholds[2], rule$operations[2])
    condition_met <- if (rule$logic == "AND") c1 & c2
                     else if (rule$logic == "OR") c1 | c2
                     else stop("Invalid logic in 2D rule")

  } else if (length(rule$dimensions) == 3 && !is.null(rule$outer_logic)) {
    # Nested rule: A outer_logic (B inner_logic C)
    # rule$outer_logic connects condition 1 with the (2 inner_logic 3) clause
    cA <- eval_cond(rule$dimensions[1], rule$thresholds[1], rule$operations[1])
    cB <- eval_cond(rule$dimensions[2], rule$thresholds[2], rule$operations[2])
    cC <- eval_cond(rule$dimensions[3], rule$thresholds[3], rule$operations[3])
    inner <- if (rule$inner_logic == "OR") cB | cC
             else if (rule$inner_logic == "AND") cB & cC
             else stop("Invalid inner_logic in nested rule")
    condition_met <- if (rule$outer_logic == "AND") cA & inner
                     else if (rule$outer_logic == "OR") cA | inner
                     else stop("Invalid outer_logic in nested rule")

  } else {
    stop("Rule dimension count not supported (only 1, 2, or 3-nested)")
  }

  ifelse(condition_met, rule$prediction_if_true, 1L - rule$prediction_if_true)
}

# --- Test ---
test_stimulus <- c(2.5, 3.8)
cat("Rule 1 Prediction:", evaluate_rule(rule_example_1d, test_stimulus),
    "(Expected: 0 — Height 2.5 is not > 3.5)\n")
cat("Rule 2 Prediction:", evaluate_rule(rule_example_2d, test_stimulus),
    "(Expected: 1 — Height > 2 TRUE, Position < 3 FALSE; AND => FALSE => predicts 1)\n")
# Nested rule test: Height > 2.5 TRUE; inner: Height > 3.5 FALSE OR Position < 2.0 FALSE => FALSE
# outer AND: TRUE AND FALSE => FALSE => predicts 0 (complementary of prediction_if_true = 1)
cat("Nested Rule Prediction:", evaluate_rule(rule_example_nested, test_stimulus),
    "(Expected: 0 — outer A=TRUE but inner B=FALSE OR C=FALSE => FALSE)\n")
```

---


## Phase 0: Two Models, Not One

The persona's Phase 0 interrogation is unusually consequential for this chapter, because rule-based categorisation has the property that the *cognitive theory we want to test* (the particle filter) and the *model we can fit with HMC* (a marginalised mixture over fixed rules) are not the same model — and pretending otherwise has cost real research conclusions in the past. We surface the structural commitments before any code is written.

> **Q1 — What is the cognitive theory?**
>
> The cognitive theory is **Model A**: a sequential agent that maintains a finite working set of $N_{\text{particles}}$ candidate rules drawn from a structured grammar, updates the weights on those rules via Bayes's rule given new feedback, resamples when effective sample size drops below threshold, and chooses categories by weighted vote. The agent's parameters are:
>
> * $\varepsilon$ — error rate (probability that the agent applies its rule incorrectly on any trial)
> * $N_{\text{particles}}$ — the number of hypotheses the agent maintains simultaneously, interpretable as a working-memory bound
> * $\text{max\_dims}$ — the maximum number of feature conditions in a rule
> * The rule grammar — which features, operators, and logical connectives are allowed
> * The resampling threshold — when the agent reorganises its hypothesis set
>
> The cognitive theory is the *whole* particle filter, including these structural choices. None of them is a nuisance.

> **Q2 — Can the cognitive theory be fitted with HMC? No. Why not?**
>
> Three reasons stack:
>
> 1. **The rule space is discrete and combinatorial.** Stan needs gradients with respect to a continuous parameter space. There is no natural notion of a derivative with respect to "which logical operator appears in condition 1" or "which feature dimension this rule uses."
> 2. **The resampling step is non-differentiable.** Even if we could embed individual rules in a continuous space, the resample operation is a stochastic discrete jump that no HMC sampler can integrate over.
> 3. **The latent particle trajectories cannot be marginalised analytically.** In Chapter 8 we marginalised mixture components by summing over them; here the "latent variable" is a high-dimensional trajectory of weighted rule sets, and the corresponding sum has no closed form for any reasonable rule grammar.
>
> The principled response is **simulator-based inference**. Neural Posterior Estimation (NPE) — a modern neural-network-based approach to simulation-based inference (SBI) — is the right tool for this exact situation: the simulator $M(\theta, V) \mapsto y$ exists (it is the `rule_particle_filter` R function), the parameters $\theta$ are well-defined, but the likelihood $p(y \mid \theta)$ has no tractable form. NPE trains a neural network (a normalising flow) on simulated $(\theta, y)$ pairs to learn the mapping from data to posterior distribution. Once trained, inference on any new dataset takes milliseconds — the upfront training cost is **amortized** across all subsequent fits (Radev et al. 2023, Deistler et al. 2025). This amortization is what makes the Chapter 15 comparison across all three cognitive theories executable: train one NPE network per model, then fit every subject in the dataset in seconds. Earlier drafts of this chapter used rejection ABC, which has three disadvantages that NPE resolves: (i) rejection ABC requires hand-crafted summary statistics and the posterior is biased by their insufficiency, (ii) the simulation budget scales linearly per subject with no reuse, and (iii) the accepted sample size is tiny (~100 draws from 10,000 simulations). NPE with a learned embedding network sidesteps all three. We implement NPE in §"Neural Posterior Estimation for Model A" via the `sbi` Python package, accessed from R through `reticulate`.

> **Q3 — What is Model B and how does it relate to Model A?**
>
> **Model B is *not* the same as Model A with a clever marginalisation trick.** It is a completely different generative model: a static, conditionally exchangeable mixture over a small fixed set of hand-picked candidate rules, with one continuous parameter $\varepsilon$ and uniform weights over $R$ rules. Its likelihood is closed-form Bernoulli, marginalised analytically with `log_sum_exp`. It has no learning dynamics, no resampling, no working-memory bound, no rule discovery, and the candidate rules are fixed by the modeller (us) by inspecting the stimulus layout.
>
> Model B fits cleanly in Stan and admits the full HMC validation pipeline. **But what it validates is Model B, not Model A.** A flat ECDF in an SBC plot for Model B means "Stan correctly inverts the marginalised mixture." It does *not* mean "Stan correctly inverts the particle filter," and it never could.
>
> What Model B *can* answer, when fitted to data generated from Model A: "given this particular candidate rule set, what value of $\varepsilon$ best explains the observed responses?" If the particle filter's converged rule happens to be in the candidate set, Model B's $\varepsilon$ posterior will track Model A's $\varepsilon$ closely. If the converged rule is *not* in the candidate set, Model B compensates by inflating $\varepsilon$ to absorb the systematic misclassifications. The $\varepsilon$ posterior is then a *biased* estimate of the cognitive parameter, biased in a direction the modeller cannot easily correct.
>
> **What we commit to here:** We treat Model B as a tractable proxy whose validation tests the proxy, not the cognitive theory. We use NPE to fit Model A. We compare the two posteriors as the chapter's main empirical result.

> **Q4 — Are there parameters being silently fixed in Model A?**
>
> Three structural choices that earlier drafts of this chapter buried in code:
>
> 1. **$N_{\text{particles}}$.** Earlier versions used 50, 100, 150, or 200 in different sub-sections without justification. We commit here to a single defended default: $N_{\text{particles}} = 100$, interpretable as a working-memory bound on the number of hypotheses an agent can maintain simultaneously. This is the value used everywhere in the chapter unless explicitly varied (in §"Sensitivity to $N_{\text{particles}}$" and in the NPE subsection where it can become a free parameter).
> 2. **$\text{max\_dims} = 2$.** The rule grammar allows up to two-condition rules (1D or 2D). We test sensitivity to this in §"Rule Grammar: Formal Specification" by also fitting under $\text{max\_dims} = 1$ and $\text{max\_dims} = 3$.
> 3. **Operator and logic sets.** $\{>, <\}$ and $\{\text{AND}, \text{OR}\}$. No negation, no equality. We support flat two-condition rules (1D or 2D) and nested three-condition rules of the form $A\ L_{\text{outer}}\ (B\ L_{\text{inner}}\ C)$; see §"Rule Grammar: Formal Specification." No deeper nesting. These are commitments to a specific rule language and they bias the inferential conclusions just as much as the prior on $\varepsilon$ does. The prior-sensitivity machinery (`priorsense::powerscale_sensitivity`) cannot detect bias from these choices because the bias is in the structural prior, not the parametric prior over $\varepsilon$.
> 4. **Resampling threshold = 0.5.** The particle filter resamples when $\text{ESS} < N_{\text{particles}} \times 0.5$. This controls how aggressively the agent reorganises its hypothesis set. We test sensitivity to this in §"Sensitivity to Resampling Threshold."

> **Q5 — Are the candidate rules in Model B truly prior to the data?**
>
> No, and this is the part of the chapter that earlier drafts admitted in a single bracketed remark and then ignored. **The seven candidate rules in Model B are picked by looking at the stimulus layout.** Strictly speaking, then, $p(\text{rule} = r) = 1/R$ is not a prior — it is an empirical approximation to the *posterior* over rules, conditional on having inspected the data-generating structure. The honest framings are either (i) treat the candidate set as a sensitivity analysis ("we test what fit a hand-picked rule set provides; results are conditional on this set"), or (ii) generate the candidate rules blind by running the particle filter on a *separately simulated* dataset and using its converged top rules. Both options are exercised in §"Discharging the Post-Hoc Rule Selection."

**Summary of Phase 0 commitments.** Two models, not one. Model A is the cognitive theory, has no Stan-tractable likelihood, and is fitted via Neural Posterior Estimation (NPE). Model B is a tractable proxy with a hand-picked rule set; it admits the full HMC validation pipeline but the validation tests the proxy, not the theory. $N_{\text{particles}}$ is fixed at 100 with an explicit cognitive interpretation; the resampling threshold is fixed at 0.5 with a sensitivity sweep. The rule grammar is documented formally below. The Stan candidate rule set is treated as a sensitivity-analysis condition, not as a prior. The chapter's main empirical contribution is the **gap between the NPE posterior on $\varepsilon$ under Model A and the Stan posterior on $\varepsilon$ under Model B** — that gap is the structural bias you accept when you replace the cognitive theory with a tractable proxy.

---

## Rule Grammar: Formal Specification

Earlier drafts of this chapter left the rule grammar inside the body of `generate_random_rule()` and let the reader infer it from the code. The persona's Phase 0 says: name the structural priors. Here is the formal grammar.

A rule $r$ in our grammar is a tuple

$$r = (D, T, O, L, c) \in \mathcal{R}$$

where:

* $D \in \mathcal{P}(\{1, 2\}) \setminus \emptyset$ with $|D| \leq \text{max\_dims}$ is the set of feature dimensions used by the rule. With $\text{max\_dims} = 2$, $|D| \in \{1, 2\}$.
* $T \in \mathbb{R}^{|D|}$ is the threshold vector, with the $k$-th element drawn uniformly from the observed range of feature $D_k$ (in our task: height $\in [1, 4]$ and position $\in [1, 4]$).
* $O \in \{>, <\}^{|D|}$ is the operator vector.
* $L \in \{\text{AND}, \text{OR}\}$ is the logical connective; only used when $|D| = 2$.
* $c \in \{0, 1\}$ is the prediction emitted when all conditions are satisfied. The complementary prediction is emitted otherwise.

**Extended grammar: nested three-condition rules.** We also support a second rule form that captures RULEX-style exception clauses (Nosofsky et al. 1994):

$$r_{\text{nested}} = (d_A, t_A, o_A, d_B, t_B, o_B, d_C, t_C, o_C, L_{\text{outer}}, L_{\text{inner}}, c)$$

evaluated as $\text{condition}_A\ L_{\text{outer}}\ (\text{condition}_B\ L_{\text{inner}}\ \text{condition}_C)$. The inner clause $(B\ L_{\text{inner}}\ C)$ acts as a disjunctive exception when $L_{\text{outer}} = \text{AND}$ and $L_{\text{inner}} = \text{OR}$: "the rule applies if A holds, but only unless the exception (B OR C) saves a stimulus from the incorrect prediction." Cognitive reading of `rule_example_nested`: "predict Cat 1 if Height > 2.5 AND (Height > 3.5 OR Position < 2.0)" — a tall-enough stimulus qualifies, *and* especially if it is very tall or at a low screen position. With only two features in the Kruschke task, the feature index $d_A$ may coincide with $d_B$ or $d_C$ without loss of generality; the thresholds differ, so the conditions are not redundant.

The number of distinct rules in this grammar is finite (although large). For $\text{max\_dims} = 2$ flat rules on a 2-feature task: $|\mathcal{R}| = 2 \cdot |T_{\text{grid}}| \cdot 2 \cdot 2 + \binom{2}{2} \cdot |T_{\text{grid}}|^2 \cdot 4 \cdot 2 \cdot 2$, where $|T_{\text{grid}}|$ is the threshold discretisation. Nested three-condition rules add $2^3 \cdot |T_{\text{grid}}|^3 \cdot 4 \cdot 2$ additional equivalence classes. With a continuous threshold space the grammar has uncountably many rules in both forms.

**Commitments encoded in this grammar:**

* No negation operator. A rule cannot say "NOT (Height > 2)." Negation can be partially recovered by flipping $c$ and choosing the complementary inequality, but compound negations cannot be expressed.
* No equality operator. Strict inequalities only.
* Maximum two conditions per flat rule, or three conditions in a nested rule. A reader who believes humans use deeper rule trees will need to extend the grammar further.
* Continuous thresholds. This defeats analytical marginalisation in Stan (see §"Phase 0: Two Models, Not One").
* Nested rules use a fixed two-level tree: one outer condition and one inner clause of two conditions. Deeper trees (e.g., A AND (B AND (C OR D))) are not implemented.

We will test sensitivity to `max_dims` in the dedicated subsection, and the operator/logic-set commitments are flagged in §"Limitations."

---

## The Challenge: Searching the Space of Rules

Evaluating a given rule is easy. The hard part of rule-based learning is finding the right rule. The number of possible rules (combinations of dimensions, thresholds, operations, logic) is vast. How can a learner efficiently search this space?

Unlike exemplar and prototype models, where the representation is straightforward, rule-based models need mechanisms for:

1. Generating candidate rules
2. Testing rules against observed data
3. Switching between rules when necessary
4. Maintaining uncertainty about which rule is correct

---

## The Bayesian Particle Filter: Tracking Rule Hypotheses (Model A)

The Bayesian particle filter provides an elegant solution to these challenges. Originally developed for tracking physical objects, particle filters have been adapted to model how humans track and update hypotheses. In our context, each "particle" represents a possible categorization rule, and the filter maintains and updates a distribution over these rules.

**This is Model A — the cognitive theory.** Everything in this section implements the cognitive model whose intractable likelihood will eventually require NPE to fit.

The core idea:

1. **Particles = Rules**: Each particle is a specific candidate rule.

2. **Weights = Belief**: Each particle has a weight $w_i$ representing the probability that this specific rule is the "true" rule. Initially, all weights are equal ($1/N$).

3. **Update Weights**: When a new stimulus and its correct category are observed, we update the weights. Rules that correctly predict the category get their weights increased; rules that predict incorrectly get their weights decreased. The amount of change depends on how likely the observation was under that rule.

4. **Resample**: Over time, many particles accumulate near-zero weight ("particle degeneracy"). To focus computational effort, we periodically resample: draw $N$ new particles with probability proportional to their weights. High-weight particles are likely to be selected multiple times; low-weight particles are likely to be dropped. After resampling, weights are reset to $1/N$.

### Why Model A's Likelihood is Intractable for HMC

Stan's HMC sampler requires a continuous, differentiable log posterior density. Individual rule particles are discrete, symbolic structures — lists containing integers, operators, and real-valued thresholds. The *space* of all possible rules is a combinatorial set of structured objects: there is no natural notion of a gradient with respect to "which logical operator appears in condition 1" or "which dimension is used." This is exactly the class of discrete latent variables that must be integrated out analytically before any Stan code is written (the same principle we applied to mixture components in Chapter 8).

For Model A, this analytical marginalisation is **not feasible in closed form** over a combinatorial rule space crossed with continuous thresholds and a non-differentiable resampling step. The particle filter therefore lives entirely in R, where it acts as its own approximate Bayesian inference engine via sequential Monte Carlo.

**This is the standard situation for which simulator-based inference was invented.** The Lintusaari et al. ABC review describes the pattern exactly:

> "Many recent models in biology describe nature to a high degree of accuracy but are not amenable to analytical treatment. The models can, however, be simulated on computers… While it is usually relatively easy to generate data from the models for any configuration of the parameters, the real interest is often focused on the inverse problem: the identification of parameter configurations that would plausibly lead to data that are sufficiently similar to the observed data."

Model A is exactly such a model: $\theta = (\varepsilon, N_{\text{particles}}, \text{max\_dims}, \ldots)$, $V$ is the random variables driving particle initialisation and resampling, $M$ is the `rule_particle_filter` function, and $y$ is the trial-by-trial response sequence. Section "Neural Posterior Estimation for Model A" implements NPE for this simulator. Until then, we treat the particle filter purely as a forward simulator.


### Step 1: Generating Initial Rule Particles

```{r ch14_generate_rules}
# Generate a single random rule.
# This is the implementation of the formal grammar from §"Rule Grammar".
generate_random_rule <- function(n_features,
                                 max_dims      = MAX_RULE_DIMS,
                                 feature_range = list(c(0, 5), c(0, 5)),
                                 nested_prob   = 0.25) {
  # With probability nested_prob, generate a nested 3-condition rule (if
  # there are >= 2 features and the grammar allows it). Otherwise generate
  # a flat 1- or 2-condition rule as before.
  use_nested <- (n_features >= 2) && (runif(1) < nested_prob)

  sample_thresh <- function(di) runif(1, feature_range[[di]][1], feature_range[[di]][2])

  if (use_nested) {
    # Three conditions; dimensions may repeat (same feature, different thresholds)
    dims_used  <- sample(seq_len(n_features), size = 3, replace = TRUE)
    thresholds <- mapply(sample_thresh, dims_used)
    operations <- sample(RULE_OPERATORS, 3, replace = TRUE)
    list(
      dimensions         = dims_used,
      thresholds         = thresholds,
      operations         = operations,
      outer_logic        = sample(RULE_LOGIC, 1),
      inner_logic        = sample(RULE_LOGIC, 1),
      prediction_if_true = sample(0:1, 1)
    )
  } else {
    num_dims  <- sample(seq_len(min(max_dims, n_features)), 1)
    dims_used <- sort(sample(seq_len(n_features), size = num_dims, replace = FALSE))
    thresholds <- numeric(num_dims)
    operations <- character(num_dims)
    for (k in seq_len(num_dims)) {
      di            <- dims_used[k]
      thresholds[k] <- sample_thresh(di)
      operations[k] <- sample(RULE_OPERATORS, 1)
    }
    list(
      dimensions        = dims_used,
      thresholds        = thresholds,
      operations        = operations,
      logic             = if (num_dims == 2) sample(RULE_LOGIC, 1) else NA,
      prediction_if_true = sample(0:1, 1)
    )
  }
}

# Initialize N particles with equal weights.
initialize_particles <- function(n_particles, n_features,
                                  max_dims = MAX_RULE_DIMS,
                                  feature_range = list(c(0, 5), c(0, 5))) {
  particles <- vector("list", n_particles)
  for (i in seq_len(n_particles))
    particles[[i]] <- generate_random_rule(n_features, max_dims, feature_range)
  list(particles = particles, weights = rep(1 / n_particles, n_particles))
}

# --- Example ---
n_particles_example <- 5
n_features_example  <- 2
initial_system <- initialize_particles(n_particles_example, n_features_example)
cat(paste("Initialized", n_particles_example, "particles. Example rule 1:\n"))
print(initial_system$particles[[1]])
cat("Initial weights:", initial_system$weights, "\n")
```

### Step 2: Updating Particle Weights

When we observe a stimulus $\vec{x}$ and its true category $y$, we update weight $w_i$ for each particle $i$. We introduce a small $\varepsilon$ (`error_prob`) to allow for occasional noise or misapplication:

$$P(y \mid \vec{x}, \text{rule}_i) = \begin{cases} 1 - \varepsilon & \text{if rule}_i \text{ predicts } y \\ \varepsilon & \text{otherwise} \end{cases}$$

The updated weight is $w_i' \propto w_i \times P(y \mid \vec{x}, \text{rule}_i)$, then renormalized.

```{r ch14_update_weights}
update_weights <- function(particles, weights, stimulus, true_category,
                           error_prob = 0.05) {
  n <- length(particles)
  new_w <- numeric(n)
  for (i in seq_len(n)) {
    pred      <- evaluate_rule(particles[[i]], stimulus)
    likelihood <- if (pred == true_category) 1 - error_prob else error_prob
    new_w[i]  <- weights[i] * likelihood
  }
  total <- sum(new_w)
  if (total > 1e-9) new_w / total else rep(1 / n, n)
}

# --- Example ---
cat("\n--- Example Weight Update ---\n")
updated_w <- update_weights(
  initial_system$particles, initial_system$weights,
  stimulus = c(4, 3), true_category = 1, error_prob = 0.1
)
preds <- sapply(initial_system$particles, evaluate_rule, stimulus = c(4, 3))
cat("Rule predictions for stimulus [4, 3]:", preds, "\n")
cat("Weights BEFORE:", round(initial_system$weights, 3), "\n")
cat("Weights AFTER: ", round(updated_w, 3), "\n")
```

### Step 3: Resampling Particles

If we only update weights, eventually most weights become near zero and only a few particles dominate — **particle degeneracy**. Resampling revitalizes the particle set by drawing $N$ new particles with probability proportional to weights, then resetting all weights to $1/N$.

A common heuristic is to resample when the **Effective Sample Size** (ESS) drops below a threshold (e.g., $N/2$). ESS is estimated as $1 / \sum_i w_i^2$. Low ESS indicates high degeneracy.

> **A note on ESS here vs. ESS in MCMC diagnostics.** ESS appears in two distinct
> contexts in this module. In Chapters 4–10 it measures how many effectively
> independent posterior draws an MCMC chain produces. Here it measures how many
> particles are effectively contributing to the filter's approximation, given
> their weights. Both are measures of statistical efficiency under a sampling
> procedure; the underlying intuition is the same but the contexts are
> different. The persona's diagnostic battery in Phase 3.0 is about the first
> kind (MCMC ESS); the resampling threshold here is about the second kind
> (particle ESS). Both names are correct and they refer to genuinely different
> things.

```{r ch14_resample_particles}
resample_particles <- function(particles, weights) {
  n          <- length(particles)
  idx        <- sample.int(n, size = n, replace = TRUE, prob = weights)
  new_p      <- particles[idx]
  new_w      <- rep(1 / n, n)
  list(particles = new_p, weights = new_w)
}

# --- Example ---
cat("\n--- Example Resampling ---\n")
skewed_w <- c(0.01, 0.60, 0.02, 0.35, 0.02)
cat("Weights BEFORE resampling:", skewed_w, "\n")
rs <- resample_particles(initial_system$particles, skewed_w)
cat("Weights AFTER resampling: ", rs$weights, "\n")
```

### Step 4: Making Predictions

The particle filter makes a categorization decision for a new stimulus $\vec{x}$ via a weighted vote across all particles:

$$P(\text{Choose Cat 1} \mid \vec{x}) = \sum_{i=1}^{N} w_i \times P(\text{Cat 1} \mid \vec{x}, \text{rule}_i)$$

where $P(\text{Cat 1} \mid \vec{x}, \text{rule}_i) = 1 - \varepsilon$ if rule $i$ predicts category 1, and $\varepsilon$ otherwise.

### Putting it Together: The Particle Filter Agent

```{r ch14_particle_filter_agent}
# Full particle filter agent — this is Model A.
# Returns a tibble of prob_cat1 and sim_response for each trial.
# Note: n_particles defaults to N_PARTICLES_FIXED (= 100), the structural
# commitment from Phase 0 Q4.
rule_particle_filter <- function(n_particles               = N_PARTICLES_FIXED,
                                 error_prob,
                                 obs,
                                 cat_one,
                                 max_dims                  = MAX_RULE_DIMS,
                                 resample_threshold_factor = RESAMPLE_THRESHOLD,
                                 quiet                     = TRUE) {
  n_trials   <- nrow(obs)
  n_features <- ncol(obs)
  fr         <- list(range(obs[, 1]), range(obs[, 2]))

  ps       <- initialize_particles(n_particles, n_features, max_dims, fr)
  particles <- ps$particles
  weights   <- ps$weights

  response_probs <- numeric(n_trials)

  for (i in seq_len(n_trials)) {
    if (!quiet && i %% 20 == 0) cat("Trial", i, "\n")

    stim <- as.numeric(obs[i, ])

    # ── Prediction ────────────────────────────────────────────────────────────
    preds        <- sapply(particles, evaluate_rule, stimulus = stim)
    p_cat1_rules <- ifelse(preds == 1, 1 - error_prob, error_prob)
    p            <- sum(weights * p_cat1_rules)
    response_probs[i] <- pmax(1e-9, pmin(1 - 1e-9, p))

    # ── Weight update ─────────────────────────────────────────────────────────
    weights <- update_weights(particles, weights, stim, cat_one[i], error_prob)

    # ── Resample if ESS is low ────────────────────────────────────────────────
    ess <- 1 / sum(weights^2)
    if (ess < n_particles * resample_threshold_factor) {
      rs        <- resample_particles(particles, weights)
      particles <- rs$particles
      weights   <- rs$weights
    }
  }

  tibble(prob_cat1 = response_probs,
         sim_response = rbinom(n_trials, 1, response_probs))
}

cat("Particle filter agent function defined.\n")
```

This implementation captures the core cognitive processes hypothesized by rule-based models: hypothesis generation, testing against evidence, belief updating, and adaptive focusing of the hypothesis set. **It is also the simulator that NPE will use in §"Neural Posterior Estimation for Model A."**

---

## Simulating Categorization Behavior

Let's simulate how Model A performs on the Kruschke (1993) task. Each agent receives its **own independently randomized stimulus order**, consistent with the path-dependence design from Chapters 11 and 12.

```{r ch14_simulate_rule_responses}
# Wrapper: generates per-agent schedule then calls rule_particle_filter.
simulate_rule_agent <- function(agent_id, n_particles, max_dims, error_prob,
                                stimulus_info, n_blocks, subject_seed) {
  schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  obs      <- as.matrix(schedule[, c("height", "position")])
  cat_one  <- schedule$category_feedback

  result <- rule_particle_filter(
    n_particles               = n_particles,
    error_prob                = error_prob,
    obs                       = obs,
    cat_one                   = cat_one,
    max_dims                  = max_dims,
    resample_threshold_factor = RESAMPLE_THRESHOLD,
    quiet                     = TRUE
  )

  schedule |>
    mutate(
      agent_id    = agent_id,
      n_particles = n_particles,
      max_dims    = max_dims,
      error_prob  = error_prob,
      prob_cat1   = result$prob_cat1,
      sim_response = result$sim_response,
      correct     = as.integer(category_feedback == sim_response)
    ) |>
    group_by(agent_id, n_particles, max_dims, error_prob) |>
    mutate(performance = cumsum(correct) / row_number()) |>
    ungroup()
}

param_df_rules <- expand_grid(
  agent_id    = 1:10,
  n_particles = c(50, 200),         # exploring sensitivity to N_particles
  max_dims    = c(1, 2),            # exploring sensitivity to grammar depth
  error_prob  = c(0.05, 0.2)
) |>
  mutate(subject_seed = agent_id)   # unique, reproducible seed per agent

rule_sim_file <- here("simdata", "ch14_rule_simulated_responses.csv")

if (regenerate_simulations || !file.exists(rule_sim_file)) {
  cat("Regenerating rule-based simulations...\n")
  rule_responses <- future_pmap_dfr(
    list(
      agent_id    = param_df_rules$agent_id,
      n_particles = param_df_rules$n_particles,
      max_dims    = param_df_rules$max_dims,
      error_prob  = param_df_rules$error_prob,
      subject_seed = param_df_rules$subject_seed
    ),
    simulate_rule_agent,
    stimulus_info = stimulus_info,
    n_blocks      = n_blocks,
    .options = furrr_options(seed = TRUE)
  )
  write_csv(rule_responses, rule_sim_file)
  cat("Simulations saved.\n")
} else {
  rule_responses <- read_csv(rule_sim_file, show_col_types = FALSE)
  cat("Simulations loaded.\n")
}

ggplot(rule_responses,
       aes(x = trial_within_subject, y = performance,
           color = factor(n_particles))) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.15,
               aes(fill = factor(n_particles)), linetype = 0) +
  facet_grid(
    max_dims ~ error_prob,
    labeller = labeller(
      max_dims   = function(x) paste("Max Dims:", x),
      error_prob = function(x) paste("Error Prob:", x)
    )
  ) +
  scale_color_viridis_d(option = "plasma", name = "N Particles") +
  scale_fill_viridis_d(option = "plasma", name = "N Particles") +
  labs(
    title    = "Rule-Based Model Learning Performance (Model A)",
    subtitle = "Facets show effect of max rule dimensions and error probability",
    x = "Trial", y = "Cumulative Accuracy (Mean ± SE)"
  ) +
  ylim(0.4, 1.0)
```

**Simulation Interpretation**: This visualization shows how Model A's performance is affected by:

* **Number of Particles**: More particles allow better exploration of the rule space, potentially leading to faster learning or higher final accuracy, though gains may diminish. **This parameter has cognitive content** (working-memory bound on hypothesis count) and is a candidate for NPE inference rather than fixing.
* **Maximum Rule Dimensions**: Allowing 2D rules (`max_dims = 2`) may be necessary if the true category boundary requires integrating information across dimensions. Setting `max_dims = 1` corresponds to a simpler grammar that cannot express conjunctive concepts.
* **Error Probability**: A higher `error_prob` makes the model more tolerant of imperfect rules, which can be helpful in noisy environments but may prevent convergence on a precise rule if one exists.


### Visualizing Rule Learning Over Time

A key feature of Model A is that the hypotheses are explicit. We can track which rules the particle filter believes are most likely at different points during learning.

```{r ch14_track_rule_learning}
track_rule_learning <- function(n_particles, error_prob, obs, cat_one,
                                max_dims                  = MAX_RULE_DIMS,
                                top_k                     = 1,
                                resample_threshold_factor = RESAMPLE_THRESHOLD) {
  n_trials  <- nrow(obs)
  n_features <- ncol(obs)
  fr         <- list(range(obs[, 1]), range(obs[, 2]))

  ps        <- initialize_particles(n_particles, n_features, max_dims, fr)
  particles <- ps$particles
  weights   <- ps$weights

  format_rule_info <- function(rule) {
    list(
      dimension1  = if (length(rule$dimensions) >= 1) rule$dimensions[1] else NA_integer_,
      operation1  = if (length(rule$operations)  >= 1) rule$operations[1]  else NA_character_,
      threshold1  = if (length(rule$thresholds)  >= 1) rule$thresholds[1]  else NA_real_,
      dimension2  = if (length(rule$dimensions) >= 2) rule$dimensions[2] else NA_integer_,
      operation2  = if (length(rule$operations)  >= 2) rule$operations[2]  else NA_character_,
      threshold2  = if (length(rule$thresholds)  >= 2) rule$thresholds[2]  else NA_real_,
      logic       = if (!is.na(rule$logic)) rule$logic else "N/A",
      prediction  = rule$prediction_if_true
    )
  }

  history <- vector("list", n_trials)
  for (i in seq_len(n_trials)) {
    stim <- as.numeric(obs[i, ])
    top_idx <- order(weights, decreasing = TRUE)[seq_len(min(top_k, n_particles))]
    history[[i]] <- map_dfr(top_idx, ~{
      tibble(trial = i, rank = which(top_idx == .x),
             weight = weights[.x], !!!format_rule_info(particles[[.x]]))
    })

    weights <- update_weights(particles, weights, stim, cat_one[i], error_prob)
    ess     <- 1 / sum(weights^2)
    if (ess < n_particles * resample_threshold_factor) {
      rs        <- resample_particles(particles, weights)
      particles <- rs$particles
      weights   <- rs$weights
    }
  }
  bind_rows(history)
}

# Run tracking with a fixed seed-1 schedule for reproducibility
track_schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = 1)

rule_tracking_data <- track_rule_learning(
  n_particles = N_PARTICLES_FIXED,
  error_prob  = 0.1,
  obs         = as.matrix(track_schedule[, c("height", "position")]),
  cat_one     = track_schedule$category_feedback,
  max_dims    = MAX_RULE_DIMS,
  top_k       = 1
)

# Format rule text for plotting
rule_to_text <- function(rule_row) {
  dim_names <- c("Height", "Position")
  if (is.na(rule_row$dimension2)) {
    sprintf("IF %s %s %.1f THEN Cat %d",
            dim_names[rule_row$dimension1], rule_row$operation1,
            rule_row$threshold1, rule_row$prediction)
  } else {
    sprintf("IF %s %s %.1f %s %s %s %.1f THEN Cat %d",
            dim_names[rule_row$dimension1], rule_row$operation1, rule_row$threshold1,
            rule_row$logic,
            dim_names[rule_row$dimension2], rule_row$operation2, rule_row$threshold2,
            rule_row$prediction)
  }
}

rule_tracking_plot_data <- rule_tracking_data |>
  filter(rank == 1) |>
  rowwise() |>
  mutate(rule_text = rule_to_text(cur_data())) |>
  ungroup() |>
  mutate(rule_block = consecutive_id(rule_text))

ggplot(rule_tracking_plot_data, aes(x = trial, y = weight)) +
  geom_line(aes(group = 1), color = "#0072B2") +
  geom_point(color = "#0072B2", size = 1.5) +
  geom_text(
    data = rule_tracking_plot_data |> distinct(rule_block, .keep_all = TRUE),
    aes(label = rule_text),
    hjust = 0, vjust = -0.5, size = 3, check_overlap = TRUE
  ) +
  scale_y_continuous(limits = c(0, 1), name = "Weight (Probability) of Top Rule") +
  scale_x_continuous(limits = c(0, total_trials + 5)) +
  labs(
    title    = "Evolution of the Top Rule Hypothesis Over Time (Model A)",
    subtitle = "Line shows the probability assigned to the single best rule at each trial.",
    x = "Trial"
  )
```

**Rule Evolution Interpretation**: This plot shows how the model's belief (weight) in its single best rule hypothesis changes over trials. Periods of high weight indicate the model has converged on a dominant rule; sharp drops followed by a new label indicate a hypothesis switch as the previous rule accumulated too many prediction errors. **Note that the converged rule is one specific rule** drawn from the grammar; the empirical fact of convergence is what makes the post-hoc selection of Model B's candidate rules tempting and dangerous.

### Visualizing the Decision Boundary

Rule-based models typically create axis-aligned, rectangular decision boundaries. Let's visualize the boundary created by the final, most probable rule found by Model A.

```{r ch14_decision_boundary}
get_final_rule_predictions <- function(n_particles, error_prob,
                                       obs_train, cat_one_train,
                                       grid_points, max_dims = MAX_RULE_DIMS,
                                       resample_threshold_factor = RESAMPLE_THRESHOLD) {
  n_features <- ncol(obs_train)
  fr         <- list(range(obs_train[, 1]), range(obs_train[, 2]))

  ps        <- initialize_particles(n_particles, n_features, max_dims, fr)
  particles <- ps$particles
  weights   <- ps$weights

  for (i in seq_len(nrow(obs_train))) {
    stim    <- as.numeric(obs_train[i, ])
    weights <- update_weights(particles, weights, stim, cat_one_train[i], error_prob)
    ess     <- 1 / sum(weights^2)
    if (ess < n_particles * resample_threshold_factor) {
      rs        <- resample_particles(particles, weights)
      particles <- rs$particles
      weights   <- rs$weights
    }
  }

  apply(as.matrix(grid_points), 1, function(pt) {
    preds <- sapply(particles, evaluate_rule, stimulus = pt)
    sum(weights * ifelse(preds == 1, 1 - error_prob, error_prob))
  })
}

grid_pts <- expand.grid(
  height   = seq(min(stimulus_info$height)   - 0.5,
                 max(stimulus_info$height)   + 0.5, length.out = 50),
  position = seq(min(stimulus_info$position) - 0.5,
                 max(stimulus_info$position) + 0.5, length.out = 50)
)

# Use the same track_schedule for boundary visualization
boundary_preds <- get_final_rule_predictions(
  n_particles   = 200,
  error_prob    = 0.1,
  obs_train     = as.matrix(track_schedule[, c("height", "position")]),
  cat_one_train = track_schedule$category_feedback,
  grid_points   = grid_pts[, c("height", "position")],
  max_dims      = MAX_RULE_DIMS
)

grid_decision_data <- grid_pts |>
  mutate(
    rule_prob_cat1 = boundary_preds,
    rule_decision  = as.integer(rule_prob_cat1 > 0.5)
  )

ggplot() +
  geom_tile(data = grid_decision_data,
            aes(x = position, y = height, fill = factor(rule_decision)),
            alpha = 0.4) +
  stat_contour(data = grid_decision_data,
               aes(x = position, y = height, z = rule_prob_cat1),
               breaks = 0.5, color = "black", linewidth = 0.8) +
  geom_point(data = stimulus_info,
             aes(x = position, y = height,
                 color = factor(category_true), shape = factor(category_true)),
             size = 3) +
  scale_fill_manual(values  = c("0" = "#E69F00", "1" = "#56B4E9"),
                    name    = "Predicted Cat") +
  scale_color_manual(values = c("0" = "#0072B2", "1" = "#D55E00"),
                     name    = "True Category") +
  scale_shape_manual(values = c("0" = 16, "1" = 17), name = "True Category") +
  labs(
    title    = "Rule-Based Model Decision Boundary (Model A, Particle Filter)",
    subtitle = "Weighted average of final particle set (N=200, error=0.1)",
    x = "Position Feature", y = "Height Feature"
  ) +
  coord_cartesian(
    xlim = range(grid_decision_data$position),
    ylim = range(grid_decision_data$height)
  )
```

**Boundary Interpretation**: The background color shows the predicted category for each point in the feature space, based on the weighted vote of the final rule particles. The black contour marks the 50% threshold. Notice the characteristic axis-aligned, rectangular character of the boundary regions — a direct consequence of the threshold-based rule structure — in contrast to the smooth, curved boundaries that would emerge from similarity-based models.

---

## Beyond Selection: Enhancing the Particle Filter for Rule Discovery

The Bayesian particle filter, as implemented above, provides a powerful framework for modeling how learners evaluate competing rule hypotheses. By maintaining a weighted set of particles and using resampling, it efficiently focuses on rules that best explain observed data.

However, a critical point: the basic filter primarily acts as a **rule selection** mechanism rather than a **rule generation** engine. It excels at identifying the most promising rules from an *initially provided set*. If the truly optimal rule is complex or unusual and not present in the initial random pool, the basic resampling step (which only duplicates existing rules) cannot discover it.

To create a model that more actively explores and discovers better rules during learning, the basic particle filter can be augmented with additional mechanisms:

1. **Mutation/Perturbation**: Introduce small, random variations to rules after resampling — slightly adjust a threshold, flip a comparison operator, or change the dimension used in a condition. This allows fine-tuning and local exploration around currently successful rules.

2. **Recombination/Crossover**: Combine elements from two successful "parent" rules to create new "child" rules — for example, the first condition of Parent A with the second condition of Parent B. This allows synthesizing features discovered independently by different rules.

3. **Targeted Rule Generation**: Generate new rules specifically designed to address stimuli consistently misclassified by the current particle set, or focus on features prevalent in high-weight rules.

4. **Complexity Adjustment**: Probabilistically add a condition to a successful 1D rule to explore whether a 2D rule performs better, or simplify a poorly performing 2D rule into its constituent 1D components.

Incorporating such exploration mechanisms transforms the particle filter from a pure selection device into a more adaptive **rule discovery system** capable of generating and testing rules that were not part of the initial random set. We do not implement these extensions here, but they represent a natural direction for more cognitively realistic models of rule learning. Note also that any of these mechanisms makes Model A's intractability *worse*, not better, from an HMC standpoint — they all add stochastic discrete jumps. NPE handles them transparently: the inference network only sees simulated $(\theta, y)$ pairs and does not care how the simulator produces them.


---

## Model B: A Tractable Stan Proxy via Marginalised Mixture

We now switch models. Everything in this section implements **Model B** — the static, conditionally exchangeable mixture over a fixed candidate rule set. Model B is **not** the same as Model A with a clever trick. It is a different generative process that admits HMC inference and the full validation pipeline. Validation of Model B tells us about Model B; it does not tell us about Model A. We will return to this distinction at every opportunity.

### Why a separate proxy model?

Model A cannot be fitted with HMC for the reasons in §"Why Model A's Likelihood is Intractable for HMC." But the persona's six-phase battery is designed around HMC and a tractable likelihood, and the model comparison framework in Chapter 15 needs ELPD scores that come from a proper Bayesian fit. Two responses:

1. **Build a tractable proxy in Stan and validate it under the standard pipeline.** This is what Model B does. The proxy is honest about being a proxy: it is not the cognitive theory, and its validation does not transfer.

2. **Fit Model A directly via simulator-based inference.** This is what NPE does. We do this in §"Neural Posterior Estimation for Model A." The chapter then compares the two posteriors as its main empirical contribution.

This subsection builds Model B. The NPE subsection comes later because pedagogically it makes sense to first see what the tractable proxy can and cannot do.

### Model B's Generative Structure

Model B enumerates a fixed set of candidate rules and marginalises over them analytically. This is identical in structure to the finite mixture models of Chapter 8, where discrete latent variables (mixture components) were summed out before any Stan code was written. Here the discrete latent variable is "which rule from the candidate set the agent follows." The marginalised likelihood for trial $i$ is:

$$p(y_i \mid x_i, \varepsilon) = \sum_{r=1}^{R} p(\text{rule} = r) \cdot p(y_i \mid x_i, \text{rule}_r, \varepsilon)$$

with a uniform prior over rules $p(\text{rule} = r) = 1/R$, implemented with `log_sum_exp` in Stan. Only the single continuous parameter $\varepsilon$ needs to be sampled.

We parameterise $\varepsilon$ on the logit scale — $\text{logit}(\varepsilon) \sim \text{Normal}(-2, 1)$ — which is unconstrained, avoids the boundary artefacts that afflict bounded parameters in HMC, and places prior mass primarily on low error rates (prior median $\varepsilon \approx 0.12$). This is the same parameterisation philosophy as `log_c` in Chapter 11.

**Critical structural facts about Model B that distinguish it from Model A:**

* **No learning dynamics.** Model B treats all trials as conditionally exchangeable given $\varepsilon$. There is no trial-by-trial weight update, no resampling, no rule discovery. The agent's behaviour is determined entirely by which rule it follows (drawn once from the candidate set) and the noise rate $\varepsilon$.
* **Fixed candidate rule set.** The $R = 7$ rules are picked by inspecting the Kruschke stimulus layout (5 flat rules + 2 nested rules of the form $A \text{ op } (B \text{ inner } C)$). Phase 0 Q5 surfaced this as a post-hoc choice; §"Discharging the Post-Hoc Rule Selection" tests sensitivity.
* **Closed-form Bernoulli likelihood.** Marginalised analytically with `log_sum_exp`. HMC handles it cleanly.
* **No path-dependence.** Standard PSIS-LOO is technically valid (see §"LFO-CV Does Not Apply Here").

### A First Caveat: The Prior Predictive Check Uses Model A's Simulator

Before fitting Model B we want to verify that the prior on `logit_error_prob` generates only scientifically plausible categorisation curves. There is a subtle issue: **the Stan model does not have a forward simulator with learning dynamics.** Model B is conditionally i.i.d. in $\varepsilon$, so its forward simulation is just "draw $\varepsilon$, draw a rule, generate Bernoulli responses." This produces flat learning curves at level $1 - \varepsilon$, which is empirically nothing like what the data look like.

What earlier drafts of this chapter did — and what we keep here, but flag explicitly — is to draw $\varepsilon$ from the Stan prior and then use **Model A's simulator** to generate the prior predictive learning curves. This is *not* a prior predictive check for Model B in the strict sense. It is a prior predictive check for the cognitive theory under Model B's $\varepsilon$ prior. The two are not the same and the reader should know which is being shown.

```{r ch14_prior_predictive_check}
n_ppc_samples <- 500
set.seed(42)

ppc_logit_eps <- rnorm(n_ppc_samples, LOGIT_ERROR_PRIOR_MEAN, LOGIT_ERROR_PRIOR_SD)
ppc_eps       <- plogis(ppc_logit_eps)
ppc_sched     <- make_subject_schedule(stimulus_info, n_blocks, seed = 999)

prior_pred_curves <- map_dfr(seq_len(n_ppc_samples), function(s) {
  sim <- rule_particle_filter(
    n_particles = N_PARTICLES_FIXED,
    error_prob  = ppc_eps[s],
    obs         = as.matrix(ppc_sched[, c("height", "position")]),
    cat_one     = ppc_sched$category_feedback,
    max_dims    = MAX_RULE_DIMS,
    quiet       = TRUE
  )
  tibble(
    sample_id = s,
    trial     = seq_len(nrow(ppc_sched)),
    correct   = as.integer(sim$sim_response == ppc_sched$category_feedback)
  ) |>
    mutate(cum_acc = cumsum(correct) / row_number())
})

ppc_summary <- prior_pred_curves |>
  group_by(trial) |>
  summarise(
    q05 = quantile(cum_acc, 0.05), q25 = quantile(cum_acc, 0.25),
    q50 = median(cum_acc),
    q75 = quantile(cum_acc, 0.75), q95 = quantile(cum_acc, 0.95),
    .groups = "drop"
  )

ggplot(ppc_summary, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#E69F00", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#E69F00", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#B07D00", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Prior Predictive Curves Under Model A's Simulator With Model B's ε Prior",
    subtitle = "WARNING: This is NOT a prior predictive check for Model B in the strict sense.\nModel B has no learning dynamics; its true forward sim would be flat at 1 - ε.",
    x = "Trial", y = "Cumulative Accuracy"
  )
```

> **⚠️ Cross-model warning explicit.** The plot above shows prior predictive learning curves drawn from the *particle filter* (Model A) using $\varepsilon$ values sampled from the Stan model's (Model B's) prior. The Stan model fitted to the same data assumes no learning, so its true prior predictive distribution would consist of flat curves at $1 - \varepsilon$ — nothing like the rising curves shown here. We display these curves anyway because they are the most informative thing to show pedagogically (they convey the range of behaviour the cognitive theory produces under Model B's $\varepsilon$ prior), but they are a *cross-model* prior predictive, not a pure within-model one. The reader who wants the strict within-model check should compute it as: for each sample $s$, draw $\varepsilon_s$, draw a rule index $r_s \in \{1, \ldots, 5\}$ uniformly, and generate Bernoulli responses with probability $\Pr(\text{cat 1} \mid x_i, r_s, \varepsilon_s)$. The result is flat. It is also not very interesting.

### Stan Model

```{r ch14_stan_rule_model}
rule_simplified_stan <- "
// Model B: Marginalised Mixture over Fixed Candidate Rules
// This is NOT the cognitive theory (Model A, the particle filter).
// It is a tractable proxy that admits HMC validation.
// Validation of Model B tests Model B, not Model A.
//
// Architecture (identical in principle to Ch. 8 mixture models):
//   - Discrete latent variable = which rule the agent follows
//   - Marginalized analytically with log_sum_exp over a fixed rule set
//   - Only continuous parameter: logit_error_prob (unconstrained)
//   - error_prob = inv_logit(logit_error_prob)
//
// No sequential learning in this model: it treats all trials as conditionally
// exchangeable given error_prob and the rule mixture.
//
// The 7 rules below cover the main decision structures visible in the
// Kruschke stimulus layout, including two nested rules of the form
// A op (B inner C) that cannot be expressed as flat 2D conjunctions/disjunctions.
// See §'Discharging the Post-Hoc Rule Selection' for the sensitivity analysis.

data {
  int<lower=1> ntrials;
  int<lower=1> nfeatures;                       // must be 2 for this rule set
  array[ntrials] int<lower=0, upper=1> y;       // observed choices
  array[ntrials, nfeatures] real obs;           // obs[,1]=Height, obs[,2]=Position

  // Prior hyperparameters (passed as data for transparency)
  real prior_logit_error_mean;
  real<lower=0> prior_logit_error_sd;
}

parameters {
  real logit_error_prob;   // unconstrained; error_prob = inv_logit(logit_error_prob)
}

transformed parameters {
  real<lower=0, upper=1> error_prob = inv_logit(logit_error_prob);

  // Per-trial choice probability under each rule, computed once per leapfrog
  // step so log_lik in generated quantities is a one-line lookup.
  // (Same architectural discipline as Ch 11 GCM and Ch 12 prototype.)
  array[ntrials] real lp_marginal;

  for (i in 1:ntrials) {
    real h = obs[i, 1];
    real p = obs[i, 2];

    // ── Flat 1D and 2D rules (same as original 5) ──────────────────────────
    int pred1 = h > 2.5 ? 1 : 0;
    int pred2 = p < 2.5 ? 1 : 0;
    int pred3 = (h > 2.0 && p > 3.0) ? 1 : 0;
    int pred4 = (h < 2.0 || p < 2.0) ? 0 : 1;
    int pred5 = (h > 2.5 && p < 3.5) ? 1 : 0;
    // ── Nested rules: A op (B inner C) ─────────────────────────────────────
    // pred6: (h > 3.5) OR ((h > 1.5) AND (p < 1.5))
    //   Cognitive reading: "very tall, OR (some height AND very low position)"
    //   Perfectly classifies the 8 Kruschke stimuli.
    int pred6 = (h > 3.5 || (h > 1.5 && p < 1.5)) ? 1 : 0;
    // pred7: (h > 1.5) AND ((p < 2.5) OR (h > 3.5))
    //   Cognitive reading: "some height AND (low position OR very tall)"
    //   Also perfectly classifies the 8 Kruschke stimuli.
    int pred7 = (h > 1.5 && (p < 2.5 || h > 3.5)) ? 1 : 0;

    int R = 7;

    vector[R] lp;
    lp[1] = bernoulli_lpmf(y[i] | pred1 == 1 ? 1 - error_prob : error_prob);
    lp[2] = bernoulli_lpmf(y[i] | pred2 == 1 ? 1 - error_prob : error_prob);
    lp[3] = bernoulli_lpmf(y[i] | pred3 == 1 ? 1 - error_prob : error_prob);
    lp[4] = bernoulli_lpmf(y[i] | pred4 == 1 ? 1 - error_prob : error_prob);
    lp[5] = bernoulli_lpmf(y[i] | pred5 == 1 ? 1 - error_prob : error_prob);
    lp[6] = bernoulli_lpmf(y[i] | pred6 == 1 ? 1 - error_prob : error_prob);
    lp[7] = bernoulli_lpmf(y[i] | pred7 == 1 ? 1 - error_prob : error_prob);

    lp_marginal[i] = log_sum_exp(lp) - log(R);
  }
}

model {
  target += normal_lpdf(logit_error_prob | prior_logit_error_mean, prior_logit_error_sd);
  for (i in 1:ntrials) target += lp_marginal[i];
}

generated quantities {
  // log_lik[i] is just a lookup of lp_marginal[i] — no recomputation.
  vector[ntrials] log_lik;
  for (i in 1:ntrials) log_lik[i] = lp_marginal[i];
}
"

stan_file_rule <- here("stan", "ch14_rule_simplified.stan")
write_stan_file(rule_simplified_stan, dir = "stan/",
                basename = "ch14_rule_simplified.stan")
mod_rule_stan <- cmdstan_model(stan_file_rule)
cat("Model B (Stan rule mixture) compiled successfully.\n")
```

**Key aspects of the Model B Stan implementation:**

* **Parameters block**: A single scalar `logit_error_prob` (unconstrained). The constraint `error_prob ∈ (0,1)` is enforced implicitly by `inv_logit`.
* **Transformed parameters**: The marginalised log-density per trial (`lp_marginal[i]`) is computed once per leapfrog step. This is the same `transformed parameters` discipline as Chapter 11 (GCM) and Chapter 12 (Kalman prototype): if a quantity depends on parameters but is otherwise deterministic given the data, it belongs in `transformed parameters` so generated quantities can read it without recomputation.
* **Model block**: Prior on `logit_error_prob` then accumulation of `lp_marginal[i]` into the target.
* **Generated quantities**: `log_lik[i]` is a one-line lookup. Architecturally consistent with the other two chapters.

### Fitting Model B

```{r ch14_fit_stan_rule_model}
# Simulate responses from one Model A agent to use as 'observed' data
fit_schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = 42)
fit_sim <- rule_particle_filter(
  n_particles = N_PARTICLES_FIXED,
  error_prob  = 0.1,
  obs         = as.matrix(fit_schedule[, c("height", "position")]),
  cat_one     = fit_schedule$category_feedback,
  max_dims    = MAX_RULE_DIMS,
  quiet       = TRUE
)

rule_data_single <- list(
  ntrials                 = nrow(fit_schedule),
  nfeatures               = 2L,
  y                       = as.integer(fit_sim$sim_response),
  obs                     = as.matrix(fit_schedule[, c("height", "position")]),
  prior_logit_error_mean  = LOGIT_ERROR_PRIOR_MEAN,
  prior_logit_error_sd    = LOGIT_ERROR_PRIOR_SD
)

fit_filepath_rule <- here("simmodels", "ch14_rule_single_fit.rds")
if (regenerate_simulations || !file.exists(fit_filepath_rule)) {
  pf_rule <- mod_rule_stan$pathfinder(
    data      = rule_data_single,
    seed      = 42,
    num_paths = 4,
    refresh   = 0
  )

  fit_rule_stan <- mod_rule_stan$sample(
    data            = rule_data_single,
    init            = pf_rule,
    seed            = 42,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1000,
    iter_sampling   = 1500,
    refresh         = 300,
    adapt_delta     = 0.9
  )
  fit_rule_stan$save_object(fit_filepath_rule)
  cat("Model B fitted and saved.\n")
} else {
  if (file.exists(fit_filepath_rule)) {
    fit_rule_stan <- readRDS(fit_filepath_rule)
    cat("Loaded existing Model B fit.\n")
  } else {
    cat("Fit file not found. Set regenerate_simulations = TRUE.\n")
    fit_rule_stan <- NULL
  }
}

if (!is.null(fit_rule_stan)) {
  param_summary <- fit_rule_stan$summary(
    variables = c("logit_error_prob", "error_prob")
  )
  print(param_summary)
}
```


### Mandatory MCMC Diagnostic Battery

The persona's Phase 3.0 says: curate the diagnostics, even when they are nearly trivial. Use the same `diagnostic_summary_table()` helper as Chapters 11 and 12.

```{r ch14_diag_battery_model_B}
if (!is.null(fit_rule_stan)) {
  diag_tbl <- diagnostic_summary_table(
    fit_rule_stan,
    params = c("logit_error_prob", "error_prob")
  )
  print(diag_tbl)

  cat("\nThreshold reference (persona Phase 3.0):\n",
      "  divergences        = 0\n",
      "  rank-normalized Rhat < 1.01\n",
      "  bulk ESS           > 400\n",
      "  tail ESS           > 400\n",
      "  E-BFMI             > 0.2\n",
      "  MCSE_mean / SD     < 0.05\n", sep = "")

  p_trace <- bayesplot::mcmc_trace(
    fit_rule_stan$draws(c("logit_error_prob", "error_prob")),
    facet_args = list(ncol = 2)
  ) +
    ggtitle("Trace plots — should look like hairy caterpillars")

  p_rank <- bayesplot::mcmc_rank_overlay(
    fit_rule_stan$draws(c("logit_error_prob", "error_prob")),
    facet_args = list(ncol = 2)
  ) +
    ggtitle("Rank histograms — should be uniform across chains")

  print(p_trace / p_rank)
}
```

### Posterior Predictive Check (Model B)

The posterior predictive check is honest about being a check on Model B, not on the cognitive theory. We sample $\varepsilon$ from the posterior, draw a rule index $r$ uniformly from the candidate set, and generate Bernoulli responses. The result is what Model B *actually* predicts: a flat-mean response sequence with whatever variation the rule mixture provides.

```{r ch14_ppc_model_B}
if (!is.null(fit_rule_stan)) {
  eps_draws  <- as_draws_df(fit_rule_stan$draws("error_prob"))$error_prob
  n_draws    <- length(eps_draws)
  obs_mat    <- as.matrix(fit_schedule[, c("height", "position")])
  R_rules    <- 7L
  n_obs      <- nrow(fit_schedule)

  # Per-rule predictions on the observed stimulus sequence (7 rules: 5 flat + 2 nested)
  rule_preds <- matrix(0L, nrow = R_rules, ncol = n_obs)
  for (i in seq_len(n_obs)) {
    h <- obs_mat[i, 1]; p <- obs_mat[i, 2]
    rule_preds[1, i] <- as.integer(h > 2.5)
    rule_preds[2, i] <- as.integer(p < 2.5)
    rule_preds[3, i] <- as.integer(h > 2.0 && p > 3.0)
    rule_preds[4, i] <- as.integer(!(h < 2.0 || p < 2.0))
    rule_preds[5, i] <- as.integer(h > 2.5 && p < 3.5)
    rule_preds[6, i] <- as.integer(h > 3.5 || (h > 1.5 && p < 1.5))  # nested: OR with inner AND
    rule_preds[7, i] <- as.integer(h > 1.5 && (p < 2.5 || h > 3.5))  # nested: AND with inner OR
  }

  set.seed(2026)
  yrep <- matrix(0L, nrow = n_draws, ncol = n_obs)
  for (s in seq_len(n_draws)) {
    r_idx     <- sample.int(R_rules, 1)
    rule_pred <- rule_preds[r_idx, ]
    p_cat1    <- ifelse(rule_pred == 1, 1 - eps_draws[s], eps_draws[s])
    yrep[s, ] <- rbinom(n_obs, 1, p_cat1)
  }

  # Cumulative-accuracy ribbons vs. observed
  obs_correct <- as.integer(fit_sim$sim_response == fit_schedule$category_feedback)
  obs_cum_acc <- cumsum(obs_correct) / seq_len(n_obs)
  yrep_correct <- sweep(yrep, 2,
                        as.numeric(fit_schedule$category_feedback),
                        FUN = function(a, b) as.numeric(a == b))
  yrep_cum_acc <- t(apply(yrep_correct, 1, function(r) cumsum(r) / seq_along(r)))

  cum_acc_summary <- tibble(
    trial = seq_len(n_obs),
    q05   = apply(yrep_cum_acc, 2, quantile, 0.05),
    q25   = apply(yrep_cum_acc, 2, quantile, 0.25),
    q50   = apply(yrep_cum_acc, 2, quantile, 0.50),
    q75   = apply(yrep_cum_acc, 2, quantile, 0.75),
    q95   = apply(yrep_cum_acc, 2, quantile, 0.95),
    obs   = obs_cum_acc
  )

  ggplot(cum_acc_summary, aes(x = trial)) +
    geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
    geom_line(aes(y = q50), color = "#0072B2", linewidth = 1) +
    geom_line(aes(y = obs), color = "#D55E00", linewidth = 1) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
    labs(
      title    = "Posterior Predictive Check (Model B)",
      subtitle = "Blue: Model B's posterior predictive\nOrange: observed (generated by Model A)\nThe gap is informative — Model B has no learning curve.",
      x = "Trial", y = "Cumulative accuracy"
    ) +
    coord_cartesian(ylim = c(0.4, 1))
}
```

**Reading this plot.** The blue ribbons are Model B's actual posterior predictive cumulative accuracy. They are **flat across trials**, because Model B has no learning dynamics. The orange line is the observed data, generated by Model A, which *does* have learning dynamics. **The systematic gap is the structural misspecification of Model B**, made visible. This is exactly what the chapter wants the reader to see: Model B can recover an $\varepsilon$ value that fits the *average* response, but it cannot reproduce the trial-by-trial structure because that structure comes from a process Model B does not have.

### Prior–Posterior Update Plot (Model B)

```{r ch14_prior_posterior_update}
if (!is.null(fit_rule_stan)) {
  draws_df <- as_draws_df(fit_rule_stan$draws(c("logit_error_prob", "error_prob")))

  set.seed(7)
  n_prior <- nrow(draws_df)
  prior_logit_eps <- rnorm(n_prior, LOGIT_ERROR_PRIOR_MEAN, LOGIT_ERROR_PRIOR_SD)
  prior_eps       <- plogis(prior_logit_eps)

  prior_post_df <- bind_rows(
    tibble(scale = "logit_error_prob (computational)",
           value = prior_logit_eps, source = "prior"),
    tibble(scale = "logit_error_prob (computational)",
           value = draws_df$logit_error_prob, source = "posterior"),
    tibble(scale = "error_prob (interpretable)",
           value = prior_eps, source = "prior"),
    tibble(scale = "error_prob (interpretable)",
           value = draws_df$error_prob, source = "posterior")
  )

  true_lines <- tibble(
    scale = c("logit_error_prob (computational)", "error_prob (interpretable)"),
    value = c(qlogis(0.1), 0.1)
  )

  ggplot(prior_post_df, aes(x = value, fill = source, color = source)) +
    geom_density(alpha = 0.4, linewidth = 0.6) +
    geom_vline(data = true_lines,
               aes(xintercept = value),
               color = "red", linetype = "dashed") +
    facet_wrap(~ scale, scales = "free") +
    scale_fill_manual(values = c(prior = "#999999", posterior = "#E69F00")) +
    scale_color_manual(values = c(prior = "#666666", posterior = "#B07D00")) +
    labs(
      title    = "Prior → Posterior Update for Model B",
      subtitle = "Red dashed line = Model A's true generating ε = 0.1\nGap between posterior centre and the dashed line is the structural-bias signature.",
      x = NULL, y = "Density"
    )
}
```

The posterior should be substantially narrower than the prior. The relationship of the posterior centre to the dashed red line (the true $\varepsilon$ used to generate the data) is the **structural bias signature**: if the dashed line falls in the bulk of the posterior, the proxy is approximately faithful for this data; if the dashed line is in the tail or outside the posterior, the proxy is biased and the size of the offset is informative about how badly the candidate rule set misses the converged rule.

### Randomised LOO-PIT (Model B)

The same calibration check we used in Chapters 11 and 12, applied to Model B.

```{r ch14_loo_pit_randomised}
if (!is.null(fit_rule_stan)) {
  log_lik_mat <- fit_rule_stan$draws("log_lik", format = "matrix")
  loo_obj     <- loo::loo(log_lik_mat, cores = 4)

  log_w <- loo::weights.importance_sampling(loo_obj$psis_object,
                                             normalize = TRUE,
                                             log = TRUE)

  # Compute LOO predictive probability of cat 1 from Model B
  # For this we need per-trial p_cat1 under each posterior draw and rule.
  # Easiest: sample rule per draw, compute p_cat1 = (1 - eps if rule says 1 else eps),
  # then average using LOO weights.
  eps_draws_full <- as_draws_df(fit_rule_stan$draws("error_prob"))$error_prob
  n_draws_full   <- length(eps_draws_full)
  n_obs          <- nrow(fit_schedule)

  set.seed(11)
  rule_idx_per_draw <- sample.int(7L, n_draws_full, replace = TRUE)

  # Per (draw, trial) p_cat1
  p_draws <- matrix(0, nrow = n_draws_full, ncol = n_obs)
  for (s in seq_len(n_draws_full)) {
    r <- rule_idx_per_draw[s]
    pred_row <- rule_preds[r, ]
    p_draws[s, ] <- ifelse(pred_row == 1,
                            1 - eps_draws_full[s],
                            eps_draws_full[s])
  }

  loo_p_cat1 <- vapply(seq_len(n_obs), function(i) {
    sum(exp(log_w[, i]) * p_draws[, i])
  }, numeric(1))

  y_obs <- as.integer(fit_sim$sim_response)
  set.seed(2026)
  rloo_pit <- vapply(seq_along(y_obs), function(i) {
    p_i <- loo_p_cat1[i]
    if (y_obs[i] == 1) runif(1, 1 - p_i, 1) else runif(1, 0, 1 - p_i)
  }, numeric(1))

  ggplot(tibble(rloo_pit = rloo_pit), aes(x = rloo_pit)) +
    geom_histogram(aes(y = after_stat(density)), bins = 20,
                   fill = "#E69F00", color = "white") +
    geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
    labs(
      title    = "Randomised LOO-PIT (Model B)",
      subtitle = "If Model B is calibrated for this data, the histogram should look uniform on [0, 1]",
      x = "Randomised PIT value", y = "Density"
    ) +
    coord_cartesian(ylim = c(0, 2.5))
}
```

A roughly uniform histogram says Model B is calibrated *for the proxy task* (predicting individual binary responses given the posterior over $\varepsilon$ and the rule mixture). It does *not* say Model B is calibrated as a model of the cognitive process — that would require the NPE pipeline to validate Model A directly.

### LOO with Pareto-k as a Misspecification Diagnostic

```{r ch14_loo_rule}
if (!is.null(fit_rule_stan)) {
  loo_rule <- loo::loo(log_lik_mat, cores = 4)
  print(loo_rule)
  plot(loo_rule, label_points = TRUE,
       main = "PSIS-LOO Pareto-k: Model B")
}
```

Unlike Chapters 11 and 12, here the LOO ELPD is **technically valid** (Model B is conditionally exchangeable — see next subsection). But the Pareto-$\hat{k}$ values are still informative as a *misspecification diagnostic*: when Model B is misspecified relative to the true generating process (Model A), $\hat{k}$ values are likely to be elevated on the trials where the proxy fails most badly. Plotting $\hat{k}$ against trial number and stimulus identity localises the misspecification.

```{r ch14_pareto_k_by_trial_and_stimulus}
if (!is.null(fit_rule_stan)) {
  k_values <- loo_rule$diagnostics$pareto_k
  k_df <- tibble(
    trial      = seq_along(k_values),
    k_hat      = k_values,
    stimulus   = fit_schedule$stimulus_id,
    height     = fit_schedule$height,
    position   = fit_schedule$position,
    true_cat   = fit_schedule$category_feedback,
    obs_y      = fit_sim$sim_response
  )

  p_trial <- ggplot(k_df, aes(x = trial, y = k_hat)) +
    geom_point(aes(color = factor(true_cat)), size = 1.5) +
    geom_hline(yintercept = c(0.5, 0.7, 1.0),
               linetype = c("dotted", "dashed", "solid"),
               color = "red", alpha = 0.6) +
    scale_color_manual(values = c("0" = "#0072B2", "1" = "#D55E00"),
                       name = "True cat") +
    labs(
      title    = "Pareto-k by trial",
      subtitle = "Red lines: 0.5 (monitor), 0.7 (problematic), 1.0 (failed)",
      x = "Trial", y = "Pareto k"
    )

  p_stim <- ggplot(k_df, aes(x = factor(stimulus), y = k_hat)) +
    geom_boxplot(fill = "#E69F00", alpha = 0.6) +
    geom_jitter(width = 0.15, alpha = 0.4, size = 1) +
    geom_hline(yintercept = 0.7, linetype = "dashed", color = "red") +
    labs(
      title    = "Pareto-k by stimulus identity",
      subtitle = "If a particular stimulus has consistently elevated k_hat,\nthat stimulus is where the rule set fails",
      x = "Stimulus ID", y = "Pareto k"
    )

  print(p_trial / p_stim)
}
```

**Reading the Pareto-k plots as a misspecification diagnostic.** If the high-$\hat{k}$ values cluster on the *boundary stimuli* of the Kruschke set — the ones whose true category cannot be determined by any single rule from the candidate set — that is direct evidence that the rule set is missing the relevant boundary structure. If the high-$\hat{k}$ values cluster at *late trials*, that is a signal that Model A's particle filter has converged to a rule the candidate set does not contain, and Model B is using $\varepsilon$ inflation to absorb the systematic late-trial errors. In both cases, the diagnostic does not just say "the model is wrong" — it says **where the model is wrong**, which is much more useful.

### LFO-CV Does Not Apply Here

> **A back-reference, not a caveat to repeat.** Chapters 11 and 12 implement LFO-CV via `psis_lfo_gcm()` and `psis_lfo_kalman()` because the GCM and the Kalman prototype are path-dependent: trial $t$'s likelihood depends on the entire history of trials $1, \ldots, t-1$ through the memory or filter state. **Model B has no such path-dependence.** All trials are conditionally independent given $\varepsilon$ and the rule mixture: leaving out trial $t$ does not affect the likelihood of any other trial. Standard PSIS-LOO is therefore *exactly correct* for Model B and the LFO-CV machinery is unnecessary. We say this once here so the reader does not encounter the same caveat for the third time after seeing it twice in the previous chapters.
>
> A subtler point: when Chapter 15 compares Model B against the GCM (Chapter 11) and the Kalman prototype (Chapter 12), the comparison should be reported under each model's *correct* validation tool — LFO-CV for the path-dependent ones, standard LOO for Model B. ELPD differences across these tools are not directly commensurable on a fine scale, but they are commensurable enough to detect large effects. The bigger concern in Chapter 15 is the cross-architecture mismatch (Model B isn't even fitting the cognitive theory), and that concern is what the NPE subsection below resolves.

### Prior Sensitivity Analysis

```{r ch14_prior_sensitivity}
if (!is.null(fit_rule_stan)) {
  ps <- priorsense::powerscale_sensitivity(fit_rule_stan)
  print(ps)
  priorsense::powerscale_plot_dens(ps, variables = c("logit_error_prob", "error_prob"))
}
```

`powerscale_sensitivity()` reports a sensitivity index $\psi$ for each parameter. **One important limitation:** it perturbs only the *explicit* prior on `logit_error_prob`. It cannot detect sensitivity to the *implicit* prior over the rule set, which is the much bigger source of structural bias for Model B. Low $\psi$ here means Model B is robust to the parametric prior on $\varepsilon$ — it does *not* mean Model B is robust to the choice of candidate rules. The next subsection addresses that.


---

## Parameter Recovery (Model B)

We test recovery systematically across the plausible range of `error_prob`, with multiple independent datasets per true value, each with its own randomized stimulus schedule. **Caveat upfront:** the data are generated by Model A, fitted by Model B. If Model A converges to a rule that is not in Model B's candidate set, recovery will be biased. This is the structural-bias signature again.

```{r ch14_parameter_recovery}
run_rule_recovery <- function(true_error_prob, iteration,
                              stimulus_info, n_blocks) {
  subject_seed <- iteration * 1000 + round(true_error_prob * 1000)
  sched        <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)

  sim <- rule_particle_filter(
    n_particles = N_PARTICLES_FIXED,
    error_prob  = true_error_prob,
    obs         = as.matrix(sched[, c("height", "position")]),
    cat_one     = sched$category_feedback,
    max_dims    = MAX_RULE_DIMS,
    quiet       = TRUE
  )

  stan_dat <- list(
    ntrials                = nrow(sched),
    nfeatures              = 2L,
    y                      = as.integer(sim$sim_response),
    obs                    = as.matrix(sched[, c("height", "position")]),
    prior_logit_error_mean = LOGIT_ERROR_PRIOR_MEAN,
    prior_logit_error_sd   = LOGIT_ERROR_PRIOR_SD
  )

  pf <- tryCatch(
    mod_rule_stan$pathfinder(data = stan_dat, num_paths = 4,
                              refresh = 0, seed = subject_seed),
    error = function(e) NULL
  )
  init_val <- if (!is.null(pf)) pf else 0.5

  fit <- tryCatch(
    mod_rule_stan$sample(
      data = stan_dat, init = init_val,
      seed = subject_seed, chains = 2, parallel_chains = 2,
      iter_warmup = 800, iter_sampling = 1000,
      refresh = 0, adapt_delta = 0.9
    ),
    error = function(e) { cat("Fit error:", e$message, "\n"); NULL }
  )
  if (is.null(fit)) return(NULL)

  draws_eps <- fit$draws("error_prob", format = "df")
  tibble(
    true_error_prob   = true_error_prob,
    iteration         = iteration,
    estimated_eps_mean = mean(draws_eps$error_prob),
    est_lower_95      = quantile(draws_eps$error_prob, 0.025),
    est_upper_95      = quantile(draws_eps$error_prob, 0.975),
    had_divergence    = any(fit$diagnostic_summary(quiet = TRUE)$num_divergent > 0)
  )
}

eps_values_to_test <- c(0.05, 0.1, 0.15, 0.2, 0.3)
n_recovery_iters   <- 5   # increase to 20–50 for publication-quality recovery

recovery_file_rule <- here("simdata", "ch14_rule_recovery_results.rds")
if (regenerate_simulations || !file.exists(recovery_file_rule)) {
  cat("Running parameter recovery for Model B...\n")
  recovery_plan <- expand_grid(
    true_error_prob = eps_values_to_test,
    iteration       = seq_len(n_recovery_iters)
  )
  rule_recovery_results <- future_pmap_dfr(
    list(true_error_prob = recovery_plan$true_error_prob,
         iteration       = recovery_plan$iteration),
    run_rule_recovery,
    stimulus_info = stimulus_info,
    n_blocks      = n_blocks,
    .options = furrr_options(seed = TRUE)
  )
  saveRDS(rule_recovery_results, recovery_file_rule)
  cat("Recovery complete and saved.\n")
} else {
  rule_recovery_results <- readRDS(recovery_file_rule)
  cat("Recovery results loaded.\n")
}

ggplot(rule_recovery_results,
       aes(x = true_error_prob, y = estimated_eps_mean)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_errorbar(aes(ymin = est_lower_95, ymax = est_upper_95,
                    color = had_divergence),
                width = 0.005, alpha = 0.7) +
  geom_point(aes(color = had_divergence), size = 3) +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00"),
                     name = "Had divergence") +
  labs(
    title    = "Parameter Recovery: Model B (error_prob)",
    subtitle = paste0("Points: posterior mean; bars: 95% CrI; n=",
                      n_recovery_iters,
                      " datasets per true value.\n",
                      "Data generated by Model A; fitted by Model B."),
    x = "True error_prob", y = "Estimated error_prob"
  )
```

**Discussion of Recovery Results.** Recovery of `error_prob` depends on whether Model B's candidate rule set contains rules that match what Model A converges to on each replicate. When the converged rule is in the candidate set, recovery is good. When it is not, Model B compensates by inflating $\varepsilon$ to absorb the systematic misclassifications, and the posterior mean shifts up — the points drift above the diagonal. This drift is the *quantitative* version of the structural-bias warning.

---

## Discharging the Post-Hoc Rule Selection

Phase 0 Q5 surfaced the fact that Model B's candidate rules are picked by inspecting the stimulus layout, which makes them empirical, not prior. We discharge this in two complementary ways.

### Option 1: Sensitivity to the Candidate Rule Set

We define three alternative candidate rule sets and refit Model B on the same data:

* **Set A** — the 7 hand-picked rules (the ones in the Stan code above: 5 flat + 2 nested).
* **Set B** — 5 rules generated *blind* by running Model A on a separately-simulated training dataset and taking its top-weighted rules at the end of training. The training data is generated with $\varepsilon = 0.1$ on a different schedule than the fitting data. This is the "honest" version of the candidate set.
* **Set C** — 5 rules deliberately chosen to be *bad* (rules that disagree with most stimuli), as a stress test.

```{r ch14_rule_set_sensitivity}
# Helper to build a Stan model from a list of (predictor function) rules
build_stan_with_rules <- function(rule_predictors, name_suffix) {
  R <- length(rule_predictors)
  rule_lines <- paste(
    sprintf("    int pred%d = %s;", seq_len(R), rule_predictors),
    collapse = "\n"
  )
  lp_lines_model <- paste(
    sprintf("    lp[%d] = bernoulli_lpmf(y[i] | pred%d == 1 ? 1 - error_prob : error_prob);",
            seq_len(R), seq_len(R)),
    collapse = "\n"
  )

  stan_code <- sprintf("
data {
  int<lower=1> ntrials;
  int<lower=1> nfeatures;
  array[ntrials] int<lower=0, upper=1> y;
  array[ntrials, nfeatures] real obs;
  real prior_logit_error_mean;
  real<lower=0> prior_logit_error_sd;
}
parameters { real logit_error_prob; }
transformed parameters {
  real<lower=0, upper=1> error_prob = inv_logit(logit_error_prob);
  array[ntrials] real lp_marginal;
  for (i in 1:ntrials) {
    real h = obs[i, 1];
    real p = obs[i, 2];
%s
    int R = %d;
    vector[R] lp;
%s
    lp_marginal[i] = log_sum_exp(lp) - log(R);
  }
}
model {
  target += normal_lpdf(logit_error_prob | prior_logit_error_mean, prior_logit_error_sd);
  for (i in 1:ntrials) target += lp_marginal[i];
}
generated quantities {
  vector[ntrials] log_lik;
  for (i in 1:ntrials) log_lik[i] = lp_marginal[i];
}
", rule_lines, R, lp_lines_model)

  fname <- here("stan", sprintf("ch14_rule_%s.stan",  name_suffix)
  write_stan_file(stan_code, dir = "stan/",
                  basename = sprintf("ch14_rule_%s.stan", name_suffix))
  cmdstan_model(fname)
}

# Set A: hand-picked rules — 5 flat + 2 nested (replicates Model B exactly)
rules_A <- c(
  "h > 2.5 ? 1 : 0",
  "p < 2.5 ? 1 : 0",
  "(h > 2.0 && p > 3.0) ? 1 : 0",
  "(h < 2.0 || p < 2.0) ? 0 : 1",
  "(h > 2.5 && p < 3.5) ? 1 : 0",
  "(h > 3.5 || (h > 1.5 && p < 1.5)) ? 1 : 0",
  "(h > 1.5 && (p < 2.5 || h > 3.5)) ? 1 : 0"
)

# Set B: blind rules from a fresh Model A run on a separate seed
get_blind_rules <- function(seed = 7) {
  blind_sched <- make_subject_schedule(stimulus_info, n_blocks, seed = 50000 + seed)
  obs_b   <- as.matrix(blind_sched[, c("height", "position")])
  cat_b   <- blind_sched$category_feedback
  res     <- track_rule_learning(
    n_particles = N_PARTICLES_FIXED, error_prob = 0.1,
    obs = obs_b, cat_one = cat_b, max_dims = MAX_RULE_DIMS, top_k = 5
  )
  # Take the top 5 distinct rules at the final trial
  final <- res |> filter(trial == max(trial)) |>
    distinct(dimension1, operation1, threshold1,
             dimension2, operation2, threshold2,
             logic, prediction, .keep_all = TRUE) |>
    head(5)

  pred_strings <- character(nrow(final))
  for (k in seq_len(nrow(final))) {
    r <- final[k, ]
    feat1 <- if (r$dimension1 == 1) "h" else "p"
    cond1 <- sprintf("%s %s %.2f", feat1, r$operation1, r$threshold1)
    if (is.na(r$dimension2)) {
      pred_strings[k] <- sprintf("(%s) ? %d : %d", cond1, r$prediction, 1 - r$prediction)
    } else {
      feat2 <- if (r$dimension2 == 1) "h" else "p"
      cond2 <- sprintf("%s %s %.2f", feat2, r$operation2, r$threshold2)
      logic_op <- if (r$logic == "AND") "&&" else "||"
      pred_strings[k] <- sprintf("(%s %s %s) ? %d : %d",
                                  cond1, logic_op, cond2,
                                  r$prediction, 1 - r$prediction)
    }
  }
  pred_strings
}

rules_B <- get_blind_rules(seed = 7)

# Set C: deliberately bad rules
rules_C <- c(
  "h > 4.5 ? 1 : 0",        # virtually never fires
  "p > 4.5 ? 1 : 0",        # virtually never fires
  "h < 0.5 ? 1 : 0",        # virtually never fires
  "(h > 0.5 && p > 0.5) ? 1 : 0",  # always fires
  "(h > 0.5 || p > 0.5) ? 1 : 0"   # always fires
)

# Compile and fit each
mod_A <- build_stan_with_rules(rules_A, "setA")
mod_B <- build_stan_with_rules(rules_B, "setB")
mod_C <- build_stan_with_rules(rules_C, "setC")

fit_with_rules <- function(model, label) {
  fit <- model$sample(
    data = rule_data_single, init = 0.5, seed = 7,
    chains = 2, parallel_chains = 2,
    iter_warmup = 800, iter_sampling = 1000, refresh = 0, adapt_delta = 0.9
  )
  draws <- as_draws_df(fit$draws("error_prob"))
  tibble(
    rule_set     = label,
    posterior_mean = mean(draws$error_prob),
    posterior_lo = quantile(draws$error_prob, 0.025),
    posterior_hi = quantile(draws$error_prob, 0.975),
    elpd_loo     = loo::loo(fit$draws("log_lik", format = "matrix"))$estimates["elpd_loo", "Estimate"]
  )
}

rule_set_sensitivity_file <- here("simdata", "ch14_rule_set_sensitivity.rds")
if (regenerate_simulations || !file.exists(rule_set_sensitivity_file)) {
  rule_set_results <- bind_rows(
    fit_with_rules(mod_A, "A: hand-picked"),
    fit_with_rules(mod_B, "B: blind from Model A"),
    fit_with_rules(mod_C, "C: deliberately bad")
  )
  saveRDS(rule_set_results, rule_set_sensitivity_file)
} else {
  rule_set_results <- readRDS(rule_set_sensitivity_file)
}

if (exists("rule_set_results")) {
  print(rule_set_results)

  ggplot(rule_set_results,
         aes(x = rule_set, y = posterior_mean)) +
    geom_pointrange(aes(ymin = posterior_lo, ymax = posterior_hi)) +
    geom_hline(yintercept = 0.1, linetype = "dashed", color = "red") +
    labs(
      title = "Sensitivity of Model B's ε posterior to candidate rule set",
      subtitle = "Red dashed: true ε = 0.1 used to generate the data",
      x = NULL, y = "Posterior mean ± 95% CrI for ε"
    ) +
    coord_flip()
}
```

**Reading the rule-set sensitivity.** If the three posteriors agree, the candidate-set choice is innocuous. If they disagree — which is the expected outcome — the gap quantifies how much of Model B's estimated $\varepsilon$ is *cognitive* (genuinely tracking the agent's error rate) vs *structural* (compensating for missing rules). Set C should produce a wildly inflated $\varepsilon$ and a much worse ELPD, providing the upper bound on how much rule-set choice can move the posterior.

### Sensitivity to `max_dims` (Rule Grammar Depth)

```{r ch14_max_dims_sensitivity}
max_dims_results_file <- here("simdata", "ch14_max_dims_sensitivity.rds")
if (regenerate_simulations || !file.exists(max_dims_results_file)) {
  cat("Sweeping max_dims...\n")
  max_dims_results <- map_dfr(c(1, 2, 3), function(md) {
    tryCatch({
      sched <- make_subject_schedule(stimulus_info, n_blocks, seed = 42)
      sim_md <- rule_particle_filter(
        n_particles = N_PARTICLES_FIXED,
        error_prob  = 0.1,
        obs         = as.matrix(sched[, c("height", "position")]),
        cat_one     = sched$category_feedback,
        max_dims    = md
      )
      cum_acc <- mean(sim_md$sim_response == sched$category_feedback)
      tibble(max_dims = md, mean_accuracy = cum_acc)
    }, error = function(e) {
      tibble(max_dims = md, mean_accuracy = NA_real_)
    })
  })
  saveRDS(max_dims_results, max_dims_results_file)
} else {
  max_dims_results <- readRDS(max_dims_results_file)
}

if (exists("max_dims_results")) {
  print(max_dims_results)
  cat("Note: max_dims = 3 will fall back to 2 in our 2-feature task.\n",
      "The point of this sweep is to make the structural choice visible,\n",
      "not to discover anything new.\n")
}
```

`max_dims` is a structural commitment of the rule grammar. In our 2-feature task it can be 1 or 2 (3 is meaningless). In a higher-dimensional task it would be a genuine free parameter and should be either fixed with explicit cognitive justification or inferred via NPE.

### Sensitivity to Resampling Threshold

Phase 0 Q4 committed to `RESAMPLE_THRESHOLD = 0.5`. This controls how aggressively the particle filter reorganises its hypothesis set: lower values allow more particle degeneracy before resampling; higher values trigger resampling earlier. The cognitive interpretation is something like "how quickly the learner abandons low-performing hypotheses."

```{r ch14_resample_threshold_sensitivity}
resample_sensitivity_file <- here("simdata", "ch14_resample_threshold_sensitivity.rds")
if (regenerate_simulations || !file.exists(resample_sensitivity_file)) {
  cat("Sweeping resampling threshold...\n")
  resample_results <- map_dfr(c(0.25, 0.50, 0.75), function(rt) {
    accs <- vapply(1:10, function(rep) {
      sched <- make_subject_schedule(stimulus_info, n_blocks, seed = 42000 + rep)
      sim <- rule_particle_filter(
        n_particles               = N_PARTICLES_FIXED,
        error_prob                = 0.1,
        obs                       = as.matrix(sched[, c("height", "position")]),
        cat_one                   = sched$category_feedback,
        max_dims                  = MAX_RULE_DIMS,
        resample_threshold_factor = rt,
        quiet                     = TRUE
      )
      mean(sim$sim_response == sched$category_feedback)
    }, numeric(1))
    tibble(resample_threshold = rt,
           mean_accuracy = mean(accs), sd_accuracy = sd(accs))
  })
  saveRDS(resample_results, resample_sensitivity_file)
} else {
  resample_results <- readRDS(resample_sensitivity_file)
}

if (exists("resample_results")) {
  print(resample_results)
  cat("If mean accuracy varies substantially across thresholds,\n",
      "the resampling threshold is a consequential structural choice.\n",
      "If it is roughly stable, the choice is innocuous for this task.\n")
}
```

### Option 2: Treat Set A as a Sensitivity Condition, Not a Canonical Model

Rather than use Set A as *the* Model B and add Set B as a "robustness check," the honest framing is to report all three sets as variants of Model B with no privileged status. The reader sees three posteriors instead of one and decides for themselves how much of the variation is structural. This is the framing we adopt for the chapter 15 model comparison, where Model B is reported as a *family* of proxies indexed by the candidate set rather than as a single point estimate.

---


## Calibration: Two Different Checks for Two Different Things

Earlier drafts of this chapter ran one block under the header "Simulation-Based Calibration (SBC)" that generated data with the particle filter (Model A) and fitted Model B. **That is not SBC.** Talts–Vehtari SBC requires the data to be drawn from the *same* generative model that is then fitted, so that the joint $(\theta, y)$ distribution of the inference model is what gets calibrated. When the generative and inferential models differ, what you are running is closer to a robustness check than a calibration check, and labelling it SBC is misleading.

We split the misnamed block into two correctly-labelled checks: (1) a **Cross-Model Calibration Check** (data from Model A, fitted by Model B) and (2) a **Real SBC for Model B** (data from Model B itself, fitted by Model B). The first is a robustness check; the second is the actual Talts–Vehtari calibration test.

### Cross-Model Calibration Check (NOT SBC)

**Procedure.** For each iteration $i = 1, \ldots, S$: draw $\varepsilon_i$ from the prior, simulate one dataset with Model A (the particle filter) using $\varepsilon_i$, fit Model B to that dataset, record the rank of $\varepsilon_i$ within Model B's posterior. Compute the rank ECDF.

**What a flat ECDF means here.** It means: under the specific misspecification of Model B for data from Model A, $\varepsilon$ happens to be recovered with the right *marginal* coverage. It does **not** mean Model B is calibrated in the Talts–Vehtari sense, because the joint $(\theta, y)$ distribution of the inference model is not the joint distribution of the data-generating process.

**Why we run it anyway.** Because it answers a useful practical question: "when I fit Model B to data that came from a richer generative process (Model A), are the resulting credible intervals approximately well-calibrated for the *cognitive* parameter $\varepsilon$?" The answer to that question is informative even though it is not what SBC is technically supposed to answer.

```{r ch14_cross_model_check}
n_cross_iters <- 500   # use >= 1000 for production

cross_generator <- SBC_generator_function(
  function() {
    logit_eps_true <- rnorm(1, LOGIT_ERROR_PRIOR_MEAN, LOGIT_ERROR_PRIOR_SD)
    eps_true       <- plogis(logit_eps_true)

    sched <- make_subject_schedule(stimulus_info, n_blocks, seed = sample.int(1e8, 1))
    # NOTE: data drawn from Model A (the particle filter)
    sim <- rule_particle_filter(
      n_particles = N_PARTICLES_FIXED,
      error_prob  = eps_true,
      obs         = as.matrix(sched[, c("height", "position")]),
      cat_one     = sched$category_feedback,
      max_dims    = MAX_RULE_DIMS,
      quiet       = TRUE
    )

    list(
      variables = list(logit_error_prob = logit_eps_true),
      generated = list(
        ntrials                = nrow(sched),
        nfeatures              = 2L,
        y                      = as.integer(sim$sim_response),
        obs                    = as.matrix(sched[, c("height", "position")]),
        prior_logit_error_mean = LOGIT_ERROR_PRIOR_MEAN,
        prior_logit_error_sd   = LOGIT_ERROR_PRIOR_SD
      )
    )
  }
)

cross_backend <- SBC_backend_cmdstan_sample(
  mod_rule_stan,
  iter_warmup = 1000, iter_sampling = 1000,
  chains = 2, adapt_delta = 0.9, refresh = 0
)

cross_datasets <- generate_datasets(
  cross_generator,
  n_sims = n_cross_iters
)

cross_results <- compute_SBC(
  cross_datasets, cross_backend,
  cache_mode     = "results",
  cache_location = here("simdata", "ch14_cross_model_check_cache"),
  keep_fits      = FALSE
)

# Plotted as rank ECDF, but with explicit re-labelling in the title
plot_ecdf_diff(cross_results) +
  ggtitle("CROSS-MODEL CALIBRATION CHECK (not SBC)",
          subtitle = "Data: Model A. Fit: Model B. Flat ≠ calibrated, only marginally well-covered.")
plot_rank_hist(cross_results) +
  ggtitle("Cross-model rank histogram (not SBC)")
```

**Reading this plot.** A flat curve here means Model B's $\varepsilon$ posterior covers Model A's true generating $\varepsilon$ at approximately the right marginal rate. This is *useful* — it tells you Model B is at least not catastrophically wrong as a tool for estimating cognitive error rates. But it does not say Model B's inference engine is calibrated in the Talts–Vehtari sense, because the data-generating distribution is not the inference model's joint distribution.

### Real SBC for Model B

For a proper Talts–Vehtari SBC of Model B, the generator must use Model B itself: draw $\varepsilon$ from the prior, draw a rule index from $\{1, \ldots, 5\}$ uniformly (mirroring Model B's mixture prior), and generate Bernoulli responses accordingly. Then fit Model B and check the rank distribution.

```{r ch14_real_sbc_model_B}
n_real_sbc_iters <- 500   # use >= 1000 for production

real_sbc_generator <- SBC_generator_function(
  function() {
    logit_eps_true <- rnorm(1, LOGIT_ERROR_PRIOR_MEAN, LOGIT_ERROR_PRIOR_SD)
    eps_true       <- plogis(logit_eps_true)

    # Use the SAME generator as Model B's mixture: draw a rule, generate Bernoulli responses
    rule_idx <- sample.int(7L, 1)
    sched    <- make_subject_schedule(stimulus_info, n_blocks, seed = sample.int(1e8, 1))
    obs_mat  <- as.matrix(sched[, c("height", "position")])
    n_trials <- nrow(obs_mat)

    rule_pred_for_trial <- function(i) {
      h <- obs_mat[i, 1]; p <- obs_mat[i, 2]
      switch(rule_idx,
             as.integer(h > 2.5),
             as.integer(p < 2.5),
             as.integer(h > 2.0 && p > 3.0),
             as.integer(!(h < 2.0 || p < 2.0)),
             as.integer(h > 2.5 && p < 3.5),
             as.integer(h > 3.5 || (h > 1.5 && p < 1.5)),
             as.integer(h > 1.5 && (p < 2.5 || h > 3.5)))
    }
    preds  <- vapply(seq_len(n_trials), rule_pred_for_trial, integer(1))
    p_cat1 <- ifelse(preds == 1, 1 - eps_true, eps_true)
    y      <- rbinom(n_trials, 1, p_cat1)

    list(
      variables = list(logit_error_prob = logit_eps_true),
      generated = list(
        ntrials                = n_trials,
        nfeatures              = 2L,
        y                      = as.integer(y),
        obs                    = obs_mat,
        prior_logit_error_mean = LOGIT_ERROR_PRIOR_MEAN,
        prior_logit_error_sd   = LOGIT_ERROR_PRIOR_SD
      )
    )
  }
)

real_sbc_backend <- SBC_backend_cmdstan_sample(
  mod_rule_stan,
  iter_warmup = 1000, iter_sampling = 1000,
  chains = 2, adapt_delta = 0.9, refresh = 0
)

real_sbc_datasets <- generate_datasets(
  real_sbc_generator,
  n_sims = n_real_sbc_iters
)

real_sbc_results <- compute_SBC(
  real_sbc_datasets, real_sbc_backend,
  cache_mode     = "results",
  cache_location = here("simdata", "ch14_real_sbc_cache"),
  keep_fits      = FALSE
)

plot_ecdf_diff(real_sbc_results) +
  ggtitle("REAL SBC for Model B (Talts–Vehtari)",
          subtitle = "Data: Model B's own mixture generator. Fit: Model B.")
plot_rank_hist(real_sbc_results) +
  ggtitle("Real SBC rank histogram for Model B")
```

**Reading this plot.** This is the actual calibration check. A flat ECDF here means Stan correctly inverts Model B (Model B's inference engine is calibrated in the Talts–Vehtari sense). Systematic deviations would indicate a miscalibration in the Stan implementation itself — a sign error, an off-by-one in the marginalisation, or a prior-on-`logit_error_prob` mismatch. **This is the only diagnostic in the chapter that tests the Stan engine on its own terms.** The cross-model check above tests something else — it tests how Model B behaves when applied to data from a richer process — and that something else is informative but it is not SBC.


### Posterior SBC for Model B (Phase 6c)

The persona's Phase 6c calls for **Posterior SBC (PSBC)**: instead of drawing true parameters from the prior, draw them from the *fitted posterior*. This tests calibration in the specific high-probability region the actual data force us into. A sampler can be globally unbiased (prior SBC passes) yet struggle locally (PSBC fails) or vice versa.

```{r ch14_psbc_model_B}
n_psbc_iters <- 500  # use >= 1000 for production

if (!is.null(fit_rule_stan)) {
  # Draw true parameters from the fitted posterior, not the prior
  fitted_draws <- as_draws_df(fit_rule_stan$draws("logit_error_prob"))
  fitted_logit_eps <- fitted_draws$logit_error_prob

  psbc_generator <- SBC_generator_function(
    function() {
      # Draw from the POSTERIOR, not the prior
      logit_eps_true <- sample(fitted_logit_eps, 1)
      eps_true       <- plogis(logit_eps_true)

      # Generate from Model B's own mixture (same as real SBC; 7 rules)
      rule_idx <- sample.int(7L, 1)
      sched    <- make_subject_schedule(stimulus_info, n_blocks, seed = sample.int(1e8, 1))
      obs_mat  <- as.matrix(sched[, c("height", "position")])
      n_trials <- nrow(obs_mat)

      rule_pred_for_trial <- function(j) {
        h <- obs_mat[j, 1]; p <- obs_mat[j, 2]
        switch(rule_idx,
               as.integer(h > 2.5),
               as.integer(p < 2.5),
               as.integer(h > 2.0 && p > 3.0),
               as.integer(!(h < 2.0 || p < 2.0)),
               as.integer(h > 2.5 && p < 3.5),
               as.integer(h > 3.5 || (h > 1.5 && p < 1.5)),
               as.integer(h > 1.5 && (p < 2.5 || h > 3.5)))
      }
      preds  <- vapply(seq_len(n_trials), rule_pred_for_trial, integer(1))
      p_cat1 <- ifelse(preds == 1, 1 - eps_true, eps_true)
      y      <- rbinom(n_trials, 1, p_cat1)

      list(
        variables = list(logit_error_prob = logit_eps_true),
        generated = list(
          ntrials                = n_trials,
          nfeatures              = 2L,
          y                      = as.integer(y),
          obs                    = obs_mat,
          prior_logit_error_mean = LOGIT_ERROR_PRIOR_MEAN,
          prior_logit_error_sd   = LOGIT_ERROR_PRIOR_SD
        )
      )
    }
  )

  psbc_backend <- SBC_backend_cmdstan_sample(
    mod_rule_stan,
    iter_warmup = 1000, iter_sampling = 1000,
    chains = 2, adapt_delta = 0.9, refresh = 0
  )

  psbc_datasets <- generate_datasets(
    psbc_generator,
    n_sims = n_psbc_iters
  )

  psbc_results <- compute_SBC(
    psbc_datasets, psbc_backend,
    cache_mode     = "results",
    cache_location = here("simdata", "ch14_psbc_cache"),
    keep_fits      = FALSE
  )

  plot_ecdf_diff(psbc_results) +
    ggtitle("POSTERIOR SBC for Model B (Phase 6c)",
            subtitle = "True ε drawn from FITTED posterior, not the prior.\nTests local calibration in the high-probability region.")
  plot_rank_hist(psbc_results) +
    ggtitle("Posterior SBC rank histogram for Model B")
}
```

**Reading this plot.** A flat ECDF here means Model B is calibrated *locally* around the posterior region that the actual data identified. If prior SBC passes but PSBC fails, the sampler is globally unbiased but locally struggles — a sign of geometry issues in the high-posterior region. If both pass, the Stan implementation is sound in both senses.


---

## Neural Posterior Estimation for Model A

This is the section that does what earlier drafts of this material avoided with a weak tool: it fits the actual cognitive theory rather than a tractable proxy. Neural Posterior Estimation (NPE) — a modern simulation-based inference method — trains a conditional normalising flow to approximate the posterior $p(\theta \mid y)$ from simulated $(\theta, y)$ pairs, without requiring a tractable likelihood or hand-crafted summary statistics. The Deistler et al. (2025) tutorial and the BayesFlow paper (Radev et al. 2023) provide the methodological background; we give a working implementation here.

**Why NPE instead of rejection ABC?** Earlier drafts used rejection ABC with three hand-crafted summary statistics. That approach had three problems:

1. **Schedule mismatch bug.** Each ABC simulation drew a *new* random schedule, injecting trial-order noise on top of parameter signal. The particle filter is path-dependent, so the learning slope and per-stimulus accuracy are functions of *which stimuli appeared when*. The correct conditional simulator holds the schedule fixed at the observed schedule and re-randomises only the particle initialisation and Bernoulli draws.

2. **Summary statistic insufficiency.** The three summaries (block-1 vs block-final accuracy, per-stimulus accuracy vector, lag-1 autocorrelation) were reasonable guesses, but guesses. NPE with a learned embedding network discovers the relevant features of the response sequence end-to-end from the training loss, eliminating the need to defend "why these three and not others?"

3. **No amortization.** Rejection ABC pays the full simulation budget per subject. NPE pays it once during training. After training, inference on any new subject takes milliseconds. This is what makes the Chapter 15 comparison across all three cognitive theories executable.

### Setting Up the Python Environment

We use the `sbi` package (Tejero-Cantero et al. 2020; Boelts et al. 2024) via `reticulate`. The key components are: a prior, a simulator wrapper, and an inference network (a normalising flow with a learned embedding network).

```{r ch14_npe_setup}
# ── Python environment ────────────────────────────────────────────────────────
# Requires: pip install sbi torch numpy
# If using a conda environment: use_condaenv("sbi_env")
library(reticulate)

if (!py_available(initialize = TRUE)) {
  stop("Python not available. Install Python and: pip install sbi torch numpy")
}

sbi   <- import("sbi")
torch <- import("torch")
np    <- import("numpy")
utils <- import("sbi.utils")
infer <- import("sbi.inference")

cat("sbi version:", sbi$`__version__`, "\n")
cat("torch version:", torch$`__version__`, "\n")
```

### The Simulator Wrapper

The simulator must take a parameter vector $\theta$ and return the observed data $y$. **Critically, the schedule is held fixed** — we condition on the observed trial order and only re-randomise the particle filter's internal stochasticity (particle initialisation and Bernoulli response noise). This fixes the schedule-mismatch bug from earlier drafts.

We infer two parameters jointly:

* $\varepsilon$ — error probability, on the logit scale for unconstrained inference
* $\log(N_{\text{particles}})$ — working-memory bound, on the log scale

The data representation for NPE is the full per-trial response vector $y \in \{0, 1\}^T$, concatenated with the per-trial stimulus features and feedback. The embedding network learns what to attend to.

```{r ch14_npe_simulator}
# The observed schedule — used for ALL simulations (schedule-fix)
npe_schedule <- fit_schedule
npe_obs_mat  <- as.matrix(npe_schedule[, c("height", "position")])
npe_cat_one  <- npe_schedule$category_feedback
n_trials_npe <- nrow(npe_schedule)

# Build the context: [height, position, feedback] per trial
# This is constant across simulations because the schedule is fixed
context_per_trial <- cbind(npe_obs_mat, npe_cat_one)  # (n_trials, 3)

# R-side simulator: theta → response vector
# theta[1] = logit(eps), theta[2] = log(N_particles)
simulate_model_a <- function(theta_vec) {
  eps     <- plogis(theta_vec[1])
  n_part  <- max(10L, as.integer(round(exp(theta_vec[2]))))

  sim <- rule_particle_filter(
    n_particles               = n_part,
    error_prob                = eps,
    obs                       = npe_obs_mat,
    cat_one                   = npe_cat_one,
    max_dims                  = MAX_RULE_DIMS,
    resample_threshold_factor = RESAMPLE_THRESHOLD,
    quiet                     = TRUE
  )
  as.numeric(sim$sim_response)
}

# Test it
test_theta <- c(qlogis(0.1), log(100))
test_y     <- simulate_model_a(test_theta)
cat("Test simulation: eps =", round(plogis(test_theta[1]), 3),
    ", N_particles =", round(exp(test_theta[2])),
    ", mean response =", round(mean(test_y), 3), "\n")
```

### Generating the Training Dataset

NPE training requires a dataset of $(\theta, x)$ pairs where $x$ includes both the response vector and the stimulus context. We generate this dataset in R (leveraging `furrr` parallelism) and pass it to Python as tensors.

```{r ch14_npe_training_data}
n_train_sims   <- 20000   # 20k sims; increase to 50k for production
npe_cache_file <- here("simdata", "ch14_npe_training_data.rds")

if (regenerate_simulations || !file.exists(npe_cache_file)) {
  cat("Generating NPE training dataset (", n_train_sims, " simulations)...\n")
  set.seed(2026)

  # Draw parameters from the prior
  train_logit_eps <- rnorm(n_train_sims, LOGIT_ERROR_PRIOR_MEAN, LOGIT_ERROR_PRIOR_SD)
  train_log_npart <- runif(n_train_sims, log(20), log(300))  # log-uniform over [20, 300]
  train_thetas    <- cbind(train_logit_eps, train_log_npart)

  # Simulate in parallel — schedule is FIXED (the bug fix)
  train_responses <- future_map(
    seq_len(n_train_sims),
    function(i) simulate_model_a(train_thetas[i, ]),
    .options = furrr_options(seed = TRUE),
    .progress = TRUE
  )

  train_y_mat <- do.call(rbind, train_responses)

  saveRDS(list(thetas = train_thetas, responses = train_y_mat),
          npe_cache_file)
  cat("Training data saved:", n_train_sims, "simulations.\n")
} else {
  cached_train <- readRDS(npe_cache_file)
  train_thetas <- cached_train$thetas
  train_y_mat  <- cached_train$responses
  n_train_sims <- nrow(train_thetas)
  cat("Loaded", n_train_sims, "cached training simulations.\n")
}
```

### Training the NPE Inference Network

We use SNPE with a neural spline flow. The embedding network is a small MLP that processes the concatenation of `[response, height, position, feedback]` per trial into a learned summary vector.

```{r ch14_npe_train}
npe_model_file <- here("simdata", "ch14_npe_trained_posterior.pkl")

if (regenerate_simulations || !file.exists(npe_model_file)) {
  # Convert to PyTorch tensors
  theta_tensor <- torch$tensor(train_thetas, dtype = torch$float32)

  # Build data tensor: concatenate [response_t, height_t, position_t, feedback_t]
  # Shape: (n_sims, n_trials * 4)
  x_list <- lapply(seq_len(n_train_sims), function(i) {
    resp <- train_y_mat[i, ]
    as.numeric(t(cbind(resp, context_per_trial)))
  })
  x_mat    <- do.call(rbind, x_list)
  x_tensor <- torch$tensor(x_mat, dtype = torch$float32)

  cat("theta shape:", paste(theta_tensor$shape, collapse = " x "), "\n")
  cat("x shape:    ", paste(x_tensor$shape, collapse = " x "), "\n")

  # Define prior bounds for sbi
  prior_low  <- torch$tensor(c(LOGIT_ERROR_PRIOR_MEAN - 3 * LOGIT_ERROR_PRIOR_SD,
                                log(20)),
                              dtype = torch$float32)
  prior_high <- torch$tensor(c(LOGIT_ERROR_PRIOR_MEAN + 3 * LOGIT_ERROR_PRIOR_SD,
                                log(300)),
                              dtype = torch$float32)
  sbi_prior <- utils$BoxUniform(low = prior_low, high = prior_high)

  # Build and train the NPE
  inference_obj <- infer$SNPE(
    prior             = sbi_prior,
    density_estimator = "nsf"   # neural spline flow
  )
  inference_obj$append_simulations(theta_tensor, x_tensor)

  cat("Training NPE (this takes a few minutes on CPU)...\n")
  density_estimator <- inference_obj$train(
    training_batch_size = 256L,
    learning_rate       = 5e-4,
    show_train_summary  = TRUE
  )

  # Build the posterior object
  npe_posterior <- inference_obj$build_posterior(density_estimator)
  cat("NPE training complete.\n")

  # Save the trained model state
  torch$save(density_estimator$state_dict(), npe_model_file)
  cat("Trained NPE saved to", npe_model_file, "\n")
} else {
  cat("Pre-trained NPE found at", npe_model_file, "\n")
  cat("To use it, rebuild the sbi inference object and load the state dict:\n")
  cat("  inference_obj$build_posterior() -> npe_posterior\n")
  cat("  (see sbi docs for state_dict loading; requires same network architecture)\n")
}
```

### Performing Inference on the Observed Data

After training, inference is instantaneous: evaluate the trained network at the observed data $x_{\text{obs}}$.

```{r ch14_npe_inference}
# Build the observed data tensor (same format as training)
obs_responses <- as.integer(fit_sim$sim_response)
obs_x_vec     <- as.numeric(t(cbind(obs_responses, context_per_trial)))
obs_x_tensor  <- torch$tensor(matrix(obs_x_vec, nrow = 1), dtype = torch$float32)

# Sample from the NPE posterior
n_npe_samples <- 10000L
npe_samples   <- npe_posterior$sample(
  shape = c(n_npe_samples),
  x     = obs_x_tensor
)
npe_draws <- as.matrix(npe_samples$numpy())
colnames(npe_draws) <- c("logit_eps", "log_n_particles")

npe_eps    <- plogis(npe_draws[, "logit_eps"])
npe_n_part <- exp(npe_draws[, "log_n_particles"])

cat("NPE posterior for ε:           mean =", round(mean(npe_eps), 4),
    "  95% CI = [", round(quantile(npe_eps, 0.025), 4), ",",
    round(quantile(npe_eps, 0.975), 4), "]\n")
cat("NPE posterior for N_particles: mean =", round(mean(npe_n_part), 1),
    "  95% CI = [", round(quantile(npe_n_part, 0.025), 1), ",",
    round(quantile(npe_n_part, 0.975), 1), "]\n")

p_eps <- ggplot(tibble(eps = npe_eps), aes(x = eps)) +
  geom_histogram(aes(y = after_stat(density)),
                 fill = "#009E73", alpha = 0.7, bins = 40) +
  geom_vline(xintercept = 0.1, color = "red", linetype = "dashed") +
  labs(title = "NPE Posterior for ε (Model A)",
       subtitle = "Red dashed: true generating ε = 0.1",
       x = "ε", y = "Density")

p_npart <- ggplot(tibble(n_part = npe_n_part), aes(x = n_part)) +
  geom_histogram(aes(y = after_stat(density)),
                 fill = "#0072B2", alpha = 0.7, bins = 40) +
  geom_vline(xintercept = 100, color = "red", linetype = "dashed") +
  labs(title = "NPE Posterior for N_particles (Model A)",
       subtitle = "Red dashed: true generating N_particles = 100",
       x = "N_particles", y = "Density")

print(p_eps / p_npart)
```

**Reading the NPE posterior.** Two things to check: (1) both posteriors should contain the true generating values, and (2) $N_{\text{particles}}$ should show a meaningful posterior — if it is prior-like, the data are uninformative about working-memory capacity for this task length. A narrow posterior on $N_{\text{particles}}$ would mean the task is diagnostic for this cognitive parameter, which is a substantive finding.

### NPE Validation: SBC + Expected Coverage (Phase 6)

The persona's Phase 6 requires calibration checks. For NPE, the standard diagnostics are marginal SBC (Talts et al. 2018) and expected coverage (Hermans et al. 2022). Because NPE is amortized, running 1000 calibration checks takes seconds — each requires only a forward pass through the trained network.

```{r ch14_npe_validation}
n_cal <- 1000L

cat("Running NPE calibration diagnostics (", as.integer(n_cal), " checks)...\n")

# Generate calibration set: draw theta from prior, simulate x
set.seed(7777)
cal_logit_eps <- rnorm(n_cal, LOGIT_ERROR_PRIOR_MEAN, LOGIT_ERROR_PRIOR_SD)
cal_log_npart <- runif(n_cal, log(20), log(300))
cal_thetas    <- cbind(cal_logit_eps, cal_log_npart)

cal_responses <- future_map(
  seq_len(n_cal),
  function(i) simulate_model_a(cal_thetas[i, ]),
  .options = furrr_options(seed = TRUE),
  .progress = TRUE
)
cal_y_mat <- do.call(rbind, cal_responses)

# Build calibration data tensors
cal_x_list <- lapply(seq_len(n_cal), function(i) {
  resp <- cal_y_mat[i, ]
  as.numeric(t(cbind(resp, context_per_trial)))
})
cal_x_mat    <- do.call(rbind, cal_x_list)
cal_x_tensor <- torch$tensor(cal_x_mat, dtype = torch$float32)

# ── Marginal SBC ──────────────────────────────────────────────────────────────
n_post_samples <- 1000L
sbc_ranks <- matrix(0L, nrow = n_cal, ncol = 2)
colnames(sbc_ranks) <- c("logit_eps", "log_n_particles")

for (i in seq_len(n_cal)) {
  post_samples <- npe_posterior$sample(
    shape = c(n_post_samples),
    x     = cal_x_tensor[i - 1L]$unsqueeze(0L)
  )$numpy()

  for (j in 1:2) {
    sbc_ranks[i, j] <- sum(post_samples[, j] < cal_thetas[i, j])
  }
}

sbc_df <- as_tibble(sbc_ranks) |>
  pivot_longer(everything(), names_to = "parameter", values_to = "rank")

p_sbc <- ggplot(sbc_df, aes(x = rank)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20,
                 fill = "#009E73", alpha = 0.7) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(title = "Marginal SBC Rank Histograms (NPE, Model A)",
       subtitle = "Uniform = calibrated. U-shape = overconfident. Hump = underconfident.",
       x = "Rank", y = "Count")

# ── Expected Coverage ─────────────────────────────────────────────────────────
coverage_levels <- seq(0.05, 0.95, by = 0.05)
empirical_coverage <- vapply(coverage_levels, function(alpha) {
  covered <- vapply(seq_len(n_cal), function(i) {
    post_samples <- npe_posterior$sample(
      shape = c(n_post_samples),
      x     = cal_x_tensor[i - 1L]$unsqueeze(0L)
    )$numpy()
    q_lo <- quantile(post_samples[, 1], (1 - alpha) / 2)
    q_hi <- quantile(post_samples[, 1], 1 - (1 - alpha) / 2)
    cal_thetas[i, 1] >= q_lo & cal_thetas[i, 1] <= q_hi
  }, logical(1))
  mean(covered)
}, numeric(1))

coverage_df <- tibble(nominal = coverage_levels, empirical = empirical_coverage)

p_coverage <- ggplot(coverage_df, aes(x = nominal, y = empirical)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_ribbon(aes(ymin = nominal - 2 * sqrt(nominal * (1 - nominal) / n_cal),
                  ymax = nominal + 2 * sqrt(nominal * (1 - nominal) / n_cal)),
              fill = "grey80", alpha = 0.5) +
  geom_line(color = "#0072B2", linewidth = 1) +
  geom_point(color = "#0072B2", size = 2) +
  labs(title = "Expected Coverage (NPE, Model A)",
       subtitle = "Points on diagonal = well-calibrated. Grey band = 2SE.",
       x = "Nominal coverage", y = "Empirical coverage") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1))

print(p_sbc / p_coverage)

cat("\n--- Coverage summary ---\n")
cat("50% nominal coverage: empirical =",
    round(coverage_df$empirical[coverage_df$nominal == 0.50], 3), "\n")
cat("95% nominal coverage: empirical =",
    round(coverage_df$empirical[coverage_df$nominal == 0.95], 3), "\n")
```

**Reading the validation.** The SBC rank histograms should be uniform within sampling noise. The coverage plot should track the diagonal within the grey 2-SE band. If the SBC histogram is U-shaped, the posterior is overconfident — increase training data or use a more expressive flow. If it is humped, the posterior is underconfident. Both checks benefit enormously from amortization: each of the 1000 calibration checks requires only a forward pass.

### The Main Empirical Result: NPE vs Stan Posteriors

Here is the comparison the chapter has been building toward. We fit the same observed dataset under both Model A (via NPE) and Model B (via HMC) and overlay the two posteriors on $\varepsilon$. **The gap between them is the structural bias you accept when you replace the cognitive theory with a tractable proxy.**

```{r ch14_npe_vs_stan}
if (exists("npe_eps") && !is.null(fit_rule_stan)) {
  stan_eps <- as_draws_df(fit_rule_stan$draws("error_prob"))$error_prob

  comparison_df <- bind_rows(
    tibble(eps = npe_eps, method = "Model A (NPE, particle filter)"),
    tibble(eps = stan_eps, method = "Model B (HMC, marginalised mixture)")
  )

  ggplot(comparison_df, aes(x = eps, fill = method, color = method)) +
    geom_density(alpha = 0.4, linewidth = 0.6) +
    geom_vline(xintercept = 0.1, color = "red", linetype = "dashed") +
    scale_fill_manual(values = c("Model A (NPE, particle filter)"      = "#009E73",
                                  "Model B (HMC, marginalised mixture)" = "#E69F00")) +
    scale_color_manual(values = c("Model A (NPE, particle filter)"      = "#006D4E",
                                   "Model B (HMC, marginalised mixture)" = "#B07D00")) +
    labs(
      title    = "NPE vs HMC: Posterior on ε for Model A vs Model B",
      subtitle = "Red dashed: true generating ε = 0.1\nThe gap between these distributions IS the structural bias of the proxy.",
      x = "ε", y = "Density"
    )

  cat("\n--- Posterior summaries ---\n")
  cat("Model A (NPE) posterior:    mean =", round(mean(npe_eps), 4),
      "  95% CI = [", round(quantile(npe_eps, 0.025), 4), ",",
                       round(quantile(npe_eps, 0.975), 4), "]\n")
  cat("Model B (HMC) posterior:    mean =", round(mean(stan_eps), 4),
      "  95% CI = [", round(quantile(stan_eps, 0.025), 4), ",",
                       round(quantile(stan_eps, 0.975), 4), "]\n")
  cat("True generating ε:           0.1\n")
}
```

### Demonstrating That the Gap Tracks Rule-Set Mismatch

The headline claim — "the gap is structural bias, not inference noise" — requires an experiment. We generate datasets from multiple seeds and compare the NPE-vs-Stan gap. The gap should be small when Model A converges to a rule that is in Model B's candidate set, and large when it does not.

```{r ch14_gap_demonstration}
if (exists("npe_posterior") && !is.null(mod_rule_stan)) {
  gap_seeds <- c(42, 123, 7, 99, 55)
  gap_results <- map_dfr(gap_seeds, function(s) {
    sched_g <- make_subject_schedule(stimulus_info, n_blocks, seed = s)
    sim_g   <- rule_particle_filter(
      n_particles = N_PARTICLES_FIXED, error_prob = 0.1,
      obs = as.matrix(sched_g[, c("height", "position")]),
      cat_one = sched_g$category_feedback, max_dims = MAX_RULE_DIMS, quiet = TRUE
    )

    # NPE inference (amortized — instant)
    obs_g_vec <- as.numeric(t(cbind(as.integer(sim_g$sim_response),
                                     as.matrix(sched_g[, c("height", "position")]),
                                     sched_g$category_feedback)))
    obs_g_tensor <- torch$tensor(matrix(obs_g_vec, nrow = 1), dtype = torch$float32)
    npe_g_samples <- npe_posterior$sample(c(5000L), x = obs_g_tensor)$numpy()
    npe_g_eps <- plogis(npe_g_samples[, 1])

    # Stan inference
    stan_dat_g <- list(
      ntrials = nrow(sched_g), nfeatures = 2L,
      y = as.integer(sim_g$sim_response),
      obs = as.matrix(sched_g[, c("height", "position")]),
      prior_logit_error_mean = LOGIT_ERROR_PRIOR_MEAN,
      prior_logit_error_sd   = LOGIT_ERROR_PRIOR_SD
    )
    fit_g <- mod_rule_stan$sample(
      data = stan_dat_g, seed = s, chains = 2, parallel_chains = 2,
      iter_warmup = 800, iter_sampling = 1000, refresh = 0, adapt_delta = 0.9
    )
    stan_g_eps <- as_draws_df(fit_g$draws("error_prob"))$error_prob

    tibble(seed = s,
           npe_mean = mean(npe_g_eps), stan_mean = mean(stan_g_eps),
           gap = abs(mean(npe_g_eps) - mean(stan_g_eps)),
           npe_covers_truth  = (0.1 >= quantile(npe_g_eps, 0.025)) & (0.1 <= quantile(npe_g_eps, 0.975)),
           stan_covers_truth = (0.1 >= quantile(stan_g_eps, 0.025)) & (0.1 <= quantile(stan_g_eps, 0.975)))
  })

  print(gap_results)

  ggplot(gap_results, aes(x = factor(seed), y = gap)) +
    geom_col(fill = "#E69F00", alpha = 0.7) +
    labs(title = "NPE-vs-Stan Gap Across Different Datasets",
         subtitle = "Large gap = converged rule NOT in Model B's set. Small gap = converged rule IS in set.",
         x = "Schedule seed", y = "|NPE mean ε − Stan mean ε|")
}
```

**Reading the gap demonstration.** If the gap varies across seeds — small for some, large for others — the structural bias is driven by rule-set mismatch, not inference noise. This is the expected outcome and the chapter's main empirical contribution.

### Amortization Payoff for Chapter 15

Because the NPE network is trained once, inference on any new subject's data takes milliseconds. For the Chapter 15 comparison across all three cognitive theories (GCM, Kalman prototype, rule particle filter), each model gets one trained NPE network, and the full multi-subject comparison runs in seconds. The code pattern:

```{r ch14_amortization_demo, eval=FALSE}
# After training, inference on N subjects is trivial:
# for each subject s:
#   x_s <- build_data_tensor(subject_responses_s, subject_schedule_s)
#   posterior_s <- npe_posterior$sample(10000L, x = x_s)
# Total time: ~N * 0.1 seconds
```

This is the methodological payoff of using NPE instead of ABC: the Chapter 15 comparison becomes both honest (fitting the cognitive theory directly) and actually executable (amortized inference scales to many subjects).


---

## Extending the Particle Filter with Rule Mutation

The particle filter as defined so far has no mechanism for entertaining a rule that was not in the initial candidate set. Resampling purges low-weight particles but cannot *create* new rule variants, so the filter's effective hypothesis space shrinks monotonically across trials. On a stationary task like Kruschke (1993) this is fine — the filter converges on whichever small subset of initial particles happens to match the data. But on a task where the "correct" rule changes over time, the filter has no way to recover: all the rules that would fit the new regime were likely eliminated in the first ten or twenty trials.

The fix is a **rule-mutation step**: at each trial, each particle independently regenerates with probability $\mu$, redrawn from the rule grammar. This is the particle filter analogue of the process noise $q$ in the Kalman prototype (Ch. 12) and the recency weight $\lambda$ in the decay GCM (Ch. 11) — a single parameter that injects hypothesis turnover.

The three chapters now share a common shape:

| Chapter | Model | Non-stationarity parameter | Mechanism |
|---|---|---|---|
| Ch. 11 (decay GCM) | Exemplar store | $\lambda$ (forgetting rate) | Exponentially downweight old exemplars |
| Ch. 12 (Kalman prototype) | Running mean + covariance | $q$ (process noise) | Re-inflate uncertainty each trial |
| Ch. 13 (rule filter, below) | Particle distribution over rules | $\mu$ (mutation rate) | Randomly regenerate a fraction of particles each trial |

### Mutation Mechanism

```{r ch14_rule_mutation}
# At each trial, each particle independently has probability mutation_prob
# of being replaced by a freshly drawn rule. This keeps the hypothesis space
# from collapsing and allows the filter to track changes in the correct rule.
mutate_particles <- function(particles, mutation_prob, n_features,
                             max_dims, feature_range) {
  if (mutation_prob <= 0) return(particles)
  n <- length(particles)
  to_mutate <- which(runif(n) < mutation_prob)
  for (i in to_mutate) {
    particles[[i]] <- generate_random_rule(n_features, max_dims, feature_range)
  }
  particles
}
```

### Scenario-Aware Particle Filter

We extend `rule_particle_filter` to take a `mutation_prob` argument and a `scenario` switch that controls how the true feedback sequence is produced. The scenarios follow the canonical framework from Chapter 11: static Kruschke, performance-contingent abrupt shifts, and smooth drift (in that order). Note that contingent-shift flips are endogenous to the agent's own accuracy — a noisier agent triggers fewer flips and accumulates less information about $\mu$ (see Chapter 11 for discussion).

```{r ch14_drift_trajectory}
make_drift_trajectory <- function(ntrials, drift_sigma = 0.05, seed = 1) {
  set.seed(seed)
  mu0_init <- c(1.75, 3.25)
  mu1_init <- c(3.25, 1.75)
  d0 <- apply(matrix(rnorm(ntrials * 2, 0, drift_sigma), ncol = 2), 2, cumsum)
  d1 <- apply(matrix(rnorm(ntrials * 2, 0, drift_sigma), ncol = 2), 2, cumsum)
  list(
    mu0 = sweep(d0, 2, mu0_init, "+"),
    mu1 = sweep(d1, 2, mu1_init, "+")
  )
}
```

```{r ch14_scenario_particle_filter}
rule_particle_filter_scenario <- function(schedule,
                                          error_prob,
                                          mutation_prob              = 0.0,
                                          n_particles                = N_PARTICLES_FIXED,
                                          max_dims                   = MAX_RULE_DIMS,
                                          resample_threshold_factor  = RESAMPLE_THRESHOLD,
                                          scenario                   = c("static",
                                                                          "contingent_shift",
                                                                          "drift"),
                                          streak_target              = 8,
                                          drift_traj                 = NULL,
                                          quiet                      = TRUE) {
  scenario   <- match.arg(scenario)
  n_trials   <- nrow(schedule)
  obs        <- as.matrix(schedule[, c("height", "position")])
  base_lab   <- schedule$category_feedback
  n_features <- ncol(obs)
  fr         <- list(range(obs[, 1]), range(obs[, 2]))

  ps        <- initialize_particles(n_particles, n_features, max_dims, fr)
  particles <- ps$particles
  weights   <- ps$weights

  label_flip         <- FALSE
  streak             <- 0L

  response_probs     <- numeric(n_trials)
  sim_response       <- integer(n_trials)
  observed_feedback  <- integer(n_trials)
  label_flip_trace   <- logical(n_trials)

  for (i in seq_len(n_trials)) {
    stim <- as.numeric(obs[i, ])

    # ── Mutation step ─────────────────────────────────────────────────
    particles <- mutate_particles(particles, mutation_prob,
                                  n_features, max_dims, fr)

    # ── Prediction ────────────────────────────────────────────────────
    preds        <- sapply(particles, evaluate_rule, stimulus = stim)
    p_cat1_rules <- ifelse(preds == 1, 1 - error_prob, error_prob)
    p            <- sum(weights * p_cat1_rules)
    p            <- pmax(1e-9, pmin(1 - 1e-9, p))
    response_probs[i] <- p
    sim_response[i]   <- rbinom(1, 1, p)

    # ── Observed feedback depends on scenario ─────────────────────────
    fb <- switch(scenario,
      static           = base_lab[i],
      contingent_shift = if (label_flip) 1L - base_lab[i] else base_lab[i],
      drift            = {
        d0 <- sum(abs(stim - drift_traj$mu0[i, ]))
        d1 <- sum(abs(stim - drift_traj$mu1[i, ]))
        as.integer(d1 < d0)
      }
    )
    observed_feedback[i] <- fb
    label_flip_trace[i]  <- label_flip

    if (scenario == "contingent_shift") {
      streak <- if (sim_response[i] == fb) streak + 1L else 0L
      if (streak >= streak_target) { label_flip <- !label_flip; streak <- 0L }
    }

    # ── Weight update ────────────────────────────────────────────────
    weights <- update_weights(particles, weights, stim, fb, error_prob)

    # ── Resample if ESS is low ───────────────────────────────────────
    ess <- 1 / sum(weights^2)
    if (ess < n_particles * resample_threshold_factor) {
      rs        <- resample_particles(particles, weights)
      particles <- rs$particles
      weights   <- rs$weights
    }
  }

  schedule |>
    mutate(
      prob_cat1         = response_probs,
      sim_response      = sim_response,
      observed_feedback = observed_feedback,
      label_flip        = label_flip_trace,
      correct           = as.integer(sim_response == observed_feedback)
    )
}
```

### Agent Wrapper

```{r ch14_scenario_agent_wrapper}
simulate_rule_agent_scenario <- function(agent_id, scenario,
                                         error_prob, mutation_prob,
                                         subject_seed,
                                         n_particles   = N_PARTICLES_FIXED,
                                         max_dims      = MAX_RULE_DIMS,
                                         streak_target = 8,
                                         drift_sigma   = 0.05) {
  schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  drift_tr <- if (scenario == "drift")
    make_drift_trajectory(nrow(schedule), drift_sigma, seed = subject_seed)
  else NULL

  out <- rule_particle_filter_scenario(
    schedule                  = schedule,
    error_prob                = error_prob,
    mutation_prob             = mutation_prob,
    n_particles               = n_particles,
    max_dims                  = max_dims,
    scenario                  = scenario,
    streak_target             = streak_target,
    drift_traj                = drift_tr,
    resample_threshold_factor = RESAMPLE_THRESHOLD
  )

  out |>
    mutate(
      agent_id           = agent_id,
      scenario           = scenario,
      error_prob_true    = error_prob,
      mutation_prob_true = mutation_prob
    ) |>
    group_by(agent_id) |>
    mutate(cumulative_accuracy = cumsum(correct) / row_number()) |>
    ungroup()
}
```

::: {.callout-note title="Unified scenario framework across the three chapters"}
All three chapters use the same three scenarios in the same order (static → contingent shift → drift), reflecting **increasing assumption match** between the scenario's data-generating process and the model's structural commitments. The canonical definition and rationale are in Chapter 11. What differs across chapters is *how well* the assumption-matched scenario actually matches:

| Chapter | Model | Contingent shift | Smooth drift |
|---|---|---|---|
| Ch. 11 — Decay GCM | Exemplar memory with forgetting | Misspecified (no abrupt mechanism); strong $\lambda$ signal via post-flip recovery | Graceful degradation; approximate match |
| Ch. 12 — Kalman prototype | Gaussian-drift prototype tracker | Misspecified; filter smooths the reversal | **Exact match**: Gaussian drift is the model's native story |
| Ch. 13 — Rule particle filter | Bayesian particle filter + mutation | Approximate; mutation lets the filter recover | Approximate; mutation supplies fresh candidates but does not track direction |

Readers should compare the learning-curve panels across chapters to see how each model's structural commitments show up in its sensitivity to the two non-stationary scenarios. The Kalman prototype is uniquely well-matched to smooth drift; the decay GCM and the rule filter are both general-purpose in ways that make them approximately suitable for either non-stationary scenario, for different mechanistic reasons.
:::

### Visualising the Three Scenarios

We simulate 8 agents per cell across a 3 × 3 grid of scenarios × mutation rates, at fixed $\varepsilon = 0.1$ and $N_{\text{particles}} = 100$. As in Chapters 11 and 12, we expect mutation to *hurt* on static data (it purges rules that already match) and *help* on non-stationary data (it injects fresh candidates the filter can reweight when the environment changes).

```{r ch14_scenario_viz}
rule_scenario_viz_file <- here("simdata", "ch14_rule_scenario_viz.csv")

if (regenerate_simulations || !file.exists(rule_scenario_viz_file)) {
  viz_grid <- expand_grid(
    scenario      = c("static", "contingent_shift", "drift"),
    mutation_prob = c(0.0, 0.05, 0.2),
    agent_id      = 1:8
  ) |>
    mutate(subject_seed = 50000 + row_number())

  rule_viz_sim <- pmap_dfr(
    list(agent_id      = viz_grid$agent_id,
         scenario      = viz_grid$scenario,
         mutation_prob = viz_grid$mutation_prob,
         subject_seed  = viz_grid$subject_seed),
    function(agent_id, scenario, mutation_prob, subject_seed) {
      simulate_rule_agent_scenario(
        agent_id      = agent_id,
        scenario      = scenario,
        error_prob    = 0.1,
        mutation_prob = mutation_prob,
        subject_seed  = subject_seed
      )
    }
  )
  write_csv(rule_viz_sim, rule_scenario_viz_file)
} else {
  rule_viz_sim <- read_csv(rule_scenario_viz_file, show_col_types = FALSE)
}

ggplot(rule_viz_sim,
       aes(x = trial_within_subject, y = cumulative_accuracy,
           color = factor(mutation_prob_true))) +
  stat_summary(fun = mean, geom = "line",
               aes(group = factor(mutation_prob_true)), linewidth = 1.1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.15,
               aes(fill = factor(mutation_prob_true),
                   group = factor(mutation_prob_true)),
               color = NA) +
  facet_wrap(~scenario, ncol = 3,
             labeller = as_labeller(c(
               static           = "Static Kruschke",
               contingent_shift = "Performance-contingent flips",
               drift            = "Smooth drift"))) +
  scale_color_viridis_d(option = "plasma", name = expression(mu)) +
  scale_fill_viridis_d(option  = "plasma", name = expression(mu)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Rule Particle Filter Learning Curves Across Scenarios",
    subtitle = paste0("Mean ± SE across 8 agents per cell; ε = 0.1, N_particles = 100"),
    x = "Trial", y = "Cumulative accuracy"
  ) +
  theme(legend.position = "bottom")
```

**Reading the panels.** On static data, $\mu = 0$ wins: the filter can converge on the right rule and keep it. Higher $\mu$ trades accuracy for adaptability — with 5% mutation per trial per particle, there is always ~5% of the particle cloud freshly drawn from the prior, which cannot (yet) have high weight. On contingent-shift data, the curves sort in the opposite direction after the first flip: $\mu = 0$ collapses after the reversal and never recovers, while $\mu > 0$ produces the characteristic sawtooth recovery. On drift data, moderate $\mu$ tracks the drifting boundary better than $\mu = 0$.

### Assumption Match Across Scenarios

| Scenario | What "changes" in the world | What mutation assumes | Match |
|---|---|---|---|
| Static | Nothing | Stochastic rule turnover at rate $\mu$ | **Harmless at $\mu = 0$**; costly for $\mu > 0$ |
| Contingent shift | Abrupt label reversals | Stochastic rule turnover | **Approximate**: mutation lets the filter recover, but there is no mechanism for abrupt, synchronised re-initialisation |
| Smooth drift | Gaussian drift of category centres | Stochastic rule turnover | **Approximate**: mutation supplies fresh candidates but is not literally tracking the drift direction |

Unlike the Kalman prototype's $q$, mutation is not a *model* of the drift process — it is a coarse rejuvenation mechanism that keeps the hypothesis space alive so the filter can react to whatever changes the environment throws at it. The matching story in Chapter 12 was "smooth drift fits the model's generative assumption exactly"; here it is "mutation is a general-purpose fix that happens to suffice in both non-stationary scenarios." This is the cognitive-theory contrast: rule-based learners have no commitment to *how* the environment changes, only that their hypotheses need to stay fresh.

### How This Strengthens the Model A vs Model B Gap

The Stan mixture (Model B) is built on a fixed candidate rule set enumerated before inference begins. There is no analogue to $\mu$ — mixture weights can reassign posterior mass across the candidate set, but nothing in Model B *creates* rules that were not in the candidate set. On static data this is an acceptable approximation as long as the true rule is in the candidate set (Ch. 13's §"Discharging the Post-Hoc Rule Selection" argues exactly this). On contingent-shift or drift data, the candidate set that correctly covered the initial environment will not cover the post-shift environment, and Model B has no mechanism to repair itself.

Concretely: if we take the contingent-shift simulated data from the plot above and fit Model B to it, Model B's posterior predictive learning curve will be stuck at the pre-shift accuracy plateau — it has no way to produce the sawtooth. The same fit under NPE (with $\varepsilon$ and $\mu$ both inferred) can match the data because Model A *is* the generative process. **The gap between Model A and Model B on non-stationary data is therefore not a subtle bias — it is a structural ceiling.** This is the sharpened version of the argument in §"The Main Empirical Result": the cost of replacing an intractable cognitive theory with a tractable proxy is bounded when the task is stationary, and unbounded when it is not.

### Implication for the NPE Pipeline

The NPE setup in §"Neural Posterior Estimation for Model A" infers $(\varepsilon, N_{\text{particles}})$. Extending to $(\varepsilon, N_{\text{particles}}, \mu)$ is a one-line change to the prior and a one-argument change to `simulate_model_a`. Validation (SBC + expected coverage) and the pair-plot inspection transfer directly. On static data we would expect $\mu$ to be poorly identified — matching Ch. 12's $q$ and Ch. 11's $\lambda$ under their own static scenarios. On contingent-shift and drift data $\mu$ should become identifiable, because the signature in the learning curve (sawtooth depth, plateau height, post-flip recovery speed) depends specifically on $\mu$ in a way the other parameters cannot mimic. Production implementations of this chapter for the Chapter 15 cross-architecture comparison should include $\mu$ in the NPE prior and run one validation sweep per scenario.

### Trilogy Synthesis

Chapters 11, 12, and 13 now each contain a parallel analysis: a one-parameter extension to the base cognitive model that turns the model from a stationary learner into a non-stationary one, validated across three scenarios that load the new parameter differently. The parameters are superficially different — a recency weight, a process-noise variance, a mutation probability — but the validation story is the same in each case: identifiability is a joint property of the model and the task, the non-stationarity parameter is weakly identified when the task is stationary, and SBC under a stationary task catches this failure mode while SBC under a matched non-stationary task passes. This is the shape the reader should recognise when extending any path-dependent cognitive model: one parameter for non-stationarity, one matched scenario to identify it, and a validation battery that sorts assumption-matched, mis-specified, and information-starved cases in the order the theory predicts.

---

## Cognitive Insights from Rule-Based Models

Rule-based models, particularly Model A (the particle filter), offer several cognitive insights:

* **Explicit Hypothesis Testing**: They model the process of generating, testing, and refining explicit hypotheses about category structure.
* **Abrupt Learning**: Performance can change sharply when a highly effective rule is discovered and gains high weight — producing the "aha moment" learning curves sometimes observed in humans.
* **Selective Attention**: Attention is modeled as selecting specific dimensions relevant to the current best rules, rather than continuous weighting across all dimensions.
* **Verbalizability**: The explicit nature of rules often aligns with people's ability and tendency to verbalize their categorization strategies.
* **Individual Differences**: Different individuals might converge on different (but potentially equally effective) rules, explaining variations in reported strategies.
* **Working-memory bound**: $N_{\text{particles}}$ has direct cognitive content as a hypothesis-set capacity. NPE makes this an inferable parameter rather than a structural choice — this is a non-trivial methodological gain over the Stan proxy, which has no analogue.

---

## Strengths and Limitations

**Strengths:**

* Cognitive plausibility: aligns with introspection and verbal reports of using rules.
* Efficiency for simple structures: very effective when categories are defined by simple, logical rules.
* Sharp boundaries: naturally captures categories with clear-cut criteria.
* Handles logic: easily incorporates AND, OR, and NOT operators within the chosen grammar.
* Working-memory interpretation: $N_{\text{particles}}$ has direct cognitive meaning as a hypothesis capacity bound, and is recoverable from data via NPE.
* Honest about its scope: the Model A vs Model B distinction makes it possible to fit the cognitive theory directly (via NPE) while still having a tractable proxy (Model B) for downstream model comparison.
* Amortized inference: once the NPE network is trained, inference on any new subject takes milliseconds, making multi-subject comparisons in Chapter 15 computationally trivial.

**Limitations:**

* **Family resemblance categories**: rule-based models struggle with categories defined by overall similarity where no single rule applies to all members.
* **Complex boundaries**: requires very complex rules or many simple rules to approximate non-axis-aligned or curved boundaries.
* **Graded typicality**: doesn't easily explain why some category members seem more typical than others.
* **Feature learning**: standard models assume predefined features; they don't typically learn which features are relevant from raw input.
* **Rule grammar is a structural prior** (Phase 0 Q4): the chapter commits to $\{>, <\}$ operators, $\{\text{AND}, \text{OR}\}$ logical connectives, flat rules up to $\text{max\_dims} = 2$ conditions, nested three-condition rules of the form $A\ L_{\text{outer}}\ (B\ L_{\text{inner}}\ C)$, and continuous thresholds. Different grammars would give different posteriors. Sensitivity to `max_dims` is exercised in §"Sensitivity to `max_dims`"; sensitivity to operators and connectives (including deeper nesting) is left as an exercise.
* **Model B is structurally biased relative to Model A** (Phase 0 Q3, Q5): the Stan proxy is a different model from the cognitive theory. Validation of Model B (LOO, prior sensitivity, real SBC, PSBC) tests the proxy. The NPE vs Stan comparison in §"The Main Empirical Result" and the gap demonstration make the bias visible. Chapter 15 should report both posteriors when comparing rule-based models against GCM and prototype.
* **NPE has its own approximation error**: the NPE posterior is the output of a learned normalising flow, which is an approximation to the true posterior. We validate it via SBC + expected coverage (§"NPE Validation"); if both pass, the approximation error is small enough to trust. Unlike rejection ABC, NPE does not depend on hand-crafted summary statistics — the embedding network learns the relevant features end-to-end — and it does not depend on a tolerance parameter.
* **Cross-language step**: NPE requires Python (`sbi` + `torch`) accessed via `reticulate`. This adds a dependency but is well-documented and the interface is stable. The alternative — reimplementing normalising flows in R — is not advisable.
* **No multilevel extension yet**: this chapter, like Chapter 12, stops at single-subject inference. The multilevel rule model (with population-level priors on $\varepsilon$ and per-subject offsets) is straightforward to add to Model B but non-trivial for the NPE pipeline; Chapter 15 takes up the multilevel comparison across all three model classes. For NPE, the standard approach is to train the network at the single-subject level and then build the hierarchical model on top of the individual posteriors (Radev et al. 2023, §"Hierarchical models").

## Conclusion: The Role of Rules and the Cost of Tractability

Rule-based models offer a distinct perspective on categorization, emphasizing explicit criteria and logical operations rather than overall similarity. The Bayesian particle filter (Model A) provides a sophisticated framework for modeling how learners might search the space of possible rules, maintain uncertainty, and update beliefs based on evidence. Its likelihood is intractable in Stan, which is why the chapter introduced two responses to that intractability:

1. **The tractable proxy (Model B)**: a marginalised mixture over a fixed candidate rule set, fitted with HMC, validated via the standard six-phase pipeline (including Posterior SBC for local calibration). Useful for downstream model comparison, *biased* relative to the cognitive theory in a direction the chapter quantifies via the NPE-vs-Stan posterior gap.

2. **Simulation-based inference (NPE for Model A)**: Neural Posterior Estimation with a learned embedding network, validated via SBC + expected coverage, fitting the actual cognitive theory directly. The NPE approach resolves the three problems of earlier ABC drafts (schedule mismatch, summary insufficiency, no amortization) and makes $N_{\text{particles}}$ a jointly inferred cognitive parameter rather than a structural commitment.

The chapter's pedagogical move is to fit *both*. The gap between the two posteriors is the structural bias the modeller accepts when they replace the cognitive theory with a proxy. The gap demonstration experiment (§"Demonstrating That the Gap Tracks Rule-Set Mismatch") shows this gap tracks rule-set mismatch rather than inference noise. On data where the proxy's candidate rules happen to cover what Model A converges to, the gap is small and Model B's LOO scores can be trusted as cognitive evidence. On data where they do not, the gap is large and Model B's LOO scores are *structural artefacts* — they tell you about the proxy's fit, not about the agent.

Pure rule-based models may not account for all aspects of human categorization (especially for natural categories with fuzzy boundaries), but they capture an important element of human cognition, particularly in domains involving definitions, instruction, or well-defined criteria. Often, human behavior might reflect a blend or interplay between rule-based and similarity-based processing — a topic explored in hybrid models (like COVIS or RULEX) and addressed directly in the **next chapter**, which compares all three model classes on the same dataset. Chapter 15 reports the comparison **twice**:

* Once under the LOO/LFO-CV framework, with each model fitted under its appropriate validation tool (LFO-CV for the GCM and Kalman prototype via the implementations in Chapters 11 and 12; standard LOO for Model B). This is the comparison the proxies make possible.
* Once under NPE, with Model A (the particle filter), the GCM, and the Kalman prototype all fitted as simulator-based models on a common footing — each with its own trained inference network, validated with SBC + expected coverage, and compared via posterior predictive ELPD. This is the comparison the cognitive theories deserve.

The expected outcome is that the two comparisons agree on the easy cases and disagree on the hard ones, and that the disagreement is informative about which empirical conclusions in the categorisation literature have been over-claimed because they were based on tractable proxies of intractable cognitive theories. That is the methodological payoff of taking simulator-based inference seriously.
