---
title: "16 — Human Performance in the Alien Categorization Task"
output:
  html_document:
    mathjax: default
    toc: true
    toc_float: true
  pdf_document:
    toc: true
date: "2025-04-28"
---

```{r 16_setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)

regenerate_fits <- FALSE   # set TRUE to refit all models to human data

pacman::p_load(
  tidyverse,   # data manipulation and visualization
  mvtnorm,     # multivariate normal density (Kalman boundary)
  patchwork,   # combining plots
  cmdstanr,    # Stan interface
  posterior,   # tidy posterior arrays
  tidybayes,   # tidy draw extraction
  bayesplot,   # MCMC diagnostics and LOO-PIT
  loo,         # LOO-CV, model comparison, stacking
  priorsense,  # prior sensitivity
  future,      # parallel processing
  furrr        # parallel mapping
)

theme_set(theme_classic())

# ── Prior hyperparameters — must match Chapters 11–13 exactly ────────────────
LOG_C_PRIOR_MEAN       <- log(2)
LOG_C_PRIOR_SD         <- 1.0
LOG_R_PRIOR_MEAN       <- 0
LOG_R_PRIOR_SD         <- 1.0
LOGIT_ERROR_PRIOR_MEAN <- -2.0
LOGIT_ERROR_PRIOR_SD   <- 1.0

for (d in c("stan", "simdata", "simmodels")) {
  if (!dir.exists(d)) dir.create(d)
}

# ── Data file path ────────────────────────────────────────────────────────────
# Update if AlienData.txt lives elsewhere
alien_data_file <- "AlienData.txt"

plan(multisession, workers = max(1, parallel::detectCores() - 1))
```

> **📍 Where we are in the Bayesian modeling workflow:**
> Chapter 15 compared the three Stan models on a common *simulated* dataset
> and established the full LOO pipeline. This chapter applies the same pipeline
> to **real human data** from Tylén et al. (2023). The structure has four parts:
> (1) load, clean, and verify the data; (2) descriptive analysis of human
> learning curves and error patterns; (3) fit all three Stan models to individual
> participant data using the multilevel GCM from Chapter 11 and single-subject
> Kalman and rule models from Chapters 12–13; (4) formal model comparison with
> LOO, stacking weights, LOO-PIT, and individual-differences analysis.
> The LFO-CV caveat introduced in Chapter 11 is handled explicitly: we flag
> all GCM and Kalman ELPD estimates as approximate and show the LFO-CV
> approximation for those models.

# Introduction

This chapter analyses human performance data from the Alien Game categorization task (Tylén et al. 2023). The goal is to: (a) characterize how people learn across sessions of increasing rule complexity; (b) fit the three computational models from Chapters 11–13 directly to individual participant data; and (c) use formal model comparison to ask which architecture best accounts for the observed patterns.

The task required participants — either individually or in pairs — to categorize visually complex alien stimuli based on feedback. Aliens varied along five binary features, and participants had to make two independent binary judgements per trial: **Dangerous** (0 = Peaceful, 1 = Dangerous) and **Nutritious** (0 = Non-Nutritious, 1 = Nutritious). The experiment comprised three sessions with increasing rule complexity.

---

# 1. Data Loading and Cleaning

```{r 16_load}
human_data <- read_csv(alien_data_file, show_col_types = FALSE)

cat("Columns:\n"); print(colnames(human_data))
cat("\nDimensions:", nrow(human_data), "rows ×", ncol(human_data), "cols\n")
cat("\nUnique values in key columns:\n")
key_cols <- c("condition", "subject", "session", "cycle",
              "test", "category", "response", "dangerous", "nutricious")
print(sapply(human_data[, key_cols], function(x) sort(unique(x))))
```

## Data Cleaning

```{r 16_clean}
human_data <- human_data |>
  # Fix column name spelling
  rename(nutritious = nutricious) |>

  # Extract the 5-digit binary feature string from the stimulus filename
  mutate(
    stimulus_str  = str_extract(stimulus, "^[01]{5}"),
    is_valid_stim = !is.na(stimulus_str)
  ) |>
  filter(is_valid_stim) |>

  # Split into individual feature columns f1–f5
  separate(stimulus_str, into = paste0("f", 1:5),
           sep = 1:4, convert = TRUE, remove = FALSE) |>

  # Map 4-choice response to independent dimension responses.
  # Assumed mapping (standard for 2×2 tasks):
  #   1 → (Dangerous=0, Nutritious=0)
  #   2 → (Dangerous=0, Nutritious=1)
  #   3 → (Dangerous=1, Nutritious=0)
  #   4 → (Dangerous=1, Nutritious=1)
  mutate(
    inferred_dangerous  = as.integer(response %in% c(3, 4)),
    inferred_nutritious = as.integer(response %in% c(2, 4))
  ) |>

  # Dimension-level correctness
  mutate(
    correct_dangerous  = as.integer(inferred_dangerous  == dangerous),
    correct_nutritious = as.integer(inferred_nutritious == nutritious),
    recalc_correct     = as.integer(correct_dangerous == 1 & correct_nutritious == 1)
  )

# Verify our response mapping against the data's own 'correct' column
mismatch <- human_data |>
  filter(!is.na(correct)) |>
  summarise(n_mismatch = sum(correct != recalc_correct, na.rm = TRUE), n = n())

if (mismatch$n_mismatch > 0) {
  warning(sprintf(
    "Response mapping mismatch: %d / %d rows. Check response coding assumption.",
    mismatch$n_mismatch, mismatch$n
  ))
} else {
  cat("Response mapping verified: 'correct' column matches inferred dimensional accuracy.\n")
}

# Add a within-subject, within-session trial counter.
# This is the temporal index that path-dependent models require.
human_data <- human_data |>
  group_by(subject, session) |>
  mutate(trial_within_subject = row_number()) |>
  ungroup()

cat("\nCleaned data — first rows:\n")
print(head(human_data |>
  dplyr::select(subject, condition, session, trial_within_subject,
                stimulus_str, f1:f5, dangerous, nutritious,
                response, inferred_dangerous, inferred_nutritious,
                correct, correct_dangerous, correct_nutritious)))
```

> **Column notes:**
> `nutricious` in the raw file is a typo; renamed to `nutritious`.
> `test == 0` = training trials with feedback (used throughout).
> `test == 1` = test trials without feedback (excluded).
> `condition == 1` = individual participants; `condition == 2` = pairs.
> `trial_within_subject` counts trials within each (subject × session) block
> and is the critical temporal index for path-dependent models.

```{r 16_training}
training_data <- human_data |>
  filter(test == 0, is_valid_stim)

cat("Training trials:", nrow(training_data), "\n")
cat("Subjects:", n_distinct(training_data$subject), "\n")
cat("Sessions:", sort(unique(training_data$session)), "\n")
```

## Verifying the Experimental Design

We check that the data matches the design described in Chapter 14: 32 unique stimuli, three blocks (96 trials) per session, and category labels consistent with the known rules.

```{r 16_design_checks}
# --- Trials per subject per session ---
trials_ss <- training_data |>
  count(condition, subject, session, name = "n_trials")

cat("Trials per subject per session (should be 96):\n")
print(summary(trials_ss$n_trials))

# --- Stimulus coverage ---
stim_coverage <- training_data |>
  group_by(session) |>
  summarise(
    n_unique = n_distinct(stimulus_str),
    min_freq = min(table(stimulus_str)),
    max_freq = max(table(stimulus_str)),
    .groups = "drop"
  )
cat("\nStimulus coverage per session (expect 32 unique):\n")
print(stim_coverage)

# --- Category rule verification ---
# Session 3 dangerous: f4 + f5 >= 2 ↔ f4 == 1 AND f5 == 1 (binary features)
rule_check <- training_data |>
  mutate(
    rule_dang = case_when(
      session == 1 ~ as.integer(f1 == 1 & f3 == 1),
      session == 2 ~ as.integer(f4 == 1),
      session == 3 ~ as.integer(f4 == 1 & f5 == 1)
    ),
    rule_nutr = case_when(
      session == 1 ~ as.integer(f4 == 1),
      session == 2 ~ as.integer((f1 + f2 + f3) >= 2),
      session == 3 ~ as.integer(
        (f2 + f3 == 2) | (f1 == 1 & ((f4 + f5) >= 1) & ((f2 + f3) >= 1))
      )
    )
  ) |>
  group_by(session) |>
  summarise(
    match_dangerous  = mean(dangerous  == rule_dang, na.rm = TRUE),
    match_nutritious = mean(nutritious == rule_nutr, na.rm = TRUE),
    .groups = "drop"
  )

cat("\nProportion of trials matching assumed category rules (expect 1.0):\n")
print(rule_check)
```

If `match_dangerous` or `match_nutritious` is not 1.0, the assumed rules for that session are incorrect and must be corrected before fitting any model. All subsequent model fitting is conditional on these rules being verified.

---

# 2. Human Performance Analysis

## Overall Learning Curves

```{r 16_overall_learning}
learning_overall <- training_data |>
  group_by(session, trial_within_subject) |>
  summarise(
    mean_acc = mean(correct, na.rm = TRUE),
    se       = sd(correct, na.rm = TRUE) / sqrt(n()),
    .groups  = "drop"
  )

ggplot(learning_overall,
       aes(x = trial_within_subject, y = mean_acc, color = factor(session))) +
  geom_ribbon(aes(ymin = mean_acc - se, ymax = mean_acc + se,
                  fill = factor(session)), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, span = 0.4,
              alpha = 0.5, linewidth = 0.7, linetype = "dashed") +
  facet_wrap(~session, labeller = labeller(session = function(s) paste("Session", s))) +
  scale_color_viridis_d(option = "plasma", name = "Session") +
  scale_fill_viridis_d(option = "plasma", name = "Session") +
  labs(
    title    = "Overall Learning Curves (Training Trials)",
    subtitle = "Mean accuracy ± SE across all participants; dashed = LOESS smoother",
    x = "Trial within session", y = "Proportion Correct"
  ) +
  ylim(0, 1)
```

## Learning Curves by Dimension

```{r 16_dimension_learning}
learning_dim <- training_data |>
  group_by(session, trial_within_subject) |>
  summarise(
    dangerous  = mean(correct_dangerous,  na.rm = TRUE),
    nutritious = mean(correct_nutritious, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(c(dangerous, nutritious),
               names_to = "dimension", values_to = "accuracy")

ggplot(learning_dim,
       aes(x = trial_within_subject, y = accuracy, color = dimension)) +
  geom_line(linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, span = 0.4,
              linewidth = 0.7, linetype = "dashed", alpha = 0.6) +
  facet_wrap(~session, labeller = labeller(session = function(s) paste("Session", s))) +
  scale_color_manual(values = c(dangerous = "#D55E00", nutritious = "#0072B2")) +
  labs(
    title    = "Learning Curves by Decision Dimension (Training Trials)",
    subtitle = "Accuracy averaged across all participants",
    x = "Trial within session", y = "Proportion Correct", color = "Dimension"
  ) +
  ylim(0, 1)
```

## Individual vs. Pair Condition

```{r 16_condition_learning}
learning_cond <- training_data |>
  mutate(
    condition_label = factor(condition, levels = c(1, 2),
                             labels = c("Individual", "Pair"))
  ) |>
  group_by(condition_label, session, trial_within_subject) |>
  summarise(
    mean_acc = mean(correct, na.rm = TRUE),
    se       = sd(correct, na.rm = TRUE) / sqrt(n()),
    .groups  = "drop"
  )

ggplot(learning_cond,
       aes(x = trial_within_subject, y = mean_acc,
           color = condition_label, fill = condition_label)) +
  geom_ribbon(aes(ymin = mean_acc - se, ymax = mean_acc + se),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, span = 0.4,
              linewidth = 0.7, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~session, labeller = labeller(session = function(s) paste("Session", s))) +
  scale_color_manual(values = c(Individual = "#0072B2", Pair = "#D55E00")) +
  scale_fill_manual(values  = c(Individual = "#0072B2", Pair = "#D55E00")) +
  labs(
    title    = "Learning Curves by Condition (Training Trials)",
    subtitle = "Individual (condition 1) vs. Pair (condition 2)",
    x = "Trial within session", y = "Proportion Correct", color = "Condition"
  ) +
  ylim(0, 1)
```

**Qualitative interpretation**: Compare these curves against the predictions from Chapter 14:

* **Session 1** (low complexity: f1 AND f3 for dangerous; f4 for nutritious) — do both dimensions approach ceiling by trial 96? The nutritious rule is a single feature and should be learnable by all models; the dangerous rule is a conjunction that benefits models with selective attention or rule representation.
* **Session 2** (intermediate: f4 for dangerous; at-least-2-of-3 for nutritious) — does nutritious learning lag dangerous? The "majority vote" rule is structurally more difficult for the Kalman filter (which expects a unimodal distribution) and for a 1D rule model.
* **Session 3** (high complexity) — performance may plateau well below ceiling, especially for the disjunctive nutritious rule.

---

# 3. Critical Error Analysis

```{r 16_error_types}
error_trials <- training_data |>
  filter(correct == 0) |>
  mutate(
    error_type = case_when(
      correct_dangerous == 0 & correct_nutritious == 1 ~ "Dangerous Only",
      correct_dangerous == 1 & correct_nutritious == 0 ~ "Nutritious Only",
      correct_dangerous == 0 & correct_nutritious == 0 ~ "Both Dimensions",
      TRUE                                             ~ "Other"
    ),
    condition_label = factor(condition, levels = c(1, 2),
                             labels = c("Individual", "Pair"))
  )

error_summary <- error_trials |>
  count(session, condition_label, error_type) |>
  group_by(session, condition_label) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

ggplot(error_summary,
       aes(x = factor(session), y = prop, fill = error_type)) +
  geom_col(position = "stack", width = 0.7) +
  facet_wrap(~condition_label) +
  scale_fill_brewer(palette = "Spectral", name = "Error type") +
  labs(
    title    = "Distribution of Error Types within Incorrect Trials",
    subtitle = "Proportion of errors on each dimension",
    x = "Session", y = "Proportion of errors"
  ) +
  theme(legend.position = "bottom")
```

```{r 16_error_stimuli}
errors_per_stim <- error_trials |>
  count(session, condition_label, stimulus_str, name = "n_errors") |>
  left_join(
    training_data |>
      distinct(stimulus_str, session, dangerous, nutritious),
    by = c("stimulus_str", "session")
  ) |>
  arrange(session, condition_label, desc(n_errors))

cat("Top 5 most error-prone stimuli per session/condition:\n")
errors_per_stim |>
  group_by(session, condition_label) |>
  slice_max(order_by = n_errors, n = 5) |>
  print(n = Inf)

# Visual: top-10 error-prone stimuli per condition × session
top_err <- errors_per_stim |>
  group_by(session, condition_label) |>
  slice_max(order_by = n_errors, n = 10) |>
  ungroup() |>
  mutate(cat_label = paste0("D", dangerous, ":N", nutritious))

ggplot(top_err,
       aes(x = reorder(stimulus_str, -n_errors), y = n_errors,
           fill = cat_label)) +
  geom_col() +
  facet_grid(condition_label ~ session, scales = "free_x") +
  scale_fill_brewer(palette = "Set2", name = "True cat (D:N)") +
  labs(
    title = "Top 10 Most Error-Prone Stimuli",
    x = "Stimulus (binary string)", y = "Total errors across subjects"
  ) +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1),
        legend.position = "bottom")
```

**Error pattern interpretation**: The most error-prone stimuli are those near rule boundaries (ambiguous under the correct rule), or those that satisfy some but not all conditions of a conjunctive or disjunctive rule. A model that predicts *exactly* these stimuli to be hardest provides the strongest cognitive account. We return to this when we examine individual-level posterior predictives below.

---

# 4. Fitting Models to Human Data

We now fit all three Stan models to the real human data. This is the fulfilment of the pipeline established in Chapters 11–15: prior predictive check → SBC → parameter recovery → single-agent fit → multilevel fit → LOO comparison, all now applied to empirical data.

## Design Choices

**Two-decision structure**: The Alien Game requires two independent binary decisions per trial (dangerous and nutritious). The three Stan models from Chapters 11–13 are single-decision models (one binary outcome per trial). We handle this by fitting each model **separately for each decision dimension** within each session, then combining the model comparison results.

**Sequential structure within sessions**: Each session has its own stimulus sequence with its own trial ordering per participant. The Kalman filter resets its prototype state at the start of each session (the categories are redefined). The GCM's exemplar memory should also be reset across sessions, since the feedback structure changes.

**Multilevel vs. single-subject for the GCM**: Chapter 11 developed both a single-subject GCM and a multilevel GCM. For human data with multiple participants we use the **multilevel GCM** (Stan model `W11_gcm_ml.stan`), which estimates population-level distributions over $c$, $w$, and $\beta$ while allowing individual variation. The Kalman and rule models are fitted as single-subject models per participant (iterating over subjects), producing individual-level posteriors that can be summarised at the population level.

```{r 16_load_models}
load_or_stop <- function(path) {
  if (!file.exists(path))
    stop("Stan model not found: ", path,
         "\nRun the corresponding chapter first.")
  cmdstan_model(path)
}

mod_gcm_ml <- load_or_stop("stan/W11_gcm_ml.stan")
mod_kalman <- load_or_stop("stan/W12_prototype_single.stan")
mod_rule   <- load_or_stop("stan/W13_rule_simplified.stan")
cat("All three Stan models loaded.\n")
```

## Preparing Human Data for Stan

### Multilevel GCM

The multilevel GCM needs the data sorted by `(subj_id_stan, trial_within_subject)` and the `subj_start` / `subj_end` slice indices (see Chapter 11 for the full explanation of why this sort is mandatory for the sequential memory update).

```{r 16_prep_gcm_data}
# We fit one multilevel GCM per (session × decision_dimension) combination.
# Feature columns for the GCM: only f1–f5 are used as 5D binary features.
# The Kruschke-task model used 2D continuous features; the Alien task uses
# 5D binary features. The Stan model is general: N_features is passed as data.

build_gcm_ml_data <- function(df, decision_col) {
  # df: training_data filtered to one session
  # decision_col: "dangerous" or "nutritious" (the true label column)
  # y is inferred_dangerous or inferred_nutritious
  resp_col <- paste0("inferred_", decision_col)

  df_sorted <- df |>
    mutate(subj_id_stan = as.integer(factor(subject))) |>
    arrange(subj_id_stan, trial_within_subject)

  bounds <- df_sorted |>
    mutate(row_idx = row_number()) |>
    group_by(subj_id_stan) |>
    summarise(
      subj_start = min(row_idx),
      subj_end   = max(row_idx),
      .groups    = "drop"
    ) |>
    arrange(subj_id_stan)

  list(
    N_total                  = nrow(df_sorted),
    N_subjects               = max(df_sorted$subj_id_stan),
    N_features               = 5L,
    max_trials_per_subject   = max(bounds$subj_end - bounds$subj_start + 1),
    subj_start               = bounds$subj_start,
    subj_end                 = bounds$subj_end,
    y                        = df_sorted[[resp_col]],
    obs                      = as.matrix(df_sorted[, paste0("f", 1:5)]),
    cat_feedback             = df_sorted[[decision_col]],
    pop_w_prior_alpha        = rep(1, 5),
    pop_log_c_mean_prior_mean  = LOG_C_PRIOR_MEAN,
    pop_log_c_mean_prior_sd    = 1.0,
    pop_log_c_sd_prior_rate    = 1.0,
    kappa_prior_rate           = 0.1,
    pop_logit_bias_sd_prior_rate = 2.0
  )
}
```

### Kalman and Rule Models (per-subject)

```{r 16_prep_single_data}
build_kalman_data <- function(df_subj, decision_col) {
  # df_subj: one subject's data for one session, sorted by trial_within_subject
  resp_col <- paste0("inferred_", decision_col)
  list(
    ntrials            = nrow(df_subj),
    nfeatures          = 5L,
    cat_one            = df_subj[[decision_col]],
    y                  = df_subj[[resp_col]],
    obs                = as.matrix(df_subj[, paste0("f", 1:5)]),
    initial_mu_cat0    = rep(0.5, 5),   # prior centre for binary features
    initial_mu_cat1    = rep(0.5, 5),
    initial_sigma_diag = 1.0,           # smaller than Ch12: binary features ∈ {0,1}
    prior_logr_mean    = LOG_R_PRIOR_MEAN,
    prior_logr_sd      = LOG_R_PRIOR_SD
  )
}

build_rule_data <- function(df_subj, decision_col) {
  resp_col <- paste0("inferred_", decision_col)
  list(
    ntrials                = nrow(df_subj),
    nfeatures              = 5L,
    y                      = df_subj[[resp_col]],
    obs                    = as.matrix(df_subj[, paste0("f", 1:5)]),
    prior_logit_error_mean = LOGIT_ERROR_PRIOR_MEAN,
    prior_logit_error_sd   = LOGIT_ERROR_PRIOR_SD
  )
}
```

> **Note on the Stan rule model with 5 features**: The Stan rule model in
> Chapter 13 hardcodes a 5-rule candidate set over a 2D feature space. For
> the 5D Alien task, the candidate rule set must be expanded to include the
> session-specific relevant rules. The session-appropriate candidate rules
> (as verified in the data exploration above) should be hardcoded into the
> Stan model's `model` and `generated quantities` blocks. A general approach
> would enumerate all single-feature and two-feature threshold rules over the
> five binary dimensions, giving $5 \times 2 + \binom{5}{2} \times 2 \times
> 2 = 10 + 40 = 50$ candidate rules. For binary features all thresholds are
> 0.5, so each rule reduces to testing whether one or two features equal 0
> or 1. We leave this extension as an exercise and fit the rule model with the
> session-specific rules derived from the rule-check above.

## Fitting

```{r 16_fit_models}
run_fit_single <- function(mod, data, seed, adapt_delta = 0.9,
                           iter_w = 800, iter_s = 1000) {
  pf <- tryCatch(
    mod$pathfinder(data = data, seed = seed, num_paths = 4, refresh = 0),
    error = function(e) NULL
  )
  mod$sample(
    data            = data,
    init            = if (!is.null(pf)) pf else 0.5,
    seed            = seed,
    chains          = 2,
    parallel_chains = 2,
    iter_warmup     = iter_w,
    iter_sampling   = iter_s,
    refresh         = 0,
    adapt_delta     = adapt_delta
  )
}

run_fit_ml <- function(mod, data, seed, adapt_delta = 0.95,
                       iter_w = 1500, iter_s = 1500) {
  pf <- tryCatch(
    mod$pathfinder(data = data, seed = seed, num_paths = 4, refresh = 0),
    error = function(e) NULL
  )
  mod$sample(
    data            = data,
    init            = if (!is.null(pf)) pf else 0.5,
    seed            = seed,
    chains          = 4,
    parallel_chains = min(4, max(1, parallel::detectCores() - 1)),
    iter_warmup     = iter_w,
    iter_sampling   = iter_s,
    refresh         = 200,
    adapt_delta     = adapt_delta,
    max_treedepth   = 12
  )
}

# ── Build fit list: one fit per (session × decision_dimension × model) ────────
sessions    <- 1:3
decisions   <- c("dangerous", "nutritious")
subjects    <- sort(unique(training_data$subject))

fits_path <- "simmodels/W16_human_fits.rds"

if (regenerate_fits || !file.exists(fits_path)) {
  cat("Fitting models to human data...\n")
  all_fits <- list()

  for (sess in sessions) {
    sess_data <- training_data |>
      filter(session == sess) |>
      arrange(subject, trial_within_subject)

    for (dec in decisions) {
      key <- paste0("s", sess, "_", dec)
      cat("  Session", sess, "—", dec, "\n")

      # ── Multilevel GCM ────────────────────────────────────────────────────
      gcm_dat <- build_gcm_ml_data(sess_data, dec)
      all_fits[[paste0(key, "_gcm")]] <- tryCatch(
        run_fit_ml(mod_gcm_ml, gcm_dat,
                   seed = sess * 100 + as.integer(dec == "nutritious")),
        error = function(e) { cat("    GCM fit error:", e$message, "\n"); NULL }
      )

      # ── Kalman (per-subject, then combine log_lik) ────────────────────────
      kalman_ll_list <- vector("list", length(subjects))
      for (ji in seq_along(subjects)) {
        subj  <- subjects[ji]
        s_dat <- sess_data |>
          filter(subject == subj) |>
          arrange(trial_within_subject)
        if (nrow(s_dat) == 0) next
        k_dat <- build_kalman_data(s_dat, dec)
        fit_k <- tryCatch(
          run_fit_single(mod_kalman, k_dat,
                         seed = sess * 1000 + ji + as.integer(dec == "nutritious")),
          error = function(e) NULL
        )
        if (!is.null(fit_k))
          kalman_ll_list[[ji]] <- fit_k$draws("log_lik", format = "matrix")
      }
      all_fits[[paste0(key, "_kalman_ll")]] <- kalman_ll_list

      # ── Rule model (per-subject) ──────────────────────────────────────────
      rule_ll_list <- vector("list", length(subjects))
      for (ji in seq_along(subjects)) {
        subj  <- subjects[ji]
        s_dat <- sess_data |>
          filter(subject == subj) |>
          arrange(trial_within_subject)
        if (nrow(s_dat) == 0) next
        r_dat <- build_rule_data(s_dat, dec)
        fit_r <- tryCatch(
          run_fit_single(mod_rule, r_dat,
                         seed = sess * 2000 + ji + as.integer(dec == "nutritious")),
          error = function(e) NULL
        )
        if (!is.null(fit_r))
          rule_ll_list[[ji]] <- fit_r$draws("log_lik", format = "matrix")
      }
      all_fits[[paste0(key, "_rule_ll")]] <- rule_ll_list
    }
  }

  saveRDS(all_fits, fits_path)
  cat("All fits saved to", fits_path, "\n")
} else {
  all_fits <- readRDS(fits_path)
  cat("Loaded fits from", fits_path, "\n")
}
```

---

# 5. Model Comparison on Human Data

## LOO per (Session × Decision × Model)

We compute LOO for each model in each condition. For the multilevel GCM we use the full `log_lik` vector; for the per-subject Kalman and rule models we concatenate the individual `log_lik` matrices.

```{r 16_loo}
# Helper: concatenate per-subject log_lik matrices (subjects as additional "observations")
concat_subj_ll <- function(ll_list) {
  valid <- Filter(Negate(is.null), ll_list)
  if (length(valid) == 0) return(NULL)
  # Each element is (n_draws × n_trials_subj); combine columns
  do.call(cbind, valid)
}

loo_results <- list()
for (sess in sessions) {
  for (dec in decisions) {
    key <- paste0("s", sess, "_", dec)

    # GCM
    gcm_fit <- all_fits[[paste0(key, "_gcm")]]
    if (!is.null(gcm_fit)) {
      ll_gcm <- gcm_fit$draws("log_lik", format = "matrix")
      loo_results[[paste0(key, "_gcm")]] <- loo::loo(ll_gcm, cores = 4)
    }

    # Kalman
    ll_kal <- concat_subj_ll(all_fits[[paste0(key, "_kalman_ll")]])
    if (!is.null(ll_kal))
      loo_results[[paste0(key, "_kalman")]] <- loo::loo(ll_kal, cores = 4)

    # Rule
    ll_rule <- concat_subj_ll(all_fits[[paste0(key, "_rule_ll")]])
    if (!is.null(ll_rule))
      loo_results[[paste0(key, "_rule")]] <- loo::loo(ll_rule, cores = 4)
  }
}

# Summarize ELPD across all conditions
loo_summary <- imap_dfr(loo_results, function(l, name) {
  tibble(
    condition      = name,
    session        = as.integer(str_extract(name, "(?<=s)\\d")),
    decision       = str_extract(name, "dangerous|nutritious"),
    model          = str_extract(name, "gcm|kalman|rule"),
    elpd           = l$estimates["elpd_loo", "Estimate"],
    se_elpd        = l$estimates["elpd_loo", "SE"],
    path_dependent = model %in% c("gcm", "kalman")
  )
})

print(loo_summary |> arrange(session, decision, model))
```

## ELPD Comparison Plot

```{r 16_elpd_plot}
loo_summary |>
  mutate(
    model_label = factor(model,
      levels = c("gcm", "kalman", "rule"),
      labels = c("GCM (exemplar)", "Kalman (prototype)", "Stan rule")
    ),
    dec_label = str_to_title(decision)
  ) |>
  ggplot(aes(x = elpd, y = model_label, color = path_dependent)) +
  geom_pointrange(
    aes(xmin = elpd - 2 * se_elpd, xmax = elpd + 2 * se_elpd),
    size = 0.7, linewidth = 1.1
  ) +
  facet_grid(dec_label ~ session,
             labeller = labeller(session = function(s) paste("Session", s))) +
  scale_color_manual(
    values  = c("FALSE" = "#0072B2", "TRUE" = "#D55E00"),
    name    = "LOO valid?",
    labels  = c("FALSE" = "Exact", "TRUE" = "Approx. (path-dep.)")
  ) +
  labs(
    title    = "ELPD Comparison: Human Data",
    subtitle = "Points: ELPD; ranges: ±2 SE.\nOrange = approximate (GCM/Kalman path-dependent; LFO-CV needed).",
    x = "ELPD", y = NULL
  ) +
  theme(strip.text = element_text(face = "bold"))
```

## Stacking Weights

```{r 16_stacking}
stacking_df <- map_dfr(sessions, function(sess) {
  map_dfr(decisions, function(dec) {
    key <- paste0("s", sess, "_", dec)
    l <- list(
      gcm    = loo_results[[paste0(key, "_gcm")]],
      kalman = loo_results[[paste0(key, "_kalman")]],
      rule   = loo_results[[paste0(key, "_rule")]]
    )
    valid <- Filter(Negate(is.null), l)
    if (length(valid) < 2) return(NULL)
    w <- loo::loo_model_weights(valid, method = "stacking")
    tibble(
      session  = sess,
      decision = dec,
      model    = names(w),
      weight   = as.numeric(w)
    )
  })
})

ggplot(stacking_df,
       aes(x = weight, y = factor(model,
         levels = c("gcm", "kalman", "rule"),
         labels = c("GCM", "Kalman", "Rule")),
           fill = model)) +
  geom_col(width = 0.6) +
  facet_grid(decision ~ session,
             labeller = labeller(session = function(s) paste("Sess.", s),
                                 decision = str_to_title)) +
  scale_x_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_fill_manual(values = c(gcm = "#0072B2", kalman = "#009E73",
                                rule = "#E69F00")) +
  labs(
    title    = "Bayesian Stacking Weights: Human Data",
    subtitle = "Higher weight = more predictive information relative to other models",
    x = "Stacking weight", y = NULL
  ) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"))
```

## LFO-CV Approximation for Path-Dependent Models

Standard LOO is approximate for the GCM and Kalman filter. The `loo` package provides an importance-sampling approximation to LFO-CV that is more appropriate for sequential models:

```{r 16_lfo_cv, eval=FALSE}
# LFO-CV for the multilevel GCM (Session 1, dangerous)
# This is computationally expensive; run selectively.
# The full LFO-CV implementation requires a refit function that
# refits the model on the first t trials and returns the log
# predictive density for trial t+1.
# Reference: Bürkner, Gabry & Vehtari (2020), Stat, doi:10.1002/sta4.323

# Approximate LFO using loo::loo with moment matching is one option;
# full LFO-CV requires re-running HMC with truncated data, which we
# sketch here for the single-session GCM fit:

lfo_gcm_s1_dang <- loo::loo_approximate_posterior(
  x        = loo_results[["s1_dangerous_gcm"]]$psis_object,
  log_p    = all_fits[["s1_dangerous_gcm"]]$draws("log_lik", format = "matrix"),
  log_g    = all_fits[["s1_dangerous_gcm"]]$draws("log_lik", format = "matrix"),
  cores    = 4
)
print(lfo_gcm_s1_dang)
```

> **Interpreting the LOO/LFO-CV results**: When the ELPD difference between two
> models exceeds approximately 4 times its SE (`elpd_diff / se_diff > 4`),
> there is substantial evidence in favor of the higher-ELPD model. Differences
> smaller than 4 SE are consistent with both models being approximately
> equivalent on this dataset.

---

# 6. Individual Differences

One of the key advantages of Bayesian hierarchical modeling is the ability to characterize individual variation in cognitive strategy. We examine the distribution of estimated parameters across participants.

## GCM: Population-Level Parameters

```{r 16_gcm_pop_params}
# Extract population parameter posteriors from the multilevel GCM fits
gcm_pop_summary <- imap_dfr(
  Filter(function(x) !is.null(x),
         all_fits[grep("_gcm$", names(all_fits))]),
  function(fit, name) {
    sess <- as.integer(str_extract(name, "(?<=s)\\d"))
    dec  <- str_extract(name, "dangerous|nutritious")
    tryCatch({
      fit$summary(variables = c("pop_w", "pop_log_c_mean",
                                "pop_log_c_sd", "kappa")) |>
        mutate(session = sess, decision = dec)
    }, error = function(e) NULL)
  }
)

if (nrow(gcm_pop_summary) > 0) {
  cat("GCM population-level parameter summaries:\n")
  print(gcm_pop_summary |>
    dplyr::select(session, decision, variable, mean, q5, q95) |>
    arrange(session, decision, variable))
}
```

## GCM: Individual Attention Weights

```{r 16_gcm_ind_weights}
gcm_ind_w <- imap_dfr(
  Filter(function(x) !is.null(x),
         all_fits[grep("_gcm$", names(all_fits))]),
  function(fit, name) {
    sess <- as.integer(str_extract(name, "(?<=s)\\d"))
    dec  <- str_extract(name, "dangerous|nutritious")
    tryCatch({
      fit$summary(variables = "subj_w") |>
        mutate(
          session    = sess,
          decision   = dec,
          subject_id = as.integer(str_extract(variable, "(?<=\\[)\\d+")),
          feature    = as.integer(str_extract(variable, "(?<=,)\\d+"))
        )
    }, error = function(e) NULL)
  }
)

if (nrow(gcm_ind_w) > 0) {
  # Average attention weight per feature across participants, by session/decision
  w_avg <- gcm_ind_w |>
    group_by(session, decision, feature) |>
    summarise(mean_w = mean(mean, na.rm = TRUE),
              se_w   = sd(mean, na.rm = TRUE) / sqrt(n()),
              .groups = "drop")

  ggplot(w_avg, aes(x = factor(feature), y = mean_w,
                     color = decision, group = decision)) +
    geom_line(linewidth = 0.8) +
    geom_pointrange(aes(ymin = mean_w - se_w, ymax = mean_w + se_w),
                    size = 0.5) +
    facet_wrap(~session, labeller = labeller(
      session = function(s) paste("Session", s)
    )) +
    scale_color_manual(values = c(dangerous = "#D55E00",
                                   nutritious = "#0072B2")) +
    labs(
      title    = "Mean Estimated GCM Attention Weights by Feature",
      subtitle = "Error bars: ±1 SE across participants.\nDashed verticals mark truly diagnostic features per session.",
      x = "Feature (f1–f5)", y = "Mean attention weight",
      color = "Decision"
    ) +
    # Annotate truly diagnostic features based on session rules
    geom_vline(
      data = tribble(
        ~session, ~feature, ~decision,
        1, 1, "dangerous",  1, 3, "dangerous",  1, 4, "nutritious",
        2, 4, "dangerous",  2, 1, "nutritious",  2, 2, "nutritious",  2, 3, "nutritious",
        3, 4, "dangerous",  3, 5, "dangerous"
      ),
      aes(xintercept = feature),
      linetype = "dashed", color = "grey40", alpha = 0.5
    )
}
```

**Interpretation**: If humans direct attention as the model predicts, estimated attention weights should be highest for features that are diagnostic for the current session's rule. In Session 1, features f1 and f3 should dominate the dangerous decision (the rule is `f1 AND f3`), while f4 should dominate nutritious. Deviation from this pattern — e.g., high weight on f2 for dangerous in Session 1 — indicates that the GCM is absorbing individual-level noise into attention weight estimates, or that participants are not fully utilizing the relevant features.

## Kalman: Individual Observation Noise

```{r 16_kalman_ind_r}
kalman_r_summary <- imap_dfr(
  all_fits[grep("_kalman_ll$", names(all_fits))],
  function(ll_list, name) {
    # ll_list is list of per-subject log_lik matrices; we need the r_value
    # posteriors from the individual fits. These were not saved separately;
    # to extract individual r_value posteriors, the per-subject fits must
    # be stored. Restructure below if full per-subject fits are needed.
    # For now, report that per-subject r estimates require re-running the
    # per-subject fitting loop with full fit objects stored.
    tibble(note = "Per-subject r_value posteriors require storing full fit objects.")
  }
)

# ── Restructured approach: store full fits for a subset of participants ─────
# For individual-differences analysis, refit Kalman for all subjects storing
# the full fit objects (not just log_lik). This is shown here as the
# correct pattern; in practice, the fitting loop above should be modified
# to save full fit objects alongside log_lik matrices.

# Example for Session 1, dangerous, first 5 subjects:
example_kalman_r <- map_dfr(
  subjects[1:min(5, length(subjects))],
  function(subj) {
    s_dat <- training_data |>
      filter(session == 1, subject == subj) |>
      arrange(trial_within_subject)
    if (nrow(s_dat) == 0) return(NULL)
    k_dat <- build_kalman_data(s_dat, "dangerous")
    fit_k <- tryCatch(
      run_fit_single(mod_kalman, k_dat, seed = as.integer(subj) * 3),
      error = function(e) NULL
    )
    if (is.null(fit_k)) return(NULL)
    fit_k$summary("r_value") |>
      mutate(subject = subj, session = 1, decision = "dangerous")
  }
)

if (nrow(example_kalman_r) > 0) {
  cat("Example individual r_value estimates (Session 1, dangerous, 5 subjects):\n")
  print(example_kalman_r |>
    dplyr::select(subject, session, decision, mean, q5, q95))
}
```

## Rule Model: Individual Error Tolerance

```{r 16_rule_ind_eps}
# Same pattern as Kalman: example for first 5 subjects, session 1, dangerous
example_rule_eps <- map_dfr(
  subjects[1:min(5, length(subjects))],
  function(subj) {
    s_dat <- training_data |>
      filter(session == 1, subject == subj) |>
      arrange(trial_within_subject)
    if (nrow(s_dat) == 0) return(NULL)
    r_dat <- build_rule_data(s_dat, "dangerous")
    fit_r <- tryCatch(
      run_fit_single(mod_rule, r_dat, seed = as.integer(subj) * 7),
      error = function(e) NULL
    )
    if (is.null(fit_r)) return(NULL)
    fit_r$summary("error_prob") |>
      mutate(subject = subj, session = 1, decision = "dangerous")
  }
)

if (nrow(example_rule_eps) > 0) {
  cat("Example individual error_prob estimates (Session 1, dangerous, 5 subjects):\n")
  print(example_rule_eps |>
    dplyr::select(subject, session, decision, mean, q5, q95))
}
```

---

# 7. Posterior Predictive Checks

We check whether the fitted models reproduce the qualitative features of the human data: the shape of the learning curves, the identity of the most error-prone stimuli, and the proportion of each error type.

```{r 16_ppc}
# ── Learning curve PPC for multilevel GCM (Session 1, dangerous) ─────────────
gcm_fit_s1d <- all_fits[["s1_dangerous_gcm"]]

if (!is.null(gcm_fit_s1d)) {
  # The multilevel GCM does not have a y_pred generated quantity by default.
  # We use $generate_quantities() to add it if needed, or we compute
  # posterior predictive accuracy from log_lik directly.
  # Here we extract p(correct) implied by the log-likelihood as a proxy PPC.

  ll_mat <- gcm_fit_s1d$draws("log_lik", format = "matrix")
  # Pareto-k plot as one form of model check
  loo_gcm_s1d <- loo::loo(ll_mat, cores = 4)
  plot(loo_gcm_s1d, label_points = TRUE,
       main = "GCM Pareto-k: Session 1, Dangerous")

  # LOO-PIT (approximate for GCM — path-dependent)
  cat("\nPareto-k summary for GCM, Session 1, Dangerous:\n")
  print(summary(loo_gcm_s1d$diagnostics$pareto_k))
}
```

---

# 8. Synthesis and Discussion

## Which Model Best Accounts for Human Categorization?

The stacking weights from Section 5 provide a summary: models with near-zero stacking weight add no predictive information beyond the other models for this session and decision. Models with high stacking weight are carrying unique predictive signal.

The key predictions from Chapter 14's simulation section that can now be evaluated against human data:

**GCM predictions** (Chapter 14):
- Optimal attention weights should outperform equal weights in sessions where only a subset of features is diagnostic.
- Learning should be gradual and monotonic.
- The attention weight estimates should concentrate on the truly diagnostic features.

*Evaluate*: Do the posterior attention weights (Section 6) align with the session's true rule structure? A GCM that correctly identifies which features are diagnostic provides the most compelling cognitive account.

**Kalman filter predictions** (Chapter 14):
- The model should struggle most with Session 2's nutritious rule (at-least-2-of-3) and Session 3's complex disjunction, where the category distribution is not unimodal.
- The observation noise $r$ should be relatively consistent across participants, reflecting a shared perceptual precision.

*Evaluate*: Does the Kalman filter's ELPD drop most sharply in the sessions where disjunctive rules apply? Is the $r$ distribution narrow or wide across participants?

**Rule model predictions** (Chapter 14):
- Performance should be highest in Session 1 where the true rules are simple conjunctions or single features expressible within the candidate rule set.
- The model may show abrupt improvements corresponding to rule discovery.
- Error tolerance $\varepsilon$ should be low for participants who report clear rule discovery and high for participants who remain uncertain.

*Evaluate*: Does the rule model's ELPD advantage (if any) concentrate in the simpler-rule sessions? Is there evidence of abrupt learning — individual learning curves with "aha moment" shapes — that the GCM and Kalman filter cannot capture?

## Individual Differences

Beyond population-level model selection, the individual-level fits tell a richer story. Participants may genuinely differ in strategy:

- Some participants may show GCM-like attention focusing (high posterior weight on diagnostic features, high estimated $c$).
- Others may show Kalman-like smooth prototype updating (low estimated $r$, stable learning curves).
- Still others may show rule-like learning (low $\varepsilon$, abrupt performance changes).

A participant-level model comparison — selecting the best-fitting model for each individual — can reveal whether strategy heterogeneity is systematic (e.g., correlated with the individual vs. pair condition) or idiosyncratic.

## The Condition Effect (Individual vs. Pair)

Do pairs show different model fit patterns from individuals? Possible hypotheses:

- **Cognitive division of labour**: Pairs may divide attention between the dangerous and nutritious dimensions, each partner tracking one rule. This would appear as narrower attention weight distributions in the multilevel GCM for pairs (each subject attending strongly to one feature set).
- **Rule verbalization**: Pairs may more easily articulate and apply explicit rules, leading to higher stacking weight for the rule model in the pair condition.
- **Error correction**: Pairs may catch each other's errors, leading to lower estimated $\varepsilon$ or lower $r$ in the corresponding models.

---

# 9. Conclusion

This chapter applied the full Bayesian modeling pipeline — developed across Chapters 11–15 — to real human data from the Alien Game. The key findings are:

1. **Data quality**: The experimental design was verified against the stimulus rules; the `nutricious` typo was corrected; the response mapping was validated against the `correct` column.

2. **Descriptive patterns**: Human learning shows the expected complexity gradient across sessions, with the nutritious dimension often learned more slowly than dangerous (especially in sessions 2 and 3 where nutritious rules are more complex). The individual vs. pair condition difference provides a natural manipulation for testing whether social interaction affects cognitive strategy.

3. **Model comparison**: The ELPD comparison and stacking weights identify which model architecture best predicts held-out trial responses for each session and dimension. The GCM's path-dependence caveat is addressed by flagging all GCM and Kalman ELPD values as approximate; the rule model's exchangeable structure makes its LOO technically valid.

4. **Individual differences**: The multilevel GCM provides a principled account of participant heterogeneity in attention weights and sensitivity. The per-subject Kalman and rule fits reveal variability in observation noise and error tolerance.

5. **Process-level validation**: The attention weight estimates can be compared directly against the session-specific diagnostic features, providing a strong test of whether the GCM's selective attention mechanism captures what humans are actually doing.

> **Looking ahead to Chapter 17**: The conclusions chapter synthesizes the
> modeling findings across the full module — from the simple Kruschke task in
> Chapters 11–13, through the simulation study in Chapter 14, the formal model
> comparison in Chapter 15, and the human data analysis here. It discusses
> what we have learned about the cognitive architecture of categorization,
> points toward hybrid models (COVIS, SUSTAIN, ATRIUM), and reflects on the
> methodological lessons of applying the Bayesian workflow to sequential
> cognitive models.
