---
title: "Strategic Dynamics in Rock-Paper-Scissors"
format:
  html:
    code-fold: true
    code-tools: true
---

```{r ch16_setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "85%",
  dpi        = 300
)

# Toggle heavy computations. Set TRUE the first time you knit on a new
# machine; subsequent knits reuse saved results from simmodels/.
regenerate_simulations <- FALSE
regenerate_fits        <- regenerate_simulations
regenerate_sbc         <- regenerate_simulations
run_intensive_checks   <- regenerate_simulations

for (d in c("stan", "simdata", "simmodels", "figures", "data")) {
  if (!dir.exists(d)) dir.create(d)
}

if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse,
  here,
  posterior,
  cmdstanr,
  tidybayes,
  patchwork,
  bayesplot,
  furrr,
  future,
  loo,
  priorsense,
  SBC,
  ggrepel,
  gtools      # rdirichlet()
)

theme_set(theme_classic())
plan(sequential)

# ── RPS helpers available globally throughout the chapter ──────────────────
# Payoff matrix (rows = focal action, cols = opponent action; 1=R,2=P,3=S)
RPS_PAYOFF <- matrix(c(0, -1, 1,
                        1,  0, -1,
                       -1,  1,  0),
                     nrow = 3, byrow = TRUE)

# Numerically stable softmax for length-3 vectors
softmax3 <- function(x) { x <- x - max(x); exp(x) / sum(exp(x)) }

# Outcome from focal player's perspective: 1=win, 2=lose, 3=draw (vectorised)
rps_outcome <- function(action, opp_action) {
  diff <- (action - opp_action) %% 3L
  ifelse(diff == 0L, 3L, ifelse(diff == 1L, 1L, 2L))
}

# Apply relative shift to a previous action (1=stay, 2=CW, 3=CCW; vectorised)
rps_apply_shift <- function(prev, rel) (prev - 1L + rel - 1L) %% 3L + 1L

# Compute relative shift from prev to next action (returns 1/2/3; vectorised)
rps_rel_shift <- function(prev, next_action) (next_action - prev) %% 3L + 1L

# Prepare transition-level data (removes first trial per player)
prep_wsls_data <- function(df) {
  df |>
    arrange(id, t) |>
    group_by(id) |>
    mutate(prev_action = lag(action), prev_opp = lag(opponent_action)) |>
    filter(!is.na(prev_action)) |>
    mutate(
      rel_shift = rps_rel_shift(prev_action, action),
      outcome   = rps_outcome(prev_action, prev_opp)
    ) |>
    ungroup()
}
```

> **📍 Where we are in the Bayesian modeling workflow:**
> Chapters 1–9 built and validated models ranging from simple biased-coin
> agents to fully recursive Theory-of-Mind learners, all in the binary
> matching-pennies paradigm. This chapter extends the toolkit to the
> **three-choice** environment of **Rock-Paper-Scissors (RPS)**. The move
> from two to three options is not merely quantitative: it introduces
> **cyclic dominance**, a non-transitive payoff structure that has no
> direct binary analogue, and it opens the door to **social cycling** —
> the systematic, directional sequences through the action space
> documented by Wang, Xu, and Zhou (2014). We contrast three increasingly
> sophisticated agents (a Nash mixer, a Win-Stay Lose-Shift heuristic
> player, and a Dirichlet-tracking 0-ToM), apply our standard six-phase
> validation battery, fit a hierarchical model to human data, and use
> PSIS-LOO to decide which architecture best explains real human play.

# Strategic Dynamics in Rock-Paper-Scissors

## Introduction

Rock-Paper-Scissors is the "Drosophila of game theory." Its Nash
Equilibrium is trivial — play each option with probability $1/3$,
rendering yourself unexploitable — yet humans consistently deviate from
it. They have a slight preference for Rock on the first move, and they
follow **conditional response rules** that create systematic patterns over
subsequent rounds. Wang et al. (2014) studied 360 participants playing 300
rounds each at Zhejiang University and identified two reliable signatures:

1. **Win-Stay**: after winning, players disproportionately repeat the
   winning move.
2. **Lose-Shift clockwise**: after losing, players tend to shift to the
   move that occupies the next position in the $R \to P \to S \to R$
   cycle — that is, to the move that would have beaten the losing move.

These two tendencies together produce **social cycles**: extended runs in
which an individual drifts through the action space in a predictable
direction. They are a cognitive fingerprint, invisible to a Nash
analysis and invisible to a model that ignores sequential structure.

## The Estimand

Before writing a single line of code, we state precisely what we want to
learn. The estimand has three layers:

1. **Description.** For a given player, what are the posterior
   distributions over their win-stay probability, lose-shift probability,
   and draw-stay probability? This is the individual-level WSLS estimand.

2. **Mechanism.** Does a player who shows social cycling do so because of
   a simple outcome-conditional heuristic (WSLS) or because they are
   actively predicting the opponent's next action on the basis of a
   generative model (0-ToM)? These two processes can produce superficially
   similar choice sequences; PSIS-LOO lets us adjudicate between them.

3. **Population.** Across players, what is the distribution of WSLS
   parameters? Is Win-Stay stronger than Lose-Shift at the population
   level? These population-level inferences require a hierarchical model.

All three layers require us to carry individual-level posterior uncertainty
forward — which is exactly the architecture built in Ch. 6.

## Learning Objectives

After completing this chapter, you will be able to:

- **Describe cyclic dominance** and explain why the Nash equilibrium
  analysis of RPS provides an incomplete picture of human play.
- **Implement a Nash baseline** in Stan as a Dirichlet-Categorical model
  and use it to quantify deviation from uniform mixing.
- **Specify an outcome-conditional transition model** for Win-Stay,
  Lose-Shift behavior, correctly encoding the cyclic shift direction.
- **Extend the Dirichlet-forgetting 0-ToM** from binary matching pennies
  to three-choice games, deriving the expected-utility computation from
  the RPS payoff matrix.
- **Sketch the 1-ToM extension** for three choices, understanding why the
  recursion generalizes structurally but introduces additional
  identification challenges.
- **Apply the six-phase validation battery** to each model, including
  prior predictive checks, parameter recovery, SBC, and posterior
  predictive reproduction of social cycling.
- **Fit a hierarchical WSLS model** in Stan with non-centered
  parameterization and interpret the population-level estimates.
- **Compare models via PSIS-LOO** and interpret the ELPD differences in
  the light of the social-cycling phenomenon.

## The Data: Wang et al. (2014)

The data come from Wang, Xu, and Zhou (2014), "Social cycling and
conditional responses in the Rock-Paper-Scissors game," a study of 360
university students playing 300 rounds of RPS against a fixed computer
opponent that played uniformly at random. The landmark finding is that
human players violate Nash equilibrium in a structured, socially
contagious way.

The dataset should be placed at `data/rps_wang2014.csv` with at minimum
the columns `id` (player integer), `round` (1–300), `action`
(1 = Rock, 2 = Paper, 3 = Scissors), and `opponent_action` (same coding).
If the file is not present we generate synthetic data with the same
structural properties for demonstration purposes.

```{r ch16_load_data}
data_path <- here::here("data", "rps_wang2014.csv")

if (file.exists(data_path)) {
  rps_raw <- read_csv(data_path, show_col_types = FALSE)
  # Harmonise column names to: id, round, action, opponent_action
  if (!all(c("id", "round", "action", "opponent_action") %in% names(rps_raw))) {
    stop("rps_wang2014.csv must contain columns: id, round, action, opponent_action")
  }
  rps <- rps_raw |>
    dplyr::select(id, round, action, opponent_action) |>
    filter(!is.na(action), !is.na(opponent_action)) |>
    arrange(id, round) |>
    group_by(id) |>
    mutate(t = row_number()) |>
    ungroup()
  cat("Players:", n_distinct(rps$id),
      "  Total trials:", nrow(rps), "\n")
  using_synthetic <- FALSE
} else {
  message("Wang et al. data not found. Generating synthetic data for demonstration.")
  using_synthetic <- TRUE
}
```

```{r ch16_synthetic_data}
# Synthetic data generation — skipped when real data are available
if (using_synthetic) {
  simulate_wsls_agent <- function(n_trials, theta_win, theta_lose, theta_draw,
                                  seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    action     <- integer(n_trials)
    opp_action <- sample(1L:3L, n_trials, replace = TRUE)
    action[1]  <- sample(1L:3L, 1L)
    for (t in 2:n_trials) {
      out <- rps_outcome(action[t - 1L], opp_action[t - 1L])
      th  <- switch(out, theta_win, theta_lose, theta_draw)
      rel <- sample(1L:3L, 1L, prob = th)
      action[t] <- rps_apply_shift(action[t - 1L], rel)
    }
    tibble(action = action, opponent_action = opp_action)
  }

  set.seed(42)
  n_players <- 60; n_trials <- 300
  rps <- map_dfr(seq_len(n_players), function(i) {
    p_stay_win  <- rbeta(1, 8, 2)   # strong Win-Stay
    p_cw_win    <- (1 - p_stay_win) * runif(1, 0.3, 0.7)
    p_ccw_win   <- 1 - p_stay_win - p_cw_win
    p_cw_lose   <- rbeta(1, 7, 3)   # strong Lose-Shift CW
    p_stay_lose <- (1 - p_cw_lose) * runif(1, 0.3, 0.7)
    p_ccw_lose  <- 1 - p_cw_lose - p_stay_lose
    p_stay_draw <- rbeta(1, 4, 4)   # roughly uniform
    p_cw_draw   <- (1 - p_stay_draw) * runif(1, 0.4, 0.6)
    p_ccw_draw  <- 1 - p_stay_draw - p_cw_draw
    simulate_wsls_agent(n_trials,
                        theta_win  = c(p_stay_win,  p_cw_win,  p_ccw_win),
                        theta_lose = c(p_stay_lose, p_cw_lose, p_ccw_lose),
                        theta_draw = c(p_stay_draw, p_cw_draw, p_ccw_draw),
                        seed = i) |>
      mutate(id = i, round = row_number(), t = round)
  })
  cat("Synthetic players:", n_distinct(rps$id),
      "  Total trials:", nrow(rps), "\n")
}
```

### Descriptive statistics

```{r ch16_descriptives, fig.cap="Left: marginal action frequencies across all players (dashed = Nash 1/3). Right: marginal win rate by player (dashed = chance 1/3 against a random opponent)."}
# Marginal choice distribution
p_marg <- rps |>
  count(action) |>
  mutate(action_lbl = factor(action, 1:3, c("Rock", "Paper", "Scissors"))) |>
  ggplot(aes(x = action_lbl, y = n / sum(n))) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_hline(yintercept = 1/3, linetype = "dashed", color = "gray40") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Proportion", title = "Marginal action distribution")

win_rates <- rps |>
  mutate(win = rps_outcome(action, opponent_action) == 1L) |>
  group_by(id) |>
  summarize(win_rate = mean(win), .groups = "drop")

p_wins <- ggplot(win_rates, aes(x = win_rate)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 1/3, linetype = "dashed", color = "gray40") +
  labs(x = "Win rate", y = "Count",
       title = "Win rate distribution across players")

p_marg | p_wins
```

### Social cycling: transition heatmaps

The key empirical signature of Wang et al. (2014) is the
**outcome-conditional transition matrix**: the probability of making each
relative shift (Stay, Clockwise, Counter-clockwise) given the previous
round's outcome (Win, Lose, Draw). If players are Nash-rational, all
transitions should be uniform. Social cycling predicts
$P(\text{Stay} \mid \text{Win}) \gg 1/3$ and
$P(\text{CW} \mid \text{Lose}) \gg 1/3$.

```{r ch16_cycling_plot, fig.cap="Observed outcome-conditional transition frequencies. Each panel is one previous outcome; bars show the proportion of Stay, Clockwise, and Counter-clockwise shifts. Dashed line = uniform 1/3. Win-Stay and Lose-Shift-CW are the signatures of social cycling."}
# Compute relative shift: (action_t - action_{t-1}) mod 3 + 1
# 1 = Stay, 2 = CW, 3 = CCW
rps_trans <- prep_wsls_data(rps)

rps_trans |>
  mutate(
    outcome_lbl   = factor(outcome,   1:3, c("Win", "Lose", "Draw")),
    rel_shift_lbl = factor(rel_shift, 1:3, c("Stay", "CW", "CCW"))
  ) |>
  count(outcome_lbl, rel_shift_lbl) |>
  group_by(outcome_lbl) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = rel_shift_lbl, y = prop, fill = rel_shift_lbl)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 1/3, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = c("steelblue", "seagreen3", "firebrick3"),
                    guide = "none") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~outcome_lbl) +
  labs(x = "Shift direction", y = "Proportion",
       title = "Observed outcome-conditional transitions",
       subtitle = "Dashed = Nash prediction (1/3 uniform)")
```

The plot is the empirical justification for this entire chapter. The
Win-Stay and Lose-Shift-CW signatures are visible. Our job is to determine
which cognitive model best accounts for them at the individual level.

---

## Model 1: The Nash Agent (Baseline)

### Theory

The Nash equilibrium of RPS is the unique mixed strategy that makes an
opponent indifferent between all three moves: play each with probability
$1/3$. The Nash Agent model asks: what is a player's *fixed* probability
distribution over moves, ignoring all sequential structure? Deviations
from $[1/3, 1/3, 1/3]$ — e.g., a "Rock preference" — are meaningful
evidence against Nash mixing, but the model cannot capture social cycling.

We model choices as i.i.d. draws from a fixed categorical distribution with
Dirichlet prior:
$$
c_t \sim \text{Categorical}(\theta),
\qquad
\theta \sim \text{Dirichlet}(\alpha_0),
\qquad
\alpha_0 = (2, 2, 2).
$$

The $\text{Dirichlet}(2, 2, 2)$ prior is weakly informative: it places
more mass near $[1/3, 1/3, 1/3]$ than a flat $\text{Dirichlet}(1,1,1)$
but allows substantial heterogeneity.

### R simulator

```{r ch16_nash_sim}
simulate_nash <- function(n_trials, theta, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  tibble(
    t      = seq_len(n_trials),
    action = sample(1L:3L, n_trials, replace = TRUE, prob = theta)
  )
}
```

### Stan model

```{r ch16_nash_stan}
stan_nash <- "
data {
  int<lower=1> N;
  array[N] int<lower=1, upper=3> action;
}
parameters {
  simplex[3] theta;
}
model {
  theta ~ dirichlet(rep_vector(2.0, 3));
  action ~ categorical(theta);
}
generated quantities {
  vector[N] log_lik;
  array[N] int action_rep;
  for (t in 1:N) {
    log_lik[t]    = categorical_lpmf(action[t] | theta);
    action_rep[t] = categorical_rng(theta);
  }
}
"
writeLines(stan_nash, here::here("stan", "ch16_nash_single.stan"))
mod_nash <- cmdstan_model(here::here("stan", "ch16_nash_single.stan"))
```

### Phase 1 — Prior predictive check

Do our priors generate plausible choice distributions? A $\text{Dirichlet}
(2,2,2)$ should produce heterogeneous but reasonable biases.

```{r ch16_nash_ppc, fig.cap="Prior predictive check for the Nash model. Each panel is one draw from Dirichlet(2,2,2), showing 120 simulated choices. Bars show action frequencies; the dashed line is 1/3. Prior generates diverse biases without saturating any single action."}
set.seed(2026)
n_pp <- 9
prior_nash <- tibble(
  draw  = 1:n_pp,
  theta = map(1:n_pp, ~ c(rdirichlet(1, c(2, 2, 2))))
) |>
  rowwise() |>
  mutate(dat = list(simulate_nash(120, theta[[1]], seed = draw))) |>
  dplyr::select(draw, dat) |>
  unnest(dat) |>
  count(draw, action) |>
  group_by(draw) |>
  mutate(prop = n / sum(n))

ggplot(prior_nash, aes(x = factor(action, 1:3, c("R","P","S")),
                       y = prop, fill = factor(action))) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 1/3, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = c("steelblue","seagreen3","firebrick3"),
                    guide = "none") +
  facet_wrap(~draw, labeller = label_both) +
  labs(x = "Action", y = "Proportion",
       title = "Nash prior predictive: 9 draws from Dirichlet(2,2,2)")
```

### Phase 2 — Parameter recovery

```{r ch16_nash_recovery}
nash_rec_path <- here::here("simmodels", "ch16_nash_recovery.rds")

if (regenerate_simulations || !file.exists(nash_rec_path)) {
  set.seed(101)
  n_agents <- 20
  truth_nash <- tibble(
    agent = 1:n_agents,
    theta = map(1:n_agents, ~ c(rdirichlet(1, c(2, 2, 2))))
  )

  fit_nash_one <- function(th, ag) {
    sim <- simulate_nash(200, th, seed = ag)
    fit <- mod_nash$sample(
      data = list(N = nrow(sim), action = sim$action),
      chains = 2, parallel_chains = 2,
      iter_warmup = 500, iter_sampling = 500,
      refresh = 0, show_messages = FALSE
    )
    fit$summary(c("theta[1]","theta[2]","theta[3]")) |>
      dplyr::select(variable, mean, q5, q95)
  }

  rec_nash <- truth_nash |>
    rowwise() |>
    mutate(post = list(fit_nash_one(theta[[1]], agent))) |>
    dplyr::select(agent, theta, post) |>
    unnest(post) |>
    rowwise() |>
    mutate(
      true_val = switch(variable,
                        "theta[1]" = theta[[1]],
                        "theta[2]" = theta[[2]],
                        "theta[3]" = theta[[3]],
                        NA_real_)
    ) |>
    ungroup()
  saveRDS(rec_nash, nash_rec_path)
} else {
  rec_nash <- readRDS(nash_rec_path)
}
```

```{r ch16_nash_recovery_plot, fig.cap="Nash parameter recovery. Each point is one simulated agent's posterior mean vs. true theta. Bars are 90% CIs. All three components recover cleanly at 200 trials."}
ggplot(rec_nash, aes(x = true_val, y = mean)) +
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0, alpha = 0.4,
                color = "steelblue") +
  geom_point(color = "midnightblue", size = 2) +
  geom_abline(linetype = "dashed", color = "gray40") +
  facet_wrap(~variable, scales = "free",
             labeller = as_labeller(c(`theta[1]` = "theta[R]",
                                      `theta[2]` = "theta[P]",
                                      `theta[3]` = "theta[S]"))) +
  labs(x = "True value", y = "Posterior mean",
       title = "Nash parameter recovery (20 agents, 200 trials)")
```

### Phase 3 — Simulation-Based Calibration (SBC)

```{r ch16_nash_sbc}
sbc_nash_path <- here::here("simmodels", "ch16_nash_sbc.rds")

gen_nash_sbc <- function(N = 200) {
  theta <- c(rdirichlet(1, c(2, 2, 2)))
  sim   <- simulate_nash(N, theta)
  list(
    variables = list(`theta[1]` = theta[1],
                     `theta[2]` = theta[2],
                     `theta[3]` = theta[3]),
    generated = list(N = N, action = sim$action)
  )
}

if (regenerate_sbc || !file.exists(sbc_nash_path)) {
  sbc_gen_n  <- SBC_generator_function(gen_nash_sbc, N = 200)
  sbc_back_n <- SBC_backend_cmdstan_sample(
    mod_nash, iter_warmup = 500, iter_sampling = 500,
    chains = 1, refresh = 0
  )
  sbc_ds_n   <- generate_datasets(sbc_gen_n, 200)
  sbc_res_n  <- compute_SBC(sbc_ds_n, sbc_back_n, keep_fits = FALSE)
  saveRDS(list(ds = sbc_ds_n, results = sbc_res_n), sbc_nash_path)
} else {
  obj        <- readRDS(sbc_nash_path)
  sbc_ds_n   <- obj$ds
  sbc_res_n  <- obj$results
}
```

```{r ch16_nash_sbc_plot, fig.width=9, fig.height=4, fig.cap="SBC for the Nash model. Rank histograms (top) and ECDF differences (bottom) for all three simplex components. Flat histograms and bands within the grey envelope confirm calibrated posteriors."}
plot_rank_hist(sbc_res_n) / plot_ecdf_diff(sbc_res_n)
```

### Phase 4 — Posterior predictive check

We fit the Nash model to each player and check whether it can reproduce
the outcome-conditional transition pattern that characterizes social
cycling. The Nash model has no memory: it cannot reproduce sequential
structure.

```{r ch16_nash_ppc_data}
nash_ppc_path <- here::here("simmodels", "ch16_nash_ppc.rds")

fit_nash_player <- function(df) {
  mod_nash$sample(
    data = list(N = nrow(df), action = df$action),
    chains = 2, parallel_chains = 2,
    iter_warmup = 500, iter_sampling = 500,
    refresh = 0, show_messages = FALSE
  )
}

# Fit to a single illustrative player for the posterior predictive
demo_id <- rps |> pull(id) |> unique() |> first()
demo_df  <- filter(rps, id == demo_id)

if (regenerate_fits || !file.exists(nash_ppc_path)) {
  fit_nash_demo <- fit_nash_player(demo_df)
  theta_draws <- fit_nash_demo$draws("theta", format = "matrix")
  ppc_nash <- map_dfr(1:100, function(s) {
    th  <- theta_draws[s, ]
    sim <- simulate_nash(nrow(demo_df), th)
    prev_a <- lag(sim$action)
    opp_a  <- demo_df$opponent_action   # same opponent sequence
    tibble(action = sim$action[-1], prev_action = prev_a[-1],
           opp_action = opp_a[-1], rep = s)
  }) |>
    mutate(
      rel_shift = rps_rel_shift(prev_action, action),
      outcome   = rps_outcome(prev_action, opp_action)
    )
  saveRDS(ppc_nash, nash_ppc_path)
} else {
  ppc_nash <- readRDS(nash_ppc_path)
}
```

```{r ch16_nash_ppc_plot, fig.cap="Nash posterior predictive check. Grey bars = posterior predictive replicates (100 draws). Coloured error bars = observed proportions with 95% CI. The Nash model cannot reproduce the Win-Stay or Lose-Shift-CW excess — evidence that sequential structure is a real feature of the data."}
obs_trans <- rps_trans |>
  filter(id == demo_id) |>
  count(outcome, rel_shift) |>
  group_by(outcome) |>
  mutate(prop_obs = n / sum(n)) |>
  mutate(
    outcome_lbl   = factor(outcome,   1:3, c("Win","Lose","Draw")),
    rel_shift_lbl = factor(rel_shift, 1:3, c("Stay","CW","CCW"))
  )

ppc_trans <- ppc_nash |>
  count(rep, outcome, rel_shift) |>
  group_by(rep, outcome) |>
  mutate(prop = n / sum(n)) |>
  mutate(
    outcome_lbl   = factor(outcome,   1:3, c("Win","Lose","Draw")),
    rel_shift_lbl = factor(rel_shift, 1:3, c("Stay","CW","CCW"))
  )

ppc_sum <- ppc_trans |>
  group_by(outcome_lbl, rel_shift_lbl) |>
  summarize(m = mean(prop), lo = quantile(prop, 0.025),
            hi = quantile(prop, 0.975), .groups = "drop")

ggplot(ppc_sum, aes(x = rel_shift_lbl)) +
  geom_col(aes(y = m), fill = "gray70", alpha = 0.8) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.2, color = "gray40") +
  geom_point(data = obs_trans, aes(y = prop_obs),
             color = "firebrick3", size = 3) +
  geom_hline(yintercept = 1/3, linetype = "dashed") +
  facet_wrap(~outcome_lbl) +
  labs(x = "Shift", y = "Proportion",
       title = "Nash posterior predictive vs. observed transitions",
       subtitle = "Grey = model, red = observed")
```

The Nash model treats all sequential transitions as i.i.d. and
therefore predicts uniform shift proportions regardless of the
previous outcome. The observed Win-Stay and Lose-Shift-CW excesses fall
outside the model's predictive envelope, establishing that sequential
structure must be modeled explicitly.

---

## Model 2: The WSLS Agent

### Theory

The Win-Stay Lose-Shift (WSLS) agent does not model the opponent at all.
It is a reactive heuristic: after observing the outcome of the last round,
it shifts through the action space according to a fixed probability
distribution that depends on whether it won, lost, or drew.

**Encoding.** We encode transitions as *relative shifts* in the
$R \to P \to S \to R$ cycle:

| Relative shift | Code | Example from Rock |
|----------------|------|-------------------|
| Stay           |   1  | R → R             |
| Clockwise (CW) |   2  | R → P             |
| Counter-CW     |   3  | R → S             |

For each previous outcome $o \in \{\text{Win}=1, \text{Lose}=2,
\text{Draw}=3\}$ the agent draws its next relative shift from
$\theta_o = [\theta_{o,\text{Stay}}, \theta_{o,\text{CW}},
\theta_{o,\text{CCW}}]$. A pure WSLS player has
$\theta_{\text{Win}} \approx [1, 0, 0]$ and
$\theta_{\text{Lose}} \approx [0, 1, 0]$. The data will tell us how
strongly this pattern holds and whether it varies across individuals.

**Prior.** We encode the WSLS intuition via outcome-specific Dirichlet
concentrations:
- $\theta_{\text{Win}} \sim \text{Dirichlet}(5, 1, 1)$ — biased toward Stay.
- $\theta_{\text{Lose}} \sim \text{Dirichlet}(1, 5, 1)$ — biased toward CW shift.
- $\theta_{\text{Draw}} \sim \text{Dirichlet}(2, 2, 2)$ — weakly uniform.

These priors are *informative* in direction but permissive in magnitude —
the posterior can still move far from the prior center if the data warrant.

### R simulator

```{r ch16_wsls_sim}
simulate_wsls <- function(n_trials, theta_win, theta_lose, theta_draw,
                          op_choices = NULL, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  if (is.null(op_choices))
    op_choices <- sample(1L:3L, n_trials, replace = TRUE)
  action    <- integer(n_trials)
  action[1] <- sample(1L:3L, 1L)
  outcome   <- integer(n_trials)
  for (t in 2:n_trials) {
    out        <- rps_outcome(action[t - 1L], op_choices[t - 1L])
    outcome[t - 1L] <- out
    th         <- switch(out, theta_win, theta_lose, theta_draw)
    rel        <- sample(1L:3L, 1L, prob = th)
    action[t]  <- rps_apply_shift(action[t - 1L], rel)
  }
  outcome[n_trials] <- rps_outcome(action[n_trials], op_choices[n_trials])
  tibble(t = seq_len(n_trials), action = action, opponent_action = op_choices,
         outcome = outcome)
}
```

### Stan model

```{r ch16_wsls_stan}
stan_wsls <- "
data {
  int<lower=1> N;                            // number of transitions
  array[N] int<lower=1, upper=3> rel_shift;  // relative shift (1=stay, 2=CW, 3=CCW)
  array[N] int<lower=1, upper=3> outcome;    // previous outcome (1=win, 2=lose, 3=draw)
}
parameters {
  array[3] simplex[3] theta_rel;  // theta_rel[outcome] = transition probs
}
model {
  // WSLS-informative priors
  theta_rel[1] ~ dirichlet([5.0, 1.0, 1.0]');  // Win  -> Stay
  theta_rel[2] ~ dirichlet([1.0, 5.0, 1.0]');  // Lose -> CW shift
  theta_rel[3] ~ dirichlet([2.0, 2.0, 2.0]');  // Draw -> uniform
  for (t in 1:N)
    rel_shift[t] ~ categorical(theta_rel[outcome[t]]);
}
generated quantities {
  vector[N] log_lik;
  array[N] int shift_rep;
  for (t in 1:N) {
    log_lik[t]   = categorical_lpmf(rel_shift[t] | theta_rel[outcome[t]]);
    shift_rep[t] = categorical_rng(theta_rel[outcome[t]]);
  }
}
"
writeLines(stan_wsls, here::here("stan", "ch16_wsls_single.stan"))
mod_wsls <- cmdstan_model(here::here("stan", "ch16_wsls_single.stan"))
```

Note that the WSLS model is a transition model: it predicts relative
shifts, not absolute actions. The likelihood conditions on the previous
outcome and marginalizes over absolute action (since $\theta_{\text{Win}}$
has the same meaning regardless of whether the agent played Rock, Paper,
or Scissors last round — a key advantage of the relative-shift encoding).

### Preparing the data for WSLS

```{r ch16_wsls_data_prep}
rps_wsls <- prep_wsls_data(rps)
```

### Phase 1 — Prior predictive check

```{r ch16_wsls_ppc1, fig.cap="WSLS prior predictive check. 9 simulated players, each with theta drawn from the WSLS-informative Dirichlet priors. Each panel shows the observed transition heatmap. The priors reliably generate Win-Stay and Lose-Shift-CW patterns without forcing perfectly deterministic behavior."}
set.seed(2026)
prior_wsls <- tibble(
  draw       = 1:9,
  theta_win  = map(1:9, ~ c(rdirichlet(1, c(5, 1, 1)))),
  theta_lose = map(1:9, ~ c(rdirichlet(1, c(1, 5, 1)))),
  theta_draw = map(1:9, ~ c(rdirichlet(1, c(2, 2, 2))))
) |>
  rowwise() |>
  mutate(dat = list({
    sim <- simulate_wsls(150, theta_win[[1]], theta_lose[[1]], theta_draw[[1]],
                         seed = draw)
    sim$id <- draw
    prep_wsls_data(sim) |> mutate(draw = draw)
  })) |>
  dplyr::select(draw, dat) |>
  unnest(dat)

prior_wsls |>
  mutate(
    outcome_lbl   = factor(outcome,   1:3, c("Win","Lose","Draw")),
    shift_lbl     = factor(rel_shift, 1:3, c("Stay","CW","CCW"))
  ) |>
  count(draw, outcome_lbl, shift_lbl) |>
  group_by(draw, outcome_lbl) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = shift_lbl, y = prop, fill = shift_lbl)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 1/3, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = c("steelblue","seagreen3","firebrick3"),
                    guide = "none") +
  facet_grid(draw ~ outcome_lbl) +
  labs(x = "Shift", y = "Proportion",
       title = "WSLS prior predictive: 9 prior draws × 3 outcomes")
```

### Phase 2 — Parameter recovery

```{r ch16_wsls_recovery}
wsls_rec_path <- here::here("simmodels", "ch16_wsls_recovery.rds")

if (regenerate_simulations || !file.exists(wsls_rec_path)) {
  set.seed(202)
  n_agents <- 20
  truth_wsls <- tibble(
    agent      = 1:n_agents,
    theta_win  = map(1:n_agents, ~ c(rdirichlet(1, c(5, 1, 1)))),
    theta_lose = map(1:n_agents, ~ c(rdirichlet(1, c(1, 5, 1)))),
    theta_draw = map(1:n_agents, ~ c(rdirichlet(1, c(2, 2, 2))))
  )

  fit_wsls_one <- function(tw, tl, td, ag) {
    sim    <- simulate_wsls(300, tw, tl, td, seed = ag)
    sim$id <- 1L
    dd     <- prep_wsls_data(sim)
    fit <- mod_wsls$sample(
      data = list(N = nrow(dd), rel_shift = dd$rel_shift,
                  outcome = dd$outcome),
      chains = 2, parallel_chains = 2,
      iter_warmup = 500, iter_sampling = 500,
      refresh = 0, show_messages = FALSE
    )
    fit$summary(c("theta_rel[1,1]","theta_rel[2,2]","theta_rel[3,1]")) |>
      dplyr::select(variable, mean, q5, q95)
  }

  rec_wsls <- truth_wsls |>
    rowwise() |>
    mutate(post = list(fit_wsls_one(theta_win[[1]], theta_lose[[1]],
                                    theta_draw[[1]], agent))) |>
    dplyr::select(agent, theta_win, theta_lose, theta_draw, post) |>
    unnest(post) |>
    rowwise() |>
    mutate(
      true_val = case_when(
        variable == "theta_rel[1,1]" ~ theta_win[[1]],   # Win-Stay
        variable == "theta_rel[2,2]" ~ theta_lose[[2]],  # Lose-CW
        variable == "theta_rel[3,1]" ~ theta_draw[[1]],  # Draw-Stay
        TRUE ~ NA_real_
      )
    ) |>
    ungroup()
  saveRDS(rec_wsls, wsls_rec_path)
} else {
  rec_wsls <- readRDS(wsls_rec_path)
}
```

```{r ch16_wsls_recovery_plot, fig.cap="WSLS parameter recovery for the three key parameters: Win-Stay probability (theta_rel[1,1]), Lose-Shift-CW probability (theta_rel[2,2]), and Draw-Stay probability (theta_rel[3,1]). Points = posterior means, bars = 90% CIs. All three recover cleanly at 300 trials."}
lbl <- c(`theta_rel[1,1]` = "Win-Stay",
          `theta_rel[2,2]` = "Lose-CW",
          `theta_rel[3,1]` = "Draw-Stay")
ggplot(rec_wsls, aes(x = true_val, y = mean)) +
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0, alpha = 0.4,
                color = "steelblue") +
  geom_point(color = "midnightblue", size = 2) +
  geom_abline(linetype = "dashed", color = "gray40") +
  facet_wrap(~variable, scales = "free", labeller = as_labeller(lbl)) +
  labs(x = "True value", y = "Posterior mean",
       title = "WSLS parameter recovery (20 agents, 300 trials)")
```

### Phase 3 — Simulation-Based Calibration

```{r ch16_wsls_sbc}
sbc_wsls_path <- here::here("simmodels", "ch16_wsls_sbc.rds")

gen_wsls_sbc <- function(N_trials = 300) {
  tw  <- c(rdirichlet(1, c(5, 1, 1)))
  tl  <- c(rdirichlet(1, c(1, 5, 1)))
  td  <- c(rdirichlet(1, c(2, 2, 2)))
  sim <- simulate_wsls(N_trials, tw, tl, td)
  sim$id <- 1L
  dd  <- prep_wsls_data(sim)
  list(
    variables = list(
      `theta_rel[1,1]` = tw[1], `theta_rel[1,2]` = tw[2],
      `theta_rel[2,1]` = tl[1], `theta_rel[2,2]` = tl[2],
      `theta_rel[3,1]` = td[1], `theta_rel[3,2]` = td[2]
    ),
    generated = list(N = nrow(dd), rel_shift = dd$rel_shift,
                     outcome = dd$outcome)
  )
}

if (regenerate_sbc || !file.exists(sbc_wsls_path)) {
  sbc_gen_w  <- SBC_generator_function(gen_wsls_sbc, N_trials = 300)
  sbc_back_w <- SBC_backend_cmdstan_sample(
    mod_wsls, iter_warmup = 500, iter_sampling = 500,
    chains = 1, refresh = 0
  )
  sbc_ds_w  <- generate_datasets(sbc_gen_w, 200)
  sbc_res_w <- compute_SBC(sbc_ds_w, sbc_back_w, keep_fits = FALSE)
  saveRDS(list(ds = sbc_ds_w, results = sbc_res_w), sbc_wsls_path)
} else {
  obj       <- readRDS(sbc_wsls_path)
  sbc_ds_w  <- obj$ds
  sbc_res_w <- obj$results
}
```

```{r ch16_wsls_sbc_plot, fig.width=9, fig.height=5, fig.cap="SBC for the WSLS model. The six free simplex parameters (two from each of the three 3-simplices) show flat rank histograms and ECDF differences within the calibration band — the Stan implementation is correct and the posteriors are well calibrated at 300 trials."}
vars_wsls <- c("theta_rel[1,1]","theta_rel[1,2]",
               "theta_rel[2,1]","theta_rel[2,2]",
               "theta_rel[3,1]","theta_rel[3,2]")
plot_rank_hist(sbc_res_w, variables = vars_wsls) /
  plot_ecdf_diff(sbc_res_w, variables = vars_wsls)
```

### Phase 4 — Posterior predictive check

We now verify that the WSLS model can reproduce the outcome-conditional
transition pattern that the Nash model failed to capture.

```{r ch16_wsls_ppc_fit}
wsls_ppc_path <- here::here("simmodels", "ch16_wsls_ppc.rds")

if (regenerate_fits || !file.exists(wsls_ppc_path)) {
  dd_demo <- rps_wsls |> filter(id == demo_id)
  fit_wsls_demo <- mod_wsls$sample(
    data = list(N = nrow(dd_demo), rel_shift = dd_demo$rel_shift,
                outcome = dd_demo$outcome),
    chains = 2, parallel_chains = 2,
    iter_warmup = 500, iter_sampling = 500,
    refresh = 0, show_messages = FALSE
  )
  theta_draws_w <- fit_wsls_demo$draws(
    c("theta_rel[1,1]","theta_rel[1,2]","theta_rel[1,3]",
      "theta_rel[2,1]","theta_rel[2,2]","theta_rel[2,3]",
      "theta_rel[3,1]","theta_rel[3,2]","theta_rel[3,3]"),
    format = "draws_df"
  )
  ppc_wsls <- map_dfr(1:100, function(s) {
    tw  <- unlist(theta_draws_w[s, c("theta_rel[1,1]","theta_rel[1,2]","theta_rel[1,3]")])
    tl  <- unlist(theta_draws_w[s, c("theta_rel[2,1]","theta_rel[2,2]","theta_rel[2,3]")])
    td  <- unlist(theta_draws_w[s, c("theta_rel[3,1]","theta_rel[3,2]","theta_rel[3,3]")])
    sim <- simulate_wsls(nrow(demo_df), tw, tl, td,
                         op_choices = demo_df$opponent_action, seed = s)
    sim$id <- 1L
    prep_wsls_data(sim) |> mutate(rep = s)
  })
  saveRDS(ppc_wsls, wsls_ppc_path)
} else {
  ppc_wsls <- readRDS(wsls_ppc_path)
}
```

```{r ch16_wsls_ppc_plot, fig.cap="WSLS posterior predictive check. Grey bars = posterior predictive replicates; red points = observed. The WSLS model closely reproduces the Win-Stay and Lose-Shift-CW pattern, validating the model as a plausible account of the sequential structure."}
ppc_wsls_sum <- ppc_wsls |>
  mutate(
    outcome_lbl = factor(outcome, 1:3, c("Win","Lose","Draw")),
    shift_lbl   = factor(rel_shift, 1:3, c("Stay","CW","CCW"))
  ) |>
  count(rep, outcome_lbl, shift_lbl) |>
  group_by(rep, outcome_lbl) |>
  mutate(prop = n / sum(n)) |>
  group_by(outcome_lbl, shift_lbl) |>
  summarize(m = mean(prop), lo = quantile(prop, 0.025),
            hi = quantile(prop, 0.975), .groups = "drop")

obs_wsls_demo <- rps_wsls |>
  filter(id == demo_id) |>
  mutate(
    outcome_lbl = factor(outcome, 1:3, c("Win","Lose","Draw")),
    shift_lbl   = factor(rel_shift, 1:3, c("Stay","CW","CCW"))
  ) |>
  count(outcome_lbl, shift_lbl) |>
  group_by(outcome_lbl) |>
  mutate(prop_obs = n / sum(n))

ggplot(ppc_wsls_sum, aes(x = shift_lbl)) +
  geom_col(aes(y = m), fill = "gray70", alpha = 0.8) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.2, color = "gray40") +
  geom_point(data = obs_wsls_demo, aes(y = prop_obs),
             color = "firebrick3", size = 3) +
  geom_hline(yintercept = 1/3, linetype = "dashed") +
  facet_wrap(~outcome_lbl) +
  labs(x = "Shift", y = "Proportion",
       title = "WSLS posterior predictive vs. observed",
       subtitle = "Grey = model predictive, red = observed")
```

---

## Model 3: The 0-ToM Agent (Dirichlet Tracker)

### Theory

The 0-ToM agent in RPS does not merely react to the previous outcome; it
maintains an evolving *generative model* of the opponent's action
distribution and best-responds to it. This is the three-choice
generalization of the Kalman-filter 0-ToM from Ch. 9.

**Belief state.** The agent maintains a Dirichlet count vector
$\alpha_t = [\alpha_{t,R}, \alpha_{t,P}, \alpha_{t,S}]$ representing
its belief about the opponent's action distribution. The Dirichlet mean
$\hat{p}^{\text{op}}_t = \alpha_t / \sum \alpha_t$ is the agent's
point-estimate prediction of the opponent's next move.

**Forgetting.** On each trial, before incorporating the new observation,
the agent applies exponential forgetting:
$$
\alpha_t \leftarrow \rho\,\alpha_{t-1} + (1-\rho)\,\mathbf{1},
\qquad
\rho = e^{-e^{\log\sigma}},
$$
where $\mathbf{1} = [1,1,1]^\top$ returns the counts toward a uniform
prior and $\sigma > 0$ controls the forgetting rate. When $\sigma = 0$,
$\rho = 1$ and the agent accumulates counts perfectly (no forgetting).
When $\sigma \to \infty$, $\rho \to 0$ and the agent resets to uniform
each step, tracking only the most recent observation.

**Expected utility.** Let $\Pi$ be the $3 \times 3$ RPS payoff matrix with
$\Pi_{k,j} \in \{-1, 0, 1\}$. The expected utility of each action $k$ is
$$
\text{EU}_{t,k} = \sum_j \Pi_{k,j}\,\hat{p}^{\text{op}}_{t,j}.
$$
Action selection uses a softmax with temperature $\beta > 0$:
$$
P(c_t = k) \propto \exp\!\left(\text{EU}_{t,k} / \beta\right).
$$

Two parameters: $\log\sigma$ (forgetting rate) and $\log\beta$
(decision temperature). Both are on the log scale to enforce positivity.

**Relationship to Ch. 9.** The binary 0-ToM in Ch. 9 used a Kalman filter
on the log-odds to track a Bernoulli opponent. The Dirichlet tracker here
is the natural multinomial analogue: it tracks a categorical opponent using
a running Dirichlet posterior with exponential forgetting. The two share
the same volatility-forgetting structure; the Dirichlet version has the
additional advantage of straightforward generalisation to $K > 3$ choices.

### R simulator

```{r ch16_tom0_sim}
simulate_tom0_rps <- function(op_choices, log_sigma, log_beta, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  rho  <- exp(-exp(log_sigma))
  beta <- exp(log_beta)
  N    <- length(op_choices)
  alpha  <- rep(1, 3)
  choice <- integer(N)
  for (t in 1:N) {
    p_op <- alpha / sum(alpha)
    EU   <- RPS_PAYOFF %*% p_op
    p_self <- softmax3(as.vector(EU) / beta)
    choice[t] <- sample(1L:3L, 1L, prob = p_self)
    alpha <- rho * alpha + (1 - rho) * rep(1, 3)
    alpha[op_choices[t]] <- alpha[op_choices[t]] + 1
  }
  tibble(t = seq_len(N), action = choice, opponent_action = op_choices)
}
```

### Stan model

```{r ch16_tom0_stan}
stan_tom0_rps <- "
data {
  int<lower=1> N;
  array[N] int<lower=1, upper=3> action;
  array[N] int<lower=1, upper=3> op_action;
}
transformed data {
  matrix[3, 3] payoff = [[0.0, -1.0,  1.0],
                          [1.0,  0.0, -1.0],
                          [-1.0, 1.0,  0.0]];
}
parameters {
  real log_sigma;
  real log_beta;
}
transformed parameters {
  matrix[N, 3] EU;  // expected utility of each action
  {
    vector[3] alpha = rep_vector(1.0, 3);
    real rho  = exp(-exp(log_sigma));
    real beta = exp(log_beta);
    for (t in 1:N) {
      vector[3] p_op = alpha / sum(alpha);
      for (k in 1:3)
        EU[t, k] = (payoff[k] * p_op) / beta;
      alpha = rho * alpha + (1.0 - rho) * rep_vector(1.0, 3);
      alpha[op_action[t]] += 1.0;
    }
  }
}
model {
  log_sigma ~ normal(-1, 1);
  log_beta  ~ normal(-1, 1);
  for (t in 1:N)
    action[t] ~ categorical_logit(to_vector(EU[t, ]));
}
generated quantities {
  vector[N] log_lik;
  array[N] int action_rep;
  real lprior = normal_lpdf(log_sigma | -1, 1)
              + normal_lpdf(log_beta  | -1, 1);
  for (t in 1:N) {
    log_lik[t]    = categorical_logit_lpmf(action[t] | to_vector(EU[t, ]));
    action_rep[t] = categorical_logit_rng(to_vector(EU[t, ]));
  }
}
"
writeLines(stan_tom0_rps, here::here("stan", "ch16_tom0_rps_single.stan"))
mod_tom0_rps <- cmdstan_model(here::here("stan", "ch16_tom0_rps_single.stan"))
```

Two remarks on the Stan code. First, the `EU` matrix is computed in
`transformed parameters` so that the gradient flows through `log_sigma`
and `log_beta` when HMC computes its leapfrog steps. Second,
`categorical_logit` applies a softmax internally, so we pass the raw
(unnormalized) EU values; the division by $\beta$ is baked in before
the Stan model block.

**Prior choice.** `log_sigma ~ normal(-1, 1)` centres the forgetting
rate at $\sigma \approx 0.37$ (moderate memory). `log_beta ~ normal(-1, 1)`
centres at $\beta \approx 0.37$ (moderately deterministic softmax). Both
are wide enough to accommodate near-perfect memory or near-random behavior.

### Phase 1 — Prior predictive check

```{r ch16_tom0_ppc1, fig.cap="0-ToM prior predictive check. 9 simulated agents against a uniform random opponent (200 trials each), with parameters drawn from the priors. The agents produce heterogeneous choice distributions and varying levels of cycling — none degenerates to a constant action."}
set.seed(2026)
n_pp <- 9
op_seq_pp <- sample(1L:3L, 200, replace = TRUE)

prior_tom0 <- tibble(
  draw      = 1:n_pp,
  log_sigma = rnorm(n_pp, -1, 1),
  log_beta  = rnorm(n_pp, -1, 1)
) |>
  rowwise() |>
  mutate(
    dat = list({
      sim <- simulate_tom0_rps(op_seq_pp, log_sigma, log_beta, seed = draw)
      sim$id <- draw
      prep_wsls_data(sim) |> mutate(draw = draw)
    })
  )

# Show marginal choice distribution per prior draw
prior_tom0 |>
  unnest(dat) |>
  count(draw, action) |>
  group_by(draw) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = factor(action, 1:3, c("R","P","S")),
             y = prop, fill = factor(action))) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 1/3, linetype = "dashed") +
  scale_fill_manual(values = c("steelblue","seagreen3","firebrick3"),
                    guide = "none") +
  facet_wrap(~draw, labeller = label_both) +
  labs(x = "Action", y = "Proportion",
       title = "0-ToM prior predictive: marginal action distributions")
```

### Phase 2 — Parameter recovery

```{r ch16_tom0_recovery}
tom0_rec_path <- here::here("simmodels", "ch16_tom0_recovery.rds")

if (regenerate_simulations || !file.exists(tom0_rec_path)) {
  set.seed(303)
  n_agents <- 20
  truth_tom0 <- tibble(
    agent     = 1:n_agents,
    log_sigma = rnorm(n_agents, -1, 1),
    log_beta  = rnorm(n_agents, -1, 1)
  )
  op_vec_rec <- sample(1L:3L, 200, replace = TRUE)

  fit_tom0_one <- function(ls, lb, ag) {
    sim <- simulate_tom0_rps(op_vec_rec, ls, lb, seed = ag)
    fit <- mod_tom0_rps$sample(
      data = list(N = nrow(sim), action = sim$action,
                  op_action = sim$opponent_action),
      chains = 2, parallel_chains = 2,
      iter_warmup = 500, iter_sampling = 500,
      refresh = 0, show_messages = FALSE
    )
    fit$summary(c("log_sigma","log_beta")) |>
      dplyr::select(variable, mean, q5, q95)
  }

  rec_tom0 <- truth_tom0 |>
    rowwise() |>
    mutate(post = list(fit_tom0_one(log_sigma, log_beta, agent))) |>
    dplyr::select(agent, log_sigma, log_beta, post) |>
    unnest(post) |>
    pivot_longer(c(log_sigma, log_beta),
                 names_to = "param", values_to = "true_val") |>
    filter(variable == param)
  saveRDS(rec_tom0, tom0_rec_path)
} else {
  rec_tom0 <- readRDS(tom0_rec_path)
}
```

```{r ch16_tom0_recovery_plot, fig.cap="0-ToM parameter recovery. log_beta recovers cleanly; log_sigma shows mild shrinkage toward the prior at 200 trials, consistent with the known difficulty of estimating forgetting rates from short sequences."}
ggplot(rec_tom0, aes(x = true_val, y = mean)) +
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0, alpha = 0.4,
                color = "steelblue") +
  geom_point(color = "midnightblue", size = 2) +
  geom_abline(linetype = "dashed", color = "gray40") +
  facet_wrap(~param, scales = "free") +
  labs(x = "True value", y = "Posterior mean",
       title = "0-ToM parameter recovery (20 agents, 200 trials)")
```

### Phase 3 — Simulation-Based Calibration

```{r ch16_tom0_sbc}
sbc_tom0_path <- here::here("simmodels", "ch16_tom0_sbc.rds")

gen_tom0_sbc <- function(N = 200) {
  ls  <- rnorm(1, -1, 1)
  lb  <- rnorm(1, -1, 1)
  op  <- sample(1L:3L, N, replace = TRUE)
  sim <- simulate_tom0_rps(op, ls, lb)
  list(
    variables = list(log_sigma = ls, log_beta = lb),
    generated = list(N = N, action = sim$action, op_action = sim$opponent_action)
  )
}

if (regenerate_sbc || !file.exists(sbc_tom0_path)) {
  sbc_gen_t  <- SBC_generator_function(gen_tom0_sbc, N = 200)
  sbc_back_t <- SBC_backend_cmdstan_sample(
    mod_tom0_rps, iter_warmup = 500, iter_sampling = 500,
    chains = 1, refresh = 0
  )
  sbc_ds_t  <- generate_datasets(sbc_gen_t, 200)
  sbc_res_t <- compute_SBC(sbc_ds_t, sbc_back_t, keep_fits = FALSE)
  saveRDS(list(ds = sbc_ds_t, results = sbc_res_t), sbc_tom0_path)
} else {
  obj       <- readRDS(sbc_tom0_path)
  sbc_ds_t  <- obj$ds
  sbc_res_t <- obj$results
}
```

```{r ch16_tom0_sbc_plot, fig.width=8, fig.height=4, fig.cap="SBC for 0-ToM. log_beta is well-calibrated. log_sigma shows mild excursions — a design diagnostic, not a sampler failure: a random opponent provides insufficient contrast to pin down the forgetting rate with high precision at 200 trials. The implication is that log_sigma posteriors should be reported with explicit uncertainty."}
plot_rank_hist(sbc_res_t) / plot_ecdf_diff(sbc_res_t)
```

### Phase 4 — Posterior predictive check

```{r ch16_tom0_ppc_fit}
tom0_ppc_path <- here::here("simmodels", "ch16_tom0_ppc.rds")

if (regenerate_fits || !file.exists(tom0_ppc_path)) {
  fit_tom0_demo <- mod_tom0_rps$sample(
    data = list(N = nrow(demo_df), action = demo_df$action,
                op_action = demo_df$opponent_action),
    chains = 2, parallel_chains = 2,
    iter_warmup = 500, iter_sampling = 500,
    refresh = 0, show_messages = FALSE
  )
  param_draws_t <- fit_tom0_demo$draws(c("log_sigma","log_beta"),
                                        format = "matrix")
  ppc_tom0 <- map_dfr(1:100, function(s) {
    sim    <- simulate_tom0_rps(
      demo_df$opponent_action,
      param_draws_t[s, "log_sigma"],
      param_draws_t[s, "log_beta"],
      seed = s
    )
    sim$id <- 1L
    prep_wsls_data(sim) |> mutate(rep = s)
  })
  saveRDS(ppc_tom0, tom0_ppc_path)
} else {
  ppc_tom0 <- readRDS(tom0_ppc_path)
}
```

```{r ch16_tom0_ppc_plot, fig.cap="0-ToM posterior predictive check. The model captures the qualitative Win-Stay pattern and partial Lose-Shift, but its predictions arise from an entirely different mechanism than WSLS: expected-utility maximization against a learned opponent distribution rather than outcome conditioning."}
ppc_tom0_sum <- ppc_tom0 |>
  mutate(
    outcome_lbl = factor(outcome, 1:3, c("Win","Lose","Draw")),
    shift_lbl   = factor(rel_shift, 1:3, c("Stay","CW","CCW"))
  ) |>
  count(rep, outcome_lbl, shift_lbl) |>
  group_by(rep, outcome_lbl) |>
  mutate(prop = n / sum(n)) |>
  group_by(outcome_lbl, shift_lbl) |>
  summarize(m = mean(prop), lo = quantile(prop, 0.025),
            hi = quantile(prop, 0.975), .groups = "drop")

ggplot(ppc_tom0_sum, aes(x = shift_lbl)) +
  geom_col(aes(y = m), fill = "gray70", alpha = 0.8) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.2, color = "gray40") +
  geom_point(data = obs_wsls_demo, aes(y = prop_obs),
             color = "firebrick3", size = 3) +
  geom_hline(yintercept = 1/3, linetype = "dashed") +
  facet_wrap(~outcome_lbl) +
  labs(x = "Shift", y = "Proportion",
       title = "0-ToM posterior predictive vs. observed")
```

### Extending to 1-ToM: the recursive structure

The 1-ToM agent in RPS believes the opponent is a 0-ToM tracker and
therefore predicts the opponent's action by simulating what a 0-ToM
would predict about the 1-ToM itself. The key new state is
$\alpha^{\text{self}}_t$ — the 1-ToM's model of what the opponent
believes about the 1-ToM's own action distribution:

$$
\alpha^{\text{self}}_t \leftarrow \rho_{\text{op}}\,\alpha^{\text{self}}_{t-1}
+ (1-\rho_{\text{op}})\,\mathbf{1}, \qquad
\alpha^{\text{self}}_{t,c^{\text{self}}_{t-1}} \mathrel{+}= 1.
$$

The opponent's predicted action distribution is then the 0-ToM response
to $\hat{p}^{\text{self}}_t = \alpha^{\text{self}}_t / \sum
\alpha^{\text{self}}_t$:

$$
\hat{p}^{\text{op}}_t \propto \exp\!\left(
  \tfrac{1}{\beta_{\text{op}}}\,(-\Pi^\top\,\hat{p}^{\text{self}}_t)
\right),
$$

where $-\Pi^\top$ is the opponent's payoff matrix (negative transpose of
the focal agent's payoff). The 1-ToM then selects the action that maximises
expected utility against $\hat{p}^{\text{op}}_t$ using its own temperature
$\beta$. The R simulator below implements this recursion:

```{r ch16_tom1_sim}
simulate_tom1_rps <- function(op_choices, log_sigma, log_sigma_op,
                               log_beta, log_beta_op, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  rho_op <- exp(-exp(log_sigma_op))
  beta_op <- exp(log_beta_op)
  beta    <- exp(log_beta)
  N       <- length(op_choices)
  alpha_self <- rep(1, 3)   # 1-ToM's model of opponent's belief about self
  choice <- integer(N)
  for (t in 1:N) {
    p_self_est <- alpha_self / sum(alpha_self)
    # Opponent (0-ToM) best-responds to p_self_est
    EU_opp   <- -t(RPS_PAYOFF) %*% p_self_est   # opp's payoff = neg of focal's
    p_op_est <- softmax3(as.vector(EU_opp) / beta_op)
    # 1-ToM best-responds to p_op_est
    EU_focal  <- RPS_PAYOFF %*% p_op_est
    p_choice  <- softmax3(as.vector(EU_focal) / beta)
    choice[t] <- sample(1L:3L, 1L, prob = p_choice)
    # Update: opponent observes focal's choice
    alpha_self <- rho_op * alpha_self + (1 - rho_op) * rep(1, 3)
    alpha_self[choice[t]] <- alpha_self[choice[t]] + 1
  }
  tibble(t = seq_len(N), action = choice, opponent_action = op_choices)
}
```

The Stan implementation of 1-ToM mirrors this structure exactly —
the `alpha_self` recursion is driven by `action[t]` (data), which
keeps the transformed-parameters block valid for HMC. It introduces
two additional parameters (`log_sigma_op`, `log_beta_op`) for the
opponent's believed volatility and temperature. As the SBC for the
binary 1-ToM in Ch. 9 showed, these opponent-model parameters are
weakly identified from short sequences; the same caveat applies here.
A complete Stan implementation follows the template of
`ch16_tom0_rps_single.stan`, substituting `alpha_self` updated with
`action[t]` for `alpha` updated with `op_action[t]`. We leave the
full implementation as an exercise and focus the model comparison on
Nash, WSLS, and 0-ToM.

---

## Hierarchical WSLS

Individual WSLS fits tell us about each player in isolation. To ask
whether Win-Stay is reliably present *at the population level* —
and to obtain stable estimates for players with few trials — we need
a hierarchical model that pools information across individuals.

### Parameterisation

The three simplices $\theta_{\text{win}}, \theta_{\text{lose}},
\theta_{\text{draw}}$ are parameterised through logit-stick-breaking:
for each outcome $o$ we introduce two free log-odds parameters

$$
\lambda_{o,1} = \log\frac{\theta_{o,\text{Stay}}}{\theta_{o,\text{CCW}}},
\qquad
\lambda_{o,2} = \log\frac{\theta_{o,\text{CW}}}{\theta_{o,\text{CCW}}},
$$

so $\theta_o = \text{softmax}([\lambda_{o,1},\,\lambda_{o,2},\,0])$.
Individual-level parameters are drawn from a population distribution:
$\lambda_{i,o,k} \sim \mathcal{N}(\mu_{o,k},\,\tau_{o,k})$,
implemented non-centered as $\lambda_{i,o,k} = \mu_{o,k} +
\tau_{o,k}\,z_{i,o,k}$ with $z_{i,o,k} \sim \mathcal{N}(0,1)$.

```{r ch16_wsls_hier_stan}
stan_wsls_hier <- "
data {
  int<lower=1> N;         // total number of transitions
  int<lower=1> I;         // number of individuals
  array[N] int<lower=1, upper=I>  id;
  array[N] int<lower=1, upper=3>  rel_shift;
  array[N] int<lower=1, upper=3>  outcome;
}
parameters {
  // Population-level log-odds for [Stay, CW] vs CCW, by outcome
  array[3] vector[2] mu_pop;
  array[3] vector<lower=0>[2] tau_pop;
  // Non-centered individual deviations: z[i, outcome, logit_index]
  array[I, 3] vector[2] z_ind;
}
transformed parameters {
  // Individual-level simplex entries via softmax of logit-stick
  array[I, 3] simplex[3] theta_ind;
  for (i in 1:I) {
    for (o in 1:3) {
      vector[2] lam = mu_pop[o] + tau_pop[o] .* z_ind[i, o];
      theta_ind[i, o] = softmax(append_row(lam, 0.0));
    }
  }
}
model {
  for (o in 1:3) {
    mu_pop[o]  ~ normal(0, 1);
    tau_pop[o] ~ exponential(2);
  }
  for (i in 1:I) {
    for (o in 1:3) {
      z_ind[i, o] ~ std_normal();
    }
  }
  for (t in 1:N)
    rel_shift[t] ~ categorical(theta_ind[id[t], outcome[t]]);
}
generated quantities {
  vector[N] log_lik;
  for (t in 1:N)
    log_lik[t] = categorical_lpmf(rel_shift[t] |
                                   theta_ind[id[t], outcome[t]]);
}
"
writeLines(stan_wsls_hier, here::here("stan", "ch16_wsls_hier.stan"))
mod_wsls_hier <- cmdstan_model(here::here("stan", "ch16_wsls_hier.stan"))
```

### Fitting to population data

```{r ch16_wsls_hier_fit}
hier_wsls_path <- here::here("simmodels", "ch16_wsls_hier_fit.rds")

# Prepare population-level data
rps_wsls_hier <- rps_wsls |>
  mutate(id_int = as.integer(factor(id)))

stan_hier_data <- list(
  N         = nrow(rps_wsls_hier),
  I         = max(rps_wsls_hier$id_int),
  id        = rps_wsls_hier$id_int,
  rel_shift = rps_wsls_hier$rel_shift,
  outcome   = rps_wsls_hier$outcome
)

if (regenerate_fits || !file.exists(hier_wsls_path)) {
  fit_hier_wsls <- mod_wsls_hier$sample(
    data            = stan_hier_data,
    chains          = 4,
    parallel_chains = 4,
    iter_warmup     = 1000,
    iter_sampling   = 1000,
    adapt_delta     = 0.95,
    seed            = 2026,
    refresh         = 200
  )
  fit_hier_wsls$save_object(hier_wsls_path)
} else {
  fit_hier_wsls <- readRDS(hier_wsls_path)
}

fit_hier_wsls$diagnostic_summary()
```

### Population-level estimates

```{r ch16_wsls_hier_summary, fig.cap="Population-level posterior distributions for the six key parameters: Win-Stay (lambda[1,1]), Win-CW (lambda[1,2]), Lose-Stay (lambda[2,1]), Lose-CW (lambda[2,2]), Draw-Stay (lambda[3,1]), Draw-CW (lambda[3,2]). Values > 0 indicate the category is preferred over CCW. Win-Stay and Lose-CW show the strongest positive signals, confirming social cycling at the population level."}
mu_draws <- fit_hier_wsls$draws(
  c("mu_pop[1,1]","mu_pop[1,2]",
    "mu_pop[2,1]","mu_pop[2,2]",
    "mu_pop[3,1]","mu_pop[3,2]"),
  format = "draws_df"
) |>
  pivot_longer(-c(.chain,.iteration,.draw),
               names_to = "param", values_to = "value") |>
  mutate(
    outcome_lbl = case_when(
      grepl("^mu_pop\\[1", param) ~ "Win",
      grepl("^mu_pop\\[2", param) ~ "Lose",
      TRUE                         ~ "Draw"
    ),
    shift_lbl = if_else(grepl(",1\\]$", param), "Stay (vs CCW)", "CW (vs CCW)")
  )

ggplot(mu_draws, aes(x = value, fill = shift_lbl)) +
  geom_density(alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = c("steelblue","seagreen3")) +
  facet_wrap(~outcome_lbl, ncol = 1) +
  labs(x = "Log-odds vs CCW shift", y = "Density", fill = NULL,
       title = "Population-level WSLS posteriors",
       subtitle = "Positive = preferred over CCW; 0 = indifferent")
```

```{r ch16_wsls_hier_individuals, fig.cap="Individual-level Win-Stay (x-axis) vs. Lose-Shift-CW (y-axis) probabilities. Players in the upper-right quadrant are the strongest WSLS adherents. The positive correlation is typically weak — Win-Stay and Lose-Shift are partially independent traits."}
# Extract posterior means for Win-Stay [outcome=1, shift=1] and Lose-CW [outcome=2, shift=2]
ind_sum <- fit_hier_wsls$draws("theta_ind", format = "draws_df") |>
  pivot_longer(-c(.chain, .iteration, .draw),
               names_to = "param", values_to = "value") |>
  filter(grepl(",1,1\\]$|,2,2\\]$", param)) |>
  mutate(
    id_int   = as.integer(sub("theta_ind\\[(\\d+),.*", "\\1", param)),
    param_lbl = if_else(grepl(",1,1\\]$", param), "win_stay", "lose_cw")
  ) |>
  group_by(id_int, param_lbl) |>
  summarize(mean_val = mean(value), .groups = "drop") |>
  pivot_wider(names_from = param_lbl, values_from = mean_val)

ggplot(ind_sum, aes(x = win_stay, y = lose_cw)) +
  geom_point(alpha = 0.7, color = "midnightblue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick3", linewidth = 0.8,
              formula = y ~ x) +
  geom_hline(yintercept = 1/3, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 1/3, linetype = "dashed", color = "gray50") +
  labs(x = "Win-Stay probability (posterior mean)",
       y = "Lose-Shift-CW probability (posterior mean)",
       title = "Individual WSLS parameters",
       subtitle = "Dashed lines = Nash baseline (1/3)")
```

---

## Model Comparison

We now ask the central question: across all players, which model — Nash,
WSLS, or 0-ToM — provides the best out-of-sample prediction of human
choices in RPS?

We fit all three single-agent models to every player and compare
PSIS-LOO expected log predictive densities (ELPD). Because the Nash
model uses absolute actions while WSLS uses relative shifts (losing one
trial per player), we standardise by dividing the summed ELPD by the
number of observations contributed by each player.

```{r ch16_model_comparison}
comp_path <- here::here("simmodels", "ch16_model_comparison.rds")

fit_all_models <- function(player_df) {
  # Nash
  fit_n <- mod_nash$sample(
    data = list(N = nrow(player_df), action = player_df$action),
    chains = 2, parallel_chains = 2,
    iter_warmup = 500, iter_sampling = 500,
    refresh = 0, show_messages = FALSE
  )
  # WSLS
  dd_w <- prep_wsls_data(player_df)
  fit_w <- mod_wsls$sample(
    data = list(N = nrow(dd_w), rel_shift = dd_w$rel_shift,
                outcome = dd_w$outcome),
    chains = 2, parallel_chains = 2,
    iter_warmup = 500, iter_sampling = 500,
    refresh = 0, show_messages = FALSE
  )
  # 0-ToM
  fit_t <- mod_tom0_rps$sample(
    data = list(N = nrow(player_df), action = player_df$action,
                op_action = player_df$opponent_action),
    chains = 2, parallel_chains = 2,
    iter_warmup = 500, iter_sampling = 500,
    refresh = 0, show_messages = FALSE
  )
  tibble(
    model = c("Nash", "WSLS", "0-ToM"),
    elpd  = c(fit_n$loo()$estimates["elpd_loo","Estimate"],
              fit_w$loo()$estimates["elpd_loo","Estimate"],
              fit_t$loo()$estimates["elpd_loo","Estimate"]),
    n_obs = c(nrow(player_df), nrow(dd_w), nrow(player_df))
  ) |>
    mutate(elpd_per_obs = elpd / n_obs)
}

# Use a subset of players for the model comparison (computational budget)
sample_ids <- rps |> pull(id) |> unique() |> head(30)

if (regenerate_fits || !file.exists(comp_path)) {
  comp_res <- map_dfr(sample_ids, function(pid) {
    df <- filter(rps, id == pid)
    fit_all_models(df) |> mutate(id = pid)
  })
  saveRDS(comp_res, comp_path)
} else {
  comp_res <- readRDS(comp_path)
}
```

```{r ch16_comparison_plot, fig.cap="Per-player ELPD (relative to Nash) for WSLS and 0-ToM. Each point is one player; positive values indicate the model outperforms Nash on that player's data. WSLS improves on Nash for most players; 0-ToM shows similar or slightly smaller gains, with greater variability."}
comp_wide <- comp_res |>
  dplyr::select(id, model, elpd_per_obs) |>
  pivot_wider(names_from = model, values_from = elpd_per_obs) |>
  mutate(
    delta_wsls = WSLS - Nash,
    delta_tom0 = `0-ToM` - Nash
  )

comp_wide |>
  pivot_longer(c(delta_wsls, delta_tom0),
               names_to = "comparison", values_to = "delta") |>
  mutate(comparison = recode(comparison,
                             delta_wsls = "WSLS vs Nash",
                             delta_tom0 = "0-ToM vs Nash")) |>
  ggplot(aes(x = comparison, y = delta)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_boxplot(fill = "gray90", outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.7, color = "midnightblue") +
  labs(x = NULL,
       y = expression(Delta * " ELPD per observation vs Nash"),
       title = "Model comparison: WSLS and 0-ToM vs Nash baseline",
       subtitle = "Positive = better out-of-sample prediction than Nash")
```

```{r ch16_comparison_wsls_vs_tom, fig.cap="Head-to-head comparison: WSLS vs 0-ToM per player. Points on the diagonal mean the two models tie. Points above = 0-ToM wins; below = WSLS wins. The distribution is roughly symmetric around the diagonal, indicating neither model consistently dominates at the individual level."}
ggplot(comp_wide, aes(x = WSLS, y = `0-ToM`)) +
  geom_abline(linetype = "dashed", color = "gray40") +
  geom_point(alpha = 0.7, color = "midnightblue", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick3", linewidth = 0.7) +
  labs(x = "WSLS ELPD per obs",
       y = "0-ToM ELPD per obs",
       title = "Head-to-head: WSLS vs 0-ToM")
```

```{r ch16_comparison_summary}
comp_res |>
  group_by(model) |>
  summarize(
    mean_elpd_per_obs = mean(elpd_per_obs),
    sd_elpd_per_obs   = sd(elpd_per_obs),
    .groups = "drop"
  ) |>
  arrange(desc(mean_elpd_per_obs)) |>
  knitr::kable(digits = 3,
               caption = "Mean per-observation ELPD by model (higher = better)")
```

The model comparison tells a story in three parts. First, *both* WSLS and
0-ToM improve on Nash: ignoring sequential structure leaves substantial
predictive power on the table, confirming that social cycling is a real
regularisation in the data, not noise. Second, WSLS and 0-ToM produce
similar aggregate ELPD, implying that the *heuristic* explanation
(outcome-conditional transition) and the *generative* explanation
(opponent-distribution tracking) fit the observable choice sequences
about equally well at the population level. Third, there is substantial
individual variation: some players are better described by WSLS, some by
0-ToM. This heterogeneity motivates future work using mixture models
(extending Ch. 8) or individual-level covariates (extending the
hierarchical architecture of Ch. 6) to identify which players are
heuristic responders versus active predictors.

---

## Conclusion

This chapter extended the Bayesian cognitive modeling toolkit from binary
to three-choice environments. The key moves were:

1. **Relative-shift encoding** for WSLS: parameterising transitions as
   Stay / CW / CCW rather than as absolute actions makes the model
   invariant to which specific action was played last round, halving the
   number of effective parameters while preserving all the information
   about social cycling.

2. **Dirichlet-forgetting 0-ToM**: generalising the binary Kalman filter
   from Ch. 9 to a multinomial belief state with exponential decay. The
   structure is identical — a belief update, a forgetting term, an
   expected-utility computation, and a softmax decision — but the domain
   is now a 3-simplex rather than a log-odds scalar.

3. **Outcome-conditioned validation**: posterior predictive checks that
   focus on the *transition pattern* (social cycling) rather than marginal
   action frequencies are necessary to distinguish models that differ in
   their sequential structure.

4. **Hierarchical pooling**: the individual variation in Win-Stay and
   Lose-Shift rates justifies a population model. The non-centered
   logit-normal hierarchy stabilises estimation for players with fewer
   trials and supports population-level inference.

The chapter stops at 0-ToM. True 1-ToM in RPS requires the same recursive
belief-state update as the binary case (the focal agent models what the
opponent believes about the focal agent's own distribution), and the R
simulator in Section 3 makes the structure explicit. The identification
challenges documented for 1-ToM in Ch. 9 — weakly constrained opponent-model
parameters with short sequences — carry over here and are amplified by the
richer action space. The comparative design lesson is the same: use the
simplest model that fits the data, validate with the full battery, and let
PSIS-LOO decide whether the recursive complexity earns its keep.
