---
title: "Chapter 6: Hierarchical Modeling and Individual Differences in Cognitive Strategies"
output: html_document
date: "2025-03-01" # Updated date
---

```{r ch6_setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  fig.align = 'center',
  out.width = "80%",
  dpi = 300
  )

regenerate_simulations <- FALSE 

# Construct the required directory infrastructure
if (!dir.exists("stan")) dir.create("stan")
if (!dir.exists("simdata")) dir.create("simdata")
if (!dir.exists("simmodels")) dir.create("simmodels")
if (!dir.exists("figures")) dir.create("figures") 

# Load the strict computational stack using pacman
pacman::p_load(
  tidyverse,       # Core suite for data manipulation and visualization
  here,            # Robust path routing relative to project root
  posterior,       # Analytical tools for posterior array manipulations
  cmdstanr,        # Direct R interface to the CmdStan C++ compiler
  tidybayes,       # Tidy extraction of draws and predictive distributions
  patchwork,       # Composition of diagnostic ggplot matrices
  bayesplot,       # Specialized geometric diagnostic plots (traces, pairs)
  furrr,           # Parallel mapping for Recovery and SBC arrays
  future,          # Asynchronous backend for parallelization
  loo,             # PSIS-LOO out-of-sample predictive cross-validation
  priorsense,      # Analytical prior sensitivity via power-scaling
  DiagrammeR,      # Generative structural visualizations (DAGs)
  SBC,             # Simulation-Based Calibration
  marginaleffects # Conditional predictive interpretations on the outcome scale
)

# Enforce a clean, data-to-ink maximized visual theme
theme_set(theme_classic())

# Establish the parallel processing topology
# We reserve 2 cores to prevent the operating system from hanging during intensive MCMC/SBC
#cores_to_use <- max(1, parallel::detectCores() - 2)
#plan(multisession, workers = cores_to_use)
plan(sequential) # to ensure proper knitting.
```

# Individual Differences in Cognitive Strategies: The Hierarchical Architecture

> **📍 Where we are in the Bayesian modeling workflow:**
> Chs. 1–5 fitted models to a single agent and validated the inference
> engine. Now we scale up: real experiments involve many participants who
> differ from each other. This chapter builds hierarchical (multilevel)
> models that estimate both individual parameters *and* the population
> distribution they are drawn from — and validates the whole pipeline
> with the same six-phase battery from Ch. 5.
> The memory model fitted here uses the **reversal learning dataset from
> Ch. 5**, which was the first dataset that cleanly separated `bias` from
> `beta`. The fitted model and its LOO score are saved for Ch. 7.

## Introduction

Our exploration of decision-making models has so far focused on single agents or averaged behavior. However, individuals differ systematically in how they approach tasks and process information. Some people may be more risk-averse, have better memory, learn faster, or employ entirely different strategies than others. Ignoring these individual differences can lead to misleading conclusions about cognitive processes. From a statistical perspective, if we assume a single set of parameters generates all data, our model is mathematically misspecified, which leads to pathological posterior geometries and misleading inferential results.

This chapter introduces hierarchical modeling (also called multilevel or mixed effects modeling) as a powerful framework for capturing these individual differences while still identifying population-level patterns.

### The Epistemic Challenge: Pooling Information

Traditional statistical approaches to handling data from multiple individuals force a false dichotomy, imposing rigid, unjustified structural assumptions on the data:

1. **Complete Pooling (Underfitting)**:  Treats all participants as identical: it estimates a single set of parameters for the entire group, completely ignoring potentially meaningful individual variation. This assumes the population variance (the average error in predicting an individual from the population) is strictly zero ($\sigma \to 0$). 
   * *The Pathology*: Systematically fails to capture cognitive heterogeneity, describing an "average" participant that does not necessarily exist.

2. **No Pooling (Overfitting)**: Analyzes each participant in complete isolation, estimating independent parameters ($\theta_j$) for each individual. No pooling corresponds to fitting $J$ separate models to $J$ individual participants. This assumes the population variance is infinite ($\sigma \to \infty$). 
   * *The Pathology*: Severely inefficient and vulnerable to noise. For individuals with sparse or noisy data, the posterior estimate will be excessively wide or completely dominated by the random fluctuations of their specific likelihood.
   
> *Why does this dichotomy matter in practice?* With 120 trials per agent,
> complete pooling produces one parameter estimate that represents nobody
> in particular. No pooling produces 50 estimates, many of which are
> dominated by noise for agents with only 20–30 trials. Partial pooling
> is not a philosophical compromise — it produces estimates that are
> demonstrably *closer to the truth* for low-data individuals, as the
> shrinkage plot at the end of this chapter will show directly.

### The Solution: Partial Pooling (Multilevel or Hierarchical Modeling)

Hierarchical modeling offers a principled, data-driven middle ground. Instead of forcing $\sigma$ to be $0$ or $\infty$, we treat the population variance as a free parameter estimated from the data. The model is structured hierarchically:

* **Individual Level**: Each individual agent $j$ has a specific parameter vector (e.g., their intrinsic bias $\theta_j$ or memory decay $\beta_j$).

* **Population Level**: These individual parameters are explicitly modeled as draws from an overarching continuous population distribution (e.g., a Gaussian characterized by a population mean $\mu$ and standard deviation $\sigma$).

* **The Bayesian Compromise**: The estimation of an individual's parameter ($\theta_j$) is a precision-weighted combination of their own empirical data (the likelihood) and the population distribution (the hierarchical prior).
* **Shrinkage**: Estimates for individuals with sparse or noisy data are "shrunk" toward the population mean $\mu$. This is a mathematical mechanism that regularizes inference, preventing overfitting and yielding more robust out-of-sample estimates.

## Learning Objectives

After completing this chapter, you will be able to:

* **Explain Partial Pooling**: Understand how multilevel modeling balances individual and group-level information through shrinkage.

* **Distinguish Pooling Approaches**: Articulate the differences and trade-offs between complete pooling ($\sigma=0$), no pooling ($\sigma=\infty$), and partial pooling.

* **Implement Hierarchical Models**: Write Stan code for hierarchical models with varying intercepts and/or slopes.

* **Reparameterize a model**: Understand why and how to reparameterize a model and in particular how to use non-centered parameterization to improve sampling efficiency in hierarchical models.

* **Model Parameter Correlations**: Implement and interpret models that estimate correlations between individual-level parameters (via  Cholesky-factorized correlation matrices $\mathbf{L}_\Omega$).

* **Perform Comprehensive Model Checks**: Apply a rigorous Bayesian workflow to hierarchical models (Predictive Checks, Prior-Posterior Update Checks, Sensitivity Analysis, Parameter Recovery, Simulation-Based Calibration, etc).

* **Apply to Cognitive Questions**: Use multilevel modeling to investigate individual differences in cognitive strategies (e.g., bias, memory, WSLS).

This chapter has two parts. The first applies the full Bayesian workflow to the **Hierarchical Biased Agent** — one parameter per person, one population distribution — and uses it to expose and fix the Centered Parameterization's pathological geometry (Neal's Funnel). The second applies the same workflow to the **Hierarchical Memory Agent**, which has two correlated parameters per person, requires a bivariate population distribution, and uses the reversal
learning dataset from Ch. 5 that we validated there. The LOO score from the fitted memory model is saved for Ch. 7's model comparison.

### Biased Agent Model (Multilevel)

In this architecture, each agent $j \in \{1, \dots, J\}$ possesses an unobserved, individual-level bias parameter $\theta_j$. This individual parameter is generated by a latent population distribution, parameterized by a global mean $\mu_\theta$ and global standard deviation $\sigma_\theta$. Finally, the empirical choices $h_{ij}$ for trial $i$ are generated conditionally on the individual parameter $\theta_j$.

Mathematically, this translates to the following generative structure on the unconstrained (logit) space:$$h_{ij} \sim \text{Bernoulli}(\text{logit}^{-1}(\theta_{\text{logit}, j}))$$$$\theta_{\text{logit}, j} \sim \text{Normal}(\mu_{\theta}, \sigma_{\theta})$$$$\mu_{\theta} \sim \text{Normal}(0, 1.5)$$$$\sigma_{\theta} \sim \text{Exponential}(1)$$

By mapping this DAG, we have strictly defined our generative assumptions. We are now ready to translate this architecture into Stan code and subject it to Prior Predictive Checks.


```{tikz hierarchical_biased_dags, echo=FALSE, fig.cap="Standard Plate Notation (Left) and Kruschke-Style Generative Notation (Right) for the Hierarchical Biased Agent.", fig.ext='png', fig.width=12, cache=TRUE}
\usetikzlibrary{positioning, shapes.geometric, arrows.meta, fit, backgrounds, calc}

\tikzset{
    stochastic/.style={->, >={Latex[length=3mm]}, thick},
    deterministic/.style={->, >={Latex[length=3mm]}, thick, dashed, draw=gray!70},
    plate/.style={draw=black, thick, rounded corners, inner sep=10pt},
    continuous/.style={circle, draw=black, thick, minimum size=1.2cm, inner sep=0pt},
    discrete/.style={rectangle, draw=black, thick, minimum size=1.2cm},
    observed/.style={fill=gray!30},
    latent/.style={fill=white},
    pics/norm_dist/.style={code={
        \draw[thin, gray] (-1.2,0) -- (1.2,0);
        \draw[thick, blue] plot[domain=-1.2:1.2, samples=50] (\x, {0.8*exp(-3*\x*\x)});
        \node[font=\footnotesize, below] at (0,0) {Normal};
    }},
    pics/exp_dist/.style={code={
        \draw[thin, gray] (-0.2,0) -- (1.5,0);
        \draw[thick, blue] plot[domain=0:1.4, samples=50] (\x, {0.8*exp(-2.5*\x)});
        \node[font=\footnotesize, below] at (0.6,0) {Exponential};
    }},
    pics/lkj_dist/.style={code={
        \draw[thin, gray] (-1.2,0) -- (1.2,0);
        \draw[thick, blue] plot[domain=-0.95:0.95, samples=50] (\x, {0.8*(1-\x*\x)});
        \node[font=\footnotesize, below] at (0,0) {LKJ};
    }},
    pics/mvn_dist/.style={code={
        \draw[thin, gray, ->] (-0.8,0) -- (0.8,0);
        \draw[thin, gray, ->] (0,-0.8) -- (0,0.8);
        \draw[thick, blue, rotate=30] (0,0) ellipse (0.6 and 0.22);
        \draw[thick, blue, rotate=30] (0,0) ellipse (0.3 and 0.11);
        \node[font=\footnotesize, below right] at (0.1,-0.15) {MVN};
    }},
    pics/bern_dist/.style={code={
        \draw[thin, gray] (-1,0) -- (1,0);
        \draw[blue, line width=4pt] (-0.4, 0) -- (-0.4, 0.3);
        \draw[blue, line width=4pt] (0.4, 0) -- (0.4, 0.8);
        \node[font=\footnotesize, below] at (0,0) {Bernoulli};
    }}
}

\begin{tikzpicture}
%% LEFT: STANDARD PLATE NOTATION
\node[font=\large\bfseries] at (0, 2.2) {Standard Plate Notation};
\node[continuous, latent] (mu_a)  at (-3,   0) {$\mu_\alpha$};
\node[continuous, latent] (sig_a) at (-1.5, 0) {$\sigma_\alpha$};
\node[continuous, latent, fill=gray!15] (omega) at (0, 0) {$\Omega$};
\node[continuous, latent] (sig_b) at (1.5,  0) {$\sigma_\beta$};
\node[continuous, latent] (mu_b)  at (3,    0) {$\mu_\beta$};
\node[continuous, latent] (alpha) at (-1.5, -3.5) {$\alpha_j$};
\node[continuous, latent] (beta)  at ( 1.5, -3.5) {$\beta_j$};
\node[discrete, observed] (m) at (-1.5, -7) {$m_{ij}$};
\node[discrete, observed] (h) at ( 1.5, -7) {$h_{ij}$};
\node[plate, fit=(m)(h), label={[anchor=south east, font=\small\bfseries]below right:$i=1\dots N$}] (pN) {};
\node[plate, inner sep=15pt, fit=(alpha)(beta)(pN), label={[anchor=south east, font=\small\bfseries]below right:$j=1\dots J$}] {};
\draw[stochastic] (mu_a) -- (alpha); \draw[stochastic] (sig_a) -- (alpha);
\draw[stochastic] (omega) -- (alpha); \draw[stochastic] (omega) -- (beta);
\draw[stochastic] (sig_b) -- (beta); \draw[stochastic] (mu_b) -- (beta);
\draw[stochastic] (alpha) -- (h);    \draw[stochastic] (beta) -- (h);
\draw[deterministic] (m) -- (h);
%% RIGHT: KRUSCHKE-STYLE NOTATION
\node[font=\large\bfseries] at (11, 2.2) {Kruschke-Style Notation};
\coordinate (k_mu_a)  at (7,  0); \draw (k_mu_a)  pic {norm_dist}; \node[above=0.7cm of k_mu_a,  font=\bfseries] {$\mu_\alpha$};
\coordinate (k_sig_a) at (9,  0); \draw (k_sig_a) pic {exp_dist};  \node[above=0.7cm of k_sig_a, font=\bfseries] {$\sigma_\alpha$};
\coordinate (k_omega) at (11, 0); \draw (k_omega) pic {lkj_dist};  \node[above=0.7cm of k_omega, font=\bfseries] {$\Omega$};
\coordinate (k_sig_b) at (13, 0); \draw (k_sig_b) pic {exp_dist};  \node[above=0.7cm of k_sig_b, font=\bfseries] {$\sigma_\beta$};
\coordinate (k_mu_b)  at (15, 0); \draw (k_mu_b)  pic {norm_dist}; \node[above=0.7cm of k_mu_b,  font=\bfseries] {$\mu_\beta$};
\coordinate (k_mvn) at (11, -3.5); \draw (k_mvn) pic {mvn_dist};
\node[above=0.9cm of k_mvn, font=\bfseries, align=center] {Agent Params \\ $(\alpha_j,\,\beta_j)$};
\node[discrete, observed, minimum size=0.9cm] (k_m) at (9, -7) {$m_{ij}$};
\node[above=0.5cm of k_m, font=\bfseries] {Memory};
\coordinate (k_h) at (13, -7); \draw (k_h) pic {bern_dist};
\node[above=0.9cm of k_h, font=\bfseries] {Choice ($h_{ij}$)};
\draw[plate, dashed]  (7.5, -5.2) rectangle (14.5, -2.2) node[below left, font=\small\bfseries] {$j=1\dots J$};
\draw[plate, dotted]  (7.5, -8.2) rectangle (14.5, -5.8) node[below left, font=\small\bfseries] {$i=1\dots N$};
\draw[stochastic] (k_mu_a) -- (k_mvn); \draw[stochastic] (k_sig_a) -- (k_mvn);
\draw[stochastic] (k_omega) -- (k_mvn) node[midway, right, font=\large] {$\sim$};
\draw[stochastic] (k_sig_b) -- (k_mvn); \draw[stochastic] (k_mu_b) -- (k_mvn);
\draw[stochastic] (k_mvn) -- (k_h) node[midway, right, font=\large] {$\sim$};
\draw[deterministic] (k_m) -- (12.5, -7);
\end{tikzpicture}
```




> **Computational Note: The Geometry of Hierarchical Models**
> The mathematical definition provided above—specifically $\theta_{\text{logit}, j} \sim \text{Normal}(\mu_{\theta}, \sigma_{\theta})$—is known as the **Centered Parameterization**. It is the most direct conceptual translation of our theory. However, Hamiltonian Monte Carlo (NUTS) often struggles with this exact form. As $\sigma_\theta$ approaches zero, the target geometry constrains into a region of extreme curvature known as "Neal's Funnel." In the next phase, we will purposefully implement this naive form to diagnose its pathological failures before introducing the coordinate-transform solution (Non-Centered Parameterization).

## Phase 1: Generative Plausibility of the Hierarchical Biased Agent

Before we write the inference algorithm in Stan, we must construct the forward generative model in R. This code must strictly mirror the mathematical dependencies formalized in our DAG.

We model the parameters on the logit (log-odds) scale. As discussed in previous chapters, probabilities are bounded between $0$ and $1$. If we attempt to place a hierarchical Normal distribution directly on the probability space, the probability mass will inevitably spill over the bounds, creating mathematically impossible values (e.g., a probability of $1.2$). The logit transform $\text{logit}(p) = \log(\frac{p}{1-p})$ maps the bounded $[0, 1]$ interval to the unbounded real line $(-\infty, \infty)$. The HMC sampler traverses this unbounded space smoothly, and we use the inverse-logit function (plogis() in R, inv_logit() in Stan) to map the inferences back to probabilities at the likelihood level.

Let us build a unified generative function. In accordance with the rigorous Bayesian workflow, this function is designed to either:

* Draw hyperparameters directly from our stated priors (for Prior Predictive Checks).

* Accept fixed "ground-truth" hyperparameters (for Parameter Recovery in Phase 2).

```{r}
#' Simulate Hierarchical Biased Agents
#'
#' @param n_agents Integer, number of agents (J).
#' @param n_trials_min Integer, minimum number of trials per agent.
#' @param n_trials_max Integer, maximum number of trials per agent.
#' @param mu_theta Numeric, population mean on the logit scale. If NULL, draws from prior.
#' @param sigma_theta Numeric, population SD on the logit scale. If NULL, draws from prior.
#' @param seed Integer, for exact computational reproducibility.
#'
#' @return A tidy tibble containing both latent parameters and observed choices.
simulate_hierarchical_biased <- function(n_agents = 100, 
                                         n_trials_min = 20, 
                                         n_trials_max = 200, 
                                         mu_theta = NULL, 
                                         sigma_theta = NULL,
                                         seed = NULL) {
  
  if (!is.null(seed)) set.seed(seed)
  
  # 1. Sample Population Hyperparameters (if not provided)
  if (is.null(mu_theta)) mu_theta <- rnorm(1, mean = 0, sd = 1.5)
  if (is.null(sigma_theta)) sigma_theta <- rexp(1, rate = 1)
  
  # 2. Sample Individual Latent Parameters and Trial Counts
  # We purposefully vary N_j to demonstrate shrinkage later.
  agent_params <- tibble(
    agent_id = 1:n_agents,
    n_trials = sample(n_trials_min:n_trials_max, n_agents, replace = TRUE),
    theta_logit = rnorm(n_agents, mean = mu_theta, sd = sigma_theta),
    theta_prob = plogis(theta_logit) 
  )
  
  # 3. Generate Observed Data (Likelihood)
  # h_ij ~ Bernoulli(inv_logit(theta_logit_j))
  sim_data <- agent_params |> 
    mutate(
      trial_data = map2(theta_prob, n_trials, function(prob, n) {
        tibble(
          trial = 1:n,
          choice = rbinom(n, size = 1, prob = prob)
        )
      })
    ) |> 
    unnest(trial_data)
  
  attr(sim_data, "hyperparameters") <- list(
    mu_theta = mu_theta,
    sigma_theta = sigma_theta
  )
  
  return(sim_data)
}
```

### Executing the Prior Predictive Check: The Logit-Normal Artifact

When placing priors on an unbounded scale (logit) that will be transformed to a bounded scale (probability), our intuition often fails. A seemingly "uninformative" prior like $\mu \sim \text{Normal}(0, 10)$ on the logit scale actually forces almost all probability mass to exactly $0$ or $1$ on the outcome scale, implying that all agents are entirely deterministic. 

Let us generate 12 datasets drawn entirely from our chosen priors ($\mu_\theta \sim N(0, 1.5), \sigma_\theta \sim \text{Exp}(1)$) to verify they generate scientifically plausible, heterogeneous populations.

```{r}
# Generate 12 distinct "universes" from the priors
prior_sims <- map_dfr(1:12, function(sim_id) {
  # We use a fixed seed per universe for stable workflow mapping
  sim_data <- simulate_hierarchical_biased(
    n_agents = 100, 
    n_trials_min = 1, n_trials_max = 1, # Only need 1 trial to extract parameters
    seed = sim_id * 10 
  )
  
  sim_data |> 
    distinct(agent_id, theta_prob) |> 
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

# Visualize the generative plausibility
ggplot(prior_sims, aes(x = theta_prob)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  facet_wrap(~simulation_id, ncol = 4) +
  scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  labs(
    title = "Prior Predictive Check: Implied Population Distributions",
    subtitle = "Healthy = populations spread across (0,1), not piled at 0 or 1. Each panel is one prior universe.",
    x = "Agent Bias (Probability Scale)",
    y = "Density"
  )
```

## Phase 2: Data Preparation and the Centered Stan Model

Before we unleash the Hamiltonian Monte Carlo (HMC) engine, we must format our simulated data structure to match the static typing of the Stan compiler. Because our agents have varying numbers of trials ($N_j$), we cannot pass a simple $J \times N$ matrix. Instead, we use a "long format" array indexing strategy, where a specific vector maps every trial $n \in \{1, \dots, N_{\text{total}}\}$ to its generating agent $j$.

```{r}
#' Prepare Data for Stan
#'
#' Maps a tidy tibble into the strictly typed list required by CmdStanR.
#' 
#' @param sim_data Tidy tibble generated by simulate_hierarchical_biased
#' @return A named list formatted for CmdStanR
prep_stan_data <- function(sim_data) {
  list(
    N = nrow(sim_data),                        # Total number of trials across all agents
    J = length(unique(sim_data$agent_id)),     # Total number of agents
    agent = sim_data$agent_id,                 # Vector of length N mapping trial to agent
    h = sim_data$choice                        # Vector of length N containing observed choices
  )
}

# Execute the preparation on a specifically generated universe for recovery
sim_data_centered <- simulate_hierarchical_biased(
  n_agents = 50, 
  n_trials_min = 30,  # Intentional sparsity to force shrinkage
  n_trials_max = 120,
  mu_theta = 1.38,# slightly below .8 in prob
  sigma_theta = 0.8,  
  seed = 1981
)
stan_data_centered <- prep_stan_data(sim_data_centered)
```

### Stan Model: Centered Parameterization (CP)

We will now implement the Centered Parameterization (CP) of the hierarchical biased agent. In CP, we explicitly model the individual latent parameters ($\theta_{\text{logit}, j}$) - determining their probability of choosing "right" versus "left" - as being drawn directly from a population distribution with mean `thetaM` and standard deviation `thetaSD`.

Mathematically, this translates to the following hierarchical structure, which strictly mirrors our Data Generating Process (DGP) from Phase 1:

* **Population Level**: $$\mu_\theta \sim \text{Normal}(0, 1.5)$$ $$\sigma_\theta \sim \text{Exponential}(1)$$

* **Individual Level**: $$\theta_{\text{logit}, j} \sim \text{Normal}(\mu_\theta, \sigma_\theta)$$

* **Data Level (Likelihood)**: $$h_{ij} \sim \text{Bernoulli}(\text{logit}^{-1}(\theta_{\text{logit}, j}))$$

This architecture mathematically instantiates partial pooling: the estimation of an individual's bias balances their specific choice history ($h_{ij}$) with the population's overarching distributional geometry ($\mu_\theta, \sigma_\theta$).

> The Anatomy of a Potential Pathology:
> Look closely at the individual-level formulation: $\theta_{\text{logit}, j} \sim \text{Normal}(\mu_\theta, \sigma_\theta)$. The permissible geometric space for $\theta_{\text{logit}, j}$ is dynamically dictated by $\sigma_\theta$. If the data suggests the population is highly homogenous, the sampler will push $\sigma_\theta \to 0$. As it does, the parameter space for $\theta_{\text{logit}, j}$ is violently crushed toward $\mu_\theta$, creating a funnel of infinite curvature. We are fitting this model precisely to watch the NUTS sampler fail in this region.

Let us write the Stan code.

```{r 06 Stan Model Biased Agent}

stan_biased_ml_cp_code <- "
// Multilevel Biased Agent Model: Centered Parameterization (CP)
data {
  int<lower=1> N;                       // Total number of observations across all agents
  int<lower=1> J;                       // Total number of agents
  array[N] int<lower=1, upper=J> agent; // Agent ID index for each trial
  array[N] int<lower=0, upper=1> h;     // Observed binary choices (1 = Right, 0 = Left)
}

parameters {
  // Population-level hyperparameters
  real mu_theta;                        // Population mean bias (logit scale)
  real<lower=0> sigma_theta;            // Population SD of bias (logit scale)

  // Individual-level parameters (CENTERED)
  vector[J] theta_logit;                // Agent-specific biases 
}

model {
  // 1. Population-level hyperpriors
  target += normal_lpdf(mu_theta | 0, 1.5);
  target += exponential_lpdf(sigma_theta | 1);

  // 2. Hierarchical prior (CENTERED)
  // The individual parameters are directly parameterized by the hyperparameters.
  // This explicitly creates the funnel geometry.
  target += normal_lpdf(theta_logit | mu_theta, sigma_theta);

  // 3. Likelihood
  // Highly vectorized evaluation mapping each trial to its specific agent
  target += bernoulli_logit_lpmf(h | theta_logit[agent]); 
}

generated quantities {
  // Transform hyperparameters back to the probability (outcome) scale for interpretation
  real<lower=0, upper=1> mu_theta_prob = inv_logit(mu_theta);
  vector<lower=0, upper=1>[J] theta_prob = inv_logit(theta_logit);

  // Initialize arrays for predictive checks, LOO-CV, and Prior Sensitivity
  vector[N] log_lik;
  array[N] int h_post_rep;
  array[N] int h_prior_rep;
  
  // Accumulator for joint prior log-density (required by 'priorsense')
  real lprior; 
  lprior = normal_lpdf(mu_theta | 0, 1.5) + 
           exponential_lpdf(sigma_theta | 1) + 
           normal_lpdf(theta_logit | mu_theta, sigma_theta);

  // Draw from the priors to establish the generative baseline
  real mu_theta_prior = normal_rng(0, 1.5);
  real sigma_theta_prior = exponential_rng(1);
  
  // We use a loop for trial-level generation to ensure array indexing stability
  for (i in 1:N) {
    // Pointwise log-likelihood for out-of-sample predictive performance (PSIS-LOO)
    log_lik[i] = bernoulli_logit_lpmf(h[i] | theta_logit[agent[i]]);
    
    // Posterior Predictive Check (PPC): Data generated from the fitted posterior
    h_post_rep[i] = bernoulli_logit_rng(theta_logit[agent[i]]);
    
    // Prior Predictive Check: Data generated entirely from the unconditioned priors
    // We sample a prior theta for the specific agent on the fly
    real theta_logit_prior_i = normal_rng(mu_theta_prior, sigma_theta_prior);
    h_prior_rep[i] = bernoulli_logit_rng(theta_logit_prior_i);
  }
}
"

# Save
stan_file_biased_ml_cp <- here::here("stan", "ch6_multilevel_biased_cp.stan")
writeLines(stan_biased_ml_cp_code, stan_file_biased_ml_cp)

# Compile the CP model utilizing the CmdStan C++ backend
mod_biased_ml_cp <- cmdstan_model(stan_file_biased_ml_cp)


```

### Fitting the model

This not really necessary yet, but a rigorous workflow demands an explicit initialization strategy, so we have all the pieces in place for more complex models. We will initialize the NUTS algorithm by drawing starting values directly from our generative priors. This ensures the chains begin in scientifically plausible regions of the parameter space, rather than Stan's arbitrary default $U[-2, 2]$ on the unconstrained space.

```{r 06_Fit_CP_Model}
fit_cp_path <- here::here("simmodels", "ch6_fit_cp.rds")

if (regenerate_simulations || !file.exists(fit_cp_path)) {
  
  # Create a plain list of lists (1 for each of the 4 chains)
  # This evaluates immediately and leaves no functions behind to capture environments!
  init_cp_list <- lapply(1:4, function(i) {
    list(
      mu_theta    = rnorm(1, 0, 1.5),
      sigma_theta = rexp(1, 1),
      theta_logit = rnorm(stan_data_centered$J, 0, 0.5)
    )
  })
  
  fit_cp <- mod_biased_ml_cp$sample(
    data            = stan_data_centered,
    seed            = 123,
    chains          = 4,
    parallel_chains = 4,
    iter_warmup     = 1000,
    iter_sampling   = 1000,
    init            = init_cp_list, 
    adapt_delta     = 0.85
  )
  
  fit_cp$save_object(fit_cp_path)
  
} else {
  fit_cp <- readRDS(fit_cp_path)
}
```

### Assessing the fitting process

Even if the Stan compiler finishes without throwing explicit warning messages, our workflow demands that we systematically verify the geometry of the posterior and the efficiency of the sampler. Hamiltonian Monte Carlo (HMC) is a physical simulation—it explores the posterior by simulating a particle gliding over a frictionless manifold dictated by the joint log-probability density. We must prove the numerical integration of this physics engine remained stable.

Let us systematically unpack the computational diagnostics of our Centered Parameterization fit.

1. The Diagnostic Summary: Integration Stability
First, we ask cmdstanr for a high-level summary of the NUTS sampler's performance.

```{r}
#fit_cp$cmdstan_diagnose()
fit_cp$diagnostic_summary()

```

We are looking for strict zeros in num_divergent. A divergent transition occurs when the numerical integration process (the leapfrog steps) fails to accurately map the curvature of the posterior manifold, causing the simulated trajectory to fly off the energy level. Even one divergence compromises the mathematical guarantee of the posterior. 

#### Chain Mixing and Stationarity ($\hat{R}$)

Did our independent Markov chains converge to the same stationary distribution? We visualize this using trace plots and quantify it using $\hat{R}$ (R-hat).

```{r}
# Extract the posterior array
draws_cp <- fit_cp$draws()

# Visual inspection of chain mixing
mcmc_trace(draws_cp, pars = c("mu_theta", "sigma_theta"), np = nuts_params(fit_cp)) +
  labs(
    title = "Trace Plots: Hyperparameters (CP Model)",
    subtitle = "Healthy = chains overlap and fluctuate around a stable band. Divergent red triangles mark sampler failures."
  )

# Quantitative assessment of convergence and efficiency
cp_summary <- draws_cp |> 
  subset_draws(variable = c("mu_theta", "sigma_theta", "theta_logit")) |> 
  summarise_draws(default_convergence_measures())

# We require R-hat < 1.01 for all parameters
cp_summary |> 
  filter(variable %in% c("mu_theta", "sigma_theta", "theta_logit[1]", "theta_logit[2]")) |>
  knitr::kable(digits = 3)
```

The potential scale reduction factor ($\hat{R}$) compares the variance within a single chain to the variance between all chains. If the chains are exploring the same typical set, this ratio converges to 1. Our strict threshold is $\hat{R} < 1.01$.

#### Effective Sample Size (ESS)

Markov chains are inherently autocorrelated. Stan might draw 4000 samples, but we need to know the Effective Sample Size—the equivalent number of completely independent draws.In your summary table, look at ess_bulk (efficiency of the posterior mean estimate) and ess_tail (efficiency of the posterior interval estimates). We strictly require ESS > 400 per parameter. If the geometry is difficult, ESS plummets, indicating the sampler is taking highly correlated, inefficient steps.

#### Energy and E-BFMI
Finally, we evaluate the Energy-Bayesian Fraction of Missing Information (E-BFMI). HMC relies on injecting kinetic energy (momentum) into the system to explore the tails of the target distribution.

```{r}
# Assess the momentum resampling efficiency
mcmc_nuts_energy(nuts_params(fit_cp)) +
  labs(
    title = "Energy Transitions",
    subtitle = "Marginal and transition energy distributions must align."
  )
```

If the model geometry is well-behaved, the energy transition distribution (the energy proposed by the momentum resampling) will neatly match the marginal energy distribution (the actual energy of the posterior). An E-BFMI below 0.2 indicates the sampler is taking tiny, localized steps and failing to explore the global geometry.

### Prior-Posterior Update Checks: Quantifying Learning
Bayesian inference is the mathematical process of allocating credibility across parameter values. The posterior is simply the prior updated by the likelihood. If the posterior is identical to the prior, the data contained no information. If the posterior is violently pushed against the boundaries of the prior, our priors were likely misspecified.

Let us visualize how much the data narrowed our uncertainty about the population parameters. Because we explicitly generated prior draws in our Stan generated quantities block, this comparison is trivial.

```{r}
# Extract prior and posterior draws for the population mean and variance
draws_df <- as_draws_df(draws_cp)

# Reshape for ggplot
pp_update_data <- draws_df |> 
  select(
    mu_posterior = mu_theta, mu_prior = mu_theta_prior,
    sigma_posterior = sigma_theta, sigma_prior = sigma_theta_prior
  ) |> 
  pivot_longer(everything(), names_to = c("parameter", "distribution"), names_sep = "_")

# Extract the true generative hyperparameters we saved as attributes
true_mu <- attr(sim_data_centered, "hyperparameters")$mu_theta
true_sigma <- attr(sim_data_centered, "hyperparameters")$sigma_theta

# Create a reference dataframe for the ground truth matching the facet names
true_pop_df <- tibble(
  parameter = c("mu", "sigma"),
  true_value = c(true_mu, true_sigma)
)

# Visualize the update
ggplot(pp_update_data, aes(x = value)) +
  geom_histogram(aes(fill = distribution), alpha = 0.6, color = NA) +
  # Add the true values as dashed lines
  geom_vline(data = true_pop_df, aes(xintercept = true_value, linetype = "True Value"), 
             color = "darkred", linewidth = 1) +
  facet_wrap(
    ~parameter, 
    scales = "free",
    # Clean up the facet labels for publication quality
    labeller = as_labeller(c("mu" = "Population Mean (\u03BC)", "sigma" = "Population SD (\u03C3)"))
  ) +
  scale_fill_manual(
    name = "Distribution",
    values = c("prior" = "gray70", "posterior" = "steelblue")
  ) +
  scale_linetype_manual(
    name = "Ground Truth", 
    values = c("True Value" = "dashed")
  ) +
  labs(
    title = "Prior-Posterior Update Checks: Population Hyperparameters",
    subtitle = "Healthy = posterior (blue) narrower than prior (grey) and centred on true value (dashed red).",
    x = "Parameter Value (Logit Scale)",
    y = "Density")


```

A healthy update shows a posterior that is significantly narrower than the prior (indicating learning) and comfortably contained within the prior's mass (indicating the prior did not inappropriately censor the likelihood).

### Posterior Parameter Correlations

Parameters in hierarchical models are structurally correlated. By plotting the joint posterior geometry of multiple parameters, we can detect pathological collinearity or explicitly visualize the boundary of possible Neal's Funnel.

We will use a pairs plot to examine the joint distribution of the population mean ($\mu_\theta$), population standard deviation ($\sigma_\theta$), and a single agent's latent bias ($\theta_{\text{logit}, 1}$).

```{r}
# Visualize joint posterior geometries and identify structural correlations
mcmc_pairs(
  draws_cp, 
  pars = c("mu_theta", "sigma_theta", "theta_logit[1]"),
  np = nuts_params(fit_cp), # Injects divergence data (red crosses) if any exist
  diag_fun = "dens",
  off_diag_fun = "hex"
)
```

### Posterior Predictive Checks (PPC): Generative Adequacy
Can our fitted model actually generate the reality we observed? If we simulate data from the posterior distribution, does it look like the empirical data? If not, the model is structurally misspecified, regardless of how well the chains mixed.

We will compare the observed choice rate for each agent against the distribution of choice rates predicted by the posterior.

```{r}
# Safely extract vectors
y_obs <- as.numeric(stan_data_centered$h)
agent_idx <- as.numeric(stan_data_centered$agent)

# Use cmdstanr's native extraction to guarantee a strict S x N matrix
y_rep <- fit_cp$draws("h_post_rep", format = "matrix")

# Select a random subset of agents to avoid overcrowded, illegible plots
n_sample <- 12
sampled_agents <- sample(unique(agent_idx), n_sample)

# Create a logical mask mapping which trials belong to the sampled agents
keep_mask <- agent_idx %in% sampled_agents

# Subset the data exactly
y_obs_sub <- y_obs[keep_mask]
agent_idx_sub <- agent_idx[keep_mask]
# For y_rep, we subset the columns (trials)
y_rep_sub <- y_rep[ , keep_mask] 

# Define a custom statistic to explicitly state our cognitive metric 
# (and simultaneously silence the bayesplot warning)
prop_right <- function(x) mean(x)

# Execute the grouped Posterior Predictive Check
ppc_stat_grouped(
  y = y_obs_sub, 
  yrep = y_rep_sub, 
  group = agent_idx_sub, 
  stat = "prop_right"
) +
  labs(
    title = "Posterior Predictive Check: Agent-Level Choice Rates",
    subtitle = paste0("Healthy = dark line (observed) inside histogram (predicted). Showing ", n_sample, " agents."),
    x = "Proportion of 'Right' Choices"
  )
```

### Calibration and Recovery

We have proven the model fits the data and generates plausible predictions. Now, we must prove that the parameters we extract are actually the true parameters that generated the data. Because this is a hierarchical model, we must assess recovery at both the population and individual levels.

#### Individual-Level Recovery

Having validated the top of the hierarchy (in the prior-posterior update check), we move to the individual agents. We plot the posterior estimates against the ground truth. We expect the estimates to align with the diagonal identity line, but crucially, we expect agents with fewer trials to deviate from the raw truth, demonstrating shrinkage toward the estimated population mean.

```{r}
# 1. Extract the full summary (which defaults to including median, q5, and q95)
recovery_summary <- draws_cp |> 
  posterior::subset_draws(variable = "theta_logit") |> 
  posterior::summarise_draws() |> 
  mutate(agent_id = as.integer(str_extract(variable, "(?<=\\[)\\d+(?=\\])")))

# 2. Extract the true generative parameters directly from the simulated dataset
true_agent_params <- sim_data_centered |> 
  distinct(agent_id, n_trials, theta_logit, theta_prob)

# 3. Join the posterior estimates with the ground truth
recovery_df <- true_agent_params |> 
  left_join(recovery_summary, by = "agent_id")

# 4. Plot Recovery with Shrinkage Dynamics
ggplot(recovery_df, aes(x = theta_logit, y = median)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0, alpha = 0.5, color = "steelblue") +
  geom_point(aes(size = n_trials), color = "midnightblue", alpha = 0.8) +
  labs(
    title = "Parameter Recovery: Hierarchical Biased Agent",
    subtitle = "Healthy = points near diagonal (y = x). Small dots shrink more — that is partial pooling working correctly.",
    x = "True Parameter Value (Logit Scale)",
    y = "Posterior Median Estimate (Logit Scale)",
    size = "Number of Trials (N_j)"
  )
```

### Prior sensitivity

```{r}
# 1. Calculate the sensitivity diagnostics
# We target the hyperparameters, as they govern the hierarchical shrinkage
sens_results <- powerscale_sensitivity(
  fit_cp, # or fit_ncp, provided lprior is defined
  variable = c("mu_theta", "sigma_theta")
)

cat("=== PRIOR SENSITIVITY DIAGNOSTICS ===\n")
print(sens_results)

ps_seq <- priorsense::powerscale_sequence(
  fit_cp, 
  variable = c("mu_theta", "sigma_theta")
)

# Plot the density shifts
priorsense::powerscale_plot_dens(ps_seq) +
  theme_classic() +
  labs(
    title = "Prior Sensitivity: Density Perturbation",
    subtitle = "Does scaling the prior/likelihood drastically alter our geometric beliefs?"
  )

# Plot the quantitative estimate shifts
priorsense::powerscale_plot_quantities(ps_seq) +
  theme_classic() +
  labs(
    title = "Prior Sensitivity: Parameter Estimates",
    subtitle = "Horizontal lines indicate the baseline posterior bounds. Deviations indicate sensitivity."
  )
```

The sensitivity analysis reveals that our population variance ($\sigma_\theta$) exhibits a prior-data conflict. This is a feature, not a bug, of partial pooling: our Exponential prior applies regularizing pressure toward zero variance (complete pooling), while the empirical data resists this pull to account for individual differences.

### Simulation-Based Calibration (SBC)

Parameter recovery on a single dataset is anecdotal. A single success might just mean we got lucky with a benign region of the parameter space. To mathematically guarantee that our NUTS sampler is unbiased and our posterior uncertainties are perfectly calibrated across the entire prior space, we must perform Simulation-Based Calibration (SBC).
The logic of SBC is structurally elegant: if we generate a ground-truth parameter $\theta_{\text{true}}$ from the prior, generate data from the likelihood conditional on $\theta_{\text{true}}$, and compute the posterior, then the posterior must not systematically over- or under-estimate $\theta_{\text{true}}$. Under a well-calibrated sampler, the rank of the true parameter value within the sorted posterior draws will be uniformly distributed:$$r \sim \text{Uniform}(0, S)$$where $S$ is the number of posterior samples.
If the model is overconfident (the posterior is too narrow and misses the truth), the rank histogram will be U-shaped. If it is underconfident (the posterior is too wide), it will form an inverted U-shape. Asymmetries indicate directional bias.

### Setting up the SBC Pipeline
The SBC package requires a strict data structure. It expects a generator function that returns a list of two components:

* variables: A named list of the true parameter values.

* generated: The formatted data list required by our Stan model.

Let us build a strict wrapper around our simulation and data preparation functions to satisfy this requirement.

```{r}
# Define the strict SBC generator wrapper
generate_sbc_dataset <- function(n_agents = 50, n_trials_min = 20, n_trials_max = 120) {
  
  # 1. Simulate the universe (draws hyperparams and agent params from priors)
  sim_data <- simulate_hierarchical_biased(
    n_agents = n_agents, 
    n_trials_min = n_trials_min, 
    n_trials_max = n_trials_max
  )
  
  # 2. Extract the true generative hyperparameters
  mu_true <- attr(sim_data, "hyperparameters")$mu_theta
  sigma_true <- attr(sim_data, "hyperparameters")$sigma_theta
  
  # 3. Extract the true individual latent parameters (must preserve agent order)
  theta_true <- sim_data |> 
    distinct(agent_id, theta_logit) |> 
    arrange(agent_id) |> 
    pull(theta_logit)
  
  # 4. Prepare the strictly typed list for CmdStanR
  stan_data <- prep_stan_data(sim_data)
  
  # 5. Return the exact structure required by the SBC package
  list(
    variables = list(
      mu_theta = mu_true,
      sigma_theta = sigma_true,
      theta_logit = theta_true
    ),
    generated = stan_data
  )
}
```

A production-grade SBC requires $N_{\text{sim}} \ge 1000$ to smooth the rank histograms. For computational feasibility in this chapter, we will run 200 simulations. Because we established our future parallel processing plan in the setup chunk, the SBC package will automatically distribute these models across your CPU cores.

```{r}

sbc_cp_path <- here::here("simmodels", "ch6_sbc_cp.rds")

generator_cp <- SBC_generator_function(
  generate_sbc_dataset,
  n_agents = 50, n_trials_min = 30, n_trials_max = 120
)

backend_cp <- SBC_backend_cmdstan_sample(
  mod_biased_ml_cp,
  iter_warmup = 1000, iter_sampling = 1000, chains = 1, refresh = 0
)

if (regenerate_simulations || !file.exists(sbc_cp_path)) {
  
  sbc_datasets_cp <- generate_datasets(generator_cp, 200)
  sbc_results_cp  <- compute_SBC(datasets = sbc_datasets_cp, backend = backend_cp,
  keep_fits = FALSE)
  saveRDS(
    list(datasets = sbc_datasets_cp, results = sbc_results_cp),
    sbc_cp_path
  )
  
} else {
  sbc_obj        <- readRDS(sbc_cp_path)
  sbc_datasets_cp <- sbc_obj$datasets
  sbc_results_cp  <- sbc_obj$results
}

# Visualize
plot_rank_overlay(sbc_results_cp, variables = c("mu_theta", "sigma_theta")) +
  labs(
    title    = "SBC Rank Histograms: Centered Parameterization",
    subtitle = "Healthy = bars flat within grey band. U-shape = overconfident. Here, U-shapes reveal the funnel pathology."
  )

plot_ecdf_diff(sbc_results_cp, variables = c("mu_theta", "sigma_theta")) +
  labs(
    title = "SBC ECDF Difference: Centered Parameterization",
    subtitle = "Deviation from the diagonal indicates miscalibration. The rank histograms\nare underpowered at N=200 for subtle distortions; divergence rates below are more sensitive."
  )

# 1. Extract the lightweight stats dataframe from the SBC object
sbc_stats_df <- sbc_results_cp$stats

# 2. Filter for the population-level parameters you care about
# (e.g., mu_alpha, mu_beta, rho, or even sigma[1])
recovery_df <- sbc_stats_df |>
  filter(variable %in% c("mu_theta"))

# 3. Create the Parameter Recovery Identity Plot
p_sbc_recovery <- ggplot(recovery_df, aes(x = simulated_value, y = median)) +
  # The Identity Line (Perfect Recovery)
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50", linewidth = 1) +
  
  # The Posterior Estimates with 90% Credible Intervals
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0, alpha = 0.3, color = "steelblue") +
  geom_point(color = "midnightblue", alpha = 0.7, size = 2) +
  
  # Facet by parameter
  facet_wrap(~variable, scales = "free") +
  
  labs(
    title = "Parameter Recovery (Extracted Directly from SBC)",
    subtitle = "Points represent the posterior median. Bars are 90% Credible Intervals.",
    x = "True Generative Value (Drawn from Prior)",
    y = "Posterior Median Estimate"
  ) +
  theme_classic() +
  theme(strip.background = element_rect(fill = "gray95", color = NA))

print(p_sbc_recovery)

# 4. Quantify the Recovery (Correlation and RMSE)
recovery_metrics <- recovery_df |>
  group_by(variable) |>
  summarize(
    Pearson_r = cor(simulated_value, median),
    Spearman_rho = cor(simulated_value, median, method = "spearman"),
    RMSE = sqrt(mean((simulated_value - median)^2)),
    .groups = "drop"
  )

cat("=== PARAMETER RECOVERY QUANTIFICATION ===\n")
print(knitr::kable(recovery_metrics, digits = 3))
```

Let's explore the warnings we got.

```{r}
# Extract diagnostics now so they are available for the targeted refit below
diags_cp <- sbc_results_cp$backend_diagnostics
cp_fail_rate <- mean(diags_cp$n_divergent > 0) * 100

# Divergence count distribution across all 200 universes
diags_cp |> 
  ggplot(aes(x = n_divergent)) +
  geom_histogram(binwidth = 1, fill = "darkred", alpha = 0.8) +
  labs(
    title = "CP Divergence Counts Across 200 SBC Universes",
    subtitle = sprintf("%.0f%% of universes produced at least one divergence. 
Even a single divergence invalidates the geometric guarantee of NUTS.", cp_fail_rate),
    x = "Number of Divergent Transitions",
    y = "Number of SBC Universes"
  )

# Locate the divergence-prone region: do they concentrate at small sigma_theta?
divergence_profile <- posterior::as_draws_df(sbc_datasets_cp$variables) |> 
  as_tibble() |> 
  mutate(sim_id = row_number()) |> 
  left_join(
    diags_cp |> mutate(sim_id = row_number()) |> select(sim_id, n_divergent),
    by = "sim_id"
  ) |> 
  mutate(had_divergence = n_divergent > 0)

ggplot(divergence_profile, aes(x = sigma_theta, y = n_divergent, color = had_divergence)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "darkred"),
    labels = c("FALSE" = "Clean", "TRUE" = "Divergent"),
    name = NULL
  ) +
  labs(
    title = "CP Divergences Concentrate at Small True sigma_theta",
    subtitle = "The funnel geometry only manifests when population variance is small.",
    x = expression("True " * sigma[theta]),
    y = "Number of Divergent Transitions"
  )

ggplot(divergence_profile, aes(x = mu_theta, y = sigma_theta, color = had_divergence)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(
    name = "Sampler Status",
    values = c("FALSE" = "steelblue", "TRUE" = "darkred"),
    labels = c("Healthy", "Diverged")
  ) +
  labs(
    title = "The Anatomy of Failure: Where does CP break?",
    x = "True Population Mean (mu_theta)",
    y = "True Population Variance (sigma_theta)"
  ) +
  theme_classic()
```

```{r}
hard_universe_path <- here::here("simmodels", "ch6_fit_cp_hard.rds")

hard_universe_idx <- divergence_profile |> 
  filter(had_divergence) |> 
  slice_min(sigma_theta, n = 1) |> 
  pull(sim_id)

hard_stan_data   <- sbc_datasets_cp$generated[[hard_universe_idx]]
hard_true_sigma  <- divergence_profile$sigma_theta[hard_universe_idx]

cat(sprintf(
  "Selected universe %d: true sigma_theta = %.3f\n",
  hard_universe_idx, hard_true_sigma
))

if (regenerate_simulations || !file.exists(hard_universe_path)) {
  fit_cp_hard <- mod_biased_ml_cp$sample(
    data            = hard_stan_data,
    seed            = 999,
    chains          = 4,
    parallel_chains = 4,
    iter_warmup     = 1000,
    iter_sampling   = 1000,
    init            = init_cp_list,
    adapt_delta     = 0.85, 
    refresh         = 0
  )
  fit_cp_hard$save_object(hard_universe_path)
} else {
  fit_cp_hard <- readRDS(hard_universe_path)
}

# 1. Generate the pairs plot 
p_funnel <- mcmc_pairs(
  fit_cp_hard$draws(),
  pars     = c("mu_theta", "sigma_theta", "theta_logit[1]"),
  np       = nuts_params(fit_cp_hard),
  np_style = pairs_style_np(div_color = "darkred", div_alpha = 0.8, div_size = 2),
  diag_fun     = "dens",
  off_diag_fun = "hex"
)

wrap_elements(p_funnel) + 
  plot_annotation(
    title = sprintf("Neal's Funnel: CP Geometry at True sigma_theta = %.3f", hard_true_sigma),
    subtitle = "Red points are divergent transitions crashing at the funnel neck."
  )
```

Here we see a funnel. As sigma goes to zero, theta_logit[1] can only have one value: the value of the population (mu_theta). The curvature becomes infinitely steep, and the NUTS sampler fails to explore this region, resulting in divergent transitions (red points). This is the geometric pathology of the Centered Parameterization.

### Non-Centered Parameterization (NCP) as the Solution

We have reached a critical juncture in our Bayesian workflow. We showed  computationally (via SBC and diagnostic pairs plots) that the Centered Parameterization (CP) creates a pathological geometry—Neal’s Funnel—that makes it difficult for the NUTS sampler to efficiently traverse the posterior space when data is sparse or population variance is small.
When the model is misbehaving, sometimes we need to reconsider our theory and/or our data. Yet, in this case, we can do something to the math underlying the model to topologically transform the posterior space the model is sampling: a coordinate transformation known as the Non-Centered Parameterization (NCP).

#### The Geometry of the Transformation

In the Centered Parameterization, the hierarchical dependency is explicit and direct:$$\theta_{\text{logit}, j} \sim \text{Normal}(\mu_\theta, \sigma_\theta)$$
This forces the sampler to navigate a space where the permissible bounds of $\theta$ are dynamically restricted by $\sigma$, creating the funnel.
In the Non-Centered Parameterization, we decouple this dependency, in a way that is not dissimilar from rescaling your predictors - a common procedure you probably learned in your statistical course to help model convergence and interpretation.

We introduce an auxiliary, standardized latent variable $z_j$ for each agent. We force the sampler to explore this $z$-space, which is simply a perfectly behaved Normal distribution:$$z_j \sim \text{Normal}(0, 1)$$
Then, we deterministically shift and scale $z_j$ to reconstruct our target cognitive parameter:$$\theta_{\text{logit}, j} = \mu_\theta + z_j \times \sigma_\theta$$

Mathematically, this is identical to the CP ($z \sim N(0,1) \implies \mu + z\sigma \sim N(\mu, \sigma)$). However, the computational geometry is entirely transformed and when the sampler traverses it, the problems it meets are likely different.

### Stan Model: Multilevel Biased Agent (NCP)

Here is the NCP version of the model. Note the changes in the parameters, transformed parameters, and model blocks compared to the CP version.

```{r}

stan_biased_ml_ncp_code <- '
// Multilevel Biased Agent Model (Non-Centered Parameterization)
// Estimates individual biases drawn from a population distribution.

data {
  int<lower=1> N;                    // Total number of observations
  int<lower=1> J;                    // Number of agents
  array[N] int<lower=1, upper=J> agent; // Agent ID for each observation (must be 1...J)
  array[N] int<lower=0, upper=1> h;    // Observed choices (0 or 1)
}

parameters {
  // Population-level parameters
  real mu_theta;                     // Population mean bias (logit scale)
  real<lower=0> sigma_theta;         // Population SD of bias (logit scale)

  // Individual-level parameters (standardized deviations, z-scores of the agents)
  vector[J] z_theta;                 // Non-centered individual effects (standard normal scale)
}

transformed parameters {
  // Deterministic reconstruction of the cognitive parameters
  // This shifts and scales the z-scores back to the target logit space
  vector[J] theta_logit = mu_theta + z_theta * sigma_theta;
}

model {
  // Priors for population-level parameters
  target += normal_lpdf(mu_theta | 0, 1.5);
  target += exponential_lpdf(sigma_theta | 1);
  
  /// Individual level prior (NON-CENTERED)
  // Notice there are no population level parameters here. The geometry is decoupled.
  target += std_normal_lpdf(z_theta);

  // Likelihood
  /// The likelihood is evaluated on the deterministically transformed parameters
  /// not the raw z-scores.
  target += bernoulli_logit_lpmf(h | theta_logit[agent]);
}

generated quantities {
  // Transform hyperparameters back to the probability (outcome) scale
  real<lower=0, upper=1> mu_theta_prob = inv_logit(mu_theta);
  vector<lower=0, upper=1>[J] theta_prob = inv_logit(theta_logit);
  
  // Initialize containers for trial-level metrics
  vector[N] log_lik;
  array[N] int h_post_rep;
  array[N] int h_prior_rep;

  // --- Prior Log-Density  ---
  real lprior;
  // Notice we score z_theta with the standard normal distribution, exactly as in the model block
  lprior = normal_lpdf(mu_theta | 0, 1.5) + 
           exponential_lpdf(sigma_theta | 1) + 
           std_normal_lpdf(z_theta);

  // --- Prior Predictive Checks: Generative Baseline ---
  // 1. Draw population hyperparameters from the EXACT priors used in the model block
  real mu_theta_prior = normal_rng(0, 1.5);
  real<lower=0> sigma_theta_prior = exponential_rng(1);
  
  // 2. Draw individual agent parameters from the non-centered prior
  vector[J] theta_logit_prior;
  for (j in 1:J) {
    real z_theta_prior = normal_rng(0, 1); 
    theta_logit_prior[j] = mu_theta_prior + z_theta_prior * sigma_theta_prior;
  }

  // --- Trial-Level Computations ---
  // We consolidate the log-likelihood and predictive checks into a single loop
  for (i in 1:N) {
    // 1. Pointwise log-likelihood (for PSIS-LOO model comparison later)
    log_lik[i] = bernoulli_logit_lpmf(h[i] | theta_logit[agent[i]]);
    
    // 2. Posterior Predictive Check: Generate choices using the fitted posterior
    h_post_rep[i] = bernoulli_logit_rng(theta_logit[agent[i]]);
    
    // 3. Prior Predictive Check: Generate choices using the unconditioned prior
    h_prior_rep[i] = bernoulli_logit_rng(theta_logit_prior[agent[i]]);
  }
}'

# Write Stan code to file
stan_file_ncp <- here::here("stan", "ch6_multilevel_biased_ncp.stan")
writeLines(stan_biased_ml_ncp_code, stan_file_ncp)

mod_biased_ml_ncp <- cmdstan_model(stan_file_ncp)

fit_ncp_path <- here::here("simmodels", "ch6_fit_ncp.rds")

# Extract J safely
  J_val_ncp <- stan_data_centered$J
  
  # The dumb list (4 chains)
  init_ncp_list <- list(
    list(mu_theta = rnorm(1, 0, 1.5), sigma_theta = rexp(1, 1), z_theta = rnorm(J_val_ncp, 0, 1)),
    list(mu_theta = rnorm(1, 0, 1.5), sigma_theta = rexp(1, 1), z_theta = rnorm(J_val_ncp, 0, 1)),
    list(mu_theta = rnorm(1, 0, 1.5), sigma_theta = rexp(1, 1), z_theta = rnorm(J_val_ncp, 0, 1)),
    list(mu_theta = rnorm(1, 0, 1.5), sigma_theta = rexp(1, 1), z_theta = rnorm(J_val_ncp, 0, 1))
  )

  if (regenerate_simulations || !file.exists(fit_ncp_path)) {
    fit_ncp <- mod_biased_ml_ncp$sample(
      data            = stan_data_centered,
      seed            = 123,
      chains          = 4,
      parallel_chains = 4,
      iter_warmup     = 1000,
      iter_sampling   = 1000,
      init            = init_ncp_list, # <--- Pass the dumb list here
      adapt_delta     = 0.85,
      refresh         = 0
    )
    fit_ncp$save_object(fit_ncp_path)
} else {
  fit_ncp <- readRDS(fit_ncp_path)
}
```

We'll skip the step-by-step model quality checks (bad!) to focus on two indices of performance: 1) the time it takes to fit (it should have improved) and 2) SBC, the part that was creating issues for the CP parameterization.

```{r 06_Time_NCP_Model}
# Timing and ESS comparison 
time_cp  <- fit_cp$time()$total
time_ncp <- fit_ncp$time()$total

ess_cp  <- summarise_draws(fit_cp$draws(c("mu_theta", "sigma_theta")),  "ess_bulk") |> rename(ess_cp  = ess_bulk)
ess_ncp <- summarise_draws(fit_ncp$draws(c("mu_theta", "sigma_theta")), "ess_bulk") |> rename(ess_ncp = ess_bulk)

efficiency_comparison <- ess_cp |> 
  left_join(ess_ncp, by = "variable") |> 
  mutate(
    ess_per_sec_cp  = ess_cp  / time_cp,
    ess_per_sec_ncp = ess_ncp / time_ncp
  )

cat("=== COMPUTATIONAL SPEED (WALL-CLOCK) ===\n")
cat("CP  Total Time:", round(time_cp,  2), "seconds\n")
cat("NCP Total Time:", round(time_ncp, 2), "seconds\n")
cat("Speedup Factor:", round(time_cp / time_ncp, 2), "x\n\n")
cat("=== SAMPLING EFFICIENCY (ESS / SECOND) ===\n")
print(knitr::kable(efficiency_comparison, digits = 1))

```

Here we are seeing that the NCP model is sometimes slightly faster, sometimes slightly slower, but always less efficient at sampling than the CP model. There goes any simple guideline that NCP is more efficient :-)

Yet, NCP might still end up exploring the space better. Let's see how it fares on the SBC assessment that was giving issues to CP.

We will generate 200 random universes from the priors, fit the NCP model to each, and plot the rank histograms.

```{r}
sbc_ncp_path <- here::here("simmodels", "ch6_sbc_ncp.rds")

backend_ncp <- SBC_backend_cmdstan_sample(
  mod_biased_ml_ncp,
  iter_warmup = 1000, iter_sampling = 1000, chains = 1, refresh = 0
)

if (regenerate_simulations || !file.exists(sbc_ncp_path)) {
  future::plan(future::sequential) # to avoid ram issues
  # Reuse the exact same universes as the CP SBC for a controlled comparison
  sbc_results_ncp <- compute_SBC(datasets = sbc_datasets_cp, backend = backend_ncp)
  saveRDS(sbc_results_ncp, sbc_ncp_path)
} else {
  sbc_results_ncp <- readRDS(sbc_ncp_path)
}

plot_rank_hist(sbc_results_ncp, variables = c("mu_theta", "sigma_theta")) +
  labs(title = "SBC Rank Histograms: Non-Centered Parameterization")

sbc_stats_df <- sbc_results_cp$stats

# 2. Filter for the population-level parameters you care about
# (e.g., mu_alpha, mu_beta, rho, or even sigma[1])
recovery_df <- sbc_stats_df |>
  filter(variable %in% c("mu_theta", "sigma_theta"))

# 3. Create the Parameter Recovery Identity Plot
p_sbc_recovery <- ggplot(recovery_df, aes(x = simulated_value, y = median)) +
  # The Identity Line (Perfect Recovery)
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50", linewidth = 1) +
  
  # The Posterior Estimates with 90% Credible Intervals
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0, alpha = 0.3, color = "steelblue") +
  geom_point(color = "midnightblue", alpha = 0.7, size = 2) +
  
  # Facet by parameter
  facet_wrap(~variable, scales = "free") +
  
  labs(
    title = "Parameter Recovery (Extracted Directly from SBC)",
    subtitle = "Points represent the posterior median. Bars are 90% Credible Intervals.",
    x = "True Generative Value (Drawn from Prior)",
    y = "Posterior Median Estimate"
  ) +
  theme_classic() +
  theme(strip.background = element_rect(fill = "gray95", color = NA))

print(p_sbc_recovery)

# 4. Quantify the Recovery (Correlation and RMSE)
recovery_metrics <- recovery_df |>
  group_by(variable) |>
  summarize(
    Pearson_r = cor(simulated_value, median),
    Spearman_rho = cor(simulated_value, median, method = "spearman"),
    RMSE = sqrt(mean((simulated_value - median)^2)),
    .groups = "drop"
  )

cat("=== PARAMETER RECOVERY QUANTIFICATION ===\n")
print(knitr::kable(recovery_metrics, digits = 3))

# Extract diagnostics now so they are available for the targeted refit below
diags_ncp <- sbc_results_ncp$backend_diagnostics
ncp_fail_rate <- mean(diags_ncp$n_divergent > 0) * 100

# Divergence count distribution across all 200 universes
diags_ncp |> 
  ggplot(aes(x = n_divergent)) +
  geom_histogram(binwidth = 1, fill = "darkred", alpha = 0.8) +
  labs(
    title = "NCP Divergence Counts Across 200 SBC Universes",
    subtitle = sprintf("%.0f%% of universes produced at least one divergence. 
Even a single divergence invalidates the geometric guarantee of NUTS.", cp_fail_rate),
    x = "Number of Divergent Transitions",
    y = "Number of SBC Universes"
  )

# Locate the divergence-prone region: do they concentrate at small sigma_theta?
divergence_profile <- posterior::as_draws_df(sbc_datasets_cp$variables) |> 
  as_tibble() |> 
  mutate(sim_id = row_number()) |> 
  left_join(
    diags_ncp |> mutate(sim_id = row_number()) |> select(sim_id, n_divergent),
    by = "sim_id"
  ) |> 
  mutate(had_divergence = n_divergent > 0)

ggplot(divergence_profile, aes(x = sigma_theta, y = n_divergent, color = had_divergence)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "darkred"),
    labels = c("FALSE" = "Clean", "TRUE" = "Divergent"),
    name = NULL
  ) +
  labs(
    title = "NCP Divergences Concentrate at Small True sigma_theta",
    subtitle = "The funnel geometry only manifests when population variance is small.",
    x = expression("True " * sigma[theta]),
    y = "Number of Divergent Transitions"
  )

ggplot(divergence_profile, aes(x = mu_theta, y = sigma_theta, color = had_divergence)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(
    name = "Sampler Status",
    values = c("FALSE" = "steelblue", "TRUE" = "darkred"),
    labels = c("Healthy", "Diverged")
  ) +
  labs(
    title = "The Anatomy of Failure: Where does NCP break?",
    x = "True Population Mean (mu_theta)",
    y = "True Population Variance (sigma_theta)"
  ) +
  theme_classic()
```

#### Posterior Simulation-Based Calibration: Probing Local Geometry

Prior SBC tested our sampler across the entire prior space — it asked whether NUTS is unbiased when the true parameters could be anything the prior assigns non-negligible mass to. This is a global guarantee, and it is necessary. But it is not sufficient. A sampler can be globally unbiased yet struggle precisely in the regions that matter most: the posterior neighborhoods that the actual observed data force us into. When the likelihood is highly informative — as it typically is once we have a few hundred trials per agent — the posterior concentrates sharply in a small corner of the prior space. The sampler must now take large numbers of tiny, precise steps in a geometrically restricted region. Prior SBC never directly stresses this regime, because most of its simulated universes place the true parameter somewhere across the full prior, and only a fraction happen to fall in the data-rich posterior neighborhood.

Posterior SBC (PSBC) is a complementary local diagnostic that directly targets this regime. The logic inverts the generative direction. Instead of drawing true parameters from the prior, we draw them from the *fitted posterior* — a distribution that has already been conditioned on our actual data and therefore concentrates in precisely the region the sampler will need to navigate in practice. Concretely: we sample a single draw $\tilde{\theta}^{(s)}$ from the posterior ($\theta \mid y_{\text{obs}}$), treat it as a ground truth, generate a synthetic dataset $\tilde{y}^{(s)} \sim p(y \mid \tilde{\theta}^{(s)})$ from the likelihood, fit the model to $\tilde{y}^{(s)}$, and record where $\tilde{\theta}^{(s)}$ falls in the resulting posterior. If the sampler is locally well-calibrated, these ranks are again uniform — but now over the concentrated geometry of the posterior, not the diffuse geometry of the prior. A U-shaped rank histogram here means the sampler is overconfident *specifically when the data are informative*, which is exactly the failure mode we need to detect. Implementing this requires us to write a *factory function*: a function that closes over a fitted model object and a data structure, and returns a new generator function that the `SBC` package can call repeatedly to produce fresh $({\tilde{\theta}}, \tilde{y})$ pairs. If this pattern is unfamiliar, read it as a function that *builds* generators rather than *being* one — `make_psbc_generator(fit_cp, stan_data_centered)` does not itself generate a dataset; it returns a self-contained callable that will do so every time the SBC machinery invokes it.

```{r}
# 1. Define a robust factory function for the PSBC generator
make_psbc_generator <- function(fit, original_data) {
  # Extract the full posterior array as a flat data frame
  draws_df <- posterior::as_draws_df(fit$draws())
  
  # Pre-compute the exact column names to guarantee flawless extraction
  theta_cols <- paste0("theta_logit[", 1:original_data$J, "]")
  y_cols <- paste0("h_post_rep[", 1:original_data$N, "]")
  
  # The generator function required by the SBC package
  function() {
    # Step A: Randomly sample one specific "state of the world" from the posterior
    idx <- sample(seq_len(nrow(draws_df)), 1)
    
    # Step B: Package the true parameters for this specific simulation
    variables <- list(
      mu_theta = as.numeric(draws_df$mu_theta[idx]),
      sigma_theta = as.numeric(draws_df$sigma_theta[idx]),
      theta_logit = as.numeric(draws_df[idx, theta_cols])
    )
    
    # Step C: Package the dataset generated by this specific state of the world
    generated <- list(
      N = original_data$N,
      J = original_data$J,
      agent = original_data$agent,
      h = as.numeric(draws_df[idx, y_cols])
    )
    
    list(variables = variables, generated = generated)
  }
}

psbc_path <- here::here("simmodels", "ch6_psbc.rds")

if (regenerate_simulations || !file.exists(psbc_path)) {
  
  gen_psbc_cp  <- SBC_generator_function(make_psbc_generator(fit_cp,  stan_data_centered))
  gen_psbc_ncp <- SBC_generator_function(make_psbc_generator(fit_ncp, stan_data_centered))
  
  set.seed(8675309)
  psbc_data_cp  <- generate_datasets(gen_psbc_cp,  100)
  psbc_data_ncp <- generate_datasets(gen_psbc_ncp, 100)
  
  backend_cp_psbc  <- SBC_backend_cmdstan_sample(mod_biased_ml_cp,  iter_warmup = 1000, iter_sampling = 1000, chains = 1, refresh = 0)
  backend_ncp_psbc <- SBC_backend_cmdstan_sample(mod_biased_ml_ncp, iter_warmup = 1000, iter_sampling = 1000, chains = 1, refresh = 0)
  future::plan(future::sequential) # to avoid ram issues
  psbc_res_cp  <- compute_SBC(psbc_data_cp,  backend_cp_psbc)
  psbc_res_ncp <- compute_SBC(psbc_data_ncp, backend_ncp_psbc)
  
  saveRDS(
    list(
      res_cp  = psbc_res_cp,
      res_ncp = psbc_res_ncp
    ),
    psbc_path
  )
  
} else {
  psbc_obj    <- readRDS(psbc_path)
  psbc_res_cp  <- psbc_obj$res_cp
  psbc_res_ncp <- psbc_obj$res_ncp
}

plot_rank_hist(psbc_res_cp,  variables = c("mu_theta", "sigma_theta")) +
  labs(title = "PSBC Rank Histograms: Centered Parameterization")

plot_rank_hist(psbc_res_ncp, variables = c("mu_theta", "sigma_theta")) +
  labs(title = "PSBC Rank Histograms: Non-Centered Parameterization")


# 1. Extract the lightweight stats dataframe from the SBC object
sbc_stats_df <- psbc_res_ncp$stats

# 2. Filter for the population-level parameters you care about
# (e.g., mu_alpha, mu_beta, rho, or even sigma[1])
recovery_df <- sbc_stats_df |>
  filter(variable %in% c("mu_theta", "sigma_theta"))

# 3. Create the Parameter Recovery Identity Plot
p_sbc_recovery <- ggplot(recovery_df, aes(x = simulated_value, y = median)) +
  # The Identity Line (Perfect Recovery)
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50", linewidth = 1) +
  
  # The Posterior Estimates with 90% Credible Intervals
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0, alpha = 0.3, color = "steelblue") +
  geom_point(color = "midnightblue", alpha = 0.7, size = 2) +
  
  # Facet by parameter
  facet_wrap(~variable, scales = "free") +
  
  labs(
    title = "Parameter Recovery (Extracted Directly from SBC)",
    subtitle = "Points represent the posterior median. Bars are 90% Credible Intervals.",
    x = "True Generative Value (Drawn from Prior)",
    y = "Posterior Median Estimate"
  ) +
  theme_classic() +
  theme(strip.background = element_rect(fill = "gray95", color = NA))

print(p_sbc_recovery)

# 4. Quantify the Recovery (Correlation and RMSE)
recovery_metrics <- recovery_df |>
  group_by(variable) |>
  summarize(
    Pearson_r = cor(simulated_value, median),
    RMSE = sqrt(mean((simulated_value - median)^2)),
    .groups = "drop"
  )

cat("=== PARAMETER RECOVERY QUANTIFICATION ===\n")
print(knitr::kable(recovery_metrics, digits = 3))

diags_psbc_cp  <- psbc_res_cp$backend_diagnostics
diags_psbc_ncp <- psbc_res_ncp$backend_diagnostics

psbc_cp_fail_rate  <- mean(diags_psbc_cp$n_divergent  > 0) * 100
psbc_ncp_fail_rate <- mean(diags_psbc_ncp$n_divergent > 0) * 100

cat("=== POSTERIOR SBC (LOCAL) DIVERGENCE RATES ===\n")
cat("CP  Failure Rate:", psbc_cp_fail_rate,  "%\n")
cat("NCP Failure Rate:", psbc_ncp_fail_rate, "%\n")
```


### Comparing CP and NCP: Global Fragility vs. Local Efficiency

The CP did not seem to perform too bad, but it broke down when doing SBC. We therefore reparameterized the model by decoupling the sampling of individual and population level parameters (from CP to NCP). This fundamentally altered how the Hamiltonian Monte Carlo (NUTS) algorithm explores the posterior manifold.

We saw that NCP took slightly longer to run and was less efficient in sampling (< ESS). We then ran extensive SBC both sampling the simulations from the priors (global calibration) and sampling the simulations from the fitted posteriors (local calibration).

By comparing global calibration (Prior SBC) against local calibration (Posterior SBC), we have mathematically mapped the boundary conditions of these parameterizations. 

* Integration Stability and Global Fragility (Divergences): In CP, sparse data forces the sampler into the infinite curvature of Neal's Funnel. As proven by our Prior SBC failure rates, CP is globally fragile; it regularly diverged across the broad generative prior space whenever the true population variance ($\sigma_\theta$) was small. NCP, conversely, decouples the parameters. The sampler explores a standard, isotropic Gaussian space ($z \sim \text{Normal}(0, 1)$) where the curvature is constant, completely immunizing the model against Neal's Funnel and dropping global divergences to zero.

* Sampling Efficiency and Local Vulnerability (ESS and Speed): Because the NCP geometry is smooth and well-behaved in data-sparse regimes, the Hamiltonian integrator can take massive, efficient step sizes, causing the Effective Sample Size (ESS) to skyrocket. However, as our Posterior SBC proved, NCP is locally vulnerable to the Data Funnel (or "Reverse Funnel"). In the specific neighborhood of our simulated data, the likelihood was highly informative and tightly constrained the posterior. CP navigated this sharp likelihood peak effortlessly. NCP, however, was forced to blindly explore a broad standard normal prior, only to find that the data violently rejected almost all of it, causing its ESS to plummet.

* Stationarity ($\hat{R}$): While both models can achieve $\hat{R} < 1.01$ under ideal local conditions, NCP achieves it reliably across the entire prior space, ensuring all Markov chains mix seamlessly regardless of the underlying data-generating universe.

* The Geometric Trade-off: When CP Wins Given the global safety of NCP, it is tempting to conclude that we should always use the Non-Centered Parameterization. This is mathematically incorrect. The choice between CP and NCP is a strict geometric trade-off governed by the density of your data.

  ** The Data-Poor Regime (Use NCP): When trial counts ($N_j$) are low or population variance ($\sigma_\theta$) is small, the prior dominates the likelihood. This creates Neal's Funnel in the CP space. NCP thrives here because it decouples the parameters, allowing the sampler to easily and safely explore the prior geometry.
  
  ** The Data-Rich Regime (Use CP): Consider a scenario where every agent completes $10,000$ trials. The likelihood becomes massively informative, crushing the uncertainty around $\theta_{\text{logit}, j}$ into a tiny, tight peak.If we use CP, the sampler effortlessly navigates this tight, data-driven likelihood space. If we use NCP, the sampler attempts to explore the broad $z \sim \text{Normal}(0,1)$ prior, only to find that almost the entire space is immediately rejected by the highly restrictive likelihood. This makes NCP highly inefficient, resulting in massive autocorrelation, low ESS, and high computational time.
  
In other words, there is no universal panacea in cognitive modeling. We do not engage in blind "tinkering" to get our models to fit. Instead, we adhere to the Iterative Bayesian Workflow: we fit the model, deploy a strict diagnostic battery to identify the exact geometric failure (prior-dominated funnels vs. data-dominated funnels), and apply the mathematically appropriate parameterization.

Because typical cognitive tasks (like our Matching Pennies game) involve relatively sparse data ($N_j \approx 100$) and substantial population-level shrinkage, we will utilize the Non-Centered Parameterization as our default architecture as we proceed to construct the more complex Memory and WSLS agents.

## The Epistemic Showdown: Visualizing Shrinkage

We have theorized about the dangers of ignoring population structure (Complete Pooling) and the dangers of ignoring noise in small samples (No Pooling). In a rigorous Bayesian workflow, we do not rely on intuition; we compute the alternatives and visualize the geometric differences.We will quickly define and fit the two extreme models using our strictly typed stan_data_centered. N.B. we should technically assess in detail all the models, but we are skipping that for brevity. The point is to visualize the shrinkage dynamics of our mathematically robust NCP model against these two extremes.

### The Complete Pooling Model (Underfitting)

This model assumes all agents are identical clones. The population variance is forced to exactly $0$. We estimate a single, universal bias parameter.

```{r}
stan_full_pooling_code <- "
data {
  int<lower=1> N;
  array[N] int<lower=0, upper=1> h;
}
parameters {
  real theta_logit; // A single parameter for the entire universe
}
model {
  target += normal_lpdf(theta_logit | 0, 1.5);
  target += bernoulli_logit_lpmf(h | theta_logit);
}
"
stan_file_full <- here::here("stan", "ch6_biased_full_pooling.stan")
writeLines(stan_full_pooling_code, stan_file_full)
mod_full <- cmdstan_model(stan_file_full)

fit_full_path <- here::here("simmodels", "ch6_fit_full_pool.rds")

if (regenerate_simulations || !file.exists(fit_full_path)) {
  fit_full <- mod_full$sample(
    data            = list(N = stan_data_centered$N, h = stan_data_centered$h),
    seed            = 123, chains = 4, parallel_chains = 4, refresh = 0
  )
  fit_full$save_object(fit_full_path)
} else {
  fit_full <- readRDS(fit_full_path)
}

full_pool_est <- posterior::summarise_draws(fit_full$draws("theta_logit"), "median")$median
```

### The No Pooling Model (Overfitting)

This model assumes every agent is a completely isolated island. The population variance is implicitly assumed to be $\infty$. We estimate an independent parameter for each agent, equipped only with a static, weakly informative prior.

```{r}
stan_no_pooling_code <- "
data {
  int<lower=1> N;
  int<lower=1> J;
  array[N] int<lower=1, upper=J> agent;
  array[N] int<lower=0, upper=1> h;
}
parameters {
  vector[J] theta_logit; // J completely independent parameters
}
model {
  // Static prior: No hierarchical learning
  target += normal_lpdf(theta_logit | 0, 1.5);
  target += bernoulli_logit_lpmf(h | theta_logit[agent]);
}
"
stan_file_no <- here::here("stan", "ch6_biased_no_pooling.stan")
writeLines(stan_no_pooling_code, stan_file_no)
mod_no <- cmdstan_model(stan_file_no)

# Fit the No Pooling model
fit_no_path <- here::here("simmodels", "ch6_fit_no_pool.rds")

if (regenerate_simulations || !file.exists(fit_no_path)) {
  fit_no <- mod_no$sample(
    data            = stan_data_centered,
    seed            = 123, chains = 4, parallel_chains = 4, refresh = 0
  )
  fit_no$save_object(fit_no_path)
} else {
  fit_no <- readRDS(fit_no_path)
}

no_pool_summary <- posterior::summarise_draws(fit_no$draws("theta_logit"), "median") |> 
  mutate(agent_id = as.integer(str_extract(variable, "(?<=\\[)\\d+(?=\\])"))) |> 
  rename(theta_no_pool = median)
```

### The Geometry of Shrinkage

We now have three distinct sets of inferences:
* No Pooling: The raw, isolated estimates (highly vulnerable to the noise of low $N_j$).
* Complete Pooling: The rigid, universal average (blind to genuine individual differences).
* Partial Pooling: Our hierarchical NCP model estimates (the precision-weighted Bayesian compromise).

To truly evaluate the epistemological value of these models, we must plot them not just against each other, but against the True Generative Parameters (the hidden reality we simulated in R). We will also extract the naive arithmetic mean of the No Pooling estimates to contrast it with the precision-weighted Partial Pooling mean.
```{r}
# 1. Extract the Partial Pooling Population and Individual Estimates (from the NCP model)
mu_partial_pool_est <- posterior::summarise_draws(fit_ncp$draws("mu_theta"), "median")$median

partial_pool_summary <- posterior::summarise_draws(fit_ncp$draws("theta_logit"), "median") |> 
  mutate(agent_id = as.integer(str_extract(variable, "(?<=\\[)\\d+(?=\\])"))) |> 
  rename(theta_partial_pool = median)

# 2. Calculate the No Pooling Mean (simple arithmetic average of the isolated estimates)
no_pool_mean_est <- mean(no_pool_summary$theta_no_pool)

# 3. Calculate the True Generative Mean (The actual center of our simulated universe)
true_mean_est <- mean(true_agent_params$theta_logit)

# 4. Combine into a single dataframe, including the TRUE generative parameters
comparison_df <- no_pool_summary |> 
  select(agent_id, theta_no_pool) |> 
  left_join(
    partial_pool_summary |> select(agent_id, theta_partial_pool), 
    by = "agent_id"
  ) |> 
  # Bring in the ground truth
  left_join(
    true_agent_params |> select(agent_id, n_trials, theta_true = theta_logit),
    by = "agent_id"
  ) |> 
  # Sort agents by their No Pooling estimate for a cleaner visual curve
  arrange(theta_no_pool) |> 
  mutate(rank = row_number())

# The Ultimate Epistemic Plot
ggplot(comparison_df) +
  # --- POPULATION AVERAGES (Mapped to linetype for a clean legend) ---
  geom_hline(aes(yintercept = no_pool_mean_est, linetype = "No Pooling Mean"), 
             color = "darkorange", linewidth = 0.8) +
  geom_hline(aes(yintercept = full_pool_est, linetype = "Complete Pooling Mean"), 
             color = "darkred", linewidth = 0.8) +
  geom_hline(aes(yintercept = mu_partial_pool_est, linetype = "Partial Pooling Mean (\U03BC)"), 
             color = "steelblue", linewidth = 1.2) +
  geom_hline(aes(yintercept = true_mean_est, linetype = "True Generative Mean"), 
             color = "black", linewidth = 0.8) +
  
  # --- INDIVIDUAL ESTIMATES ---
  # Shrinkage Vectors (from No Pool to Partial Pool)
  geom_segment(aes(x = rank, xend = rank, 
                   y = theta_no_pool, yend = theta_partial_pool), 
               color = "gray65", arrow = arrow(length = unit(0.1, "cm"))) +
  
  # True Generative Parameters (Black Crosses)
  geom_point(aes(x = rank, y = theta_true, color = "True Value"), 
             shape = 4, size = 2, stroke = 1) +
             
  # No Pooling Estimates (Open Circles)
  geom_point(aes(x = rank, y = theta_no_pool, color = "No Pooling", size = n_trials), 
             shape = 1, stroke = 1.2) +
             
  # Partial Pooling Estimates (Solid Points)
  geom_point(aes(x = rank, y = theta_partial_pool, color = "Partial Pooling", size = n_trials), 
             alpha = 0.8) +
  
  # --- SCALES & THEMES ---
  scale_linetype_manual(
    name = "Population Averages",
    values = c(
      "True Generative Mean" = "solid",
      "Partial Pooling Mean (\U03BC)" = "dotted",
      "Complete Pooling Mean" = "dashed",
      "No Pooling Mean" = "dotdash"
    )
  ) +
  scale_color_manual(
    name = "Individual Estimates",
    values = c(
      "True Value" = "black",
      "No Pooling" = "darkorange", 
      "Partial Pooling" = "steelblue"
    )
  ) +
  labs(
    title = "The Epistemic Compromise: Visualizing Bayesian Shrinkage",
    x = "Agents (Sorted by No-Pooling Estimate)",
    y = "Estimated Bias Parameter (Logit Scale)",
    size = "Trial Count (N_j)"
  ) +
  theme(
    axis.text.x = element_blank(), 
    axis.ticks.x = element_blank(),
    legend.position = "right"
  )
```

Study this plot carefully. It contains the entire epistemological justification for hierarchical modeling, visible across two dimensions: the horizontal population averages and the individual agent estimates.

### The Horizontal Lines: The Quartet of Population Means

* True Generative Mean (Solid Black): The actual center of our simulated universe. This is the structural reality we are trying to recover.

* No Pooling Mean (Dot-dash Orange): The unweighted arithmetic average of the isolated estimates. Because it treats all agents equally, it is highly susceptible to being dragged away from the truth by extreme outliers with very little data.

* Complete Pooling Mean (Dashed Red): The unweighted average of all trials. Because it ignores agent boundaries, it allows participants who completed massive numbers of trials to disproportionately pull the center of gravity toward themselves.

* Partial Pooling Mean ($\mu_\theta$, Dotted Blue): The precision-weighted compromise. By mathematically down-weighting the noisy, low-$N_j$ agents, it reliably lands closest to the true generative mean.

### The Points and Arrows: Information-Weighted Regularization

* The Overfitting Extremes: Look at the tails of the graph. The orange open circles (No Pooling) are extreme. If we trusted No Pooling, we would conclude these agents have massive, almost deterministic biases.

* The Power of the Prior: The solid blue dots (Partial Pooling) are systematically pulled inward. The hierarchical model "knows" that extremely high or low values are mathematically unlikely given the behavior of the rest of the cohort.

* The Calibration of Skepticism: Look at the size of the dots, representing the trial count ($N_j$). The longest arrows (most aggressive shrinkage) almost exclusively belong to the smallest dots. When an agent has sparse data, the posterior relies heavily on the population prior. The largest dots barely move; when an agent has massive amounts of data, the likelihood overwhelms the prior, allowing them to remain safely extreme.

## The Memory Agent: Two Correlated Individual Differences

### Introduction

The Biased Agent established the full Bayesian workflow for a single individual-level parameter: simulate, fit, diagnose, recover, calibrate. It also taught us, the hard way, that the Centered Parameterization fails when population variance is small, and that the Non-Centered Parameterization resolves this by decoupling individual and population parameters geometrically. We proceed to the Memory Agent with both of those lessons in hand.

The Memory Agent is structurally richer. Recall from Chapter 3 its decision rule. On each trial $i$, agent $j$ maintains a running average of the opponent's past choices $m_{ij}$ and uses it as a predictive signal:

$$h_{ij} \sim \text{Bernoulli}(\text{logit}^{-1}(\alpha_j + \beta_j \cdot \text{logit}(m_{ij})))$$

$$m_{ij} = \frac{1}{i-1}\sum_{k=1}^{i-1} o_{kj}$$

There are now two individual-level parameters per agent: $\alpha_j$, the baseline directional bias, and $\beta_j$, the sensitivity to opponent history. They need not vary independently across the population. An agent who has a strong directional tendency might also track the opponent more aggressively — or might not. Fitting two separate univariate hierarchical models forecloses this question before the data can answer it. We model the parameters jointly.

#### The Bivariate Population Distribution

$$\begin{pmatrix} \alpha_j \\ \beta_j \end{pmatrix} \sim \mathcal{N}_2\!\left(\boldsymbol{\mu},\, \boldsymbol{\Sigma}\right), \quad \boldsymbol{\Sigma} = \operatorname{diag}(\boldsymbol{\sigma})\, \Omega\, \operatorname{diag}(\boldsymbol{\sigma}), \quad \Omega = \begin{pmatrix} 1 & \rho \\ \rho & 1 \end{pmatrix}$$

This decomposition separates how much individuals differ on each cognitive dimension ($\sigma_\alpha$, $\sigma_\beta$) from how those dimensions covary ($\rho$). The scalar $\rho \in (-1, 1)$ is the key new inferential target.

#### The LKJ Prior

We need a prior on the correlation matrix $\Omega$. Stan provides the LKJ distribution (Lewandowski, Kurowicka, and Joe, 2009), with a single shape parameter $\eta > 0$:

$$\Omega \sim \text{LKJ}(\eta) \quad \propto \quad (\det \Omega)^{\eta - 1}$$

When $\eta = 1$ the distribution is uniform over valid correlation matrices. When $\eta > 1$ mass concentrates toward the identity, encoding the prior belief that parameters tend to vary independently unless the data argue otherwise. For a $2 \times 2$ matrix, the marginal on $\rho$ reduces to a rescaled Beta: $\rho \sim 2 \cdot \text{Beta}(\eta, \eta) - 1$. With our default $\eta = 2$, the prior is unimodal at zero and weakly regularizes against extreme correlations.

Stan samples the lower-triangular **Cholesky factor** $\mathbf{L}_\Omega$ (declared as `cholesky_factor_corr[2]`) under the `lkj_corr_cholesky(eta)` prior, recovering $\Omega = \mathbf{L}_\Omega \mathbf{L}_\Omega^\top$ in `generated quantities`. This parameterization is unconstrained on its natural manifold.

#### Non-Centered Parameterization

The scalar NCP from the biased agent ($\theta_j = \mu + z_j \cdot \sigma$) generalizes directly. We introduce a $2 \times J$ matrix of independent standard normals:

$$z_{kj} \sim \mathcal{N}(0, 1) \text{ i.i.d.}, \qquad \begin{pmatrix}\alpha_j \\ \beta_j\end{pmatrix} = \begin{pmatrix}\mu_\alpha \\ \mu_\beta\end{pmatrix} + \operatorname{diag}(\boldsymbol{\sigma})\,\mathbf{L}_\Omega\,\mathbf{z}_j$$

The sampler traverses the flat, isotropic space of $\mathbf{Z}$, whose geometry is entirely independent of $\boldsymbol{\sigma}$ and $\Omega$. Neal's Funnel is structurally impossible.

```{tikz hierarchical_memory_dag, echo=FALSE, fig.cap="Standard Plate Notation (left) and Kruschke-Style Notation (right) for the Hierarchical Memory Agent. The correlation node Omega and the bivariate MVN distinguish this model from the biased agent.", fig.ext='png', fig.width=12, cache=TRUE}
\usetikzlibrary{positioning, shapes.geometric, arrows.meta, fit, backgrounds, calc}
\begin{tikzpicture}[
    stochastic/.style={->, >={Latex[length=3mm]}, thick},
    deterministic/.style={->, >={Latex[length=3mm]}, thick, dashed, draw=gray!70},
    plate/.style={draw=black, thick, rounded corners, inner sep=10pt},
    continuous/.style={circle, draw=black, thick, minimum size=1.2cm, inner sep=0pt},
    discrete/.style={rectangle, draw=black, thick, minimum size=1.2cm},
    observed/.style={fill=gray!30},
    latent/.style={fill=white},
    pics/norm_dist/.style={code={
        \draw[thin, gray] (-1.2,0) -- (1.2,0);
        \draw[thick, blue] plot[domain=-1.2:1.2, samples=50] (\x, {0.8*exp(-3*\x*\x)});
        \node[font=\footnotesize, below] at (0,0) {Normal};
    }},
    pics/exp_dist/.style={code={
        \draw[thin, gray] (-0.2,0) -- (1.5,0);
        \draw[thick, blue] plot[domain=0:1.4, samples=50] (\x, {0.8*exp(-2.5*\x)});
        \node[font=\footnotesize, below] at (0.6,0) {Exponential};
    }},
    pics/lkj_dist/.style={code={
        \draw[thin, gray] (-1.2,0) -- (1.2,0);
        \draw[thick, blue] plot[domain=-0.95:0.95, samples=50] (\x, {0.8*(1-\x*\x)});
        \node[font=\footnotesize, below] at (0,0) {LKJ};
    }},
    pics/mvn_dist/.style={code={
        \draw[thin, gray, ->] (-0.8,0) -- (0.8,0);
        \draw[thin, gray, ->] (0,-0.8) -- (0,0.8);
        \draw[thick, blue, rotate=30] (0,0) ellipse (0.6 and 0.22);
        \draw[thick, blue, rotate=30] (0,0) ellipse (0.3 and 0.11);
        \node[font=\footnotesize, below right] at (0.1,-0.15) {MVN};
    }},
    pics/bern_dist/.style={
        code={
            \draw[thin, gray] (-1,0) -- (1,0);
            % Replaced ycomb plot with explicit thick lines for compiler stability
            \draw[blue, line width=4pt] (-0.4, 0) -- (-0.4, 0.3);
            \draw[blue, line width=4pt] (0.4, 0) -- (0.4, 0.8);
            \node[font=\footnotesize, below] at (0,0) {Bernoulli};
        }
    }
]

%% LEFT: STANDARD PLATE NOTATION
\node[font=\large\bfseries] at (0, 2.2) {Standard Plate Notation};

\node[continuous, latent] (mu_a)  at (-3,   0) {$\mu_\alpha$};
\node[continuous, latent] (sig_a) at (-1.5, 0) {$\sigma_\alpha$};
\node[continuous, latent, fill=gray!15] (omega) at (0, 0) {$\Omega$};
\node[continuous, latent] (sig_b) at (1.5,  0) {$\sigma_\beta$};
\node[continuous, latent] (mu_b)  at (3,    0) {$\mu_\beta$};

\node[continuous, latent] (alpha) at (-1.5, -3.5) {$\alpha_j$};
\node[continuous, latent] (beta)  at ( 1.5, -3.5) {$\beta_j$};

\node[discrete, observed] (m) at (-1.5, -7) {$m_{ij}$};
\node[discrete, observed] (h) at ( 1.5, -7) {$h_{ij}$};

\node[plate, fit=(m)(h), label={[anchor=south east, font=\small\bfseries]below right:$i=1\dots N$}] (pN) {};
\node[plate, inner sep=15pt, fit=(alpha)(beta)(pN), label={[anchor=south east, font=\small\bfseries]below right:$j=1\dots J$}] {};

\draw[stochastic] (mu_a) -- (alpha); \draw[stochastic] (sig_a) -- (alpha);
\draw[stochastic] (omega) -- (alpha); \draw[stochastic] (omega) -- (beta);
\draw[stochastic] (sig_b) -- (beta); \draw[stochastic] (mu_b) -- (beta);
\draw[stochastic] (alpha) -- (h);    \draw[stochastic] (beta) -- (h);
\draw[deterministic] (m) -- (h);

%% RIGHT: KRUSCHKE-STYLE NOTATION
\node[font=\large\bfseries] at (11, 2.2) {Kruschke-Style Notation};

\coordinate (k_mu_a)  at (7,  0); \draw (k_mu_a)  pic {norm_dist}; \node[above=0.7cm of k_mu_a,  font=\bfseries] {$\mu_\alpha$};
\coordinate (k_sig_a) at (9,  0); \draw (k_sig_a) pic {exp_dist};  \node[above=0.7cm of k_sig_a, font=\bfseries] {$\sigma_\alpha$};
\coordinate (k_omega) at (11, 0); \draw (k_omega) pic {lkj_dist};  \node[above=0.7cm of k_omega, font=\bfseries] {$\Omega$};
\coordinate (k_sig_b) at (13, 0); \draw (k_sig_b) pic {exp_dist};  \node[above=0.7cm of k_sig_b, font=\bfseries] {$\sigma_\beta$};
\coordinate (k_mu_b)  at (15, 0); \draw (k_mu_b)  pic {norm_dist}; \node[above=0.7cm of k_mu_b,  font=\bfseries] {$\mu_\beta$};

\coordinate (k_mvn) at (11, -3.5); \draw (k_mvn) pic {mvn_dist};
\node[above=0.9cm of k_mvn, font=\bfseries, align=center] {Agent Params \\ $(\alpha_j,\,\beta_j)$};

\node[discrete, observed, minimum size=0.9cm] (k_m) at (9, -7) {$m_{ij}$};
\node[above=0.5cm of k_m, font=\bfseries] {Memory};
\coordinate (k_h) at (13, -7); \draw (k_h) pic {bern_dist};
\node[above=0.9cm of k_h, font=\bfseries] {Choice ($h_{ij}$)};

\draw[plate, dashed]  (7.5, -5.2) rectangle (14.5, -2.2) node[below left, font=\small\bfseries] {$j=1\dots J$};
\draw[plate, dotted]  (7.5, -8.2) rectangle (14.5, -5.8) node[below left, font=\small\bfseries] {$i=1\dots N$};

\draw[stochastic] (k_mu_a) -- (k_mvn); \draw[stochastic] (k_sig_a) -- (k_mvn);
\draw[stochastic] (k_omega) -- (k_mvn) node[midway, right, font=\large] {$\sim$};
\draw[stochastic] (k_sig_b) -- (k_mvn); \draw[stochastic] (k_mu_b) -- (k_mvn);
\draw[stochastic] (k_mvn) -- (k_h) node[midway, right, font=\large] {$\sim$};
\draw[deterministic] (k_m) -- (12.5, -7);
\end{tikzpicture}
```

### Phase 1: Generative Plausibility of the Memory Agent

Before writing any Stan code, we construct the forward generative model in R. This code must strictly mirror the mathematical hierarchy above.

One design decision deserves explicit justification. The memory signal $m_{ij}$ is a deterministic function of the opponent's past choices. We could compute it inside Stan's `transformed data` block, but this executes a nested loop on every leapfrog step of every chain. Instead, we pre-compute the vector `opp_rate_prev` in R once and pass it as data. This reduces the memory calculation to a single vectorized pass before fitting begins, at zero loss of model fidelity.

The simulation function must also mirror the NCP Stan model exactly. We draw $\mathbf{z}_j \sim \mathcal{N}(\mathbf{0}, \mathbf{I}_2)$ and apply the Cholesky transform, exactly as the Stan model will. For a $2 \times 2$ correlation matrix with off-diagonal $\rho$, the lower Cholesky factor is:

$$\mathbf{L}_\Omega = \begin{pmatrix} 1 & 0 \\ \rho & \sqrt{1-\rho^2} \end{pmatrix}$$

We sample $\rho$ from the LKJ(2) marginal: $\rho \sim 2 \cdot \text{Beta}(2, 2) - 1$.

In accordance with the rigorous Bayesian workflow, the simulation function is designed to either:

* Draw all hyperparameters directly from the stated priors (for Prior Predictive Checks).

* Accept fixed ground-truth hyperparameters (for Parameter Recovery).

```{r}
#' Simulate Hierarchical Memory Agents
#'
#' @param n_agents Integer, number of agents (J).
#' @param n_trials_min Integer, minimum trials per agent.
#' @param n_trials_max Integer, maximum trials per agent.
#' @param mu_alpha Numeric, population mean baseline bias (logit scale). NULL draws from prior.
#' @param mu_beta Numeric, population mean memory sensitivity. NULL draws from prior.
#' @param sigma_alpha Numeric, population SD of alpha. NULL draws from prior.
#' @param sigma_beta Numeric, population SD of beta. NULL draws from prior.
#' @param rho Numeric, population correlation alpha-beta. NULL draws from LKJ(2).
#' @param opponent_rate Numeric, true probability of opponent choosing "right".
#' @param seed Integer, for exact computational reproducibility.
#'
#' @return A tidy tibble containing both latent parameters and observed choices.
simulate_hierarchical_memory <- function(
  n_agents      = 50,
  n_trials_min  = 20,
  n_trials_max  = 120,
  mu_alpha      = NULL,
  mu_beta       = NULL,
  sigma_alpha   = NULL,
  sigma_beta    = NULL,
  rho           = NULL,
  opponent_rate = 0.7,
  seed          = NULL
) {
  if (!is.null(seed)) set.seed(seed)

  # 1. Sample hyperparameters from priors if not provided
  if (is.null(mu_alpha))    mu_alpha    <- rnorm(1, 0, 1)
  if (is.null(mu_beta))     mu_beta     <- rnorm(1, 0, 1)
  if (is.null(sigma_alpha)) sigma_alpha <- rexp(1, 1)
  if (is.null(sigma_beta))  sigma_beta  <- rexp(1, 1)
  # LKJ(2) marginal on rho for 2x2 = 2*Beta(2,2) - 1
  if (is.null(rho))         rho         <- 2 * rbeta(1, 2, 2) - 1

  # 2. Construct the lower-triangular Cholesky factor L_Omega
  L_Omega <- matrix(c(1, rho, 0, sqrt(1 - rho^2)), nrow = 2, ncol = 2)

  # 3. Sample individual parameters via the NCP transform (mirrors Stan exactly)
  # Z is 2 x J; Scale = diag(sigma) %*% L_Omega
  Z            <- matrix(rnorm(2 * n_agents), nrow = 2)
  Scale        <- diag(c(sigma_alpha, sigma_beta)) %*% L_Omega
  # indiv_params is J x 2: column 1 = alpha_j, column 2 = beta_j
  indiv_params <- t(c(mu_alpha, mu_beta) + Scale %*% Z)

  # 4. Simulate trial-level data
  n_trials_vec <- sample(n_trials_min:n_trials_max, n_agents, replace = TRUE)

  sim_data <- map_dfr(seq_len(n_agents), function(j) {
    n_t     <- n_trials_vec[j]
    alpha_j <- indiv_params[j, 1]
    beta_j  <- indiv_params[j, 2]

    opp_choices <- rbinom(n_t, 1, opponent_rate)

    # Pre-compute running opponent rate BEFORE each trial (strict t-1 history).
    # Trial 1 has no history: initialize to 0.5 (uninformative).
    # Clip to [0.01, 0.99] so logit(opp_rate_prev) is always finite.
    opp_rate_prev    <- numeric(n_t)
    opp_rate_prev[1] <- 0.5
    if (n_t > 1) {
      raw_rates         <- cumsum(opp_choices[-n_t]) / seq_len(n_t - 1)
      opp_rate_prev[-1] <- pmax(0.01, pmin(0.99, raw_rates))
    }

    logit_p       <- alpha_j + beta_j * qlogis(opp_rate_prev)
    agent_choices <- rbinom(n_t, 1, plogis(logit_p))

    tibble(
      agent_id      = j,
      trial         = seq_len(n_t),
      n_trials      = n_t,
      choice        = agent_choices,
      opp_rate_prev = opp_rate_prev,
      true_alpha    = alpha_j,
      true_beta     = beta_j
    )
  })

  attr(sim_data, "hyperparameters") <- list(
    mu_alpha    = mu_alpha,    mu_beta     = mu_beta,
    sigma_alpha = sigma_alpha, sigma_beta  = sigma_beta,
    rho         = rho
  )

  sim_data
}
```

#### Executing the Prior Predictive Check: The Joint Population Cloud

For a bivariate model, the most revealing prior predictive display is a scatter of individual $(\alpha_j, \beta_j)$ pairs across multiple prior universes. This simultaneously reveals how dispersed individual differences are ($\sigma_\alpha$, $\sigma_\beta$), how correlated the two cognitive dimensions tend to be ($\rho$), and whether any prior universe generates implausible populations (e.g., $|\beta_j| > 8$, meaning the memory signal completely overrides choice).

```{r}
# Generate 12 prior universes
prior_sims_mem <- map_dfr(1:12, function(sim_id) {
  sim <- simulate_hierarchical_memory(
    n_agents = 80, n_trials_min = 1, n_trials_max = 1,
    seed = sim_id * 13
  )
  hp <- attr(sim, "hyperparameters")
  sim |>
    distinct(agent_id, true_alpha, true_beta) |>
    mutate(simulation_id = paste("Prior Universe", sim_id))
})

ggplot(prior_sims_mem, aes(x = true_alpha, y = true_beta)) +
  geom_point(alpha = 0.4, size = 0.9, color = "steelblue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60", linewidth = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60", linewidth = 0.4) +
  facet_wrap(~simulation_id, ncol = 4) +
  labs(
    title    = "Prior Predictive Checks: Implied Joint Population Distributions",
    subtitle = "Populations generated by mu ~ N(0,1), sigma ~ Exp(1), Omega ~ LKJ(2)",
    x        = expression("Baseline bias  " * alpha[j] * "  (logit scale)"),
    y        = expression("Memory sensitivity  " * beta[j])
  )
```

A healthy prior generates a range of cloud shapes — some circular (low $|\rho|$), some elongated and tilted (high $|\rho|$), some compact (low $\sigma$), some diffuse (high $\sigma$). No prior universe should collapse all agents to a point or place all agents at implausibly extreme values.

### Phase 2: Data Preparation and the NCP Stan Model

Having established from the biased agent workflow that NCP is our default architecture for cognitive tasks with sparse per-agent data, we implement it directly. We skip the Centered Parameterization entirely.

```{r, 06_stan_memory_data_prep }
#' Prepare Memory Agent Data for Stan
#'
#' @param sim_data Tibble from simulate_hierarchical_memory().
#' @return A named list formatted for CmdStanR.
prep_stan_data_memory <- function(sim_data) {
  list(
    N             = nrow(sim_data),
    J             = length(unique(sim_data$agent_id)),
    agent         = sim_data$agent_id,
    h             = sim_data$choice,
    opp_rate_prev = sim_data$opp_rate_prev
  )
}

# Execute the preparation on a specifically generated universe for recovery
sim_data_memory <- simulate_hierarchical_memory(
  n_agents      = 50,
  n_trials_min  = 20,
  n_trials_max  = 120,
  mu_alpha      =  0.0,  # No systematic directional baseline
  mu_beta       =  1.5,  # Moderate positive memory tracking
  sigma_alpha   =  0.3,
  sigma_beta    =  0.5,
  rho           =  0.3,  # Mild positive correlation
  opponent_rate =  0.7,
  seed          = 1981
)

stan_data_memory <- prep_stan_data_memory(sim_data_memory)
```

#### Stan Model: Hierarchical Memory Agent (NCP)

The only new structural element relative to the biased agent NCP is the bivariate extension. Instead of a scalar $z_j$, we have a $2 \times J$ matrix $\mathbf{Z}$. Instead of scaling by a single $\sigma$, we apply `diag_pre_multiply(sigma, L_Omega)` — which computes $\operatorname{diag}(\boldsymbol{\sigma})\mathbf{L}_\Omega$ and produces the full Cholesky scale matrix encoding both marginal SDs and the correlation structure.

```{r, 06_stan_memory_ncp_code }
stan_memory_ncp_code <- "
// Hierarchical Memory Agent Model (Non-Centered Parameterization)
// -------------------------------------------------------------------
// alpha_j (baseline bias) and beta_j (memory sensitivity) are jointly
// drawn from a bivariate Normal. NCP decouples sampling geometry from
// the population parameters, preventing Neal's Funnel.

data {
  int<lower=1> N;                          // Total observations across all agents
  int<lower=1> J;                          // Number of agents
  array[N] int<lower=1, upper=J>  agent;   // Agent index per trial
  array[N] int<lower=0, upper=1>  h;       // Observed binary choices
  // Pre-computed running average of opponent choices, clipped to [0.01, 0.99].
  vector<lower=0.01, upper=0.99>[N] opp_rate_prev;
}

parameters {
  // Population-level parameters
  real mu_alpha;              // Population mean baseline bias (logit scale)
  real mu_beta;               // Population mean memory sensitivity
  vector<lower=0>[2] sigma;   // Marginal SDs: sigma[1] = sigma_alpha, sigma[2] = sigma_beta

  // Cholesky factor of the 2x2 correlation matrix. Prior: Omega ~ LKJ(2).
  cholesky_factor_corr[2] L_Omega;

  // Individual-level parameters (NCP): 2 x J matrix of standard normals.
  // Row 1 = alpha offsets, row 2 = beta offsets.
  matrix[2, J] z;
}

transformed parameters {
  // Reconstruct cognitive parameters from z-space via the Cholesky transform.
  // diag_pre_multiply(sigma, L_Omega) = diag(sigma) * L_Omega  (2x2 matrix)
  // Multiplying by z (2xJ) and transposing gives J x 2.
  matrix[J, 2] indiv_params;
  {
    matrix[2, J] deviations = diag_pre_multiply(sigma, L_Omega) * z;
    for (j in 1:J) {
      indiv_params[j, 1] = mu_alpha + deviations[1, j];
      indiv_params[j, 2] = mu_beta  + deviations[2, j];
    }
  }
}

model {
  // Hyperpriors
  target += normal_lpdf(mu_alpha | 0, 1);
  target += normal_lpdf(mu_beta  | 0, 1);
  target += exponential_lpdf(sigma | 1);
  target += lkj_corr_cholesky_lpdf(L_Omega | 2);

  // Individual-level prior (NCP): geometry fully decoupled from population params
  target += std_normal_lpdf(to_vector(z));

  // Likelihood
  vector[N] logit_p;
  for (i in 1:N) {
    logit_p[i] = indiv_params[agent[i], 1]
               + indiv_params[agent[i], 2] * logit(opp_rate_prev[i]);
  }
  target += bernoulli_logit_lpmf(h | logit_p);
}

generated quantities {
  // Recover the full correlation matrix and the scalar rho
  matrix[2, 2] Omega = multiply_lower_tri_self_transpose(L_Omega);
  real rho = Omega[1, 2];

  // Individual parameters on natural scales
  vector[J] alpha    = indiv_params[, 1];
  vector[J] beta_mem = indiv_params[, 2];

  // Prior log-density (required by priorsense)
  real lprior = normal_lpdf(mu_alpha | 0, 1)
              + normal_lpdf(mu_beta  | 0, 1)
              + exponential_lpdf(sigma | 1)
              + lkj_corr_cholesky_lpdf(L_Omega | 2)
              + std_normal_lpdf(to_vector(z));

  // Prior draws for prior-posterior update plots
  real mu_alpha_prior    = normal_rng(0, 1);
  real mu_beta_prior     = normal_rng(0, 1);
  real sigma_alpha_prior = exponential_rng(1);
  real sigma_beta_prior  = exponential_rng(1);
  // LKJ(2) marginal on rho: 2*Beta(2,2) - 1
  real rho_prior         = 2.0 * beta_rng(2.0, 2.0) - 1.0;

  // Predictive checks and pointwise log-likelihood
  vector[N] log_lik;
  array[N] int h_post_rep;
  array[N] int h_prior_rep;

  // Build a prior-drawn population for h_prior_rep
  cholesky_factor_corr[2] L_Omega_prior = lkj_corr_cholesky_rng(2, 2.0);
  vector[2] sigma_prior = [sigma_alpha_prior, sigma_beta_prior]';
  matrix[2, J] z_prior;
  for (j in 1:J) { z_prior[1,j] = normal_rng(0,1); z_prior[2,j] = normal_rng(0,1); }
  matrix[J, 2] indiv_prior;
  {
    matrix[2, J] dev_prior = diag_pre_multiply(sigma_prior, L_Omega_prior) * z_prior;
    for (j in 1:J) {
      indiv_prior[j, 1] = mu_alpha_prior + dev_prior[1, j];
      indiv_prior[j, 2] = mu_beta_prior  + dev_prior[2, j];
    }
  }

  for (i in 1:N) {
    int j = agent[i];
    real lp_post  = indiv_params[j,1] + indiv_params[j,2] * logit(opp_rate_prev[i]);
    real lp_prior = indiv_prior[j,1]  + indiv_prior[j,2]  * logit(opp_rate_prev[i]);

    log_lik[i]     = bernoulli_logit_lpmf(h[i] | lp_post);
    h_post_rep[i]  = bernoulli_logit_rng(lp_post);
    h_prior_rep[i] = bernoulli_logit_rng(lp_prior);
  }
}
"

stan_file_memory_ncp <- here::here("stan", "ch6_multilevel_memory_ncp.stan")
writeLines(stan_memory_ncp_code, stan_file_memory_ncp)
mod_memory_ncp <- cmdstan_model(stan_file_memory_ncp)
```

Two Stan details are worth noting. First, the local block `{ matrix[2,J] deviations = ...; }` in `transformed parameters` confines the intermediate matrix to local scope, preventing it from appearing in the parameter output. Second, `lkj_corr_cholesky_rng(2, 2.0)` in `generated quantities` draws a full Cholesky factor from the LKJ(2) prior. This is the correct way to generate prior predictive data — unlike the Beta marginal trick, it generalizes to the $3 \times 3$ case we will need for the WSLS agent.

#### Fitting the model

```{r 06_Fit_Memory_NCP}
fit_memory_path <- here::here("simmodels", "ch6_fit_memory_ncp.rds")

# Initialize at prior means. L_Omega starts at the identity Cholesky (rho = 0).
# This is a neutral, numerically stable starting point for the correlation geometry.
J_val_mem <- stan_data_memory$J
  
  # A 4-chain dumb list for the Memory Model
  init_memory_list <- list(
    list(mu_alpha = rnorm(1, 0, 0.3), mu_beta = rnorm(1, 0, 0.3), sigma = abs(rnorm(2, 0, 0.3)) + 0.1, L_Omega = matrix(c(1, 0, 0, 1), nrow=2), z = matrix(0, nrow=2, ncol=J_val_mem)),
    list(mu_alpha = rnorm(1, 0, 0.3), mu_beta = rnorm(1, 0, 0.3), sigma = abs(rnorm(2, 0, 0.3)) + 0.1, L_Omega = matrix(c(1, 0, 0, 1), nrow=2), z = matrix(0, nrow=2, ncol=J_val_mem)),
    list(mu_alpha = rnorm(1, 0, 0.3), mu_beta = rnorm(1, 0, 0.3), sigma = abs(rnorm(2, 0, 0.3)) + 0.1, L_Omega = matrix(c(1, 0, 0, 1), nrow=2), z = matrix(0, nrow=2, ncol=J_val_mem)),
    list(mu_alpha = rnorm(1, 0, 0.3), mu_beta = rnorm(1, 0, 0.3), sigma = abs(rnorm(2, 0, 0.3)) + 0.1, L_Omega = matrix(c(1, 0, 0, 1), nrow=2), z = matrix(0, nrow=2, ncol=J_val_mem))
  )

if (regenerate_simulations || !file.exists(fit_memory_path)) {
  fit_memory <- mod_memory_ncp$sample(
    data            = stan_data_memory,
    seed            = 123,
    chains          = 4,
    parallel_chains = 4,
    iter_warmup     = 1000,
    iter_sampling   = 1000,
    init            = init_memory_list, 
    adapt_delta     = 0.90,   
    max_treedepth   = 12,
    refresh         = 0
  )
  fit_memory$save_object(fit_memory_path)
} else {
  fit_memory <- readRDS(fit_memory_path)
}
```

We raise `adapt_delta` to 0.90. The bivariate NCP is well-behaved globally, but the correlation parameter $\rho$ introduces a curved direction that can require longer leapfrog trajectories. The higher target acceptance rate forces the dual-averaging algorithm to find a smaller step size, reducing integration error along this direction. If `max_treedepth` warnings appear (distinct from divergences), raise `max_treedepth` further — do not raise `adapt_delta` in response to those.

#### Assessing the fitting process

Even if the Stan compiler finishes without explicit warnings, our workflow demands that we systematically verify the geometry of the posterior and the efficiency of the sampler.

```{r 06-memory-cmddiagnostics}
#fit_memory$cmdstan_diagnose()
fit_memory$diagnostic_summary()
```

We are looking for strict zeros in `num_divergent`. With the NCP, this should hold. Any divergences here are a signal that `adapt_delta` needs further adjustment, the priors are misspecified, or there is a structural model issue.

#### Chain Mixing and Stationarity ($\hat{R}$)

```{r 06-memory-chain-mixing}
draws_memory <- fit_memory$draws()

mcmc_trace(
  draws_memory,
  pars = c("mu_alpha", "mu_beta", "sigma[1]", "sigma[2]", "rho"),
  np   = nuts_params(fit_memory)
) +
  labs(
    title    = "Trace Plots: Population Parameters (Memory Agent NCP)",
    subtitle = "All five population-level parameters, including rho. Stationary chains expected throughout."
  )

draws_memory |>
  subset_draws(variable = c("mu_alpha", "mu_beta", "sigma[1]", "sigma[2]", "rho",
                            "alpha[1]", "alpha[2]", "beta_mem[1]", "beta_mem[2]")) |>
  summarise_draws(default_convergence_measures()) |>
  knitr::kable(digits = 3)
```

The potential scale reduction factor ($\hat{R}$) must be below 1.01 for all parameters. The correlation parameter `rho` is informed only by covariation *across* agents, not within any individual agent's trial sequence, so its ESS will be lower than the marginal parameters. ESS > 400 remains our threshold. An ESS below 200 for `rho` warrants investigation of the correlation manifold geometry.

##### Effective Sample Size (ESS)

Monitor `ess_tail` for `sigma[1]` and `sigma[2]`. Population SDs can develop mild funnel-like behavior even under NCP when the likelihood is highly informative about individual parameters, because the $z_j$ draws must simultaneously explain many agents. If `ess_tail` is low for the sigmas but `ess_bulk` is acceptable, raising `adapt_delta` further or running more warmup iterations is the correct response.

##### Energy (E-BFMI)

```{r 06-memory-energy}
mcmc_nuts_energy(nuts_params(fit_memory)) +
  labs(
    title    = "Energy Transitions: Memory Agent NCP",
    subtitle = "Marginal and transition energy distributions must align. E-BFMI > 0.2 required."
  )
```

#### Posterior Parameter Correlations

```{r 06-memory-posterior-correlations}
mcmc_pairs(
  draws_memory,
  pars         = c("mu_alpha", "mu_beta", "sigma[1]", "sigma[2]", "rho"),
  np           = nuts_params(fit_memory),
  diag_fun     = "dens",
  off_diag_fun = "hex"
)
```

The critical panels are $(\sigma_\alpha, \mu_\alpha)$ and $(\sigma_\beta, \mu_\beta)$. Under the CP biased agent, these showed triangular wedges — the Neal's Funnel signature. Under NCP, they should be roughly elliptical. Any triangular shape here would be unexpected and would require investigation.

### Prior-Posterior Update Checks: Quantifying Learning

```{r 06-memory-prior-posterior-update}
draws_df_memory <- as_draws_df(draws_memory)

# Extract true generative hyperparameters saved as attributes
true_hp <- attr(sim_data_memory, "hyperparameters")

true_pop_mem <- tibble(
  parameter  = c("mu_alpha", "mu_beta", "sigma_alpha", "sigma_beta", "rho"),
  true_value = c(true_hp$mu_alpha, true_hp$mu_beta,
                 true_hp$sigma_alpha, true_hp$sigma_beta, true_hp$rho)
)

# Reshape: split on the underscore immediately before "posterior" or "prior"
pp_update_mem <- draws_df_memory |>
  select(
    mu_alpha_posterior  = mu_alpha,        mu_alpha_prior,
    mu_beta_posterior   = mu_beta,         mu_beta_prior,
    sigma_alpha_posterior = `sigma[1]`,    sigma_alpha_prior,
    sigma_beta_posterior  = `sigma[2]`,    sigma_beta_prior,
    rho_posterior       = rho,             rho_prior
  ) |>
  pivot_longer(
    everything(),
    names_to  = c("parameter", "distribution"),
    names_sep = "_(?=posterior|prior)"
  )

ggplot(pp_update_mem, aes(x = value)) +
  geom_histogram(aes(fill = distribution), alpha = 0.6, color = NA, bins = 60) +
  geom_vline(
    data = true_pop_mem,
    aes(xintercept = true_value, linetype = "True Value"),
    color = "darkred", linewidth = 1
  ) +
  facet_wrap(~parameter, scales = "free") +
  scale_fill_manual(
    name   = "Distribution",
    values = c("prior" = "gray70", "posterior" = "steelblue")
  ) +
  scale_linetype_manual(
    name   = "Ground Truth",
    values = c("True Value" = "dashed")
  ) +
  labs(
    title = "Prior-Posterior Update Checks: Memory Agent",
    x     = "Parameter Value",
    y     = "Count"
  )
```

A healthy update shows a posterior significantly narrower than the prior and comfortably contained within the prior's mass. The `rho` panel carries the central new scientific information. If the posterior on $\rho$ is essentially indistinguishable from the LKJ(2) prior, the data do not contain sufficient information to identify the population-level correlation — this can happen with small $J$ or short trial sequences. In that case, report the parameter as unidentified, not the prior as the conclusion. A clearly narrowed posterior constitutes genuine evidence about the cognitive architecture of the population.

3### Posterior Predictive Checks (PPC): Generative Adequacy

```{r 06-memory-posterior-pc}
y_obs_mem     <- as.numeric(stan_data_memory$h)
agent_idx_mem <- as.numeric(stan_data_memory$agent)
y_rep_mem     <- fit_memory$draws("h_post_rep", format = "matrix")

n_sample_mem       <- 12
sampled_agents_mem <- sample(unique(agent_idx_mem), n_sample_mem)
keep_mask_mem      <- agent_idx_mem %in% sampled_agents_mem

prop_right <- function(x) mean(x)

ppc_stat_grouped(
  y     = y_obs_mem[keep_mask_mem],
  yrep  = y_rep_mem[, keep_mask_mem],
  group = agent_idx_mem[keep_mask_mem],
  stat  = "prop_right"
) +
  labs(
    title    = "Posterior Predictive Checks: Agent-Level Choice Rates (Subset)",
    subtitle = paste("Showing", n_sample_mem, "randomly sampled agents. Dark line: Observed. Histogram: Predicted."),
    x        = "Proportion of 'Right' Choices"
  )
```

#### Calibration and Recovery

We have proven the model fits the data and generates plausible predictions. Now we must prove the parameters we extract are actually the true parameters that generated the data. For a bivariate model, recovery must be assessed at both levels: all five population parameters, and the individual $(\alpha_j, \beta_j)$ pairs jointly.

##### Individual-Level Recovery

```{r 06-memory-individual-recovery}
recovery_alpha <- draws_memory |>
  subset_draws(variable = "alpha") |>
  summarise_draws() |>
  mutate(agent_id = as.integer(str_extract(variable, "(?<=\\[)\\d+(?=\\])")))

recovery_beta <- draws_memory |>
  subset_draws(variable = "beta_mem") |>
  summarise_draws() |>
  mutate(agent_id = as.integer(str_extract(variable, "(?<=\\[)\\d+(?=\\])")))

true_agent_params_mem <- sim_data_memory |>
  distinct(agent_id, n_trials, true_alpha, true_beta)

recovery_mem_df <- true_agent_params_mem |>
  left_join(recovery_alpha |> select(agent_id, alpha_est = median, alpha_q5 = q5, alpha_q95 = q95), by = "agent_id") |>
  left_join(recovery_beta  |> select(agent_id, beta_est  = median, beta_q5  = q5, beta_q95  = q95),  by = "agent_id")

p_rec_alpha <- ggplot(recovery_mem_df, aes(x = true_alpha, y = alpha_est)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_errorbar(aes(ymin = alpha_q5, ymax = alpha_q95), width = 0, alpha = 0.5, color = "steelblue") +
  geom_point(aes(size = n_trials), color = "midnightblue", alpha = 0.8) +
  labs(
    title = expression("Recovery: Baseline Bias " * alpha[j]),
    x = "True alpha (logit scale)", y = "Posterior Median", size = "N trials"
  )

p_rec_beta <- ggplot(recovery_mem_df, aes(x = true_beta, y = beta_est)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_errorbar(aes(ymin = beta_q5, ymax = beta_q95), width = 0, alpha = 0.5, color = "steelblue") +
  geom_point(aes(size = n_trials), color = "midnightblue", alpha = 0.8) +
  labs(
    title = expression("Recovery: Memory Sensitivity " * beta[j]),
    x = "True beta", y = "Posterior Median", size = "N trials"
  )

p_rec_alpha + p_rec_beta + plot_layout(guides = "collect")
```

Recovery for $\beta_j$ is typically noisier than for $\alpha_j$ at any given trial count. The memory signal $\text{logit}(m_{ij})$ starts at zero (the $\text{logit}(0.5) = 0$ initialization) and becomes informative only as the cumulative opponent history diverges from 0.5. Agents with few trials show stronger shrinkage toward $\mu_\beta = 1.5$ — not toward zero, but toward the estimated population mean, which is the appropriate regularization given what the rest of the cohort tells us about the population.

#### Prior sensitivity

```{r 06-memory-priorsense}
sens_results_mem <- powerscale_sensitivity(
  fit_memory,
  variable = c("mu_alpha", "mu_beta", "sigma[1]", "sigma[2]", "rho")
)
cat("=== PRIOR SENSITIVITY DIAGNOSTICS: MEMORY AGENT ===\n")
print(sens_results_mem)

ps_seq_mem <- powerscale_sequence(
  fit_memory,
  variable = c("mu_alpha", "mu_beta", "sigma[1]", "sigma[2]", "rho")
)

powerscale_plot_dens(ps_seq_mem) +
  labs(
    title    = "Prior Sensitivity: Density Perturbation (Memory Agent)",
    subtitle = "Does scaling the prior or likelihood drastically alter our beliefs about the population?"
  )

powerscale_plot_quantities(ps_seq_mem) +
  labs(
    title    = "Prior Sensitivity: Parameter Estimates (Memory Agent)",
    subtitle = "Horizontal lines: baseline posterior bounds. Deviations indicate sensitivity."
  )
```

Pay particular attention to the `rho` panel. Because the correlation is identified only from across-agent covariation, it has fewer effective observations than the marginal population parameters. If `rho` shows strong sensitivity to prior scaling, the data contain limited information about the correlation structure and the LKJ prior is doing meaningful regularization.

#### Simulation-Based Calibration (SBC)

Parameter recovery on a single dataset is anecdotal. We must prove that the sampler is globally unbiased across the entire prior space.

```{r 06-memory-sbc}
sbc_memory_path <- here::here("simmodels", "ch6_sbc_memory.rds")

generate_sbc_dataset_memory <- function(n_agents = 50, n_trials_min = 20, n_trials_max = 120) {
  sim  <- simulate_hierarchical_memory(n_agents = n_agents,
                                        n_trials_min = n_trials_min,
                                        n_trials_max = n_trials_max)
  hp   <- attr(sim, "hyperparameters")

  list(
    variables = list(
      mu_alpha   = hp$mu_alpha,
      mu_beta    = hp$mu_beta,
      `sigma[1]` = hp$sigma_alpha,
      `sigma[2]` = hp$sigma_beta,
      rho        = hp$rho,
      alpha      = sim |> distinct(agent_id, true_alpha) |> arrange(agent_id) |> pull(true_alpha),
      beta_mem   = sim |> distinct(agent_id, true_beta)  |> arrange(agent_id) |> pull(true_beta)
    ),
    generated = prep_stan_data_memory(sim)
  )
}

generator_memory <- SBC_generator_function(
  generate_sbc_dataset_memory,
  n_agents = 50, n_trials_min = 30, n_trials_max = 120
)

backend_memory <- SBC_backend_cmdstan_sample(
  mod_memory_ncp,
  iter_warmup = 1000, iter_sampling = 1000, chains = 1, refresh = 0
)

if (regenerate_simulations || !file.exists(sbc_memory_path)) {
  sbc_datasets_memory <- generate_datasets(generator_memory, 100)
  plan(sequential)
  #future::plan(future::multisession, workers = 4)
  sbc_results_memory  <- compute_SBC(datasets = sbc_datasets_memory, backend = backend_memory,
    keep_fits = FALSE)
  saveRDS(
    list(datasets = sbc_datasets_memory, results = sbc_results_memory),
    sbc_memory_path
  )
} else {
  sbc_mem_obj         <- readRDS(sbc_memory_path)
  sbc_datasets_memory <- sbc_mem_obj$datasets
  sbc_results_memory  <- sbc_mem_obj$results
}

plot_rank_hist(sbc_results_memory, variables = c("mu_alpha", "mu_beta", "sigma[1]", "sigma[2]", "rho")) +
  labs(title = "SBC Rank Histograms: Memory Agent NCP")

plot_ecdf_diff(sbc_results_memory, variables = c("mu_alpha", "mu_beta", "sigma[1]", "sigma[2]", "rho")) +
  labs(
    title    = "SBC ECDF Difference: Memory Agent NCP",
    subtitle = "Deviation from the diagonal indicates miscalibration."
  )

# 1. Extract the lightweight stats dataframe from the SBC object
sbc_stats_df <- sbc_results_memory$stats

# 2. Filter for the population-level parameters you care about
# (e.g., mu_alpha, mu_beta, rho, or even sigma[1])
recovery_df <- sbc_stats_df |>
  filter(variable %in% c("mu_alpha", "mu_beta", "sigma[1]", "sigma[2]", "rho"))

# 3. Create the Parameter Recovery Identity Plot
p_sbc_recovery <- ggplot(recovery_df, aes(x = simulated_value, y = median)) +
  # The Identity Line (Perfect Recovery)
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50", linewidth = 1) +
  
  # The Posterior Estimates with 90% Credible Intervals
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0, alpha = 0.3, color = "steelblue") +
  geom_point(color = "midnightblue", alpha = 0.7, size = 2) +
  
  # Facet by parameter
  facet_wrap(~variable, scales = "free") +
  
  labs(
    title = "Parameter Recovery (Extracted Directly from SBC)",
    subtitle = "Points represent the posterior median. Bars are 90% Credible Intervals.",
    x = "True Generative Value (Drawn from Prior)",
    y = "Posterior Median Estimate"
  ) +
  theme_classic() +
  theme(strip.background = element_rect(fill = "gray95", color = NA))

print(p_sbc_recovery)

# 4. Quantify the Recovery (Correlation and RMSE)
recovery_metrics <- recovery_df |>
  group_by(variable) |>
  summarize(
    Pearson_r = cor(simulated_value, median),
    RMSE = sqrt(mean((simulated_value - median)^2)),
    .groups = "drop"
  )

cat("=== PARAMETER RECOVERY QUANTIFICATION ===\n")
print(knitr::kable(recovery_metrics, digits = 3))
```

```{r 06-memory-sbc-diagnostics}
diags_memory  <- sbc_results_memory$backend_diagnostics
mem_fail_rate <- mean(diags_memory$n_divergent > 0) * 100

diags_memory |>
  ggplot(aes(x = n_divergent)) +
  geom_histogram(binwidth = 1, fill = "darkred", alpha = 0.8) +
  labs(
    title    = "Memory NCP Divergence Counts Across 200 SBC Universes",
    subtitle = sprintf("%.0f%% of universes produced at least one divergence.", mem_fail_rate),
    x        = "Number of Divergent Transitions",
    y        = "Number of SBC Universes"
  )

# Does any failure concentrate at small sigma values?
posterior::as_draws_df(sbc_datasets_memory$variables) |>
  as_tibble() |>
  mutate(sim_id = row_number()) |>
  left_join(
    diags_memory |> mutate(sim_id = row_number()) |> select(sim_id, n_divergent),
    by = "sim_id"
  ) |>
  mutate(had_divergence = n_divergent > 0) |>
  ggplot(aes(x = `sigma[1]`, y = `sigma[2]`, color = had_divergence)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "darkred"),
    labels = c("FALSE" = "Clean", "TRUE" = "Divergent"),
    name   = NULL
  ) +
  labs(
    title    = "Memory NCP: Where Do Divergences Occur?",
    subtitle = "Any divergences should not concentrate at small sigma values (that would indicate residual funnel geometry).",
    x        = expression(sigma[alpha]), y = expression(sigma[beta])
  )
```

The `rho` rank histogram will typically have wider confidence bands than the marginal mean and SD histograms at $N = 200$. This is expected: $\rho$ is identified only from covariation across $J = 50$ agents, so each SBC universe contributes one effective draw of $\rho$ to the calibration check. Mild irregularity within the confidence envelope is statistical noise. Systematic U-shape or strong asymmetry would indicate a genuine calibration failure.

#### Cognitive Interpretation of $\hat{\rho}$

The posterior on $\rho$ is the inferential result that distinguishes this model from two independent univariate hierarchical models.

* **$\rho \approx 0$**: Baseline tendency and memory-based opponent tracking are orthogonal individual differences. Studying one tells us nothing about the other.

* **$\rho > 0$**: Agents with a stronger directional baseline also tend to weight memory more heavily — consistent with a general exploitation tendency, where both parameters index willingness to depart from 50/50 in response to a signal.

* **$\rho < 0$**: Agents with a strong directional baseline tend to suppress memory use — a compensatory account in which prior commitment crowds out sensitivity to the opponent's history.

None of these interpretations follow from the model alone. The model provides a calibrated posterior over $\rho$. Cognitive theory provides the framework for deciding what that value means.

#### LOO for Model Comparison

We retain the LOO-CV score here for the model comparison section at the end of the chapter.

```{r 06-loo-memory}
loo_memory <- loo(fit_memory$draws("log_lik", format = "matrix"))
print(loo_memory)
```

