---
title: "10-CategorizationModels"
output: html_document
date: "2023-03-13"
---

```{r}
pacman::p_load(
    tidyverse,
    future,
    purrr,
    furrr,
    patchwork,
    brms,
    cmdstanr
)
```


# Categorization models

In this module the course covers formal models of categorization. In particular, here we implement a version of the Generalized Context Model (from now on GCM). 

[MISSING: INTRODUCTION TO THE GCM]

In order to understand the GCM, we will rely on a setup from Kruschke 1993. Participants go through 8 blocks of trials. Within each block, participants see 8 stimuli, the same across every block. Each stimulus has to be categorised (classified) as belonging to either category A or category B. After categorizing a stimulus, feedback is given, so that the participant knows which category the stimulus truly belongs to.

The stimuli differ along two continuous dimensions (more specifically the height of a square and the position of a line with it).

Following on the structure in Kruschke 1993, we need to start simulating data. Let's start with the stimuli.

```{r}

# Defining the stimuli, their height and position features, and their category
stimulus <- c(5,3,7,1,8,2,6,4)
height <- c(1,1, 2,2, 3,3, 4,4)
position <- c(2,3, 1,4, 1,4, 2,3)
category <- as.factor(c(0,0, 1,0, 1,0, 1,1))

# Making this into a tibble
stimuli <- tibble(stimulus, height, position, category)

# Plotting to make sure all looks right
ggplot(stimuli, aes(position, height, color = category, label = stimulus)) +
  geom_point(shape = 16, size = 3) +
  geom_label() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  theme_bw()

# Generating the sequence of stimuli in the full experiment
sequence <- c()
for (i in 1:8) {
  temp <- sample(seq(8), 8, replace = F)
  sequence <- append(sequence, temp)
}

experiment <- tibble(stimulus = sequence, height = NA, position = NA, category = NA)
for (i in seq(nrow(experiment))) {
  experiment$height[i] <- stimuli$height[stimuli$stimulus == experiment$stimulus[i]]
  experiment$position[i] <- stimuli$position[stimuli$stimulus == experiment$stimulus[i]]
  experiment$category[i] <- as.numeric(as.character(stimuli$category[stimuli$stimulus == experiment$stimulus[i]]))
}

```

We then have to simulate an agent using the GCM to assess the stimuli. 
N.B. code based on work by Sara Østergaard

First we define functions for distance and similarity and assess them.
Distance takes feature values along n dimensions and calculate their euclidean distance,
weighing dimensions according to pregiven weights (w).
Similarity decays exponentially with distance, with a pre-given exponent (c). See plot.

```{r}
# Distance 
distance <- function(vect1, vect2, w) {
  return(sum(w * abs(vect1 - vect2)))
}

# Similarity
similarity <- function(distance, c) {
  return(exp(-c * distance))
}

# Let's assess similarity
dd <- tibble(
  expand_grid(
    distance = c(0,.1,.2, .3,.4,.5,1,1.5,2,3,4,5,6), 
    c = c(0.1, 0.2, 0.5, 0.7, 1, 1.5, 2, 3, 4, 5, 6))) %>% 
  mutate(
    similarity = similarity(distance, c)
  )

dd %>% mutate(c = factor(c)) %>%
  ggplot() +
  geom_line(aes(distance, similarity, group = c, color = c)) + 
  theme_bw()

#
```


With distance and similarity in place, we now need to implement an agent that can observe stimuli,
put them into categories according to feedback and - once enough stimuli are collected to have at
least one exemplar in each category - compare stimuli to exemplars in each category and assess
which category is more likely.
In other words: 1) if not enough stimuli have already been observed, the agent picks a category at random,
receives feedback and adjusts accordingly the category; otherwise, 2) it assesses average distance (according to weights) from observed exemplars by category; calculates similarity (according to c); and uses the relative similarity to the two categories to produce a probability of choosing category one. then it receives feedback and adjusts category accordingly.

As usual we make the agent a function, so that we can more easily deploy it.

```{r}
### generative model ###
gcm <- function(w, c, obs, cat_one, quiet = TRUE) {
  # create an empty list to save probability of saying "1" for each trial
  r <- c()
  
  ntrials <- nrow(obs)
  
  for (i in 1:ntrials) {
    # If quiet is FALSE, print every ten trials
    if (!quiet && i %% 10 == 0) {
      print(paste("i =", i))
    }
    # if this is the first trial, or there any category with no exemplars seen yet, set the choice to random
    if (i == 1 || sum(cat_one[1:(i - 1)]) == 0 || sum(cat_one[1:(i - 1)]) == (i - 1)) {
      r <- c(r, .5)
    } else {
      similarities <- c()
      # for each previously seen stimulus assess distance and similarity
      for (e in 1:(i - 1)) {
        sim <- similarity(distance(obs[i, ], obs[e, ], w), c)
        similarities <- c(similarities, sim)
      }
      # Calculate prob of saying "1" by dividing similarity to 1 by the sum of similarity to 1 and to 2
      numerator <- 0.5 * sum(similarities[cat_one[1:(i - 1)] == 1])
      denominator <- 0.5 * sum(similarities[cat_one[1:(i - 1)] == 1]) + 0.5 * sum(similarities[cat_one[1:(i - 1)] == 0])
      r <- c(r, numerator / denominator)
    }
  }

  return(rbinom(ntrials, 1, r))
}
```

With the agent in place, we can now simulate behavior on our simulated stimuli and observe the impact of different weights and scaling parameters c.

```{r}
# function for simulation responses
simulate_responses <- function(agent, w, c) {
    
    observations <- experiment %>%
        select(c("height", "position"))
    
    category <- experiment$category
    
    if (w == "equal") {
        weight <- rep(1 / 2, 2)
    } else if (w == "skewed1") {
        weight <- c(0, 1)
    } else if (w == "skewed2") {
        weight <- c(0.1, 0.9)
    }

    # simulate responses
    responses <- gcm(
        weight,
        c,
        observations,
        category
    )
    
    tmp_simulated_responses <- experiment %>%
        mutate(
            trial = seq(nrow(experiment)),
            sim_response = responses,
            correct = ifelse(category == sim_response, 1, 0),
            performance = cumsum(correct) / seq_along(correct),
            c = c,
            w = w,
            agent = agent
        )

    return(tmp_simulated_responses)
}


# simulate responses
plan(multisession, workers = availableCores())

param_df <- dplyr::tibble(
    expand_grid(
        agent = 1:10,
        c = seq(.1, 2, 0.2),
        w = c("equal", "skewed1", "skewed2")
    )
)

simulated_responses <- future_pmap_dfr(param_df,
    simulate_responses,
    .options = furrr_options(seed = TRUE)
)

```

And now we can try a few plots to better understand how the model fares and the difference made by weights and scaling factors.

```{r}

p3 <- simulated_responses %>%
  mutate(w = as.factor(w)) %>%
  ggplot(aes(trial, performance, group = w, color = w)) +
  geom_smooth() +
  theme_bw() +
  facet_wrap(c ~ .)

p4 <- simulated_responses %>%
  mutate(c = as.factor(c)) %>%
  ggplot(aes(trial, performance, group = c, color = c)) +
  geom_smooth() +
  theme_bw() +
  facet_wrap(w ~ .)

p3 + p4

p5 <- simulated_responses %>%
  mutate(c = as.factor(c)) %>%
  ggplot(aes(st, performance, group = c, color = c)) +
  geom_smooth() +
  theme_bw() +
  facet_wrap(w ~ .)
```

[MISSING: Discussion of how to make sense of the plots: Equal Weights (matching the actual rule) generating better decisions as long as c is above 1]

While we could try different scenarios, let us now move on to inference. We need to build the same GCM model in Stan.

[MISSING: MORE DETAILED EXPLANATION]

```{r}

gcm_model <- "
// Generalized Context Model (GCM)

data {
    int<lower=1> ntrials;  // number of trials
    int<lower=1> nfeatures;  // number of predefined relevant features
    array[ntrials] int<lower=0, upper=1> cat_one; // true responses on a trial by trial basis
    array[ntrials] int<lower=0, upper=1> y;  // decisions on a trial by trial basis
    array[ntrials, nfeatures] real obs; // stimuli as vectors of features
    real<lower=0, upper=1> b;  // initial bias for category one over two

    // priors
    vector[nfeatures] w_prior_values;  // concentration parameters for dirichlet distribution <lower=1>
    array[2] real c_prior_values;  // mean and variance for logit-normal distribution
}

transformed data {
    array[ntrials] int<lower=0, upper=1> cat_two; // dummy variable for category two over cat 1
    array[sum(cat_one)] int<lower=1, upper=ntrials> cat_one_idx; // array of which stimuli are cat 1
    array[ntrials-sum(cat_one)] int<lower=1, upper=ntrials> cat_two_idx; //  array of which stimuli are cat 2
    int idx_one = 1; // Initializing 
    int idx_two = 1;
    for (i in 1:ntrials){
        cat_two[i] = abs(cat_one[i]-1);

        if (cat_one[i]==1){
            cat_one_idx[idx_one] = i;
            idx_one +=1;
        } else {
            cat_two_idx[idx_two] = i;
            idx_two += 1;
        }
    }
}

parameters {
    simplex[nfeatures] w;  // simplex means sum(w)=1
    real logit_c;
}

transformed parameters {
    // parameter c 
    real<lower=0, upper=2> c = inv_logit(logit_c)*2;  // times 2 as c is bounded between 0 and 2

    // parameter r (probability of response = category 1)
    array[ntrials] real<lower=0.0001, upper=0.9999> r;
    array[ntrials] real rr;

    for (i in 1:ntrials) {

        // calculate distance from obs to all exemplars
        array[(i-1)] real exemplar_sim;
        for (e in 1:(i-1)){
            array[nfeatures] real tmp_dist;
            for (j in 1:nfeatures) {
                tmp_dist[j] = w[j]*abs(obs[e,j] - obs[i,j]);
            }
            exemplar_sim[e] = exp(-c * sum(tmp_dist));
        }

        if (sum(cat_one[:(i-1)])==0 || sum(cat_two[:(i-1)])==0){  // if there are no examplars in one of the categories
            r[i] = 0.5;

        } else {
            // calculate similarity
            array[2] real similarities;
            
            array[sum(cat_one[:(i-1)])] int tmp_idx_one = cat_one_idx[:sum(cat_one[:(i-1)])];
            array[sum(cat_two[:(i-1)])] int tmp_idx_two = cat_two_idx[:sum(cat_two[:(i-1)])];
            similarities[1] = sum(exemplar_sim[tmp_idx_one]);
            similarities[2] = sum(exemplar_sim[tmp_idx_two]);

            // calculate r[i]
            rr[i] = (b*similarities[1]) / (b*similarities[1] + (1-b)*similarities[2]);

            // to make the sampling work
            if (rr[i] > 0.9999){
                r[i] = 0.9999;
            } else if (rr[i] < 0.0001) {
                r[i] = 0.0001;
            } else if (rr[i] > 0.0001 && rr[i] < 0.9999) {
                r[i] = rr[i];
            } else {
                r[i] = 0.5;
            }
        }
    }
}

model {
    // Priors
    target += dirichlet_lpdf(w | w_prior_values);
    target += normal_lpdf(logit_c | c_prior_values[1], c_prior_values[2]);
    
    
    // Decision Data
    target += bernoulli_lpmf(y | r);
}

generated quantities {
    // priors
    simplex[nfeatures] w_prior = dirichlet_rng(w_prior_values);
    real logit_c_prior = normal_rng(c_prior_values[1], c_prior_values[2]);
    real<lower=0, upper=2> c_prior = inv_logit(logit_c_prior)*2;

    // prior pred
    array[ntrials] real<lower=0, upper=1> r_prior;
    array[ntrials] real rr_prior;
    for (i in 1:ntrials) {

        // calculate distance from obs to all exemplars
        array[(i-1)] real exemplar_dist;
        for (e in 1:(i-1)){
            array[nfeatures] real tmp_dist;
            for (j in 1:nfeatures) {
                tmp_dist[j] = w_prior[j]*abs(obs[e,j] - obs[i,j]);
            }
            exemplar_dist[e] = sum(tmp_dist);
        }

        if (sum(cat_one[:(i-1)])==0 || sum(cat_two[:(i-1)])==0){  // if there are no examplars in one of the categories
            r_prior[i] = 0.5;

        } else {
            // calculate similarity
            array[2] real similarities;
            
            array[sum(cat_one[:(i-1)])] int tmp_idx_one = cat_one_idx[:sum(cat_one[:(i-1)])];
            array[sum(cat_two[:(i-1)])] int tmp_idx_two = cat_two_idx[:sum(cat_two[:(i-1)])];
            similarities[1] = exp(-c_prior * sum(exemplar_dist[tmp_idx_one]));
            similarities[2] = exp(-c_prior * sum(exemplar_dist[tmp_idx_two]));

            // calculate r[i]
            rr_prior[i] = (b*similarities[1]) / (b*similarities[1] + (1-b)*similarities[2]);

            // to make the sampling work
            if (rr_prior[i] == 1){
                r_prior[i] = 0.9999;
            } else if (rr_prior[i] == 0) {
                r_prior[i] = 0.0001;
            } else if (rr_prior[i] > 0 && rr_prior[i] < 1) {
                r_prior[i] = rr_prior[i];
            } else {
                r_prior[i] = 0.5;
            }
        }
    }

    array[ntrials] int<lower=0, upper=1> priorpred = bernoulli_rng(r_prior);


    // posterior pred
    array[ntrials] int<lower=0, upper=1> posteriorpred = bernoulli_rng(r);
    array[ntrials] int<lower=0, upper=1> posteriorcorrect;
    for (i in 1:ntrials) {
        if (posteriorpred[i] == cat_one[i]) {
            posteriorcorrect[i] = 1;
        } else {
            posteriorcorrect[i] = 0;
        }
    }
    
    
    // log likelihood
   array[ntrials] real log_lik;

   for (i in 1:ntrials) {
        log_lik[i] = bernoulli_lpmf(y[i] | r[i]);
   }

}"

write_stan_file(
  gcm_model,
  dir = "stan/",
  basename = "W10_GCM.stan")

file <- file.path("stan/W10_GCM.stan")
mod_GCM <- cmdstan_model(file, cpp_options = list(stan_threads = TRUE),
                     stanc_options = list("O1"))
```

Now it's time to fit the data

```{r}

d <- simulated_responses %>% subset(
  c == "1.1" & w == "equal"
)

gcm_data <- list(
  ntrials = nrow(d),
  nfeatures = 2,
  cat_one = d$category,
  y = d$sim_response,
  obs = as.matrix(d[, c("height", "position")]),
  b = 0.5,
  w_prior_values = c(1, 1),
  c_prior_values = c(0, 1)
)

samples_gcm <- mod_GCM$sample(
  data = gcm_data,
  seed = 123,
  chains = 1,
  parallel_chains = 1,
  threads_per_chain = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  refresh = 500
)

```

Now we need to check the model

```{r}
samples_gcm$cmdstan_diagnose()
samples_gcm$summary()

draws_df <- as_draws_df(samples_gcm$draws())

ggplot(draws_df, aes(.iteration, c, group = .chain, color = .chain)) +
  geom_line(alpha = 0.5) +
  theme_classic()
ggplot(draws_df, aes(.iteration, logit_c, group = .chain, color = .chain)) +
  geom_line(alpha = 0.5) +
  theme_classic()

ggplot(draws_df, aes(.iteration, `w[1]`, group = .chain, color = .chain)) +
  geom_line(alpha = 0.5) +
  theme_classic()

ggplot(draws_df, aes(.iteration, `w[2]`, group = .chain, color = .chain)) +
  geom_line(alpha = 0.5) +
  theme_classic()

ggplot(draws_df) +
  geom_histogram(aes(c), alpha = 0.6, fill = "lightblue") +
  geom_histogram(aes(c_prior), alpha = 0.6, fill = "pink") +
  geom_vline(xintercept = d$c[1]) +
  theme_bw()

ggplot(draws_df) +
  geom_histogram(aes(`w[1]`), alpha = 0.6, fill = "lightblue") +
  geom_histogram(aes(`w_prior[1]`), alpha = 0.6, fill = "pink") +
  geom_vline(xintercept = 0.5) +
  theme_bw()

ggplot(draws_df) +
  geom_histogram(aes(`w[2]`), alpha = 0.6, fill = "lightblue") +
  geom_histogram(aes(`w_prior[2]`), alpha = 0.6, fill = "pink") +
  geom_vline(xintercept = 0.5) +
  theme_bw()

ggplot(draws_df) +
  geom_point(aes(c, `w[2]`), alpha = 0.6, color = "lightblue") +
  theme_bw()
ggplot(draws_df) +
  geom_point(aes(c, `w[1]`), alpha = 0.6, color = "lightblue") +
  theme_bw()
ggplot(draws_df) +
  geom_point(aes(`w[1]`, `w[2]`), alpha = 0.6, color = "lightblue") +
  theme_bw()

```

C looks like crap, while weights seem decently recovered. Let's look at the larger picture, but looping through the different combinations of parameter values.


```{r, eval = F}
pacman::p_load(future, purrr, furrr)
plan(multisession, workers = availableCores())

sim_d_and_fit <- function(agent, scaling, weights) {
  
    
    temp <- simulated_responses %>% subset(
      c == scaling & w == weights & agent == agent
      )
    
    data <- list(
      ntrials = nrow(temp),
      nfeatures = 2,
      cat_one = temp$category,
      y = temp$sim_response,
      obs = as.matrix(temp[, c("height", "position")]),
      b = 0.5,
      w_prior_values = c(1, 1),
      c_prior_values = c(0, 1)
    )
    
    samples_gcm <- mod_GCM$sample(
      data = data,
      seed = 123,
      chains = 1,
      parallel_chains = 1,
      threads_per_chain = 4,
      iter_warmup = 1000,
      iter_sampling = 1000,
      refresh = 500
    )
    
    draws_df <- as_draws_df(samples_gcm$draws()) 
    temp <- tibble(trueC = scaling, trueW = weights, agent = agent,
                   estC = draws_df$c, 
                   estW1 = draws_df$`w[1]`,
                   estW2 = draws_df$`w[2]`
                   )
    
    return(temp)
  
}


temp <- tibble(unique(simulated_responses[,c("agent", "c", "w")])) %>%
  rename(
    scaling = c,
    weights = w
  )

recovery_df <- future_pmap_dfr(temp, sim_d_and_fit, .options = furrr_options(seed = TRUE))

write_csv(recovery_df, "simdata/W10_GCM_recoverydf.csv")
```

Time to visualize

```{r}
recovery_df <- read_csv("simdata/W10_GCM_recoverydf.csv")

p1 <- ggplot(recovery_df) +
  geom_density(aes(estC, group = trueW, fill = trueW), alpha = 0.3) +
  #geom_vline(xintercept = trueC) +
  facet_wrap(. ~ trueC) +
  theme_bw()

p2 <- ggplot(recovery_df) +
  geom_density(aes(estW1, group = trueW, fill = trueW), alpha = 0.3) +
  #geom_vline(xintercept = trueC) +
  facet_wrap(trueC ~ .) +
  theme_bw()

p1 / p2
```

## Multilevel GCM

[MISSING: quick reminder of what it means to make GCM multilevel: weights and can vary by participants, but they do not do so in correlated ways]

### Simulate hierarchical simplices

We need weights that sum up to 1 at the population level and at the individual level. Let's simulate the dirichlet distribution to see how it does it.

```{r}

simulate_dirichlet <- function(weights, kappa, agents){
  
  w_n <- length(weights)
  
  w_ind <- rdirichlet(agents, weights * kappa)
  
  w_ind_df <- tibble(
    agent = as.factor(rep(seq(agents), each = w_n)),
    value = c(w_ind),
    weight = rep(seq(w_n), agents)
  )
  
  return(w_ind_df)
}

d <- simulate_dirichlet(weights = c(0.5, 0.5), kappa = 100, agents = 10)

p1 <- ggplot(d, aes(weight, value, group = agent, color = agent)) +
  geom_point() +
  geom_line(linetype = "dashed", alpha = 0.5) +
  ylim(0,1) +
  theme_bw()

d <- simulate_dirichlet(weights = c(0.5, 0.5), kappa = 1, agents = 10)

p2 <- ggplot(d, aes(weight, value, group = agent, color = agent)) +
  geom_point() +
  geom_line(linetype = "dashed", alpha = 0.5) +
  ylim(0,1) +
  theme_bw()

p1 + p2


```

### Set up the data simulation

We need to feed: 
- agents
- scaling (M and SD)
- weights (weights and concentration)

```{r}
agents <- 10
scalingM <- 1
scalingSD <- 0.1
weights <- c(0.5,0.5)
kappa <- 1

# First one agent, just to make sure it works
simulate_responses <- function(agent, weights, c) {
    
    observations <- experiment %>%
        select(c("height", "position"))
    
    category <- experiment$category
    
    # simulate responses
    responses <- gcm(
        weights,
        c,
        observations,
        category
    )
    
    tmp_simulated_responses <- experiment %>%
        mutate(
            trial = seq(nrow(experiment)),
            sim_response = responses,
            correct = ifelse(category == sim_response, 1, 0),
            performance = cumsum(correct) / seq_along(correct),
            c = c,
            w1 = weights[1],
            w2 = weights[2],
            agent = agent
        )

    return(tmp_simulated_responses)
}

d <- simulate_responses(agents, weights, scalingM)

## Then we make sure to simulate n agents

simulate_ml_responses <- function(agents, scalingM, scalingSD, weights, kappa) {
  
  w_ind <- rdirichlet(agents, weights * kappa)
  c_ind <- rnorm(agents, scalingM, scalingSD)
  
  for (i in 1:agents) {
    tmp <- simulate_responses(i, weights = c(w_ind[i,1:2]), c = c_ind[i])
    if (i == 1) {
      simulated_responses <- tmp
    } else {
      simulated_responses <- rbind(simulated_responses, tmp)
    }
  }
  
  return(simulated_responses)
}

# Simulate and visualize

d <- simulate_ml_responses(agents, scalingM, scalingSD, weights, kappa)

ggplot(d, aes(trial, performance)) +
  geom_smooth() +
  geom_line(aes(group = agent, color = agent)) +
  theme_bw()

```


### Set up the Stan model

```{r}
gcm_ml_model <- "
// Generalized Context Model (GCM) - multilevel version

data {
    int<lower=1> nsubjects;  // number of subjects
    int<lower=1> ntrials;  // number of trials
    int<lower=1> nfeatures;  // number of predefined relevant features
    array[ntrials] int<lower=0, upper=1> cat_one; // true responses on a trial by trial basis
    array[ntrials, nsubjects] int<lower=0, upper=1> y;  // decisions on a trial by trial basis
    array[ntrials, nfeatures] real obs; // stimuli as vectors of features assuming all participants get the same sequence
    real<lower=0, upper=1> b;  // initial bias for category one over two

    // priors
    vector[nfeatures] w_prior_values;  // concentration parameters for dirichlet distribution <lower=1>
    array[2] real c_prior_values;  // mean and variance for logit-normal distribution
}

transformed data { // assuming all participants get the same sequence
    
    array[ntrials] int<lower=0, upper=1> cat_two; // dummy variable for category two over cat 1
    array[sum(cat_one)] int<lower=1, upper = ntrials> cat_one_idx; // array of which stimuli are cat 1
    array[ntrials - sum(cat_one)] int<lower = 1, upper = ntrials> cat_two_idx; //  array of which stimuli are cat 2
    int idx_one = 1; // Initializing 
    int idx_two = 1;
    for (i in 1:ntrials){
        cat_two[i] = abs(cat_one[i]-1);

        if (cat_one[i]==1){
            cat_one_idx[idx_one] = i;
            idx_one +=1;
        } else {
            cat_two_idx[idx_two] = i;
            idx_two += 1;
        }
    }
}

parameters {
    real logit_c_M;    // Pop Mean of the scaling parameter (how fast similarity decrease with distance). 
    real<lower = 0> logit_c_SD;    // Pop SD of the scaling parameter (how fast similarity decrease with distance). 
    vector[nsubjects] logit_c;    // scaling parameter (how fast similarity decrease with distance). 
    
    simplex[nfeatures] weight;  // simplex means sum(w)=1
    real<lower=0> kappa;
    array[nsubjects] simplex[nfeatures] w_ind;    // weight parameter (how much attention should be paid to feature 1 related to feature 2 - summing up to 1)
}

transformed parameters {
    // parameter w
    vector[nfeatures] alpha = kappa * weight;
    
    // parameter c 
    vector<lower=0,upper=2>[nsubjects] c = inv_logit(logit_c)*2;  // times 2 as c is bounded between 0 and 2

    // parameter r (probability of response = category 1)
    array[ntrials, nsubjects] real<lower=0.0001, upper=0.9999> r;
    array[ntrials, nsubjects] real rr;

    for (sub in 1:nsubjects) {
      for (trial in 1:ntrials) {

        // calculate distance from obs to all exemplars
        array[(trial-1)] real exemplar_sim;
        for (e in 1:(trial-1)){
            array[nfeatures] real tmp_dist;
            for (feature in 1:nfeatures) {
                tmp_dist[feature] = w_ind[sub,feature]*abs(obs[e,feature] - obs[trial,feature]);
            }
            exemplar_sim[e] = exp(-c[sub] * sum(tmp_dist));
        }

        if (sum(cat_one[:(trial-1)])==0 || sum(cat_two[:(trial-1)])==0){  // if there are no examplars in one of the categories
            r[trial,sub] = 0.5;

        } else {
            // calculate similarity
            array[2] real similarities;
            
            array[sum(cat_one[:(trial-1)])] int tmp_idx_one = cat_one_idx[:sum(cat_one[:(trial-1)])];
            array[sum(cat_two[:(trial-1)])] int tmp_idx_two = cat_two_idx[:sum(cat_two[:(trial-1)])];
            similarities[1] = sum(exemplar_sim[tmp_idx_one]);
            similarities[2] = sum(exemplar_sim[tmp_idx_two]);

            // calculate r
            rr[trial,sub] = (b*similarities[1]) / (b*similarities[1] + (1-b)*similarities[2]);

            // to make the sampling work
            if (rr[trial,sub] > 0.9999){
                r[trial,sub] = 0.9999;
            } else if (rr[trial,sub] < 0.0001) {
                r[trial,sub] = 0.0001;
            } else if (rr[trial,sub] > 0.0001 && rr[trial,sub] < 0.9999) {
                r[trial,sub] = rr[trial,sub];
            } else {
                r[trial,sub] = 0.5;
            }}}}}

model {
    // Priors
    target += exponential_lpdf(kappa | 0.1);
    target += dirichlet_lpdf(weight | w_prior_values);
    
    target += normal_lpdf(logit_c_M | c_prior_values[1], c_prior_values[2]);
    target += normal_lpdf(logit_c_SD | 0, 1) - normal_lccdf(0 | 0, 1);
    
    target += normal_lpdf(logit_c | logit_c_M, logit_c_SD);
    
    // Decision Data
    for (sub in 1:nsubjects){
      
      target += dirichlet_lpdf(w_ind[sub] | alpha);
      
      for (trial in 1:ntrials){
        target += bernoulli_lpmf(y[trial,sub] | r[trial,sub]);
      }
      
    }
    
}

// generated quantities {
//     // priors
//     simplex[nfeatures] w_prior = dirichlet_rng(w_prior_values);
//     real logit_c_prior = normal_rng(c_prior_values[1], c_prior_values[2]);
//     real<lower=0, upper=2> c_prior = inv_logit(logit_c_prior)*2;
// 
//     // prior pred
//     array[ntrials] real<lower=0, upper=1> r_prior;
//     array[ntrials] real rr_prior;
//     for (i in 1:ntrials) {
// 
//         // calculate distance from obs to all exemplars
//         array[(i-1)] real exemplar_dist;
//         for (e in 1:(i-1)){
//             array[nfeatures] real tmp_dist;
//             for (j in 1:nfeatures) {
//                 tmp_dist[j] = w_prior[j]*abs(obs[e,j] - obs[i,j]);
//             }
//             exemplar_dist[e] = sum(tmp_dist);
//         }
// 
//         if (sum(cat_one[:(i-1)])==0 || sum(cat_two[:(i-1)])==0){  // if there are no examplars in one of the categories
//             r_prior[i] = 0.5;
// 
//         } else {
//             // calculate similarity
//             array[2] real similarities;
//             
//             array[sum(cat_one[:(i-1)])] int tmp_idx_one = cat_one_idx[:sum(cat_one[:(i-1)])];
//             array[sum(cat_two[:(i-1)])] int tmp_idx_two = cat_two_idx[:sum(cat_two[:(i-1)])];
//             similarities[1] = exp(-c_prior * sum(exemplar_dist[tmp_idx_one]));
//             similarities[2] = exp(-c_prior * sum(exemplar_dist[tmp_idx_two]));
// 
//             // calculate r[i]
//             rr_prior[i] = (b*similarities[1]) / (b*similarities[1] + (1-b)*similarities[2]);
// 
//             // to make the sampling work
//             if (rr_prior[i] == 1){
//                 r_prior[i] = 0.9999;
//             } else if (rr_prior[i] == 0) {
//                 r_prior[i] = 0.0001;
//             } else if (rr_prior[i] > 0 && rr_prior[i] < 1) {
//                 r_prior[i] = rr_prior[i];
//             } else {
//                 r_prior[i] = 0.5;
//             }
//         }
//     }
// 
//     array[ntrials] int<lower=0, upper=1> priorpred = bernoulli_rng(r_prior);
// 
// 
//     // posterior pred
//     array[ntrials] int<lower=0, upper=1> posteriorpred = bernoulli_rng(r);
//     array[ntrials] int<lower=0, upper=1> posteriorcorrect;
//     for (i in 1:ntrials) {
//         if (posteriorpred[i] == cat_one[i]) {
//             posteriorcorrect[i] = 1;
//         } else {
//             posteriorcorrect[i] = 0;
//         }
//     }
//     
//     
//     // log likelihood
//    array[ntrials] real log_lik;
// 
//    for (i in 1:ntrials) {
//         log_lik[i] = bernoulli_lpmf(y[i] | r[i]);
//    }
// 
// }

"

write_stan_file(
  gcm_ml_model,
  dir = "stan/",
  basename = "W10_GCM_ml.stan")

file <- file.path("stan/W10_GCM_ml.stan")
mod_GCM_ml <- cmdstan_model(file, cpp_options = list(stan_threads = TRUE),
                     stanc_options = list("O1"))
```


```{r}

## Fit the simulated data
d1 <- d[,c("agent","trial","sim_response")] %>% pivot_wider(
  names_from = agent,
  values_from = c(sim_response))

gcm_ml_data <- list(
  nsubjects = agents,
  ntrials = nrow(experiment),
  nfeatures = 2,
  cat_one = experiment$category,
  y = as.matrix(d1[, 2:(agents + 1)]),
  obs = as.matrix(experiment[, c("height", "position")]),
  b = 0.5,
  w_prior_values = c(1, 1),
  c_prior_values = c(0, 1)
)

samples_gcm_ml <- mod_GCM_ml$sample(
  data = gcm_ml_data,
  seed = 123,
  chains = 2,
  parallel_chains = 2,
  threads_per_chain = 2,
  iter_warmup = 1000,
  iter_sampling = 1000,
  refresh = 1000
)

draws <- as_draws_df(samples_gcm_ml)

ggplot(draws, aes(.iteration, logit_c_M, group = .chain, color = .chain)) +
  geom_line(alpha = 0.5) +
  theme_classic()

ggplot(draws, aes(.iteration, logit_c_SD, group = .chain, color = .chain)) +
  geom_line(alpha = 0.5) +
  theme_classic()

ggplot(draws, aes(.iteration, `weight[1]`, group = .chain, color = .chain)) +
  geom_line(alpha = 0.5) +
  theme_classic()

ggplot(draws, aes(.iteration, `weight[2]`, group = .chain, color = .chain)) +
  geom_line(alpha = 0.5) +
  theme_classic()

ggplot(draws, aes(.iteration, `kappa`, group = .chain, color = .chain)) +
  geom_line(alpha = 0.5) +
  theme_classic()


draws %>% mutate(c_prior = rnorm(nrow(draws), 1,1)) %>%
  ggplot() +
  geom_histogram(aes(logit_c_M), alpha = 0.6, fill = "lightblue") +
  geom_histogram(aes(c_prior), alpha = 0.6, fill = "pink") +
  geom_vline(xintercept = logit_scaled(scalingM,0, 2)) +
  theme_bw()

library(msm)

draws %>% mutate(c_prior = rtnorm(nrow(draws), 0, 1, lower = 0)) %>%
  ggplot() +
  geom_histogram(aes(logit_c_SD), alpha = 0.6, fill = "lightblue") +
  geom_histogram(aes(c_prior), alpha = 0.6, fill = "pink") +
  geom_vline(xintercept = 0.5) +
  theme_bw()


draws %>% mutate(`w_prior[1]` = rdirichlet(nrow(draws), c(1, 1))[,1]) %>%
  ggplot() +
  geom_histogram(aes(`weight[1]`), alpha = 0.6, fill = "lightblue") +
  geom_histogram(aes(`w_prior[1]`), alpha = 0.6, fill = "pink") +
  geom_vline(xintercept = 0.5) +
  theme_bw()

draws %>% mutate(`w_prior[2]` = rdirichlet(nrow(draws), c(1, 1))[,1]) %>%
  ggplot() +
  geom_histogram(aes(`weight[2]`), alpha = 0.6, fill = "lightblue") +
  geom_histogram(aes(`w_prior[2]`), alpha = 0.6, fill = "pink") +
  geom_vline(xintercept = 0.5) +
  theme_bw()

draws %>% mutate(`kappa_prior` = rexp(nrow(draws), 0.1)) %>%
  ggplot() +
  geom_histogram(aes(`kappa`), alpha = 0.6, fill = "lightblue") +
  geom_histogram(aes(`kappa_prior`), alpha = 0.6, fill = "pink") +
  geom_vline(xintercept = 1) +
  theme_bw()

ggplot(draws) +
  geom_point(aes(`weight[1]`, `weight[2]`), alpha = 0.6, color = "lightblue") +
  theme_bw()

ggplot(draws) +
  geom_point(aes(logit_c_M, logit_c_SD), alpha = 0.6, color = "lightblue") +
  theme_bw()

ggplot(draws) +
  geom_point(aes(kappa, `weight[1]`), alpha = 0.6, color = "lightblue") +
  theme_bw()

ggplot(draws) +
  geom_point(aes(kappa, `weight[2]`), alpha = 0.6, color = "lightblue") +
  theme_bw()

ggplot(draws) +
  geom_point(aes(kappa, `logit_c_M`), alpha = 0.6, color = "lightblue") +
  theme_bw()

```

## Now we parallelize [N.B. NOT FINISHED]

```{r}
# ## Then we create the function that simulates data and fits the model, so we can quickly go through multiple parameter values in the simulation
# 
# sim_ml_and_fit <- function(agents, scalingM, scalingSD, weights, kappa) {
#   
#   d <- simulate_ml_responses(agents, scalingM, scalingSD, weights, kappa)
#   
#   d1 <- d[,c("agent","trial","sim_response")] %>% pivot_wider(
#     names_from = agent,
#     values_from = c(sim_response))
#   
#   gcm_ml_data <- list(
#     nsubjects = agents,
#     ntrials = nrow(experiment),
#     nfeatures = 2,
#     cat_one = experiment$category,
#     y = as.matrix(d1[, 2:(agents + 1)]),
#     obs = as.matrix(experiment[, c("height", "position")]),
#     b = 0.5,
#     w_prior_values = c(1, 1),
#     c_prior_values = c(0, 1)
#   )
#   
#   samples_gcm_ml <- mod_GCM_ml$sample(
#     data = gcm_ml_data,
#     seed = 123,
#     chains = 2,
#     parallel_chains = 2,
#     threads_per_chain = 2,
#     iter_warmup = 1000,
#     iter_sampling = 1000,
#     refresh = 1000
#   )
#   
#   
#   draws_df <- as_draws_df(samples_gcm$draws()) 
#   temp <- tibble(trueC_M = scalingM, 
#                  trueC_SD = scalingSD, 
#                  trueW1 = weights[1], 
#                  trueW2 = weights[2], 
#                  trueKappa = kappa, 
#                  agents = agents,
#                  # FROM HERE
#                  estC = draws_df$c, 
#                  estW1 = draws_df$`w[1]`,
#                  estW2 = draws_df$`w[2]`
#   )
#   
#   return(temp)
#   
# }
# 
# ## Now I need to make a list of lists
# 
# 
# temp <- tibble(unique(simulated_responses[,c("agent", "c", "w")])) %>%
#   rename(
#     scaling = c,
#     weights = w
#   )
# 
# recovery_df <- future_pmap_dfr(temp, sim_d_and_fit, .options = furrr_options(seed = TRUE))
# 
# write_csv(recovery_df, "simdata/W10_GCM_recoverydf.csv")
```



### Student-produced models of categorization

* Attempting to identify features

* Attempting to associate 1 of the features to one of the outcome dimensions (e.g. green as danger), test hypothesis, stay if positive, try different trait if negative. [missing details on how to define the hypothesized association, and how much evidence needed to shift]

* Remembering the rule and/or remembering the single exemplars

* Focusing on 1 feature 1 outcome vs trying to incorporate additional features

* Using arbitrary associations btw features and outcomes vs. relying on intuition (raised arms are dangerous)

