# The Generalized Context Model {#sec-categorization-exemplars}

```{r ch12_setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "80%",
  dpi        = 300
)

# Set to TRUE to rerun all Stan fits and heavy simulations; FALSE loads saved results.
regenerate_simulations <- FALSE
# Set to TRUE to rerun SBC, extensive recovery, exact LFO validation.
run_intensive_checks   <- regenerate_simulations

# Directory scaffolding
for (d in c("stan", "simdata", "simmodels")) {
  if (!dir.exists(d)) dir.create(d)
}

pacman::p_load(
  tidyverse,    # data manipulation and visualization
  here,         # robust project-relative paths
  future,       # parallel processing setup
  purrr,        # functional programming tools
  furrr,        # parallel functional programming
  patchwork,    # combining plots
  cmdstanr,     # interface to Stan
  posterior,    # tidy posterior arrays
  tidybayes,    # tidy extraction of draws
  bayesplot,    # Bayesian visualization
  loo,          # LOO-CV model comparison
  SBC,          # Simulation-Based Calibration Checks (ECDF-diff plots)
  priorsense,   # prior sensitivity via power-scaling
  MCMCpack      # rdirichlet
)

theme_set(theme_classic())

# Prior hyperparameters for sensitivity c
LOG_C_PRIOR_MEAN <- log(2)   # prior median c = 2
LOG_C_PRIOR_SD   <- 1.0      # covers c ≈ 0.27–14.8 at ±2 SD

# ── Shared diagnostic helper used throughout the chapter ─────────────────────
diagnostic_summary_table <- function(fit, params = NULL) {
  if (is.null(fit)) return(NULL)
  diag <- fit$diagnostic_summary(quiet = TRUE)

  s <- fit$summary(
    variables = params,
    "mean", "sd", "rhat", "ess_bulk", "ess_tail", "mcse_mean"
  )

  tibble(
    metric = c(
      "Divergences (zero tolerance)",
      "Max rank-normalised R-hat",
      "Min bulk ESS",
      "Min tail ESS",
      "Min E-BFMI",
      "Max MCSE / posterior SD"
    ),
    value = c(
      sum(diag$num_divergent),
      max(s$rhat,     na.rm = TRUE),
      min(s$ess_bulk, na.rm = TRUE),
      min(s$ess_tail, na.rm = TRUE),
      min(diag$ebfmi, na.rm = TRUE),
      max(s$mcse_mean / s$sd, na.rm = TRUE)
    ),
    threshold = c("== 0", "< 1.01", "> 400", "> 400", "> 0.2", "< 0.05"),
    pass = c(
      sum(diag$num_divergent)               == 0,
      max(s$rhat, na.rm = TRUE)             <  1.01,
      min(s$ess_bulk, na.rm = TRUE)         >  400,
      min(s$ess_tail, na.rm = TRUE)         >  400,
      min(diag$ebfmi, na.rm = TRUE)         >  0.2,
      max(s$mcse_mean / s$sd, na.rm = TRUE) <  0.05
    )
  )
}
```

> **📍 Where we are in the Bayesian modeling workflow:**
> Chapters 1–10 built the complete inference toolkit: parameter estimation
> (Ch. 4), the six-phase validation pipeline (Ch. 5), hierarchical models
> (Ch. 6), model comparison via LOO (Ch. 7), mixture models with discrete
> marginalization (Ch. 8), and hierarchical Bayesian cognition (Ch. 9–10).
> This and the following 2 chapters apply the full toolkit to *categorization*, 
> asking not only what parameters describe a participant's behavior, but which 
> *cognitive mechanism* best accounts for how people learn to assign objects
> to kinds.
> We begin with the exemplar approach: the Generalized Context Model (GCM).
> The following chapters will introduce prototype and rule-based models,
> and then compare all three approaches head-to-head using the full validation
> battery.
>
> Two important caveats.
> First, the categorization models learn, that is, are **path-dependent** —
> the likelihood at trial $t$ depends on the full sequence of stimuli on
> trials $1, \ldots, t-1$ through the exemplar memory store. Standard LOO-CV
> is therefore technically invalid here. We met this issue when discussing 
> sequential Bayes. Here we introduce leave-future-out cross-validation (LFO-CV;
> Bürkner, Gabry, & Vehtari, 2020).
> Second, the rule-based models we will discuss cannot really be fit using
> MCMC (we'll see why), so we need to introduce a new slower method: Approximate
> Bayesian Computation (ABC).

## Introduction to Categorization Models

### The Fundamental Problem of Categorization

Categorization is one of the most fundamental cognitive abilities that humans possess. From early childhood, we learn to organize the world around us into meaningful categories: distinguishing food from non-food, safe from dangerous, or one letter from another. This process of assigning objects to categories is complex, involving perception, memory, attention, and decision-making.

How do humans learn categories and make categorization decisions? This question has generated decades of research and theoretical debate in cognitive science.

### Why Model Categorization?

As we have seen in previous chapters, computational models offer powerful ways to formalize theories about categorization. By implementing these theories as computer algorithms, we can:

* Explore the consequences of different theoretical assumptions

* Make precise predictions about human behavior in categorization tasks

* Test competing theories against empirical data

In this chapter, we will explore three major approaches to modeling categorization:

* **Exemplar models** — which propose that categories are represented by storing individual examples

* **Prototype models** — which suggest categories are represented by an abstract "average" or central tendency

* **Rule-based models** — which posit that categories are represented by explicit rules or decision boundaries

Each approach captures different aspects of human categorization and has generated substantial empirical research. We'll implement computational versions of each model type, allowing us to compare their behavior and predictions.

### Historical Development of Categorization Models

#### Early Views: The Classical Approach

Early theories of categorization followed what is now called the classical view, where categories were defined by necessary and sufficient features. In this view, category membership is an all-or-nothing affair: an object either satisfies the logical criteria or it doesn't. Some examples from Aristotle: 

* a human is a rational animal. 

* a bird has feathers, wings, and a beak; and lays eggs.

* a fish lives in water; has scales; and breathes through gills.

You need all of these features to be part of one category, and having them makes you part of that category.

While intuitive, this approach struggled to explain many aspects of human categorization, such as:

* Graded category membership (some members seem "more typical" than others)

* Unclear boundaries between categories.

* Context-dependent categorization

* Family resemblance structures (where no single feature is necessary)

Think of the traditional definition of a species: a set of organisms that can interbreed and produce fertile offspring. However, there are many cases of organism A being able to interbreed with B, and B with C, but A and C cannot interbreed (for instance Californian Ensatina Salamanders have such phenomena). This creates a fuzzy boundary that the classical view cannot easily accommodate.

#### The Prototype Revolution

In the 1970s, Eleanor Rosch's pioneering work on prototypes challenged the classical view. She demonstrated that categories appear to be organized around central tendencies or "prototypes," with membership determined by similarity to these prototypes. Objects closer to the prototype are categorized more quickly and consistently.

#### The Exemplar Alternative

In the late 1970s and 1980s, researchers like Douglas Medin and Robert Nosofsky proposed that rather than abstracting prototypes, people might store individual exemplars of categories and make judgments based on similarity to these stored examples.

The Generalized Context Model (GCM), developed by Nosofsky, became the standard exemplar model. It proposes that categorization decisions are based on the summed similarity of a new stimulus to all stored exemplars of each category, weighted by attention to different stimulus dimensions.

#### Rule-Based Models

While similarity-based approaches gained prominence, other researchers argued that people sometimes use explicit rules for categorization (e.g. in scientific reasoning, but not only). Models like Bayesian particle filters, multinomial trees, COVIS and RULEX formalize these ideas, suggesting that rule learning and application form a core part of human categorization, particularly for well-defined categories.

#### Hybrid Approaches

More recent work has focused on hybrid models that incorporate elements of multiple approaches. Models like SUSTAIN [@love2004sustain] and ATRIUM [@erickson1998rules] propose that humans can flexibly switch between strategies or that different systems operate in parallel.

### A Systematization Framework for Categorization Models

To pedagogically navigate and compare the different approaches, we can deconstruct the cognitive process of categorization into a sequential six-step pipeline:

1. **Input Representation:** How are stimuli represented? (e.g., continuous geometry vs. discrete binary features).

2. **Attentional Processes:** How are features prioritized? Not all features are equally relevant given a task, and attention is often modeled as weights summing to 1.

3. **Intermediate Representations:** How is the category internally stored? (e.g., as raw exemplars, abstract prototypes, or explicit rules).

4. **Evidential Mechanisms:** How is the incoming stimulus compared to the representation to gather evidence? (e.g., calculating average similarity with an exponential decay over distance, or evidence accumulation).

5. **Decision Rules:** How is evidence translated into a choice? (e.g., maximum similarity, or a probabilistic Luce choice axiom).

6. **Learning:** How do representations, attentional weights, or rule structures update over time through exposure, memory accumulation, and feedback?

#### The Three Model Classes Through the Pipeline

We can map our three theoretical model classes directly onto this processing pipeline:

| Pipeline Stage | Exemplar Models (e.g., GCM) | Prototype Models | Rule-Based Models |
| :--- | :--- | :--- | :--- |
| **1. Input Representation** | Continuous spatial geometry or discrete features. | Continuous spatial geometry or discrete features. | Often discrete, logical features. |
| **2. Attentional Processes** | Continuous attention weights to dimensions. | Continuous attention weights to dimensions. | Selective attention weights allocated to specific rules or dimensions. |
| **3. Intermediate Representations**| **Raw/None:** All encountered exemplars are stored. | **Abstract:** Integrated into a summary centroid/average. | **Abstract:** Continuous space partitioned into discrete rule boundaries. |
| **4. Evidential Mechanisms** | Exponentially decaying similarity to *all* stored exemplars. | Exponentially decaying similarity to the *central prototype*. | Probability that the exemplar is generated by the current rule. |
| **5. Decision Rules** | Luce choice axiom (probabilistic assignment). | Luce choice axiom (probabilistic assignment). | Maximum similarity or probabilistic rule combination. |
| **6. Learning** | Accumulation of distinct exemplars in memory (path-dependent). | Updating the centroid/summary representation (e.g., via a Kalman filter). | Updating weights for sampled rules based on feedback. |

### Our Implementation Approach

We'll implement three models representing each major theoretical approach:

* **Generalized Context Model (GCM)** — A canonical exemplar model where categorization is based on summed similarity to all stored examples

* **Kalman Filter Prototype Model** — A dynamic prototype model (Ch. 12)

* **Bayesian Particle Filter for Rules** — A rule-based model (Ch. 13)

Each implementation follows a similar structure: core model function as a forward simulator, Stan implementation for parameter estimation, the MCMC diagnostic battery, prior/posterior predictive checks, parameter recovery, prior sensitivity, and SBC.

::: {.callout-note title="The Conceptual Need for Shared Implementations"}

A major challenge in evaluating cognitive theories is the "hidden implementation" problem. If different research groups write their own custom code for a GCM or a prototype model, subtle differences in how the math is translated into an algorithm can lead to conflicting theoretical conclusions. To genuinely compare models, the field conceptually requires shared, open-source implementations tested against standardized empirical benchmarks. 

A pioneering example of this movement is the work by @wills2017progress, who developed the `catlearn` R package to provide formal, collaboratively verified implementations of classic categorization models and canonical datasets.

These course notes are designed as a partial, pedagogical contribution to this open-science ethos. We do not adopt the specific software architecture or strict standardization of packages like `catlearn`, but we share the same fundamental goal: transparent, reproducible science. We show how to create, document and validate transparent bespoke scripts and Stan-based Bayesian inference. This means that you will not just adopt some secretly shared pre-made script, or struggle implementing an equation and then never share the script as you are secretely afraid it contains errors. Rather, you'll be able to implement and develop different algorithms, validate and document them, and share them with the world, so that others will be able to build on your work and we can do better science as a collective enterprise.

:::

### Categorization (Cognitive Science) vs. Classification (Machine Learning)

Because the mathematical tools are often identical, it is tempting to view cognitive models of categorization simply as early, inefficient machine learning classifiers. Both fields rely on the exact same underlying mechanics to map inputs to outputs: calculating distances in multi-dimensional feature space, applying functions (like Luce or softmax) to translate evidence into probabilities, and using gradient-based weight updates or Bayesian filters to learn from feedback. 

However, despite this shared mathematical framework, they serve fundamentally different goals:

* **Classification (Machine Learning):** The primary goal is *engineering*. It focuses on drawing the optimal decision boundary to maximize predictive accuracy and minimize loss on a given dataset. The internal mechanism (e.g., a deep neural network's hidden layers) doesn't need to resemble human thought—it just needs to work efficiently and accurately.

* **Categorization (Cognitive Modeling):** The primary goal is *reverse-engineering*. Categorization is about understanding the knowledge structures that enable a biological agent to make predictive inferences. We ask not only what parameters describe behavior, but which *cognitive mechanism* best accounts for how people actually learn, including the specific mistakes they make.

We can summarize this distinction along three key dimensions:

| Dimension | Classification (Machine Learning) | Categorization (Cognitive Modeling) |
| :--- | :--- | :--- |
| **Primary Goal** | Maximize out-of-sample accuracy. | Explain human behavior, including systematic human errors. |
| **Data Processing** | Often batch-processed over massive, static datasets. | Inherently **path-dependent**; learns sequentially trial-by-trial. |
| **Constraints** | Hardware limits and computational efficiency. | Psychological plausibility (bounded memory, selective attention, cognitive load). |

In short: A machine learning classifier asks, *"What is the best mathematical way to separate Category A from Category B?"* A cognitive model asks, *"How do humans separate A from B, and what are the internal mental representations that allow them to do so?"*

---

## The Generalized Context Model (GCM)

### Mathematical Foundations and Cognitive Principles

The Generalized Context Model (GCM), developed by Robert Nosofsky [@nosofsky1986attention] in the 1980s, represents one of the most influential exemplar-based approaches to categorization. Unlike models that abstract category information into prototypes or rules, the GCM proposes that people store individual exemplars in memory and make categorization decisions based on similarity to these stored examples.

#### The GCM Mapped to the Categorization Pipeline

The GCM translates the 6-step cognitive pipeline into a formal mathematical structure as follows:

* **1. Input Representation (Geometric Space):** Stimuli are represented as continuous points in a multi-dimensional psychological space.

* **2. Attentional Processes (Selective Attention):** Per-dimension weights $w$ (summing to 1) rescale the psychological space before distance is computed, prioritizing task-relevant features.

* **3. Intermediate Representations (Memory for Instances):** The model explicitly skips abstraction. Every encountered example is stored directly in memory with its category label.

* **4. Evidential Mechanisms (Similarity Computation):** Distance is calculated using a city-block metric (Manhattan distance). Similarity between a new stimulus and each stored exemplar is an exponentially decreasing function of their distance, governed by a sensitivity parameter $c$. Evidence is then aggregated as average similarity for each category.

* **5. Decision Rules (Choice Probability):** The decision relies on the Luce choice axiom. The probability of assigning a stimulus to a category is proportional to its summed similarity to exemplars of that category, modified by a baseline category bias $\beta$.
* **6. Learning (Memory Accumulation):** The model learns sequentially by adding each newly categorized and feedback-evaluated stimulus directly into its exemplar memory store, altering the evidential landscape for future trials.

::: {.callout-note}

### Historical Context: From Binary Logic to Psychological Space

The evolution of the Generalized Context Model (GCM) reflects a broader shift in cognitive science: moving from models constrained by computational simplicity toward those more capable of capturing the "messy" richness of human perception.

#### The Era of Binary Constraints
In the late 1970s, categorization research was dominated by the "small-set" paradigm. Stimuli were often constructed from a few discrete, binary features (e.g., "large vs. small," "red vs. blue"). The pioneering **Context Model** [@medin1978context] was designed for these environments, calculating similarity based on simple feature overlap. While revolutionary, it was computationally and theoretically "flat": it couldn't account for the fact that some dimensions are more perceptually salient than others, or that the psychological distance between features isn't always linear.

#### The Geometric Revolution

The introduction of the GCM [@nosofsky1986attention] broke these constraints by synthesizing exemplar theory with **Multidimensional Scaling (MDS)**. Instead of binary features, stimuli became points in a continuous, multi-dimensional psychological space. The core innovation was the weighted Minkowski metric:
$$d_{ij} = c \left[ \sum w_k |x_{ik} - x_{jk}|^r \right]^{1/r}$$
This formula allowed the model to "stretch" or "shrink" psychological space based on selective attention ($w_k$). It transformed categorization from a logical matching process into a geometric one, where the "geometry of thought" could be empirically mapped.

#### The "Ecological" Turn: Summed vs. Mean Similarity

As the field moved toward more "ecological" experiments with larger and unbalanced category sizes, a heated debate emerged. The original GCM relies on **summed similarity**: the total evidence for Category A is the sum of similarity to *every* stored exemplar. Critics [@smith1998prototypes; @smith2000thirty] argued that this creates a "category-size bias," where massive categories eventually overwhelm smaller ones regardless of typicality. 

Nosofsky [-@nosofsky1991tests] countered that humans actually *do* show base-rate biases, arguing that response-bias parameters ($\beta$) were mathematically sufficient to manage these frequency effects. However, a static $\beta$ can only compensate for an *average* frequency imbalance; it cannot fix the imbalance dynamically as the memory store grows and shrinks trial-by-trial. This conceptual limitation is exactly why our implementation adopts the **mean-similarity** formulation, which naturally neutralizes unbounded category-size inflation while keeping $\beta$ as a clean index of pure response bias.

#### The Process Revolution: Sequential Accumulation and Path-Dependency

Early models were "static": they were used to fit terminal, steady-state data from the end of an experiment. The conceptual shift toward process models began by acknowledging that categorization is a sequential learning experience. The critical first step is simply modeling memory as it grows: exemplars are accumulated into the memory store one trial at a time. 

This transition highlights the fundamentally **path-dependent** nature of cognition: your response at trial $t$ depends strictly on the specific sequence of exemplars and feedback you encountered in trials $1 \dots t-1$. While later models like **ALCOVE** [@kruschke1992alcove] attempted to model trial-by-trial, error-driven updates to the attention weights themselves, in my experience such extensions often yield terrible parameter recovery and fail stringent modern validation checks. Consequently, our approach grounds the "Process Revolution" in robust sequential exemplar accumulation. We then conceptually extend this path-dependency by introducing a memory decay parameter, providing a mechanism for agents to unlearn and adapt in non-stationary environments without sacrificing model identifiability.

:::

#### Parameters: Cognitive Commitment, Algebraic Role, Stan Parameterization

**Attention weights $w$.** Cognitively, $w$ is a simplex of dimension-specific attention weights. Algebraically, $w$ enters the distance formula as the per-dimension coefficient in the attention-weighted city-block metric. At the Stan level $w$ is declared as `simplex[N_features]`, which enforces $w_m \ge 0$ and $\sum_m w_m = 1$ directly.

**Sensitivity $c$.** Cognitively, $c$ represents perceptual discriminability. Algebraically, $c$ is the exponential decay rate in $\eta_{ij} = e^{-c\, d_{ij}}$. At the Stan level, we sample an unconstrained `real log_c` and set `c = exp(log_c)` in `transformed parameters`.

**Response bias $\beta$.** Cognitively, $\beta$ is a baseline preference for category A independent of stimulus similarity; $\beta = 0.5$ is unbiased. In this chapter $\beta$ also serves as the cold-start fallback. At the Stan level $\beta$ is declared as `real<lower=0, upper=1>` with a Beta prior.

### Mathematical Formulation

**1. Distance Calculation**

$$d_{ij} = \left[ \sum_{m=1}^{M} w_m |x_{im} - x_{jm}|^r \right]^{\frac{1}{r}}$$

We use the city-block metric ($r = 1$), appropriate for separable stimulus dimensions:

```{r ch12_distance_formula}
# City-block distance (r = 1), attention-weighted
distance <- function(vect1, vect2, w) {
  sum(w * abs(vect1 - vect2))
}
```

**2. Similarity Computation**

$$\eta_{ij} = e^{-c \cdot d_{ij}}$$

```{r ch12_similarity_formula}
# Exponential similarity kernel
similarity <- function(dist_val, c) {
  exp(-c * dist_val)
}
```

Let's visualize how similarity changes with distance for different values of $c$:

```{r ch12_visualizing_sensitivity}
dd <- expand_grid(
  distance  = seq(0, 6, by = 0.1),
  c         = c(0.3, 0.7, 1.5, 3.0, 6.0, 10.0)
) |>
  mutate(
    sim     = similarity(distance, c),
    c_label = factor(c)
  )

ggplot(dd, aes(x = distance, y = sim, color = c_label, group = c_label)) +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(option = "plasma", name = "Sensitivity (c)") +
  labs(
    title    = "Effect of Sensitivity Parameter (c) on Similarity",
    subtitle = "c is on the natural (original) scale; prior places most mass on c ∈ (0.3, 15)",
    x        = "Psychological Distance",
    y        = "Similarity"
  ) +
  theme(legend.position = "bottom")
```

**3. Category Response Probability**

$$P(A|i) = \frac{\beta \; \bar\eta_{i,A}}{\beta \; \bar\eta_{i,A} + (1-\beta) \; \bar\eta_{i,B}}, \qquad \bar\eta_{i,A} = \frac{1}{N_A} \sum_{j \in A} \eta_{ij}$$

where $\bar\eta_{i,A}$ is the **mean** similarity of the current stimulus to the exemplars stored under category $A$, and $N_A$ is the count of those stored exemplars. The cold-start fallback applies when either category has no stored exemplars yet: $P(A|i) = \beta$.

### Summed vs. Mean Similarity: A Caveat on the Canonical Formulation

The original GCM [@nosofsky1986attention] and the ALCOVE extensions [@kruschke1992alcove] use the *summed* form $\sum_{j \in A} \eta_{ij}$ rather than the mean. That choice carries a subtle but consequential assumption: whichever category has more stored exemplars acquires a systematic advantage, independent of how similar any of those exemplars is to the current stimulus.

A concrete failure. Suppose category A has 12 stored exemplars at distance roughly $d = 2$ from the current stimulus (so each contributes $\eta_{ij} \approx e^{-2c}$) and category B has 1 stored exemplar at distance $d = 0.2$ (contributing $\eta_{ij} \approx e^{-0.2c}$). At a moderate sensitivity $c = 1$:

* Summed: $S_A = 12 \cdot e^{-2} \approx 1.62$ vs $S_B = 1 \cdot e^{-0.2} \approx 0.82$. Category A wins.
* Mean: $\bar\eta_A = e^{-2} \approx 0.14$ vs $\bar\eta_B = e^{-0.2} \approx 0.82$. Category B wins, which is what our intuition expects: the closest neighbour belongs to B.

The bias parameter $\beta$ cannot repair this. Fitting $\beta$ to the data will compensate for the *average* frequency imbalance, but cannot fix it trial-by-trial because the imbalance itself grows and shrinks as the memory store evolves.

This chapter therefore uses the **mean-similarity** formulation throughout, which is the form adopted by @ashby1993relations and by several later papers that explicitly address this point. At $N_A = N_B$ (the canonical Kruschke task with balanced training) the two formulations produce effectively identical behaviour, so replication of published Kruschke-era results is unaffected. On data with unequal base rates, partial trial sequences, or the non-stationary scenarios introduced later in this chapter — where memory composition changes meaningfully over trials — the difference becomes both measurable and interpretable: with mean similarity, $\beta$ becomes a clean response-bias parameter rather than a fitted compensator for implicit base-rate effects.

```{r ch12_sum_vs_mean_demo}
# Toy demonstration of the summed-vs-mean choice with deliberately
# unbalanced memory contents. 12 cat-A exemplars far away, 1 cat-B
# exemplar close by. We show choice probabilities under both forms.
demo_memory <- tibble(
  category = c(rep(0, 12), 1),
  sim      = c(rep(exp(-2), 12), exp(-0.2))  # pre-computed for c = 1
)
demo_sums <- demo_memory |>
  group_by(category) |>
  summarise(sum_sim = sum(sim), mean_sim = mean(sim), .groups = "drop")
demo_sums |>
  mutate(
    p_cat1_summed = sum_sim  / sum(sum_sim),
    p_cat1_mean   = mean_sim / sum(mean_sim)
  ) |>
  print()
# Summed: P(cat 1) ≈ 0.34 — the 12 distant exemplars in category 0 win.
# Mean:   P(cat 1) ≈ 0.85 — the 1 close exemplar in category 1 correctly wins.
```

The demonstration above uses bias $\beta = 0.5$ (neutral) so the asymmetry comes purely from the aggregation choice. All figures, recovery results, and SBC checks in the rest of this chapter use the mean-similarity form. Readers who want to reproduce a published summed-similarity study can swap the aggregation line in `gcm_simulate()` (replace each `mean()` with `sum()`) and the corresponding two lines in the Stan model (remove the division by `n1` and `n0`) without any other changes.

### The GCM Agent Implementation (R Simulation)

```{r ch12_generative_gcm}
# Generative GCM model function.
gcm_simulate <- function(w,            # attention weights (sums to 1)
                         c,            # sensitivity (natural scale, strictly positive)
                         bias,         # response bias towards category 1
                         obs,          # matrix of stimulus features (trials x features)
                         cat_feedback, # true category labels (0 or 1)
                         quiet = TRUE) {
  ntrials   <- nrow(obs)
  nfeatures <- ncol(obs)

  prob_cat1    <- numeric(ntrials)
  sim_response <- numeric(ntrials)

  memory_obs <- matrix(NA_real_, nrow = 0, ncol = nfeatures)
  memory_cat <- numeric(0)

  for (i in seq_len(ntrials)) {
    if (!quiet && i %% 20 == 0) cat("Simulating trial:", i, "\n")

    current_stim <- as.numeric(obs[i, ])
    n_mem        <- nrow(memory_obs)
    has_cat0     <- any(memory_cat == 0)
    has_cat1     <- any(memory_cat == 1)

    if (n_mem == 0 || !has_cat0 || !has_cat1) {
      prob_cat1[i] <- bias
    } else {
      sims <- vapply(seq_len(n_mem), function(e) {
        similarity(distance(current_stim, memory_obs[e, ], w), c)
      }, numeric(1))

      # Mean similarity per category — removes category-frequency bias.
      m1  <- mean(sims[memory_cat == 1])
      m0  <- mean(sims[memory_cat == 0])
      num <- bias * m1
      den <- num + (1 - bias) * m0

      prob_cat1[i] <- if (den > 1e-9) num / den else bias
      prob_cat1[i] <- pmax(1e-9, pmin(1 - 1e-9, prob_cat1[i]))
    }

    sim_response[i] <- rbinom(1, 1, prob_cat1[i])

    memory_obs <- rbind(memory_obs, current_stim)
    memory_cat <- c(memory_cat, cat_feedback[i])
  }

  tibble(prob_cat1 = prob_cat1, sim_response = sim_response)
}
```

---

### Experimental Setup: Why @kruschke1993human?

To test our GCM implementation, we need a stimulus environment that is more complex than a simple linear separation but structured enough to reveal cognitive principles. We adopt the setup from @kruschke1993human for three specific reasons:

1.  **Continuous Dimension Benchmarking:** Unlike earlier binary-feature tasks (e.g., @medin1978context), Kruschke uses stimuli (rectangles varying in height and position) that occupy a continuous 2D psychological space. This is the natural environment for the GCM's Minkowski distance metrics.

2.  **Testing Selective Attention:** The task is designed so that both dimensions (height and position) are relevant, but their importance depends on the specific stimuli. This allows us to observe whether our model can "discover" the optimal attention weights ($w$) to minimize categorization error—a core feature of the ALCOVE extension to the GCM.

3.  **Non-Linearity & Abstraction:** The category structure (plotted below) is not perfectly separable by a single horizontal or vertical line. It requires the agent to either learn a diagonal boundary or to store the specific exemplars that form the "corners" of each category. This makes it a diagnostic benchmark for comparing exemplar models against the prototype models we'll see in Chapter 12.

```{r ch12_experimental_setup}
stimulus_info <- tibble(
  stimulus      = c(5, 3, 7, 1, 8, 2, 6, 4),
  height        = c(1, 1, 2, 2, 3, 3, 4, 4),
  position      = c(2, 3, 1, 4, 1, 4, 2, 3),
  category_true = c(0, 0, 1, 0, 1, 0, 1, 1)
)

ggplot(stimulus_info, aes(x = position, y = height,
                          color = factor(category_true), label = stimulus)) +
  geom_point(size = 5) +
  geom_text(nudge_y = 0.18) +
  scale_color_manual(values = c("0" = "#0072B2", "1" = "#D55E00"), name = "Category") +
  labs(title = "@kruschke1993human Stimulus Structure", x = "Position", y = "Height") +
  coord_fixed()

n_blocks         <- 8
n_stim_per_block <- nrow(stimulus_info)
total_trials     <- n_blocks * n_stim_per_block
```

---

### Per-Subject Schedule Generation

```{r ch12_subject_schedule}
make_subject_schedule <- function(stimulus_info, n_blocks, seed) {
  set.seed(seed)
  n_stim <- nrow(stimulus_info)

  sequence <- unlist(lapply(seq_len(n_blocks), function(b) {
    sample(stimulus_info$stimulus, n_stim, replace = FALSE)
  }))

  tibble(
    trial_within_subject = seq_along(sequence),
    block                = rep(seq_len(n_blocks), each = n_stim),
    stimulus_id          = sequence
  ) |>
    left_join(stimulus_info, by = c("stimulus_id" = "stimulus")) |>
    rename(category_feedback = category_true) |>
    dplyr::select(trial_within_subject, block, stimulus_id,
                  height, position, category_feedback)
}
```

---

### Simulating Behavior with the GCM

```{r ch12_gcm_agent_wrapper}
simulate_gcm_agent <- function(agent_id, w_true, c_true, bias_true,
                               stimulus_info, n_blocks, subject_seed) {
  schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  obs      <- schedule |> dplyr::select(height, position)
  feedback <- schedule$category_feedback

  sim_results <- gcm_simulate(
    w = w_true, c = c_true, bias = bias_true,
    obs = obs, cat_feedback = feedback, quiet = TRUE
  )

  schedule |>
    mutate(
      agent_id     = agent_id,
      w1_true      = w_true[1],
      w2_true      = w_true[2],
      c_true       = c_true,
      log_c_true   = log(c_true),
      bias_true    = bias_true,
      prob_cat1    = sim_results$prob_cat1,
      sim_response = sim_results$sim_response,
      correct      = as.integer(category_feedback == sim_response)
    ) |>
    group_by(agent_id) |>
    mutate(performance = cumsum(correct) / row_number()) |>
    ungroup()
}
```

```{r ch12_gcm_simulations}
param_grid <- expand_grid(
  w_setting    = list(c(0.5, 0.5), c(0.7, 0.3), c(0.9, 0.1)),
  c_setting    = c(0.5, 1.0, 2.0, 4.0, 5.0, 15.0),
  bias_setting = 0.5
) |>
  mutate(
    setting_id = row_number(),
    w_label    = map_chr(w_setting, ~paste(round(.x, 1), collapse = ",")),
    c_label    = paste0("c≈", round(c_setting, 1)),
    label      = paste0("w=(", w_label, "), ", c_label)
  )

n_agents_per_setting <- 5

simulation_plan <- param_grid |>
  uncount(n_agents_per_setting) |>
  mutate(agent_id = row_number())

sim_file <- here("simdata", "ch12_gcm_simulated_responses.csv")

if (regenerate_simulations || !file.exists(sim_file)) {
  plan(multisession, workers = max(1, availableCores() - 1))

  simulated_responses <- future_pmap_dfr(
    list(
      agent_id  = simulation_plan$agent_id,
      w_true    = simulation_plan$w_setting,
      c_true    = simulation_plan$c_setting,
      bias_true = simulation_plan$bias_setting
    ),
    simulate_gcm_agent,
    stimulus_info = stimulus_info,
    n_blocks      = n_blocks,
    subject_seed  = simulation_plan$agent_id,
    .options = furrr_options(seed = TRUE)
  )

  plan(sequential)
  write_csv(simulated_responses, sim_file)
  cat("Simulation complete. Results saved.\n")
} else {
  simulated_responses <- read_csv(sim_file, show_col_types = FALSE)
  cat("Loaded existing simulation results.\n")
}
```

```{r ch12_visualize_parameter_effects}
simulated_responses_labeled <- simulated_responses |>
  mutate(
    w_label = paste0("w=(", round(w1_true, 1), ",", round(w2_true, 1), ")"),
    c_label = paste0("c≈", round(c_true, 1)),
    # Order the c_label factor based on the numeric c_true values
    c_label = forcats::fct_reorder(c_label, c_true) 
  )

p_c_effect <- ggplot(
  simulated_responses_labeled,
  aes(x = trial_within_subject, y = performance,
      color = factor(round(c_true, 1)),
      group = interaction(agent_id, round(c_true, 1)))
) +
  stat_summary(fun = mean, geom = "line",
               aes(group = factor(round(c_true, 1))), linewidth = 1.2) +
  facet_wrap(~w_label) +
  scale_color_viridis_d(option = "plasma", name = "Sensitivity (c)") +
  labs(
    title    = "Effect of Sensitivity (c) on GCM Learning",
    subtitle = "Mean across simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme(legend.position = "bottom")

p_w_effect <- ggplot(
  simulated_responses_labeled,
  aes(x = trial_within_subject, y = performance,
      color = w_label, group = interaction(agent_id, w_label))
) +
  stat_summary(fun = mean, geom = "line", aes(group = w_label), linewidth = 1.2) +
  facet_wrap(~c_label) +
  scale_color_brewer(palette = "Set1", name = "Attention Weights") +
  labs(
    title    = "Effect of Attention Weights (w) on GCM Learning",
    subtitle = "Mean across simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme(legend.position = "bottom")

p_c_effect
p_w_effect
```

---

### Prior Predictive Check

We need to define the priors for the Stan model already here:

- $w \sim \text{Dirichlet}(1, 1)$ — uniform over the simplex
- $\log c \sim \text{Normal}(\log(2), 1)$ — implies $c \in (0.27, 14.8)$ at $\pm 2$ SD
- $\text{bias} \sim \text{Beta}(1, 1)$ — uniform over $(0, 1)$

```{r ch12_prior_predictive}
n_prior_samples <- 500

prior_draws <- tibble(
  sample_id = seq_len(n_prior_samples),
  log_c     = rnorm(n_prior_samples, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD),
  w1        = runif(n_prior_samples),
  bias      = rbeta(n_prior_samples, 1, 1)
) |>
  mutate(c_true = exp(log_c), w2 = 1 - w1)

ppc_schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = 999)

prior_pred_curves <- pmap_dfr(
  list(
    sample_id = prior_draws$sample_id,
    log_c     = prior_draws$log_c,
    w1        = prior_draws$w1,
    bias      = prior_draws$bias
  ),
  function(sample_id, log_c, w1, bias) {
    c <- exp(log_c)
    sim <- gcm_simulate(
      w = c(w1, 1 - w1), c = c, bias = bias,
      obs          = ppc_schedule |> dplyr::select(height, position),
      cat_feedback = ppc_schedule$category_feedback
    )
    tibble(
      sample_id = sample_id,
      trial     = seq_len(nrow(ppc_schedule)),
      correct   = as.integer(sim$sim_response == ppc_schedule$category_feedback)
    ) |>
      mutate(cum_acc = cumsum(correct) / row_number())
  }
)

prior_pred_summary <- prior_pred_curves |>
  group_by(trial) |>
  summarise(
    q05 = quantile(cum_acc, 0.05),
    q25 = quantile(cum_acc, 0.25),
    q50 = median(cum_acc),
    q75 = quantile(cum_acc, 0.75),
    q95 = quantile(cum_acc, 0.95),
    .groups = "drop"
  )

ggplot(prior_pred_summary, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#0072B2", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Prior Predictive Check: GCM Learning Curves",
    subtitle = "Ribbons: 50% and 90% prior predictive intervals; line: median\nPriors: log(c) ~ Normal(log 2, 1),  w ~ Dirichlet(1,1),  bias ~ Beta(1,1)",
    x = "Trial", y = "Cumulative Accuracy"
  )
```

---

### Implementing the GCM in Stan for Parameter Estimation

#### Architectural choice: `prob_cat1[i]` in `transformed parameters`

Because the GCM's per-trial choice probability $p_i$ is a deterministic function of the parameters $(w, c, \beta)$ and the data sequence, it belongs in `transformed parameters` as a single source of truth. The `model{}` block then becomes a one-line vectorised Bernoulli likelihood, and `generated quantities{}` becomes a lookup against the already-computed `prob_cat1` vector.

```{tikz gcm-single-dag, fig.cap = "DAG for the Single-Subject GCM. Hyperparameters (dots) feed into latent parameters (circles): attention weights $\\mathbf{w}$, log sensitivity $\\log c$, and response bias $\\beta$. Inside the plate, the stimulus $\\mathbf{x}_i$ and accumulated exemplar memory $z_{<i}$ (past category feedback) combine with the parameters to yield the deterministic choice probability $p_i$ (double circle), from which the response $y_i$ is drawn.", fig.align = "center", out.width = "70%"}
\usetikzlibrary{bayesnet}
\begin{tikzpicture}
  % ── Hyperparameters (const) ─────────────────────────────
  \node[const] (alpha_w) at (0, 0)   {$\boldsymbol{\alpha}_w$};
  \node[const] (hyp_c)   at (4, 0)   {$\mu_c,\,\sigma_c$};
  \node[const] (hyp_b)   at (8, 0)   {$\alpha_\beta,\,\beta_\beta$};

  % ── Latent parameters ───────────────────────────────────
  \node[latent] (w)    at (0, -2)  {$\mathbf{w}$};
  \node[latent] (logc) at (4, -2)  {$\log c$};
  \node[latent] (bias) at (8, -2)  {$\beta$};

  % ── Trial-level nodes (inside plate) ────────────────────
  \node[obs] (xi) at (2, -4.5) {$\mathbf{x}_i$};
  \node[obs] (zi) at (4, -4.5) {$z_{<i}$};
  \node[det] (pi) at (6, -4.5) {$p_i$};
  \node[obs] (yi) at (6, -6.5) {$y_i$};

  % ── Edges ───────────────────────────────────────────────
  \edge {alpha_w}        {w};
  \edge {hyp_c}          {logc};
  \edge {hyp_b}          {bias};
  \edge {w, logc, bias}  {pi};
  \edge {xi, zi}         {pi};
  \edge {pi}             {yi};

  % ── Plate ───────────────────────────────────────────────
  \plate {trial} {(xi)(zi)(pi)(yi)} {$N$ trials};
\end{tikzpicture}
```

The generative relationships for the Single-Subject GCM define how an agent's internal parameters and past experiences combine to produce a choice on each trial. We can break this down into four conceptual stages:

#### 1. Global Latent Parameters (Priors)

First, we define the "mental setting" of our agent. These parameters are constant throughout the experiment:

$$
\begin{aligned}
\mathbf{w} &\sim \text{Dirichlet}(\boldsymbol{\alpha}_w) \\
\log c &\sim \text{Normal}(\mu_c, \sigma_c) \\
\beta &\sim \text{Beta}(\alpha_\beta, \beta_\beta)
\end{aligned}
$$

*   **Attention Weights ($\mathbf{w}$):** A vector (simplex) that represents how much the agent cares about each stimulus dimension (e.g., height vs. position). The Dirichlet prior ensures these weights sum to 1.
*   **Sensitivity ($c$):** How quickly similarity drops off as distance increases. We sample $\log c$ to ensure $c$ remains strictly positive.
*   **Response Bias ($\beta$):** A baseline preference for Category 1. A $\beta = 0.5$ means no bias.

#### 2. The Perceptual Geometry (Distance & Similarity)

For every new stimulus $i$ on trial $t$, the agent compares it to every previously encountered exemplar $e$ stored in memory:

$$
\begin{aligned}
d_{ie} &= \sum_f w_f |x_{if} - x_{ef}| \\
s_{ie} &= \exp(-c \cdot d_{ie})
\end{aligned}
$$

*   **Attention-Weighted Distance ($d_{ie}$):** The agent calculates the distance between the current stimulus (with coordinates $x_{if}$) and a stored exemplar (with coordinates $x_{ef}$) across all dimensions $f$. Crucially, each dimension $f$ is weighted by $w_f$. If the agent ignores a dimension ($w_f \approx 0$), differences on that dimension don't contribute to the distance.
*   **The Similarity Kernel ($s_{ie}$):** Distance is converted into similarity using an exponential decay. If the distance is 0, similarity is 1. As distance grows, similarity decays towards 0 at a rate governed by sensitivity $c$.

#### 3. Evidence Aggregation (Mean Similarity)

The agent then tallies up the evidence for each category $k \in \{0, 1\}$ by averaging the similarity of the current stimulus to all stored exemplars of that category ($C_k$):

$$
m_{k,i} = \frac{1}{n_{k,i}} \sum_{e \in C_k} s_{ie}
$$

*   **Mean Similarity ($m_{k,i}$):** This is the average "resonance" the current stimulus has with Category $k$. 
*   **Indices:** Here, $C_k$ is the set of all exemplars $e$ that the agent has previously seen and categorized as belonging to Category $k$, and $n_{k,i}$ is the number of such exemplars currently in memory.
*   **Why Mean?** By using the mean (rather than the sum), we ensure the model isn't biased simply toward the category that has more exemplars in memory (see the "Summed vs. Mean" caveat above).

#### 4. The Decision Rule (Outcome)

Finally, the agent translates this evidence into a probability $p_i$ and makes a choice $y_i$:

$$
\begin{aligned}
p_i &= \frac{\beta \cdot m_{1,i}}{\beta \cdot m_{1,i} + (1 - \beta) \cdot m_{0,i}} \\
y_i &\sim \text{Bernoulli}(p_i)
\end{aligned}
$$

*   **Luce Choice Rule ($p_i$):** The probability of choosing Category 1 is the ratio of its evidence to the total evidence, weighted by the response bias $\beta$.
*   **The Choice ($y_i$):** The actual response is a coin flip (Bernoulli draw) with probability $p_i$.

```{r ch12_gcm_stan_model}
gcm_single_stan <- "
// Generalized Context Model — Single Subject (refactored architecture)

data {
  int<lower=1> N_total;
  int<lower=1> N_features;
  array[N_total] int<lower=0, upper=1> y;
  array[N_total, N_features] real obs;
  array[N_total] int<lower=0, upper=1> cat_feedback;

  vector[N_features] prior_w_alpha;
  real prior_log_c_mu;
  real<lower=0> prior_log_c_sigma;
  real<lower=0> prior_bias_alpha;
  real<lower=0> prior_bias_beta;
  int<lower=0, upper=1> run_diagnostics;
}

transformed data {
  array[N_total, N_total, N_features] real abs_diff;
  for (i in 1:N_total) {
    for (j in 1:N_total) {
      for (f in 1:N_features) {
        abs_diff[i, j, f] = abs(obs[i, f] - obs[j, f]);
      }
    }
  }
}

parameters {
  simplex[N_features] w;
  real log_c;
  real<lower=0, upper=1> bias;
}

transformed parameters {
  real<lower=0> c = exp(log_c);
  vector<lower=1e-9, upper=1-1e-9>[N_total] prob_cat1;

  {
    array[N_total] int memory_trial_idx;
    array[N_total] int memory_cat;
    int n_mem = 0;

    for (i in 1:N_total) {
      real p_i;
      int has_cat0 = 0;
      int has_cat1 = 0;

      for (k in 1:n_mem) {
        if (memory_cat[k] == 0) has_cat0 = 1;
        if (memory_cat[k] == 1) has_cat1 = 1;
      }

      if (n_mem == 0 || has_cat0 == 0 || has_cat1 == 0) {
        p_i = bias;
      } else {
        real s1 = 0; real s0 = 0;
        int n1 = 0;  int n0 = 0;
        for (e in 1:n_mem) {
          real d = 0;
          int past_i = memory_trial_idx[e];
          for (f in 1:N_features)
            d += w[f] * abs_diff[i, past_i, f];
          real sim = exp(-c * d);
          if (memory_cat[e] == 1) { s1 += sim; n1 += 1; }
          else                    { s0 += sim; n0 += 1; }
        }
        real m1 = s1 / n1;
        real m0 = s0 / n0;
        real num = bias * m1;
        real den = num + (1 - bias) * m0;
        p_i = (den > 1e-9) ? num / den : bias;
      }

      prob_cat1[i] = fmax(1e-9, fmin(1 - 1e-9, p_i));

      n_mem += 1;
      memory_trial_idx[n_mem] = i;
      memory_cat[n_mem] = cat_feedback[i];
    }
  }
}

model {
  target += dirichlet_lpdf(w    | prior_w_alpha);
  target += normal_lpdf(log_c   | prior_log_c_mu, prior_log_c_sigma);
  target += beta_lpdf(bias      | prior_bias_alpha,  prior_bias_beta);

  target += bernoulli_lpmf(y | prob_cat1);
}

generated quantities {
  vector[N_total] log_lik;
  real lprior;
  if (run_diagnostics) {
    for (i in 1:N_total)
      log_lik[i] = bernoulli_lpmf(y[i] | prob_cat1[i]);
  }
  lprior = dirichlet_lpdf(w | prior_w_alpha) +
           normal_lpdf(log_c | prior_log_c_mu, prior_log_c_sigma) +
           beta_lpdf(bias | prior_bias_alpha, prior_bias_beta);
}
"

stan_file_gcm_single <- "stan/ch12_gcm_single.stan"
write_stan_file(gcm_single_stan, dir = "stan/", basename = "ch12_gcm_single.stan")
mod_gcm_single <- cmdstan_model(stan_file_gcm_single)
```

#### What each block does:

* **`transformed data`**: Builds the `abs_diff[i, j, f]` tensor. Evaluated exactly once per chain.
* **`parameters`**: `simplex[N_features] w`, unconstrained `real log_c`, and `real<lower=0, upper=1> bias`.
* **`transformed parameters`**: Caches `prob_cat1[i]` in a single forward pass over trials.
* **`model`**: Priors followed by a vectorised one-line likelihood.
* **`generated quantities`**: `log_lik[i]` as a per-trial lookup, and `lprior` for `priorsense`.

> A note on reduce_sum and parallelization: The GCM's inner loop cannot be parallelized
> with reduce_sum because trial $t$'s contribution to target depends on the memory state
> accumulated sequentially from trials $1, \ldots, t-1$.

#### Fitting the Single-Subject GCM

```{r ch12_fitting_gcm_single}
# Select one simulated agent: w ≈ (0.9, 0.1), c ≈ 2
agent_to_fit <- simulated_responses |>
  filter(
    round(w1_true, 1) == 0.9,
    abs(log_c_true - log(2.0)) == min(abs(log_c_true - log(2.0)))
  ) |>
  filter(agent_id == min(agent_id)) |>
  slice(1:total_trials)

stopifnot(nrow(agent_to_fit) == total_trials)

gcm_data_single <- list(
  N_total           = nrow(agent_to_fit),
  N_features        = 2L,
  y                 = as.integer(agent_to_fit$sim_response),
  obs               = as.matrix(agent_to_fit[, c("height", "position")]),
  cat_feedback      = as.integer(agent_to_fit$category_feedback),
  prior_w_alpha     = c(1, 1),
  prior_log_c_mu    = LOG_C_PRIOR_MEAN,
  prior_log_c_sigma = LOG_C_PRIOR_SD,
  prior_bias_alpha  = 1,
  prior_bias_beta   = 1,
  run_diagnostics   = 1L
)

fit_filepath_single <- here("simmodels", "ch12_gcm_single_fit.rds")

if (regenerate_simulations || !file.exists(fit_filepath_single)) {
  pf_single <- tryCatch(
    mod_gcm_single$pathfinder(
      data      = gcm_data_single,
      seed      = 123,
      num_paths = 4,
      refresh   = 0
    ),
    error = function(e) {
      message("Pathfinder failed — falling back to default initialisation.")
      NULL
    }
  )

  fit_gcm_single <- mod_gcm_single$sample(
    data            = gcm_data_single,
    init            = pf_single,
    seed            = 123,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1000,
    iter_sampling   = 1500,
    refresh         = 300,
    adapt_delta     = 0.9
  )
  fit_gcm_single$save_object(fit_filepath_single)
  cat("Single-agent model fitted and saved.\n")
} else {
  fit_gcm_single <- readRDS(fit_filepath_single)
  cat("Loaded existing single-agent fit.\n")
}
```

#### MCMC Diagnostic Battery

```{r ch12_single_diagnostic_battery}
if (!is.null(fit_gcm_single)) {
  diag_tbl <- diagnostic_summary_table(
    fit_gcm_single,
    params = c("w[1]", "w[2]", "log_c", "c", "bias")
  )
  print(diag_tbl)

  if (!all(diag_tbl$pass)) {
    warning("Diagnostic battery has failures — interpret posteriors with caution.")
  }

  bayesplot::mcmc_trace(
    fit_gcm_single$draws(c("log_c", "w[1]", "bias")),
    facet_args = list(ncol = 1)
  )

  bayesplot::mcmc_rank_overlay(
    fit_gcm_single$draws(c("log_c", "w[1]", "bias"))
  )
}
```

#### Posterior Predictive Check (single subject)

```{r ch12_ppc_single}
if (!is.null(fit_gcm_single)) {
  prob_draws <- fit_gcm_single$draws("prob_cat1", format = "matrix")
  n_draws    <- nrow(prob_draws)
  n_trials   <- ncol(prob_draws)

  set.seed(2025)
  yrep_post <- matrix(rbinom(n_draws * n_trials, 1, as.vector(prob_draws)),
                      nrow = n_draws)

  obs_acc <- cummean(agent_to_fit$sim_response == agent_to_fit$category_feedback)
  rep_acc <- t(apply(yrep_post, 1, function(yr) {
    cummean(yr == agent_to_fit$category_feedback)
  }))

  ppc_summary <- tibble(
    trial = seq_len(n_trials),
    obs   = obs_acc,
    q05   = apply(rep_acc, 2, quantile, 0.05),
    q25   = apply(rep_acc, 2, quantile, 0.25),
    q50   = apply(rep_acc, 2, quantile, 0.50),
    q75   = apply(rep_acc, 2, quantile, 0.75),
    q95   = apply(rep_acc, 2, quantile, 0.95)
  )

  ggplot(ppc_summary, aes(x = trial)) +
    geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
    geom_line(aes(y = q50), color = "#0072B2", linewidth = 0.8) +
    geom_line(aes(y = obs), color = "#D55E00", linewidth = 1) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
      title    = "Posterior Predictive Check: Cumulative Accuracy",
      subtitle = "Blue ribbons: 50% and 90% posterior predictive intervals\nOrange line: observed cumulative accuracy",
      x = "Trial", y = "Cumulative Accuracy"
    )
}
```

#### Prior–Posterior Update Plot

```{r ch12_prior_posterior_single}
if (!is.null(fit_gcm_single)) {
  draws_df <- as_draws_df(
    fit_gcm_single$draws(variables = c("w[1]", "log_c", "c", "bias"))
  )

  true_w1    <- agent_to_fit$w1_true[1]
  true_log_c <- agent_to_fit$log_c_true[1]
  true_c     <- agent_to_fit$c_true[1]
  true_bias  <- agent_to_fit$bias_true[1]

  set.seed(2026)
  n_prior <- 4000
  prior_df <- tibble(
    `w[1]` = runif(n_prior),
    log_c  = rnorm(n_prior, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD),
    c      = exp(log_c),
    bias   = rbeta(n_prior, 1, 1)
  )

  pp_panel <- function(prior_x, post_x, true_val, label) {
    tibble(
      x      = c(prior_x, post_x),
      source = c(rep("Prior", length(prior_x)),
                 rep("Posterior", length(post_x)))
    ) |>
      ggplot(aes(x = x, fill = source, color = source)) +
      geom_density(alpha = 0.4, linewidth = 0.5) +
      geom_vline(xintercept = true_val, color = "red",
                 linetype = "dashed", linewidth = 0.7) +
      scale_fill_manual(values = c(Prior = "grey60", Posterior = "#0072B2")) +
      scale_color_manual(values = c(Prior = "grey40", Posterior = "#0072B2")) +
      labs(title = label, x = NULL, y = "Density") +
      theme(legend.position = "bottom")
  }

  p1 <- pp_panel(prior_df$`w[1]`, draws_df$`w[1]`, true_w1, "w[1]")
  p2 <- pp_panel(prior_df$log_c,  draws_df$log_c,  true_log_c, "log(c) — computational scale")
  p3 <- pp_panel(prior_df$c,      draws_df$c,      true_c,     "c — natural scale")
  p4 <- pp_panel(prior_df$bias,   draws_df$bias,   true_bias,  "bias")

  (p1 | p2) / (p3 | p4) +
    plot_annotation(
      title    = "Prior–Posterior Update for the Single-Subject GCM",
      subtitle = "Red dashed line = true generating value"
    )
}
```

#### Joint Posterior Geometry: $c$ vs $w$

```{r ch12_cw_identifiability_pairs}
if (!is.null(fit_gcm_single)) {
  bayesplot::mcmc_pairs(
    fit_gcm_single$draws(c("log_c", "w[1]", "bias")),
    diag_fun     = "dens",
    off_diag_fun = "hex"
  )
}
```

---

#### LOO with Pareto-$\hat{k}$ Diagnostics

Standard PSIS-LOO is computed for completeness and for comparison against LFO-CV below. We do not interpret it as a valid measure of out-of-sample predictive accuracy for a sequential model — see the next subsection for why — but the Pareto-$\hat{k}$ diagnostic still localises trials where the path-dependence concern bites hardest.

```{r ch12_loo_single}
if (!is.null(fit_gcm_single)) {
  log_lik_mat    <- fit_gcm_single$draws("log_lik", format = "matrix")
  loo_gcm_single <- loo::loo(log_lik_mat, cores = 4)
  print(loo_gcm_single)

  plot(loo_gcm_single, label_points = TRUE,
       main = "PSIS-LOO Pareto-k diagnostics: Single-subject GCM")
}
```

#### Leave-Future-Out Cross-Validation

**Why standard LOO is invalid here.** Standard PSIS-LOO assumes observations are conditionally exchangeable given the parameters. In the GCM, this assumption is violated: leaving out observation $t$ also conceptually removes the memory trace that observation $t$ contributed to trials $t+1, \ldots, T$.

**What LFO-CV does instead.** LFO-CV (Bürkner, Gabry, & Vehtari, 2020) estimates the *one-step-ahead* predictive density. The PSIS approximation refits only when the importance sampling deteriorates ($\hat{k} > k_\text{thresh}$) and uses importance reweighting otherwise.

```{r ch12_lfo_cv_implementation}
# ─────────────────────────────────────────────────────────────────────────────
# psis_lfo_gcm: PSIS-approximated leave-future-out CV for the GCM.
# Follows Bürkner, Gabry, & Vehtari (2020), Algorithm 1.
# ─────────────────────────────────────────────────────────────────────────────
psis_lfo_gcm <- function(stan_model,
                         full_data,
                         L = 16,
                         k_thresh = 0.7,
                         seed = 1,
                         verbose = FALSE) {

  fit_subset <- function(n_train, seed) {
    sub_data <- full_data
    sub_data$N_total      <- n_train
    sub_data$y            <- sub_data$y[1:n_train]
    sub_data$obs          <- sub_data$obs[1:n_train, , drop = FALSE]
    sub_data$cat_feedback <- sub_data$cat_feedback[1:n_train]

    pf  <- stan_model$pathfinder(data = sub_data, num_paths = 4, refresh = 0,
                                  seed = seed)
    fit <- stan_model$sample(
      data = sub_data, init = pf, seed = seed,
      chains = 2, parallel_chains = 2,
      iter_warmup = 600, iter_sampling = 800,
      refresh = 0, adapt_delta = 0.9
    )
    fit
  }

  predict_future <- function(fit, full_data) {
    full_n   <- length(full_data$y)
    ext_data <- full_data
    ext_data$N_total <- full_n

    gq_fit <- stan_model$generate_quantities(
      fitted_params   = fit,
      data            = ext_data,
      seed            = seed,
      parallel_chains = 2
    )
    gq_fit$draws("log_lik", format = "matrix")
  }

  T_total      <- length(full_data$y)
  elpd_lfo     <- numeric(T_total)
  elpd_lfo[]   <- NA_real_
  refits       <- integer(0)
  k_history    <- numeric(0)

  fit_curr     <- fit_subset(L, seed = seed)
  ll_full_curr <- predict_future(fit_curr, full_data)

  refits       <- c(refits, L)
  last_fit_n   <- L
  n_draws      <- nrow(ll_full_curr)

  ll_train_curr <- rowSums(ll_full_curr[, 1:L, drop = FALSE])

  for (t in (L + 1):T_total) {
    if (t - 1 == last_fit_n) {
      k_t          <- 0
      k_history[t] <- k_t
      lw           <- rep(-log(n_draws), n_draws)
    } else {
      ll_train_target <- rowSums(ll_full_curr[, 1:(t - 1), drop = FALSE])
      log_w           <- ll_train_target - ll_train_curr

      psis_obj     <- loo::psis(log_w, r_eff = NA)
      k_t          <- loo::pareto_k_values(psis_obj)
      k_history[t] <- k_t

      if (k_t > k_thresh) {
        if (verbose) cat("Refit at t =", t, "(k =", round(k_t, 3), ")\n")
        fit_curr      <- fit_subset(t - 1, seed = seed + t)
        ll_full_curr  <- predict_future(fit_curr, full_data)
        ll_train_curr <- rowSums(ll_full_curr[, 1:(t - 1), drop = FALSE])
        last_fit_n    <- t - 1
        refits        <- c(refits, last_fit_n)
        lw            <- rep(-log(n_draws), n_draws)
      } else {
        lw <- as.numeric(weights(psis_obj))
      }
    }

    lp          <- ll_full_curr[, t]
    elpd_lfo[t] <- matrixStats::logSumExp(lw + lp)
  }

  list(
    elpd_lfo       = sum(elpd_lfo, na.rm = TRUE),
    elpd_pointwise = elpd_lfo,
    refits         = refits,
    k_history      = k_history,
    L              = L
  )
}

lfo_filepath <- here("simdata", "ch12_lfo_single.rds")
if (regenerate_simulations || !file.exists(lfo_filepath)) {
  lfo_single <- psis_lfo_gcm(
    stan_model = mod_gcm_single,
    full_data  = gcm_data_single,
    L          = 16,
    k_thresh   = 0.7,
    seed       = 11,
    verbose    = TRUE
  )
  saveRDS(lfo_single, lfo_filepath)
} else {
  lfo_single <- readRDS(lfo_filepath)
}

cat("LFO-CV ELPD (single subject):", round(lfo_single$elpd_lfo, 2), "\n")
cat("Number of refits:",             length(lfo_single$refits),     "\n")
cat("Refit triggered at trials:",    lfo_single$refits,             "\n")

if (!is.null(fit_gcm_single)) {
  loo_pw  <- loo_gcm_single$pointwise[, "elpd_loo"]
  comp_df <- tibble(
    trial    = seq_along(loo_pw),
    elpd_loo = loo_pw,
    elpd_lfo = lfo_single$elpd_pointwise
  ) |>
    pivot_longer(starts_with("elpd_"), names_to = "method", values_to = "elpd")

  ggplot(comp_df, aes(x = trial, y = elpd, color = method)) +
    geom_line(alpha = 0.7) +
    geom_point(alpha = 0.5) +
    scale_color_manual(values = c(elpd_loo = "#0072B2", elpd_lfo = "#D55E00"),
                       labels = c("PSIS-LFO (correct)",
                                  "PSIS-LOO (invalid for sequential)")) +
    labs(
      title    = "Per-trial ELPD: LOO vs LFO-CV",
      subtitle = "Differences are largest where the path-dependence assumption fails most",
      x = "Trial", y = "Pointwise ELPD"
    ) +
    theme(legend.position = "bottom")
}
```

**Validation against exact one-step-ahead refits.**

```{r ch12_exact_lfo_validation}
exact_lfo_grid     <- seq(16, length(gcm_data_single$y) - 1, by = 8)
exact_lfo_filepath <- here("simdata", "ch12_exact_lfo_single.rds")

if (regenerate_simulations || !file.exists(exact_lfo_filepath)) {
  exact_results <- map_dfr(exact_lfo_grid, function(n_train) {
    sub_data <- gcm_data_single
    sub_data$N_total      <- n_train
    sub_data$y            <- sub_data$y[1:n_train]
    sub_data$obs          <- sub_data$obs[1:n_train, , drop = FALSE]
    sub_data$cat_feedback <- sub_data$cat_feedback[1:n_train]

    pf  <- mod_gcm_single$pathfinder(data = sub_data, num_paths = 4,
                                      refresh = 0, seed = n_train)
    fit <- mod_gcm_single$sample(
      data = sub_data, init = pf, seed = n_train,
      chains = 2, parallel_chains = 2,
      iter_warmup = 600, iter_sampling = 800,
      refresh = 0, adapt_delta = 0.9
    )

    ext <- gcm_data_single
    gq  <- mod_gcm_single$generate_quantities(
      fitted_params = fit, data = ext, seed = n_train, parallel_chains = 2
    )
    ll_target  <- gq$draws("log_lik", format = "matrix")[, n_train + 1]
    elpd_exact <- matrixStats::logSumExp(ll_target) - log(length(ll_target))

    tibble(trial = n_train + 1, elpd_exact = elpd_exact)
  })
  saveRDS(exact_results, exact_lfo_filepath)
} else {
  exact_results <- readRDS(exact_lfo_filepath)
}

psis_at_grid <- tibble(
  trial     = exact_results$trial,
  elpd_psis = lfo_single$elpd_pointwise[exact_results$trial]
)

agreement_df <- exact_results |>
  left_join(psis_at_grid, by = "trial")

ggplot(agreement_df, aes(x = elpd_exact, y = elpd_psis)) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_point(size = 3, color = "#0072B2") +
  labs(
    title    = "PSIS-LFO vs Exact One-Step-Ahead Refits",
    subtitle = "Points should fall on the y = x line if PSIS-LFO is calibrated",
    x = "Exact ELPD (refit per point)",
    y = "PSIS-LFO ELPD"
  ) +
  coord_fixed()
```

#### Randomised LOO-PIT

```{r ch12_randomized_loo_pit}
if (!is.null(fit_gcm_single)) {
  prob_draws  <- fit_gcm_single$draws("prob_cat1", format = "matrix")
  log_lik_mat <- fit_gcm_single$draws("log_lik",   format = "matrix")

  psis_obj <- loo::psis(-log_lik_mat, r_eff = NA)
  lw       <- weights(psis_obj, normalize = TRUE)

  y_obs <- gcm_data_single$y

  loo_p_cat1 <- vapply(seq_along(y_obs), function(t) {
    w <- exp(lw[, t] - matrixStats::logSumExp(lw[, t]))
    sum(w * prob_draws[, t])
  }, numeric(1))

  set.seed(2027)
  pit_vals <- ifelse(
    y_obs == 1,
    runif(length(y_obs), 1 - loo_p_cat1, 1),
    runif(length(y_obs), 0,              1 - loo_p_cat1)
  )

  ggplot(tibble(pit = pit_vals), aes(x = pit)) +
    geom_histogram(aes(y = after_stat(density)),
                   bins = 20, fill = "#0072B2", alpha = 0.7,
                   boundary = 0) +
    geom_hline(yintercept = 1, color = "red", linetype = "dashed") +
    scale_x_continuous(limits = c(0, 1)) +
    labs(
      title    = "Randomised LOO-PIT for the Single-Subject GCM",
      subtitle = "If the model is calibrated, PIT values are Uniform(0,1) (red dashed line)",
      x = "Randomised LOO-PIT", y = "Density"
    )
}
```

#### Prior Sensitivity Analysis

```{r ch12_prior_sensitivity_single}
if (!is.null(fit_gcm_single)) {
  ps <- priorsense::powerscale_sensitivity(fit_gcm_single)
  print(ps)

  priorsense::powerscale_plot_dens(
    fit_gcm_single,
    variables = c("log_c", "w[1]", "bias")
  )
}
```

---

### Single-Subject Simulation-Based Calibration Checks (SBC)

SBC checks the internal consistency and calibration of the entire Bayesian inference procedure across the *whole* prior. It asks: "If the true parameters really came from our specified priors, would our inference procedure produce posteriors that are statistically consistent with those priors?"

```{r ch12_sbc_single}
sbc_single_filepath <- here("simdata", "ch12_sbc_single_results.rds")

if (regenerate_simulations || !file.exists(sbc_single_filepath)) {
  n_sbc_iterations_single <- 500   # use >= 1000 for publication-quality SBC

  gcm_single_sbc_generator <- SBC_generator_function(
    function() {
      w_true     <- as.numeric(MCMCpack::rdirichlet(1, c(1, 1)))
      log_c_true <- rnorm(1, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD)
      c_true     <- exp(log_c_true)
      bias_true  <- rbeta(1, 1, 1)

      sbc_schedule <- make_subject_schedule(stimulus_info, n_blocks,
                                            seed = sample.int(1e8, 1))

      sim <- gcm_simulate(
        w            = w_true,
        c            = c_true,
        bias         = bias_true,
        obs          = sbc_schedule |> dplyr::select(height, position),
        cat_feedback = sbc_schedule$category_feedback
      )

      list(
        variables = list(
          `w[1]` = w_true[1],
          `w[2]` = w_true[2],
          log_c  = log_c_true,
          bias   = bias_true
        ),
        generated = list(
          N_total           = nrow(sbc_schedule),
          N_features        = 2L,
          y                 = as.integer(sim$sim_response),
          obs               = as.matrix(sbc_schedule[, c("height", "position")]),
          cat_feedback      = as.integer(sbc_schedule$category_feedback),
          prior_w_alpha     = c(1, 1),
          prior_log_c_mu    = LOG_C_PRIOR_MEAN,
          prior_log_c_sigma = LOG_C_PRIOR_SD,
          prior_bias_alpha  = 1,
          prior_bias_beta   = 1,
          run_diagnostics   = 1L
        )
      )
    }
  )

  gcm_single_sbc_backend <- SBC_backend_cmdstan_sample(
    mod_gcm_single,
    iter_warmup   = 1000,
    iter_sampling = 1000,
    chains        = 2,
    adapt_delta   = 0.9,
    refresh       = 0
  )

  gcm_single_datasets <- generate_datasets(
    gcm_single_sbc_generator,
    n_sims = n_sbc_iterations_single
  )

  sbc_results_single <- compute_SBC(
    gcm_single_datasets,
    gcm_single_sbc_backend,
    cache_mode     = "results",
    cache_location = here("simdata", "ch12_sbc_single_cache"),
    keep_fits      = FALSE
  )
  saveRDS(sbc_results_single, sbc_single_filepath)
  cat("SBC single-subject results computed and saved.\n")
} else {
  sbc_results_single <- readRDS(sbc_single_filepath)
  cat("Loaded existing SBC single-subject results.\n")
}

plot_ecdf_diff(sbc_results_single)
plot_rank_hist(sbc_results_single)

# ── Diagnostic issue investigation ───────────────────────────────────────────
backend_diags  <- sbc_results_single$backend_diagnostics
default_diags  <- sbc_results_single$default_diagnostics |>
  dplyr::select(sim_id, max_rhat, min_ess_to_rank)
true_params_wide <- sbc_results_single$stats |>
  dplyr::select(sim_id, variable, simulated_value) |>
  tidyr::pivot_wider(names_from = variable, values_from = simulated_value)

diag_df <- backend_diags |>
  dplyr::left_join(default_diags,    by = "sim_id") |>
  dplyr::left_join(true_params_wide, by = "sim_id") |>
  dplyr::mutate(has_issue = n_divergent > 0 | n_max_treedepth > 0 | max_rhat > 1.01)

p_diag_c_bias <- ggplot(diag_df, aes(x = log_c, y = bias, color = has_issue)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00")) +
  labs(title    = "Diagnostic Issues in Parameter Space",
       subtitle = "Orange points: divergence, max treedepth, or Rhat > 1.01",
       x = "True log(c)", y = "True Bias") +
  theme(legend.position = "bottom")

p_diag_w_c <- ggplot(diag_df, aes(x = `w[1]`, y = log_c, color = has_issue)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00")) +
  labs(title = "Attention vs Sensitivity", x = "True w[1]", y = "True log(c)") +
  theme(legend.position = "none")

print(p_diag_c_bias | p_diag_w_c)

# ── Parameter recovery on SBC simulations ────────────────────────────────────
recovery_df <- sbc_results_single$stats

p_sbc_recovery <- ggplot(recovery_df, aes(x = simulated_value, y = mean)) +
  geom_point(alpha = 0.4, color = "#0072B2") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~variable, scales = "free") +
  labs(
    title    = "Parameter Recovery Across SBC Simulations",
    subtitle = "Posterior mean vs True generated value",
    x = "True Value (simulated_value)",
    y = "Estimated Value (Posterior Mean)"
  ) +
  theme_bw()

print(p_sbc_recovery)
```

When the true log(c) is above 0 the model computes similarity using exponential decay: `exp(-c * distance)`. When c gets large, similarity drops to exactly zero for anything except perfectly identical stimuli, causing gradient collapse and HMC failures. The diagnostic scatter plots above reveal that divergences and Rhat violations cluster at high log(c) values, confirming that the prior mass above log(c) ≈ 2 (c ≈ 7–8) is unsupported by the sampler geometry.

**Practical constraint from SBC.** We treat c ≈ 12 as a hard ceiling: a sensitivity this extreme makes stimuli that differ by any measurable amount effectively identical, which is psychologically implausible for the feature ranges used here. We therefore require that the prior places negligible mass above c = 12, i.e., that the 99.85th percentile (mean + 3 SD on the log scale) stays below log(12) ≈ 2.49. For the single-subject model this is satisfied with `Normal(log(2), 1.0)`, whose +3 SD upper bound is exp(0.69 + 3.0) ≈ 20 — already at the boundary of acceptability. We carry this constraint forward explicitly into the multilevel parameterisation.

---

## Extending the GCM with Memory Decay

The canonical GCM weights every stored exemplar equally — a perfect, unbounded memory. That is pedagogically clean but psychologically implausible, and it becomes a real liability as soon as the learning environment is non-stationary: if the category structure changes halfway through the task, the unweighted sum of exemplars anchors the agent to contradictions it has no mechanism to discard. We add a single forgetting-rate parameter, $\lambda$, that downweights older exemplars by $e^{-\lambda (t - t_e)}$ before entering the similarity average:

$$\eta_{ie} = e^{-\lambda (i - t_e)} \cdot e^{-c\, d_{ie}}, \qquad m_{k,i} = \frac{\sum_{e \in C_k} \eta_{ie}}{\sum_{e \in C_k} e^{-\lambda(i - t_e)}}, \qquad p_{i} = \frac{\beta \cdot m_{1,i}}{\beta \cdot m_{1,i} + (1 - \beta) \cdot m_{0,i}}.$$

At $\lambda = 0$ the model reduces to the canonical GCM with mean similarity. At large $\lambda$, only the last few exemplars matter. This parallels the process-noise term $q$ in the Kalman prototype (Ch. 12): both are mechanisms for "unlearning" — the Kalman filter unlearns by letting covariance re-grow, the decay GCM unlearns by shrinking old weights.

Decay also complements the mean-similarity aggregation we adopted earlier in the chapter: the effective category "size" in the mean becomes $\sum_{e \in C_k} e^{-\lambda (i - t_e)}$ — a decay-weighted effective sample size, rather than a raw count. A category whose exemplars are mostly old gets a smaller effective count, which is exactly the right behaviour in a non-stationary environment where old evidence is less trustworthy. The canonical summed-similarity GCM has no such protection: a category whose exemplars are all old and now mostly irrelevant would still dominate by sheer count. Mean similarity + decay is the minimal combination that keeps the model honest on both dimensions.

We will test this extended model against three generating scenarios that load the decay parameter differently:

1. **Static Kruschke (baseline).** Fixed category labels, as in the canonical GCM analysis above. A low-$\lambda$ model is correct for this data; we expect the posterior to be pulled toward the left tail of the $\log\lambda$ prior and for the $c$/$\lambda$ pair to be partially non-identified (both can make the similarity average narrower).
2. **Smooth drift.** Category centres in feature space drift as an independent random walk. Labels are assigned trial-by-trial by proximity to the current (drifted) centres, so the category boundary migrates smoothly through the stimulus set. Moderate $\lambda$ is appropriate: agents that forget too slowly lag behind the moving boundary; agents that forget too fast throw away evidence the drift has not yet invalidated.
3. **Performance-contingent abrupt shifts.** After the agent reaches a streak of $k$ consecutive correct responses, all category labels flip. The agent must unlearn the old mapping and relearn the new one. A high-$\lambda$ model is optimal: it forgets the now-contradicted exemplars faster and recovers faster. Note that these shifts are *endogenous* to the agent's own behaviour — a noisier agent triggers shifts less often and accumulates less information about $\lambda$.

**Assumption match.** The GCM's similarity-to-exemplar architecture does not explicitly model drift or reversal — it just reweights *any* evidence in memory. That makes the static scenario a natural home (no drift to account for), the smooth-drift scenario a graceful degradation (decay is a reasonable approximation to "exponentially forget stale evidence"), and the abrupt-shift scenario a misspecified-but-useful case (no *abrupt* mechanism exists in the model; decay smooths the reversal into gradual forgetting, which is suboptimal for the task but still lets $\lambda$ be recovered from how fast the agent bounces back after each flip). We will see all three signatures in the validation results below.

### Prior on the Decay Parameter

```{r ch12_decay_prior_constants}
# Prior hyperparameters for log(lambda).
# Normal(-2, 1) gives median lambda ≈ 0.135 (half-life ≈ 5 trials).
# ±2 SD covers lambda ∈ (0.018, 1.0), spanning "effectively no decay"
# to "only the last trial or two matters".
LOG_LAMBDA_PRIOR_MEAN <- -2.0
LOG_LAMBDA_PRIOR_SD   <- 1.0
```

### R Simulator With Scenario Logic

The simulator loops over trials, computes the decay-weighted choice probability, draws a response, and determines the *observed* feedback according to the scenario. Memory stores the trial index at which each exemplar was encoded, so ages are exact at every step. For the contingent-shift scenario, the label-flip state machine updates *after* the response is drawn (so the current trial's feedback uses the label map that was in force at decision time).

```{r ch12_decay_simulator}
simulate_decay_scenario <- function(w, c, lambda, bias, schedule,
                                    scenario = c("static", "contingent_shift", "drift"),
                                    streak_target = 8,
                                    drift_traj    = NULL) {
  scenario <- match.arg(scenario)
  ntrials   <- nrow(schedule)
  obs_mat   <- as.matrix(schedule[, c("height", "position")])
  base_lab  <- schedule$category_feedback

  memory_obs   <- matrix(NA_real_, nrow = 0, ncol = 2)
  memory_cat   <- integer(0)
  memory_trial <- integer(0)

  label_flip <- FALSE
  streak     <- 0L

  prob_cat1         <- numeric(ntrials)
  sim_response      <- integer(ntrials)
  observed_feedback <- integer(ntrials)
  label_flip_trace  <- logical(ntrials)

  for (i in seq_len(ntrials)) {
    current_stim <- as.numeric(obs_mat[i, ])
    n_mem        <- nrow(memory_obs)

    # ── Decay-weighted *mean* similarity per category ───────────────────
    # Numerator:   Σ (decay_j · similarity_j) over exemplars in that category.
    # Denominator: Σ decay_j over exemplars in that category  (effective N).
    # Collapses to plain mean similarity when lambda = 0.
    if (n_mem == 0 || !any(memory_cat == 0) || !any(memory_cat == 1)) {
      p <- bias
    } else {
      d     <- as.numeric(abs(sweep(memory_obs, 2, current_stim, "-")) %*% w)
      age   <- i - memory_trial
      decay <- exp(-lambda * age)
      sim   <- exp(-c * d)
      wt    <- decay * sim
      s1    <- sum(wt[memory_cat == 1]);    s0 <- sum(wt[memory_cat == 0])
      w1    <- sum(decay[memory_cat == 1]); w0 <- sum(decay[memory_cat == 0])
      m1    <- if (w1 > 1e-12) s1 / w1 else 0
      m0    <- if (w0 > 1e-12) s0 / w0 else 0
      num   <- bias * m1
      den   <- num + (1 - bias) * m0
      p     <- if (den > 1e-12) num / den else bias
    }
    p <- pmax(1e-9, pmin(1 - 1e-9, p))
    prob_cat1[i]    <- p
    sim_response[i] <- rbinom(1, 1, p)

    # ── Observed feedback depends on the scenario ───────────────────────
    fb <- switch(scenario,
      static           = base_lab[i],
      contingent_shift = if (label_flip) 1L - base_lab[i] else base_lab[i],
      drift            = {
        d0 <- sum(abs(current_stim - drift_traj$mu0[i, ]))
        d1 <- sum(abs(current_stim - drift_traj$mu1[i, ]))
        as.integer(d1 < d0)
      }
    )
    observed_feedback[i] <- fb
    label_flip_trace[i]  <- label_flip

    # Update the streak/flip state for the contingent scenario
    if (scenario == "contingent_shift") {
      streak <- if (sim_response[i] == fb) streak + 1L else 0L
      if (streak >= streak_target) { label_flip <- !label_flip; streak <- 0L }
    }

    # Add the current trial to memory
    memory_obs   <- rbind(memory_obs, current_stim)
    memory_cat   <- c(memory_cat, fb)
    memory_trial <- c(memory_trial, i)
  }

  schedule |>
    mutate(
      prob_cat1         = prob_cat1,
      sim_response      = sim_response,
      observed_feedback = observed_feedback,
      label_flip        = label_flip_trace,
      correct           = as.integer(sim_response == observed_feedback)
    )
}
```

### Drift Trajectory Generator

The smooth-drift scenario needs a pre-computed trajectory of category centres. Each centre follows an independent bivariate Gaussian random walk with per-trial step SD `drift_sigma`. Defaults place the initial centres at the @kruschke1993human category means; `drift_sigma = 0.05` gives a cumulative SD of about 0.4 units over 64 trials — enough to migrate the boundary through roughly one stimulus cell.

```{r ch12_decay_drift_trajectory}
make_drift_trajectory <- function(ntrials, drift_sigma = 0.05, seed = 1) {
  set.seed(seed)
  mu0_init <- c(1.75, 3.25)   # Kruschke cat-0 mean (height, position)
  mu1_init <- c(3.25, 1.75)   # Kruschke cat-1 mean
  d0 <- apply(matrix(rnorm(ntrials * 2, 0, drift_sigma), ncol = 2), 2, cumsum)
  d1 <- apply(matrix(rnorm(ntrials * 2, 0, drift_sigma), ncol = 2), 2, cumsum)
  list(
    mu0 = sweep(d0, 2, mu0_init, "+"),
    mu1 = sweep(d1, 2, mu1_init, "+")
  )
}
```

### Scenario Agent Wrappers

```{r ch12_decay_agent_wrappers}
simulate_decay_agent <- function(agent_id, scenario, w, c, lambda, bias,
                                 subject_seed, streak_target = 8,
                                 drift_sigma = 0.05) {
  schedule  <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  drift_tr  <- if (scenario == "drift")
    make_drift_trajectory(nrow(schedule), drift_sigma, seed = subject_seed)
  else NULL

  out <- simulate_decay_scenario(
    w = w, c = c, lambda = lambda, bias = bias,
    schedule      = schedule,
    scenario      = scenario,
    streak_target = streak_target,
    drift_traj    = drift_tr
  )

  out |>
    mutate(
      agent_id       = agent_id,
      scenario       = scenario,
      w1_true        = w[1],
      c_true         = c,
      lambda_true    = lambda,
      log_c_true     = log(c),
      log_lambda_true = log(lambda),
      bias_true      = bias
    ) |>
    group_by(agent_id) |>
    mutate(cumulative_accuracy = cumsum(correct) / row_number()) |>
    ungroup()
}
```

### Visualising the Three Scenarios

We simulate 10 agents per cell of a grid that crosses three $\lambda$ values with the three scenarios and plot cumulative accuracy. Static data is where low $\lambda$ wins; drift data favours moderate $\lambda$ (track the moving boundary without throwing away still-relevant evidence); the contingent-shift data is where moderate-to-high $\lambda$ wins (decay speeds reversal recovery).

```{r ch12_decay_scenario_viz}
decay_viz_file <- here("simdata", "ch12_decay_scenario_viz.csv")

if (regenerate_simulations || !file.exists(decay_viz_file)) {
  viz_grid <- expand_grid(
    scenario    = c("static", "drift", "contingent_shift"),
    lambda_true = c(0.02, 0.2, 0.6),
    agent_id    = 1:10
  ) |>
    mutate(subject_seed = 10000 + row_number())

  viz_sim <- pmap_dfr(
    list(agent_id     = viz_grid$agent_id,
         scenario     = viz_grid$scenario,
         lambda       = viz_grid$lambda_true,
         subject_seed = viz_grid$subject_seed),
    function(agent_id, scenario, lambda, subject_seed) {
      simulate_decay_agent(
        agent_id     = agent_id,
        scenario     = scenario,
        w            = c(0.5, 0.5),
        c            = 2.0,
        lambda       = lambda,
        bias         = 0.5,
        subject_seed = subject_seed
      )
    }
  )
  write_csv(viz_sim, decay_viz_file)
} else {
  viz_sim <- read_csv(decay_viz_file, show_col_types = FALSE)
}

ggplot(viz_sim |> 
         mutate(scenario = factor(scenario, 
                                 levels = c("static", "drift", "contingent_shift"),
                                 labels = c("Static Kruschke", "Smooth drift", "Performance-contingent flips"))),
       aes(x = trial_within_subject, y = cumulative_accuracy,
           color = factor(lambda_true), group = interaction(agent_id, lambda_true))) +
  stat_summary(fun = mean, geom = "line",
               aes(group = factor(lambda_true)), linewidth = 1.1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.15,
               aes(fill = factor(lambda_true), group = factor(lambda_true)),
               color = NA) +
  facet_wrap(~scenario, ncol = 3) +
  scale_color_viridis_d(option = "plasma", name = expression(lambda)) +
  scale_fill_viridis_d(option  = "plasma", name = expression(lambda)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Learning Curves Across the Three Scenarios",
    subtitle = paste0("Mean ± SE across 10 agents per cell; c = 2, bias = 0.5, ",
                      "uniform attention"),
    x = "Trial", y = "Cumulative accuracy"
  ) +
  theme(legend.position = "bottom")
```

**Reading the scenario panels.** In the static panel, accuracy climbs monotonically to near-ceiling for all $\lambda$, with the lowest $\lambda$ reaching the highest asymptote—unsurprising, because any decay discards still-valid evidence. In the drift panel, the curves plateau below ceiling; the plateau is higher for agents with a $\lambda$ tuned to the drift rate, allowing them to track the moving boundary without being "anchored" by obsolete history.

In the contingent-shift panel, we see a characteristic sawtooth pattern, but it is somewhat muted by the averaging. High-$\lambda$ agents rebound faster after a flip, which means they reach the criterion earlier and flip more often. Because these flips happen at different trial numbers for different agents, averaging across them smooths out the sharp drops and recoveries into a noisy plateau.

**The Alignment Critique.** To see the true effect of $\lambda$ on cognitive flexibility, we must align the data to the moment of the shift. In non-stationary environments, averaging across absolute trial numbers often masks the very process we are interested in—the speed of adaptation.

```{r ch12_decay_reversal_alignment}
# Align contingent-shift data to the start of each new phase
reversal_viz <- viz_sim |>
  filter(scenario == "contingent_shift") |>
  group_by(agent_id, lambda_true) |>
  mutate(
    # Detect when the label mapping flipped
    shift_event = label_flip != lag(label_flip, default = first(label_flip)),
    phase_id    = cumsum(shift_event)
  ) |>
  group_by(agent_id, lambda_true, phase_id) |>
  mutate(
    trial_in_phase = row_number(),
    phase_accuracy = cumsum(correct) / trial_in_phase
  ) |>
  ungroup() |>
  # Focus on the first few reversals to see the learning dynamics
  filter(phase_id > 0, phase_id <= 4, trial_in_phase <= 40)

ggplot(reversal_viz, aes(x = trial_in_phase, y = phase_accuracy, 
                         color = factor(lambda_true), group = factor(lambda_true))) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.1, color = NA) +
  facet_wrap(~paste("Phase", phase_id), ncol = 4) +
  scale_color_viridis_d(option = "plasma", name = expression(lambda)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Learning Performance Aligned to Category Shifts",
    subtitle = "Cumulative accuracy reset at each reversal; contingent-shift scenario only",
    x = "Trials Since Shift", y = "Cumulative Phase Accuracy"
  ) +
  theme(legend.position = "bottom")
```

When aligned this way, the "recovery advantage" of high $\lambda$ becomes stark. While all agents collapse to chance (or below) immediately after a shift, agents with higher decay rates ($\lambda = 0.6$) shed their obsolete memory faster and return to high accuracy significantly earlier than those with low decay.

### The Decay GCM in Stan

The Stan implementation mirrors the canonical GCM chunk above, adding `log_lambda` as an unconstrained scalar parameter and multiplying the similarity term by the decay weight `exp(-lambda * age)` where `age = i - past_trial_idx`.

```{r ch12_decay_stan_compile}
stan_file_gcm_decay_single <- here("stan", "ch12_gcm_decay_single.stan")
mod_gcm_decay_single <- cmdstan_model(stan_file_gcm_decay_single)
```

See `stan/ch12_gcm_decay_single.stan` in the repository for the full source; the only structural change relative to `ch12_gcm_single.stan` is the decay term in the similarity average and the addition of `log_lambda` (with its Normal prior) to the parameters and model blocks.

### Fitting One Agent Per Scenario

We pick one representative agent per scenario with $\lambda = 0.2$, $c = 2$, bias $= 0.5$, and uniform attention, and fit the decay GCM to each.

```{r ch12_decay_fit_per_scenario}
decay_fit_file <- here("simmodels", "ch12_gcm_decay_single_fits.rds")

decay_single_fit_data <- list(
  static           = simulate_decay_agent(1, "static",
                                          w = c(0.5, 0.5), c = 2.0,
                                          lambda = 0.2, bias = 0.5,
                                          subject_seed = 20001),
  drift            = simulate_decay_agent(1, "drift",
                                          w = c(0.5, 0.5), c = 2.0,
                                          lambda = 0.2, bias = 0.5,
                                          subject_seed = 20002),
  contingent_shift = simulate_decay_agent(1, "contingent_shift",
                                          w = c(0.5, 0.5), c = 2.0,
                                          lambda = 0.2, bias = 0.5,
                                          subject_seed = 20003)
)

stan_data_from_agent <- function(agent_df) {
  list(
    N_total               = nrow(agent_df),
    N_features             = 2L,
    y                     = as.integer(agent_df$sim_response),
    obs                   = as.matrix(agent_df[, c("height", "position")]),
    cat_feedback          = as.integer(agent_df$observed_feedback),
    prior_w_alpha         = c(1, 1),
    prior_log_c_mu        = LOG_C_PRIOR_MEAN,
    prior_log_c_sigma     = LOG_C_PRIOR_SD,
    prior_log_lambda_mu   = LOG_LAMBDA_PRIOR_MEAN,
    prior_log_lambda_sigma = LOG_LAMBDA_PRIOR_SD,
    prior_bias_alpha      = 1,
    prior_bias_beta       = 1,
    run_diagnostics       = 1
  )
}

if (regenerate_simulations || !file.exists(decay_fit_file)) {
  decay_fits <- purrr::imap(decay_single_fit_data, function(agent_df, scen) {
    dat <- stan_data_from_agent(agent_df)
    pf  <- tryCatch(
      mod_gcm_decay_single$pathfinder(data = dat, num_paths = 4,
                                      refresh = 0, seed = 500),
      error = function(e) NULL
    )
    fit <- mod_gcm_decay_single$sample(
      data            = dat,
      init            = if (!is.null(pf)) pf else 0.5,
      seed            = 500,
      chains          = 4,
      parallel_chains = min(4, availableCores()),
      iter_warmup     = 1000,
      iter_sampling   = 1500,
      refresh         = 0,
      adapt_delta     = 0.9
    )
    fit$save_object(here("simmodels",
                         paste0("ch12_gcm_decay_single_", scen, ".rds")))
    fit
  })
  saveRDS(lapply(decay_fits, function(f) f$output_files()), decay_fit_file)
  cat("Decay GCM fits computed and saved.\n")
} else {
  decay_fits <- setNames(
    lapply(names(decay_single_fit_data), function(scen) {
      readRDS(here("simmodels",
                   paste0("ch12_gcm_decay_single_", scen, ".rds")))
    }),
    names(decay_single_fit_data)
  )
  cat("Loaded existing decay GCM fits.\n")
}
```

### MCMC Diagnostic Battery Across Scenarios

```{r ch12_decay_diagnostics}
decay_diag_tbls <- purrr::imap_dfr(decay_fits, function(fit, scen) {
  tbl <- diagnostic_summary_table(
    fit,
    params = c("w[1]", "log_c", "log_lambda", "bias")
  )
  tbl |> mutate(scenario = scen)
})

print(decay_diag_tbls |>
        dplyr::select(scenario, metric, value, threshold, pass))

walk(names(decay_fits), function(scen) {
  cat("=== Pair plot:", scen, "===\n")
  print(bayesplot::mcmc_pairs(
    decay_fits[[scen]]$draws(c("log_c", "log_lambda", "w[1]", "bias")),
    diag_fun     = "dens",
    off_diag_fun = "hex"
  ))
})
```

**What to watch for in the pair plot.** A `log_c`/`log_lambda` ridge is expected on static data — both narrow the effective similarity average, so the data under-constrain their difference. On drift and contingent-shift data, boundary-tracking and post-shift reversal dynamics inject trial-level evidence that the two parameters tension against, and the ridge should shrink toward a circle.

### Prior-Posterior Update Plots Across Scenarios

```{r ch12_decay_prior_posterior}
decay_pp_df <- purrr::imap_dfr(decay_fits, function(fit, scen) {
  d <- as_draws_df(fit$draws(c("log_c", "log_lambda", "w[1]", "bias")))
  bind_rows(
    tibble(scenario = scen, parameter = "log_c",      value = d$log_c,      source = "posterior"),
    tibble(scenario = scen, parameter = "log_lambda", value = d$log_lambda, source = "posterior"),
    tibble(scenario = scen, parameter = "w[1]",       value = d$`w[1]`,     source = "posterior"),
    tibble(scenario = scen, parameter = "bias",       value = d$bias,       source = "posterior")
  )
})

set.seed(11)
n_prior <- 4000
prior_ref <- bind_rows(
  tibble(parameter = "log_c",      value = rnorm(n_prior, LOG_C_PRIOR_MEAN,      LOG_C_PRIOR_SD),
         source = "prior"),
  tibble(parameter = "log_lambda", value = rnorm(n_prior, LOG_LAMBDA_PRIOR_MEAN, LOG_LAMBDA_PRIOR_SD),
         source = "prior"),
  tibble(parameter = "w[1]",       value = runif(n_prior),         source = "prior"),
  tibble(parameter = "bias",       value = rbeta(n_prior, 1, 1),   source = "prior")
) |>
  tidyr::crossing(scenario = c("static", "drift", "contingent_shift"))

truth_lines <- tibble(
  parameter = c("log_c", "log_lambda", "w[1]", "bias"),
  value     = c(log(2.0), log(0.2), 0.5, 0.5)
)

ggplot(bind_rows(prior_ref, decay_pp_df |> 
                   mutate(scenario = factor(scenario, levels = c("static", "drift", "contingent_shift")))),
       aes(x = value, fill = source, color = source)) +
  geom_density(alpha = 0.4, linewidth = 0.5) +
  geom_vline(data = truth_lines, aes(xintercept = value),
             color = "red", linetype = "dashed") +
  facet_grid(scenario ~ parameter, scales = "free") +
  scale_fill_manual(values  = c(prior = "grey70", posterior = "#009E73")) +
  scale_color_manual(values = c(prior = "grey50", posterior = "#006D4E")) +
  labs(
    title    = "Prior → Posterior Update: Decay GCM Across Scenarios",
    subtitle = "Red dashed line = true generating value used for the single-agent fit",
    x = NULL, y = "Density"
  )
```

**What we expect to see.** On static data, the `log_lambda` posterior is pulled *leftward* from its prior (data favour small $\lambda$) but remains wide — the truth ($\log 0.2 \approx -1.6$) is inside the posterior but not sharply peaked there, because everything a moderate $\lambda$ can do on stationary data can be approximated by slightly larger $c$. On drift data, the result is a moderately informative `log_lambda` posterior whose width depends on the drift amplitude. On contingent-shift data, the `log_lambda` posterior is sharply peaked near the truth and its correlation with `log_c` drops — the post-flip recovery rate depends specifically on how fast old exemplars are forgotten, an effect `c` cannot mimic.

### One-Shot Parameter Recovery Across Scenarios

We sweep a 3 × 3 grid of $(c, \lambda)$ truths, with 5 independent simulated datasets per cell per scenario (135 fits total), and report recovery scatters.

```{r ch12_decay_recovery_sweep}
decay_recovery_file <- here("simdata", "ch12_decay_recovery_results.rds")

run_decay_recovery_iter <- function(scenario, c_true, lambda_true, iteration) {
  subject_seed <- iteration * 10000 +
                   round(c_true * 100) +
                   round(lambda_true * 1000) +
                   switch(scenario, static = 1, drift = 2, contingent_shift = 3)

  agent_df <- simulate_decay_agent(
    agent_id     = iteration,
    scenario     = scenario,
    w            = c(0.5, 0.5),
    c            = c_true,
    lambda       = lambda_true,
    bias         = 0.5,
    subject_seed = subject_seed
  )
  dat <- stan_data_from_agent(agent_df)
  pf  <- tryCatch(
    mod_gcm_decay_single$pathfinder(data = dat, num_paths = 4,
                                    refresh = 0, seed = subject_seed),
    error = function(e) NULL
  )
  fit <- tryCatch(
    mod_gcm_decay_single$sample(
      data            = dat,
      init            = if (!is.null(pf)) pf else 0.5,
      seed            = subject_seed,
      chains          = 2,
      parallel_chains = 2,
      iter_warmup     = 800,
      iter_sampling   = 1000,
      refresh         = 0,
      adapt_delta     = 0.9
    ),
    error = function(e) { message("Fit error: ", e$message); NULL }
  )
  if (is.null(fit)) return(NULL)

  s <- fit$summary(variables = c("log_c", "log_lambda", "c", "lambda"))
  diag <- fit$diagnostic_summary(quiet = TRUE)
  tibble(
    scenario    = scenario,
    c_true      = c_true,
    lambda_true = lambda_true,
    iteration   = iteration,
    log_c_mean       = s$mean[s$variable == "log_c"],
    log_lambda_mean  = s$mean[s$variable == "log_lambda"],
    c_mean           = s$mean[s$variable == "c"],
    lambda_mean      = s$mean[s$variable == "lambda"],
    log_c_q5         = s$q5[s$variable == "log_c"],
    log_c_q95        = s$q95[s$variable == "log_c"],
    log_lambda_q5    = s$q5[s$variable == "log_lambda"],
    log_lambda_q95   = s$q95[s$variable == "log_lambda"],
    had_divergence   = any(diag$num_divergent > 0)
  )
}

if (regenerate_simulations || !file.exists(decay_recovery_file)) {
  recovery_plan <- expand_grid(
    scenario    = c("static", "drift", "contingent_shift"),
    c_true      = c(0.5, 2.0, 5.0),
    lambda_true = c(0.02, 0.2, 0.6),
    iteration   = 1:5
  )

  plan(multisession, workers = max(1, availableCores() - 1))
  decay_recovery <- future_pmap_dfr(
    list(
      scenario    = recovery_plan$scenario,
      c_true      = recovery_plan$c_true,
      lambda_true = recovery_plan$lambda_true,
      iteration   = recovery_plan$iteration
    ),
    run_decay_recovery_iter,
    .options = furrr_options(seed = TRUE, scheduling = 1)
  )
  plan(sequential)
  saveRDS(decay_recovery, decay_recovery_file)
  cat("Decay recovery sweep complete.\n")
} else {
  decay_recovery <- readRDS(decay_recovery_file)
  cat("Loaded existing decay recovery sweep.\n")
}

decay_recovery_long <- decay_recovery |>
  dplyr::select(scenario, c_true, lambda_true, iteration, had_divergence,
                log_c_mean, log_lambda_mean) |>
  pivot_longer(
    cols      = c(log_c_mean, log_lambda_mean),
    names_to  = "variable",
    values_to = "estimated"
  ) |>
  mutate(
    parameter = if_else(variable == "log_c_mean", "log_c", "log_lambda"),
    truth     = if_else(parameter == "log_c", log(c_true), log(lambda_true)),
    scenario  = factor(scenario, levels = c("static", "drift", "contingent_shift"))
  )

ggplot(decay_recovery_long, aes(x = truth, y = estimated, color = had_divergence)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.7, size = 2) +
  facet_grid(parameter ~ scenario, scales = "free") +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00"),
                     name = "Had divergence") +
  labs(
    title    = "One-Shot Parameter Recovery for the Decay GCM",
    subtitle = "Posterior mean vs true generating value; 5 iterations per (c, lambda) cell, 3 scenarios",
    x = "True value (log scale)", y = "Estimated value (posterior mean)"
  ) +
  theme(legend.position = "bottom")
```

**Interpretation.** The `log_c` row is near-diagonal across all three scenarios — the sensitivity parameter is well-identified regardless of task structure. The `log_lambda` row is the diagnostic: scatter under static data (the decay parameter is weakly identified, so posterior means regress toward the prior mean), considerably tighter under drift, and tightest under contingent-shift — the reversal dynamics load directly on $\lambda$. This is the empirical answer to the identifiability question from the chapter-12 planning discussion: the parameter that is partially confounded under a stationary task is cleanly recovered as soon as the task exercises the mechanism the parameter encodes.

### Simulation-Based Calibration Checks Across Scenarios

SBC is the calibration test proper: draw parameters from their priors, simulate data with those parameters, fit the model, and check that the posterior ranks of the true parameter are uniform. We run a separate SBC sweep for each scenario.

```{r ch12_decay_sbc_generator}
make_decay_sbc_generator <- function(scenario) {
  SBC_generator_function(
    function() {
      log_c_true      <- rnorm(1, LOG_C_PRIOR_MEAN,      LOG_C_PRIOR_SD)
      log_lambda_true <- rnorm(1, LOG_LAMBDA_PRIOR_MEAN, LOG_LAMBDA_PRIOR_SD)
      c_true          <- exp(log_c_true)
      lambda_true     <- exp(log_lambda_true)
      w1_true         <- runif(1)
      bias_true       <- rbeta(1, 1, 1)

      subject_seed <- sample.int(1e8, 1)
      agent_df <- simulate_decay_agent(
        agent_id     = 1,
        scenario     = scenario,
        w            = c(w1_true, 1 - w1_true),
        c            = c_true,
        lambda       = lambda_true,
        bias         = bias_true,
        subject_seed = subject_seed
      )

      list(
        variables = list(
          log_c      = log_c_true,
          log_lambda = log_lambda_true,
          `w[1]`     = w1_true,
          bias       = bias_true
        ),
        generated = stan_data_from_agent(agent_df)
      )
    }
  )
}

run_decay_sbc <- function(scenario, n_sims = 200) {
  cache_file <- here("simdata", paste0("ch12_decay_sbc_", scenario, ".rds"))
  if (!regenerate_simulations && file.exists(cache_file)) {
    cat("Loaded cached decay SBC for scenario:", scenario, "\n")
    return(readRDS(cache_file))
  }

  gen <- make_decay_sbc_generator(scenario)
  backend <- SBC_backend_cmdstan_sample(
    mod_gcm_decay_single,
    iter_warmup   = 800,
    iter_sampling = 1000,
    chains        = 2,
    adapt_delta   = 0.9,
    refresh       = 0
  )
  datasets <- generate_datasets(gen, n_sims = n_sims)
  res <- compute_SBC(
    datasets,
    backend,
    cache_mode     = "results",
    cache_location = here("simdata", paste0("ch12_decay_sbc_cache_", scenario)),
    keep_fits      = FALSE
  )
  saveRDS(res, cache_file)
  res
}

decay_sbc_static <- run_decay_sbc("static",           n_sims = 200)
decay_sbc_drift  <- run_decay_sbc("drift",            n_sims = 200)
decay_sbc_shift  <- run_decay_sbc("contingent_shift", n_sims = 200)
```

```{r ch12_decay_sbc_ecdf}
print(plot_ecdf_diff(decay_sbc_static) + ggtitle("SBC: static"))
print(plot_ecdf_diff(decay_sbc_drift)  + ggtitle("SBC: smooth drift"))
print(plot_ecdf_diff(decay_sbc_shift)  + ggtitle("SBC: performance-contingent shifts"))

print(plot_rank_hist(decay_sbc_static) + ggtitle("Rank histograms — static"))
print(plot_rank_hist(decay_sbc_drift)  + ggtitle("Rank histograms — drift"))
print(plot_rank_hist(decay_sbc_shift)  + ggtitle("Rank histograms — contingent shifts"))
```

### Parameter Recovery Across SBC Simulations

```{r ch12_decay_sbc_recovery_plots}
combine_sbc_stats <- function(sbc_obj, scenario) {
  sbc_obj$stats |> mutate(scenario = scenario)
}

decay_sbc_stats <- bind_rows(
  combine_sbc_stats(decay_sbc_static, "static"),
  combine_sbc_stats(decay_sbc_drift,  "drift"),
  combine_sbc_stats(decay_sbc_shift,  "contingent_shift")
) |>
  dplyr::filter(variable %in% c("log_c", "log_lambda", "w[1]", "bias")) |>
  mutate(scenario = factor(scenario, levels = c("static", "drift", "contingent_shift")))

ggplot(decay_sbc_stats, aes(x = simulated_value, y = mean)) +
  geom_point(alpha = 0.35, color = "#0072B2") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_grid(variable ~ scenario, scales = "free") +
  labs(
    title    = "SBC-Level Parameter Recovery: Decay GCM",
    subtitle = "Posterior mean vs true generating value, pooled across SBC simulations",
    x = "True value", y = "Posterior mean"
  )

decay_sbc_stats |>
  group_by(scenario, variable) |>
  summarise(
    sbc_mean_bias = mean(mean - simulated_value, na.rm = TRUE),
    sbc_rmse      = sqrt(mean((mean - simulated_value)^2, na.rm = TRUE)),
    sbc_corr      = cor(mean, simulated_value,
                        use = "complete.obs", method = "pearson"),
    n_sims        = dplyr::n(),
    .groups       = "drop"
  ) |>
  print(n = Inf)
```

### Discussion: Task Structure Governs Identifiability

The three SBC panels give the cleanest summary of the argument that opens Ch. 12. Under static data, the `log_lambda` ECDF-difference curve drifts outside the simultaneous band and the rank histogram is non-uniform — the posterior does not contain the truth at the nominal rate because the data carry almost no information about forgetting. The `log_c` curve is well-behaved on the same data: the sensitivity parameter is identified regardless. Under drift and contingent-shift data, both curves return inside the band and the rank histograms are flat.

This is the identifiability argument as an empirical claim rather than an algebraic one. The decay GCM's *parameter* is always the same. Its *identifiability* is a joint property of the model and the task: when the task exercises forgetting — whether through a drifting boundary or through performance-contingent reversals — the model recovers $\lambda$ cleanly; when it does not, $\lambda$ is indistinguishable from a slightly wider similarity kernel.

A secondary observation: the scatter of posterior means on the static panel still clusters near the identity line in the SBC recovery plot, even though the SBC calibration curve fails. Point recovery and coverage calibration are different checks — a model can be right "on average" while miscalibrating its uncertainty. SBC is the one that catches the latter, which is why we run it rather than stopping at the recovery plot.

**How the three scenarios match the model's assumptions.** The decay GCM explicitly assumes exemplars decay exponentially as a function of trial age and have no other way of being overridden. The *static* scenario violates nothing but offers no signal for $\lambda$. The *smooth-drift* scenario is a reasonable match: exponentially downweighting past exemplars is a defensible approximation to "evidence collected under a different category mean is stale." The *contingent-shift* scenario is a genuine misspecification — the true data-generating process has abrupt reversals that no smooth decay can exactly reproduce. What saves recovery there is that the post-shift dynamics depend specifically on how fast old evidence can be effectively ignored, and the decay rate is the only parameter in the model that controls that speed. The posterior predictive checks at scenario transitions (not plotted here) should reveal systematic underperformance right after each flip — the reader is encouraged to add one and confirm. That residual misfit is the handle for a future extension with an explicit change-point mechanism.

### The $c$/$\lambda$ Joint Posterior: Ridge Geometry and Task Resolution

The pair plots from the single-agent scenario fits revealed a `log_c`/`log_lambda` ridge on static data. Understanding the geometry of this ridge — and why task structure resolves it — is worth making explicit, because Chapter 12 faces a structurally similar problem ($q$/$r$ in the Kalman filter) and resolves it differently.

**Why does the ridge form?** Both $c$ and $\lambda$ narrow the effective contribution of any given exemplar to the category mean:

* A high $c$ shrinks contributions from spatially distant exemplars (similarity decays faster with distance).
* A high $\lambda$ shrinks contributions from temporally distant exemplars (similarity decays faster with trial age).

Under a stationary task with balanced categories, distance and age are nearly uncorrelated — older exemplars are not systematically farther from the current stimulus. This means that increasing $c$ and decreasing $\lambda$ (or vice versa) can produce nearly the same observed choice probabilities, creating the elongated posterior ridge you see in the pair plots.

**Why is the GCM ridge less severe than the Kalman filter's $q$/$r$ ridge?** The two parameters operate on orthogonal dimensions: $c$ operates on the spatial axis (feature distances), $\lambda$ on the temporal axis (trial age). Even under static data, choice probabilities have some leverage on each independently — $c$ is constrained by how performance differs across stimuli at varying spatial distances, while $\lambda$ is partially constrained by the early-learning transient (the model must form initial memories before it can categorize, and the learning rate is $\lambda$-dependent). In the Kalman filter, $q$ and $r$ both scale the *same* likelihood at decision time (the multivariate normal variance at trial $t$ is a function of both), so their directions are more nearly collinear and the ridge is typically more severe. Chapter 12 introduces an explicit reparameterization (the SNR parameter $\rho = q/r$) to address this; the GCM does not need it here.

**Task structure as the primary handle.** The SBC results above confirm that the contingent-shift scenario cleanly resolves $\lambda$ without any reparameterization. The reversal dynamics create a temporal structure that spatial sensitivity cannot account for: the *time* at which performance drops is determined by $\lambda$ (how fast old, now-wrong exemplars are forgotten) in a way that is independent of $c$ (which only governs how much nearby vs. distant exemplars are weighted). This is the key empirical signature of a parameter that is theoretically identifiable but under-constrained by a particular task design: swap in a task that exercises the mechanism, and identifiability follows.

**When would reparameterization be needed for the GCM?** If you needed to run the decay GCM exclusively on static data — for example, when fitting archival datasets collected before non-stationary paradigms were standard — the ridge would be a genuine sampling problem. In that case, a reparameterization analogous to Ch. 12's SNR approach could help: write $\log s = \log c + \log \lambda$ (the overall "forgetting rate" in combined spatial-temporal units) and $\log d = \log c - \log \lambda$ (the relative weighting of spatial vs. temporal decay). The joint posterior of $(\log s, \log d)$ would have a narrower marginal in the $\log s$ direction (identified by overall accuracy) and a wider marginal in the $\log d$ direction (the degenerate direction). On non-stationary data the present parameterisation performs well without this change, so we do not implement it here — but the reader should recognise the family resemblance with Ch. 12's solution.

---

## The Multilevel Decay GCM: Scaling to populations and changing environments

Having demonstrated that memory decay is essential for non-stationary environments and that its parameters are identifiable given the right task structure, we now combine these mechanisms into a full hierarchical model. The **Multilevel Decay GCM** allows us to estimate individual cognitive differences while sharing information across the population to constrain our estimates. This is our "Final Boss" model: population-aware, psychologically realistic, and robust to environmental change.

### Mathematical Formulation: The Hierarchical Architecture

This model uses log-normal hierarchies for sensitivity and forgetting, and logit-normal hierarchies for attention and response bias. With two features, the simplex-valued attention vector $\mathbf{w}_j$ is fully determined by $w_{j,1}$, so the natural simplex hierarchy (Dirichlet around a population mean $\bar{\mathbf{w}}$ with concentration $\kappa$) collapses to a Beta on $w_{j,1}$. We use a **logit-normal** in its place — equivalent in support and shape to the Beta, but its non-centered parameterisation samples much more cleanly than a Dirichlet hierarchy in NUTS, which is what `ch12_gcm_ml_decay.stan` implements.

**Population-level distributions (Priors):**

$$
\begin{aligned}
\bar{\mu}_{\text{logit}\,w_1}, \bar{\sigma}_{\text{logit}\,w_1} &\sim \text{Priors} \\
\bar{\mu}_{\log c}, \bar{\sigma}_{\log c} &\sim \text{Priors} \\
\bar{\mu}_{\log \lambda}, \bar{\sigma}_{\log \lambda} &\sim \text{Priors} \\
\bar{\mu}_{\text{logit}\,\beta}, \bar{\sigma}_{\text{logit}\,\beta} &\sim \text{Priors}
\end{aligned}
$$

**Individual-level parameters (for subject $j$, with $w_{j,2} = 1 - w_{j,1}$):**

$$
\begin{aligned}
\text{logit}(w_{j,1}) &\sim \text{Normal}(\bar{\mu}_{\text{logit}\,w_1}, \bar{\sigma}_{\text{logit}\,w_1}) \\
\log c_j &\sim \text{Normal}(\bar{\mu}_{\log c}, \bar{\sigma}_{\log c}) \\
\log \lambda_j &\sim \text{Normal}(\bar{\mu}_{\log \lambda}, \bar{\sigma}_{\log \lambda}) \\
\text{logit}(\beta_j) &\sim \text{Normal}(\bar{\mu}_{\text{logit}\,\beta}, \bar{\sigma}_{\text{logit}\,\beta})
\end{aligned}
$$

Each individual-level scalar uses a non-centered offset (`z ~ N(0,1)`, with the actual parameter equal to `pop_mean + pop_sd * z`) to avoid the funnel that direct centred parameterisations produce when subject-level data are weakly informative.

**Trial-level Generative Process (for trial $i$ of subject $j$):**

$$
\begin{aligned}
d_{iej} &= \sum_f w_{jf} |x_{if} - x_{ef}| \\
\eta_{iej} &= e^{-\lambda_j(i - t_e)} \cdot e^{-c_j d_{iej}} \\
m_{k,ij} &= \frac{\sum_{e \in C_k} \eta_{iej}}{\sum_{e \in C_k} e^{-\lambda_j(i - t_e)}} \\
p_{ij} &= \frac{\beta_j \cdot m_{1,ij}}{\beta_j \cdot m_{1,ij} + (1 - \beta_j) \cdot m_{0,ij}} \\
y_{ij} &\sim \text{Bernoulli}(p_{ij})
\end{aligned}
$$

*   **Shrinkage:** The model uses the population means to "pull" uncertain individual estimates toward the group average. This is especially useful for $\lambda_j$ in scenarios where an individual's data might be locally uninformative.
*   **Effective Sample Size:** The denominator of $m_{k,ij}$ is now the **decay-weighted effective number of exemplars**, ensuring that old, forgotten evidence doesn't inflate the mean similarity.

### Simulating Data for the Multilevel Decay GCM

We simulate three small populations of agents — one per scenario (static, smooth drift, performance-contingent shifts) — using the **same** population-level parameters across scenarios. Subject-level parameters are drawn around the population means using the logit-normal/log-normal hierarchies that match the Stan model exactly. Because the simulator and the inference model now share a parameterisation, recovery scatters can be read directly without translating between Dirichlet concentration and logit SD.

```{r ch12_simulate_multilevel_decay_data}
simulate_multilevel_decay_gcm <- function(n_subjects, stimulus_info, n_blocks,
                                          scenario,
                                          base_seed,
                                          pop_logit_w_mean,    pop_logit_w_sd,
                                          pop_log_c_mean,      pop_log_c_sd,
                                          pop_log_lambda_mean, pop_log_lambda_sd,
                                          pop_logit_bias_mean, pop_logit_bias_sd) {
  set.seed(base_seed)
  ind_logit_w    <- rnorm(n_subjects, pop_logit_w_mean,    pop_logit_w_sd)
  ind_w1         <- plogis(ind_logit_w)
  ind_log_c      <- rnorm(n_subjects, pop_log_c_mean,      pop_log_c_sd)
  ind_log_lambda <- rnorm(n_subjects, pop_log_lambda_mean, pop_log_lambda_sd)
  ind_logit_bias <- rnorm(n_subjects, pop_logit_bias_mean, pop_logit_bias_sd)

  true_params <- tibble(
    agent_id        = seq_len(n_subjects),
    scenario        = scenario,
    logit_w_true    = ind_logit_w,
    w1_true         = ind_w1,
    w2_true         = 1 - ind_w1,
    log_c_true      = ind_log_c,
    c_true          = exp(ind_log_c),
    log_lambda_true = ind_log_lambda,
    lambda_true     = exp(ind_log_lambda),
    logit_bias_true = ind_logit_bias,
    bias_true       = plogis(ind_logit_bias)
  )

  all_agent_data <- vector("list", n_subjects)
  for (j in seq_len(n_subjects)) {
    subject_seed <- base_seed * 1000 + j
    all_agent_data[[j]] <- simulate_decay_agent(
      agent_id     = j,
      scenario     = scenario,
      w            = c(ind_w1[j], 1 - ind_w1[j]),
      c            = exp(ind_log_c[j]),
      lambda       = exp(ind_log_lambda[j]),
      bias         = plogis(ind_logit_bias[j]),
      subject_seed = subject_seed
    )
  }

  list(
    data        = bind_rows(all_agent_data),
    true_params = true_params
  )
}

n_subjects_sim <- 10
pop_params_sim <- list(
  pop_logit_w_mean    = 0,        # median w1 ≈ 0.5
  pop_logit_w_sd      = 0.7,      # ±2SD covers w1 ∈ (0.20, 0.80)
  pop_log_c_mean      = log(2.0), # population median c = 2
  pop_log_c_sd        = 0.4,
  pop_log_lambda_mean = log(0.2), # population median λ = 0.2
  pop_log_lambda_sd   = 0.5,
  pop_logit_bias_mean = 0,
  pop_logit_bias_sd   = 0.3
)

ml_scenarios <- c("static", "drift", "contingent_shift")

ml_decay_files <- map(ml_scenarios, function(scen) list(
  data  = here("simdata", paste0("ch12_gcm_ml_decay_data_",  scen, ".csv")),
  truth = here("simdata", paste0("ch12_gcm_ml_decay_truth_", scen, ".csv"))
)) |> setNames(ml_scenarios)

ml_files_exist <- all(file.exists(unlist(map(ml_decay_files, ~ c(.x$data, .x$truth)))))

if (regenerate_simulations || !ml_files_exist) {
  ml_decay_sims <- map(ml_scenarios, function(scen) {
    out <- simulate_multilevel_decay_gcm(
      n_subjects          = n_subjects_sim,
      stimulus_info       = stimulus_info,
      n_blocks            = n_blocks,
      scenario            = scen,
      base_seed           = match(scen, ml_scenarios) * 1000 + 42,
      pop_logit_w_mean    = pop_params_sim$pop_logit_w_mean,
      pop_logit_w_sd      = pop_params_sim$pop_logit_w_sd,
      pop_log_c_mean      = pop_params_sim$pop_log_c_mean,
      pop_log_c_sd        = pop_params_sim$pop_log_c_sd,
      pop_log_lambda_mean = pop_params_sim$pop_log_lambda_mean,
      pop_log_lambda_sd   = pop_params_sim$pop_log_lambda_sd,
      pop_logit_bias_mean = pop_params_sim$pop_logit_bias_mean,
      pop_logit_bias_sd   = pop_params_sim$pop_logit_bias_sd
    )
    write_csv(out$data,        ml_decay_files[[scen]]$data)
    write_csv(out$true_params, ml_decay_files[[scen]]$truth)
    out
  }) |> setNames(ml_scenarios)
} else {
  ml_decay_sims <- map(ml_scenarios, function(scen) list(
    data        = read_csv(ml_decay_files[[scen]]$data,  show_col_types = FALSE),
    true_params = read_csv(ml_decay_files[[scen]]$truth, show_col_types = FALSE)
  )) |> setNames(ml_scenarios)
}

ml_decay_all <- map_dfr(ml_scenarios, function(scen)
  ml_decay_sims[[scen]]$data |> mutate(scenario = scen)
) |>
  mutate(scenario = factor(scenario, levels = ml_scenarios,
                           labels = c("Static", "Smooth drift",
                                      "Performance-contingent")))

ggplot(ml_decay_all,
       aes(x = trial_within_subject, y = cumulative_accuracy, group = agent_id)) +
  geom_line(alpha = 0.25, color = "grey30") +
  stat_summary(fun = mean, geom = "line", aes(group = 1),
               color = "#0072B2", linewidth = 1.4) +
  facet_wrap(~scenario, ncol = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Multilevel Decay GCM: simulated learning curves across scenarios",
    subtitle = paste0(n_subjects_sim, " agents per scenario; thin grey = individuals, ",
                      "thick blue = population mean"),
    x = "Trial", y = "Cumulative accuracy"
  )
```

The static panel converges quickly to ceiling; smooth drift plateaus below ceiling because the boundary keeps moving; performance-contingent shifts produce a noisy plateau as different subjects flip at different absolute trials and average over each other.

### Stan Implementation

The `stan/ch12_gcm_ml_decay.stan` model implements the hierarchy described above, with three architectural choices that matter for performance:

* **The trial-by-trial likelihood lives in the `model` block, not in `transformed parameters`.** Memory is held in plain `int` arrays during the loop, so the autodiff tape never sees the per-trial scan. `prob_cat1` and (optionally gated) `log_lik` are recomputed once per saved draw in `generated quantities`.
* **All hierarchies are non-centered** (logit-normal NCP for `w_1`; standard z-offset for `log_c`, `log_lambda`, and `logit_bias`).
* **Pairwise feature distances are precomputed in `transformed data`**, indexed by `stim_id` instead of read from per-trial features.

```{r ch12_ml_decay_stan_compile}
stan_file_gcm_ml_decay <- here("stan", "ch12_gcm_ml_decay.stan")
mod_gcm_ml_decay       <- cmdstan_model(stan_file_gcm_ml_decay)
```

### Building the Stan Data List

```{r ch12_ml_decay_stan_data_builder}
build_ml_stan_data <- function(ml_data, stimulus_info, run_diagnostics = 0L) {
  ml_data <- ml_data |>
    arrange(agent_id, trial_within_subject) |>
    mutate(row = row_number())

  unique_stim <- stimulus_info |>
    arrange(stimulus) |>
    mutate(unique_idx = row_number())
  stim_lookup <- setNames(unique_stim$unique_idx, unique_stim$stimulus)

  subj_idx <- ml_data |>
    group_by(agent_id) |>
    summarise(start = min(row), end = max(row), .groups = "drop") |>
    arrange(agent_id)

  list(
    N_total                = nrow(ml_data),
    N_subjects             = nrow(subj_idx),
    N_features             = 2L,
    N_unique_stim          = nrow(unique_stim),
    max_trials_per_subject = as.integer(max(subj_idx$end - subj_idx$start + 1L)),

    subj_start             = as.integer(subj_idx$start),
    subj_end               = as.integer(subj_idx$end),

    stim_id                = as.integer(stim_lookup[as.character(ml_data$stimulus_id)]),
    y                      = as.integer(ml_data$sim_response),
    unique_obs             = as.matrix(unique_stim[, c("height", "position")]),
    cat_feedback           = as.integer(ml_data$observed_feedback),

    prior_pop_logit_w_mean      = 0,
    prior_pop_logit_w_sigma     = 1.0,
    prior_pop_logit_w_sd_lambda = 2.0,
    prior_pop_log_c_mu          = LOG_C_PRIOR_MEAN,
    prior_pop_log_c_sigma       = LOG_C_PRIOR_SD,
    prior_pop_log_c_lambda      = 2.0,
    prior_pop_log_lambda_mu     = LOG_LAMBDA_PRIOR_MEAN,
    prior_pop_log_lambda_sigma  = LOG_LAMBDA_PRIOR_SD,
    prior_pop_log_lambda_lambda = 2.0,
    prior_pop_logit_bias_mu     = 0,
    prior_pop_logit_bias_sigma  = 1.0,
    prior_pop_logit_bias_lambda = 2.0,

    run_diagnostics             = as.integer(run_diagnostics)
  )
}
```

The `Exponential(2)` priors on the population SDs (`*_lambda = 2`) have median 0.35 and 95th percentile ≈ 1.5 — wide enough to allow real between-subject variation, tight enough to keep NUTS off the worst part of the funnel when only 10 subjects are observed. We keep `run_diagnostics = 0` for the main fits because we will not run LFO at the multilevel level here; flip it to `1` only when computing per-trial `log_lik` is needed.

### Fitting the Multilevel Decay GCM Across Scenarios

We compile once and fit each scenario in turn. With 10 subjects × 64 trials × 4 chains × 2000 iterations, expect each fit to take a few minutes on a modern laptop.

```{r ch12_ml_decay_fit_per_scenario}
ml_decay_fit_files <- map(ml_scenarios, function(scen)
  here("simmodels", paste0("ch12_gcm_ml_decay_fit_", scen, ".rds"))
) |> setNames(ml_scenarios)

if (regenerate_simulations || !all(file.exists(unlist(ml_decay_fit_files)))) {
  ml_decay_fits <- map(ml_scenarios, function(scen) {
    cat("Fitting multilevel decay GCM:", scen, "\n")
    dat <- build_ml_stan_data(ml_decay_sims[[scen]]$data, stimulus_info,
                              run_diagnostics = 0L)
    fit <- mod_gcm_ml_decay$sample(
      data            = dat,
      seed            = 700 + match(scen, ml_scenarios),
      chains          = 4,
      parallel_chains = min(4, availableCores()),
      iter_warmup     = 1000,
      iter_sampling   = 1000,
      adapt_delta     = 0.95,
      max_treedepth   = 12,
      refresh         = 200,
      init            = 0.5
    )
    fit$save_object(ml_decay_fit_files[[scen]])
    fit
  }) |> setNames(ml_scenarios)
  cat("Multilevel decay GCM fits computed and saved.\n")
} else {
  ml_decay_fits <- map(ml_scenarios, function(scen)
    readRDS(ml_decay_fit_files[[scen]])
  ) |> setNames(ml_scenarios)
  cat("Loaded existing multilevel decay GCM fits.\n")
}
```

### MCMC Diagnostic Battery

```{r ch12_ml_decay_diagnostics}
ml_decay_diag <- map_dfr(ml_scenarios, function(scen) {
  diagnostic_summary_table(
    ml_decay_fits[[scen]],
    params = c("pop_logit_w_mean",    "pop_log_c_mean",
               "pop_log_lambda_mean", "pop_logit_bias_mean",
               "pop_logit_w_sd",      "pop_log_c_sd",
               "pop_log_lambda_sd",   "pop_logit_bias_sd")
  ) |> mutate(scenario = scen)
})

print(ml_decay_diag |>
        dplyr::select(scenario, metric, value, threshold, pass))
```

If the model is healthy you will see no divergences, R̂ < 1.01 across all population parameters, and ESS comfortably above 400 for the population means. Subject-level scalars typically have somewhat lower ESS, especially for `log_lambda` under the static scenario — the data carry little information about forgetting when the boundary never moves, so the posterior is partly pinned by the prior and the partial pooling toward the population mean.

### Population-Level Recovery

```{r ch12_ml_decay_population_recovery}
pop_truth <- with(pop_params_sim, tibble(
  variable = c("pop_logit_w_mean", "pop_log_c_mean",
               "pop_log_lambda_mean", "pop_logit_bias_mean"),
  truth    = c(pop_logit_w_mean, pop_log_c_mean,
               pop_log_lambda_mean, pop_logit_bias_mean)
))

pop_summary <- map_dfr(ml_scenarios, function(scen) {
  ml_decay_fits[[scen]]$summary(variables = pop_truth$variable) |>
    mutate(scenario = scen)
}) |>
  left_join(pop_truth, by = "variable") |>
  mutate(scenario = factor(scenario, levels = ml_scenarios,
                           labels = c("Static", "Drift", "Contingent shift")),
         variable = factor(variable, levels = pop_truth$variable))

ggplot(pop_summary, aes(y = scenario, x = mean, color = scenario)) +
  geom_pointrange(aes(xmin = q5, xmax = q95), size = 0.6) +
  geom_vline(aes(xintercept = truth), linetype = "dashed", color = "red") +
  facet_wrap(~variable, scales = "free_x", ncol = 2) +
  scale_color_viridis_d(option = "plasma", end = 0.85) +
  labs(
    title    = "Population-level parameter recovery",
    subtitle = "Posterior mean ± 90% CI; red dashed = simulation truth",
    x = NULL, y = NULL
  ) +
  theme(legend.position = "none")
```

The expected pattern: `pop_log_c_mean` and `pop_logit_bias_mean` recover cleanly across all three scenarios; `pop_log_lambda_mean` is well-recovered under drift and contingent-shift but pulled toward the prior under static (the data simply lack a forgetting signal); `pop_logit_w_mean` is recovered everywhere but with wider posteriors when attention has little leverage on accuracy.

### Subject-Level Recovery and Shrinkage

```{r ch12_ml_decay_subject_recovery}
subj_recovery <- map_dfr(ml_scenarios, function(scen) {
  fit   <- ml_decay_fits[[scen]]
  truth <- ml_decay_sims[[scen]]$true_params

  s <- fit$summary(variables = c("subj_log_c", "subj_log_lambda")) |>
    mutate(
      param    = if_else(stringr::str_detect(variable, "log_c"),
                         "log_c", "log_lambda"),
      agent_id = as.integer(stringr::str_extract(variable, "\\d+"))
    ) |>
    dplyr::select(agent_id, param, est_mean = mean, est_q5 = q5, est_q95 = q95)

  truth_long <- truth |>
    dplyr::select(agent_id, log_c_true, log_lambda_true) |>
    pivot_longer(cols = c(log_c_true, log_lambda_true),
                 names_to = "param", values_to = "truth") |>
    mutate(param = if_else(param == "log_c_true", "log_c", "log_lambda"))

  s |>
    left_join(truth_long, by = c("agent_id", "param")) |>
    mutate(scenario = scen)
}) |>
  mutate(scenario = factor(scenario, levels = ml_scenarios,
                           labels = c("Static", "Drift", "Contingent shift")))

ggplot(subj_recovery, aes(x = truth, y = est_mean, color = scenario)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = est_q5, ymax = est_q95), alpha = 0.7) +
  facet_grid(param ~ scenario, scales = "free") +
  scale_color_viridis_d(option = "plasma", end = 0.85) +
  labs(
    title    = "Subject-level parameter recovery (with partial pooling)",
    subtitle = "Posterior mean ± 90% CI vs simulation truth",
    x = "True value", y = "Posterior estimate"
  ) +
  theme(legend.position = "none")
```

The `log_c` row should be tight along the identity across scenarios. The `log_lambda` row is the diagnostic: scattered and pulled toward the population mean under static (visible shrinkage when an individual's data is uninformative), tighter along the diagonal under drift, and tightest under contingent-shift — consistent with the single-subject recovery results from earlier in the chapter.

### What changes when you scale up

Three things to keep in mind when adapting this model to real data:

1. **Sample size enters twice.** With more subjects, the population SDs become better identified, which shrinks subject-level estimates more aggressively toward the population mean — desirable when individual-level data is noisy, but worth checking visually.
2. **Sequential-trial costs scale linearly.** Doubling trials per subject doubles per-leapfrog cost; doubling subjects doubles it again. Going from this 10×64 toy to a 100×200 dataset is a ~30× increase in wall time, not 30 different fits to manage.
3. **Static-scenario behaviour is the canary.** If `pop_log_lambda_mean` and `pop_log_lambda_sd` start to diverge or drag at `max_treedepth` on real data, your task is probably structurally closer to "static" than you think — there is not enough non-stationarity for $\lambda$ to be identified at the population level. Reduce the model or accept that $\lambda$ is pinned by the prior and report it as such.



### GCM in Context: Strengths and Limitations

#### Strengths

* **Empirical support**: The GCM has successfully fit human categorization data across numerous experiments.
* **Psychological plausibility**: The model's parameters correspond to psychologically meaningful constructs (attention, sensitivity/generalization, bias, and now forgetting).
* **Flexibility**: It can account for a wide range of category structures and learning phenomena, including non-stationary environments.
* **Individual differences**: The multilevel framework allows principled modeling of variations across people.

#### Limitations

* **Memory requirements**: Storing all exemplars becomes computationally demanding for large numbers of stimuli. How memory is bounded over long timescales is not fully specified.
* **Conflation of generalisation and decisional precision**: The standard GCM has no separate softmax temperature; sensitivity $c$ doubles as a generalisation parameter and as a decisional precision parameter.
* **Complex categories**: The standard GCM struggles with categories defined by abstract rules, relations, or logical combinations of features.
* **Sequential structure and LOO**: Because the GCM is path-dependent, standard LOO-CV is technically invalid — but this chapter implements LFO-CV (Bürkner et al., 2020) as the canonical sequential-model cross-validation.

In the next two chapters we explore alternative approaches — prototype models (Ch. 12) and rule-based models (Ch. 13) — that address some of these limitations while introducing their own strengths and challenges.

### Looking Ahead to Chapter 12: The Six-Step Pipeline in the Prototype Model

Chapter 12 applies the same six-step pipeline to a prototype-based model. Before moving on, it is worth tracing explicitly how each stage maps from the GCM to the Kalman filter prototype model — because the chapters are designed as a matched pair, and the contrasts at each stage are the core theoretical content.

| Pipeline Stage | GCM (this chapter) | Kalman Filter Prototype (Ch. 12) |
| :--- | :--- | :--- |
| **1. Input representation** | Stimuli as points in continuous feature space (height, position). | Same feature space; stimuli used as Kalman observations. |
| **2. Attention** | Explicit attention weights $w_f$ summing to 1; estimated from data. | No explicit attention weights; covariance structure implicitly weights dimensions. |
| **3. Intermediate representation** | **All** past exemplars stored verbatim; memory grows trial by trial. | One prototype $\vec{\mu}_k$ per category, updated by each observation; memory is constant-size. |
| **4. Evidential mechanism** | Similarity to each stored exemplar, exponentially decaying with distance and (in the extended model) trial age. | Mahalanobis-like distance from the current stimulus to each prototype, scaled by the running covariance $\Sigma_k$. |
| **5. Decision rule** | Luce choice axiom: $p(\text{cat}_1) = \frac{\beta \cdot \bar{\eta}_1}{\beta \cdot \bar{\eta}_1 + (1-\beta) \cdot \bar{\eta}_0}$. | Same Luce choice axiom; the per-trial probability is computed from the Kalman-filter likelihood ratio. |
| **6. Learning** | Accumulate: each trial's exemplar enters permanent (but decaying) memory. No structural update. | Update: the Kalman gain $K$ shifts $\vec{\mu}_k$ toward the new observation and reduces $\Sigma_k$; the prediction step (process noise $Q$) can re-inflate $\Sigma_k$ between trials. |

The key tension is at **Stage 3 and Stage 6**. The GCM treats every trial as permanently informative (modulo forgetting); the prototype model treats each trial as an observation that refines a running summary. As a result:

* The GCM is maximally sensitive to specific exemplars and their spatial arrangement. The prototype model is sensitive only to the category centroid and spread.
* The GCM's non-stationarity mechanism (decay $\lambda$) operates *retrospectively* — old exemplars become less influential as time passes. The Kalman filter's mechanism (process noise $q$) operates *prospectively* — future observations are expected to diverge from the current prototype.
* Both mechanisms are phenomenologically similar under smooth drift. They diverge most sharply under abrupt shifts: the GCM gradually forgets old contradicting exemplars, while the Kalman filter re-inflates uncertainty after surprising observations and quickly reacquires the new prototype.

::: {.callout-note title="What to watch for when reading Chapter 12"}
1. **The $q$/$r$ identifiability problem.** The Kalman filter's two parameters ($r$ = observation noise, $q$ = process noise) are more tightly confounded under static data than the GCM's $c$/$\lambda$ pair — they both scale the *same* decision-time likelihood rather than operating on orthogonal axes (see §"The $c$/$\lambda$ Joint Posterior" above). Ch. 12 introduces an explicit reparameterization (the SNR parameter $\rho = q/r$) to diagnose and partially resolve this.
2. **The scenario order is deliberately reversed.** Ch. 12 arranges scenarios as static → contingent shift → drift (increasing assumption match for the Kalman filter), while this chapter used static → drift → contingent shift (increasing $\lambda$ preference). The three scenarios are identical; only the ordering logic differs.
3. **No attention weights.** Removing $w$ and $\beta$ from the parameter set makes the prototype model simpler to diagnose — SBC rank histograms have two curves instead of four, and pair plots have one panel instead of six. That simplicity is *why* Ch. 12 is the right place to introduce LFO-CV in detail: the algorithm is most legible when the model is lean.
:::

---
