# The Generalized Context Model

```{r ch12_setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "80%",
  dpi        = 300
)

# Set to TRUE to rerun all Stan fits and heavy simulations; FALSE loads saved results.
regenerate_simulations <- FALSE
# Set to TRUE to rerun SBC, extensive recovery, exact LFO validation.
run_intensive_checks   <- FALSE

# Directory scaffolding
for (d in c("stan", "simdata", "simmodels")) {
  if (!dir.exists(d)) dir.create(d)
}

pacman::p_load(
  tidyverse,    # data manipulation and visualization
  here,         # robust project-relative paths
  future,       # parallel processing setup
  purrr,        # functional programming tools
  furrr,        # parallel functional programming
  patchwork,    # combining plots
  cmdstanr,     # interface to Stan
  posterior,    # tidy posterior arrays
  tidybayes,    # tidy extraction of draws
  bayesplot,    # Bayesian visualization
  loo,          # LOO-CV model comparison
  SBC,          # Simulation-Based Calibration (ECDF-diff plots)
  priorsense,   # prior sensitivity via power-scaling
  MCMCpack      # rdirichlet
)

theme_set(theme_classic())

# Prior hyperparameters for sensitivity c
LOG_C_PRIOR_MEAN <- log(2)   # prior median c = 2
LOG_C_PRIOR_SD   <- 1.0      # covers c ≈ 0.27–14.8 at ±2 SD

# ── Shared diagnostic helper used throughout the chapter ─────────────────────
diagnostic_summary_table <- function(fit, params = NULL) {
  if (is.null(fit)) return(NULL)
  diag <- fit$diagnostic_summary(quiet = TRUE)

  s <- fit$summary(
    variables = params,
    "mean", "sd", "rhat", "ess_bulk", "ess_tail", "mcse_mean"
  )

  tibble(
    metric = c(
      "Divergences (zero tolerance)",
      "Max rank-normalised R-hat",
      "Min bulk ESS",
      "Min tail ESS",
      "Min E-BFMI",
      "Max MCSE / posterior SD"
    ),
    value = c(
      sum(diag$num_divergent),
      max(s$rhat,     na.rm = TRUE),
      min(s$ess_bulk, na.rm = TRUE),
      min(s$ess_tail, na.rm = TRUE),
      min(diag$ebfmi, na.rm = TRUE),
      max(s$mcse_mean / s$sd, na.rm = TRUE)
    ),
    threshold = c("== 0", "< 1.01", "> 400", "> 400", "> 0.2", "< 0.05"),
    pass = c(
      sum(diag$num_divergent)               == 0,
      max(s$rhat, na.rm = TRUE)             <  1.01,
      min(s$ess_bulk, na.rm = TRUE)         >  400,
      min(s$ess_tail, na.rm = TRUE)         >  400,
      min(diag$ebfmi, na.rm = TRUE)         >  0.2,
      max(s$mcse_mean / s$sd, na.rm = TRUE) <  0.05
    )
  )
}
```

> **📍 Where we are in the Bayesian modeling workflow:**
> Chapters 1–10 built the complete inference toolkit: parameter estimation
> (Ch. 4), the six-phase validation pipeline (Ch. 5), hierarchical models
> (Ch. 6), model comparison via LOO (Ch. 7), mixture models with discrete
> marginalization (Ch. 8), and hierarchical Bayesian cognition (Ch. 9–10).
> This chapter applies the full toolkit to *categorization* — asking not only
> what parameters describe a participant's behavior, but which *cognitive
> mechanism* best accounts for how people learn to assign objects to kinds.
> We begin with the exemplar approach: the Generalized Context Model (GCM).
> The following chapters will introduce prototype and rule-based models,
> and then compare all three approaches head-to-head using the full validation
> battery.
>
> Two important caveats that previous chapters did not face.
> First, the categorization models learn, that is, are **path-dependent** —
> the likelihood at trial $t$ depends on the full sequence of stimuli on
> trials $1, \ldots, t-1$ through the exemplar memory store. Standard LOO-CV
> is therefore technically invalid here. Rather than wave at this and hope it
> doesn't matter, we introduce leave-future-out cross-validation (LFO-CV;
> Bürkner, Gabry, & Vehtari, 2020).
> Second, the rule-based models we will discuss cannot really be fit using
> MCMC (we'll see why), so we need to introduce a new slower method: Approximate
> Bayesian Computation (ABC).

# Introduction to Categorization Models

## The Fundamental Problem of Categorization

Categorization is one of the most fundamental cognitive abilities that humans possess. From early childhood, we learn to organize the world around us into meaningful categories: distinguishing food from non-food, safe from dangerous, or one letter from another. This process of assigning objects to categories is complex, involving perception, memory, attention, and decision-making.

How do humans learn categories and make categorization decisions? This seemingly simple question has generated decades of research and theoretical debate in cognitive science.

## Why Model Categorization?

As we have seen in previous chapters, computational models offer powerful ways to formalize theories about categorization. By implementing these theories as computer algorithms, we can:

* Explore the consequences of different theoretical assumptions

* Make precise predictions about human behavior in categorization tasks

* Test competing theories against empirical data

In this chapter, we will explore three major approaches to modeling categorization:

* **Exemplar models** — which propose that categories are represented by storing individual examples

* **Prototype models** — which suggest categories are represented by an abstract "average" or central tendency

* **Rule-based models** — which posit that categories are represented by explicit rules or decision boundaries

Each approach captures different aspects of human categorization and has generated substantial empirical research. We'll implement computational versions of each model type, allowing us to compare their behavior and predictions.

## Historical Development of Categorization Models

### Early Views: The Classical Approach

Early theories of categorization followed what is now called the classical view, where categories were defined by necessary and sufficient features. In this view, category membership is an all-or-nothing affair: an object either satisfies the logical criteria or it doesn't. While intuitive, this approach struggled to explain many aspects of human categorization, such as:

* Graded category membership (some members seem "more typical" than others)

* Unclear boundaries between categories

* Context-dependent categorization

* Family resemblance structures (where no single feature is necessary)

### The Prototype Revolution

In the 1970s, Eleanor Rosch's pioneering work on prototypes challenged the classical view. She demonstrated that categories appear to be organized around central tendencies or "prototypes," with membership determined by similarity to these prototypes. Objects closer to the prototype are categorized more quickly and consistently.

### The Exemplar Alternative

In the late 1970s and 1980s, researchers like Douglas Medin and Robert Nosofsky proposed that rather than abstracting prototypes, people might store individual exemplars of categories and make judgments based on similarity to these stored examples.

The Generalized Context Model (GCM), developed by Nosofsky, became the standard exemplar model. It proposes that categorization decisions are based on the **summed** similarity of a new stimulus to all stored exemplars of each category, weighted by attention to different stimulus dimensions.

### Rule-Based Models

While similarity-based approaches gained prominence, other researchers argued that people sometimes use explicit rules for categorization. Models like Bayesian particle filters, multinomial trees, COVIS and RULEX formalize these ideas, suggesting that rule learning and application form a core part of human categorization, particularly for well-defined categories.

### Hybrid Approaches

More recent work has focused on hybrid models that incorporate elements of multiple approaches. Models like SUSTAIN and ATRIUM propose that humans can flexibly switch between strategies or that different systems operate in parallel.

## A Systematization Framework for Categorization Models

To pedagogically navigate and compare the different approaches, we can deconstruct the cognitive process of categorization into a sequential six-step pipeline:

1. **Input Representation:** How are stimuli represented? (e.g., continuous geometry vs. discrete binary features).
2. **Attentional Processes:** How are features prioritized? Not all features are equally relevant given a task, and attention is often modeled as weights summing to 1.
3. **Intermediate Representations:** How is the category internally stored? (e.g., as raw exemplars, abstract prototypes, or explicit rules).
4. **Evidential Mechanisms:** How is the incoming stimulus compared to the representation to gather evidence? (e.g., calculating average similarity with an exponential decay over distance, or evidence accumulation).
5. **Decision Rules:** How is evidence translated into a choice? (e.g., maximum similarity, or a probabilistic Luce choice axiom).
6. **Learning:** How do representations, attentional weights, or rule structures update over time through exposure, memory accumulation, and feedback?

### The Three Model Classes Through the Pipeline

We can map our three theoretical model classes directly onto this processing pipeline:

| Pipeline Stage | Exemplar Models (e.g., GCM) | Prototype Models | Rule-Based Models |
| :--- | :--- | :--- | :--- |
| **1. Input Representation** | Continuous spatial geometry or discrete features. | Continuous spatial geometry or discrete features. | Often discrete, logical features. |
| **2. Attentional Processes** | Continuous attention weights to dimensions. | Continuous attention weights to dimensions. | Selective attention weights allocated to specific rules or dimensions. |
| **3. Intermediate Representations**| **Raw/None:** All encountered exemplars are stored. | **Abstract:** Integrated into a summary centroid/average. | **Abstract:** Continuous space partitioned into discrete rule boundaries. |
| **4. Evidential Mechanisms** | Exponentially decaying similarity to *all* stored exemplars. | Exponentially decaying similarity to the *central prototype*. | Probability that the exemplar is generated by the current rule. |
| **5. Decision Rules** | Luce choice axiom (probabilistic assignment). | Luce choice axiom (probabilistic assignment). | Maximum similarity or probabilistic rule combination. |
| **6. Learning** | Accumulation of distinct exemplars in memory (path-dependent). | Updating the centroid/summary representation (e.g., via a Kalman filter). | Updating weights for sampled rules based on feedback. |

## Our Implementation Approach

We'll implement three models representing each major theoretical approach:

* **Generalized Context Model (GCM)** — A canonical exemplar model where categorization is based on summed similarity to all stored examples

* **Kalman Filter Prototype Model** — A dynamic prototype model (Ch. 12)

* **Bayesian Particle Filter for Rules** — A rule-based model (Ch. 13)

Each implementation follows a similar structure: core model function as a forward simulator, Stan implementation for parameter estimation, the mandatory MCMC diagnostic battery, prior/posterior predictive checks, parameter recovery, prior sensitivity, and SBC.

::: {.callout-note title="The Conceptual Need for Shared Implementations"}
A major challenge in evaluating cognitive theories is the "hidden implementation" problem. If different research groups write their own custom code for a GCM or a prototype model, subtle differences in how the math is translated into an algorithm can lead to conflicting theoretical conclusions. To genuinely compare models, the field conceptually requires shared, open-source implementations tested against standardized empirical benchmarks. 

A pioneering example of this movement is the work by Wills et al. (2017), who developed the `catlearn` R package to provide formal, collaboratively verified implementations of classic categorization models and canonical datasets. 

These course notes are designed as a partial, pedagogical contribution to this open-science ethos. We do not adopt the specific software architecture or strict standardization of packages like `catlearn`, but we share the same fundamental goal: transparent, reproducible science. We show how to create, document and validate transparent bespoke scripts and Stan-based Bayesian inference. This means that you will not just adopt some secretly shared pre-made script, or struggle implementing an equation and then never share the script as you are secretely afraid it contains errors. Rather, you'll be able to implement and develop different algorithms, validate and document them, and share them with the world, so that others will be able to build on your work and we can do better science as a collective enterprise.

*Reference:* Wills, A. J., O'Connell, G., Edmunds, C. E., & Inkster, A. B. (2017). Progress in modeling through distributed collaboration: Concepts, tools and category-learning examples. *Psychology of learning and motivation* (Vol. 66, pp. 79-115). Academic Press.
:::


## Categorization (Cognitive Science) vs. Classification (Machine Learning)

Because the mathematical tools are often identical, it is tempting to view cognitive models of categorization simply as early, inefficient machine learning classifiers. Both fields rely on the exact same underlying mechanics to map inputs to outputs: calculating distances in multi-dimensional feature space, applying functions (like Luce or softmax) to translate evidence into probabilities, and using gradient-based weight updates or Bayesian filters to learn from feedback. 

However, despite this shared mathematical DNA, they serve fundamentally different goals:

* **Classification (Machine Learning):** The primary goal is *engineering*. It focuses on drawing the optimal decision boundary to maximize predictive accuracy and minimize loss on a given dataset. The internal mechanism (e.g., a deep neural network's hidden layers) doesn't need to resemble human thought—it just needs to work efficiently and accurately.
* **Categorization (Cognitive Modeling):** The primary goal is *reverse-engineering*. Categorization is about understanding the knowledge structures that enable a biological agent to make predictive inferences. We ask not only what parameters describe behavior, but which *cognitive mechanism* best accounts for how people actually learn, including the specific mistakes they make.

We can summarize this distinction along three key dimensions:

| Dimension | Classification (Machine Learning) | Categorization (Cognitive Modeling) |
| :--- | :--- | :--- |
| **Primary Goal** | Maximize out-of-sample accuracy. | Explain human behavior, including systematic human errors. |
| **Data Processing** | Often batch-processed over massive, static datasets. | Inherently **path-dependent**; learns sequentially trial-by-trial. |
| **Constraints** | Hardware limits and computational efficiency. | Psychological plausibility (bounded memory, selective attention, cognitive load). |

In short: A machine learning classifier asks, *"What is the best mathematical way to separate Category A from Category B?"* A cognitive model asks, *"How do humans separate A from B, and what are the internal mental representations that allow them to do so?"*

---

# The Generalized Context Model (GCM)

## Mathematical Foundations and Cognitive Principles

The Generalized Context Model (GCM), developed by Robert Nosofsky in the 1980s, represents one of the most influential exemplar-based approaches to categorization. Unlike models that abstract category information into prototypes or rules, the GCM proposes that people store individual exemplars in memory and make categorization decisions based on similarity to these stored examples.

### The GCM Mapped to the Categorization Pipeline

The GCM translates the 6-step cognitive pipeline into a formal mathematical structure as follows:

* **1. Input Representation (Geometric Space):** Stimuli are represented as continuous points in a multi-dimensional psychological space.
* **2. Attentional Processes (Selective Attention):** Per-dimension weights $w$ (summing to 1) rescale the psychological space before distance is computed, prioritizing task-relevant features.
* **3. Intermediate Representations (Memory for Instances):** The model explicitly skips abstraction. Every encountered example is stored directly in memory with its category label.
* **4. Evidential Mechanisms (Similarity Computation):** Distance is calculated using a city-block metric (Manhattan distance). Similarity between a new stimulus and each stored exemplar is an exponentially decreasing function of their distance, governed by a sensitivity parameter $c$. Evidence is then aggregated as average similarity for each category.
* **5. Decision Rules (Choice Probability):** The decision relies on the Luce choice axiom. The probability of assigning a stimulus to a category is proportional to its summed similarity to exemplars of that category, modified by a baseline category bias $\beta$.
* **6. Learning (Memory Accumulation):** The model learns sequentially by adding each newly categorized and feedback-evaluated stimulus directly into its exemplar memory store, altering the evidential landscape for future trials.

### Parameters: Cognitive Commitment, Algebraic Role, Stan Parameterization

**Attention weights $w$.** Cognitively, $w$ is a simplex of dimension-specific attention weights. Algebraically, $w$ enters the distance formula as the per-dimension coefficient in the attention-weighted city-block metric. At the Stan level $w$ is declared as `simplex[nfeatures]`, which enforces $w_m \ge 0$ and $\sum_m w_m = 1$ directly.

**Sensitivity $c$.** Cognitively, $c$ represents perceptual discriminability. Algebraically, $c$ is the exponential decay rate in $\eta_{ij} = e^{-c\, d_{ij}}$. At the Stan level, we sample an unconstrained `real log_c` and set `c = exp(log_c)` in `transformed parameters`.

**Response bias $\beta$.** Cognitively, $\beta$ is a baseline preference for category A independent of stimulus similarity; $\beta = 0.5$ is unbiased. In this chapter $\beta$ also serves as the cold-start fallback. At the Stan level $\beta$ is declared as `real<lower=0, upper=1>` with a Beta prior.

## Mathematical Formulation

**1. Distance Calculation**

$$d_{ij} = \left[ \sum_{m=1}^{M} w_m |x_{im} - x_{jm}|^r \right]^{\frac{1}{r}}$$

We use the city-block metric ($r = 1$), appropriate for separable stimulus dimensions:

```{r ch12_distance_formula}
# City-block distance (r = 1), attention-weighted
distance <- function(vect1, vect2, w) {
  sum(w * abs(vect1 - vect2))
}
```

**2. Similarity Computation**

$$\eta_{ij} = e^{-c \cdot d_{ij}}$$

```{r ch12_similarity_formula}
# Exponential similarity kernel
similarity <- function(dist_val, c) {
  exp(-c * dist_val)
}
```

Let's visualize how similarity changes with distance for different values of $c$:

```{r ch12_visualizing_sensitivity}
dd <- expand_grid(
  distance  = seq(0, 6, by = 0.1),
  c         = c(0.3, 0.7, 1.5, 3.0, 6.0, 10.0)
) |>
  mutate(
    sim     = similarity(distance, c),
    c_label = factor(c)
  )

ggplot(dd, aes(x = distance, y = sim, color = c_label, group = c_label)) +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(option = "plasma", name = "Sensitivity (c)") +
  labs(
    title    = "Effect of Sensitivity Parameter (c) on Similarity",
    subtitle = "c is on the natural (original) scale; prior places most mass on c ∈ (0.3, 15)",
    x        = "Psychological Distance",
    y        = "Similarity"
  ) +
  theme(legend.position = "bottom")
```

**3. Category Response Probability**

$$P(A|i) = \frac{\beta \sum_{j \in A} \eta_{ij}}{\beta \sum_{j \in A} \eta_{ij} + (1-\beta) \sum_{k \in B} \eta_{ik}}$$

The cold-start fallback applies whenever one of the summed similarities is undefined: if either $\sum_{j \in A}$ or $\sum_{k \in B}$ has no terms yet, the response probability is $\beta$.

### The GCM Agent Implementation (R Simulation)

```{r ch12_generative_gcm}
# Generative GCM model function.
gcm_simulate <- function(w,            # attention weights (sums to 1)
                         c,            # sensitivity (natural scale, strictly positive)
                         bias,         # response bias towards category 1
                         obs,          # matrix of stimulus features (trials x features)
                         cat_feedback, # true category labels (0 or 1)
                         quiet = TRUE) {
  ntrials   <- nrow(obs)
  nfeatures <- ncol(obs)

  prob_cat1    <- numeric(ntrials)
  sim_response <- numeric(ntrials)

  memory_obs <- matrix(NA_real_, nrow = 0, ncol = nfeatures)
  memory_cat <- numeric(0)

  for (i in seq_len(ntrials)) {
    if (!quiet && i %% 20 == 0) cat("Simulating trial:", i, "\n")

    current_stim <- as.numeric(obs[i, ])
    n_mem        <- nrow(memory_obs)
    has_cat0     <- any(memory_cat == 0)
    has_cat1     <- any(memory_cat == 1)

    if (n_mem == 0 || !has_cat0 || !has_cat1) {
      prob_cat1[i] <- bias
    } else {
      sims <- vapply(seq_len(n_mem), function(e) {
        similarity(distance(current_stim, memory_obs[e, ], w), c)
      }, numeric(1))

      s1  <- sum(sims[memory_cat == 1])
      s0  <- sum(sims[memory_cat == 0])
      num <- bias * s1
      den <- num + (1 - bias) * s0

      prob_cat1[i] <- if (den > 1e-9) num / den else bias
      prob_cat1[i] <- pmax(1e-9, pmin(1 - 1e-9, prob_cat1[i]))
    }

    sim_response[i] <- rbinom(1, 1, prob_cat1[i])

    memory_obs <- rbind(memory_obs, current_stim)
    memory_cat <- c(memory_cat, cat_feedback[i])
  }

  tibble(prob_cat1 = prob_cat1, sim_response = sim_response)
}
```

---

## Experimental Setup: Kruschke (1993)

```{r ch12_experimental_setup}
stimulus_info <- tibble(
  stimulus      = c(5, 3, 7, 1, 8, 2, 6, 4),
  height        = c(1, 1, 2, 2, 3, 3, 4, 4),
  position      = c(2, 3, 1, 4, 1, 4, 2, 3),
  category_true = c(0, 0, 1, 0, 1, 0, 1, 1)
)

ggplot(stimulus_info, aes(x = position, y = height,
                          color = factor(category_true), label = stimulus)) +
  geom_point(size = 5) +
  geom_text(nudge_y = 0.18) +
  scale_color_manual(values = c("0" = "#0072B2", "1" = "#D55E00"), name = "Category") +
  labs(title = "Kruschke (1993) Stimulus Structure", x = "Position", y = "Height") +
  coord_fixed()

n_blocks         <- 8
n_stim_per_block <- nrow(stimulus_info)
total_trials     <- n_blocks * n_stim_per_block
```

---

## Per-Subject Schedule Generation

```{r ch12_subject_schedule}
make_subject_schedule <- function(stimulus_info, n_blocks, seed) {
  set.seed(seed)
  n_stim <- nrow(stimulus_info)

  sequence <- unlist(lapply(seq_len(n_blocks), function(b) {
    sample(stimulus_info$stimulus, n_stim, replace = FALSE)
  }))

  tibble(
    trial_within_subject = seq_along(sequence),
    block                = rep(seq_len(n_blocks), each = n_stim),
    stimulus_id          = sequence
  ) |>
    left_join(stimulus_info, by = c("stimulus_id" = "stimulus")) |>
    rename(category_feedback = category_true) |>
    dplyr::select(trial_within_subject, block, stimulus_id,
                  height, position, category_feedback)
}
```

---

## Simulating Behavior with the GCM

```{r ch12_gcm_agent_wrapper}
simulate_gcm_agent <- function(agent_id, w_true, c_true, bias_true,
                               stimulus_info, n_blocks, subject_seed) {
  schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  obs      <- schedule |> dplyr::select(height, position)
  feedback <- schedule$category_feedback

  sim_results <- gcm_simulate(
    w = w_true, c = c_true, bias = bias_true,
    obs = obs, cat_feedback = feedback, quiet = TRUE
  )

  schedule |>
    mutate(
      agent_id     = agent_id,
      w1_true      = w_true[1],
      w2_true      = w_true[2],
      c_true       = c_true,
      log_c_true   = log(c_true),
      bias_true    = bias_true,
      prob_cat1    = sim_results$prob_cat1,
      sim_response = sim_results$sim_response,
      correct      = as.integer(category_feedback == sim_response)
    ) |>
    group_by(agent_id) |>
    mutate(performance = cumsum(correct) / row_number()) |>
    ungroup()
}
```

```{r ch12_gcm_simulations}
param_grid <- expand_grid(
  w_setting    = list(c(0.5, 0.5), c(0.7, 0.3), c(0.9, 0.1)),
  c_setting    = c(0.5, 1.0, 2.0, 4.0, 5.0, 15.0),
  bias_setting = 0.5
) |>
  mutate(
    setting_id = row_number(),
    w_label    = map_chr(w_setting, ~paste(round(.x, 1), collapse = ",")),
    c_label    = paste0("c≈", round(c_setting, 1)),
    label      = paste0("w=(", w_label, "), ", c_label)
  )

n_agents_per_setting <- 5

simulation_plan <- param_grid |>
  uncount(n_agents_per_setting) |>
  mutate(agent_id = row_number())

sim_file <- here("simdata", "ch12_gcm_simulated_responses.csv")

if (regenerate_simulations || !file.exists(sim_file)) {
  plan(multisession, workers = max(1, availableCores() - 1))

  simulated_responses <- future_pmap_dfr(
    list(
      agent_id  = simulation_plan$agent_id,
      w_true    = simulation_plan$w_setting,
      c_true    = simulation_plan$c_setting,
      bias_true = simulation_plan$bias_setting
    ),
    simulate_gcm_agent,
    stimulus_info = stimulus_info,
    n_blocks      = n_blocks,
    subject_seed  = simulation_plan$agent_id,
    .options = furrr_options(seed = TRUE)
  )

  plan(sequential)
  write_csv(simulated_responses, sim_file)
  cat("Simulation complete. Results saved.\n")
} else {
  simulated_responses <- read_csv(sim_file, show_col_types = FALSE)
  cat("Loaded existing simulation results.\n")
}
```

```{r ch12_visualize_parameter_effects}
simulated_responses_labeled <- simulated_responses |>
  mutate(
    w_label = paste0("w=(", round(w1_true, 1), ",", round(w2_true, 1), ")"),
    c_label = paste0("c≈", round(c_true, 1)),
    # Order the c_label factor based on the numeric c_true values
    c_label = forcats::fct_reorder(c_label, c_true) 
  )

p_c_effect <- ggplot(
  simulated_responses_labeled,
  aes(x = trial_within_subject, y = performance,
      color = factor(round(c_true, 1)),
      group = interaction(agent_id, round(c_true, 1)))
) +
  stat_summary(fun = mean, geom = "line",
               aes(group = factor(round(c_true, 1))), linewidth = 1.2) +
  facet_wrap(~w_label) +
  scale_color_viridis_d(option = "plasma", name = "Sensitivity (c)") +
  labs(
    title    = "Effect of Sensitivity (c) on GCM Learning",
    subtitle = "Mean across simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme(legend.position = "bottom")

p_w_effect <- ggplot(
  simulated_responses_labeled,
  aes(x = trial_within_subject, y = performance,
      color = w_label, group = interaction(agent_id, w_label))
) +
  stat_summary(fun = mean, geom = "line", aes(group = w_label), linewidth = 1.2) +
  facet_wrap(~c_label) +
  scale_color_brewer(palette = "Set1", name = "Attention Weights") +
  labs(
    title    = "Effect of Attention Weights (w) on GCM Learning",
    subtitle = "Mean across simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme(legend.position = "bottom")

p_c_effect
p_w_effect
```

---

## Prior Predictive Check

We need to define the priors for the Stan model already here:

- $w \sim \text{Dirichlet}(1, 1)$ — uniform over the simplex
- $\log c \sim \text{Normal}(\log(2), 1)$ — implies $c \in (0.27, 14.8)$ at $\pm 2$ SD
- $\text{bias} \sim \text{Beta}(1, 1)$ — uniform over $(0, 1)$

```{r ch12_prior_predictive}
n_prior_samples <- 500

prior_draws <- tibble(
  sample_id = seq_len(n_prior_samples),
  log_c     = rnorm(n_prior_samples, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD),
  w1        = runif(n_prior_samples),
  bias      = rbeta(n_prior_samples, 1, 1)
) |>
  mutate(c_true = exp(log_c), w2 = 1 - w1)

ppc_schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = 999)

prior_pred_curves <- pmap_dfr(
  list(
    sample_id = prior_draws$sample_id,
    log_c     = prior_draws$log_c,
    w1        = prior_draws$w1,
    bias      = prior_draws$bias
  ),
  function(sample_id, log_c, w1, bias) {
    c <- exp(log_c)
    sim <- gcm_simulate(
      w = c(w1, 1 - w1), c = c, bias = bias,
      obs          = ppc_schedule |> dplyr::select(height, position),
      cat_feedback = ppc_schedule$category_feedback
    )
    tibble(
      sample_id = sample_id,
      trial     = seq_len(nrow(ppc_schedule)),
      correct   = as.integer(sim$sim_response == ppc_schedule$category_feedback)
    ) |>
      mutate(cum_acc = cumsum(correct) / row_number())
  }
)

prior_pred_summary <- prior_pred_curves |>
  group_by(trial) |>
  summarise(
    q05 = quantile(cum_acc, 0.05),
    q25 = quantile(cum_acc, 0.25),
    q50 = median(cum_acc),
    q75 = quantile(cum_acc, 0.75),
    q95 = quantile(cum_acc, 0.95),
    .groups = "drop"
  )

ggplot(prior_pred_summary, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#0072B2", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Prior Predictive Check: GCM Learning Curves",
    subtitle = "Ribbons: 50% and 90% prior predictive intervals; line: median\nPriors: log(c) ~ Normal(log 2, 1),  w ~ Dirichlet(1,1),  bias ~ Beta(1,1)",
    x = "Trial", y = "Cumulative Accuracy"
  )
```

---

## Implementing the GCM in Stan for Parameter Estimation

### Architectural choice: `prob_cat1[i]` in `transformed parameters`

Because the GCM's per-trial choice probability $p_i$ is a deterministic function of the parameters $(w, c, \beta)$ and the data sequence, it belongs in `transformed parameters` as a single source of truth. The `model{}` block then becomes a one-line vectorised Bernoulli likelihood, and `generated quantities{}` becomes a lookup against the already-computed `prob_cat1` vector.

```{r ch12_gcm_stan_model}
gcm_single_stan <- "
// Generalized Context Model — Single Subject (refactored architecture)
//
// Key design:
//   prob_cat1[i] is the per-trial choice probability. It is deterministic
//   given (w, log_c, bias) and the data sequence, so it lives in
//   transformed parameters. model{} reads it once via bernoulli_lpmf.
//   generated quantities{} reads the same vector for log_lik[i].
//
// Parameterization:
//   w      : attention weights (simplex)
//   log_c  : log sensitivity (unconstrained); c = exp(log_c)
//   bias   : response bias towards category 1 (0–1)

data {
  int<lower=1> ntrials;
  int<lower=1> nfeatures;
  array[ntrials] int<lower=0, upper=1> y;
  array[ntrials, nfeatures] real obs;
  array[ntrials] int<lower=0, upper=1> cat_feedback;

  vector[nfeatures] w_prior_alpha;
  real log_c_prior_mean;
  real<lower=0> log_c_prior_sd;
  real<lower=0> bias_prior_alpha;
  real<lower=0> bias_prior_beta;
}

transformed data {
  array[ntrials, ntrials, nfeatures] real abs_diff;

  for (i in 1:ntrials) {
    for (j in 1:ntrials) {
      for (f in 1:nfeatures) {
        abs_diff[i, j, f] = abs(obs[i, f] - obs[j, f]);
      }
    }
  }
}

parameters {
  simplex[nfeatures] w;
  real log_c;
  real<lower=0, upper=1> bias;
}

transformed parameters {
  real<lower=0> c = exp(log_c);

  vector<lower=1e-9, upper=1-1e-9>[ntrials] prob_cat1;

  {
    array[ntrials] int memory_trial_idx;
    array[ntrials] int memory_cat;
    int n_mem = 0;

    for (i in 1:ntrials) {
      real p_i;
      int has_cat0 = 0;
      int has_cat1 = 0;

      for (k in 1:n_mem) {
        if (memory_cat[k] == 0) has_cat0 = 1;
        if (memory_cat[k] == 1) has_cat1 = 1;
      }

      if (n_mem == 0 || has_cat0 == 0 || has_cat1 == 0) {
        p_i = bias;
      } else {
        vector[n_mem] sims;
        for (e in 1:n_mem) {
          real d = 0;
          int past_i = memory_trial_idx[e];
          for (f in 1:nfeatures)
            d += w[f] * abs_diff[i, past_i, f];
          sims[e] = exp(-c * d);
        }
        real s1 = 0;
        real s0 = 0;
        for (e in 1:n_mem) {
          if (memory_cat[e] == 1) s1 += sims[e];
          else                    s0 += sims[e];
        }
        real num = bias * s1;
        real den = num + (1 - bias) * s0;
        p_i = (den > 1e-9) ? num / den : bias;
      }

      prob_cat1[i] = fmax(1e-9, fmin(1 - 1e-9, p_i));

      n_mem += 1;
      memory_trial_idx[n_mem] = i;
      memory_cat[n_mem] = cat_feedback[i];
    }
  }
}

model {
  target += dirichlet_lpdf(w    | w_prior_alpha);
  target += normal_lpdf(log_c   | log_c_prior_mean, log_c_prior_sd);
  target += beta_lpdf(bias      | bias_prior_alpha, bias_prior_beta);

  target += bernoulli_lpmf(y | prob_cat1);
}

generated quantities {
  vector[ntrials] log_lik;
  real lprior;
  for (i in 1:ntrials)
    log_lik[i] = bernoulli_lpmf(y[i] | prob_cat1[i]);
  lprior = dirichlet_lpdf(w | w_prior_alpha) +
           normal_lpdf(log_c | log_c_prior_mean, log_c_prior_sd) +
           beta_lpdf(bias | bias_prior_alpha, bias_prior_beta);
}
"

stan_file_gcm_single <- "stan/ch12_gcm_single.stan"
write_stan_file(gcm_single_stan, dir = "stan/", basename = "ch12_gcm_single.stan")
mod_gcm_single <- cmdstan_model(stan_file_gcm_single)
```

### What each block does:

* **`transformed data`**: Builds the `abs_diff[i, j, f]` tensor. Evaluated exactly once per chain.
* **`parameters`**: `simplex[nfeatures] w`, unconstrained `real log_c`, and `real<lower=0, upper=1> bias`.
* **`transformed parameters`**: Caches `prob_cat1[i]` in a single forward pass over trials.
* **`model`**: Priors followed by a vectorised one-line likelihood.
* **`generated quantities`**: `log_lik[i]` as a per-trial lookup, and `lprior` for `priorsense`.

> A note on reduce_sum and parallelization: The GCM's inner loop cannot be parallelized
> with reduce_sum because trial $t$'s contribution to target depends on the memory state
> accumulated sequentially from trials $1, \ldots, t-1$.

### Fitting the Single-Subject GCM

```{r ch12_fitting_gcm_single}
# Select one simulated agent: w ≈ (0.9, 0.1), c ≈ 2
agent_to_fit <- simulated_responses |>
  filter(
    round(w1_true, 1) == 0.9,
    abs(log_c_true - log(2.0)) == min(abs(log_c_true - log(2.0)))
  ) |>
  filter(agent_id == min(agent_id)) |>
  slice(1:total_trials)

stopifnot(nrow(agent_to_fit) == total_trials)

gcm_data_single <- list(
  ntrials          = nrow(agent_to_fit),
  nfeatures        = 2,
  y                = agent_to_fit$sim_response,
  obs              = as.matrix(agent_to_fit[, c("height", "position")]),
  cat_feedback     = agent_to_fit$category_feedback,
  w_prior_alpha    = c(1, 1),
  log_c_prior_mean = LOG_C_PRIOR_MEAN,
  log_c_prior_sd   = LOG_C_PRIOR_SD,
  bias_prior_alpha = 1,
  bias_prior_beta  = 1
)

fit_filepath_single <- here("simmodels", "ch12_gcm_single_fit.rds")

if (regenerate_simulations || !file.exists(fit_filepath_single)) {
  pf_single <- tryCatch(
    mod_gcm_single$pathfinder(
      data      = gcm_data_single,
      seed      = 123,
      num_paths = 4,
      refresh   = 0
    ),
    error = function(e) {
      message("Pathfinder failed — falling back to default initialisation.")
      NULL
    }
  )

  fit_gcm_single <- mod_gcm_single$sample(
    data            = gcm_data_single,
    init            = pf_single,
    seed            = 123,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1000,
    iter_sampling   = 1500,
    refresh         = 300,
    adapt_delta     = 0.9
  )
  fit_gcm_single$save_object(fit_filepath_single)
  cat("Single-agent model fitted and saved.\n")
} else {
  fit_gcm_single <- readRDS(fit_filepath_single)
  cat("Loaded existing single-agent fit.\n")
}
```

### MCMC Diagnostic Battery

```{r ch12_single_diagnostic_battery}
if (!is.null(fit_gcm_single)) {
  diag_tbl <- diagnostic_summary_table(
    fit_gcm_single,
    params = c("w[1]", "w[2]", "log_c", "c", "bias")
  )
  print(diag_tbl)

  if (!all(diag_tbl$pass)) {
    warning("Diagnostic battery has failures — interpret posteriors with caution.")
  }

  bayesplot::mcmc_trace(
    fit_gcm_single$draws(c("log_c", "w[1]", "bias")),
    facet_args = list(ncol = 1)
  )

  bayesplot::mcmc_rank_overlay(
    fit_gcm_single$draws(c("log_c", "w[1]", "bias"))
  )
}
```

### Posterior Predictive Check (single subject)

```{r ch12_ppc_single}
if (!is.null(fit_gcm_single)) {
  prob_draws <- fit_gcm_single$draws("prob_cat1", format = "matrix")
  n_draws    <- nrow(prob_draws)
  n_trials   <- ncol(prob_draws)

  set.seed(2025)
  yrep_post <- matrix(rbinom(n_draws * n_trials, 1, as.vector(prob_draws)),
                      nrow = n_draws)

  obs_acc <- cummean(agent_to_fit$sim_response == agent_to_fit$category_feedback)
  rep_acc <- t(apply(yrep_post, 1, function(yr) {
    cummean(yr == agent_to_fit$category_feedback)
  }))

  ppc_summary <- tibble(
    trial = seq_len(n_trials),
    obs   = obs_acc,
    q05   = apply(rep_acc, 2, quantile, 0.05),
    q25   = apply(rep_acc, 2, quantile, 0.25),
    q50   = apply(rep_acc, 2, quantile, 0.50),
    q75   = apply(rep_acc, 2, quantile, 0.75),
    q95   = apply(rep_acc, 2, quantile, 0.95)
  )

  ggplot(ppc_summary, aes(x = trial)) +
    geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
    geom_line(aes(y = q50), color = "#0072B2", linewidth = 0.8) +
    geom_line(aes(y = obs), color = "#D55E00", linewidth = 1) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
      title    = "Posterior Predictive Check: Cumulative Accuracy",
      subtitle = "Blue ribbons: 50% and 90% posterior predictive intervals\nOrange line: observed cumulative accuracy",
      x = "Trial", y = "Cumulative Accuracy"
    )
}
```

### Prior–Posterior Update Plot

```{r ch12_prior_posterior_single}
if (!is.null(fit_gcm_single)) {
  draws_df <- as_draws_df(
    fit_gcm_single$draws(variables = c("w[1]", "log_c", "c", "bias"))
  )

  true_w1    <- agent_to_fit$w1_true[1]
  true_log_c <- agent_to_fit$log_c_true[1]
  true_c     <- agent_to_fit$c_true[1]
  true_bias  <- agent_to_fit$bias_true[1]

  set.seed(2026)
  n_prior <- 4000
  prior_df <- tibble(
    `w[1]` = runif(n_prior),
    log_c  = rnorm(n_prior, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD),
    c      = exp(log_c),
    bias   = rbeta(n_prior, 1, 1)
  )

  pp_panel <- function(prior_x, post_x, true_val, label) {
    tibble(
      x      = c(prior_x, post_x),
      source = c(rep("Prior", length(prior_x)),
                 rep("Posterior", length(post_x)))
    ) |>
      ggplot(aes(x = x, fill = source, color = source)) +
      geom_density(alpha = 0.4, linewidth = 0.5) +
      geom_vline(xintercept = true_val, color = "red",
                 linetype = "dashed", linewidth = 0.7) +
      scale_fill_manual(values = c(Prior = "grey60", Posterior = "#0072B2")) +
      scale_color_manual(values = c(Prior = "grey40", Posterior = "#0072B2")) +
      labs(title = label, x = NULL, y = "Density") +
      theme(legend.position = "bottom")
  }

  p1 <- pp_panel(prior_df$`w[1]`, draws_df$`w[1]`, true_w1, "w[1]")
  p2 <- pp_panel(prior_df$log_c,  draws_df$log_c,  true_log_c, "log(c) — computational scale")
  p3 <- pp_panel(prior_df$c,      draws_df$c,      true_c,     "c — natural scale")
  p4 <- pp_panel(prior_df$bias,   draws_df$bias,   true_bias,  "bias")

  (p1 | p2) / (p3 | p4) +
    plot_annotation(
      title    = "Prior–Posterior Update for the Single-Subject GCM",
      subtitle = "Red dashed line = true generating value"
    )
}
```

### Joint Posterior Geometry: $c$ vs $w$

```{r ch12_cw_identifiability_pairs}
if (!is.null(fit_gcm_single)) {
  bayesplot::mcmc_pairs(
    fit_gcm_single$draws(c("log_c", "w[1]", "bias")),
    diag_fun     = "dens",
    off_diag_fun = "hex"
  )
}
```

---

### LOO with Pareto-$\hat{k}$ Diagnostics

Standard PSIS-LOO is computed for completeness and for comparison against LFO-CV below. We do not interpret it as a valid measure of out-of-sample predictive accuracy for a sequential model — see the next subsection for why — but the Pareto-$\hat{k}$ diagnostic still localises trials where the path-dependence concern bites hardest.

```{r ch12_loo_single}
if (!is.null(fit_gcm_single)) {
  log_lik_mat    <- fit_gcm_single$draws("log_lik", format = "matrix")
  loo_gcm_single <- loo::loo(log_lik_mat, cores = 4)
  print(loo_gcm_single)

  plot(loo_gcm_single, label_points = TRUE,
       main = "PSIS-LOO Pareto-k diagnostics: Single-subject GCM")
}
```

### Leave-Future-Out Cross-Validation

**Why standard LOO is invalid here.** Standard PSIS-LOO assumes observations are conditionally exchangeable given the parameters. In the GCM, this assumption is violated: leaving out observation $t$ also conceptually removes the memory trace that observation $t$ contributed to trials $t+1, \ldots, T$.

**What LFO-CV does instead.** LFO-CV (Bürkner, Gabry, & Vehtari, 2020) estimates the *one-step-ahead* predictive density. The PSIS approximation refits only when the importance sampling deteriorates ($\hat{k} > k_\text{thresh}$) and uses importance reweighting otherwise.

```{r ch12_lfo_cv_implementation, eval=run_intensive_checks}
# ─────────────────────────────────────────────────────────────────────────────
# psis_lfo_gcm: PSIS-approximated leave-future-out CV for the GCM.
# Follows Bürkner, Gabry, & Vehtari (2020), Algorithm 1.
# ─────────────────────────────────────────────────────────────────────────────
psis_lfo_gcm <- function(stan_model,
                         full_data,
                         L = 16,
                         k_thresh = 0.7,
                         seed = 1,
                         verbose = FALSE) {

  fit_subset <- function(n_train, seed) {
    sub_data <- full_data
    sub_data$ntrials      <- n_train
    sub_data$y            <- sub_data$y[1:n_train]
    sub_data$obs          <- sub_data$obs[1:n_train, , drop = FALSE]
    sub_data$cat_feedback <- sub_data$cat_feedback[1:n_train]

    pf  <- stan_model$pathfinder(data = sub_data, num_paths = 4, refresh = 0,
                                  seed = seed)
    fit <- stan_model$sample(
      data = sub_data, init = pf, seed = seed,
      chains = 2, parallel_chains = 2,
      iter_warmup = 600, iter_sampling = 800,
      refresh = 0, adapt_delta = 0.9
    )
    fit
  }

  predict_future <- function(fit, full_data) {
    full_n   <- length(full_data$y)
    ext_data <- full_data
    ext_data$ntrials <- full_n

    gq_fit <- stan_model$generate_quantities(
      fitted_params   = fit,
      data            = ext_data,
      seed            = seed,
      parallel_chains = 2
    )
    gq_fit$draws("log_lik", format = "matrix")
  }

  T_total      <- length(full_data$y)
  elpd_lfo     <- numeric(T_total)
  elpd_lfo[]   <- NA_real_
  refits       <- integer(0)
  k_history    <- numeric(0)

  fit_curr     <- fit_subset(L, seed = seed)
  ll_full_curr <- predict_future(fit_curr, full_data)

  refits       <- c(refits, L)
  last_fit_n   <- L
  n_draws      <- nrow(ll_full_curr)

  ll_train_curr <- rowSums(ll_full_curr[, 1:L, drop = FALSE])

  for (t in (L + 1):T_total) {
    if (t - 1 == last_fit_n) {
      k_t          <- 0
      k_history[t] <- k_t
      lw           <- rep(-log(n_draws), n_draws)
    } else {
      ll_train_target <- rowSums(ll_full_curr[, 1:(t - 1), drop = FALSE])
      log_w           <- ll_train_target - ll_train_curr

      psis_obj     <- loo::psis(log_w, r_eff = NA)
      k_t          <- loo::pareto_k_values(psis_obj)
      k_history[t] <- k_t

      if (k_t > k_thresh) {
        if (verbose) cat("Refit at t =", t, "(k =", round(k_t, 3), ")\n")
        fit_curr      <- fit_subset(t - 1, seed = seed + t)
        ll_full_curr  <- predict_future(fit_curr, full_data)
        ll_train_curr <- rowSums(ll_full_curr[, 1:(t - 1), drop = FALSE])
        last_fit_n    <- t - 1
        refits        <- c(refits, last_fit_n)
        lw            <- rep(-log(n_draws), n_draws)
      } else {
        lw <- as.numeric(weights(psis_obj))
      }
    }

    lp          <- ll_full_curr[, t]
    elpd_lfo[t] <- matrixStats::logSumExp(lw + lp)
  }

  list(
    elpd_lfo       = sum(elpd_lfo, na.rm = TRUE),
    elpd_pointwise = elpd_lfo,
    refits         = refits,
    k_history      = k_history,
    L              = L
  )
}

lfo_filepath <- here("simdata", "ch12_lfo_single.rds")
if (regenerate_simulations || !file.exists(lfo_filepath)) {
  lfo_single <- psis_lfo_gcm(
    stan_model = mod_gcm_single,
    full_data  = gcm_data_single,
    L          = 16,
    k_thresh   = 0.7,
    seed       = 11,
    verbose    = TRUE
  )
  saveRDS(lfo_single, lfo_filepath)
} else {
  lfo_single <- readRDS(lfo_filepath)
}

cat("LFO-CV ELPD (single subject):", round(lfo_single$elpd_lfo, 2), "\n")
cat("Number of refits:",             length(lfo_single$refits),     "\n")
cat("Refit triggered at trials:",    lfo_single$refits,             "\n")

if (!is.null(fit_gcm_single)) {
  loo_pw  <- loo_gcm_single$pointwise[, "elpd_loo"]
  comp_df <- tibble(
    trial    = seq_along(loo_pw),
    elpd_loo = loo_pw,
    elpd_lfo = lfo_single$elpd_pointwise
  ) |>
    pivot_longer(starts_with("elpd_"), names_to = "method", values_to = "elpd")

  ggplot(comp_df, aes(x = trial, y = elpd, color = method)) +
    geom_line(alpha = 0.7) +
    geom_point(alpha = 0.5) +
    scale_color_manual(values = c(elpd_loo = "#0072B2", elpd_lfo = "#D55E00"),
                       labels = c("PSIS-LFO (correct)",
                                  "PSIS-LOO (invalid for sequential)")) +
    labs(
      title    = "Per-trial ELPD: LOO vs LFO-CV",
      subtitle = "Differences are largest where the path-dependence assumption fails most",
      x = "Trial", y = "Pointwise ELPD"
    ) +
    theme(legend.position = "bottom")
}
```

**Validation against exact one-step-ahead refits.**

```{r ch12_exact_lfo_validation, eval=run_intensive_checks}
exact_lfo_grid     <- seq(16, length(gcm_data_single$y) - 1, by = 8)
exact_lfo_filepath <- here("simdata", "ch12_exact_lfo_single.rds")

if (regenerate_simulations || !file.exists(exact_lfo_filepath)) {
  exact_results <- map_dfr(exact_lfo_grid, function(n_train) {
    sub_data <- gcm_data_single
    sub_data$ntrials      <- n_train
    sub_data$y            <- sub_data$y[1:n_train]
    sub_data$obs          <- sub_data$obs[1:n_train, , drop = FALSE]
    sub_data$cat_feedback <- sub_data$cat_feedback[1:n_train]

    pf  <- mod_gcm_single$pathfinder(data = sub_data, num_paths = 4,
                                      refresh = 0, seed = n_train)
    fit <- mod_gcm_single$sample(
      data = sub_data, init = pf, seed = n_train,
      chains = 2, parallel_chains = 2,
      iter_warmup = 600, iter_sampling = 800,
      refresh = 0, adapt_delta = 0.9
    )

    ext <- gcm_data_single
    gq  <- mod_gcm_single$generate_quantities(
      fitted_params = fit, data = ext, seed = n_train, parallel_chains = 2
    )
    ll_target  <- gq$draws("log_lik", format = "matrix")[, n_train + 1]
    elpd_exact <- matrixStats::logSumExp(ll_target) - log(length(ll_target))

    tibble(trial = n_train + 1, elpd_exact = elpd_exact)
  })
  saveRDS(exact_results, exact_lfo_filepath)
} else {
  exact_results <- readRDS(exact_lfo_filepath)
}

psis_at_grid <- tibble(
  trial     = exact_results$trial,
  elpd_psis = lfo_single$elpd_pointwise[exact_results$trial]
)

agreement_df <- exact_results |>
  left_join(psis_at_grid, by = "trial")

ggplot(agreement_df, aes(x = elpd_exact, y = elpd_psis)) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_point(size = 3, color = "#0072B2") +
  labs(
    title    = "PSIS-LFO vs Exact One-Step-Ahead Refits",
    subtitle = "Points should fall on the y = x line if PSIS-LFO is calibrated",
    x = "Exact ELPD (refit per point)",
    y = "PSIS-LFO ELPD"
  ) +
  coord_fixed()
```

### Randomised LOO-PIT

```{r ch12_randomized_loo_pit}
if (!is.null(fit_gcm_single)) {
  prob_draws  <- fit_gcm_single$draws("prob_cat1", format = "matrix")
  log_lik_mat <- fit_gcm_single$draws("log_lik",   format = "matrix")

  psis_obj <- loo::psis(-log_lik_mat, r_eff = NA)
  lw       <- weights(psis_obj, normalize = TRUE)

  y_obs <- gcm_data_single$y

  loo_p_cat1 <- vapply(seq_along(y_obs), function(t) {
    w <- exp(lw[, t] - matrixStats::logSumExp(lw[, t]))
    sum(w * prob_draws[, t])
  }, numeric(1))

  set.seed(2027)
  pit_vals <- ifelse(
    y_obs == 1,
    runif(length(y_obs), 1 - loo_p_cat1, 1),
    runif(length(y_obs), 0,              1 - loo_p_cat1)
  )

  ggplot(tibble(pit = pit_vals), aes(x = pit)) +
    geom_histogram(aes(y = after_stat(density)),
                   bins = 20, fill = "#0072B2", alpha = 0.7,
                   boundary = 0) +
    geom_hline(yintercept = 1, color = "red", linetype = "dashed") +
    scale_x_continuous(limits = c(0, 1)) +
    labs(
      title    = "Randomised LOO-PIT for the Single-Subject GCM",
      subtitle = "If the model is calibrated, PIT values are Uniform(0,1) (red dashed line)",
      x = "Randomised LOO-PIT", y = "Density"
    )
}
```

### Prior Sensitivity Analysis

```{r ch12_prior_sensitivity_single}
if (!is.null(fit_gcm_single)) {
  ps <- priorsense::powerscale_sensitivity(fit_gcm_single)
  print(ps)

  priorsense::powerscale_plot_dens(
    fit_gcm_single,
    variables = c("log_c", "w[1]", "bias")
  )
}
```

---

## Single-Subject Simulation-Based Calibration (SBC)

SBC checks the internal consistency and calibration of the entire Bayesian inference procedure across the *whole* prior. It asks: "If the true parameters really came from our specified priors, would our inference procedure produce posteriors that are statistically consistent with those priors?"

```{r ch12_sbc_single, eval=run_intensive_checks}
sbc_single_filepath <- here("simdata", "ch12_sbc_single_results.rds")

if (regenerate_simulations || !file.exists(sbc_single_filepath)) {
  n_sbc_iterations_single <- 500   # use >= 1000 for publication-quality SBC

  gcm_single_sbc_generator <- SBC_generator_function(
    function() {
      w_true     <- as.numeric(MCMCpack::rdirichlet(1, c(1, 1)))
      log_c_true <- rnorm(1, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD)
      c_true     <- exp(log_c_true)
      bias_true  <- rbeta(1, 1, 1)

      sbc_schedule <- make_subject_schedule(stimulus_info, n_blocks,
                                            seed = sample.int(1e8, 1))

      sim <- gcm_simulate(
        w            = w_true,
        c            = c_true,
        bias         = bias_true,
        obs          = sbc_schedule |> dplyr::select(height, position),
        cat_feedback = sbc_schedule$category_feedback
      )

      list(
        variables = list(
          `w[1]` = w_true[1],
          `w[2]` = w_true[2],
          log_c  = log_c_true,
          bias   = bias_true
        ),
        generated = list(
          ntrials          = nrow(sbc_schedule),
          nfeatures        = 2L,
          y                = as.integer(sim$sim_response),
          obs              = as.matrix(sbc_schedule[, c("height", "position")]),
          cat_feedback     = as.integer(sbc_schedule$category_feedback),
          w_prior_alpha    = c(1, 1),
          log_c_prior_mean = LOG_C_PRIOR_MEAN,
          log_c_prior_sd   = LOG_C_PRIOR_SD,
          bias_prior_alpha = 1,
          bias_prior_beta  = 1
        )
      )
    }
  )

  gcm_single_sbc_backend <- SBC_backend_cmdstan_sample(
    mod_gcm_single,
    iter_warmup   = 1000,
    iter_sampling = 1000,
    chains        = 2,
    adapt_delta   = 0.9,
    refresh       = 0
  )

  gcm_single_datasets <- generate_datasets(
    gcm_single_sbc_generator,
    n_sims = n_sbc_iterations_single
  )

  sbc_results_single <- compute_SBC(
    gcm_single_datasets,
    gcm_single_sbc_backend,
    cache_mode     = "results",
    cache_location = here("simdata", "ch12_sbc_single_cache"),
    keep_fits      = FALSE
  )
  saveRDS(sbc_results_single, sbc_single_filepath)
  cat("SBC single-subject results computed and saved.\n")
} else {
  sbc_results_single <- readRDS(sbc_single_filepath)
  cat("Loaded existing SBC single-subject results.\n")
}

plot_ecdf_diff(sbc_results_single)
plot_rank_hist(sbc_results_single)

# ── Diagnostic issue investigation ───────────────────────────────────────────
backend_diags  <- sbc_results_single$backend_diagnostics
default_diags  <- sbc_results_single$default_diagnostics |>
  dplyr::select(sim_id, max_rhat, min_ess_to_rank)
true_params_wide <- sbc_results_single$stats |>
  dplyr::select(sim_id, variable, simulated_value) |>
  tidyr::pivot_wider(names_from = variable, values_from = simulated_value)

diag_df <- backend_diags |>
  dplyr::left_join(default_diags,    by = "sim_id") |>
  dplyr::left_join(true_params_wide, by = "sim_id") |>
  dplyr::mutate(has_issue = n_divergent > 0 | n_max_treedepth > 0 | max_rhat > 1.01)

p_diag_c_bias <- ggplot(diag_df, aes(x = log_c, y = bias, color = has_issue)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00")) +
  labs(title    = "Diagnostic Issues in Parameter Space",
       subtitle = "Orange points: divergence, max treedepth, or Rhat > 1.01",
       x = "True log(c)", y = "True Bias") +
  theme(legend.position = "bottom")

p_diag_w_c <- ggplot(diag_df, aes(x = `w[1]`, y = log_c, color = has_issue)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00")) +
  labs(title = "Attention vs Sensitivity", x = "True w[1]", y = "True log(c)") +
  theme(legend.position = "none")

print(p_diag_c_bias | p_diag_w_c)

# ── Parameter recovery on SBC simulations ────────────────────────────────────
recovery_df <- sbc_results_single$stats

p_sbc_recovery <- ggplot(recovery_df, aes(x = simulated_value, y = mean)) +
  geom_point(alpha = 0.4, color = "#0072B2") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~variable, scales = "free") +
  labs(
    title    = "Parameter Recovery Across SBC Simulations",
    subtitle = "Posterior mean vs True generated value",
    x = "True Value (simulated_value)",
    y = "Estimated Value (Posterior Mean)"
  ) +
  theme_bw()

print(p_sbc_recovery)
```

When the true log(c) is above 0 the model computes similarity using exponential decay: `exp(-c * distance)`. When c gets large, similarity drops to exactly zero for anything except perfectly identical stimuli, causing gradient collapse and HMC failures. The diagnostic scatter plots above reveal that divergences and Rhat violations cluster at high log(c) values, confirming that the prior mass above log(c) ≈ 2 (c ≈ 7–8) is unsupported by the sampler geometry.

**Practical constraint from SBC.** We treat c ≈ 12 as a hard ceiling: a sensitivity this extreme makes stimuli that differ by any measurable amount effectively identical, which is psychologically implausible for the feature ranges used here. We therefore require that the prior places negligible mass above c = 12, i.e., that the 99.85th percentile (mean + 3 SD on the log scale) stays below log(12) ≈ 2.49. For the single-subject model this is satisfied with `Normal(log(2), 1.0)`, whose +3 SD upper bound is exp(0.69 + 3.0) ≈ 20 — already at the boundary of acceptability. We carry this constraint forward explicitly into the multilevel parameterisation.

---

# Multilevel Generalized Context Model (GCM)

## The Need for a Multilevel Approach

The multilevel (hierarchical) approach assumes that individual participant parameters are drawn from overarching group-level distributions, allowing us to estimate parameters for each individual informed by the group trend (partial pooling).

### Motivation: Visualizing Individual Differences

```{r ch12_motivate_multilevel, eval=run_intensive_checks}
motivate_ml_filepath <- here("simdata", "ch12_motivate_multilevel.rds")

if (regenerate_simulations || !file.exists(motivate_ml_filepath)) {
  n_demo_agents       <- 10
  demo_pop_log_c_mean <- log(2.0)
  demo_pop_log_c_sd   <- 0.5

  demo_true_log_c <- rnorm(n_demo_agents, demo_pop_log_c_mean, demo_pop_log_c_sd)

  demo_agent_data <- purrr::map_dfr(seq_len(n_demo_agents), function(j) {
    simulate_gcm_agent(
      agent_id      = j,
      w_true        = c(0.5, 0.5),
      c_true        = exp(demo_true_log_c[j]),
      bias_true     = 0.5,
      stimulus_info = stimulus_info,
      n_blocks      = n_blocks,
      subject_seed  = 8000 + j
    )
  })

  demo_fits <- purrr::map(
    seq_len(n_demo_agents),
    function(agent_id) {
      agent_dat <- demo_agent_data |> filter(agent_id == !!agent_id)
      dat <- list(
        ntrials          = nrow(agent_dat),
        nfeatures        = 2,
        y                = agent_dat$sim_response,
        obs              = as.matrix(agent_dat[, c("height", "position")]),
        cat_feedback     = agent_dat$category_feedback,
        w_prior_alpha    = c(1, 1),
        log_c_prior_mean = LOG_C_PRIOR_MEAN,
        log_c_prior_sd   = LOG_C_PRIOR_SD,
        bias_prior_alpha = 1,
        bias_prior_beta  = 1
      )

      pf  <- mod_gcm_single$pathfinder(data = dat, seed = agent_id,
                                       num_paths = 4, refresh = 0)
      fit <- mod_gcm_single$sample(data = dat, init = pf, seed = agent_id,
                                    chains = 2, parallel_chains = 2,
                                    iter_warmup = 800, iter_sampling = 1000,
                                    refresh = 0, adapt_delta = 0.9)
      list(agent_id = agent_id, fit = fit)
    }
  )

  demo_estimates <- map_dfr(demo_fits, function(x) {
    x$fit$summary(variables = c("c", "bias")) |>
      dplyr::select(variable, mean, q5, q95) |>
      mutate(agent_id = x$agent_id)
  })

  demo_draws <- purrr::map_dfr(demo_fits, function(x) {
    draws <- as_draws_df(x$fit$draws(variables = c("c", "w[1]", "bias")))
    draws$agent_id <- factor(x$agent_id)
    return(draws)
  })

  saveRDS(list(demo_estimates = demo_estimates, demo_draws = demo_draws),
          motivate_ml_filepath)
  cat("Multilevel motivation fits computed and saved.\n")
} else {
  saved <- readRDS(motivate_ml_filepath)
  demo_estimates <- saved$demo_estimates
  demo_draws     <- saved$demo_draws
  cat("Loaded existing multilevel motivation fits.\n")
}

ggplot(demo_estimates, aes(x = factor(agent_id), y = mean,
                           ymin = q5, ymax = q95, color = variable)) +
  geom_pointrange(position = position_dodge(0.4)) +
  facet_wrap(~variable, scales = "free_y") +
  labs(
    title    = "Individual Posterior Estimates: 10 Agents Fitted Separately",
    subtitle = "Points: posterior mean; ranges: 90% CrI. Spread motivates the multilevel approach.",
    x = "Agent", y = "Estimated Value"
  ) +
  theme(legend.position = "none")

p_joint_cw <- ggplot(demo_draws, aes(x = `w[1]`, y = c, group = agent_id)) +
  stat_ellipse(geom = "polygon", aes(fill = agent_id),
               level = 0.50, alpha = 0.15, color = NA) +
  stat_ellipse(geom = "path", aes(color = agent_id),
               level = 0.50, alpha = 0.8, linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = 0.5, linetype = "dotted", color = "grey50") +
  scale_x_continuous(limits = c(0, 1)) +
  labs(title    = "Joint Posterior: w[1] vs c",
       subtitle = "50% credible ellipses.",
       x = "Attention Weight w[1]", y = "Sensitivity (c)") +
  theme(legend.position = "none")

p_joint_cbias <- ggplot(demo_draws, aes(x = bias, y = c, group = agent_id)) +
  stat_ellipse(geom = "polygon", aes(fill = agent_id),
               level = 0.50, alpha = 0.15, color = NA) +
  stat_ellipse(geom = "path", aes(color = agent_id),
               level = 0.50, alpha = 0.8, linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = 0.5, linetype = "dotted", color = "grey50") +
  scale_x_continuous(limits = c(0, 1)) +
  labs(title    = "Joint Posterior: Bias vs c",
       subtitle = "Distinct clusters motivate the need for a population model.",
       x = "Response Bias", y = "Sensitivity (c)") +
  guides(color = guide_legend(title = "Agent"),
         fill  = guide_legend(title = "Agent")) +
  theme(legend.position = "right")

print(p_joint_cw | p_joint_cbias)
```

## Mathematical Formulation of the Multilevel GCM

**Population-level distributions:**

$$\log c_j \sim \text{Normal}(\mu_{\log c}, \sigma_{\log c})$$
$$w_j \sim \text{Dirichlet}(\kappa \cdot \omega)$$
$$\text{logit}(\beta_j) \sim \text{Normal}(\mu_{\text{logit}\,\beta}, \sigma_{\text{logit}\,\beta})$$

**A note on NCP and the reverse funnel.** Non-centred parameterisation is the safer default for cognitive experiments with ~100 trials per participant. But NCP can develop a *reverse funnel* when the data are very informative for individual parameters. The multilevel SBC at the end of this chapter is the diagnostic that catches this.

### Simulating Data for the Multilevel GCM

```{r ch12_simulate_hierarchical_weights}
simulate_hierarchical_dirichlet <- function(pop_weights, kappa, n_agents) {
  n_features   <- length(pop_weights)
  alpha_params <- pmax(1e-6, kappa * pop_weights)
  ind_weights  <- MCMCpack::rdirichlet(n_agents, alpha_params)

  as_tibble(ind_weights) |>
    setNames(paste0("w", seq_len(n_features))) |>
    mutate(agent_id = seq_len(n_agents), kappa_true = kappa) |>
    pivot_longer(starts_with("w"), names_to = "weight_dim",
                 values_to = "weight_value", names_prefix = "w") |>
    mutate(weight_dim = as.integer(weight_dim))
}

bind_rows(
  simulate_hierarchical_dirichlet(c(0.5, 0.5), kappa = 50, n_agents = 20) |>
    mutate(kappa_level = "High (50)", pop_mean = "Equal (0.5, 0.5)"),
  simulate_hierarchical_dirichlet(c(0.8, 0.2), kappa = 50, n_agents = 20) |>
    mutate(kappa_level = "High (50)", pop_mean = "Skewed (0.8, 0.2)"),
  simulate_hierarchical_dirichlet(c(0.5, 0.5), kappa =  5, n_agents = 20) |>
    mutate(kappa_level = "Low (5)",  pop_mean = "Equal (0.5, 0.5)"),
  simulate_hierarchical_dirichlet(c(0.8, 0.2), kappa =  5, n_agents = 20) |>
    mutate(kappa_level = "Low (5)",  pop_mean = "Skewed (0.8, 0.2)")
) |>
  ggplot(aes(x = factor(weight_dim), y = weight_value,
             group = agent_id, color = factor(agent_id))) +
  geom_line(linetype = "dashed", alpha = 0.5) +
  geom_point(alpha = 0.8) +
  facet_grid(kappa_level ~ pop_mean) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Simulated Individual Attention Weights: Effect of Kappa",
       x = "Dimension", y = "Attention Weight") +
  theme(legend.position = "none")
```

```{r ch12_simulate_multilevel_data}
simulate_multilevel_gcm <- function(n_subjects, stimulus_info, n_blocks,
                                    base_seed,
                                    pop_w, kappa,
                                    pop_log_c_mean, pop_log_c_sd,
                                    pop_logit_bias_mean, pop_logit_bias_sd) {
  n_features   <- length(pop_w)
  alpha_params <- pmax(1e-9, kappa * pop_w)

  ind_w          <- MCMCpack::rdirichlet(n_subjects, alpha_params)
  ind_log_c      <- rnorm(n_subjects, pop_log_c_mean, pop_log_c_sd)
  ind_c          <- exp(ind_log_c)
  ind_logit_bias <- rnorm(n_subjects, pop_logit_bias_mean, pop_logit_bias_sd)
  ind_bias       <- plogis(ind_logit_bias)

  true_params <- tibble(
    agent_id        = seq_len(n_subjects),
    log_c_true      = ind_log_c,
    c_true          = ind_c,
    logit_bias_true = ind_logit_bias,
    bias_true       = ind_bias
  )
  for (f in seq_len(n_features))
    true_params[[paste0("w", f, "_true")]] <- ind_w[, f]

  all_agent_data <- vector("list", n_subjects)
  for (j in seq_len(n_subjects)) {
    subject_seed <- base_seed * 1000 + j
    all_agent_data[[j]] <- simulate_gcm_agent(
      agent_id      = j,
      w_true        = as.numeric(ind_w[j, ]),
      c_true        = ind_c[j],
      bias_true     = ind_bias[j],
      stimulus_info = stimulus_info,
      n_blocks      = n_blocks,
      subject_seed  = subject_seed
    )
  }

  list(data = bind_rows(all_agent_data), true_params = true_params)
}

n_subjects_sim <- 20
pop_params_sim <- list(
  pop_w               = c(0.8, 0.2),
  kappa               = 10,
  pop_log_c_mean      = log(1.5),
  pop_log_c_sd        = 0.5,
  pop_logit_bias_mean = 0,
  pop_logit_bias_sd   = 0.3
)

sim_file_ml        <- here("simdata", "ch12_gcm_multilevel_sim_data.csv")
true_params_file_ml <- here("simdata", "ch12_gcm_multilevel_true_params.csv")

if (regenerate_simulations || !file.exists(sim_file_ml) || !file.exists(true_params_file_ml)) {
  multilevel_sim_output <- simulate_multilevel_gcm(
    n_subjects          = n_subjects_sim,
    stimulus_info       = stimulus_info,
    n_blocks            = n_blocks,
    base_seed           = 42,
    pop_w               = pop_params_sim$pop_w,
    kappa               = pop_params_sim$kappa,
    pop_log_c_mean      = pop_params_sim$pop_log_c_mean,
    pop_log_c_sd        = pop_params_sim$pop_log_c_sd,
    pop_logit_bias_mean = pop_params_sim$pop_logit_bias_mean,
    pop_logit_bias_sd   = pop_params_sim$pop_logit_bias_sd
  )
  multilevel_sim_data    <- multilevel_sim_output$data
  multilevel_true_params <- multilevel_sim_output$true_params
  write_csv(multilevel_sim_data,    sim_file_ml)
  write_csv(multilevel_true_params, true_params_file_ml)
  cat("Multilevel simulation complete. Results saved.\n")
} else {
  multilevel_sim_data    <- read_csv(sim_file_ml,         show_col_types = FALSE)
  multilevel_true_params <- read_csv(true_params_file_ml, show_col_types = FALSE)
  cat("Loaded existing multilevel simulation results.\n")
}

ggplot(multilevel_sim_data,
       aes(x = trial_within_subject, y = performance, group = agent_id)) +
  geom_line(alpha = 0.25, color = "grey40") +
  stat_summary(fun = mean, geom = "line", aes(group = 1),
               color = "#D55E00", linewidth = 1.5) +
  labs(
    title    = "Simulated Learning Curves: Multilevel GCM",
    subtitle = "Grey lines: individual agents; orange: group mean",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  ylim(0, 1)
```

### Implementing the Multilevel GCM in Stan

```{r ch12_multilevel_gcm_stan}
gcm_ml_stan <- "
// Generalized Context Model — Multilevel (refactored architecture)
//
// New relative to the single-subject model:
//   1. log_c parameterization with logit-normal hierarchy for bias.
//   2. Subject-indexed loop: outer over subjects, inner over each subject's
//      contiguous slice [subj_start[j]:subj_end[j]].
//   3. prob_cat1[i] in transformed parameters; model{} and generated quantities
//      both read from it without re-running the loop.
//   4. ALL prior hyperparameters passed through the data block.

data {
  int<lower=1> N_total;
  int<lower=1> N_subjects;
  int<lower=1> N_features;
  int<lower=1> max_trials_per_subject;

  array[N_subjects] int<lower=1, upper=N_total> subj_start;
  array[N_subjects] int<lower=1, upper=N_total> subj_end;

  array[N_total] int<lower=0, upper=1> y;
  array[N_total, N_features] real obs;
  array[N_total] int<lower=0, upper=1> cat_feedback;

  vector[N_features] pop_w_prior_alpha;
  real pop_log_c_mean_prior_mean;
  real<lower=0> pop_log_c_mean_prior_sd;
  real<lower=0> pop_log_c_sd_prior_rate;
  real<lower=0> kappa_prior_rate;
  real pop_logit_bias_mean_prior_mean;
  real<lower=0> pop_logit_bias_mean_prior_sd;
  real<lower=0> pop_logit_bias_sd_prior_rate;
}

parameters {
  simplex[N_features] pop_w;
  real<lower=0> kappa;
  real pop_log_c_mean;
  real<lower=0> pop_log_c_sd;
  real pop_logit_bias_mean;
  real<lower=0> pop_logit_bias_sd;

  vector[N_subjects] z_log_c;
  vector[N_subjects] z_logit_bias;

  array[N_subjects] simplex[N_features] subj_w;
}

transformed parameters {
  vector[N_subjects] subj_log_c;
  vector<lower=0>[N_subjects] subj_c;
  vector<lower=0, upper=1>[N_subjects] subj_bias;

  vector<lower=1e-9, upper=1-1e-9>[N_total] prob_cat1;

  for (j in 1:N_subjects) {
    subj_log_c[j] = pop_log_c_mean + z_log_c[j] * pop_log_c_sd;
    subj_c[j]     = exp(subj_log_c[j]);
    subj_bias[j]  = inv_logit(pop_logit_bias_mean + z_logit_bias[j] * pop_logit_bias_sd);
  }

  {
    array[N_subjects, max_trials_per_subject, N_features] real memory_obs;
    array[N_subjects, max_trials_per_subject] int memory_cat;
    array[N_subjects] int n_mem_subj;

    for (j in 1:N_subjects) n_mem_subj[j] = 0;

    for (j in 1:N_subjects) {
      int n_mem = 0;

      for (i in subj_start[j]:subj_end[j]) {
        real p_i;
        int has_cat0 = 0;
        int has_cat1 = 0;

        for (k in 1:n_mem) {
          if (memory_cat[j, k] == 0) has_cat0 = 1;
          if (memory_cat[j, k] == 1) has_cat1 = 1;
        }

        if (n_mem == 0 || has_cat0 == 0 || has_cat1 == 0) {
          p_i = subj_bias[j];
        } else {
          vector[n_mem] sims;
          for (e in 1:n_mem) {
            real d = 0;
            for (f in 1:N_features)
              d += subj_w[j][f] * abs(obs[i, f] - memory_obs[j, e, f]);
            sims[e] = exp(-subj_c[j] * d);
          }
          real s1 = 0;
          real s0 = 0;
          for (e in 1:n_mem) {
            if (memory_cat[j, e] == 1) s1 += sims[e];
            else                       s0 += sims[e];
          }
          real num = subj_bias[j] * s1;
          real den = num + (1 - subj_bias[j]) * s0;
          p_i = (den > 1e-9) ? num / den : subj_bias[j];
        }

        prob_cat1[i] = fmax(1e-9, fmin(1 - 1e-9, p_i));

        n_mem += 1;
        for (f in 1:N_features) memory_obs[j, n_mem, f] = obs[i, f];
        memory_cat[j, n_mem] = cat_feedback[i];
      }
    }
  }
}

model {
  target += dirichlet_lpdf(pop_w           | pop_w_prior_alpha);
  target += exponential_lpdf(kappa         | kappa_prior_rate);
  target += normal_lpdf(pop_log_c_mean     | pop_log_c_mean_prior_mean,
                                              pop_log_c_mean_prior_sd);
  target += exponential_lpdf(pop_log_c_sd  | pop_log_c_sd_prior_rate);
  target += normal_lpdf(pop_logit_bias_mean | pop_logit_bias_mean_prior_mean,
                                              pop_logit_bias_mean_prior_sd);
  target += exponential_lpdf(pop_logit_bias_sd | pop_logit_bias_sd_prior_rate);

  target += std_normal_lpdf(z_log_c);
  target += std_normal_lpdf(z_logit_bias);

  for (j in 1:N_subjects) {
    vector[N_features] alpha = kappa * pop_w;
    for (f in 1:N_features) alpha[f] = fmax(1e-9, alpha[f]);
    target += dirichlet_lpdf(subj_w[j] | alpha);
  }

  target += bernoulli_lpmf(y | prob_cat1);
}

generated quantities {
  vector[N_total] log_lik;
  for (i in 1:N_total)
    log_lik[i] = bernoulli_lpmf(y[i] | prob_cat1[i]);
}
"

stan_file_gcm_ml <- "stan/ch12_gcm_ml.stan"
write_stan_file(gcm_ml_stan, dir = "stan/", basename = "ch12_gcm_ml.stan")
mod_gcm_ml <- cmdstan_model(stan_file_gcm_ml)
```

### Fitting the Multilevel GCM

```{r ch12_fit_gcm_ml}
gcm_ml_data_sorted <- multilevel_sim_data |>
  mutate(subj_id_stan = as.integer(factor(agent_id))) |>
  arrange(subj_id_stan, trial_within_subject)

subj_bounds <- gcm_ml_data_sorted |>
  mutate(row_idx = row_number()) |>
  group_by(subj_id_stan) |>
  summarise(
    subj_start = min(row_idx),
    subj_end   = max(row_idx),
    .groups    = "drop"
  ) |>
  arrange(subj_id_stan)

n_subjects_in_data  <- max(gcm_ml_data_sorted$subj_id_stan)
max_trials_per_subj <- max(subj_bounds$subj_end - subj_bounds$subj_start + 1)

# ── Prior design note on sensitivity c ───────────────────────────────────────
# Single-agent SBC identified gradient collapse for c > ~7 (log_c > ~2).
# We carry this constraint into the multilevel model via a variance-budget
# argument: the marginal prior on individual log_c_j integrates over both the
# population-mean prior and the population-SD prior.
#
#   Var[log_c_j] ≈ pop_log_c_mean_prior_sd^2  +  E[pop_log_c_sd^2]
#
# For pop_log_c_sd ~ Exponential(λ), E[X^2] = 2/λ^2.
# Choosing pop_log_c_mean_prior_sd = 0.5 and λ = 5 gives:
#   marginal SD = sqrt(0.25 + 0.08) ≈ 0.574
#   +3 SD upper bound on c: exp(log(2) + 3 × 0.574) ≈ 11.2  [< 12 ✓]
# This is tighter than the single-agent prior (SD = 1.0; +3 SD ≈ 20) but
# consistent with the constraint established by the SBC analysis above.
gcm_ml_data <- list(
  N_total                         = nrow(gcm_ml_data_sorted),
  N_subjects                      = n_subjects_in_data,
  N_features                      = 2,
  max_trials_per_subject          = max_trials_per_subj,
  subj_start                      = subj_bounds$subj_start,
  subj_end                        = subj_bounds$subj_end,
  y                               = gcm_ml_data_sorted$sim_response,
  obs                             = as.matrix(gcm_ml_data_sorted[, c("height", "position")]),
  cat_feedback                    = gcm_ml_data_sorted$category_feedback,
  pop_w_prior_alpha               = c(1, 1),
  pop_log_c_mean_prior_mean       = LOG_C_PRIOR_MEAN,
  pop_log_c_mean_prior_sd         = 0.5,   # variance budget: see note above
  pop_log_c_sd_prior_rate         = 5.0,   # variance budget: see note above
  kappa_prior_rate                = 0.1,
  pop_logit_bias_mean_prior_mean  = 0,
  pop_logit_bias_mean_prior_sd    = 1.5,
  pop_logit_bias_sd_prior_rate    = 2.0
)

fit_filepath_ml <- here("simmodels", "ch12_gcm_ml_fit.rds")

if (regenerate_simulations || !file.exists(fit_filepath_ml)) {
  pf_ml <- tryCatch(
    mod_gcm_ml$pathfinder(
      data      = gcm_ml_data,
      seed      = 42,
      num_paths = 4,
      refresh   = 0
    ),
    error = function(e) {
      message("Pathfinder failed — falling back to default initialisation.")
      NULL
    }
  )

  fit_gcm_ml <- mod_gcm_ml$sample(
    data            = gcm_ml_data,
    init            = pf_ml,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1500,
    iter_sampling   = 1500,
    refresh         = 200,
    adapt_delta     = 0.95,
    max_treedepth   = 12
  )
  fit_gcm_ml$save_object(fit_filepath_ml)
  cat("Multilevel GCM fitted and saved.\n")
} else {
  fit_gcm_ml <- readRDS(fit_filepath_ml)
  cat("Loaded existing multilevel GCM fit.\n")
}
```

### Mandatory MCMC Diagnostic Battery (Multilevel)

```{r ch12_multilevel_diagnostic_battery}
if (!is.null(fit_gcm_ml)) {
  pop_vars <- c("pop_w", "kappa",
                "pop_log_c_mean", "pop_log_c_sd",
                "pop_logit_bias_mean", "pop_logit_bias_sd")

  diag_tbl_ml <- diagnostic_summary_table(fit_gcm_ml, params = pop_vars)
  print(diag_tbl_ml)

  if (!all(diag_tbl_ml$pass)) {
    warning("Multilevel diagnostic battery has failures — likely a geometry issue. ",
            "Inspect mcmc_pairs() before interpreting posteriors.")
  }

  bayesplot::mcmc_pairs(
    fit_gcm_ml$draws(c("pop_log_c_mean", "pop_log_c_sd",
                       "pop_logit_bias_mean", "pop_logit_bias_sd")),
    diag_fun     = "dens",
    off_diag_fun = "hex"
  )

  pop_summary <- fit_gcm_ml$summary(variables = pop_vars)
  print("Population Parameter Summary:")
  print(pop_summary)

  cat("\nTrue population parameters used for simulation:\n")
  print(pop_params_sim)
}
```

```{r ch12_multilevel_posterior_visualisation}
if (!is.null(fit_gcm_ml)) {
  draws_ml <- as_draws_df(fit_gcm_ml$draws(variables = c(
    "pop_w", "kappa", "pop_log_c_mean", "pop_log_c_sd",
    "pop_logit_bias_mean", "pop_logit_bias_sd"
  )))

  n_prior <- nrow(draws_ml)
  prior_ml <- tibble(
    `pop_w[1]`          = runif(n_prior),
    `pop_w[2]`          = 1 - `pop_w[1]`,
    kappa               = rexp(n_prior, 0.1),
    pop_log_c_mean      = rnorm(n_prior, LOG_C_PRIOR_MEAN, 0.5),
    pop_log_c_sd        = rexp(n_prior, 5.0),
    pop_logit_bias_mean = rnorm(n_prior, 0, 1.5),
    pop_logit_bias_sd   = rexp(n_prior, 2.0)
  )

  plot_pop_update <- function(param_name, prior_vec, post_vec, true_val, fill_color) {
    df <- tibble(
      value        = c(prior_vec, post_vec),
      Distribution = rep(c("Prior", "Posterior"), each = length(prior_vec))
    )
    ggplot(df, aes(x = value, fill = Distribution, color = Distribution)) +
      geom_density(alpha = 0.5, linewidth = 0.5) +
      geom_vline(xintercept = true_val, color = "red", linetype = "dashed", linewidth = 1) +
      scale_fill_manual(values  = c("Prior" = "grey70", "Posterior" = fill_color)) +
      scale_color_manual(values = c("Prior" = "grey50", "Posterior" = fill_color)) +
      labs(title = param_name, x = NULL, y = "Density") +
      theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
  }

  p_pop_w1 <- plot_pop_update("Pop w[1]", prior_ml$`pop_w[1]`, draws_ml$`pop_w[1]`,
                              pop_params_sim$pop_w[1], "#56B4E9")
  p_pop_w2 <- plot_pop_update("Pop w[2]", prior_ml$`pop_w[2]`, draws_ml$`pop_w[2]`,
                              pop_params_sim$pop_w[2], "#56B4E9")
  p_kappa  <- plot_pop_update("Kappa",    prior_ml$kappa,      draws_ml$kappa,
                              pop_params_sim$kappa, "#E69F00")
  p_logcm  <- plot_pop_update("Pop log(c) mean", prior_ml$pop_log_c_mean,
                              draws_ml$pop_log_c_mean, pop_params_sim$pop_log_c_mean, "#009E73")
  p_logcsd <- plot_pop_update("Pop log(c) SD", prior_ml$pop_log_c_sd,
                              draws_ml$pop_log_c_sd, pop_params_sim$pop_log_c_sd, "#009E73")
  p_biasm  <- plot_pop_update("Pop logit(bias) mean", prior_ml$pop_logit_bias_mean,
                              draws_ml$pop_logit_bias_mean,
                              pop_params_sim$pop_logit_bias_mean, "#CC79A7")
  p_biassd <- plot_pop_update("Pop logit(bias) SD", prior_ml$pop_logit_bias_sd,
                              draws_ml$pop_logit_bias_sd,
                              pop_params_sim$pop_logit_bias_sd, "#CC79A7")

  layout_grid <- (p_pop_w1 | p_pop_w2 | p_kappa) /
                 (p_logcm  | p_logcsd | plot_spacer()) /
                 (p_biasm  | p_biassd | plot_spacer()) +
                 plot_layout(guides = "collect") & theme(legend.position = "bottom")

  cat("=== Population Parameter Prior-Posterior Updates ===\n")
  print(layout_grid)

  p_pairs <- bayesplot::mcmc_pairs(
    fit_gcm_ml$draws(c("pop_w[1]", "kappa", "pop_log_c_mean",
                       "pop_log_c_sd", "pop_logit_bias_sd")),
    diag_fun     = "dens",
    off_diag_fun = "hex"
  )
  print(p_pairs)

  subj_c_summary <- fit_gcm_ml$summary(variables = "subj_c") |>
    mutate(agent_id = readr::parse_number(
      stringr::str_extract(variable, "\\[\\d+\\]")
    )) |>
    left_join(multilevel_true_params |> dplyr::select(agent_id, c_true),
              by = "agent_id")

  p_recov_c <- ggplot(subj_c_summary, aes(x = c_true, y = mean)) +
    geom_point(alpha = 0.7) +
    geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
    geom_smooth(method = lm, color = "#0072B2", linewidth = 0.7, se = TRUE) +
    labs(title = "Individual Recovery: Sensitivity c",
         x = "True c", y = "Estimated c (posterior mean)") +
    coord_cartesian(xlim = c(0, 8), ylim = c(0, 8))

  subj_w_summary <- fit_gcm_ml$summary(variables = "subj_w") |>
    mutate(
      agent_id = readr::parse_number(stringr::str_extract(variable, "\\[\\d+")),
      feature  = readr::parse_number(stringr::str_extract(variable, ",\\d+"))
    )

  true_w_long <- multilevel_true_params |>
    dplyr::select(agent_id, w1_true, w2_true) |>
    pivot_longer(c(w1_true, w2_true),
                 names_to = "feature", values_to = "w_true",
                 names_pattern = "w(\\d+)_true") |>
    mutate(feature = as.integer(feature))

  recovery_w <- true_w_long |>
    left_join(subj_w_summary |> dplyr::select(agent_id, feature, mean),
              by = c("agent_id", "feature"))

  p_recov_w <- ggplot(recovery_w, aes(x = w_true, y = mean, color = factor(feature))) +
    geom_point(alpha = 0.7) +
    geom_abline(intercept = 0, slope = 1, color = "grey30", linetype = "dashed") +
    scale_color_manual(values = c("1" = "#0072B2", "2" = "#D55E00"),
                       name = "Dimension") +
    labs(title = "Individual Recovery: Attention Weights",
         x = "True w", y = "Estimated w (posterior mean)") +
    coord_fixed(xlim = c(0, 1), ylim = c(0, 1))

  cat("\n=== Individual Parameter Recovery ===\n")
  print(p_recov_c | p_recov_w)
}
```

---

## Extensive Parameter Recovery for the Multilevel GCM

```{r ch12_extensive_recovery, eval=run_intensive_checks}
param_grid_ml <- expand_grid(
  pop_log_c_mean = c(log(1.0), log(1.5), log(2.5)),
  pop_log_c_sd   = c(0.1, 0.3),   # within constrained prior (Exp(5), 39th–78th pctile)
  pop_w1         = c(0.2, 0.35, 0.5),
  kappa          = c(5, 20)
) |>
  mutate(
    pop_w   = map(pop_w1, ~c(.x, 1 - .x)),
    grid_id = row_number()
  )

n_sims_per_grid  <- 3
n_subjects_recov <- 15

run_recovery_iteration <- function(grid_id, pop_log_c_mean, pop_log_c_sd,
                                   pop_w, kappa, sim_iteration) {
  cat("Grid", grid_id, "Iter", sim_iteration, "\n")
  mod_gcm_ml <- cmdstanr::cmdstan_model("stan/ch12_gcm_ml.stan")
  sim_out <- tryCatch(
    simulate_multilevel_gcm(
      n_subjects          = n_subjects_recov,
      stimulus_info       = stimulus_info,
      n_blocks            = n_blocks,
      base_seed           = grid_id * 100 + sim_iteration,
      pop_w               = pop_w,
      kappa               = kappa,
      pop_log_c_mean      = pop_log_c_mean,
      pop_log_c_sd        = pop_log_c_sd,
      pop_logit_bias_mean = 0,
      pop_logit_bias_sd   = 0.3
    ),
    error = function(e) { cat("Simulation error:", e$message, "\n"); NULL }
  )
  if (is.null(sim_out)) return(NULL)

  sim_dat  <- sim_out$data
  sim_true <- sim_out$true_params

  sorted_dat <- sim_dat |>
    mutate(subj_id_stan = as.integer(factor(agent_id))) |>
    arrange(subj_id_stan, trial_within_subject)

  bounds <- sorted_dat |>
    mutate(row_idx = row_number()) |>
    group_by(subj_id_stan) |>
    summarise(subj_start = min(row_idx), subj_end = max(row_idx), .groups = "drop") |>
    arrange(subj_id_stan)

  stan_dat <- list(
    N_total                         = nrow(sorted_dat),
    N_subjects                      = n_subjects_recov,
    N_features                      = 2,
    max_trials_per_subject          = max(bounds$subj_end - bounds$subj_start + 1),
    subj_start                      = bounds$subj_start,
    subj_end                        = bounds$subj_end,
    y                               = sorted_dat$sim_response,
    obs                             = as.matrix(sorted_dat[, c("height", "position")]),
    cat_feedback                    = sorted_dat$category_feedback,
    pop_w_prior_alpha               = c(1, 1),
    pop_log_c_mean_prior_mean       = LOG_C_PRIOR_MEAN,
    pop_log_c_mean_prior_sd         = 0.5,
    pop_log_c_sd_prior_rate         = 5.0,
    kappa_prior_rate                = 0.1,
    pop_logit_bias_mean_prior_mean  = 0,
    pop_logit_bias_mean_prior_sd    = 1.5,
    pop_logit_bias_sd_prior_rate    = 2.0
  )

  pf_r <- tryCatch(
    mod_gcm_ml$pathfinder(data = stan_dat, num_paths = 4, refresh = 0,
                           seed = grid_id * 100 + sim_iteration),
    error = function(e) NULL
  )
  init_r <- if (!is.null(pf_r)) pf_r else 0.5

  fit_r <- tryCatch(
    mod_gcm_ml$sample(
      data = stan_dat, init = init_r,
      seed = grid_id * 100 + sim_iteration,
      chains = 2, parallel_chains = 2,
      iter_warmup = 800, iter_sampling = 1000,
      refresh = 0, adapt_delta = 0.9, max_treedepth = 10
    ),
    error = function(e) { cat("Fit error:", e$message, "\n"); NULL }
  )
  if (is.null(fit_r)) return(NULL)

  diag <- fit_r$diagnostic_summary(quiet = TRUE)
  pop_est <- fit_r$summary(
    variables = c("pop_w", "kappa", "pop_log_c_mean", "pop_log_c_sd"),
    "mean"
  ) |>
    mutate(
      true_value = case_when(
        variable == "pop_w[1]"       ~ pop_w[1],
        variable == "pop_w[2]"       ~ pop_w[2],
        variable == "kappa"          ~ kappa,
        variable == "pop_log_c_mean" ~ pop_log_c_mean,
        variable == "pop_log_c_sd"   ~ pop_log_c_sd,
        TRUE                         ~ NA_real_
      ),
      grid_id       = grid_id,
      sim_iteration = sim_iteration,
      had_divergence = any(diag$num_divergent > 0)
    )

  pop_est
}

recovery_plan <- param_grid_ml |>
  crossing(sim_iteration = seq_len(n_sims_per_grid))

recovery_file <- here("simdata", "ch12_gcm_ml_recovery_raw.csv")

if (regenerate_simulations || !file.exists(recovery_file)) {
  plan(multisession, workers = max(1, availableCores() %/% 2))
  recovery_results_raw <- future_pmap_dfr(
    list(
      grid_id        = recovery_plan$grid_id,
      pop_log_c_mean = recovery_plan$pop_log_c_mean,
      pop_log_c_sd   = recovery_plan$pop_log_c_sd,
      pop_w          = recovery_plan$pop_w,
      kappa          = recovery_plan$kappa,
      sim_iteration  = recovery_plan$sim_iteration
    ),
    run_recovery_iteration,
    .options = furrr_options(seed = TRUE, scheduling = 1)
  )
  plan(sequential)
  write_csv(recovery_results_raw, recovery_file)
} else {
  recovery_results_raw <- read_csv(recovery_file, show_col_types = FALSE)
}

recovery_labeled <- recovery_results_raw |>
  left_join(param_grid_ml |> dplyr::select(grid_id, pop_log_c_mean, pop_log_c_sd,
                                            pop_w1, kappa), by = "grid_id")

ggplot(recovery_labeled |> filter(!is.na(true_value)),
       aes(x = true_value, y = mean)) +
  geom_point(aes(color = had_divergence), alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_smooth(method = lm, color = "#0072B2", linewidth = 0.7) +
  scale_color_manual(values = c("FALSE" = "grey30", "TRUE" = "#D55E00"),
                     name = "Had divergence") +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Population Parameter Recovery: Multilevel GCM",
       x = "True value", y = "Estimated (posterior mean)") +
  theme(legend.position = "bottom")
```

---

## Multilevel SBC

```{r ch12_sbc_multilevel, eval=run_intensive_checks}
sbc_ml_filepath <- here("simdata", "ch12_sbc_multilevel_results.rds")

if (regenerate_simulations || !file.exists(sbc_ml_filepath)) {
  n_sbc_iterations_ml <- 200   # use >= 1000 for publication-quality SBC
  n_subjects_sbc      <- 10

  gcm_ml_sbc_generator <- SBC_generator_function(
    function() {
      pop_log_c_mean      <- rnorm(1, LOG_C_PRIOR_MEAN, 0.5)
      pop_log_c_sd        <- rexp(1, 5.0)
      pop_w               <- as.numeric(MCMCpack::rdirichlet(1, c(1, 1)))
      kappa               <- rexp(1, 0.1)
      pop_logit_bias_mean <- rnorm(1, 0, 1.5)
      pop_logit_bias_sd   <- rexp(1, 2.0)

      z_log_c      <- rnorm(n_subjects_sbc, 0, 1)
      z_logit_bias <- rnorm(n_subjects_sbc, 0, 1)
      subj_log_c   <- pop_log_c_mean + pop_log_c_sd * z_log_c
      subj_c       <- exp(subj_log_c)
      subj_bias <- plogis(pop_logit_bias_mean + pop_logit_bias_sd * z_logit_bias)
      
      # Safe Dirichlet sampling to prevent rgamma underflow (0/0 = NaN)
      alpha_w <- pmax(1e-9, kappa * pop_w)
      subj_w  <- MCMCpack::rdirichlet(n_subjects_sbc, alpha_w)
      
      for (s_idx in seq_len(n_subjects_sbc)) {
        if (any(is.na(subj_w[s_idx, ]))) {
          # As alpha approaches 0, Dirichlet concentrates on the boundaries.
          # If R underflows to NaN, we safely sample a strict corner.
          corner <- rep(0, length(pop_w))
          corner[sample(length(pop_w), 1, prob = pop_w)] <- 1
          subj_w[s_idx, ] <- corner
        }
      }

      all_dat <- map_dfr(seq_len(n_subjects_sbc), function(j) {
        sched <- make_subject_schedule(stimulus_info, n_blocks,
                                       seed = sample.int(1e8, 1))
        sim   <- gcm_simulate(
          w            = as.numeric(subj_w[j, ]),
          c            = subj_c[j],
          bias         = subj_bias[j],
          obs          = sched |> dplyr::select(height, position),
          cat_feedback = sched$category_feedback
        )
        sched |>
          mutate(agent_id = j, sim_response = sim$sim_response)
      })

      sorted <- all_dat |>
        mutate(subj_id_stan = as.integer(factor(agent_id))) |>
        arrange(subj_id_stan, trial_within_subject)

      bounds <- sorted |>
        mutate(row_idx = row_number()) |>
        group_by(subj_id_stan) |>
        summarise(subj_start = min(row_idx), subj_end = max(row_idx),
                  .groups = "drop") |>
        arrange(subj_id_stan)

      list(
          variables = list(
            pop_log_c_mean      = pop_log_c_mean,
            pop_log_c_sd        = pop_log_c_sd,
            `pop_w[1]`          = pop_w[1],
            `pop_w[2]`          = pop_w[2],
            kappa               = kappa,
            pop_logit_bias_mean = pop_logit_bias_mean,
            pop_logit_bias_sd   = pop_logit_bias_sd,
            # Explicitly track Subject 1 for individual-level recovery
            `subj_c[1]`         = subj_c[1],
            `subj_w[1,1]`       = subj_w[1, 1]
          ),
        generated = list(
          N_total                         = nrow(sorted),
          N_subjects                      = n_subjects_sbc,
          N_features                      = 2L,
          max_trials_per_subject          = max(bounds$subj_end - bounds$subj_start + 1),
          subj_start                      = bounds$subj_start,
          subj_end                        = bounds$subj_end,
          y                               = as.integer(sorted$sim_response),
          obs                             = as.matrix(sorted[, c("height", "position")]),
          cat_feedback                    = as.integer(sorted$category_feedback),
          pop_w_prior_alpha               = c(1, 1),
          pop_log_c_mean_prior_mean       = LOG_C_PRIOR_MEAN,
          pop_log_c_mean_prior_sd         = 0.5,
          pop_log_c_sd_prior_rate         = 5.0,
          kappa_prior_rate                = 0.1,
          pop_logit_bias_mean_prior_mean  = 0,
          pop_logit_bias_mean_prior_sd    = 1.5,
          pop_logit_bias_sd_prior_rate    = 2.0
        )
      )
    }
  )

  gcm_ml_sbc_backend <- SBC_backend_cmdstan_sample(
    mod_gcm_ml,
    iter_warmup   = 1000,
    iter_sampling = 1000,
    chains        = 2,
    adapt_delta   = 0.95,
    refresh       = 0
  )

  gcm_ml_datasets <- generate_datasets(
    gcm_ml_sbc_generator,
    n_sims = n_sbc_iterations_ml
  )

  sbc_results_ml <- compute_SBC(
    gcm_ml_datasets,
    gcm_ml_sbc_backend,
    cache_mode     = "results",
    cache_location = here("simdata", "ch12_sbc_multilevel_cache"),
    keep_fits      = FALSE
  )
  saveRDS(sbc_results_ml, sbc_ml_filepath)
  cat("SBC multilevel results computed and saved.\n")
} else {
  sbc_results_ml <- readRDS(sbc_ml_filepath)
  cat("Loaded existing SBC multilevel results.\n")
}

plot_ecdf_diff(sbc_results_ml)
plot_rank_hist(sbc_results_ml)

# ── Parameter recovery on Multilevel SBC simulations ───────────────────────
recovery_df_ml <- sbc_results_ml$stats |>
  filter(variable %in% c("pop_log_c_mean", "pop_log_c_sd", "pop_w[1]", 
                         "kappa", "pop_logit_bias_mean", "pop_logit_bias_sd"))

p_sbc_recovery_ml <- ggplot(recovery_df_ml, aes(x = simulated_value, y = mean)) +
  geom_point(alpha = 0.5, color = "#0072B2") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~variable, scales = "free") +
  labs(
    title    = "Population Parameter Recovery Across SBC Simulations",
    subtitle = "Posterior mean vs True generated value",
    x = "True Value (simulated_value)",
    y = "Estimated Value (Posterior Mean)"
  ) +
  theme_bw()

print(p_sbc_recovery_ml)

# ── Individual-Level Parameter Recovery on Multilevel SBC ────────────────────
recovery_df_indiv <- sbc_results_ml$stats |>
  filter(variable %in% c("subj_c[1]", "subj_w[1,1]"))

p_sbc_recovery_indiv <- ggplot(recovery_df_indiv, aes(x = simulated_value, y = mean)) +
  geom_point(alpha = 0.5, color = "#D55E00") +
  geom_abline(intercept = 0, slope = 1, color = "grey30", linetype = "dashed") +
  facet_wrap(~variable, scales = "free") +
  labs(
    title    = "Individual Parameter Recovery (Subject 1) Across SBC",
    subtitle = "Posterior mean vs True generated value",
    x = "True Value (simulated_value)",
    y = "Estimated Value (Posterior Mean)"
  ) +
  theme_bw()

print(p_sbc_recovery_indiv)
```

**How to read the multilevel SBC**. First, look at the ECDF and rank histograms. Pay special attention to the curves for pop_log_c_sd and pop_logit_bias_sd — these are the population scale parameters where the Non-Centered Parameterisation (NCP) can develop a reverse funnel. A U-shaped rank histogram means the posterior is overconfident. If that happens, the standard modeling prescription is to switch the affected parameter from a non-centred to a centred parameterisation. Second, look at the parameter recovery scatter plots. While the rank histograms tell us if our uncertainties (credible intervals) are statistically calibrated, the scatter plots tell us if the model is actually retrieving the correct data-generating parameters on average. Points falling neatly along the red dashed line ($y = x$) confirm that our multilevel architecture successfully recovers the group-level cognitive mechanisms.

---

## GCM in Context: Strengths and Limitations

### Strengths

* **Empirical support**: The GCM has successfully fit human categorization data across numerous experiments.
* **Psychological plausibility**: The model's parameters correspond to psychologically meaningful constructs (attention, sensitivity/generalization, bias).
* **Flexibility**: It can account for a wide range of category structures and learning phenomena.
* **Integration with memory**: The model naturally connects categorization to memory for specific instances.
* **Individual differences**: The multilevel framework allows principled modeling of variations across people.

### Limitations

* **Memory requirements**: Storing all exemplars becomes computationally demanding for large numbers of stimuli. How memory is bounded over long timescales is not fully specified.
* **Conflation of generalisation and decisional precision**: The standard GCM has no separate softmax temperature; sensitivity $c$ doubles as a generalisation parameter and as a decisional precision parameter.
* **Complex categories**: The standard GCM struggles with categories defined by abstract rules, relations, or logical combinations of features.
* **Sequential structure and LOO**: Because the GCM is path-dependent, standard LOO-CV is technically invalid — but this chapter implements LFO-CV (Bürkner et al., 2020) as the canonical sequential-model cross-validation. The implementation is reused by Chapters 12, 13, and 15.

In the next two chapters we explore alternative approaches — prototype models (Ch. 12) and rule-based models (Ch. 13) — that address some of these limitations while introducing their own strengths and challenges.

---
