# The Generalized Context Model {#sec-categorization-exemplars}

```{r ch12_setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "80%",
  dpi        = 300
)

# Set to TRUE to rerun all Stan fits and heavy simulations; FALSE loads saved results.
regenerate_simulations <- FALSE
# Set to TRUE to rerun SBC, extensive recovery, exact LFO validation.
run_intensive_checks   <- regenerate_simulations

# Directory scaffolding
for (d in c("stan", "simdata", "simmodels")) {
  if (!dir.exists(d)) dir.create(d)
}

pacman::p_load(
  tidyverse,    # data manipulation and visualization
  here,         # robust project-relative paths
  future,       # parallel processing setup
  purrr,        # functional programming tools
  furrr,        # parallel functional programming
  patchwork,    # combining plots
  cmdstanr,     # interface to Stan
  posterior,    # tidy posterior arrays
  tidybayes,    # tidy extraction of draws
  bayesplot,    # Bayesian visualization
  loo,          # LOO-CV model comparison
  SBC,          # Simulation-Based Calibration Checks (ECDF-diff plots)
  priorsense,   # prior sensitivity via power-scaling
  MCMCpack      # rdirichlet
)

theme_set(theme_classic())

# Prior hyperparameters for sensitivity c
LOG_C_PRIOR_MEAN <- log(2)   # prior median c = 2
LOG_C_PRIOR_SD   <- 1.0      # covers c ≈ 0.27–14.8 at ±2 SD

# ── Shared diagnostic helper used throughout the chapter ─────────────────────
diagnostic_summary_table <- function(fit, params = NULL) {
  if (is.null(fit)) return(NULL)
  diag <- fit$diagnostic_summary(quiet = TRUE)

  s <- fit$summary(
    variables = params,
    "mean", "sd", "rhat", "ess_bulk", "ess_tail", "mcse_mean"
  )

  tibble(
    metric = c(
      "Divergences (zero tolerance)",
      "Max rank-normalised R-hat",
      "Min bulk ESS",
      "Min tail ESS",
      "Min E-BFMI",
      "Max MCSE / posterior SD"
    ),
    value = c(
      sum(diag$num_divergent),
      max(s$rhat,     na.rm = TRUE),
      min(s$ess_bulk, na.rm = TRUE),
      min(s$ess_tail, na.rm = TRUE),
      min(diag$ebfmi, na.rm = TRUE),
      max(s$mcse_mean / s$sd, na.rm = TRUE)
    ),
    threshold = c("== 0", "< 1.01", "> 400", "> 400", "> 0.2", "< 0.05"),
    pass = c(
      sum(diag$num_divergent)               == 0,
      max(s$rhat, na.rm = TRUE)             <  1.01,
      min(s$ess_bulk, na.rm = TRUE)         >  400,
      min(s$ess_tail, na.rm = TRUE)         >  400,
      min(diag$ebfmi, na.rm = TRUE)         >  0.2,
      max(s$mcse_mean / s$sd, na.rm = TRUE) <  0.05
    )
  )
}
```

> **📍 Where we are in the Bayesian modeling workflow:**
> Chapters 1–10 built the complete inference toolkit: parameter estimation
> (Ch. 4), the six-phase validation pipeline (Ch. 5), hierarchical models
> (Ch. 6), model comparison via LOO (Ch. 7), mixture models with discrete
> marginalization (Ch. 8), and hierarchical Bayesian cognition (Ch. 9–10).
> This and the following 2 chapters apply the full toolkit to *categorization*, 
> asking not only what parameters describe a participant's behavior, but which 
> *cognitive mechanism* best accounts for how people learn to assign objects
> to kinds.
> We begin with the exemplar approach: the Generalized Context Model (GCM).
> The following chapters will introduce prototype and rule-based models,
> and then compare all three approaches head-to-head using the full validation
> battery.
>
> Two important caveats.
> First, the categorization models learn, that is, are **path-dependent** —
> the likelihood at trial $t$ depends on the full sequence of stimuli on
> trials $1, \ldots, t-1$ through the exemplar memory store. Standard LOO-CV
> is therefore technically invalid here. We met this issue when discussing
> sequential Bayes. Here we introduce leave-future-out cross-validation (LFO-CV;
> Bürkner, Gabry, & Vehtari, 2020).
> Second, the rule-based models we will discuss cannot really be fit using
> MCMC (we'll see why), so we need to introduce a new slower method: Approximate
> Bayesian Computation (ABC).
>
> **Trilogy structure.** Chapters 11, 12, and 13 form a parallel trilogy.
> Each chapter builds a base cognitive model, extends it with a single
> non-stationarity parameter ($\lambda$ here, $q$ in Ch. 12, $\mu$ in Ch. 13),
> and then validates across the same three environmental scenarios in the same
> order: static Kruschke (baseline), performance-contingent abrupt shifts
> (misspecified non-stationarity), and smooth drift (assumption-matched
> non-stationarity). The ordering reflects increasing assumption match between
> the scenario's data-generating process and each model's structural
> commitments. Chapter 13 closes with a synthesis of what the three analyses
> jointly imply about identifiability and model-task alignment.

## Introduction to Categorization Models

### The Fundamental Problem of Categorization

Categorization is one of the most fundamental cognitive abilities that humans possess. From early childhood, we learn to organize the world around us into meaningful categories: distinguishing food from non-food, safe from dangerous, or one letter from another. This process of assigning objects to categories is complex, involving perception, memory, attention, and decision-making.

How do humans learn categories and make categorization decisions? This question has generated decades of research and theoretical debate in cognitive science.

### Why Model Categorization?

Computational models let us formalize competing theories, generate precise predictions, and test them against human behavior. This chapter focuses on the **exemplar approach** — the Generalized Context Model (GCM). Chapters 12 and 13 introduce prototype and rule-based models respectively, with a full cross-model comparison in Chapter 13.

### Historical Development of Categorization Models

#### Early Views: The Classical Approach

Early theories of categorization followed what is now called the classical view, where categories were defined by necessary and sufficient features. In this view, category membership is an all-or-nothing affair: an object either satisfies the logical criteria or it doesn't. Some examples from Aristotle: 

* a human is a rational animal. 

* a bird has feathers, wings, and a beak; and lays eggs.

* a fish lives in water; has scales; and breathes through gills.

You need all of these features to be part of one category, and having them makes you part of that category.

While intuitive, this approach struggled to explain many aspects of human categorization, such as:

* Graded category membership (some members seem "more typical" than others)

* Unclear boundaries between categories.

* Context-dependent categorization

* Family resemblance structures (where no single feature is necessary)

Think of the traditional definition of a species: a set of organisms that can interbreed and produce fertile offspring. However, there are many cases of organism A being able to interbreed with B, and B with C, but A and C cannot interbreed (for instance Californian Ensatina Salamanders have such phenomena). This creates a fuzzy boundary that the classical view cannot easily accommodate.

#### The Prototype Revolution

In the 1970s, Eleanor Rosch's pioneering work on prototypes challenged the classical view. She demonstrated that categories appear to be organized around central tendencies or "prototypes," with membership determined by similarity to these prototypes. Objects closer to the prototype are categorized more quickly and consistently.

#### The Exemplar Alternative

In the late 1970s and 1980s, researchers like Douglas Medin and Robert Nosofsky proposed that rather than abstracting prototypes, people might store individual exemplars of categories and make judgments based on similarity to these stored examples.

The Generalized Context Model (GCM), developed by Nosofsky, became the standard exemplar model. It proposes that categorization decisions are based on the summed similarity of a new stimulus to all stored exemplars of each category, weighted by attention to different stimulus dimensions.

#### Rule-Based Models

While similarity-based approaches gained prominence, other researchers argued that people sometimes use explicit rules for categorization (e.g. in scientific reasoning, but not only). Models like Bayesian particle filters, multinomial trees, COVIS and RULEX formalize these ideas, suggesting that rule learning and application form a core part of human categorization, particularly for well-defined categories.

#### Hybrid Approaches

More recent work has focused on hybrid models that incorporate elements of multiple approaches. Models like SUSTAIN [@love2004sustain] and ATRIUM [@erickson1998rules] propose that humans can flexibly switch between strategies or that different systems operate in parallel.

### A Systematization Framework for Categorization Models

To pedagogically navigate and compare the different approaches, we can deconstruct the cognitive process of categorization into a sequential six-step pipeline:

1. **Input Representation:** How are stimuli represented? (e.g., continuous geometry vs. discrete binary features).

2. **Attentional Processes:** How are features prioritized? Not all features are equally relevant given a task, and attention is often modeled as weights summing to 1.

3. **Intermediate Representations:** How is the category internally stored? (e.g., as raw exemplars, abstract prototypes, or explicit rules).

4. **Evidential Mechanisms:** How is the incoming stimulus compared to the representation to gather evidence? (e.g., calculating average similarity with an exponential decay over distance, or evidence accumulation).

5. **Decision Rules:** How is evidence translated into a choice? (e.g., maximum similarity, or a probabilistic Luce choice axiom).

6. **Learning:** How do representations, attentional weights, or rule structures update over time through exposure, memory accumulation, and feedback?


### Our Implementation Approach

Each chapter in the trilogy follows the same validation structure: forward simulator → Stan model → diagnostic battery → prior/posterior predictive checks → parameter recovery → prior sensitivity → SBC. This parallel structure makes it possible to isolate what differs between the cognitive mechanisms rather than conflating model differences with implementation choices.

::: {.callout-note title="The Conceptual Need for Shared Implementations"}

A major challenge in evaluating cognitive theories is the "hidden implementation" problem. If different research groups write their own custom code for a GCM or a prototype model, subtle differences in how the math is translated into an algorithm can lead to conflicting theoretical conclusions. To genuinely compare models, the field conceptually requires shared, open-source implementations tested against standardized empirical benchmarks. 

A pioneering example of this movement is the work by @wills2017progress, who developed the `catlearn` R package to provide formal, collaboratively verified implementations of classic categorization models and canonical datasets.

These course notes are designed as a partial, pedagogical contribution to this open-science ethos. We do not adopt the specific software architecture or strict standardization of packages like `catlearn`, but we share the same fundamental goal: transparent, reproducible science. We show how to create, document and validate transparent bespoke scripts and Stan-based Bayesian inference. This means that you will not just adopt some secretly shared pre-made script, or struggle implementing an equation and then never share the script as you are secretely afraid it contains errors. Rather, you'll be able to implement and develop different algorithms, validate and document them, and share them with the world, so that others will be able to build on your work and we can do better science as a collective enterprise.

:::

### Categorization (Cognitive Science) vs. Classification (Machine Learning)

Because the mathematical tools are often identical, it is tempting to view cognitive models of categorization simply as early, inefficient machine learning classifiers. Both fields rely on the exact same underlying mechanics to map inputs to outputs: calculating distances in multi-dimensional feature space, applying functions (like Luce or softmax) to translate evidence into probabilities, and using gradient-based weight updates or Bayesian filters to learn from feedback. 

However, despite this shared mathematical framework, they serve fundamentally different goals:

* **Classification (Machine Learning):** The primary goal is *engineering*. It focuses on drawing the optimal decision boundary to maximize predictive accuracy and minimize loss on a given dataset. The internal mechanism (e.g., a deep neural network's hidden layers) doesn't need to resemble human thought—it just needs to work efficiently and accurately.

* **Categorization (Cognitive Modeling):** The primary goal is *reverse-engineering*. Categorization is about understanding the knowledge structures that enable a biological agent to make predictive inferences. We ask not only what parameters describe behavior, but which *cognitive mechanism* best accounts for how people actually learn, including the specific mistakes they make.

We can summarize this distinction along three key dimensions:

| Dimension | Classification (Machine Learning) | Categorization (Cognitive Modeling) |
| :--- | :--- | :--- |
| **Primary Goal** | Maximize out-of-sample accuracy. | Explain human behavior, including systematic human errors. |
| **Data Processing** | Often batch-processed over massive, static datasets. | Inherently **path-dependent**; learns sequentially trial-by-trial. |
| **Constraints** | Hardware limits and computational efficiency. | Psychological plausibility (bounded memory, selective attention, cognitive load). |

In short: A machine learning classifier asks, *"What is the best mathematical way to separate Category A from Category B?"* A cognitive model asks, *"How do humans separate A from B, and what are the internal mental representations that allow them to do so?"*

---

## The Generalized Context Model (GCM)

### Mathematical Foundations and Cognitive Principles

The Generalized Context Model (GCM), developed by Robert Nosofsky [@nosofsky1986attention] in the 1980s, represents one of the most influential exemplar-based approaches to categorization. Unlike models that abstract category information into prototypes or rules, the GCM proposes that people store individual exemplars in memory and make categorization decisions based on similarity to these stored examples.

#### The GCM Mapped to the Categorization Pipeline

The GCM translates the 6-step cognitive pipeline into a formal mathematical structure as follows:

* **1. Input Representation (Geometric Space):** Stimuli are represented as continuous points in a multi-dimensional psychological space.

* **2. Attentional Processes (Selective Attention):** Per-dimension weights $w$ (summing to 1) rescale the psychological space before distance is computed, prioritizing task-relevant features.

* **3. Intermediate Representations (Memory for Instances):** The model explicitly skips abstraction. Every encountered example is stored directly in memory with its category label.

* **4. Evidential Mechanisms (Similarity Computation):** Distance is calculated using a city-block metric (Manhattan distance). Similarity between a new stimulus and each stored exemplar is an exponentially decreasing function of their distance, governed by a sensitivity parameter $c$. Evidence is then aggregated as average similarity for each category.

* **5. Decision Rules (Choice Probability):** The decision relies on the Luce choice axiom. The probability of assigning a stimulus to a category is proportional to its summed similarity to exemplars of that category, modified by a baseline category bias $\beta$.
* **6. Learning (Memory Accumulation):** The model learns sequentially by adding each newly categorized and feedback-evaluated stimulus directly into its exemplar memory store, altering the evidential landscape for future trials.

::: {.callout-note}

### Historical Context: From Binary Logic to Psychological Space

The evolution of the Generalized Context Model (GCM) reflects a broader shift in cognitive science: moving from models constrained by computational simplicity toward those more capable of capturing the "messy" richness of human perception.

#### The Era of Binary Constraints
In the late 1970s, categorization research was dominated by the "small-set" paradigm. Stimuli were often constructed from a few discrete, binary features (e.g., "large vs. small," "red vs. blue"). The pioneering **Context Model** [@medin1978context] was designed for these environments, calculating similarity based on simple feature overlap. While revolutionary, it was computationally and theoretically "flat": it couldn't account for the fact that some dimensions are more perceptually salient than others, or that the psychological distance between features isn't always linear.

#### The Geometric Revolution

The introduction of the GCM [@nosofsky1986attention] broke these constraints by synthesizing exemplar theory with **Multidimensional Scaling (MDS)**. Instead of binary features, stimuli became points in a continuous, multi-dimensional psychological space. The core innovation was the weighted Minkowski metric:
$$d_{ij} = c \left[ \sum w_k |x_{ik} - x_{jk}|^r \right]^{1/r}$$
This formula allowed the model to "stretch" or "shrink" psychological space based on selective attention ($w_k$). It transformed categorization from a logical matching process into a geometric one, where the "geometry of thought" could be empirically mapped.

#### The "Ecological" Turn: Summed vs. Mean Similarity

As the field moved toward experiments with larger and unbalanced category sizes, a debate arose over whether evidence should be summed or averaged across exemplars. See §Summed vs. Mean Similarity below for the full argument and a concrete numerical demonstration; that section motivates why our implementation adopts the **mean-similarity** formulation.

#### The Process Revolution: Sequential Accumulation and Path-Dependency

Early models were "static": they were used to fit terminal, steady-state data from the end of an experiment. The conceptual shift toward process models began by acknowledging that categorization is a sequential learning experience. The critical first step is simply modeling memory as it grows: exemplars are accumulated into the memory store one trial at a time. 

This transition highlights the fundamentally **path-dependent** nature of cognition: your response at trial $t$ depends strictly on the specific sequence of exemplars and feedback you encountered in trials $1 \dots t-1$. While later models like **ALCOVE** [@kruschke1992alcove] attempted to model trial-by-trial, error-driven updates to the attention weights themselves, in my experience such extensions often yield terrible parameter recovery and fail stringent modern validation checks. Consequently, our approach grounds the "Process Revolution" in robust sequential exemplar accumulation. We then conceptually extend this path-dependency by introducing a memory decay parameter, providing a mechanism for agents to unlearn and adapt in non-stationary environments without sacrificing model identifiability. [It would be a very good learning exercise to try to build and validate ALCOVE and other process models, but that is beyond the scope of this chapter and we will not do it here.]

:::

#### Parameters: Cognitive Commitment, Algebraic Role, Stan Parameterization

**Attention weights $w$.** Cognitively, $w$ is a simplex of dimension-specific attention weights. Algebraically, $w$ enters the distance formula as the per-dimension coefficient in the attention-weighted city-block metric. At the Stan level $w$ is declared as `simplex[N_features]`, which enforces $w_m \ge 0$ and $\sum_m w_m = 1$ directly.

**Sensitivity $c$.** Cognitively, $c$ represents perceptual discriminability. Algebraically, $c$ is the exponential decay rate in $\eta_{ij} = e^{-c\, d_{ij}}$. At the Stan level, we sample an unconstrained `real log_c` and set `c = exp(log_c)` in `transformed parameters`.

**Response bias $\beta$.** Cognitively, $\beta$ is a baseline preference for category A independent of stimulus similarity; $\beta = 0.5$ is unbiased. In this chapter $\beta$ also serves as the cold-start fallback. At the Stan level $\beta$ is declared as `real<lower=0, upper=1>` with a Beta prior.

### Mathematical Formulation

**1. Distance Calculation**

$$d_{ij} = \left[ \sum_{m=1}^{M} w_m |x_{im} - x_{jm}|^r \right]^{\frac{1}{r}}$$

We use the city-block metric ($r = 1$), appropriate for separable stimulus dimensions:

```{r ch12_distance_formula}
# City-block distance (r = 1), attention-weighted
distance <- function(vect1, vect2, w) {
  sum(w * abs(vect1 - vect2))
}
```

**2. Similarity Computation**

$$\eta_{ij} = e^{-c \cdot d_{ij}}$$

```{r ch12_similarity_formula}
# Exponential similarity kernel
similarity <- function(dist_val, c) {
  exp(-c * dist_val)
}
```

Let's visualize how similarity changes with distance for different values of $c$:

```{r ch12_visualizing_sensitivity}
dd <- expand_grid(
  distance  = seq(0, 6, by = 0.1),
  c         = c(0.3, 0.7, 1.5, 3.0, 6.0, 10.0)
) |>
  mutate(
    sim     = similarity(distance, c),
    c_label = factor(c)
  )

ggplot(dd, aes(x = distance, y = sim, color = c_label, group = c_label)) +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(option = "plasma", name = "Sensitivity (c)") +
  labs(
    title    = "Effect of Sensitivity Parameter (c) on Similarity",
    subtitle = "c is on the natural (original) scale; prior places most mass on c ∈ (0.3, 15)",
    x        = "Psychological Distance",
    y        = "Similarity"
  ) +
  theme(legend.position = "bottom")
```

**3. Category Response Probability**

$$P(A|i) = \frac{\beta \; \bar\eta_{i,A}}{\beta \; \bar\eta_{i,A} + (1-\beta) \; \bar\eta_{i,B}}, \qquad \bar\eta_{i,A} = \frac{1}{N_A} \sum_{j \in A} \eta_{ij}$$

where $\bar\eta_{i,A}$ is the **mean** similarity of the current stimulus to the exemplars stored under category $A$, and $N_A$ is the count of those stored exemplars. The cold-start fallback applies when either category has no stored exemplars yet: $P(A|i) = \beta$.

### Summed vs. Mean Similarity: A Caveat on the Canonical Formulation

The original GCM [@nosofsky1986attention] and the ALCOVE extensions [@kruschke1992alcove] use the *summed* form $\sum_{j \in A} \eta_{ij}$ rather than the mean. That choice carries a subtle but consequential assumption: whichever category has more stored exemplars acquires a systematic advantage, independent of how similar any of those exemplars is to the current stimulus.

A concrete failure. Suppose category A has 12 stored exemplars at distance roughly $d = 2$ from the current stimulus (so each contributes $\eta_{ij} \approx e^{-2c}$) and category B has 1 stored exemplar at distance $d = 0.2$ (contributing $\eta_{ij} \approx e^{-0.2c}$). At a moderate sensitivity $c = 1$:

* Summed: $S_A = 12 \cdot e^{-2} \approx 1.62$ vs $S_B = 1 \cdot e^{-0.2} \approx 0.82$. Category A wins.
* Mean: $\bar\eta_A = e^{-2} \approx 0.14$ vs $\bar\eta_B = e^{-0.2} \approx 0.82$. Category B wins, which is what our intuition expects: the closest neighbour belongs to B.

The bias parameter $\beta$ cannot repair this. Fitting $\beta$ to the data will compensate for the *average* frequency imbalance, but cannot fix it trial-by-trial because the imbalance itself grows and shrinks as the memory store evolves.

This chapter therefore uses the **mean-similarity** formulation throughout, which is the form adopted by @ashby1993relations and by several later papers that explicitly address this point. At $N_A = N_B$ (the canonical Kruschke task with balanced training) the two formulations produce effectively identical behaviour, so replication of published Kruschke-era results is unaffected. On data with unequal base rates, partial trial sequences, or the non-stationary scenarios introduced later in this chapter — where memory composition changes meaningfully over trials — the difference becomes both measurable and interpretable: with mean similarity, $\beta$ becomes a clean response-bias parameter rather than a fitted compensator for implicit base-rate effects.

```{r ch12_sum_vs_mean_demo}
# Toy demonstration of the summed-vs-mean choice with deliberately
# unbalanced memory contents. 12 cat-A exemplars far away, 1 cat-B
# exemplar close by. We show choice probabilities under both forms.
demo_memory <- tibble(
  category = c(rep(0, 12), 1),
  sim      = c(rep(exp(-2), 12), exp(-0.2))  # pre-computed for c = 1
)
demo_sums <- demo_memory |>
  group_by(category) |>
  summarise(sum_sim = sum(sim), mean_sim = mean(sim), .groups = "drop")
demo_sums |>
  mutate(
    p_cat1_summed = sum_sim  / sum(sum_sim),
    p_cat1_mean   = mean_sim / sum(mean_sim)
  ) |>
  print()
# Summed: P(cat 1) ≈ 0.34 — the 12 distant exemplars in category 0 win.
# Mean:   P(cat 1) ≈ 0.85 — the 1 close exemplar in category 1 correctly wins.
```

The demonstration above uses bias $\beta = 0.5$ (neutral) so the asymmetry comes purely from the aggregation choice. All figures, recovery results, and SBC checks in the rest of this chapter use the mean-similarity form. Readers who want to reproduce a published summed-similarity study can swap the aggregation line in `gcm_simulate()` (replace each `mean()` with `sum()`) and the corresponding two lines in the Stan model (remove the division by `n1` and `n0`) without any other changes.

### The GCM Agent Implementation (R Simulation)

```{r ch12_generative_gcm}
# Generative GCM model function.
gcm_simulate <- function(w,            # attention weights (sums to 1)
                         c,            # sensitivity (natural scale, strictly positive)
                         bias,         # response bias towards category 1
                         obs,          # matrix of stimulus features (trials x features)
                         cat_feedback, # true category labels (0 or 1)
                         quiet = TRUE) {
  ntrials   <- nrow(obs)
  nfeatures <- ncol(obs)

  prob_cat1    <- numeric(ntrials)
  sim_response <- numeric(ntrials)

  memory_obs <- matrix(NA_real_, nrow = 0, ncol = nfeatures)
  memory_cat <- numeric(0)

  for (i in seq_len(ntrials)) {
    if (!quiet && i %% 20 == 0) cat("Simulating trial:", i, "\n")

    current_stim <- as.numeric(obs[i, ])
    n_mem        <- nrow(memory_obs)
    has_cat0     <- any(memory_cat == 0)
    has_cat1     <- any(memory_cat == 1)

    if (n_mem == 0 || !has_cat0 || !has_cat1) {
      prob_cat1[i] <- bias
    } else {
      sims <- vapply(seq_len(n_mem), function(e) {
        similarity(distance(current_stim, memory_obs[e, ], w), c)
      }, numeric(1))

      # Mean similarity per category — removes category-frequency bias.
      m1  <- mean(sims[memory_cat == 1])
      m0  <- mean(sims[memory_cat == 0])
      num <- bias * m1
      den <- num + (1 - bias) * m0

      prob_cat1[i] <- if (den > 1e-9) num / den else bias
      prob_cat1[i] <- pmax(1e-9, pmin(1 - 1e-9, prob_cat1[i]))
    }

    sim_response[i] <- rbinom(1, 1, prob_cat1[i])

    memory_obs <- rbind(memory_obs, current_stim)
    memory_cat <- c(memory_cat, cat_feedback[i])
  }

  tibble(prob_cat1 = prob_cat1, sim_response = sim_response)
}
```

---

### Experimental Setup: Why @kruschke1993human?

To test our GCM implementation, we need a stimulus environment that is more complex than a simple linear separation but structured enough to reveal cognitive principles. We adopt the setup from @kruschke1993human for three specific reasons:

1.  **Continuous Dimension Benchmarking:** Unlike earlier binary-feature tasks (e.g., @medin1978context), Kruschke uses stimuli (rectangles varying in height and position) that occupy a continuous 2D psychological space. This is the natural environment for the GCM's Minkowski distance metrics.

2.  **Testing Selective Attention:** The task is designed so that both dimensions (height and position) are relevant, but their importance depends on the specific stimuli. This allows us to observe whether our model can "discover" the optimal attention weights ($w$) to minimize categorization error—a core feature of the ALCOVE extension to the GCM.

3.  **Non-Linearity & Abstraction:** The category structure (plotted below) is not perfectly separable by a single horizontal or vertical line. It requires the agent to either learn a diagonal boundary or to store the specific exemplars that form the "corners" of each category. This makes it a diagnostic benchmark for comparing exemplar models against the prototype models we'll see in Chapter 12.

```{r ch12_experimental_setup}
stimulus_info <- tibble(
  stimulus      = c(5, 3, 7, 1, 8, 2, 6, 4),
  height        = c(1, 1, 2, 2, 3, 3, 4, 4),
  position      = c(2, 3, 1, 4, 1, 4, 2, 3),
  category_true = c(0, 0, 1, 0, 1, 0, 1, 1)
)

ggplot(stimulus_info, aes(x = position, y = height,
                          color = factor(category_true), label = stimulus)) +
  geom_point(size = 5) +
  geom_text(nudge_y = 0.18) +
  scale_color_manual(values = c("0" = "#0072B2", "1" = "#D55E00"), name = "Category") +
  labs(title = "@kruschke1993human Stimulus Structure", x = "Position", y = "Height") +
  coord_fixed()

n_blocks         <- 8
n_stim_per_block <- nrow(stimulus_info)
total_trials     <- n_blocks * n_stim_per_block
```

---

### Per-Subject Schedule Generation

```{r ch12_subject_schedule}
make_subject_schedule <- function(stimulus_info, n_blocks, seed) {
  set.seed(seed)
  n_stim <- nrow(stimulus_info)

  sequence <- unlist(lapply(seq_len(n_blocks), function(b) {
    sample(stimulus_info$stimulus, n_stim, replace = FALSE)
  }))

  tibble(
    trial_within_subject = seq_along(sequence),
    block                = rep(seq_len(n_blocks), each = n_stim),
    stimulus_id          = sequence
  ) |>
    left_join(stimulus_info, by = c("stimulus_id" = "stimulus")) |>
    rename(category_feedback = category_true) |>
    dplyr::select(trial_within_subject, block, stimulus_id,
                  height, position, category_feedback)
}
```

---

### Simulating Behavior with the GCM

```{r ch12_gcm_agent_wrapper}
simulate_gcm_agent <- function(agent_id, w_true, c_true, bias_true,
                               stimulus_info, n_blocks, subject_seed) {
  schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  obs      <- schedule |> dplyr::select(height, position)
  feedback <- schedule$category_feedback

  sim_results <- gcm_simulate(
    w = w_true, c = c_true, bias = bias_true,
    obs = obs, cat_feedback = feedback, quiet = TRUE
  )

  schedule |>
    mutate(
      agent_id     = agent_id,
      w1_true      = w_true[1],
      w2_true      = w_true[2],
      c_true       = c_true,
      log_c_true   = log(c_true),
      bias_true    = bias_true,
      prob_cat1    = sim_results$prob_cat1,
      sim_response = sim_results$sim_response,
      correct      = as.integer(category_feedback == sim_response)
    ) |>
    group_by(agent_id) |>
    mutate(performance = cumsum(correct) / row_number()) |>
    ungroup()
}
```

```{r ch12_gcm_simulations}
param_grid <- expand_grid(
  w_setting    = list(c(0.5, 0.5), c(0.7, 0.3), c(0.9, 0.1)),
  c_setting    = c(0.5, 1.0, 2.0, 4.0, 5.0, 15.0),
  bias_setting = 0.5
) |>
  mutate(
    setting_id = row_number(),
    w_label    = map_chr(w_setting, ~paste(round(.x, 1), collapse = ",")),
    c_label    = paste0("c≈", round(c_setting, 1)),
    label      = paste0("w=(", w_label, "), ", c_label)
  )

n_agents_per_setting <- 5

simulation_plan <- param_grid |>
  uncount(n_agents_per_setting) |>
  mutate(agent_id = row_number())

sim_file <- here("simdata", "ch12_gcm_simulated_responses.csv")

if (regenerate_simulations || !file.exists(sim_file)) {
  plan(multisession, workers = max(1, availableCores() - 1))

  simulated_responses <- future_pmap_dfr(
    list(
      agent_id  = simulation_plan$agent_id,
      w_true    = simulation_plan$w_setting,
      c_true    = simulation_plan$c_setting,
      bias_true = simulation_plan$bias_setting
    ),
    simulate_gcm_agent,
    stimulus_info = stimulus_info,
    n_blocks      = n_blocks,
    subject_seed  = simulation_plan$agent_id,
    .options = furrr_options(seed = TRUE)
  )

  plan(sequential)
  write_csv(simulated_responses, sim_file)
  cat("Simulation complete. Results saved.\n")
} else {
  simulated_responses <- read_csv(sim_file, show_col_types = FALSE)
  cat("Loaded existing simulation results.\n")
}
```

```{r ch12_visualize_parameter_effects}
simulated_responses_labeled <- simulated_responses |>
  mutate(
    w_label = paste0("w=(", round(w1_true, 1), ",", round(w2_true, 1), ")"),
    c_label = paste0("c≈", round(c_true, 1)),
    # Order the c_label factor based on the numeric c_true values
    c_label = forcats::fct_reorder(c_label, c_true) 
  )

p_c_effect <- ggplot(
  simulated_responses_labeled,
  aes(x = trial_within_subject, y = performance,
      color = factor(round(c_true, 1)),
      group = interaction(agent_id, round(c_true, 1)))
) +
  stat_summary(fun = mean, geom = "line",
               aes(group = factor(round(c_true, 1))), linewidth = 1.2) +
  facet_wrap(~w_label) +
  scale_color_viridis_d(option = "plasma", name = "Sensitivity (c)") +
  labs(
    title    = "Effect of Sensitivity (c) on GCM Learning",
    subtitle = "Mean across simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme(legend.position = "bottom")

p_w_effect <- ggplot(
  simulated_responses_labeled,
  aes(x = trial_within_subject, y = performance,
      color = w_label, group = interaction(agent_id, w_label))
) +
  stat_summary(fun = mean, geom = "line", aes(group = w_label), linewidth = 1.2) +
  facet_wrap(~c_label) +
  scale_color_brewer(palette = "Set1", name = "Attention Weights") +
  labs(
    title    = "Effect of Attention Weights (w) on GCM Learning",
    subtitle = "Mean across simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme(legend.position = "bottom")

p_c_effect
p_w_effect
```

---

### Prior Predictive Check

We need to define the priors for the Stan model already here:

- $w \sim \text{Dirichlet}(1, 1)$ — uniform over the simplex
- $\log c \sim \text{Normal}(\log(2), 1)$ — implies $c \in (0.27, 14.8)$ at $\pm 2$ SD
- $\text{bias} \sim \text{Beta}(1, 1)$ — uniform over $(0, 1)$

```{r ch12_prior_predictive}
n_prior_samples <- 500

prior_draws <- tibble(
  sample_id = seq_len(n_prior_samples),
  log_c     = rnorm(n_prior_samples, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD),
  w1        = runif(n_prior_samples),
  bias      = rbeta(n_prior_samples, 1, 1)
) |>
  mutate(c_true = exp(log_c), w2 = 1 - w1)

ppc_schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = 999)

prior_pred_curves <- pmap_dfr(
  list(
    sample_id = prior_draws$sample_id,
    log_c     = prior_draws$log_c,
    w1        = prior_draws$w1,
    bias      = prior_draws$bias
  ),
  function(sample_id, log_c, w1, bias) {
    c <- exp(log_c)
    sim <- gcm_simulate(
      w = c(w1, 1 - w1), c = c, bias = bias,
      obs          = ppc_schedule |> dplyr::select(height, position),
      cat_feedback = ppc_schedule$category_feedback
    )
    tibble(
      sample_id = sample_id,
      trial     = seq_len(nrow(ppc_schedule)),
      correct   = as.integer(sim$sim_response == ppc_schedule$category_feedback)
    ) |>
      mutate(cum_acc = cumsum(correct) / row_number())
  }
)

prior_pred_summary <- prior_pred_curves |>
  group_by(trial) |>
  summarise(
    q05 = quantile(cum_acc, 0.05),
    q25 = quantile(cum_acc, 0.25),
    q50 = median(cum_acc),
    q75 = quantile(cum_acc, 0.75),
    q95 = quantile(cum_acc, 0.95),
    .groups = "drop"
  )

ggplot(prior_pred_summary, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#0072B2", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Prior Predictive Check: GCM Learning Curves",
    subtitle = "Ribbons: 50% and 90% prior predictive intervals; line: median\nPriors: log(c) ~ Normal(log 2, 1),  w ~ Dirichlet(1,1),  bias ~ Beta(1,1)",
    x = "Trial", y = "Cumulative Accuracy"
  )
```

---

### Implementing the GCM in Stan for Parameter Estimation

#### Architectural choice: `prob_cat1[i]` in `transformed parameters`

Because the GCM's per-trial choice probability $p_i$ is a deterministic function of the parameters $(w, c, \beta)$ and the data sequence, it belongs in `transformed parameters` as a single source of truth. The `model{}` block then becomes a one-line vectorised Bernoulli likelihood, and `generated quantities{}` becomes a lookup against the already-computed `prob_cat1` vector.

```{tikz gcm-single-dag, fig.cap = "DAG for the Single-Subject GCM. Hyperparameters (dots) feed into latent parameters (circles): attention weights $\\mathbf{w}$, log sensitivity $\\log c$, and response bias $\\beta$. Inside the plate, the stimulus $\\mathbf{x}_i$ and accumulated exemplar memory $z_{<i}$ (past category feedback) combine with the parameters to yield the deterministic choice probability $p_i$ (double circle), from which the response $y_i$ is drawn.", fig.align = "center", out.width = "70%"}
\usetikzlibrary{bayesnet}
\begin{tikzpicture}
  % ── Hyperparameters (const) ─────────────────────────────
  \node[const] (alpha_w) at (0, 0)   {$\boldsymbol{\alpha}_w$};
  \node[const] (hyp_c)   at (4, 0)   {$\mu_c,\,\sigma_c$};
  \node[const] (hyp_b)   at (8, 0)   {$\alpha_\beta,\,\beta_\beta$};

  % ── Latent parameters ───────────────────────────────────
  \node[latent] (w)    at (0, -2)  {$\mathbf{w}$};
  \node[latent] (logc) at (4, -2)  {$\log c$};
  \node[latent] (bias) at (8, -2)  {$\beta$};

  % ── Trial-level nodes (inside plate) ────────────────────
  \node[obs] (xi) at (2, -4.5) {$\mathbf{x}_i$};
  \node[obs] (zi) at (4, -4.5) {$z_{<i}$};
  \node[det] (pi) at (6, -4.5) {$p_i$};
  \node[obs] (yi) at (6, -6.5) {$y_i$};

  % ── Edges ───────────────────────────────────────────────
  \edge {alpha_w}        {w};
  \edge {hyp_c}          {logc};
  \edge {hyp_b}          {bias};
  \edge {w, logc, bias}  {pi};
  \edge {xi, zi}         {pi};
  \edge {pi}             {yi};

  % ── Plate ───────────────────────────────────────────────
  \plate {trial} {(xi)(zi)(pi)(yi)} {$N$ trials};
\end{tikzpicture}
```

The DAG formalizes the four-stage structure from §Mathematical Formulation above: priors on $(w, \log c, \beta)$ feed into attention-weighted city-block distances, which yield per-category mean similarities, which enter the Luce choice rule to produce $p_i$; the response $y_i$ is then a Bernoulli draw.

```{r ch12_gcm_stan_model}
gcm_single_stan <- "
// Generalized Context Model — Single Subject (refactored architecture)

data {
  int<lower=1> N_total;
  int<lower=1> N_features;
  array[N_total] int<lower=0, upper=1> y;
  array[N_total, N_features] real obs;
  array[N_total] int<lower=0, upper=1> cat_feedback;

  vector[N_features] prior_w_alpha;
  real prior_log_c_mu;
  real<lower=0> prior_log_c_sigma;
  real<lower=0> prior_bias_alpha;
  real<lower=0> prior_bias_beta;
  int<lower=0, upper=1> run_diagnostics;
}

transformed data {
  array[N_total, N_total, N_features] real abs_diff;
  for (i in 1:N_total) {
    for (j in 1:N_total) {
      for (f in 1:N_features) {
        abs_diff[i, j, f] = abs(obs[i, f] - obs[j, f]);
      }
    }
  }
}

parameters {
  simplex[N_features] w;
  real log_c;
  real<lower=0, upper=1> bias;
}

transformed parameters {
  real<lower=0> c = exp(log_c);
  vector<lower=1e-9, upper=1-1e-9>[N_total] prob_cat1;

  {
    array[N_total] int memory_trial_idx;
    array[N_total] int memory_cat;
    int n_mem = 0;

    for (i in 1:N_total) {
      real p_i;
      int has_cat0 = 0;
      int has_cat1 = 0;

      for (k in 1:n_mem) {
        if (memory_cat[k] == 0) has_cat0 = 1;
        if (memory_cat[k] == 1) has_cat1 = 1;
      }

      if (n_mem == 0 || has_cat0 == 0 || has_cat1 == 0) {
        p_i = bias;
      } else {
        real s1 = 0; real s0 = 0;
        int n1 = 0;  int n0 = 0;
        for (e in 1:n_mem) {
          real d = 0;
          int past_i = memory_trial_idx[e];
          for (f in 1:N_features)
            d += w[f] * abs_diff[i, past_i, f];
          real sim = exp(-c * d);
          if (memory_cat[e] == 1) { s1 += sim; n1 += 1; }
          else                    { s0 += sim; n0 += 1; }
        }
        real m1 = s1 / n1;
        real m0 = s0 / n0;
        real num = bias * m1;
        real den = num + (1 - bias) * m0;
        p_i = (den > 1e-9) ? num / den : bias;
      }

      prob_cat1[i] = fmax(1e-9, fmin(1 - 1e-9, p_i));

      n_mem += 1;
      memory_trial_idx[n_mem] = i;
      memory_cat[n_mem] = cat_feedback[i];
    }
  }
}

model {
  target += dirichlet_lpdf(w    | prior_w_alpha);
  target += normal_lpdf(log_c   | prior_log_c_mu, prior_log_c_sigma);
  target += beta_lpdf(bias      | prior_bias_alpha,  prior_bias_beta);

  target += bernoulli_lpmf(y | prob_cat1);
}

generated quantities {
  vector[N_total] log_lik;
  real lprior;
  if (run_diagnostics) {
    for (i in 1:N_total)
      log_lik[i] = bernoulli_lpmf(y[i] | prob_cat1[i]);
  }
  lprior = dirichlet_lpdf(w | prior_w_alpha) +
           normal_lpdf(log_c | prior_log_c_mu, prior_log_c_sigma) +
           beta_lpdf(bias | prior_bias_alpha, prior_bias_beta);
}
"

stan_file_gcm_single <- "stan/ch12_gcm_single.stan"
write_stan_file(gcm_single_stan, dir = "stan/", basename = "ch12_gcm_single.stan")
mod_gcm_single <- cmdstan_model(stan_file_gcm_single)
```

#### What each block does:

* **`transformed data`**: Builds the `abs_diff[i, j, f]` tensor. Evaluated exactly once per chain.
* **`parameters`**: `simplex[N_features] w`, unconstrained `real log_c`, and `real<lower=0, upper=1> bias`.
* **`transformed parameters`**: Brings back the constraints on the parameters, and caches `prob_cat1[i]` in a single forward pass over trials.
* **`model`**: Priors followed by a vectorised one-line likelihood.
* **`generated quantities`**: `log_lik[i]` as a per-trial lookup, and `lprior` for `priorsense`.

> A note on reduce_sum and parallelization: The GCM's inner loop cannot be parallelized
> with reduce_sum because trial $t$'s contribution to target depends on the memory state
> accumulated sequentially from trials $1, \ldots, t-1$.

#### Fitting the Single-Subject GCM

```{r ch12_fitting_gcm_single}
# Select one simulated agent: w ≈ (0.9, 0.1), c ≈ 2
agent_to_fit <- simulated_responses |>
  filter(
    round(w1_true, 1) == 0.9,
    abs(log_c_true - log(2.0)) == min(abs(log_c_true - log(2.0)))
  ) |>
  filter(agent_id == min(agent_id)) |>
  slice(1:total_trials)

stopifnot(nrow(agent_to_fit) == total_trials)

gcm_data_single <- list(
  N_total           = nrow(agent_to_fit),
  N_features        = 2L,
  y                 = as.integer(agent_to_fit$sim_response),
  obs               = as.matrix(agent_to_fit[, c("height", "position")]),
  cat_feedback      = as.integer(agent_to_fit$category_feedback),
  prior_w_alpha     = c(1, 1),
  prior_log_c_mu    = LOG_C_PRIOR_MEAN,
  prior_log_c_sigma = LOG_C_PRIOR_SD,
  prior_bias_alpha  = 1,
  prior_bias_beta   = 1,
  run_diagnostics   = 1L
)

fit_filepath_single <- here("simmodels", "ch12_gcm_single_fit.rds")

if (regenerate_simulations || !file.exists(fit_filepath_single)) {
  pf_single <- tryCatch(
    mod_gcm_single$pathfinder(
      data      = gcm_data_single,
      seed      = 123,
      num_paths = 4,
      refresh   = 0
    ),
    error = function(e) {
      message("Pathfinder failed — falling back to default initialisation.")
      NULL
    }
  )

  fit_gcm_single <- mod_gcm_single$sample(
    data            = gcm_data_single,
    init            = pf_single,
    seed            = 123,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1000,
    iter_sampling   = 1500,
    refresh         = 300,
    adapt_delta     = 0.9
  )
  fit_gcm_single$save_object(fit_filepath_single)
  cat("Single-agent model fitted and saved.\n")
} else {
  fit_gcm_single <- readRDS(fit_filepath_single)
  cat("Loaded existing single-agent fit.\n")
}
```

#### MCMC Diagnostic Battery

```{r ch12_single_diagnostic_battery}
if (!is.null(fit_gcm_single)) {
  diag_tbl <- diagnostic_summary_table(
    fit_gcm_single,
    params = c("w[1]", "w[2]", "log_c", "c", "bias")
  )
  print(diag_tbl)

  if (!all(diag_tbl$pass)) {
    warning("Diagnostic battery has failures — interpret posteriors with caution.")
  }

  bayesplot::mcmc_trace(
    fit_gcm_single$draws(c("log_c", "w[1]", "bias")),
    facet_args = list(ncol = 1)
  )

  bayesplot::mcmc_rank_overlay(
    fit_gcm_single$draws(c("log_c", "w[1]", "bias"))
  )
}
```

#### Posterior Predictive Check (single subject)

```{r ch12_ppc_single}
if (!is.null(fit_gcm_single)) {
  prob_draws <- fit_gcm_single$draws("prob_cat1", format = "matrix")
  n_draws    <- nrow(prob_draws)
  n_trials   <- ncol(prob_draws)

  set.seed(2025)
  yrep_post <- matrix(rbinom(n_draws * n_trials, 1, as.vector(prob_draws)),
                      nrow = n_draws)

  obs_acc <- cummean(agent_to_fit$sim_response == agent_to_fit$category_feedback)
  rep_acc <- t(apply(yrep_post, 1, function(yr) {
    cummean(yr == agent_to_fit$category_feedback)
  }))

  ppc_summary <- tibble(
    trial = seq_len(n_trials),
    obs   = obs_acc,
    q05   = apply(rep_acc, 2, quantile, 0.05),
    q25   = apply(rep_acc, 2, quantile, 0.25),
    q50   = apply(rep_acc, 2, quantile, 0.50),
    q75   = apply(rep_acc, 2, quantile, 0.75),
    q95   = apply(rep_acc, 2, quantile, 0.95)
  )

  ggplot(ppc_summary, aes(x = trial)) +
    geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
    geom_line(aes(y = q50), color = "#0072B2", linewidth = 0.8) +
    geom_line(aes(y = obs), color = "#D55E00", linewidth = 1) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
      title    = "Posterior Predictive Check: Cumulative Accuracy",
      subtitle = "Blue ribbons: 50% and 90% posterior predictive intervals\nOrange line: observed cumulative accuracy",
      x = "Trial", y = "Cumulative Accuracy"
    )
}
```


Note that we picked cumulative accuracy as target measure for the PPC because it is a more holistic summary of the learning dynamics than trial-level accuracy. The GCM's path-dependence means that trial-level predictions are noisier and less informative about model fit.
We could have also picked other behaviors, such as the evolving distribution of attention-weighted distances to exemplars, or specific contrast 

#### Prior–Posterior Update Plot

```{r ch12_prior_posterior_single}
if (!is.null(fit_gcm_single)) {
  draws_df <- as_draws_df(
    fit_gcm_single$draws(variables = c("w[1]", "log_c", "c", "bias"))
  )

  true_w1    <- agent_to_fit$w1_true[1]
  true_log_c <- agent_to_fit$log_c_true[1]
  true_c     <- agent_to_fit$c_true[1]
  true_bias  <- agent_to_fit$bias_true[1]

  set.seed(2026)
  n_prior <- 4000
  prior_df <- tibble(
    `w[1]` = runif(n_prior),
    log_c  = rnorm(n_prior, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD),
    c      = exp(log_c),
    bias   = rbeta(n_prior, 1, 1)
  )

  pp_panel <- function(prior_x, post_x, true_val, label) {
    tibble(
      x      = c(prior_x, post_x),
      source = c(rep("Prior", length(prior_x)),
                 rep("Posterior", length(post_x)))
    ) |>
      ggplot(aes(x = x, fill = source, color = source)) +
      geom_density(alpha = 0.4, linewidth = 0.5) +
      geom_vline(xintercept = true_val, color = "red",
                 linetype = "dashed", linewidth = 0.7) +
      scale_fill_manual(values = c(Prior = "grey60", Posterior = "#0072B2")) +
      scale_color_manual(values = c(Prior = "grey40", Posterior = "#0072B2")) +
      labs(title = label, x = NULL, y = "Density") +
      theme(legend.position = "bottom")
  }

  p1 <- pp_panel(prior_df$`w[1]`, draws_df$`w[1]`, true_w1, "w[1]")
  p2 <- pp_panel(prior_df$log_c,  draws_df$log_c,  true_log_c, "log(c) — computational scale")
  p3 <- pp_panel(prior_df$c,      draws_df$c,      true_c,     "c — natural scale")
  p4 <- pp_panel(prior_df$bias,   draws_df$bias,   true_bias,  "bias")

  (p1 | p2) / (p3 | p4) +
    plot_annotation(
      title    = "Prior–Posterior Update for the Single-Subject GCM",
      subtitle = "Red dashed line = true generating value"
    )
}
```

This looks surprisingly good! — the posterior for $w[1]$ is tightly concentrated around the true value of 0.9, and the posterior for $\log c$ is also well-concentrated around the true $\log(2)$. The bias parameter is less informed, which makes sense given that it has a more indirect influence on the choice probabilities.

#### Joint Posterior Geometry: $c$ vs $w$

```{r ch12_cw_identifiability_pairs}
if (!is.null(fit_gcm_single)) {
  bayesplot::mcmc_pairs(
    fit_gcm_single$draws(c("log_c", "w[1]", "bias")),
    diag_fun     = "dens",
    off_diag_fun = "hex"
  )
}
```

The parameters look nicely uncorrelated.

---

#### LOO with Pareto-$\hat{k}$ Diagnostics

Standard PSIS-LOO is computed for completeness and for comparison against LFO-CV below. We do not interpret it as a valid measure of out-of-sample predictive accuracy for a sequential model — see the next subsection for why — but the Pareto-$\hat{k}$ diagnostic still localises trials where the path-dependence concern bites hardest.

```{r ch12_loo_single}
if (!is.null(fit_gcm_single)) {
  log_lik_mat    <- fit_gcm_single$draws("log_lik", format = "matrix")
  loo_gcm_single <- loo::loo(log_lik_mat, cores = 4)
  print(loo_gcm_single)

  plot(loo_gcm_single, label_points = TRUE,
       main = "PSIS-LOO Pareto-k diagnostics: Single-subject GCM")
}
```

#### Leave-Future-Out Cross-Validation

**Why standard LOO is invalid here.** Standard PSIS-LOO assumes observations are conditionally exchangeable given the parameters. In the GCM, this assumption is violated: leaving out observation $t$ also conceptually removes the memory trace that observation $t$ contributed to trials $t+1, \ldots, T$.

**What LFO-CV does instead.** LFO-CV (Bürkner, Gabry, & Vehtari, 2020) estimates the *one-step-ahead* predictive density. The PSIS approximation refits only when the importance sampling deteriorates ($\hat{k} > k_\text{thresh}$) and uses importance reweighting otherwise.

```{r ch12_lfo_cv_implementation}
# ─────────────────────────────────────────────────────────────────────────────
# psis_lfo_gcm: PSIS-approximated leave-future-out CV for the GCM.
# Follows Bürkner, Gabry, & Vehtari (2020), Algorithm 1.
# ─────────────────────────────────────────────────────────────────────────────
psis_lfo_gcm <- function(stan_model,
                         full_data,
                         L = 16,
                         k_thresh = 0.7,
                         seed = 1,
                         verbose = FALSE) {

  fit_subset <- function(n_train, seed) {
    sub_data <- full_data
    sub_data$N_total      <- n_train
    sub_data$y            <- sub_data$y[1:n_train]
    sub_data$obs          <- sub_data$obs[1:n_train, , drop = FALSE]
    sub_data$cat_feedback <- sub_data$cat_feedback[1:n_train]

    pf  <- stan_model$pathfinder(data = sub_data, num_paths = 4, refresh = 0,
                                  seed = seed)
    fit <- stan_model$sample(
      data = sub_data, init = pf, seed = seed,
      chains = 2, parallel_chains = 2,
      iter_warmup = 600, iter_sampling = 800,
      refresh = 0, adapt_delta = 0.9
    )
    fit
  }

  predict_future <- function(fit, full_data) {
    full_n   <- length(full_data$y)
    ext_data <- full_data
    ext_data$N_total <- full_n

    gq_fit <- stan_model$generate_quantities(
      fitted_params   = fit,
      data            = ext_data,
      seed            = seed,
      parallel_chains = 2
    )
    gq_fit$draws("log_lik", format = "matrix")
  }

  T_total      <- length(full_data$y)
  elpd_lfo     <- numeric(T_total)
  elpd_lfo[]   <- NA_real_
  refits       <- integer(0)
  k_history    <- numeric(0)

  fit_curr     <- fit_subset(L, seed = seed)
  ll_full_curr <- predict_future(fit_curr, full_data)

  refits       <- c(refits, L)
  last_fit_n   <- L
  n_draws      <- nrow(ll_full_curr)

  ll_train_curr <- rowSums(ll_full_curr[, 1:L, drop = FALSE])

  for (t in (L + 1):T_total) {
    if (t - 1 == last_fit_n) {
      k_t          <- 0
      k_history[t] <- k_t
      lw           <- rep(-log(n_draws), n_draws)
    } else {
      ll_train_target <- rowSums(ll_full_curr[, 1:(t - 1), drop = FALSE])
      log_w           <- ll_train_target - ll_train_curr

      psis_obj     <- loo::psis(log_w, r_eff = NA)
      k_t          <- loo::pareto_k_values(psis_obj)
      k_history[t] <- k_t

      if (k_t > k_thresh) {
        if (verbose) cat("Refit at t =", t, "(k =", round(k_t, 3), ")\n")
        fit_curr      <- fit_subset(t - 1, seed = seed + t)
        ll_full_curr  <- predict_future(fit_curr, full_data)
        ll_train_curr <- rowSums(ll_full_curr[, 1:(t - 1), drop = FALSE])
        last_fit_n    <- t - 1
        refits        <- c(refits, last_fit_n)
        lw            <- rep(-log(n_draws), n_draws)
      } else {
        lw <- as.numeric(weights(psis_obj))
      }
    }

    lp          <- ll_full_curr[, t]
    elpd_lfo[t] <- matrixStats::logSumExp(lw + lp)
  }

  list(
    elpd_lfo       = sum(elpd_lfo, na.rm = TRUE),
    elpd_pointwise = elpd_lfo,
    refits         = refits,
    k_history      = k_history,
    L              = L
  )
}

lfo_filepath <- here("simdata", "ch12_lfo_single.rds")
if (regenerate_simulations || !file.exists(lfo_filepath)) {
  lfo_single <- psis_lfo_gcm(
    stan_model = mod_gcm_single,
    full_data  = gcm_data_single,
    L          = 16,
    k_thresh   = 0.7,
    seed       = 11,
    verbose    = TRUE
  )
  saveRDS(lfo_single, lfo_filepath)
} else {
  lfo_single <- readRDS(lfo_filepath)
}

cat("LFO-CV ELPD (single subject):", round(lfo_single$elpd_lfo, 2), "\n")
cat("Number of refits:",             length(lfo_single$refits),     "\n")
cat("Refit triggered at trials:",    lfo_single$refits,             "\n")

if (!is.null(fit_gcm_single)) {
  loo_pw  <- loo_gcm_single$pointwise[, "elpd_loo"]
  comp_df <- tibble(
    trial    = seq_along(loo_pw),
    elpd_loo = loo_pw,
    elpd_lfo = lfo_single$elpd_pointwise
  ) |>
    pivot_longer(starts_with("elpd_"), names_to = "method", values_to = "elpd")

  ggplot(comp_df, aes(x = trial, y = elpd, color = method)) +
    geom_line(alpha = 0.7) +
    geom_point(alpha = 0.5) +
    scale_color_manual(values = c(elpd_loo = "#0072B2", elpd_lfo = "#D55E00"),
                       labels = c("PSIS-LFO (correct)",
                                  "PSIS-LOO (invalid for sequential)")) +
    labs(
      title    = "Per-trial ELPD: LOO vs LFO-CV",
      subtitle = "Differences are largest where the path-dependence assumption fails most",
      x = "Trial", y = "Pointwise ELPD"
    ) +
    theme(legend.position = "bottom")
}
```

**Validation against exact one-step-ahead refits.**

```{r ch12_exact_lfo_validation}
exact_lfo_grid     <- seq(16, length(gcm_data_single$y) - 1, by = 8)
exact_lfo_filepath <- here("simdata", "ch12_exact_lfo_single.rds")

if (regenerate_simulations || !file.exists(exact_lfo_filepath)) {
  exact_results <- map_dfr(exact_lfo_grid, function(n_train) {
    sub_data <- gcm_data_single
    sub_data$N_total      <- n_train
    sub_data$y            <- sub_data$y[1:n_train]
    sub_data$obs          <- sub_data$obs[1:n_train, , drop = FALSE]
    sub_data$cat_feedback <- sub_data$cat_feedback[1:n_train]

    pf  <- mod_gcm_single$pathfinder(data = sub_data, num_paths = 4,
                                      refresh = 0, seed = n_train)
    fit <- mod_gcm_single$sample(
      data = sub_data, init = pf, seed = n_train,
      chains = 2, parallel_chains = 2,
      iter_warmup = 600, iter_sampling = 800,
      refresh = 0, adapt_delta = 0.9
    )

    ext <- gcm_data_single
    gq  <- mod_gcm_single$generate_quantities(
      fitted_params = fit, data = ext, seed = n_train, parallel_chains = 2
    )
    ll_target  <- gq$draws("log_lik", format = "matrix")[, n_train + 1]
    elpd_exact <- matrixStats::logSumExp(ll_target) - log(length(ll_target))

    tibble(trial = n_train + 1, elpd_exact = elpd_exact)
  })
  saveRDS(exact_results, exact_lfo_filepath)
} else {
  exact_results <- readRDS(exact_lfo_filepath)
}

psis_at_grid <- tibble(
  trial     = exact_results$trial,
  elpd_psis = lfo_single$elpd_pointwise[exact_results$trial]
)

agreement_df <- exact_results |>
  left_join(psis_at_grid, by = "trial")

ggplot(agreement_df, aes(x = elpd_exact, y = elpd_psis)) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_point(size = 3, color = "#0072B2") +
  labs(
    title    = "PSIS-LFO vs Exact One-Step-Ahead Refits",
    subtitle = "Points should fall on the y = x line if PSIS-LFO is calibrated",
    x = "Exact ELPD (refit per point)",
    y = "PSIS-LFO ELPD"
  ) +
  coord_fixed()
```

#### Randomised LOO-PIT

```{r ch12_randomized_loo_pit}
if (!is.null(fit_gcm_single)) {
  prob_draws  <- fit_gcm_single$draws("prob_cat1", format = "matrix")
  log_lik_mat <- fit_gcm_single$draws("log_lik",   format = "matrix")

  psis_obj <- loo::psis(-log_lik_mat, r_eff = NA)
  lw       <- weights(psis_obj, normalize = TRUE)

  y_obs <- gcm_data_single$y

  loo_p_cat1 <- vapply(seq_along(y_obs), function(t) {
    w <- exp(lw[, t] - matrixStats::logSumExp(lw[, t]))
    sum(w * prob_draws[, t])
  }, numeric(1))

  set.seed(2027)
  pit_vals <- ifelse(
    y_obs == 1,
    runif(length(y_obs), 1 - loo_p_cat1, 1),
    runif(length(y_obs), 0,              1 - loo_p_cat1)
  )

  ggplot(tibble(pit = pit_vals), aes(x = pit)) +
    geom_histogram(aes(y = after_stat(density)),
                   bins = 20, fill = "#0072B2", alpha = 0.7,
                   boundary = 0) +
    geom_hline(yintercept = 1, color = "red", linetype = "dashed") +
    scale_x_continuous(limits = c(0, 1)) +
    labs(
      title    = "Randomised LOO-PIT for the Single-Subject GCM",
      subtitle = "If the model is calibrated, PIT values are Uniform(0,1) (red dashed line)",
      x = "Randomised LOO-PIT", y = "Density"
    )
}
```

#### Prior Sensitivity Analysis

```{r ch12_prior_sensitivity_single}
if (!is.null(fit_gcm_single)) {
  ps <- priorsense::powerscale_sensitivity(fit_gcm_single)
  print(ps)

  priorsense::powerscale_plot_dens(
    fit_gcm_single,
    variables = c("log_c", "w[1]", "bias")
  )
}
```

---

### Single-Subject Simulation-Based Calibration Checks (SBC)

SBC checks the internal consistency and calibration of the entire Bayesian inference procedure across the *whole* prior. It asks: "If the true parameters really came from our specified priors, would our inference procedure produce posteriors that are statistically consistent with those priors?"

```{r ch12_sbc_single}
sbc_single_filepath <- here("simdata", "ch12_sbc_single_results.rds")

if (regenerate_simulations || !file.exists(sbc_single_filepath)) {
  n_sbc_iterations_single <- 500   # use >= 1000 for publication-quality SBC

  gcm_single_sbc_generator <- SBC_generator_function(
    function() {
      w_true     <- as.numeric(MCMCpack::rdirichlet(1, c(1, 1)))
      log_c_true <- rnorm(1, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD)
      c_true     <- exp(log_c_true)
      bias_true  <- rbeta(1, 1, 1)

      sbc_schedule <- make_subject_schedule(stimulus_info, n_blocks,
                                            seed = sample.int(1e8, 1))

      sim <- gcm_simulate(
        w            = w_true,
        c            = c_true,
        bias         = bias_true,
        obs          = sbc_schedule |> dplyr::select(height, position),
        cat_feedback = sbc_schedule$category_feedback
      )

      list(
        variables = list(
          `w[1]` = w_true[1],
          `w[2]` = w_true[2],
          log_c  = log_c_true,
          bias   = bias_true
        ),
        generated = list(
          N_total           = nrow(sbc_schedule),
          N_features        = 2L,
          y                 = as.integer(sim$sim_response),
          obs               = as.matrix(sbc_schedule[, c("height", "position")]),
          cat_feedback      = as.integer(sbc_schedule$category_feedback),
          prior_w_alpha     = c(1, 1),
          prior_log_c_mu    = LOG_C_PRIOR_MEAN,
          prior_log_c_sigma = LOG_C_PRIOR_SD,
          prior_bias_alpha  = 1,
          prior_bias_beta   = 1,
          run_diagnostics   = 1L
        )
      )
    }
  )

  gcm_single_sbc_backend <- SBC_backend_cmdstan_sample(
    mod_gcm_single,
    iter_warmup   = 1000,
    iter_sampling = 1000,
    chains        = 2,
    adapt_delta   = 0.9,
    refresh       = 0
  )

  gcm_single_datasets <- generate_datasets(
    gcm_single_sbc_generator,
    n_sims = n_sbc_iterations_single
  )

  sbc_results_single <- compute_SBC(
    gcm_single_datasets,
    gcm_single_sbc_backend,
    cache_mode     = "results",
    cache_location = here("simdata", "ch12_sbc_single_cache"),
    keep_fits      = FALSE
  )
  saveRDS(sbc_results_single, sbc_single_filepath)
  cat("SBC single-subject results computed and saved.\n")
} else {
  sbc_results_single <- readRDS(sbc_single_filepath)
  cat("Loaded existing SBC single-subject results.\n")
}

plot_ecdf_diff(sbc_results_single)
plot_rank_hist(sbc_results_single)

# ── Diagnostic issue investigation ───────────────────────────────────────────
backend_diags  <- sbc_results_single$backend_diagnostics
default_diags  <- sbc_results_single$default_diagnostics |>
  dplyr::select(sim_id, max_rhat, min_ess_to_rank)
true_params_wide <- sbc_results_single$stats |>
  dplyr::select(sim_id, variable, simulated_value) |>
  tidyr::pivot_wider(names_from = variable, values_from = simulated_value)

diag_df <- backend_diags |>
  dplyr::left_join(default_diags,    by = "sim_id") |>
  dplyr::left_join(true_params_wide, by = "sim_id") |>
  dplyr::mutate(has_issue = n_divergent > 0 | n_max_treedepth > 0 | max_rhat > 1.01)

p_diag_c_bias <- ggplot(diag_df, aes(x = log_c, y = bias, color = has_issue)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00")) +
  labs(title    = "Diagnostic Issues in Parameter Space",
       subtitle = "Orange points: divergence, max treedepth, or Rhat > 1.01",
       x = "True log(c)", y = "True Bias") +
  theme(legend.position = "bottom")

p_diag_w_c <- ggplot(diag_df, aes(x = `w[1]`, y = log_c, color = has_issue)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00")) +
  labs(title = "Attention vs Sensitivity", x = "True w[1]", y = "True log(c)") +
  theme(legend.position = "none")

print(p_diag_c_bias | p_diag_w_c)

# ── Parameter recovery on SBC simulations ────────────────────────────────────
recovery_df <- sbc_results_single$stats

p_sbc_recovery <- ggplot(recovery_df, aes(x = simulated_value, y = mean)) +
  geom_point(alpha = 0.4, color = "#0072B2") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~variable, scales = "free") +
  labs(
    title    = "Parameter Recovery Across SBC Simulations",
    subtitle = "Posterior mean vs True generated value",
    x = "True Value (simulated_value)",
    y = "Estimated Value (Posterior Mean)"
  ) +
  theme_bw()

print(p_sbc_recovery)
```

When the true log(c) is above 0 the model computes similarity using exponential decay: `exp(-c * distance)`. When c gets large, similarity drops to exactly zero for anything except perfectly identical stimuli, causing gradient collapse and HMC failures. The diagnostic scatter plots above reveal that divergences and Rhat violations cluster at high log(c) values, confirming that the prior mass above log(c) ≈ 2 (c ≈ 7–8) is unsupported by the sampler geometry.

However, calibration and parameter recovery look otherwise good.

**Practical constraint from SBC.** We treat c ≈ 12 as a hard ceiling: a sensitivity this extreme makes stimuli that differ by any measurable amount effectively identical, which is psychologically implausible for the feature ranges used here. We therefore require that the prior places negligible mass above c = 12, i.e., that the 99.85th percentile (mean + 3 SD on the log scale) stays below log(12) ≈ 2.49. For the single-subject model this is satisfied with `Normal(log(2), 1.0)`, whose +3 SD upper bound is exp(0.69 + 3.0) ≈ 20 — already at the boundary of acceptability. We carry this constraint forward explicitly into the multilevel parameterisation.

---

## Extending the GCM with Memory Decay

The canonical GCM weights every stored exemplar equally — a perfect, unbounded memory. That is pedagogically clean but psychologically implausible. Indeed, many early experiments had participants being extensively trained on a set of stimuli, repeated many times, before they actually got tested, so that they would resemble better the model's assumptions. Further, an unbounded memory becomes a real liability as soon as the learning environment is non-stationary: if the category structure changes halfway through the task, the unweighted sum of exemplars anchors the agent to contradictions it has no mechanism to discard. We add a single forgetting-rate parameter, $\lambda$, that downweights older exemplars by $e^{-\lambda (t - t_e)}$ before entering the similarity average:

$$\eta_{ie} = e^{-\lambda (i - t_e)} \cdot e^{-c\, d_{ie}}, \qquad m_{k,i} = \frac{\sum_{e \in C_k} \eta_{ie}}{\sum_{e \in C_k} e^{-\lambda(i - t_e)}}, \qquad p_{i} = \frac{\beta \cdot m_{1,i}}{\beta \cdot m_{1,i} + (1 - \beta) \cdot m_{0,i}}.$$

   * $\eta_{ie}$ (Activation/Similarity): The total strength of the connection
     between the current stimulus $i$ and a stored exemplar $e$. It combines how
     similar they are in "feature space" (spatial) and how recent the memory is
     (temporal).
   * $i$: The current trial index (the "now").
   * $t_e$: The trial index when exemplar $e$ was first encountered and stored
     in memory.
   * $(i - t_e)$: The age of the exemplar—how many trials have passed since it
     was seen.
   * $d_{ie}$: The psychological distance between the current stimulus $i$ and
     the stored exemplar $e$ in feature space.
   * $c$: The sensitivity parameter. It controls how quickly similarity drops as
     distance increases. A high $c$ means only very similar items count; a low
     $c$ means the agent generalizes broadly.
   * $\lambda$: The forgetting-rate (decay) parameter. It controls how quickly
     memories fade with time.
   * $m_{k,i}$: The match (or evidence) for category $k$. This is a weighted
     average of similarities for all exemplars belonging to that category.
   * $C_k$: The set of all exemplars that the agent has previously assigned to
     category $k$.
   * $p_i$: The probability that the agent classifies the current stimulus $i$
     into Category 1.
   * $\beta$: The bias parameter. It represents a baseline preference for
     Category 1 ($\beta > 0.5$) or Category 0 ($\beta < 0.5$) regardless of the
     evidence.


The parameter $\lambda$ governs the Temporal Decay term: $e^{-\lambda (i - t_e)}$.

1. The "Memory Window": Think of $\lambda$ as controlling the size of the agent's moving window of attention. 

  * If $\lambda = 0$: The term $e^{0} = 1$. The age of the memory has no effect. Every exemplar ever seen is weighted equally, representing a "perfect," infinite memory.
       
  * If $\lambda$ is large: The term $e^{-\lambda \cdot age}$ drops to near-zero very quickly as age increases. The agent only "remembers" or cares about the last few trials.

  * The Half-Life: You can calculate the "half-life" of a memory (the number of trials it takes for its influence to drop by half) as $t_{1/2} = \frac{\ln(2)}{\lambda} \approx \frac{0.69}{\lambda}$. For example, if $\lambda = 0.1$, a memory's influence is halved every ~7 trials.

2. Effective Sample Size: The denominator in the $m_{k,i}$ formula, $\sum_{e \in C_k} e^{-\lambda(i - t_e)}$, is the decay-weighted effective sample size. 

  * Instead of just counting how many items are in Category $k$ (which would be the raw count $|C_k|$), the model counts how many "fresh" items are in Category $k$. 
       
  * If a category has 100 exemplars but they were all seen 500 trials ago, and $\lambda$ is high, the "effective count" for that category will be near zero. This prevents old, potentially obsolete evidence from dominating the decision.

3. Why use it? $\lambda$ is essential for non-stationary environments. If the rules of the task change (e.g., the category boundary shifts), an agent with $\lambda=0$ will be stuck trying to reconcile new evidence with a massive mountain of old, contradictory evidence. An agent with a well-tuned $\lambda$  can "forget" the old rules and adapt to the new ones by letting the old evidence fade away.

::: {.callout-note title="Canonical scenario framework (Chapters 11–13)"}
All three categorization chapters test their non-stationarity extension against the same three environmental scenarios, presented in the same order. The ordering reflects **increasing assumption match** between the scenario's data-generating process and the model's structural commitments. The scenarios are defined here once; Chapters 12 and 13 adopt the same definitions and ordering.

| Scenario | What changes in the world | Structural role |
|---|---|---|
| **Static Kruschke** | Nothing — labels are fixed | Baseline: no signal for the non-stationarity parameter |
| **Performance-contingent abrupt shifts** | Labels flip after a streak of $k$ correct responses | Misspecified: abrupt reversals that no smooth-decay / Gaussian-drift / stochastic-mutation mechanism can exactly represent |
| **Smooth drift** | Category centres follow a bivariate Gaussian random walk | Assumption-matched (to varying degrees per model): gradually stale evidence is what all three mechanisms are built to handle |

Comparing the three scenarios across chapters reveals how structural differences between model families show up in their sensitivity to non-stationarity. The synthesis is in the Trilogy Synthesis section at the end of Chapter 13.
:::

We will test this extended model against the three scenarios:

1. **Static Kruschke (baseline).** Fixed category labels, as in the canonical GCM analysis above. A low-$\lambda$ model is correct for this data; we expect the posterior to be pulled toward the left tail of the $\log\lambda$ prior and for the $c$/$\lambda$ pair to be partially non-identified (both can make the similarity average narrower).

2. **Performance-contingent abrupt shifts.** After the agent reaches a streak of $k$ consecutive correct responses, all category labels flip. The agent must unlearn the old mapping and relearn the new one. A high-$\lambda$ model recovers fastest: it forgets the now-contradicted exemplars before they can anchor decisions. Note that these shifts are *endogenous* to the agent's own behaviour — a noisier agent triggers shifts less often and accumulates less information about $\lambda$. Although this scenario is structurally misspecified (the decay GCM has no explicit abrupt-reversal mechanism), the post-flip recovery dynamics impose a strong constraint on $\lambda$ that produces the clearest identifiability signal of the three scenarios.

3. **Smooth drift.** Category centres in feature space drift as an independent random walk. Labels are assigned trial-by-trial by proximity to the current (drifted) centres, so the category boundary migrates smoothly through the stimulus set. This is the **assumption-matched** scenario for the decay GCM: exponentially downweighting past exemplars is a defensible approximation to "evidence collected under a different category mean is stale." Moderate $\lambda$ is appropriate here — agents that forget too slowly lag behind the moving boundary; agents that forget too fast discard still-valid recent evidence.

### Assumption Match Across Scenarios

| Scenario | What drifts in the world | What decay GCM assumes | Match |
|---|---|---|---|
| Static | Nothing | Exponential forgetting (correct at $\lambda \to 0$) | **Information-starved**: $\lambda$ weakly identified |
| Contingent shift | Abrupt label reversals | Smooth exponential forgetting | **Misspecified**: no abrupt mechanism, but post-flip recovery provides strong $\lambda$ signal |
| Smooth drift | Gaussian centre drift | Smooth exponential downweighting of stale exemplars | **Graceful degradation**: approximate but reasonable match |

An important subtlety emerges from this table: *assumption match and parameter identifiability do not always align*. The contingent-shift scenario is more misspecified than smooth drift, yet it produces the cleanest $\lambda$ recovery — because abrupt reversals impose strong temporal constraints that only $\lambda$ can explain. The drift scenario recovers $\lambda$ at the correct calibration rate (SBC passes) but without the "bonus" identifiability that misspecification accidentally provides. We will see both signatures in the validation results below.

### Prior on the Decay Parameter

```{r ch12_decay_prior_constants}
# Prior hyperparameters for log(lambda).
# Normal(-2, 1) gives median lambda ≈ 0.135 (half-life ≈ 5 trials).
# ±2 SD covers lambda ∈ (0.018, 1.0), spanning "effectively no decay"
# to "only the last trial or two matters".
LOG_LAMBDA_PRIOR_MEAN <- -2.0
LOG_LAMBDA_PRIOR_SD   <- 1.0
```

### R Simulator With Scenario Logic

The simulator loops over trials, computes the decay-weighted choice probability, draws a response, and determines the *observed* feedback according to the scenario. Memory stores the trial index at which each exemplar was encoded, so ages are exact at every step. For the contingent-shift scenario, the label-flip state machine updates *after* the response is drawn (so the current trial's feedback uses the label map that was in force at decision time).

```{r ch12_decay_simulator}
simulate_decay_scenario <- function(w, c, lambda, bias, schedule,
                                    scenario = c("static", "contingent_shift", "drift"),
                                    streak_target = 8,
                                    drift_traj    = NULL) {
  scenario <- match.arg(scenario)
  ntrials   <- nrow(schedule)
  obs_mat   <- as.matrix(schedule[, c("height", "position")])
  base_lab  <- schedule$category_feedback

  memory_obs   <- matrix(NA_real_, nrow = 0, ncol = 2)
  memory_cat   <- integer(0)
  memory_trial <- integer(0)

  label_flip <- FALSE
  streak     <- 0L

  prob_cat1         <- numeric(ntrials)
  sim_response      <- integer(ntrials)
  observed_feedback <- integer(ntrials)
  label_flip_trace  <- logical(ntrials)

  for (i in seq_len(ntrials)) {
    current_stim <- as.numeric(obs_mat[i, ])
    n_mem        <- nrow(memory_obs)

    # ── Decay-weighted *mean* similarity per category ───────────────────
    # Numerator:   Σ (decay_j · similarity_j) over exemplars in that category.
    # Denominator: Σ decay_j over exemplars in that category  (effective N).
    # Collapses to plain mean similarity when lambda = 0.
    if (n_mem == 0 || !any(memory_cat == 0) || !any(memory_cat == 1)) {
      p <- bias
    } else {
      d     <- as.numeric(abs(sweep(memory_obs, 2, current_stim, "-")) %*% w)
      age   <- i - memory_trial
      decay <- exp(-lambda * age)
      sim   <- exp(-c * d)
      wt    <- decay * sim
      s1    <- sum(wt[memory_cat == 1]);    s0 <- sum(wt[memory_cat == 0])
      w1    <- sum(decay[memory_cat == 1]); w0 <- sum(decay[memory_cat == 0])
      m1    <- if (w1 > 1e-12) s1 / w1 else 0
      m0    <- if (w0 > 1e-12) s0 / w0 else 0
      num   <- bias * m1
      den   <- num + (1 - bias) * m0
      p     <- if (den > 1e-12) num / den else bias
    }
    p <- pmax(1e-9, pmin(1 - 1e-9, p))
    prob_cat1[i]    <- p
    sim_response[i] <- rbinom(1, 1, p)

    # ── Observed feedback depends on the scenario ───────────────────────
    fb <- switch(scenario,
      static           = base_lab[i],
      contingent_shift = if (label_flip) 1L - base_lab[i] else base_lab[i],
      drift            = {
        d0 <- sum(abs(current_stim - drift_traj$mu0[i, ]))
        d1 <- sum(abs(current_stim - drift_traj$mu1[i, ]))
        as.integer(d1 < d0)
      }
    )
    observed_feedback[i] <- fb
    label_flip_trace[i]  <- label_flip

    # Update the streak/flip state for the contingent scenario
    if (scenario == "contingent_shift") {
      streak <- if (sim_response[i] == fb) streak + 1L else 0L
      if (streak >= streak_target) { label_flip <- !label_flip; streak <- 0L }
    }

    # Add the current trial to memory
    memory_obs   <- rbind(memory_obs, current_stim)
    memory_cat   <- c(memory_cat, fb)
    memory_trial <- c(memory_trial, i)
  }

  schedule |>
    mutate(
      prob_cat1         = prob_cat1,
      sim_response      = sim_response,
      observed_feedback = observed_feedback,
      label_flip        = label_flip_trace,
      correct           = as.integer(sim_response == observed_feedback)
    )
}
```

### Drift Trajectory Generator

The smooth-drift scenario needs a pre-computed trajectory of category centres. Each centre follows an independent bivariate Gaussian random walk with per-trial step SD `drift_sigma`. Defaults place the initial centres at the @kruschke1993human category means; `drift_sigma = 0.05` gives a cumulative SD of about 0.4 units over 64 trials — enough to migrate the boundary through roughly one stimulus cell.

```{r ch12_decay_drift_trajectory}
make_drift_trajectory <- function(ntrials, drift_sigma = 0.05, seed = 1) {
  set.seed(seed)
  mu0_init <- c(1.75, 3.25)   # Kruschke cat-0 mean (height, position)
  mu1_init <- c(3.25, 1.75)   # Kruschke cat-1 mean
  d0 <- apply(matrix(rnorm(ntrials * 2, 0, drift_sigma), ncol = 2), 2, cumsum)
  d1 <- apply(matrix(rnorm(ntrials * 2, 0, drift_sigma), ncol = 2), 2, cumsum)
  list(
    mu0 = sweep(d0, 2, mu0_init, "+"),
    mu1 = sweep(d1, 2, mu1_init, "+")
  )
}
```

1. The "Ground Truth" is a Random Walk
  The function make_drift_trajectory creates two paths ($mu_0$ and $mu_1$). 
   
   * Initial State: On Trial 1, the categories start at the standard Kruschke positions (e.g., Cat 0 is at top-left, Cat 1 is at bottom-right).
   
   * The Drift: Every single trial, both centers take a small random step in a random direction (North, South, East, or West). Because of the cumsum() (cumulative sum), these small steps add up.
   * The Result: By Trial 64, the centers might have wandered quite far from where they started. The "boundary" between the categories—the line where you are equally likely to call something Cat 0 or Cat 1—is constantly shifting.

  2. How Exemplars are Generated
  When the simulation runs, it uses these drifting centers to create the stimuli the agent actually sees:
   1. On Trial $t$, the computer decides to show a "Category 0" stimulus.
   2. It looks up the current coordinates of the center: $mu_{0,t}$.
   3. It generates a stimulus near that point (usually by adding a bit of noise).
   4. The agent sees this stimulus and is told: "This is a Category 0."


### Scenario Agent Wrappers

```{r ch12_decay_agent_wrappers}
simulate_decay_agent <- function(agent_id, scenario, w, c, lambda, bias,
                                 subject_seed, streak_target = 8,
                                 drift_sigma = 0.05) {
  schedule  <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  drift_tr  <- if (scenario == "drift")
    make_drift_trajectory(nrow(schedule), drift_sigma, seed = subject_seed)
  else NULL

  out <- simulate_decay_scenario(
    w = w, c = c, lambda = lambda, bias = bias,
    schedule      = schedule,
    scenario      = scenario,
    streak_target = streak_target,
    drift_traj    = drift_tr
  )

  out |>
    mutate(
      agent_id       = agent_id,
      scenario       = scenario,
      w1_true        = w[1],
      c_true         = c,
      lambda_true    = lambda,
      log_c_true     = log(c),
      log_lambda_true = log(lambda),
      bias_true      = bias
    ) |>
    group_by(agent_id) |>
    mutate(cumulative_accuracy = cumsum(correct) / row_number()) |>
    ungroup()
}
```

### Visualising the Three Scenarios

We simulate 10 agents per cell of a grid that crosses three $\lambda$ values with the three scenarios and plot cumulative accuracy. Static data is where low $\lambda$ wins. The contingent-shift data is where moderate-to-high $\lambda$ wins (faster forgetting speeds reversal recovery, even though the model is misspecified for abrupt shifts). The drift data rewards moderate $\lambda$ tuned to the drift rate — agents that forget too slowly are anchored by obsolete exemplars; agents that forget too fast discard still-valid recent evidence.

```{r ch12_decay_scenario_viz}
decay_viz_file <- here("simdata", "ch12_decay_scenario_viz.csv")

if (regenerate_simulations || !file.exists(decay_viz_file)) {
  viz_grid <- expand_grid(
    scenario    = c("static", "contingent_shift", "drift"),
    lambda_true = c(0.02, 0.2, 0.6),
    agent_id    = 1:10
  ) |>
    mutate(subject_seed = 10000 + row_number())

  viz_sim <- pmap_dfr(
    list(agent_id     = viz_grid$agent_id,
         scenario     = viz_grid$scenario,
         lambda       = viz_grid$lambda_true,
         subject_seed = viz_grid$subject_seed),
    function(agent_id, scenario, lambda, subject_seed) {
      simulate_decay_agent(
        agent_id     = agent_id,
        scenario     = scenario,
        w            = c(0.5, 0.5),
        c            = 2.0,
        lambda       = lambda,
        bias         = 0.5,
        subject_seed = subject_seed
      )
    }
  )
  write_csv(viz_sim, decay_viz_file)
} else {
  viz_sim <- read_csv(decay_viz_file, show_col_types = FALSE)
}

ggplot(viz_sim |> 
         mutate(scenario = factor(scenario, 
                                 levels = c("static", "contingent_shift", "drift"),
                                 labels = c("Static Kruschke", "Performance-contingent flips", "Smooth drift"))),
       aes(x = trial_within_subject, y = cumulative_accuracy,
           color = factor(lambda_true), group = interaction(agent_id, lambda_true))) +
  stat_summary(fun = mean, geom = "line",
               aes(group = factor(lambda_true)), linewidth = 1.1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.15,
               aes(fill = factor(lambda_true), group = factor(lambda_true)),
               color = NA) +
  facet_wrap(~scenario, ncol = 3) +
  scale_color_viridis_d(option = "plasma", name = expression(lambda)) +
  scale_fill_viridis_d(option  = "plasma", name = expression(lambda)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Learning Curves Across the Three Scenarios",
    subtitle = paste0("Mean ± SE across 10 agents per cell; c = 2, bias = 0.5, ",
                      "uniform attention"),
    x = "Trial", y = "Cumulative accuracy"
  ) +
  theme(legend.position = "bottom")
```

**Reading the scenario panels.** In the static panel, accuracy climbs monotonically to near-ceiling for all $\lambda$, with the lowest $\lambda$ reaching the highest asymptote—unsurprising, because any decay discards still-valid evidence. In the drift panel, the curves plateau below ceiling; the plateau is higher for agents with a $\lambda$ tuned to the drift rate, allowing them to track the moving boundary without being "anchored" by obsolete history.

In the contingent-shift panel, we would see a characteristic sawtooth pattern if we were plotting individuals, but this is somewhat muted by the averaging. High-$\lambda$ agents rebound faster after a flip, which means they reach the criterion earlier and flip more often. Because these flips happen at different trial numbers for different agents, averaging across them smooths out the sharp drops and recoveries into a noisy plateau.

**The Alignment Critique.** To see the true effect of $\lambda$ on cognitive flexibility, we must align the data to the moment of the shift. In non-stationary environments, averaging across absolute trial numbers often masks the very process we are interested in—the speed of adaptation.

```{r ch12_decay_reversal_alignment}
# Align contingent-shift data to the start of each new phase
reversal_viz <- viz_sim |>
  filter(scenario == "contingent_shift") |>
  group_by(agent_id, lambda_true) |>
  mutate(
    # Detect when the label mapping flipped
    shift_event = label_flip != lag(label_flip, default = first(label_flip)),
    phase_id    = cumsum(shift_event)
  ) |>
  group_by(agent_id, lambda_true, phase_id) |>
  mutate(
    trial_in_phase = row_number(),
    phase_accuracy = cumsum(correct) / trial_in_phase
  ) |>
  ungroup() |>
  # Focus on the first few reversals to see the learning dynamics
  filter(phase_id > 0, phase_id <= 4, trial_in_phase <= 40)

ggplot(reversal_viz, aes(x = trial_in_phase, y = phase_accuracy, 
                         color = factor(lambda_true), group = factor(lambda_true))) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.1, color = NA) +
  facet_wrap(~paste("Phase", phase_id), ncol = 4) +
  scale_color_viridis_d(option = "plasma", name = expression(lambda)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Learning Performance Aligned to Category Shifts",
    subtitle = "Cumulative accuracy reset at each reversal; contingent-shift scenario only",
    x = "Trials Since Shift", y = "Cumulative Phase Accuracy"
  ) +
  theme(legend.position = "bottom")
```

When aligned this way, the "recovery advantage" of high $\lambda$ becomes stark. While all agents collapse to chance (or below) immediately after a shift, agents with higher decay rates ($\lambda = 0.6$) shed their obsolete memory faster and return to high accuracy significantly earlier than those with low decay.

### The Decay GCM in Stan

The Stan implementation adds `log_lambda` as an unconstrained scalar parameter and multiplies the similarity term by the decay weight `exp(-lambda * age)` where `age = i - past_trial_idx`. We also use a decay-weighted mean similarity to ensure that the relative evidence for each category is properly scaled as old exemplars fade.

```{r ch12_decay_stan_compile}
gcm_decay_single_stan <- "
// Generalized Context Model with Memory Decay — Single Subject
data {
  int<lower=1> N_total;
  int<lower=1> N_features;
  array[N_total] int<lower=0, upper=1> y;
  array[N_total, N_features] real obs;
  array[N_total] int<lower=0, upper=1> cat_feedback;

  vector[N_features] prior_w_alpha;
  real prior_log_c_mu;
  real<lower=0> prior_log_c_sigma;
  real prior_log_lambda_mu;
  real<lower=0> prior_log_lambda_sigma;
  real<lower=0> prior_bias_alpha;
  real<lower=0> prior_bias_beta;
  int<lower=0, upper=1> run_diagnostics;
}

transformed data {
  array[N_total, N_total, N_features] real abs_diff;
  for (i in 1:N_total) {
    for (j in 1:N_total) {
      for (f in 1:N_features) {
        abs_diff[i, j, f] = abs(obs[i, f] - obs[j, f]);
      }
    }
  }
}

parameters {
  simplex[N_features] w;
  real log_c;
  real log_lambda;
  real<lower=0, upper=1> bias;
}

transformed parameters {
  real<lower=0> c      = exp(log_c);
  real<lower=0> lambda = exp(log_lambda);

  vector<lower=1e-9, upper=1-1e-9>[N_total] prob_cat1;

  {
    array[N_total] int memory_trial_idx;
    array[N_total] int memory_cat;
    int n_mem = 0;

    for (i in 1:N_total) {
      real p_i;
      int has_cat0 = 0;
      int has_cat1 = 0;

      for (k in 1:n_mem) {
        if (memory_cat[k] == 0) has_cat0 = 1;
        if (memory_cat[k] == 1) has_cat1 = 1;
      }

      if (n_mem == 0 || has_cat0 == 0 || has_cat1 == 0) {
        p_i = bias;
      } else {
        real s0 = 0.0;   // sum of decay * similarity for cat 0
        real s1 = 0.0;   // sum of decay * similarity for cat 1
        real w0 = 0.0;   // sum of decay weights for cat 0 (effective N)
        real w1 = 0.0;   // sum of decay weights for cat 1 (effective N)
        for (e in 1:n_mem) {
          real d = 0.0;
          int past_i = memory_trial_idx[e];
          for (f in 1:N_features)
            d += w[f] * abs_diff[i, past_i, f];
          real age          = i - past_i;
          real decay_weight = exp(-lambda * age);
          real sim          = exp(-c * d);
          real weight       = decay_weight * sim;
          if (memory_cat[e] == 1) { s1 += weight; w1 += decay_weight; }
          else                    { s0 += weight; w0 += decay_weight; }
        }
        real m1 = (w1 > 1e-12) ? s1 / w1 : 0.0;
        real m0 = (w0 > 1e-12) ? s0 / w0 : 0.0;
        real num = bias * m1;
        real den = num + (1.0 - bias) * m0;
        p_i = (den > 1e-12) ? num / den : bias;
      }

      prob_cat1[i] = fmax(1e-9, fmin(1.0 - 1e-9, p_i));

      n_mem += 1;
      memory_trial_idx[n_mem] = i;
      memory_cat[n_mem] = cat_feedback[i];
    }
  }
}

model {
  target += dirichlet_lpdf(w | prior_w_alpha);
  target += normal_lpdf(log_c      | prior_log_c_mu,      prior_log_c_sigma);
  target += normal_lpdf(log_lambda | prior_log_lambda_mu, prior_log_lambda_sigma);
  target += beta_lpdf(bias         | prior_bias_alpha,    prior_bias_beta);

  target += bernoulli_lpmf(y | prob_cat1);
}

generated quantities {
  vector[N_total] log_lik;
  real lprior;
  
  if (run_diagnostics) {
    for (i in 1:N_total)
      log_lik[i] = bernoulli_lpmf(y[i] | prob_cat1[i]);
  }
  
  lprior =
    dirichlet_lpdf(w         | prior_w_alpha) +
    normal_lpdf(log_c        | prior_log_c_mu,      prior_log_c_sigma) +
    normal_lpdf(log_lambda   | prior_log_lambda_mu, prior_log_lambda_sigma) +
    beta_lpdf(bias           | prior_bias_alpha,    prior_bias_beta);
}
"

stan_file_gcm_decay_single <- "stan/ch12_gcm_decay_single.stan"
write_stan_file(gcm_decay_single_stan, dir = "stan/", basename = "ch12_gcm_decay_single.stan")
mod_gcm_decay_single <- cmdstan_model(stan_file_gcm_decay_single)
```

### Fitting One Agent Per Scenario

We pick one representative agent per scenario with $\lambda = 0.2$, $c = 2$, bias $= 0.5$, and uniform attention, and fit the decay GCM to each.

```{r ch12_decay_fit_per_scenario}
decay_fit_file <- here("simmodels", "ch12_gcm_decay_single_fits.rds")

decay_single_fit_data <- list(
  static           = simulate_decay_agent(1, "static",
                                          w = c(0.5, 0.5), c = 2.0,
                                          lambda = 0.2, bias = 0.5,
                                          subject_seed = 20001),
  contingent_shift = simulate_decay_agent(1, "contingent_shift",
                                          w = c(0.5, 0.5), c = 2.0,
                                          lambda = 0.2, bias = 0.5,
                                          subject_seed = 20003),
  drift            = simulate_decay_agent(1, "drift",
                                          w = c(0.5, 0.5), c = 2.0,
                                          lambda = 0.2, bias = 0.5,
                                          subject_seed = 20002)
)

stan_data_from_agent <- function(agent_df) {
  list(
    N_total               = nrow(agent_df),
    N_features             = 2L,
    y                     = as.integer(agent_df$sim_response),
    obs                   = as.matrix(agent_df[, c("height", "position")]),
    cat_feedback          = as.integer(agent_df$observed_feedback),
    prior_w_alpha         = c(1, 1),
    prior_log_c_mu        = LOG_C_PRIOR_MEAN,
    prior_log_c_sigma     = LOG_C_PRIOR_SD,
    prior_log_lambda_mu   = LOG_LAMBDA_PRIOR_MEAN,
    prior_log_lambda_sigma = LOG_LAMBDA_PRIOR_SD,
    prior_bias_alpha      = 1,
    prior_bias_beta       = 1,
    run_diagnostics       = 1
  )
}

if (regenerate_simulations || !file.exists(decay_fit_file)) {
  decay_fits <- purrr::imap(decay_single_fit_data, function(agent_df, scen) {
    dat <- stan_data_from_agent(agent_df)
    pf  <- tryCatch(
      mod_gcm_decay_single$pathfinder(data = dat, num_paths = 4,
                                      refresh = 0, seed = 500),
      error = function(e) NULL
    )
    fit <- mod_gcm_decay_single$sample(
      data            = dat,
      init            = if (!is.null(pf)) pf else 0.5,
      seed            = 500,
      chains          = 4,
      parallel_chains = min(4, availableCores()),
      iter_warmup     = 1000,
      iter_sampling   = 1500,
      refresh         = 0,
      adapt_delta     = 0.9
    )
    fit$save_object(here("simmodels",
                         paste0("ch12_gcm_decay_single_", scen, ".rds")))
    fit
  })
  saveRDS(lapply(decay_fits, function(f) f$output_files()), decay_fit_file)
  cat("Decay GCM fits computed and saved.\n")
} else {
  decay_fits <- setNames(
    lapply(names(decay_single_fit_data), function(scen) {
      readRDS(here("simmodels",
                   paste0("ch12_gcm_decay_single_", scen, ".rds")))
    }),
    names(decay_single_fit_data)
  )
  cat("Loaded existing decay GCM fits.\n")
}
```

### MCMC Diagnostic Battery Across Scenarios

```{r ch12_decay_diagnostics}
decay_diag_tbls <- purrr::imap_dfr(decay_fits, function(fit, scen) {
  tbl <- diagnostic_summary_table(
    fit,
    params = c("w[1]", "log_c", "log_lambda", "bias")
  )
  tbl |> mutate(scenario = scen)
})

print(decay_diag_tbls |>
        dplyr::select(scenario, metric, value, threshold, pass))

walk(names(decay_fits), function(scen) {
  cat("=== Pair plot:", scen, "===\n")
  print(bayesplot::mcmc_pairs(
    decay_fits[[scen]]$draws(c("log_c", "log_lambda", "w[1]", "bias")),
    diag_fun     = "dens",
    off_diag_fun = "hex"
  ))
})
```

**What to watch for in the pair plot.** A `log_c`/`log_lambda` ridge is expected on static data — both narrow the effective similarity average, so the data under-constrain their difference. On contingent-shift data, post-flip reversal dynamics inject strong temporal evidence that tensions the two parameters against each other, and the ridge should shrink markedly. On drift data, the ridge also narrows (boundary tracking exercises $\lambda$) but typically less sharply, consistent with the weaker identifiability signal that assumption-matched tasks provide.

### Prior-Posterior Update Plots Across Scenarios

```{r ch12_decay_prior_posterior}
decay_pp_df <- purrr::imap_dfr(decay_fits, function(fit, scen) {
  d <- as_draws_df(fit$draws(c("log_c", "log_lambda", "w[1]", "bias")))
  bind_rows(
    tibble(scenario = scen, parameter = "log_c",      value = d$log_c,      source = "posterior"),
    tibble(scenario = scen, parameter = "log_lambda", value = d$log_lambda, source = "posterior"),
    tibble(scenario = scen, parameter = "w[1]",       value = d$`w[1]`,     source = "posterior"),
    tibble(scenario = scen, parameter = "bias",       value = d$bias,       source = "posterior")
  )
})

set.seed(11)
n_prior <- 4000
prior_ref <- bind_rows(
  tibble(parameter = "log_c",      value = rnorm(n_prior, LOG_C_PRIOR_MEAN,      LOG_C_PRIOR_SD),
         source = "prior"),
  tibble(parameter = "log_lambda", value = rnorm(n_prior, LOG_LAMBDA_PRIOR_MEAN, LOG_LAMBDA_PRIOR_SD),
         source = "prior"),
  tibble(parameter = "w[1]",       value = runif(n_prior),         source = "prior"),
  tibble(parameter = "bias",       value = rbeta(n_prior, 1, 1),   source = "prior")
) |>
  tidyr::crossing(scenario = c("static", "contingent_shift", "drift"))

truth_lines <- tibble(
  parameter = c("log_c", "log_lambda", "w[1]", "bias"),
  value     = c(log(2.0), log(0.2), 0.5, 0.5)
)

ggplot(bind_rows(prior_ref, decay_pp_df |> 
                   mutate(scenario = factor(scenario, levels = c("static", "contingent_shift", "drift")))),
       aes(x = value, fill = source, color = source)) +
  geom_density(alpha = 0.4, linewidth = 0.5) +
  geom_vline(data = truth_lines, aes(xintercept = value),
             color = "red", linetype = "dashed") +
  facet_grid(scenario ~ parameter, scales = "free") +
  scale_fill_manual(values  = c(prior = "grey70", posterior = "#009E73")) +
  scale_color_manual(values = c(prior = "grey50", posterior = "#006D4E")) +
  labs(
    title    = "Prior → Posterior Update: Decay GCM Across Scenarios",
    subtitle = "Red dashed line = true generating value used for the single-agent fit",
    x = NULL, y = "Density"
  )
```

### One-Shot Parameter Recovery Across Scenarios

We sweep a 3 × 3 grid of $(c, \lambda)$ truths, with 5 independent simulated datasets per cell per scenario (135 fits total), and report recovery scatters.

```{r ch12_decay_recovery_sweep}
decay_recovery_file <- here("simdata", "ch12_decay_recovery_results.rds")

run_decay_recovery_iter <- function(scenario, c_true, lambda_true, iteration) {
  subject_seed <- iteration * 10000 +
                   round(c_true * 100) +
                   round(lambda_true * 1000) +
                   switch(scenario, static = 1, drift = 2, contingent_shift = 3)

  agent_df <- simulate_decay_agent(
    agent_id     = iteration,
    scenario     = scenario,
    w            = c(0.5, 0.5),
    c            = c_true,
    lambda       = lambda_true,
    bias         = 0.5,
    subject_seed = subject_seed
  )
  dat <- stan_data_from_agent(agent_df)
  pf  <- tryCatch(
    mod_gcm_decay_single$pathfinder(data = dat, num_paths = 4,
                                    refresh = 0, seed = subject_seed),
    error = function(e) NULL
  )
  fit <- tryCatch(
    mod_gcm_decay_single$sample(
      data            = dat,
      init            = if (!is.null(pf)) pf else 0.5,
      seed            = subject_seed,
      chains          = 2,
      parallel_chains = 2,
      iter_warmup     = 800,
      iter_sampling   = 1000,
      refresh         = 0,
      adapt_delta     = 0.9
    ),
    error = function(e) { message("Fit error: ", e$message); NULL }
  )
  if (is.null(fit)) return(NULL)

  s <- fit$summary(variables = c("log_c", "log_lambda", "c", "lambda"))
  diag <- fit$diagnostic_summary(quiet = TRUE)
  tibble(
    scenario    = scenario,
    c_true      = c_true,
    lambda_true = lambda_true,
    iteration   = iteration,
    log_c_mean       = s$mean[s$variable == "log_c"],
    log_lambda_mean  = s$mean[s$variable == "log_lambda"],
    c_mean           = s$mean[s$variable == "c"],
    lambda_mean      = s$mean[s$variable == "lambda"],
    log_c_q5         = s$q5[s$variable == "log_c"],
    log_c_q95        = s$q95[s$variable == "log_c"],
    log_lambda_q5    = s$q5[s$variable == "log_lambda"],
    log_lambda_q95   = s$q95[s$variable == "log_lambda"],
    had_divergence   = any(diag$num_divergent > 0)
  )
}

if (regenerate_simulations || !file.exists(decay_recovery_file)) {
  recovery_plan <- expand_grid(
    scenario    = c("static", "contingent_shift", "drift"),
    c_true      = c(0.5, 2.0, 5.0),
    lambda_true = c(0.02, 0.2, 0.6),
    iteration   = 1:5
  )

  plan(multisession, workers = max(1, availableCores() - 1))
  decay_recovery <- future_pmap_dfr(
    list(
      scenario    = recovery_plan$scenario,
      c_true      = recovery_plan$c_true,
      lambda_true = recovery_plan$lambda_true,
      iteration   = recovery_plan$iteration
    ),
    run_decay_recovery_iter,
    .options = furrr_options(seed = TRUE, scheduling = 1)
  )
  plan(sequential)
  saveRDS(decay_recovery, decay_recovery_file)
  cat("Decay recovery sweep complete.\n")
} else {
  decay_recovery <- readRDS(decay_recovery_file)
  cat("Loaded existing decay recovery sweep.\n")
}

decay_recovery_long <- decay_recovery |>
  dplyr::select(scenario, c_true, lambda_true, iteration, had_divergence,
                log_c_mean, log_lambda_mean) |>
  pivot_longer(
    cols      = c(log_c_mean, log_lambda_mean),
    names_to  = "variable",
    values_to = "estimated"
  ) |>
  mutate(
    parameter = if_else(variable == "log_c_mean", "log_c", "log_lambda"),
    truth     = if_else(parameter == "log_c", log(c_true), log(lambda_true)),
    scenario  = factor(scenario, levels = c("static", "contingent_shift", "drift"))
  )

ggplot(decay_recovery_long, aes(x = truth, y = estimated, color = had_divergence)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.7, size = 2) +
  facet_grid(parameter ~ scenario, scales = "free") +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00"),
                     name = "Had divergence") +
  labs(
    title    = "One-Shot Parameter Recovery for the Decay GCM",
    subtitle = "Posterior mean vs true generating value; 5 iterations per (c, lambda) cell, 3 scenarios",
    x = "True value (log scale)", y = "Estimated value (posterior mean)"
  ) +
  theme(legend.position = "bottom")
```

**Interpretation.** The `log_c` row is near-diagonal across all three scenarios — the sensitivity parameter is well-identified regardless of task structure. The `log_lambda` row is the diagnostic: scatter under static data (the decay parameter is weakly identified, so posterior means regress toward the prior mean), tighter under contingent-shift (post-flip recovery dynamics directly constrain $\lambda$), and well-recovered but more diffuse under smooth drift (the assumption-matched scenario recovers $\lambda$ at the nominal calibration rate, but without the "bonus" constraint the abrupt reversals provide). As the assumption match table above predicts, better assumption match does not automatically mean stronger identifiability — the misspecified contingent-shift task exercises the forgetting mechanism more sharply than the matched drift task does.

### Simulation-Based Calibration Checks Across Scenarios

SBC is the calibration test proper: draw parameters from their priors, simulate data with those parameters, fit the model, and check that the posterior ranks of the true parameter are uniform. We run a separate SBC sweep for each scenario.

```{r ch12_decay_sbc_generator}
make_decay_sbc_generator <- function(scenario) {
  SBC_generator_function(
    function() {
      log_c_true      <- rnorm(1, LOG_C_PRIOR_MEAN,      LOG_C_PRIOR_SD)
      log_lambda_true <- rnorm(1, LOG_LAMBDA_PRIOR_MEAN, LOG_LAMBDA_PRIOR_SD)
      c_true          <- exp(log_c_true)
      lambda_true     <- exp(log_lambda_true)
      w1_true         <- runif(1)
      bias_true       <- rbeta(1, 1, 1)

      subject_seed <- sample.int(1e8, 1)
      agent_df <- simulate_decay_agent(
        agent_id     = 1,
        scenario     = scenario,
        w            = c(w1_true, 1 - w1_true),
        c            = c_true,
        lambda       = lambda_true,
        bias         = bias_true,
        subject_seed = subject_seed
      )

      list(
        variables = list(
          log_c      = log_c_true,
          log_lambda = log_lambda_true,
          `w[1]`     = w1_true,
          bias       = bias_true
        ),
        generated = stan_data_from_agent(agent_df)
      )
    }
  )
}

run_decay_sbc <- function(scenario, n_sims = 200) {
  cache_file <- here("simdata", paste0("ch12_decay_sbc_", scenario, ".rds"))
  if (!regenerate_simulations && file.exists(cache_file)) {
    cat("Loaded cached decay SBC for scenario:", scenario, "\n")
    return(readRDS(cache_file))
  }

  gen <- make_decay_sbc_generator(scenario)
  backend <- SBC_backend_cmdstan_sample(
    mod_gcm_decay_single,
    iter_warmup   = 800,
    iter_sampling = 1000,
    chains        = 2,
    adapt_delta   = 0.9,
    refresh       = 0
  )
  datasets <- generate_datasets(gen, n_sims = n_sims)
  res <- compute_SBC(
    datasets,
    backend,
    cache_mode     = "results",
    cache_location = here("simdata", paste0("ch12_decay_sbc_cache_", scenario)),
    keep_fits      = FALSE
  )
  saveRDS(res, cache_file)
  res
}

decay_sbc_static <- run_decay_sbc("static",           n_sims = 200)
decay_sbc_drift  <- run_decay_sbc("drift",            n_sims = 200)
decay_sbc_shift  <- run_decay_sbc("contingent_shift", n_sims = 200)
```

```{r ch12_decay_sbc_ecdf}
print(plot_ecdf_diff(decay_sbc_static) + ggtitle("SBC: static"))
print(plot_ecdf_diff(decay_sbc_drift)  + ggtitle("SBC: smooth drift"))
print(plot_ecdf_diff(decay_sbc_shift)  + ggtitle("SBC: performance-contingent shifts"))

print(plot_rank_hist(decay_sbc_static) + ggtitle("Rank histograms — static"))
print(plot_rank_hist(decay_sbc_drift)  + ggtitle("Rank histograms — drift"))
print(plot_rank_hist(decay_sbc_shift)  + ggtitle("Rank histograms — contingent shifts"))
```

### Parameter Recovery Across SBC Simulations

```{r ch12_decay_sbc_recovery_plots}
combine_sbc_stats <- function(sbc_obj, scenario) {
  sbc_obj$stats |> mutate(scenario = scenario)
}

decay_sbc_stats <- bind_rows(
  combine_sbc_stats(decay_sbc_static, "static"),
  combine_sbc_stats(decay_sbc_drift,  "drift"),
  combine_sbc_stats(decay_sbc_shift,  "contingent_shift")
) |>
  dplyr::filter(variable %in% c("log_c", "log_lambda", "w[1]", "bias")) |>
  mutate(scenario = factor(scenario, levels = c("static", "contingent_shift", "drift")))

ggplot(decay_sbc_stats, aes(x = simulated_value, y = mean)) +
  geom_point(alpha = 0.35, color = "#0072B2") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_grid(variable ~ scenario, scales = "free") +
  labs(
    title    = "SBC-Level Parameter Recovery: Decay GCM",
    subtitle = "Posterior mean vs true generating value, pooled across SBC simulations",
    x = "True value", y = "Posterior mean"
  )

decay_sbc_stats |>
  group_by(scenario, variable) |>
  summarise(
    sbc_mean_bias = mean(mean - simulated_value, na.rm = TRUE),
    sbc_rmse      = sqrt(mean((mean - simulated_value)^2, na.rm = TRUE)),
    sbc_corr      = cor(mean, simulated_value,
                        use = "complete.obs", method = "pearson"),
    n_sims        = dplyr::n(),
    .groups       = "drop"
  ) |>
  print(n = Inf)
```

### Discussion: Task Structure Governs Identifiability

Under static data, the `log_lambda` parameter cannot really be identified — the posterior does not contain the truth at the nominal rate because the data carry almost no information about forgetting. The `log_c` curve is well-behaved on the same data: the sensitivity parameter is identified regardless. Under contingent-shift and drift data, both parameters recover well, with contingent-shift providing the sharpest $\lambda$ constraint (abrupt reversals) and smooth drift providing the best-matched but less forceful identifiability.

This is the identifiability argument as an empirical claim rather than an algebraic one. The decay GCM's *parameter* is always the same. Its *identifiability* is a joint property of the model and the task: when the task exercises forgetting — whether through a drifting boundary or through performance-contingent reversals — the model recovers $\lambda$ cleanly; when it does not, $\lambda$ is indistinguishable from a slightly wider similarity kernel.

A secondary observation emerges when we look closer at the calibration checks: while parameter recovery (the scatter plot) looks excellent for `log_c` across all scenarios, the SBC rank histograms reveal a subtle "worry" for `log_c` specifically in the drift scenario. This highlights a fundamental distinction in Bayesian validation:

*   **Point Recovery (The "What"):** This asks if the posterior *mean* is close to the truth. If your recovery plot is diagonal, your model is right "on average."
*   **Coverage Calibration (The "How Sure"):** This asks if the posterior *width* (the uncertainty) is honest. If the model is 90% confident, is it actually right 90% of the time?

SBC is designed to catch the latter. In the drift scenario, the model is right about the value of `log_c` (good point recovery), but it is slightly overconfident—it underestimates its own uncertainty. One might wonder: **since the model has $\lambda$, why doesn't the decay parameter just "map" the drift and leave $c$ alone?**

The answer lies in the **Smeared Memory** problem. Even with an optimal $\lambda$, the agent doesn't categorise based on only the single last trial; they need a "window" of recent exemplars to be effective. Because the boundary is drifting *during* that window, the exemplars in memory become spatially "smeared" — a location that was Category 0 five trials ago might be Category 1 now.

While $\lambda$ can tell the model to trust recent trials more, it is a blunt temporal instrument. It cannot resolve the spatial conflict that arises when two different category labels occupy the same coordinates in the recent memory window. The model's likelihood, having no parameter for "environmental volatility," is forced to explain this residual friction using spatial sensitivity ($c$). It performs a form of **attribution error**, interpreting this temporal volatility as spatial overlap and settling on an overly precise estimate of $c$. Because the model is using the wrong "tool" to explain the noise, it converges on a narrow posterior that doesn't cover the true value as often as it should. If we only looked at the recovery plot, we would miss this overconfidence entirely. This is why SBC is the "gold standard" check: it ensures the model isn't just accurate, but also appropriately humble about what it knows.

**How the three scenarios match the model's assumptions.** The decay GCM assumes exemplars decay exponentially as a function of trial age, with no separate mechanism for abrupt reversals or directed drift. The *static* scenario violates nothing but offers no signal for $\lambda$. The *contingent-shift* scenario is a genuine misspecification — the true data-generating process has abrupt reversals that no smooth decay can exactly reproduce — yet it provides the sharpest $\lambda$ signal because post-flip recovery speed depends specifically on how fast old, now-wrong exemplars are discounted. The posterior predictive checks at scenario transitions (not plotted here) should reveal systematic underperformance right after each flip; that residual misfit is the handle for a future extension with an explicit change-point mechanism. The *smooth-drift* scenario is the assumption-matched case: exponentially downweighting past exemplars is a defensible approximation to "evidence collected under a different category mean is stale." SBC passes cleanly here, but $\lambda$ recovery is less sharp than under contingent-shift because drift does not impose the stark temporal discontinuities that most strongly constrain the forgetting rate.

### The $c$/$\lambda$ Joint Posterior: Ridge Geometry and Task Resolution

The pair plots from the single-agent scenario fits revealed a `log_c`/`log_lambda` ridge on static data. Understanding the geometry of this ridge — and why task structure resolves it — is worth making explicit, because Chapter 12 faces a structurally similar problem ($q$/$r$ in the Kalman filter) and resolves it differently.

**Why does the ridge form?** Both $c$ and $\lambda$ narrow the effective contribution of any given exemplar to the category mean:

* A high $c$ shrinks contributions from spatially distant exemplars (similarity decays faster with distance).
* A high $\lambda$ shrinks contributions from temporally distant exemplars (similarity decays faster with trial age).

Under a stationary task with balanced categories, distance and age are nearly uncorrelated — older exemplars are not systematically farther from the current stimulus. This means that increasing $c$ and decreasing $\lambda$ (or vice versa) can produce nearly the same observed choice probabilities, creating the elongated posterior ridge you see in the pair plots.

**Why is the GCM ridge less severe than the Kalman filter's $q$/$r$ ridge?** The two parameters operate on orthogonal dimensions: $c$ operates on the spatial axis (feature distances), $\lambda$ on the temporal axis (trial age). Even under static data, choice probabilities have some leverage on each independently — $c$ is constrained by how performance differs across stimuli at varying spatial distances, while $\lambda$ is partially constrained by the early-learning transient (the model must form initial memories before it can categorize, and the learning rate is $\lambda$-dependent). In the Kalman filter, $q$ and $r$ both scale the *same* likelihood at decision time (the multivariate normal variance at trial $t$ is a function of both), so their directions are more nearly collinear and the ridge is typically more severe. Chapter 12 introduces an explicit reparameterization (the SNR parameter $\rho = q/r$) to address this; the GCM does not need it here.

**Task structure as the primary handle.** The SBC results above confirm that the contingent-shift scenario cleanly resolves $\lambda$ without any reparameterization. The reversal dynamics create a temporal structure that spatial sensitivity cannot account for: the *time* at which performance drops is determined by $\lambda$ (how fast old, now-wrong exemplars are forgotten) in a way that is independent of $c$ (which only governs how much nearby vs. distant exemplars are weighted). This is the key empirical signature of a parameter that is theoretically identifiable but under-constrained by a particular task design: swap in a task that exercises the mechanism, and identifiability follows.

**When would reparameterization be needed for the GCM?** If you needed to run the decay GCM exclusively on static data, the ridge would be a genuine sampling problem. In that case, a reparameterization could help: write $\log s = \log c + \log \lambda$ (the overall "forgetting rate" in combined spatial-temporal units) and $\log d = \log c - \log \lambda$ (the relative weighting of spatial vs. temporal decay). The joint posterior of $(\log s, \log d)$ would have a narrower marginal in the $\log s$ direction (identified by overall accuracy) and a wider marginal in the $\log d$ direction (the degenerate direction).

---

## The Multilevel Decay GCM: Scaling to populations and changing environments

Having demonstrated that memory decay is essential for non-stationary environments and that its parameters are identifiable given the right task structure, we now combine these mechanisms into a full hierarchical model. The **Multilevel Decay GCM** allows us to estimate individual cognitive differences while sharing information across the population to constrain our estimates. This is our "Final Boss" model: population-aware, psychologically realistic, and robust to environmental change.

### Mathematical Formulation: The Hierarchical Architecture

This model uses log-normal hierarchies for sensitivity and forgetting, and logit-normal hierarchies for attention and response bias. With two features, the simplex-valued attention vector $\mathbf{w}_j$ is fully determined by $w_{j,1}$, so the natural simplex hierarchy (Dirichlet around a population mean $\bar{\mathbf{w}}$ with concentration $\kappa$) collapses to a Beta on $w_{j,1}$. We use a **logit-normal** in its place — equivalent in support and shape to the Beta, but its non-centered parameterisation samples much more cleanly than a Dirichlet hierarchy in NUTS, which is what `ch12_gcm_ml_decay.stan` implements.

**Population-level distributions (Priors):**

$$
\begin{aligned}
\bar{\mu}_{\text{logit}\,w_1}, \bar{\sigma}_{\text{logit}\,w_1} &\sim \text{Priors} \\
\bar{\mu}_{\log c}, \bar{\sigma}_{\log c} &\sim \text{Priors} \\
\bar{\mu}_{\log \lambda}, \bar{\sigma}_{\log \lambda} &\sim \text{Priors} \\
\bar{\mu}_{\text{logit}\,\beta}, \bar{\sigma}_{\text{logit}\,\beta} &\sim \text{Priors}
\end{aligned}
$$

**Individual-level parameters (for subject $j$, with $w_{j,2} = 1 - w_{j,1}$):**

$$
\begin{aligned}
\text{logit}(w_{j,1}) &\sim \text{Normal}(\bar{\mu}_{\text{logit}\,w_1}, \bar{\sigma}_{\text{logit}\,w_1}) \\
\log c_j &\sim \text{Normal}(\bar{\mu}_{\log c}, \bar{\sigma}_{\log c}) \\
\log \lambda_j &\sim \text{Normal}(\bar{\mu}_{\log \lambda}, \bar{\sigma}_{\log \lambda}) \\
\text{logit}(\beta_j) &\sim \text{Normal}(\bar{\mu}_{\text{logit}\,\beta}, \bar{\sigma}_{\text{logit}\,\beta})
\end{aligned}
$$

Each individual-level scalar uses a non-centered offset (`z ~ N(0,1)`, with the actual parameter equal to `pop_mean + pop_sd * z`) to avoid the funnel that direct centred parameterisations produce when subject-level data are weakly informative.

**Trial-level Generative Process (for trial $i$ of subject $j$):**

$$
\begin{aligned}
d_{iej} &= \sum_f w_{jf} |x_{if} - x_{ef}| \\
\eta_{iej} &= e^{-\lambda_j(i - t_e)} \cdot e^{-c_j d_{iej}} \\
m_{k,ij} &= \frac{\sum_{e \in C_k} \eta_{iej}}{\sum_{e \in C_k} e^{-\lambda_j(i - t_e)}} \\
p_{ij} &= \frac{\beta_j \cdot m_{1,ij}}{\beta_j \cdot m_{1,ij} + (1 - \beta_j) \cdot m_{0,ij}} \\
y_{ij} &\sim \text{Bernoulli}(p_{ij})
\end{aligned}
$$

*   **Shrinkage:** The model uses the population means to "pull" uncertain individual estimates toward the group average. This is especially useful for $\lambda_j$ in scenarios where an individual's data might be locally uninformative.
*   **Effective Sample Size:** The denominator of $m_{k,ij}$ is now the **decay-weighted effective number of exemplars**, ensuring that old, forgotten evidence doesn't inflate the mean similarity.

### Simulating Data for the Multilevel Decay GCM

We simulate three small populations of agents — one per scenario (static, smooth drift, performance-contingent shifts) — using the **same** population-level parameters across scenarios. Subject-level parameters are drawn around the population means using the logit-normal/log-normal hierarchies that match the Stan model exactly. Because the simulator and the inference model now share a parameterisation, recovery scatters can be read directly without translating between Dirichlet concentration and logit SD.

```{r ch12_simulate_multilevel_decay_data}
simulate_multilevel_decay_gcm <- function(n_subjects, stimulus_info, n_blocks,
                                          scenario,
                                          base_seed,
                                          pop_logit_w_mean,    pop_logit_w_sd,
                                          pop_log_c_mean,      pop_log_c_sd,
                                          pop_log_lambda_mean, pop_log_lambda_sd,
                                          pop_logit_bias_mean, pop_logit_bias_sd) {
  set.seed(base_seed)
  ind_logit_w    <- rnorm(n_subjects, pop_logit_w_mean,    pop_logit_w_sd)
  ind_w1         <- plogis(ind_logit_w)
  ind_log_c      <- rnorm(n_subjects, pop_log_c_mean,      pop_log_c_sd)
  ind_log_lambda <- rnorm(n_subjects, pop_log_lambda_mean, pop_log_lambda_sd)
  ind_logit_bias <- rnorm(n_subjects, pop_logit_bias_mean, pop_logit_bias_sd)

  true_params <- tibble(
    agent_id        = seq_len(n_subjects),
    scenario        = scenario,
    logit_w_true    = ind_logit_w,
    w1_true         = ind_w1,
    w2_true         = 1 - ind_w1,
    log_c_true      = ind_log_c,
    c_true          = exp(ind_log_c),
    log_lambda_true = ind_log_lambda,
    lambda_true     = exp(ind_log_lambda),
    logit_bias_true = ind_logit_bias,
    bias_true       = plogis(ind_logit_bias)
  )

  all_agent_data <- vector("list", n_subjects)
  for (j in seq_len(n_subjects)) {
    subject_seed <- base_seed * 1000 + j
    all_agent_data[[j]] <- simulate_decay_agent(
      agent_id     = j,
      scenario     = scenario,
      w            = c(ind_w1[j], 1 - ind_w1[j]),
      c            = exp(ind_log_c[j]),
      lambda       = exp(ind_log_lambda[j]),
      bias         = plogis(ind_logit_bias[j]),
      subject_seed = subject_seed
    )
  }

  list(
    data        = bind_rows(all_agent_data),
    true_params = true_params
  )
}

n_subjects_sim <- 10
pop_params_sim <- list(
  pop_logit_w_mean    = 0,        # median w1 ≈ 0.5
  pop_logit_w_sd      = 0.7,      # ±2SD covers w1 ∈ (0.20, 0.80)
  pop_log_c_mean      = log(2.0), # population median c = 2
  pop_log_c_sd        = 0.4,
  pop_log_lambda_mean = log(0.2), # population median λ = 0.2
  pop_log_lambda_sd   = 0.5,
  pop_logit_bias_mean = 0,
  pop_logit_bias_sd   = 0.3
)

ml_scenarios <- c("static", "contingent_shift", "drift")

ml_decay_files <- map(ml_scenarios, function(scen) list(
  data  = here("simdata", paste0("ch12_gcm_ml_decay_data_",  scen, ".csv")),
  truth = here("simdata", paste0("ch12_gcm_ml_decay_truth_", scen, ".csv"))
)) |> setNames(ml_scenarios)

ml_files_exist <- all(file.exists(unlist(map(ml_decay_files, ~ c(.x$data, .x$truth)))))

if (regenerate_simulations || !ml_files_exist) {
  ml_decay_sims <- map(ml_scenarios, function(scen) {
    out <- simulate_multilevel_decay_gcm(
      n_subjects          = n_subjects_sim,
      stimulus_info       = stimulus_info,
      n_blocks            = n_blocks,
      scenario            = scen,
      base_seed           = match(scen, ml_scenarios) * 1000 + 42,
      pop_logit_w_mean    = pop_params_sim$pop_logit_w_mean,
      pop_logit_w_sd      = pop_params_sim$pop_logit_w_sd,
      pop_log_c_mean      = pop_params_sim$pop_log_c_mean,
      pop_log_c_sd        = pop_params_sim$pop_log_c_sd,
      pop_log_lambda_mean = pop_params_sim$pop_log_lambda_mean,
      pop_log_lambda_sd   = pop_params_sim$pop_log_lambda_sd,
      pop_logit_bias_mean = pop_params_sim$pop_logit_bias_mean,
      pop_logit_bias_sd   = pop_params_sim$pop_logit_bias_sd
    )
    write_csv(out$data,        ml_decay_files[[scen]]$data)
    write_csv(out$true_params, ml_decay_files[[scen]]$truth)
    out
  }) |> setNames(ml_scenarios)
} else {
  ml_decay_sims <- map(ml_scenarios, function(scen) list(
    data        = read_csv(ml_decay_files[[scen]]$data,  show_col_types = FALSE),
    true_params = read_csv(ml_decay_files[[scen]]$truth, show_col_types = FALSE)
  )) |> setNames(ml_scenarios)
}

ml_decay_all <- map_dfr(ml_scenarios, function(scen)
  ml_decay_sims[[scen]]$data |> mutate(scenario = scen)
) |>
  mutate(scenario = factor(scenario, levels = ml_scenarios,
                           labels = c("Static", "Smooth drift",
                                      "Performance-contingent")))

ggplot(ml_decay_all,
       aes(x = trial_within_subject, y = cumulative_accuracy, group = agent_id)) +
  geom_line(alpha = 0.25, color = "grey30") +
  stat_summary(fun = mean, geom = "line", aes(group = 1),
               color = "#0072B2", linewidth = 1.4) +
  facet_wrap(~scenario, ncol = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Multilevel Decay GCM: simulated learning curves across scenarios",
    subtitle = paste0(n_subjects_sim, " agents per scenario; thin grey = individuals, ",
                      "thick blue = population mean"),
    x = "Trial", y = "Cumulative accuracy"
  )
```

The static panel converges quickly to ceiling; smooth drift plateaus below ceiling because the boundary keeps moving; performance-contingent shifts produce a noisy plateau as different subjects flip at different absolute trials and average over each other.

### Stan Implementation

The Stan model implements the hierarchy described above, with three architectural choices that matter for performance:

* **The trial-by-trial likelihood lives in the `model` block, not in `transformed parameters`.** Memory is held in plain `int` arrays during the loop, so the autodiff tape never sees the per-trial scan. `prob_cat1` and (optionally gated) `log_lik` are recomputed once per saved draw in `generated quantities`.
* **All hierarchies are non-centered** (logit-normal NCP for `w_1`; standard z-offset for `log_c`, `log_lambda`, and `logit_bias`).
* **Pairwise feature distances are precomputed in `transformed data`**, indexed by `stim_id` instead of read from per-trial features.

```{r ch12_ml_decay_stan_compile}
gcm_ml_decay_stan <- "
// Multilevel Decay Generalized Context Model
// All individual parameters use non-centred parameterisation:
//   - w_1 via logit-normal NCP (eliminates Dirichlet funnel geometry)
//   - log_c, log_lambda, logit_bias via standard z-offset NCP
// Sequential likelihood is computed directly in the model block (no
// prob_cat1 stored in transformed parameters), avoiding a large var-vector
// on the autodiff tape. prob_cat1 and log_lik are recomputed once per saved
// draw in generated quantities.

data {
  int<lower=1> N_total;
  int<lower=1> N_subjects;
  int<lower=1> N_features;
  int<lower=1> N_unique_stim;
  int<lower=1> max_trials_per_subject;

  array[N_subjects] int<lower=1, upper=N_total> subj_start;
  array[N_subjects] int<lower=1, upper=N_total> subj_end;

  array[N_total] int<lower=1, upper=N_unique_stim> stim_id;
  array[N_total] int<lower=0, upper=1> y;
  array[N_unique_stim, N_features] real unique_obs;
  array[N_total] int<lower=0, upper=1> cat_feedback;

  // Hyperparameters
  real prior_pop_logit_w_mean;
  real<lower=0> prior_pop_logit_w_sigma;
  real<lower=0> prior_pop_logit_w_sd_lambda;
  real prior_pop_log_c_mu;
  real<lower=0> prior_pop_log_c_sigma;
  real<lower=0> prior_pop_log_c_lambda;
  real prior_pop_log_lambda_mu;
  real<lower=0> prior_pop_log_lambda_sigma;
  real<lower=0> prior_pop_log_lambda_lambda;
  real prior_pop_logit_bias_mu;
  real<lower=0> prior_pop_logit_bias_sigma;
  real<lower=0> prior_pop_logit_bias_lambda;
  int<lower=0, upper=1> run_diagnostics;
}

transformed data {
  array[N_unique_stim, N_unique_stim, N_features] real stimulus_abs_diff;
  for (i in 1:N_unique_stim)
    for (k in 1:N_unique_stim)
      for (f in 1:N_features)
        stimulus_abs_diff[i, k, f] = abs(unique_obs[i, f] - unique_obs[k, f]);
}

parameters {
  // Population attention weight: logit-normal NCP for w_1 (w_2 = 1 - w_1)
  real pop_logit_w_mean;
  real<lower=0> pop_logit_w_sd;
  vector[N_subjects] z_w;

  // Population c, lambda, bias
  real pop_log_c_mean;
  real<lower=0> pop_log_c_sd;
  real pop_log_lambda_mean;
  real<lower=0> pop_log_lambda_sd;
  real pop_logit_bias_mean;
  real<lower=0> pop_logit_bias_sd;

  // Non-centred individual offsets
  vector[N_subjects] z_log_c;
  vector[N_subjects] z_log_lambda;
  vector[N_subjects] z_logit_bias;
}

transformed parameters {
  vector[N_subjects] subj_log_c;
  vector[N_subjects] subj_log_lambda;
  vector<lower=0, upper=1>[N_subjects] subj_bias;
  array[N_subjects] vector[N_features] subj_w;

  for (j in 1:N_subjects) {
    subj_log_c[j]      = pop_log_c_mean + pop_log_c_sd * z_log_c[j];
    subj_log_lambda[j] = pop_log_lambda_mean + pop_log_lambda_sd * z_log_lambda[j];
    subj_bias[j]       = inv_logit(pop_logit_bias_mean + pop_logit_bias_sd * z_logit_bias[j]);
    real logit_w       = pop_logit_w_mean + pop_logit_w_sd * z_w[j];
    subj_w[j][1]       = inv_logit(logit_w);
    subj_w[j][2]       = 1.0 - subj_w[j][1];
  }
}

model {
  // Population priors
  pop_logit_w_mean    ~ normal(prior_pop_logit_w_mean, prior_pop_logit_w_sigma);
  pop_logit_w_sd      ~ exponential(prior_pop_logit_w_sd_lambda);
  pop_log_c_mean      ~ normal(prior_pop_log_c_mu, prior_pop_log_c_sigma);
  pop_log_c_sd        ~ exponential(prior_pop_log_c_lambda);
  pop_log_lambda_mean ~ normal(prior_pop_log_lambda_mu, prior_pop_log_lambda_sigma);
  pop_log_lambda_sd   ~ exponential(prior_pop_log_lambda_lambda);
  pop_logit_bias_mean ~ normal(prior_pop_logit_bias_mu, prior_pop_logit_bias_sigma);
  pop_logit_bias_sd   ~ exponential(prior_pop_logit_bias_lambda);

  // Non-centred priors
  z_w          ~ std_normal();
  z_log_c      ~ std_normal();
  z_log_lambda ~ std_normal();
  z_logit_bias ~ std_normal();

  // Sequential likelihood — local workspace, one subject at a time
  {
    array[max_trials_per_subject] int mem_stim;
    array[max_trials_per_subject] int mem_cat;
    array[max_trials_per_subject] int mem_trial;

    for (j in 1:N_subjects) {
      int n_mem  = 0;
      real c      = exp(subj_log_c[j]);
      real lambda = exp(subj_log_lambda[j]);
      real bias   = subj_bias[j];

      for (i in subj_start[j]:subj_end[j]) {
        real p_i;
        if (n_mem == 0) {
          p_i = bias;
        } else {
          real s1 = 0.0; real s0 = 0.0;
          real w1 = 0.0; real w0 = 0.0;
          int curr_stim = stim_id[i];
          for (e in 1:n_mem) {
            real dist = 0.0;
            int past_stim = mem_stim[e];
            for (f in 1:N_features)
              dist += subj_w[j][f] * stimulus_abs_diff[curr_stim, past_stim, f];
            real age   = (i - subj_start[j] + 1.0) - mem_trial[e];
            real decay = exp(-lambda * age);
            real sim   = exp(-c * dist);
            real wt    = decay * sim;
            if (mem_cat[e] == 1) { s1 += wt; w1 += decay; }
            else                 { s0 += wt; w0 += decay; }
          }
          real m1  = (w1 > 1e-9) ? s1 / w1 : 0.0;
          real m0  = (w0 > 1e-9) ? s0 / w0 : 0.0;
          real num = bias * m1;
          real den = num + (1.0 - bias) * m0;
          p_i = (den > 1e-9) ? num / den : bias;
        }
        target += bernoulli_lpmf(y[i] | fmax(1e-9, fmin(1.0 - 1e-9, p_i)));

        n_mem += 1;
        mem_stim[n_mem]  = stim_id[i];
        mem_cat[n_mem]   = cat_feedback[i];
        mem_trial[n_mem] = i - subj_start[j] + 1;
      }
    }
  }
}

generated quantities {
  // Recompute once per saved draw (not on every leapfrog step).
  // prob_cat1 is always filled (needed for PPC); log_lik only when run_diagnostics=1.
  vector[N_total] prob_cat1;
  vector[N_total] log_lik = rep_vector(0.0, N_total);

  {
    array[max_trials_per_subject] int mem_stim;
    array[max_trials_per_subject] int mem_cat;
    array[max_trials_per_subject] int mem_trial;

    for (j in 1:N_subjects) {
      int n_mem  = 0;
      real c      = exp(subj_log_c[j]);
      real lambda = exp(subj_log_lambda[j]);
      real bias   = subj_bias[j];

      for (i in subj_start[j]:subj_end[j]) {
        real p_i;
        if (n_mem == 0) {
          p_i = bias;
        } else {
          real s1 = 0.0; real s0 = 0.0;
          real w1 = 0.0; real w0 = 0.0;
          int curr_stim = stim_id[i];
          for (e in 1:n_mem) {
            real dist = 0.0;
            int past_stim = mem_stim[e];
            for (f in 1:N_features)
              dist += subj_w[j][f] * stimulus_abs_diff[curr_stim, past_stim, f];
            real age   = (i - subj_start[j] + 1.0) - mem_trial[e];
            real decay = exp(-lambda * age);
            real sim   = exp(-c * dist);
            real wt    = decay * sim;
            if (mem_cat[e] == 1) { s1 += wt; w1 += decay; }
            else                 { s0 += wt; w0 += decay; }
          }
          real m1  = (w1 > 1e-9) ? s1 / w1 : 0.0;
          real m0  = (w0 > 1e-9) ? s0 / w0 : 0.0;
          real num = bias * m1;
          real den = num + (1.0 - bias) * m0;
          p_i = (den > 1e-9) ? num / den : bias;
        }
        real p_clamp  = fmax(1e-9, fmin(1.0 - 1e-9, p_i));
        prob_cat1[i]  = p_clamp;
        if (run_diagnostics)
          log_lik[i]  = bernoulli_lpmf(y[i] | p_clamp);

        n_mem += 1;
        mem_stim[n_mem]  = stim_id[i];
        mem_cat[n_mem]   = cat_feedback[i];
        mem_trial[n_mem] = i - subj_start[j] + 1;
      }
    }
  }
}
"

stan_file_gcm_ml_decay <- "stan/ch12_gcm_ml_decay.stan"
write_stan_file(gcm_ml_decay_stan, dir = "stan/", basename = "ch12_gcm_ml_decay.stan")
mod_gcm_ml_decay       <- cmdstan_model(stan_file_gcm_ml_decay)
```

### Building the Stan Data List

```{r ch12_ml_decay_stan_data_builder}
build_ml_stan_data <- function(ml_data, stimulus_info, run_diagnostics = 0L) {
  ml_data <- ml_data |>
    arrange(agent_id, trial_within_subject) |>
    mutate(row = row_number())

  unique_stim <- stimulus_info |>
    arrange(stimulus) |>
    mutate(unique_idx = row_number())
  stim_lookup <- setNames(unique_stim$unique_idx, unique_stim$stimulus)

  subj_idx <- ml_data |>
    group_by(agent_id) |>
    summarise(start = min(row), end = max(row), .groups = "drop") |>
    arrange(agent_id)

  list(
    N_total                = nrow(ml_data),
    N_subjects             = nrow(subj_idx),
    N_features             = 2L,
    N_unique_stim          = nrow(unique_stim),
    max_trials_per_subject = as.integer(max(subj_idx$end - subj_idx$start + 1L)),

    subj_start             = as.integer(subj_idx$start),
    subj_end               = as.integer(subj_idx$end),

    stim_id                = as.integer(stim_lookup[as.character(ml_data$stimulus_id)]),
    y                      = as.integer(ml_data$sim_response),
    unique_obs             = as.matrix(unique_stim[, c("height", "position")]),
    cat_feedback           = as.integer(ml_data$observed_feedback),

    prior_pop_logit_w_mean      = 0,
    prior_pop_logit_w_sigma     = 1.0,
    prior_pop_logit_w_sd_lambda = 2.0,
    prior_pop_log_c_mu          = LOG_C_PRIOR_MEAN,
    prior_pop_log_c_sigma       = LOG_C_PRIOR_SD,
    prior_pop_log_c_lambda      = 2.0,
    prior_pop_log_lambda_mu     = LOG_LAMBDA_PRIOR_MEAN,
    prior_pop_log_lambda_sigma  = LOG_LAMBDA_PRIOR_SD,
    prior_pop_log_lambda_lambda = 2.0,
    prior_pop_logit_bias_mu     = 0,
    prior_pop_logit_bias_sigma  = 1.0,
    prior_pop_logit_bias_lambda = 2.0,

    run_diagnostics             = as.integer(run_diagnostics)
  )
}
```

The `Exponential(2)` priors on the population SDs (`*_lambda = 2`) have median 0.35 and 95th percentile ≈ 1.5 — wide enough to allow real between-subject variation, tight enough to keep NUTS off the worst part of the funnel when only 10 subjects are observed. We keep `run_diagnostics = 0` for the main fits because we will not run LFO at the multilevel level here; flip it to `1` only when computing per-trial `log_lik` is needed.

### Fitting the Multilevel Decay GCM Across Scenarios

We compile once and fit each scenario in turn. With 10 subjects × 64 trials × 4 chains × 2000 iterations, expect each fit to take a few minutes on a modern laptop.

```{r ch12_ml_decay_fit_per_scenario}
ml_decay_fit_files <- map(ml_scenarios, function(scen)
  here("simmodels", paste0("ch12_gcm_ml_decay_fit_", scen, ".rds"))
) |> setNames(ml_scenarios)

if (regenerate_simulations || !all(file.exists(unlist(ml_decay_fit_files)))) {
  ml_decay_fits <- map(ml_scenarios, function(scen) {
    cat("Fitting multilevel decay GCM:", scen, "\n")
    dat <- build_ml_stan_data(ml_decay_sims[[scen]]$data, stimulus_info,
                              run_diagnostics = 0L)
    fit <- mod_gcm_ml_decay$sample(
      data            = dat,
      seed            = 700 + match(scen, ml_scenarios),
      chains          = 4,
      parallel_chains = min(4, availableCores()),
      iter_warmup     = 1000,
      iter_sampling   = 1000,
      adapt_delta     = 0.95,
      max_treedepth   = 12,
      refresh         = 200,
      init            = 0.5
    )
    fit$save_object(ml_decay_fit_files[[scen]])
    fit
  }) |> setNames(ml_scenarios)
  cat("Multilevel decay GCM fits computed and saved.\n")
} else {
  ml_decay_fits <- map(ml_scenarios, function(scen)
    readRDS(ml_decay_fit_files[[scen]])
  ) |> setNames(ml_scenarios)
  cat("Loaded existing multilevel decay GCM fits.\n")
}
```

### MCMC Diagnostic Battery

```{r ch12_ml_decay_diagnostics}
ml_decay_diag <- map_dfr(ml_scenarios, function(scen) {
  diagnostic_summary_table(
    ml_decay_fits[[scen]],
    params = c("pop_logit_w_mean",    "pop_log_c_mean",
               "pop_log_lambda_mean", "pop_logit_bias_mean",
               "pop_logit_w_sd",      "pop_log_c_sd",
               "pop_log_lambda_sd",   "pop_logit_bias_sd")
  ) |> mutate(scenario = scen)
})

print(ml_decay_diag |>
        dplyr::select(scenario, metric, value, threshold, pass))
```

If the model is healthy you will see no divergences, R̂ < 1.01 across all population parameters, and ESS comfortably above 400 for the population means. Subject-level scalars typically have somewhat lower ESS, especially for `log_lambda` under the static scenario — the data carry little information about forgetting when the boundary never moves, so the posterior is partly pinned by the prior and the partial pooling toward the population mean.

### Population-Level Recovery

```{r ch12_ml_decay_population_recovery}
pop_truth <- with(pop_params_sim, tibble(
  variable = c("pop_logit_w_mean", "pop_log_c_mean",
               "pop_log_lambda_mean", "pop_logit_bias_mean"),
  truth    = c(pop_logit_w_mean, pop_log_c_mean,
               pop_log_lambda_mean, pop_logit_bias_mean)
))

# Extract all posterior draws for the population means
pop_draws <- map_dfr(ml_scenarios, function(scen) {
  as_draws_df(ml_decay_fits[[scen]]$draws(pop_truth$variable)) |>
    pivot_longer(cols = all_of(pop_truth$variable), 
                 names_to = "variable", values_to = "value") |>
    mutate(scenario = scen)
}) |>
  mutate(scenario = factor(scenario, levels = ml_scenarios,
                           labels = c("Static", "Drift", "Contingent shift")),
         variable = factor(variable, levels = pop_truth$variable))

# Precompute prior densities for overlay
prior_params <- tibble(
  variable = factor(pop_truth$variable, levels = pop_truth$variable),
  mu    = c(0, LOG_C_PRIOR_MEAN, LOG_LAMBDA_PRIOR_MEAN, 0),
  sigma = c(1, LOG_C_PRIOR_SD, LOG_LAMBDA_PRIOR_SD, 1)
)

prior_density <- prior_params |>
  group_by(variable) |>
  reframe(
    x = seq(mu - 3.5*sigma, mu + 3.5*sigma, length.out = 200),
    y = dnorm(x, mu, sigma)
  )

ggplot() +
  # Prior distribution
  geom_line(data = prior_density, aes(x = x, y = y), 
            linetype = "dotted", color = "black", linewidth = 0.8) +
  # Posterior distributions per scenario
  geom_density(data = pop_draws, aes(x = value, fill = scenario), 
               alpha = 0.4, color = NA) +
  # Simulation truth
  geom_vline(data = pop_truth, aes(xintercept = truth), 
             linetype = "dashed", color = "red", linewidth = 0.8) +
  facet_wrap(~variable, scales = "free", ncol = 2) +
  scale_fill_viridis_d(option = "plasma", end = 0.85) +
  labs(
    title    = "Population-level parameter recovery",
    subtitle = "Posterior densities vs Simulation Truth (red dashed) and Prior (black dotted)",
    x = "Parameter value (transformed scale)", y = "Density"
  ) +
  theme(legend.position = "bottom")
```


### Subject-Level Recovery and Shrinkage

```{r ch12_ml_decay_subject_recovery}
subj_recovery <- map_dfr(ml_scenarios, function(scen) {
  fit   <- ml_decay_fits[[scen]]
  truth <- ml_decay_sims[[scen]]$true_params

  s <- fit$summary(variables = c("subj_log_c", "subj_log_lambda")) |>
    mutate(
      param    = if_else(stringr::str_detect(variable, "log_c"),
                         "log_c", "log_lambda"),
      agent_id = as.integer(stringr::str_extract(variable, "\\d+"))
    ) |>
    dplyr::select(agent_id, param, est_mean = mean, est_q5 = q5, est_q95 = q95)

  truth_long <- truth |>
    dplyr::select(agent_id, log_c_true, log_lambda_true) |>
    pivot_longer(cols = c(log_c_true, log_lambda_true),
                 names_to = "param", values_to = "truth") |>
    mutate(param = if_else(param == "log_c_true", "log_c", "log_lambda"))

  s |>
    left_join(truth_long, by = c("agent_id", "param")) |>
    mutate(scenario = scen)
}) |>
  mutate(scenario = factor(scenario, levels = ml_scenarios,
                           labels = c("Static", "Drift", "Contingent shift")))

ggplot(subj_recovery, aes(x = truth, y = est_mean, color = scenario)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = est_q5, ymax = est_q95), alpha = 0.7) +
  facet_grid(param ~ scenario, scales = "free") +
  scale_color_viridis_d(option = "plasma", end = 0.85) +
  labs(
    title    = "Subject-level parameter recovery (with partial pooling)",
    subtitle = "Posterior mean ± 90% CI vs simulation truth",
    x = "True value", y = "Posterior estimate"
  ) +
  theme(legend.position = "none")
```

The `log_c` row should be tight along the identity across scenarios. The `log_lambda` row is the diagnostic: scattered and pulled toward the population mean under static (visible shrinkage when an individual's data is uninformative), tighter along the diagonal under drift, and tightest under contingent-shift — consistent with the single-subject recovery results from earlier in the chapter.

### Multilevel Simulation-Based Calibration (SBC)

Multilevel SBC is the most rigorous way to check that the hierarchical priors and the non-centered parameterisation are working as intended. Because fitting a multilevel model hundreds of times is computationally intensive, we provide the code here as a template for students wishing to perform an exhaustive validation of their own hierarchical extensions.

```{r ch12_ml_decay_sbc, eval=FALSE}
# 1. Generator: Draws population parameters from priors, then subject 
#    parameters from those hierarchies, and finally simulates trial data.
make_ml_decay_sbc_generator <- function(scenario, n_subjects = 5) {
  SBC_generator_function(
    function() {
      # Draw population hyperparameters from the priors
      pop_logit_w_mean    <- rnorm(1, 0, 1)
      pop_logit_w_sd      <- rexp(1, 2)
      pop_log_c_mean      <- rnorm(1, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD)
      pop_log_c_sd        <- rexp(1, 2)
      pop_log_lambda_mean <- rnorm(1, LOG_LAMBDA_PRIOR_MEAN, LOG_LAMBDA_PRIOR_SD)
      pop_log_lambda_sd   <- rexp(1, 2)
      pop_logit_bias_mean <- rnorm(1, 0, 1)
      pop_logit_bias_sd   <- rexp(1, 2)
      
      # Simulate one multilevel dataset
      sim <- simulate_multilevel_decay_gcm(
        n_subjects          = n_subjects,
        stimulus_info       = stimulus_info,
        n_blocks            = 2, # Fewer blocks for faster SBC
        scenario            = scenario,
        base_seed           = sample.int(1e6, 1),
        pop_logit_w_mean    = pop_logit_w_mean,
        pop_logit_w_sd      = pop_logit_w_sd,
        pop_log_c_mean      = pop_log_c_mean,
        pop_log_c_sd        = pop_log_c_sd,
        pop_log_lambda_mean = pop_log_lambda_mean,
        pop_log_lambda_sd   = pop_log_lambda_sd,
        pop_logit_bias_mean = pop_logit_bias_mean,
        pop_logit_bias_sd   = pop_logit_bias_sd
      )
      
      # Return parameters and the list for Stan
      list(
        parameters = list(
          pop_logit_w_mean    = pop_logit_w_mean,
          pop_log_c_mean      = pop_log_c_mean,
          pop_log_lambda_mean = pop_log_lambda_mean,
          pop_logit_bias_mean = pop_logit_bias_mean
        ),
        generated = build_ml_stan_data(sim$data, stimulus_info)
      )
    }
  )
}

# 2. Backend: Uses the compiled multilevel Stan model
ml_decay_sbc_backend <- SBC_backend_cmdstan_sample(
  mod_gcm_ml_decay,
  iter_warmup = 800, iter_sampling = 800, chains = 2
)

# 3. Compute SBC (Uncomment to run)
# ml_sbc_results <- compute_SBC(
#   make_ml_decay_sbc_generator("drift", n_subjects = 5),
#   ml_decay_sbc_backend,
#   n_sims = 50,
#   cache_location = here("simdata", "ch12_ml_decay_sbc_cache")
# )

# 4. Visualize Results
# plot_rank_hist(ml_sbc_results)
# plot_ecdf_diff(ml_sbc_results)
```

### What changes when you scale up

Three things to keep in mind when adapting this model to real data:

1. **Sample size enters twice.** With more subjects, the population SDs become better identified, which shrinks subject-level estimates more aggressively toward the population mean — desirable when individual-level data is noisy, but worth checking visually.
2. **Sequential-trial costs scale linearly.** Doubling trials per subject doubles per-leapfrog cost; doubling subjects doubles it again. Going from this 10×64 toy to a 100×200 dataset is a ~30× increase in wall time, not 30 different fits to manage.
3. **Static-scenario behaviour is the canary.** If `pop_log_lambda_mean` and `pop_log_lambda_sd` start to diverge or drag at `max_treedepth` on real data, your task is probably structurally closer to "static" than you think — there is not enough non-stationarity for $\lambda$ to be identified at the population level. Reduce the model or accept that $\lambda$ is pinned by the prior and report it as such.



### GCM in Context: Strengths and Limitations

#### Strengths

* **Empirical support**: The GCM has successfully fit human categorization data across numerous experiments.
* **Psychological plausibility**: The model's parameters correspond to psychologically meaningful constructs (attention, sensitivity/generalization, bias, and now forgetting).
* **Flexibility**: It can account for a wide range of category structures and learning phenomena, including non-stationary environments.
* **Individual differences**: The multilevel framework allows principled modeling of variations across people.

#### Limitations

* **Memory requirements**: Storing all exemplars becomes computationally demanding for large numbers of stimuli. How memory is bounded over long timescales is not fully specified.
* **Conflation of generalisation and decisional precision**: The standard GCM has no separate softmax temperature; sensitivity $c$ doubles as a generalisation parameter and as a decisional precision parameter.
* **Complex categories**: The standard GCM struggles with categories defined by abstract rules, relations, or logical combinations of features.
* **Sequential structure and LOO**: Because the GCM is path-dependent, standard LOO-CV is technically invalid — but this chapter implements LFO-CV (Bürkner et al., 2020) as the canonical sequential-model cross-validation.

In the next two chapters we explore alternative approaches — prototype models (Ch. 12) and rule-based models (Ch. 13) — that address some of these limitations while introducing their own strengths and challenges.

### Looking Ahead to Chapter 12: The Six-Step Pipeline in the Prototype Model

Chapter 12 applies the same six-step pipeline to a prototype-based model. Before moving on, it is worth tracing explicitly how each stage maps from the GCM to the Kalman filter prototype model — because the chapters are designed as a matched pair, and the contrasts at each stage are the core theoretical content.

| Pipeline Stage | GCM (this chapter) | Kalman Filter Prototype (Ch. 12) |
| :--- | :--- | :--- |
| **1. Input representation** | Stimuli as points in continuous feature space (height, position). | Same feature space; stimuli used as Kalman observations. |
| **2. Attention** | Explicit attention weights $w_f$ summing to 1; estimated from data. | No explicit attention weights; covariance structure implicitly weights dimensions. |
| **3. Intermediate representation** | **All** past exemplars stored verbatim; memory grows trial by trial. | One prototype $\vec{\mu}_k$ per category, updated by each observation; memory is constant-size. |
| **4. Evidential mechanism** | Similarity to each stored exemplar, exponentially decaying with distance and (in the extended model) trial age. | Mahalanobis-like distance from the current stimulus to each prototype, scaled by the running covariance $\Sigma_k$. |
| **5. Decision rule** | Luce choice axiom: $p(\text{cat}_1) = \frac{\beta \cdot \bar{\eta}_1}{\beta \cdot \bar{\eta}_1 + (1-\beta) \cdot \bar{\eta}_0}$. | Same Luce choice axiom; the per-trial probability is computed from the Kalman-filter likelihood ratio. |
| **6. Learning** | Accumulate: each trial's exemplar enters permanent (but decaying) memory. No structural update. | Update: the Kalman gain $K$ shifts $\vec{\mu}_k$ toward the new observation and reduces $\Sigma_k$; the prediction step (process noise $Q$) can re-inflate $\Sigma_k$ between trials. |

The key tension is at **Stage 3 and Stage 6**. The GCM treats every trial as permanently informative (modulo forgetting); the prototype model treats each trial as an observation that refines a running summary. As a result:

* The GCM is maximally sensitive to specific exemplars and their spatial arrangement. The prototype model is sensitive only to the category centroid and spread.
* The GCM's non-stationarity mechanism (decay $\lambda$) operates *retrospectively* — old exemplars become less influential as time passes. The Kalman filter's mechanism (process noise $q$) operates *prospectively* — future observations are expected to diverge from the current prototype.
* Both mechanisms are phenomenologically similar under smooth drift. They diverge most sharply under abrupt shifts: the GCM gradually forgets old contradicting exemplars, while the Kalman filter re-inflates uncertainty after surprising observations and quickly reacquires the new prototype.

::: {.callout-note title="What to watch for when reading Chapter 12"}
1. **The $q$/$r$ identifiability problem.** The Kalman filter's two parameters ($r$ = observation noise, $q$ = process noise) are more tightly confounded under static data than the GCM's $c$/$\lambda$ pair — they both scale the *same* decision-time likelihood rather than operating on orthogonal axes (see §"The $c$/$\lambda$ Joint Posterior" above). Ch. 12 introduces an explicit reparameterization (the SNR parameter $\rho = q/r$) to diagnose and partially resolve this.
2. **The scenario order is deliberately reversed.** Ch. 12 arranges scenarios as static → contingent shift → drift (increasing assumption match for the Kalman filter), while this chapter used static → drift → contingent shift (increasing $\lambda$ preference). The three scenarios are identical; only the ordering logic differs.
3. **No attention weights.** Removing $w$ and $\beta$ from the parameter set makes the prototype model simpler to diagnose — SBC rank histograms have two curves instead of four, and pair plots have one panel instead of six. That simplicity is *why* Ch. 12 is the right place to introduce LFO-CV in detail: the algorithm is most legible when the model is lean.
:::

---
