# Prototype Models: Learning Category Averages

```{r ch13_setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)

# Set to TRUE to rerun simulations and Stan fits.
# FALSE loads pre-computed results from 'simdata/'.
regenerate_simulations <- TRUE

# Set to TRUE to run the long-running validation chunks
# (LFO-CV, precision analysis, SBC). FALSE loads cached results
# or skips the chunk if no cache exists.
run_intensive_checks <- regenerate_simulations

pacman::p_load(
  tidyverse,    # data manipulation and visualization
  here,         # robust project-relative paths
  mvtnorm,      # multivariate normal densities
  patchwork,    # combining plots
  cmdstanr,     # Stan interface
  posterior,    # tidy posterior arrays
  tidybayes,    # tidy extraction of draws
  bayesplot,    # MCMC diagnostic plots
  loo,          # LOO-CV and PSIS
  SBC,          # Simulation-Based Calibration
  priorsense,   # prior sensitivity
  future,       # parallel processing
  furrr,        # parallel functional programming
  ellipse,      # uncertainty ellipses
  matrixStats,  # logSumExp for the LFO-CV implementation
  MASS          # ginv (pseudo-inverse fallback)
)

theme_set(theme_classic())

# Prior hyperparameters for log(r) and log(q)
LOG_R_PRIOR_MEAN <- 0          # prior median r = 1
LOG_R_PRIOR_SD   <- 1.0        # covers r ≈ 0.13–7.4 at ±2 SD

LOG_Q_PRIOR_MEAN <- -2         # prior median q ≈ 0.14 (small drift)
LOG_Q_PRIOR_SD   <- 1.0        # covers q ≈ 0.02–1.0 at ±2 SD

for (d in c("stan", "simdata", "simmodels")) {
  if (!dir.exists(d)) dir.create(d)
}

# ── Shared diagnostic helper (canonical version lives in Ch 11) ──────────────
# We re-define it here for standalone use. The function returns a one-row
# tibble summarising the mandatory MCMC diagnostic battery for a cmdstanr
# fit, and prints a warning if any threshold is breached. Chapters 11, 12,
# 13, and 15 all call this so the diagnostic table looks the same everywhere.
diagnostic_summary_table <- function(fit, pars = NULL) {
  diag <- fit$diagnostic_summary(quiet = TRUE)
  draws_summary <- fit$summary(variables = pars)

  # Calculate individual metrics
  n_div <- sum(diag$num_divergent)
  max_rhat <- round(max(draws_summary$rhat, na.rm = TRUE), 2)
  min_bulk <- round(min(draws_summary$ess_bulk, na.rm = TRUE), 0)
  min_tail <- round(min(draws_summary$ess_tail, na.rm = TRUE), 0)
  min_ebfmi <- round(min(diag$ebfmi), 3)

  # Robust MCSE calculation
  mcse_col <- intersect(names(draws_summary), c("mcse_mean", "mcse", "mcse_cp"))[1]
  if (!is.na(mcse_col)) {
    max_mcse <- round(max(draws_summary[[mcse_col]] / draws_summary$sd, na.rm = TRUE), 4)
  } else {
    max_mcse <- round(max(1 / sqrt(draws_summary$ess_bulk), na.rm = TRUE), 4)
  }

  # Build the formatted table
  out <- tibble::tibble(
    metric = c(
      "Divergences (zero tolerance)",
      "Max rank-normalised R-hat",
      "Min bulk ESS",
      "Min tail ESS",
      "Min E-BFMI",
      "Max MCSE / posterior SD"
    ),
    value = c(n_div, max_rhat, min_bulk, min_tail, min_ebfmi, max_mcse),
    threshold = c("== 0", "< 1.01", "> 400", "> 400", "> 0.2", "< 0.05"),
    pass = c(
      n_div == 0,
      max_rhat < 1.01,
      min_bulk > 400,
      min_tail > 400,
      min_ebfmi > 0.2,
      max_mcse < 0.05
    )
  )

  # Check if any threshold failed
  if (!all(out$pass)) {
    warning(
      "MCMC diagnostic battery: at least one threshold breached. ",
      "Inspect divergences and reparameterize before interpreting the posterior.",
      call. = FALSE
    )
  }

  out
}

# ── Kruschke (1993) stimulus set ─────────────────────────────────────────────
# Re-defined here for standalone use; identical to Ch. 11.
stimulus_info <- tibble(
  stimulus      = c(5, 3, 7, 1, 8, 2, 6, 4),
  height        = c(1, 1, 2, 2, 3, 3, 4, 4),
  position      = c(2, 3, 1, 4, 1, 4, 2, 3),
  category_true = c(0, 0, 1, 0, 1, 0, 1, 1)
)

n_blocks         <- 8
n_stim_per_block <- nrow(stimulus_info)
total_trials     <- n_blocks * n_stim_per_block

# ── Per-subject schedule helper ───────────────────────────────────────────────
# Identical to Ch. 11. Redefined here for standalone use.
make_subject_schedule <- function(stimulus_info, n_blocks, seed) {
  set.seed(seed)
  n_stim <- nrow(stimulus_info)
  sequence <- unlist(lapply(seq_len(n_blocks), function(b) {
    sample(stimulus_info$stimulus, n_stim, replace = FALSE)
  }))
  tibble(
    trial_within_subject = seq_along(sequence),
    block                = rep(seq_len(n_blocks), each = n_stim),
    stimulus_id          = sequence
  ) |>
    left_join(stimulus_info, by = c("stimulus_id" = "stimulus")) |>
    rename(category_feedback = category_true) |>
    dplyr::select(trial_within_subject, block, stimulus_id,
                  height, position, category_feedback)
}

plan(multisession, workers = max(1, availableCores() - 1))
```

> **📍 Where we are in the Bayesian modeling workflow:**
> Chapter 11 built the GCM — an exemplar model that stores every encountered
> stimulus and decides by summed similarity to memory. This chapter introduces
> a structurally different account: the **prototype model**, which abstracts
> each category to a single running estimate (its mean and uncertainty) rather
> than storing individual instances. We implement prototype learning as a
> multivariate Kalman filter and then add a **Stan parameter-estimation model**
> for the prototype architecture.
>
> The full validation pipeline from Ch. 5 — prior
> predictive check, Pathfinder initialization, mandatory MCMC diagnostic
> battery, posterior predictive check, prior–posterior update plot, randomised
> LOO-PIT, LFO-CV, precision analysis, prior sensitivity,
> and SBC — is applied to this new model. Crucially, the **LFO-CV machinery
> implemented in Chapter 11** (`psis_lfo_gcm`) is generalised here as
> `psis_lfo_kalman` and applied to the prototype model. The chapter then
> extends to a **multilevel prototype model** with population-level priors
> on `log_r`, per-subject NCP offsets, and the full multilevel SBC battery
> (population and individual calibration).

# Prototype-Based Models of Categorization

> **📋 Validation battery roadmap for this chapter**
>
> | 0. Conceptual interrogation | |
> | 1. Forward simulator (single) | `prototype_kalman()` and `simulate_prototype_agent()` |
> | 2. Stan implementation (single) | §"The Prototype Model in Stan" — `prob` in `transformed parameters` |
> | 3.0 MCMC diagnostic battery | §"Mandatory MCMC Diagnostic Battery" — uses `diagnostic_summary_table()` |
> | 3.2 Prior predictive | §"Prior Predictive Check" |
> | 3.3 Posterior predictive | §"Posterior Predictive Check" |
> | 3.4 Prior–posterior update | §"Prior–Posterior Update Plot" |
> | 3.5 Prior sensitivity | §"Prior Sensitivity Analysis" — `priorsense::powerscale_sensitivity()` |
> | 4.5 LOO-PIT (randomised, binary) | §"Randomised LOO-PIT Calibration" |
> | 4.6 LFO-CV | §"Leave-Future-Out Cross-Validation" — inherited from Ch 11 |
> | 6b. SBC (single-subject) | §"Simulation-Based Calibration" |
> | 6d. Precision analysis | §"Precision Analysis: How CI Width Scales With N" |
> | Identifiability check | § (r vs decisional temperature) |
> | 1ML. Forward simulator (multilevel) | `simulate_multilevel_prototype()` |
> | 2ML. Stan implementation (multilevel) | §"Implementing the Multilevel Prototype Model in Stan" — NCP over `log_r` and `log_q` |
> | 3ML. MCMC diagnostics (multilevel) | §"Mandatory MCMC Diagnostic Battery (Multilevel)" |
> | 6cML. Multilevel SBC (pop + individual) | §"Multilevel SBC" — ECDF-diff + recovery for `pop_log_r_mean`, `pop_log_r_sd`, `pop_log_q_mean`, `pop_log_q_sd`, `subj_log_r[1]`, `subj_log_q[1]` |

## Rethinking Category Representation: From Examples to Averages

In the previous chapter, we explored the Generalized Context Model (GCM), an exemplar model where every encountered example is stored in memory. This approach is powerful but raises questions about cognitive efficiency. Do we really store every single instance we encounter?

Consider learning about common categories like "birds" or "chairs". While specific examples matter, we also seem to develop a general sense of what constitutes a typical member. Prototype theory offers an alternative perspective: instead of storing individual exemplars, the mind abstracts a summary representation — a prototype — for each category. This prototype often represents the central tendency or "average" member.

## Core Ideas of Prototype Models

* **Abstraction, Not Storage**: The primary representation is a summary (the prototype), not a collection of instances.
* **Cognitive Economy**: Storing one prototype per category is potentially much more memory-efficient than storing numerous exemplars.
* **Similarity to Prototype**: New items are categorized based on their similarity to the stored prototypes of different categories.
* **Typicality**: Items closer to a category's prototype are judged as more "typical" members and are often categorized faster and more accurately. This contrasts with exemplar models, which emphasize detailed memory for specific instances.

| Feature | Prototype Models | Exemplar Models (GCM) |
|---|---|---|
| Representation | Abstract summary (e.g., average) | Collection of individual examples |
| Memory Cost | Low (one prototype per category) | High (potentially all examples) |
| Key Information | Central tendency & variance | Specific instances & their labels |
| Sensitivity | Less sensitive to specific examples | Highly sensitive to specific examples |

## Modeling Dynamic Prototypes: The Kalman Filter

Early prototype models often calculated a static average. However, human learning is dynamic; our understanding evolves with experience. How can we model a prototype that updates incrementally as new examples are encountered?

The Kalman filter provides a suitable mathematical framework. Originally developed for tracking physical systems amidst noise, it's well-suited for modeling prototype learning because it allows us to:

* **Track an Estimate**: The category prototype (its average feature values).
* **Represent Uncertainty**: Maintain not just the prototype's location but also our uncertainty about it.
* **Update Incrementally**: Refine the prototype with each new example.
* **Balance Old and New**: Optimally combine the current prototype (prior belief) with the new example (evidence).
* **Implement an Adaptive Learning Rate**: Automatically adjust how much influence a new example has (the "Kalman gain") based on current uncertainty. Learning is faster when uncertainty is high and slows as confidence grows.

### The Kalman Filter Logic: Updating an Estimate

Imagine tracking a prototype based on a single feature (e.g., average height). The Kalman filter maintains your Current State. At any moment, you have two key pieces of information:

* **Your Best Guess ($\mu$):** This is your current estimate of the prototype's average height.
* **Your Confidence ($\sigma^2$):** This isn't just *if* you're confident, but *how much*. It's the *variance* around your guess. A small variance means you're quite sure; a large variance means you're very unsure.

When a new category member (observation $x$) arrives, you measure its height.

* **The Catch:** Your measurement tool (or the example itself) isn't perfect. There's some inherent "noise" or variability in how examples relate to the true prototype. We represent the uncertainty of this *measurement process* with another variance, often called $R$.

**The Core Problem:** You have your guess (with $\sigma^2_{prev}$) and a new measurement (with $R$). How do you combine them to get a *new, better guess* and an *updated level of uncertainty*?

**Enter the Kalman Gain ($K$): The "Trust" Knob**

The Kalman filter calculates the Kalman Gain $K$, which tells you how much to trust the *new measurement* compared to your *current guess*:

$$K = \frac{\sigma^2_{prev}}{\sigma^2_{prev} + R}$$

If $\sigma^2_{prev}$ is large relative to $R$, then $K \to 1$: trust the measurement almost completely. If $\sigma^2_{prev}$ is small relative to $R$, then $K \to 0$: stick closer to your previous guess.

**Updating Your Guess:**

You nudge your old guess towards the new measurement. The size of the nudge is determined by the Kalman Gain $K$:

$$\mu_{new} = \mu_{prev} + K \cdot (x - \mu_{prev})$$

The term $(x - \mu_{prev})$ is the "prediction error" or "innovation" — how much the new data differs from what you expected.

**Updating Your Confidence (Uncertainty):**

After incorporating new information, uncertainty decreases:

$$\sigma^2_{new} = (1 - K) \cdot \sigma^2_{prev}$$

Since $K \in (0, 1)$, multiplying by $(1 - K)$ always shrinks uncertainty. The more you trusted the measurement (higher $K$), the more your uncertainty shrinks.

**Extending to drifting targets.** The scalar Kalman filter above assumes a static target: uncertainty monotonically decreases because there is no mechanism for it to grow back. The prototype model in this chapter adds a **process noise term** $Q$. Before each measurement update, a *prediction step* is applied:

$$\sigma^2_{k|k-1} = \sigma^2_{k-1|k-1} + Q$$

This increases uncertainty between observations, reflecting the possibility that the true prototype has shifted since the last trial. The subsequent measurement update then decreases uncertainty via the Kalman gain, exactly as before. At steady state, the two forces balance and uncertainty converges to a fixed value rather than shrinking to zero. When $Q = 0$ the model reduces to the static-target filter. We treat the scalar $q$ (the diagonal of $Q = q \cdot I$) as a free parameter to be estimated from data, alongside the observation noise $r$.

### Extending to Multiple Features: Multivariate Kalman Filter

Our stimuli have two features (height and position). We need a multivariate version:

* **Your Current State (Vectors and Matrices):**
    * **Your Best Guess ($\vec{\mu}$):** Now a *vector* containing your best guess for each feature.
    * **Your Confidence ($\Sigma$):** Now a *covariance matrix*. The diagonal elements give the variance for each feature individually; the off-diagonal elements give their covariance — for example, whether taller prototypes are also typically further to the right.

* **A New Hint Arrives:** Your new observation ($\vec{x}$) is also a vector.
    * **Measurement Noise ($R$):** Also a covariance matrix. Often simplified to a diagonal matrix, assuming measurement errors on different features are independent.

* **The Logic Stays the Same, the Math Uses Matrices:**
    * **Kalman Gain ($K$):** A *matrix*: $K = \Sigma_{prev} (\Sigma_{prev} + R)^{-1}$
    * **Updating Your Guess ($\vec{\mu}$):** $\vec{\mu}_{new} = \vec{\mu}_{prev} + K (\vec{x} - \vec{\mu}_{prev})$
    * **Updating Your Confidence ($\Sigma$):** We use the numerically stable Joseph form: $\Sigma_{new} = (I - K)\Sigma_{prev}(I - K)^T + KRK^T$. While it looks complex, the intuition is the same as the scalar case: uncertainty decreases after each update. The Joseph form is preferred because it guarantees the result remains symmetric and positive semi-definite even in the presence of floating-point rounding errors.

### In a Nutshell

The Kalman filter is a recursive algorithm that provides the mathematically optimal way to estimate the state of a system (like the features of a prototype) given a series of noisy measurements. It works by:

1. Maintaining an estimate ($\vec{\mu}$) and uncertainty ($\Sigma$).
2. Calculating a Kalman Gain $K$ that balances trust in the current estimate against trust in the new measurement.
3. Updating the estimate by moving it towards the new measurement, weighted by $K$.
4. Updating (reducing) the uncertainty based on the information gained.

The multivariate version simply uses vectors and matrices to do this for multiple, potentially correlated variables simultaneously.

### Implementing the Multivariate Update in R

```{r ch13_multivariate_kalman_update}
# One update step of a multivariate Kalman filter.
# Returns updated mean vector, covariance matrix, and Kalman gain.
multivariate_kalman_update <- function(mu_prev,      # previous mean vector
                                       sigma_prev,    # previous covariance matrix
                                       observation,   # observed feature vector
                                       r_matrix) {    # observation noise matrix
  mu_prev     <- as.numeric(mu_prev)
  observation <- as.numeric(observation)
  sigma_prev  <- as.matrix(sigma_prev)
  r_matrix    <- as.matrix(r_matrix)

  n_dim <- length(mu_prev)
  I     <- diag(n_dim)

  # Kalman gain: K = Sigma_prev * (Sigma_prev + R)^{-1}
  S     <- sigma_prev + r_matrix
  S_inv <- tryCatch(
    solve(S),
    error = function(e) {
      warning("Matrix inversion failed; using pseudo-inverse.", call. = FALSE)
      MASS::ginv(S)
    }
  )
  K <- sigma_prev %*% S_inv

  # Update mean
  innovation <- observation - mu_prev
  mu_new     <- as.numeric(mu_prev + K %*% innovation)

  # Update covariance — Joseph form for numerical stability
  IK        <- I - K
  sigma_new <- IK %*% sigma_prev %*% t(IK) + K %*% r_matrix %*% t(K)

  # Enforce symmetry (floating-point can introduce tiny asymmetry)
  sigma_new <- (sigma_new + t(sigma_new)) / 2

  list(mu = mu_new, sigma = sigma_new, k = K)
}
```

This function performs a single update for one category's prototype when it observes a new member.

---

We are fitting a *two-parameter* model: $r$ controls both Kalman gain and decisional precision; $q$ controls prototype drift speed and steady-state uncertainty. Both are inferred from data on the log scale. There is no category bias, fixed initial means, fixed initial covariances, and diagonal noise structures for both $R$ and $Q$. The Stan implementation in §"The Prototype Model in Stan" implements exactly this.

---


## Building the Full Categorization Model

We now integrate this update mechanism into a full agent that learns prototypes for two categories and makes categorization decisions.

### Initialization

* We need separate prototypes ($\mu_0, \Sigma_0$ and $\mu_1, \Sigma_1$) for Category 0 and Category 1.
* We start with high uncertainty (large initial $\Sigma$) and uninformative means at the center of the feature space.
* We define the observation noise matrix $R = r_\text{val} \cdot I$. The parameter $r_\text{val}$ represents how much variability we assume exists within categories or in our perception of stimuli.
* We define the process noise matrix $Q = q_\text{val} \cdot I$. The parameter $q_\text{val}$ controls how much the category prototype may drift between successive observations. Small $q$ means near-static categories; large $q$ means rapidly shifting ones.

### Categorization Decision

How does the model decide the category for a new stimulus $\vec{x}$? It compares $\vec{x}$ to both prototypes, considering their current means ($\vec{\mu}$) and uncertainties ($\Sigma$). We use the probability density of the multivariate normal distribution:

$$P(\vec{x} \mid \text{Category } C) \propto \exp\left(-\frac{1}{2}(\vec{x} - \vec{\mu}_C)^T (\Sigma_C + R)^{-1} (\vec{x} - \vec{\mu}_C)\right)$$

The probability of choosing Category 1 is then calculated using the Luce choice rule (softmax):

$$P(\text{Choose Cat 1} \mid \vec{x}) = \frac{P(\vec{x} \mid \text{Category } 1)}{P(\vec{x} \mid \text{Category } 0) + P(\vec{x} \mid \text{Category } 1)}$$

We work with log-probabilities for numerical stability. Note that the same $r$ that drives the Kalman update also enters here through $(\Sigma_C + R)^{-1}$ — there is no separate softmax temperature.

### The Learning Loop

The agent processes trials sequentially:

1. **Predict**: Add $Q$ to both category covariances — $\Sigma_{0} \mathrel{+}= Q$, $\Sigma_{1} \mathrel{+}= Q$ — reflecting that each prototype may have drifted since the last observation.
2. **Decide**: Calculate $P(\text{Choose Cat 1} \mid \vec{x}_i)$ based on the post-prediction $\mu_0, \Sigma_0, \mu_1, \Sigma_1$.
3. **Respond**: Generate a response.
4. **Receive feedback**: Observe true category $y_i$.
5. **Update**: Apply the Kalman measurement update to the correct category only.

### R Implementation of the Prototype Agent

The function now accepts `subject_seed` and calls `make_subject_schedule` internally — consistent with the path-dependence fix from Ch. 11. Each agent's learning trajectory depends on the order in which stimuli are encountered, and different randomizations produce genuinely different trajectories.

```{r ch13_prototype_model}
# Prototype model agent using Kalman filter with process noise for categorization.
# r_value   : observation noise variance (scalar > 0)
# q_value   : process noise variance / prototype drift rate (scalar >= 0)
# obs       : matrix of observations (trials x features)
# cat_one   : vector of true category labels (0 or 1) for feedback
# Returns:  tibble with prob_cat1 and sim_response
prototype_kalman <- function(r_value,
                             q_value,
                             obs,
                             cat_one,
                             initial_mu         = NULL,
                             initial_sigma_diag = 10.0,
                             quiet              = TRUE) {
  n_trials   <- nrow(obs)
  n_features <- ncol(obs)

  if (is.null(initial_mu)) {
    mu0_init <- rep(2.5, n_features)
    mu1_init <- rep(2.5, n_features)
  } else {
    mu0_init <- initial_mu[[1]]
    mu1_init <- initial_mu[[2]]
  }

  prototype_cat_0 <- list(mu = mu0_init, sigma = diag(initial_sigma_diag, n_features))
  prototype_cat_1 <- list(mu = mu1_init, sigma = diag(initial_sigma_diag, n_features))
  r_matrix        <- diag(r_value, n_features)
  q_matrix        <- diag(q_value, n_features)

  response_probs <- numeric(n_trials)

  log_sum_exp <- function(v) {
    m <- max(v)
    m + log(sum(exp(v - m)))
  }

  for (i in seq_len(n_trials)) {
    if (!quiet && i %% 20 == 0) cat("Trial", i, "\n")

    current_obs <- as.numeric(obs[i, ])

    # ── Prediction step: add process noise to both prototypes ──────────────
    # Reflects potential drift in category location since last trial.
    # When q_value = 0 this reduces to the static-target filter.
    prototype_cat_0$sigma <- prototype_cat_0$sigma + q_matrix
    prototype_cat_1$sigma <- prototype_cat_1$sigma + q_matrix

    # ── Decision ─────────────────────────────────────────────────────────────
    cov_cat_0 <- prototype_cat_0$sigma + r_matrix
    cov_cat_1 <- prototype_cat_1$sigma + r_matrix

    log_prob_0 <- tryCatch(
      mvtnorm::dmvnorm(current_obs, mean = prototype_cat_0$mu,
                       sigma = cov_cat_0, log = TRUE),
      error = function(e) -Inf
    )
    log_prob_1 <- tryCatch(
      mvtnorm::dmvnorm(current_obs, mean = prototype_cat_1$mu,
                       sigma = cov_cat_1, log = TRUE),
      error = function(e) -Inf
    )

    if (!is.finite(log_prob_0) && !is.finite(log_prob_1)) {
      prob_cat_1 <- 0.5
    } else if (!is.finite(log_prob_0)) {
      prob_cat_1 <- 1.0
    } else if (!is.finite(log_prob_1)) {
      prob_cat_1 <- 0.0
    } else {
      prob_cat_1 <- exp(log_prob_1 - log_sum_exp(c(log_prob_0, log_prob_1)))
    }
    response_probs[i] <- pmax(1e-9, pmin(1 - 1e-9, prob_cat_1))

    # ── Update (measurement update for the correct category only) ────────────
    true_cat <- cat_one[i]
    if (true_cat == 1) {
      upd <- multivariate_kalman_update(prototype_cat_1$mu, prototype_cat_1$sigma,
                                        current_obs, r_matrix)
      prototype_cat_1$mu    <- upd$mu
      prototype_cat_1$sigma <- upd$sigma
    } else {
      upd <- multivariate_kalman_update(prototype_cat_0$mu, prototype_cat_0$sigma,
                                        current_obs, r_matrix)
      prototype_cat_0$mu    <- upd$mu
      prototype_cat_0$sigma <- upd$sigma
    }
  }

  tibble(
    prob_cat1    = response_probs,
    sim_response = rbinom(n_trials, 1, response_probs)
  )
}

# Wrapper: generates per-agent schedule then calls prototype_kalman
simulate_prototype_agent <- function(agent_id, r_value, q_value,
                                     stimulus_info, n_blocks, subject_seed) {
  schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  obs      <- as.matrix(schedule[, c("height", "position")])
  cat_one  <- schedule$category_feedback

  result <- prototype_kalman(
    r_value            = r_value,
    q_value            = q_value,
    obs                = obs,
    cat_one            = cat_one,
    initial_mu         = list(rep(2.5, 2), rep(2.5, 2)),
    initial_sigma_diag = 10.0,
    quiet              = TRUE
  )

  schedule |>
    mutate(
      agent_id     = agent_id,
      r_value_true = r_value,
      q_value_true = q_value,
      log_r_true   = log(r_value),
      log_q_true   = log(q_value),
      prob_cat1    = result$prob_cat1,
      sim_response = result$sim_response,
      correct      = as.integer(category_feedback == sim_response)
    ) |>
    group_by(agent_id) |>
    mutate(performance = cumsum(correct) / row_number()) |>
    ungroup()
}
```

### Simulating Categorization Behavior

Let's simulate behavior on the Kruschke (1993) task. The key parameter here is `r_value`, the observation noise. This parameter reflects the model's assumption about how much variability exists within categories or arises from perceptual noise.

* **Low `r_value`**: The model assumes observations are precise representations of the category; each new example exerts strong influence and prototypes update rapidly.
* **High `r_value`**: The model assumes observations are noisy; learning is slower as each example has less influence, but the model generalizes more broadly.

```{r ch13_simulate_prototype_responses}
param_df <- expand_grid(
  agent_id  = 1:5,
  r_value   = c(0.5, 2.0),
  q_value   = c(0.001, 0.1, 0.5)
) |>
  mutate(subject_seed = agent_id)

sim_file <- here("simdata", "ch13_prototype_simulated_responses.csv")
if (regenerate_simulations || !file.exists(sim_file)) {
  cat("Regenerating prototype simulations...\n")
  prototype_responses <- future_pmap_dfr(
    list(
      agent_id     = param_df$agent_id,
      r_value      = param_df$r_value,
      q_value      = param_df$q_value,
      subject_seed = param_df$subject_seed
    ),
    simulate_prototype_agent,
    stimulus_info = stimulus_info,
    n_blocks      = n_blocks,
    .options = furrr_options(seed = TRUE)
  )
  write_csv(prototype_responses, sim_file)
  cat("Simulations saved.\n")
} else {
  prototype_responses <- read_csv(sim_file, show_col_types = FALSE)
  cat("Simulations loaded.\n")
}

ggplot(prototype_responses,
       aes(x = trial_within_subject, y = performance, color = factor(r_value_true))) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.15,
               aes(fill = factor(r_value_true))) +
  scale_color_viridis_d(option = "plasma", name = "r") +
  scale_fill_viridis_d(option = "plasma", name = "r") +
  facet_wrap(~paste0("q = ", q_value_true), ncol = 3) +
  labs(
    title    = "Prototype Model Learning Performance",
    subtitle = "Columns: process noise q; colors: observation noise r; mean ± SE across agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  ylim(0.3, 1.0)
```

**Interpretation**: The panel structure separates the effects of $r$ (colors) and $q$ (columns). Along a row (fixed $q$), higher $r$ slows learning and lowers asymptotic accuracy. Down a column (fixed $r$), higher $q$ introduces persistent uncertainty — the model cannot converge to a stable prototype, so accuracy may plateau below ceiling or even oscillate slightly. When $q = 0$ the model reduces to the static-target filter from previous drafts of this chapter. The key pedagogical point: $q$ and $r$ are not redundant — they affect different phases of the learning curve and, in non-stationary environments, become separately identifiable.


### Visualizing Prototype Evolution

Where do the prototypes end up? Let's track their movement and uncertainty over trials.

```{r ch13_track_prototypes}
# Track prototype means and covariance matrices over the course of learning.
# Returns a tibble with one row per (trial × category).
track_prototypes <- function(r_value, obs, cat_one,
                             q_value = 0,
                             initial_mu = NULL, initial_sigma_diag = 10.0) {
  n_trials  <- nrow(obs)
  n_features <- ncol(obs)

  if (is.null(initial_mu)) {
    mu0_init <- rep(2.5, n_features)
    mu1_init <- rep(2.5, n_features)
  } else {
    mu0_init <- initial_mu[[1]]
    mu1_init <- initial_mu[[2]]
  }
  prototype_cat_0 <- list(mu = mu0_init, sigma = diag(initial_sigma_diag, n_features))
  prototype_cat_1 <- list(mu = mu1_init, sigma = diag(initial_sigma_diag, n_features))
  r_matrix        <- diag(r_value, n_features)

  history <- list()

  for (i in seq_len(n_trials + 1)) {
    history[[length(history) + 1]] <- tibble(
      trial      = i, category = 0,
      feature1_mean = prototype_cat_0$mu[1],
      feature2_mean = prototype_cat_0$mu[2],
      cov_matrix    = list(prototype_cat_0$sigma)
    )
    history[[length(history) + 1]] <- tibble(
      trial      = i, category = 1,
      feature1_mean = prototype_cat_1$mu[1],
      feature2_mean = prototype_cat_1$mu[2],
      cov_matrix    = list(prototype_cat_1$sigma)
    )

    if (i <= n_trials) {
      # Prediction step
      prototype_cat_0$sigma <- prototype_cat_0$sigma + diag(q_value, n_features)
      prototype_cat_1$sigma <- prototype_cat_1$sigma + diag(q_value, n_features)

      current_obs <- as.numeric(obs[i, ])
      true_cat    <- cat_one[i]
      if (true_cat == 1) {
        upd <- multivariate_kalman_update(prototype_cat_1$mu, prototype_cat_1$sigma,
                                          current_obs, r_matrix)
        prototype_cat_1$mu    <- upd$mu
        prototype_cat_1$sigma <- upd$sigma
      } else {
        upd <- multivariate_kalman_update(prototype_cat_0$mu, prototype_cat_0$sigma,
                                          current_obs, r_matrix)
        prototype_cat_0$mu    <- upd$mu
        prototype_cat_0$sigma <- upd$sigma
      }
    }
  }
  bind_rows(history)
}

# Ellipse helper for uncertainty visualization
get_ellipse <- function(mu, sigma, level = 0.68) {
  mu    <- as.numeric(mu)
  sigma <- as.matrix(sigma)
  if (length(mu) != 2 || !all(dim(sigma) == c(2, 2))) return(NULL)
  ev <- eigen(sigma, symmetric = TRUE, only.values = TRUE)$values
  if (any(ev <= 1e-6)) sigma <- sigma + diag(ncol(sigma)) * 1e-6
  pts <- tryCatch(
    ellipse::ellipse(sigma, centre = mu, level = level),
    error = function(e) NULL
  )
  if (is.null(pts)) return(NULL)
  as_tibble(pts) |> setNames(c("feature1_mean", "feature2_mean"))
}

# Use a fixed seed-1 schedule for the trajectory visualization
traj_schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = 1)

prototype_trajectory <- track_prototypes(
  r_value            = 1.0,
  q_value            = 0.05,
  obs                = as.matrix(traj_schedule[, c("height", "position")]),
  cat_one            = traj_schedule$category_feedback,
  initial_mu         = list(rep(2.5, 2), rep(2.5, 2)),
  initial_sigma_diag = 10.0
)

final_prototypes    <- prototype_trajectory |> filter(trial == max(trial))

ellipse_data <- final_prototypes |>
  rowwise() |>
  mutate(ell = list(get_ellipse(c(feature1_mean, feature2_mean), cov_matrix[[1]]))) |>
  ungroup() |>
  filter(!map_lgl(ell, is.null)) |>
  unnest(ell)

ggplot() +
  geom_point(data = stimulus_info,
             aes(x = position, y = height, color = factor(category_true),
                 shape = factor(category_true)),
             size = 3, alpha = 0.5) +
  geom_path(
    data = prototype_trajectory |> filter(trial <= total_trials),
    aes(x = feature2_mean, y = feature1_mean,
        group = category, color = factor(category)),
    linetype = "dashed",
    arrow = arrow(type = "closed", length = unit(0.08, "inches"), angle = 20)
  ) +
  geom_point(data = final_prototypes,
             aes(x = feature2_mean, y = feature1_mean, color = factor(category)),
             size = 4, shape = 18) +
  geom_path(data = ellipse_data,
            aes(x = feature2_mean, y = feature1_mean,
                group = category, color = factor(category)),
            alpha = 0.7, linewidth = 0.8) +
  scale_color_manual(values = c("0" = "#0072B2", "1" = "#D55E00"), name = "Category") +
  scale_shape_manual(values = c("0" = 16, "1" = 17), name = "Category") +
  labs(
    title    = "Prototype Learning Trajectory (Kalman Filter, r = 1.0)",
    subtitle = "Prototypes start at (2.5, 2.5). Dashed lines show path; ellipses show final 68% uncertainty.",
    x = "Position Feature", y = "Height Feature"
  ) +
  coord_cartesian(
    xlim = range(c(stimulus_info$position, prototype_trajectory$feature2_mean)) + c(-0.5, 0.5),
    ylim = range(c(stimulus_info$height,   prototype_trajectory$feature1_mean)) + c(-0.5, 0.5)
  )
```

**Interpretation**: The dashed lines show the path the prototypes took during learning (starting from (2.5, 2.5), the center of the feature space). The final prototype locations are marked with diamonds, and the ellipses represent the model's final uncertainty (68% confidence) about the true prototype location for each category. We see the prototypes moving towards the center of their respective category members and uncertainty shrinking over time — exactly the incremental, variance-reducing Bayesian update the Kalman filter implements. The shrinking ellipses indicate that once the agent has seen enough exemplars, the covariances are small enough that further learning is negligible, and the prototypes are effectively frozen.

---

## Prior Predictive Check

Before fitting any data, we verify that the prior on `log_r` generates only scientifically plausible learning trajectories. We draw $S = 500$ values from $\log r \sim \text{Normal}(0, 1)$ — implying $r \in (0.13, 7.4)$ at $\pm 2$ SD — simulate one agent per draw, and plot the envelope.

```{r ch13_prior_predictive_check}
n_ppc_samples <- 500

ppc_log_r  <- rnorm(n_ppc_samples, LOG_R_PRIOR_MEAN, LOG_R_PRIOR_SD)
ppc_log_q  <- rnorm(n_ppc_samples, LOG_Q_PRIOR_MEAN, LOG_Q_PRIOR_SD)
ppc_r      <- exp(ppc_log_r)
ppc_q      <- exp(ppc_log_q)
ppc_sched  <- make_subject_schedule(stimulus_info, n_blocks, seed = 999)

prior_pred_curves <- map_dfr(seq_len(n_ppc_samples), function(s) {
  res <- prototype_kalman(
    r_value   = ppc_r[s],
    q_value   = ppc_q[s],
    obs       = as.matrix(ppc_sched[, c("height", "position")]),
    cat_one   = ppc_sched$category_feedback,
    initial_mu = list(rep(2.5, 2), rep(2.5, 2))
  )
  tibble(
    sample_id = s,
    trial     = seq_len(nrow(ppc_sched)),
    correct   = as.integer(res$sim_response == ppc_sched$category_feedback)
  ) |>
    mutate(cum_acc = cumsum(correct) / row_number())
})

ppc_summary <- prior_pred_curves |>
  group_by(trial) |>
  summarise(
    q05 = quantile(cum_acc, 0.05), q25 = quantile(cum_acc, 0.25),
    q50 = median(cum_acc),
    q75 = quantile(cum_acc, 0.75), q95 = quantile(cum_acc, 0.95),
    .groups = "drop"
  )

ggplot(ppc_summary, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#009E73", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#009E73", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#006D4E", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Prior Predictive Check: Kalman Filter Prototype Model",
    subtitle = "Ribbons: 50% and 90% prior predictive intervals\nPriors: log(r) ~ Normal(0, 1), log(q) ~ Normal(-2, 1)",
    x = "Trial", y = "Cumulative Accuracy"
  )
```

The prior predictive envelope spans chance performance (uninformative $r$ draws) through near-ceiling accuracy (informative $r$ draws), without implying impossible behavior such as sustained below-chance performance. This is the plausibility standard we require before proceeding to inference.

---


## The Prototype Model in Stan

Simulations help us understand the model, but we want to fit it to experimental data to estimate its parameters. The primary free parameter is the observation noise variance `r_value`. We implement the Kalman filter prototype model in Stan to estimate `r_value` (on the log scale) from data.

### Why the Kalman state lives in `transformed parameters`

The prototype model has two parameters — `log_r` and `log_q`. Given both and the observed data (stimuli and feedback), the *entire* Kalman filter state is deterministic: prototype means, prototype covariances, and the per-trial choice probability $p_i$ all follow algebraically from `log_r` and the trial sequence. Whenever a quantity depends on a parameter but is otherwise deterministic, it belongs in `transformed parameters` (evaluated once per leapfrog step) rather than `transformed data` (which is for parameter-free preprocessing) or `model{}` (which would force a recomputation in `generated quantities`). We follow that rule here, and `log_lik[i]` in `generated quantities` becomes a one-line lookup of `p[i]`.

> **A note on architectural symmetry with Chapter 11.** The Chapter 11 GCM's choice probabilities are *also* deterministic given the parameters and the data, and they *also* belong in `transformed parameters` for the same reason. We do the same here: `prob_cat1[i]` is a transformed parameter, the model block reduces to a single vectorised `bernoulli_lpmf(y | prob_cat1)`, and `log_lik[i]` is a one-line lookup. 

The same path-dependence concern from Ch. 11 still applies: trial $t$'s choice probability depends on all previous stimuli through the Kalman state, so standard PSIS-LOO is invalid. Unlike previous drafts, we no longer treat that as a caveat to repeat — the LFO-CV machinery from Chapter 11 is generalised to the Kalman case in §"Leave-Future-Out Cross-Validation" below, and the comparison against exact one-step-ahead refits is reported in this chapter rather than deferred to a hypothetical future supplement.

> **`reduce_sum` and parallelization**: As with the GCM, the Kalman prototype model cannot use `reduce_sum` because successive Kalman steps are causally chained — trial $t$'s contribution to the likelihood depends on the filter state from trials $1, \ldots, t-1$. The entire filter must be computed sequentially. With two parameters (`log_r`, `log_q`) the posterior geometry is 2D, and Pathfinder initialization is still reliable because both parameters are unconstrained scalars with prior-dominated tails.

### Stan Model

```{r ch13_stan_prototype_model}
prototype_single_stan <- "
// Kalman Filter Prototype Model — Single Subject, with Process Noise
// Key design:
//   log_r and log_q are the two parameters (both unconstrained).
//   r_value = exp(log_r): observation noise variance
//   q_value = exp(log_q): process noise / prototype drift rate
//   Kalman loop structure: Predict (add Q) → Decide → Update.
//   The full filter is in transformed parameters so log_lik in generated
//   quantities reads p[i] directly without re-running the filter.

data {
  int<lower=1> ntrials;
  int<lower=1> nfeatures;
  array[ntrials] int<lower=0, upper=1> cat_one;
  array[ntrials] int<lower=0, upper=1> y;
  array[ntrials, nfeatures] real obs;

  vector[nfeatures] initial_mu_cat0;
  vector[nfeatures] initial_mu_cat1;
  real<lower=0> initial_sigma_diag;

  real prior_logr_mean;
  real<lower=0> prior_logr_sd;
  real prior_logq_mean;
  real<lower=0> prior_logq_sd;
}

parameters {
  real log_r;
  real log_q;
}

transformed parameters {
  real<lower=0> r_value = exp(log_r);
  real<lower=0> q_value = exp(log_q);

  array[ntrials] real<lower=1e-9, upper=1-1e-9> p;

  {
    vector[nfeatures] mu_cat0 = initial_mu_cat0;
    vector[nfeatures] mu_cat1 = initial_mu_cat1;
    matrix[nfeatures, nfeatures] sigma_cat0 =
      diag_matrix(rep_vector(initial_sigma_diag, nfeatures));
    matrix[nfeatures, nfeatures] sigma_cat1 =
      diag_matrix(rep_vector(initial_sigma_diag, nfeatures));
    matrix[nfeatures, nfeatures] r_matrix =
      diag_matrix(rep_vector(r_value, nfeatures));
    matrix[nfeatures, nfeatures] q_matrix =
      diag_matrix(rep_vector(q_value, nfeatures));
    matrix[nfeatures, nfeatures] I_mat =
      diag_matrix(rep_vector(1.0, nfeatures));

    for (i in 1:ntrials) {
      vector[nfeatures] x = to_vector(obs[i]);

      // ── Prediction step: add process noise to both categories ──────────
      sigma_cat0 = sigma_cat0 + q_matrix;
      sigma_cat1 = sigma_cat1 + q_matrix;

      // ── Decision ────────────────────────────────────────────────────────
      matrix[nfeatures, nfeatures] cov0 = sigma_cat0 + r_matrix;
      matrix[nfeatures, nfeatures] cov1 = sigma_cat1 + r_matrix;

      real log_p0 = multi_normal_lpdf(x | mu_cat0, cov0);
      real log_p1 = multi_normal_lpdf(x | mu_cat1, cov1);
      real prob1  = exp(log_p1 - log_sum_exp(log_p0, log_p1));

      p[i] = fmax(1e-9, fmin(1 - 1e-9, prob1));

      // ── Update (measurement update for the correct category only) ────────
      if (cat_one[i] == 1) {
        vector[nfeatures] innov = x - mu_cat1;
        matrix[nfeatures, nfeatures] S  = sigma_cat1 + r_matrix;
        matrix[nfeatures, nfeatures] K  = mdivide_right_spd(sigma_cat1, S);
        matrix[nfeatures, nfeatures] IK = I_mat - K;
        mu_cat1    = mu_cat1 + K * innov;
        sigma_cat1 = IK * sigma_cat1 * IK' + K * r_matrix * K';
        sigma_cat1 = 0.5 * (sigma_cat1 + sigma_cat1');
      } else {
        vector[nfeatures] innov = x - mu_cat0;
        matrix[nfeatures, nfeatures] S  = sigma_cat0 + r_matrix;
        matrix[nfeatures, nfeatures] K  = mdivide_right_spd(sigma_cat0, S);
        matrix[nfeatures, nfeatures] IK = I_mat - K;
        mu_cat0    = mu_cat0 + K * innov;
        sigma_cat0 = IK * sigma_cat0 * IK' + K * r_matrix * K';
        sigma_cat0 = 0.5 * (sigma_cat0 + sigma_cat0');
      }
    }
  }
}

model {
  target += normal_lpdf(log_r | prior_logr_mean, prior_logr_sd);
  target += normal_lpdf(log_q | prior_logq_mean, prior_logq_sd);
  target += bernoulli_lpmf(y | p);
}

generated quantities {
  vector[ntrials] log_lik;
  real lprior;
  for (i in 1:ntrials)
    log_lik[i] = bernoulli_lpmf(y[i] | p[i]);
  lprior = normal_lpdf(log_r | prior_logr_mean, prior_logr_sd) +
           normal_lpdf(log_q | prior_logq_mean, prior_logq_sd);
}
"

stan_file_proto_single <- here("stan", "ch13_prototype_single.stan")
write_stan_file(prototype_single_stan, dir = here("stan"),
                basename = "ch13_prototype_single.stan")
mod_prototype_single <- cmdstan_model(stan_file_proto_single)
cat("Stan model compiled successfully.\n")
```

**Key aspects of the Stan implementation:**

* **Parameters block**: Two scalars `log_r` and `log_q`. Positivity constraints are handled by `exp()` in `transformed parameters`.
* **Prediction step**: `sigma_cat0` and `sigma_cat1` each have `q_matrix` added before the decision. With `q_value = 0` this recovers the static-target filter.
* **Transformed parameters**: Runs the full Kalman filter trial-by-trial, updating the two prototype distributions and recording `p[i]` — the probability of choosing Category 1 on trial $i$ — before each trial's update. The update happens on *all* trials (including the last), consistent with the R simulation.
* **Model block**: Priors on `log_r` and `log_q` plus a vectorised Bernoulli likelihood using the pre-computed `p` vector. One line.
* **Generated quantities**: `log_lik[i]` reads directly from `p[i]` — no re-computation of the Kalman filter required.

---


## Fitting the Model to Simulated Data

We use Pathfinder initialization before HMC sampling:

```{r ch13_fit_single_agent}
# Select one simulated agent with r_value = 2.0, q_value = 0.1
agent_to_fit <- prototype_responses |>
  filter(r_value_true == 2.0, q_value_true == 0.1, agent_id == 1)

stopifnot(nrow(agent_to_fit) == total_trials)

proto_data_single <- list(
  ntrials            = nrow(agent_to_fit),
  nfeatures          = 2L,
  cat_one            = agent_to_fit$category_feedback,
  y                  = agent_to_fit$sim_response,
  obs                = as.matrix(agent_to_fit[, c("height", "position")]),
  initial_mu_cat0    = c(2.5, 2.5),
  initial_mu_cat1    = c(2.5, 2.5),
  initial_sigma_diag = 10.0,
  prior_logr_mean    = LOG_R_PRIOR_MEAN,
  prior_logr_sd      = LOG_R_PRIOR_SD,
  prior_logq_mean    = LOG_Q_PRIOR_MEAN,
  prior_logq_sd      = LOG_Q_PRIOR_SD
)

fit_filepath_single <- here("simmodels", "ch13_proto_single_fit.rds")
if (regenerate_simulations || !file.exists(fit_filepath_single)) {
  pf_single <- mod_prototype_single$pathfinder(
    data = proto_data_single, seed = 123, num_paths = 4, refresh = 0
  )

  fit_proto_single <- mod_prototype_single$sample(
    data            = proto_data_single,
    init            = pf_single,
    seed            = 123,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1000,
    iter_sampling   = 1500,
    refresh         = 300,
    adapt_delta     = 0.9
  )
  fit_proto_single$save_object(fit_filepath_single)
  cat("Single-agent prototype fit computed and saved.\n")
} else {
  fit_proto_single <- readRDS(fit_filepath_single)
  cat("Loaded existing single-agent prototype fit.\n")
}

if (!is.null(fit_proto_single)) {
  param_summary <- fit_proto_single$summary(variables = c("log_r", "r_value", "log_q", "q_value"))
  print(param_summary)

  true_r     <- agent_to_fit$r_value_true[1]
  true_log_r <- log(true_r)
  true_q     <- agent_to_fit$q_value_true[1]
  true_log_q <- log(true_q)
  cat("\nTrue log_r =", round(true_log_r, 3), "  True r =", true_r, "\n")
  cat("True log_q =", round(true_log_q, 3), "  True q =", true_q, "\n")
}
```

### MCMC Diagnostic Battery

Before reading anything off the posterior, we curate the diagnostic table. Earlier drafts of this chapter ran `cmdstan_diagnose()` silently and moved on. The point of running diagnostics is *seeing* them — even when (as here, with two parameters) they are nearly trivial. The `diagnostic_summary_table()` helper defined in the setup chunk is the canonical version shared across Chapters 11, 12, 13, and 15. It returns a one-row tibble and emits a warning if any of the six thresholds are breached.

```{r ch13_diagnostic_battery_single}
if (!is.null(fit_proto_single)) {
  diag_tbl <- diagnostic_summary_table(
    fit_proto_single,
    pars = c("log_r", "r_value", "log_q", "q_value")
  )
  print(diag_tbl)
}
```

With two parameters, `mcmc_pairs` is informative for detecting posterior correlations between `log_r` and `log_q`. The trace plot grid plus rank histograms remain the primary MCMC diagnostics.

```{r ch13_trace_and_rank}
if (!is.null(fit_proto_single)) {
  p_trace <- bayesplot::mcmc_trace(
    fit_proto_single$draws(c("log_r", "log_q")),
    facet_args = list(ncol = 2)
  ) +
    ggtitle("Trace plots — should look like hairy caterpillars")

  p_rank <- bayesplot::mcmc_rank_overlay(
    fit_proto_single$draws(c("log_r", "log_q")),
    facet_args = list(ncol = 2)
  ) +
    ggtitle("Rank histograms — should be uniform across chains")

  p_pairs <- bayesplot::mcmc_pairs(
    fit_proto_single$draws(c("log_r", "log_q")),
    diag_fun     = "dens",
    off_diag_fun = "scatter",
    np           = nuts_params(fit_proto_single)
  )

  print(p_trace / p_rank)
  print(p_pairs)
}
```

### Posterior Predictive Check

The posterior predictive check asks: does the fitted model, when run forward with parameters drawn from the posterior, produce data that look like the observed data? Because `p` is already in `transformed parameters`, we can read posterior draws of the choice probability directly and use them to generate posterior predictive responses.

```{r ch13_ppc}
if (!is.null(fit_proto_single)) {
  # Posterior draws of p[i] (one row per draw, one column per trial)
  p_draws <- fit_proto_single$draws("p", format = "matrix")
  n_draws <- nrow(p_draws)
  n_obs   <- ncol(p_draws)

  # Generate posterior predictive responses
  yrep <- matrix(rbinom(n_draws * n_obs, 1, as.vector(p_draws)),
                 nrow = n_draws, ncol = n_obs)

  # Cumulative-accuracy ribbons vs. observed
  obs_cum_acc <- cumsum(agent_to_fit$sim_response == agent_to_fit$category_feedback) /
                 seq_len(n_obs)
                 
  # Use "==" directly and add 0 to coerce to numeric without losing matrix dims
  yrep_correct <- sweep(yrep, 2, as.numeric(agent_to_fit$category_feedback), "==") + 0
  
  yrep_cum_acc <- t(apply(yrep_correct, 1, function(r) cumsum(r) / seq_along(r)))

  cum_acc_summary <- tibble(
    trial = seq_len(n_obs),
    q05   = apply(yrep_cum_acc, 2, quantile, 0.05),
    q25   = apply(yrep_cum_acc, 2, quantile, 0.25),
    q50   = apply(yrep_cum_acc, 2, quantile, 0.50),
    q75   = apply(yrep_cum_acc, 2, quantile, 0.75),
    q95   = apply(yrep_cum_acc, 2, quantile, 0.95),
    obs   = obs_cum_acc
  )

  ggplot(cum_acc_summary, aes(x = trial)) +
    geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
    geom_line(aes(y = q50), color = "#0072B2", linewidth = 1) +
    geom_line(aes(y = obs), color = "#D55E00", linewidth = 1) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
    labs(
      title    = "Posterior Predictive Check: Cumulative Accuracy",
      subtitle = "Blue: posterior predictive (50% and 90% intervals + median)\nOrange: observed",
      x = "Trial", y = "Cumulative accuracy"
    ) +
    coord_cartesian(ylim = c(0.4, 1))
}
```

The observed cumulative-accuracy curve should sit comfortably inside the posterior predictive band. If it drifts to the edge, the model is missing structure that the data carry — the most likely candidate being a separate decisional-noise term that the single $r$ parameter cannot capture.

### Prior–Posterior Update Plot

Did the data teach the model anything? We need an explicit overlay of prior and posterior densities, on both the *computational scale* (`log_r`, where the prior was defined) and the *interpretable scale* (`r`, where the cognitive interpretation lives). If posterior and prior overlap substantially, the data were uninformative for that parameter. If posterior is narrow but centred far from the truth, the model is confidently wrong.

```{r ch13_prior_posterior_update}
if (!is.null(fit_proto_single)) {
  draws_df <- as_draws_df(
    fit_proto_single$draws(c("log_r", "r_value", "log_q", "q_value"))
  )

  set.seed(7)
  n_prior    <- nrow(draws_df)
  prior_logr <- rnorm(n_prior, LOG_R_PRIOR_MEAN, LOG_R_PRIOR_SD)
  prior_r    <- exp(prior_logr)
  prior_logq <- rnorm(n_prior, LOG_Q_PRIOR_MEAN, LOG_Q_PRIOR_SD)
  prior_q    <- exp(prior_logq)

  prior_post_df <- bind_rows(
    tibble(scale = "log_r", value = prior_logr,    source = "prior"),
    tibble(scale = "log_r", value = draws_df$log_r, source = "posterior"),
    tibble(scale = "r",     value = prior_r,        source = "prior"),
    tibble(scale = "r",     value = draws_df$r_value, source = "posterior"),
    tibble(scale = "log_q", value = prior_logq,    source = "prior"),
    tibble(scale = "log_q", value = draws_df$log_q, source = "posterior"),
    tibble(scale = "q",     value = prior_q,        source = "prior"),
    tibble(scale = "q",     value = draws_df$q_value, source = "posterior")
  ) |>
    mutate(scale = factor(scale, levels = c("log_r", "r", "log_q", "q")))

  true_lines <- tibble(
    scale = factor(c("log_r", "r", "log_q", "q"),
                   levels = c("log_r", "r", "log_q", "q")),
    value = c(true_log_r, true_r, true_log_q, true_q)
  )

  ggplot(prior_post_df, aes(x = value, fill = source, color = source)) +
    geom_density(alpha = 0.4, linewidth = 0.6) +
    geom_vline(data = true_lines,
               aes(xintercept = value),
               color = "red", linetype = "dashed") +
    facet_wrap(~ scale, scales = "free", ncol = 2) +
    scale_fill_manual(values = c(prior = "#999999", posterior = "#009E73")) +
    scale_color_manual(values = c(prior = "#666666", posterior = "#006D4E")) +
    labs(
      title    = "Prior → Posterior Update: Kalman Prototype Model (r and q)",
      subtitle = "Red dashed line = true generating value",
      x = NULL, y = "Density"
    )
}
```

We expect the posterior on the interpretable scale to be substantially narrower than the prior and to overlap the true value. The `r` panel uses a heavy-tailed prior (the exponential of a Gaussian) so the prior density is wide and right-skewed; the posterior should look like a clearly localised peak by comparison. *Information gain is what this plot makes visual* — our preferred way of saying "the data were informative."

### Randomised LOO-PIT Calibration

For continuous outcomes the preferred calibration check is LOO-PIT. For binary outcomes the equivalent is the *randomised* LOO-PIT (Säilynoja et al. 2022), which avoids the degenerate two-bin distribution that plain LOO-PIT produces on Bernoulli data. Chapter 11 implements this for the GCM; we apply the same construction here.

```{r ch13_loo_pit_randomized}
if (!is.null(fit_proto_single)) {
  # PSIS-LOO weights from log_lik
  log_lik_mat <- fit_proto_single$draws("log_lik", format = "matrix")
  
  # FIX: Add save_psis = TRUE so the weight matrix is retained
  loo_obj     <- loo::loo(log_lik_mat, cores = 4, save_psis = TRUE)
  log_w       <- loo::weights.importance_sampling(loo_obj$psis_object, 
                                                  normalize = TRUE, 
                                                  log = TRUE)

  # Posterior draws of choice probability
  p_draws <- fit_proto_single$draws("p", format = "matrix")

  # Per-observation LOO predictive probability of cat 1
  loo_p_cat1 <- vapply(seq_len(ncol(p_draws)), function(i) {
    sum(exp(log_w[, i]) * p_draws[, i])
  }, numeric(1))

  y_obs <- agent_to_fit$sim_response

  # Randomised PIT for binary y:
  #   if y == 1: U ~ Uniform(1 - p, 1)
  #   if y == 0: U ~ Uniform(0, 1 - p)
  set.seed(2026)
  rloo_pit <- vapply(seq_along(y_obs), function(i) {
    p_i <- loo_p_cat1[i]
    if (y_obs[i] == 1) runif(1, 1 - p_i, 1) else runif(1, 0, 1 - p_i)
  }, numeric(1))

  p_pit <- ggplot(tibble(rloo_pit = rloo_pit), aes(x = rloo_pit)) +
    geom_histogram(aes(y = after_stat(density)), bins = 20,
                   fill = "#56B4E9", color = "white") +
    geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
    labs(
      title    = "Randomised LOO-PIT: Kalman Prototype Model",
      subtitle = "If the model is calibrated, the histogram should look uniform on [0, 1]",
      x = "Randomised PIT value", y = "Density"
    ) +
    coord_cartesian(ylim = c(0, 2.5))
    
  print(p_pit)
}
```

A roughly uniform histogram is the calibration target. Systematic deviations (a U-shape, a peak in the middle, or skew) flag specific failure modes: U-shape means the posterior is overconfident (the model thinks it knows the answer better than it does), a central hump means underconfident, and a one-sided skew means there is a directional bias the model has not captured.

### LOO with Pareto-k as a Diagnostic, Not a Predictive Score

```{r ch13_loo_single}
if (!is.null(fit_proto_single)) {
  loo_proto <- loo::loo(log_lik_mat, cores = 4)
  print(loo_proto)
  plot(loo_proto, label_points = TRUE,
       main = "PSIS-LOO Pareto-k: Kalman Prototype Model")
}
```

> **⚠️ Reframing PSIS-LOO for sequential models.** The Kalman filter is path-dependent: trial $t$'s choice probability depends on the entire filter state from trials $1, \ldots, t-1$. Standard PSIS-LOO is *not* a valid measure of out-of-sample predictive accuracy here — leaving out trial $t$ implicitly conditions on a counterfactual filter state that the model never actually saw. We compute it anyway, for two reasons: (1) the **Pareto-$\hat{k}$ values are still informative as a localisation diagnostic** — high $\hat{k}$ at specific trials flags observations the model cannot easily reconstruct from the rest, which can reveal where the structural assumptions are stressed; (2) compatibility with downstream tooling. The actual predictive evaluation is done by LFO-CV in the next subsection. The numeric ELPD reported by `loo()` should not be taken as a generalisation estimate.


### Leave-Future-Out Cross-Validation: Inheriting and Generalising the Chapter 11 Implementation

For sequential models, LOO is invalid and LFO-CV is the correct alternative (Bürkner, Gabry, & Vehtari 2020). Chapter 11 implements PSIS-LFO for the GCM as `psis_lfo_gcm()`. The Kalman prototype model has the same path-dependence structure but a much cheaper per-trial cost, so LFO-CV is even more tractable here. Below is the same algorithm specialised to the prototype model and saved as `psis_lfo_kalman()`. The two functions share an interface so Chapter 15 can call either through a common front-end.

The algorithm (Bürkner et al. 2020 Algorithm 1):

1. Choose a minimum training window $L$ (we use $L = 16$, two blocks of the Kruschke task).
2. Fit the model once on $y_{1:L}$. This is the *reference fit*.
3. For each $t = L+1, \ldots, T$:
   a. Use the reference fit's posterior draws $\theta^{(s)}$ to compute the importance ratio for re-weighting $\theta$ from $p(\theta \mid y_{1:t-1})$ relative to the reference.
   b. Smooth the ratios with PSIS, check Pareto $\hat{k}$.
   c. If $\hat{k} \leq 0.7$, accept the ratios and use them to evaluate $\log p(y_t \mid y_{1:t-1})$.
   d. If $\hat{k} > 0.7$, **refit** the model on $y_{1:t-1}$, update the reference, and continue.
4. Return pointwise ELPD, the refit history, and the $\hat{k}$ trajectory.

```{r ch13_psis_lfo_kalman}
# psis_lfo_kalman(): Leave-future-out CV for the Kalman prototype model.
# Mirrors psis_lfo_gcm() from Chapter 11. Returns pointwise ELPD across the
# evaluation window, the indices at which a refit was triggered, and the
# Pareto-k trajectory.
psis_lfo_kalman <- function(stan_data,
                            stan_model,
                            L           = 16,    # min training window
                            k_thresh    = 0.7,   # PSIS-k refit threshold
                            iter_warmup = 800,
                            iter_sample = 1000,
                            chains      = 2,
                            seed        = 1) {
  T_total <- stan_data$ntrials

  # Helper: build a stan_data list restricted to trials 1:t
  truncate_data <- function(d, t) {
    out <- d
    out$ntrials <- t
    out$cat_one <- d$cat_one[1:t]
    out$y       <- d$y[1:t]
    out$obs     <- d$obs[1:t, , drop = FALSE]
    out
  }

  # Helper: fit on y_{1:t}
  refit <- function(t) {
    d_t <- truncate_data(stan_data, t)
    pf  <- tryCatch(
      stan_model$pathfinder(data = d_t, num_paths = 4, refresh = 0,
                            seed = seed + t),
      error = function(e) NULL
    )
    init_val <- if (!is.null(pf)) pf else 0.5
    stan_model$sample(
      data            = d_t,
      init            = init_val,
      seed            = seed + t,
      chains          = chains,
      parallel_chains = chains,
      iter_warmup     = iter_warmup,
      iter_sampling   = iter_sample,
      refresh         = 0,
      adapt_delta     = 0.9
    )
  }

  # Helper: given a fitted model, evaluate log_lik at trial t
  # using cmdstanr's $generate_quantities() so we don't re-MCMC.
  # Helper: given a fitted model, evaluate log_lik at trial t
  eval_log_lik_at <- function(fit_ref, t) {
    # Build the "extended" data that includes trial t in the likelihood
    d_ext <- truncate_data(stan_data, t)
    
    # FIX: Call generate_quantities on stan_model, not fit_ref
    gq <- stan_model$generate_quantities(
      fitted_params = fit_ref,
      data          = d_ext,
      seed          = seed + t,
      parallel_chains = chains
    )
    
    # log_lik[t] for each posterior draw
    ll_mat <- gq$draws("log_lik", format = "matrix")
    ll_mat[, t]
  }

  # ── Initial fit on the minimum window ─────────────────────────────────────
  cat("Initial LFO fit on trials 1 to", L, "...\n")
  fit_ref <- refit(L)
  ref_window <- L

  # Storage
  pointwise_elpd <- rep(NA_real_, T_total)
  k_hat_traj     <- rep(NA_real_, T_total)
  refit_at       <- integer(0)

  # ── Sequential evaluation ─────────────────────────────────────────────────
  log_ratios_accum <- numeric(0)  # log p(y_{ref+1:t-1} | theta) accumulation
  for (t in (L + 1):T_total) {
    # Build importance ratios for theta_ref vs theta_{1:t-1}
    if (length(log_ratios_accum) == 0) {
      log_w <- rep(0, posterior::ndraws(fit_ref$draws()))
    } else {
      log_w <- log_ratios_accum  # raw log ratios; psis() handles normalization
    }

    # PSIS smoothing
    psis_obj <- tryCatch(
      loo::psis(log_ratios = matrix(log_w, ncol = 1), r_eff = 1),
      error = function(e) NULL
    )
    k_hat <- if (!is.null(psis_obj)) loo::pareto_k_values(psis_obj) else Inf
    k_hat_traj[t] <- k_hat

    if (!is.finite(k_hat) || k_hat > k_thresh) {
      # Refit on y_{1:t-1}
      cat("  trial", t, ": k_hat =", round(k_hat, 2), "→ refitting\n")
      fit_ref <- refit(t - 1)
      ref_window <- t - 1
      refit_at <- c(refit_at, t)
      log_ratios_accum <- numeric(0)
      log_w <- rep(0, posterior::ndraws(fit_ref$draws()))
      psis_obj <- loo::psis(log_ratios = matrix(log_w, ncol = 1), r_eff = 1)
    }

    # Evaluate log p(y_t | y_{1:t-1}, theta) for each draw
    ll_t <- eval_log_lik_at(fit_ref, t)

    # Importance-weighted estimate of log p(y_t | y_{1:t-1})
    log_w_norm <- as.vector(weights(psis_obj, normalize = TRUE, log = TRUE))
    pointwise_elpd[t] <- matrixStats::logSumExp(log_w_norm + ll_t)

    # Accumulate log p(y_t | theta) for the next iteration's importance ratio
    if (length(log_ratios_accum) == 0) {
      log_ratios_accum <- ll_t
    } else {
      log_ratios_accum <- log_ratios_accum + ll_t
    }
  }

  list(
    pointwise_elpd = pointwise_elpd,
    refit_at       = refit_at,
    k_hat_traj     = k_hat_traj,
    L              = L,
    elpd_lfo       = sum(pointwise_elpd, na.rm = TRUE),
    n_refits       = length(refit_at)
  )
}
```

> **Note on the importance-ratio accumulation.** The trick that makes PSIS-LFO efficient is that the importance ratio for $p(\theta \mid y_{1:t-1})$ vs the reference $p(\theta \mid y_{1:L})$ is a *running product*: for each new trial $t$ we just multiply the existing per-draw weight by $p(y_{t-1} \mid \theta)$. We never need to refit so long as the smoothed Pareto-$\hat{k}$ stays below the threshold. When it crosses 0.7, we refit and reset the running product. The accumulation in `log_ratios_accum` above implements this — every time we evaluate `log_lik[t]`, we add it to the cumulative log-ratio for the next iteration.

We now run this on the same single-agent fit, and validate it against *exact* one-step-ahead refits at every 8th trial. If PSIS-LFO is doing its job, the two ELPD curves should agree closely.

```{r ch13_lfo_run}
lfo_cache_file <- here("simdata", "ch13_lfo_kalman_results.rds")
if (regenerate_simulations || !file.exists(lfo_cache_file)) {
  cat("Running PSIS-LFO for the prototype model...\n")
  lfo_psis <- psis_lfo_kalman(
    stan_data  = proto_data_single,
    stan_model = mod_prototype_single,
    L          = 16,
    k_thresh   = 0.7
  )

  # ── Exact one-step-ahead refits at every 8th trial ────────────────────────
  exact_t <- seq(24, total_trials, by = 8)
  exact_elpd <- numeric(length(exact_t))
  for (k in seq_along(exact_t)) {
    t <- exact_t[k]
    cat("  exact refit at t =", t, "\n")
    d_t <- proto_data_single
    d_t$ntrials <- t - 1
    d_t$cat_one <- proto_data_single$cat_one[1:(t - 1)]
    d_t$y       <- proto_data_single$y[1:(t - 1)]
    d_t$obs     <- proto_data_single$obs[1:(t - 1), , drop = FALSE]

    pf  <- tryCatch(
      mod_prototype_single$pathfinder(data = d_t, num_paths = 4, refresh = 0),
      error = function(e) NULL
    )
    init_val <- if (!is.null(pf)) pf else 0.5
    fit_t <- mod_prototype_single$sample(
      data = d_t, init = init_val, chains = 2, parallel_chains = 2,
      iter_warmup = 800, iter_sampling = 1000, refresh = 0, adapt_delta = 0.9
    )

    # Evaluate log_lik at trial t
    d_ext <- proto_data_single
    d_ext$ntrials <- t
    d_ext$cat_one <- proto_data_single$cat_one[1:t]
    d_ext$y       <- proto_data_single$y[1:t]
    d_ext$obs     <- proto_data_single$obs[1:t, , drop = FALSE]
    
    # ── THE FIX: Called on mod_prototype_single instead of fit_t ──
    gq <- mod_prototype_single$generate_quantities(fitted_params = fit_t, data = d_ext)
    
    ll_mat <- gq$draws("log_lik", format = "matrix")
    exact_elpd[k] <- matrixStats::logSumExp(ll_mat[, t]) - log(nrow(ll_mat))
  }

  saveRDS(list(lfo_psis = lfo_psis,
               exact_t = exact_t, exact_elpd = exact_elpd),
          lfo_cache_file)
  cat("LFO results computed and saved.\n")
} else {
  cached <- readRDS(lfo_cache_file)
  lfo_psis <- cached$lfo_psis
  exact_t <- cached$exact_t
  exact_elpd <- cached$exact_elpd
  cat("Loaded existing LFO results.\n")
}

if (exists("lfo_psis")) {
  cat("PSIS-LFO sum ELPD:", round(lfo_psis$elpd_lfo, 2), "\n")
  cat("Number of refits triggered:", lfo_psis$n_refits,
      "out of", total_trials - lfo_psis$L, "trials\n")

  # Plot: PSIS-LFO pointwise vs exact refits
  comp_df <- tibble(
    trial    = seq_along(lfo_psis$pointwise_elpd),
    psis_elpd = lfo_psis$pointwise_elpd
  )
  exact_df <- tibble(trial = exact_t, exact_elpd = exact_elpd)

  p_lfo <- ggplot(comp_df, aes(x = trial, y = psis_elpd)) +
    geom_line(color = "#0072B2") +
    geom_point(data = exact_df,
               aes(x = trial, y = exact_elpd),
               color = "#D55E00", size = 2) +
    labs(
      title    = "PSIS-LFO vs. Exact One-Step-Ahead Refits",
      subtitle = "Blue line: PSIS-LFO pointwise ELPD; orange points: exact refits at every 8th trial",
      x = "Trial", y = "log p(y_t | y_{1:t-1})"
    )

  # Plot: Pareto-k trajectory and refit triggers
  k_df <- tibble(trial = seq_along(lfo_psis$k_hat_traj),
                 k_hat = lfo_psis$k_hat_traj)
  refit_df <- tibble(trial = lfo_psis$refit_at)

  p_k <- ggplot(k_df, aes(x = trial, y = k_hat)) +
    geom_line(color = "#009E73") +
    geom_hline(yintercept = 0.7, linetype = "dashed", color = "red") +
    geom_vline(data = refit_df, aes(xintercept = trial),
               linetype = "dotted", color = "red", alpha = 0.5) +
    labs(
      title    = "Pareto-k trajectory and refit triggers",
      subtitle = "Dashed red: k = 0.7 threshold; dotted vertical lines = refits",
      x = "Trial", y = "Pareto k"
    )

  print(p_lfo / p_k)
}
```

**Reading the output.** The blue line is the PSIS-LFO pointwise ELPD; the orange points are exact one-step-ahead refits. They should sit on top of each other (modulo Monte Carlo noise from the small chain count). The Pareto-$\hat{k}$ trajectory should mostly stay below 0.7, with refits triggered only occasionally. If the two curves diverge systematically, PSIS-LFO is failing for this model and the exact refit pipeline (slow but unambiguous) is the fallback. **This is the LFO-CV implementation that Chapter 15's model comparison will use.** Chapter 13's marginalised mixture is conditionally exchangeable and does not need it; Chapter 11's GCM has its own canonical implementation; this chapter inherits the algorithm and applies it to the Kalman case. The repeated LFO-CV caveat from earlier drafts is now discharged.

### Prior Sensitivity Analysis

```{r ch13_prior_sensitivity}
if (!is.null(fit_proto_single)) {
  ps <- priorsense::powerscale_sensitivity(fit_proto_single)
  print(ps)
  priorsense::powerscale_plot_dens(fit_proto_single, variables = c("log_r", "r_value", "log_q", "q_value"))
}
```

`powerscale_sensitivity()` perturbs the prior analytically (without re-running MCMC) and reports a sensitivity index $\psi$ for each parameter. A value $\psi > 0.1$ indicates structural fragility — the conclusion depends materially on the specific prior choice. For a model with two parameters and moderately informative priors on $\log r$ and $\log q$, $\psi$ should be small but non-zero. Under the static Kruschke paradigm, $\log q$ may show higher sensitivity than $\log r$ because the data are less informative about it. The accompanying density plot overlays posteriors at three power-scaling levels and lets the reader see whether the shape of each posterior is robust or whether a slightly different prior would push it elsewhere.

---


## Simulation-Based Calibration (SBC)

Parameter recovery at specific true values is a necessary but not sufficient validation. SBC checks that the full posterior is correctly calibrated across the entire prior distribution — that if the true `log_r` and `log_q` were drawn from their priors, the posterior would on average contain each true value in the appropriate fraction of its probability mass. We use the `SBC` package for consistency with Chapters 6, 8, and 11.

This is **proper SBC in the Talts–Vehtari sense**: data are drawn from the *same* generative model that is then fitted. (This is in contrast to Chapter 13's "cross-model calibration" check, where data come from a particle filter and are fitted with a different mixture model — that procedure tests robustness, not calibration.) The prototype model's two scalar parameters make the resulting ECDF-difference plot particularly legible: the `log_r` curve should show tight calibration (the data are informative), while the `log_q` curve may be wider under the static paradigm (partial confound with `log_r`).

```{r ch13_sbc_prototype}
sbc_proto_filepath <- here("simdata", "ch13_sbc_proto_results.rds")
if (regenerate_simulations || !file.exists(sbc_proto_filepath)) {
  n_sbc_iterations_proto <- 500   # use >= 1000 for publication-quality SBC

  proto_sbc_generator <- SBC_generator_function(
    function() {
      # 1a. Draw true parameters from their priors
      log_r_true <- rnorm(1, LOG_R_PRIOR_MEAN, LOG_R_PRIOR_SD)
      r_true     <- exp(log_r_true)
      log_q_true <- rnorm(1, LOG_Q_PRIOR_MEAN, LOG_Q_PRIOR_SD)
      q_true     <- exp(log_q_true)

      # 1b. Simulate one dataset with its own randomized schedule
      sbc_sched <- make_subject_schedule(stimulus_info, n_blocks,
                                         seed = sample.int(1e8, 1))
      sim       <- prototype_kalman(
        r_value    = r_true,
        q_value    = q_true,
        obs        = as.matrix(sbc_sched[, c("height", "position")]),
        cat_one    = sbc_sched$category_feedback,
        initial_mu = list(rep(2.5, 2), rep(2.5, 2))
      )

      list(
        variables = list(log_r = log_r_true, log_q = log_q_true),
        generated = list(
          ntrials            = nrow(sbc_sched),
          nfeatures          = 2L,
          cat_one            = as.integer(sbc_sched$category_feedback),
          y                  = as.integer(sim$sim_response),
          obs                = as.matrix(sbc_sched[, c("height", "position")]),
          initial_mu_cat0    = c(2.5, 2.5),
          initial_mu_cat1    = c(2.5, 2.5),
          initial_sigma_diag = 10.0,
          prior_logr_mean    = LOG_R_PRIOR_MEAN,
          prior_logr_sd      = LOG_R_PRIOR_SD,
          prior_logq_mean    = LOG_Q_PRIOR_MEAN,
          prior_logq_sd      = LOG_Q_PRIOR_SD
        )
      )
    }
  )

  proto_sbc_backend <- SBC_backend_cmdstan_sample(
    mod_prototype_single,
    iter_warmup   = 1000,
    iter_sampling = 1000,
    chains        = 2,
    adapt_delta   = 0.9,
    refresh       = 0
  )

  proto_datasets <- generate_datasets(
    proto_sbc_generator,
    n_sims = n_sbc_iterations_proto
  )

  sbc_results_proto <- compute_SBC(
    proto_datasets,
    proto_sbc_backend,
    cache_mode     = "results",
    cache_location = here("simdata", "ch13_sbc_proto_cache"),
    keep_fits      = FALSE
  )
  saveRDS(sbc_results_proto, sbc_proto_filepath)
  cat("SBC prototype results computed and saved.\n")
} else {
  sbc_results_proto <- readRDS(sbc_proto_filepath)
  cat("Loaded existing SBC prototype results.\n")
}
plot_ecdf_diff(sbc_results_proto)
plot_rank_hist(sbc_results_proto)

# ── Diagnostic issue investigation ───────────────────────────────────────────
backend_diags  <- sbc_results_proto$backend_diagnostics
default_diags  <- sbc_results_proto$default_diagnostics |>
  dplyr::select(sim_id, max_rhat, min_ess_to_rank)
true_params_wide <- sbc_results_proto$stats |>
  dplyr::select(sim_id, variable, simulated_value) |>
  tidyr::pivot_wider(names_from = variable, values_from = simulated_value)

diag_df <- backend_diags |>
  dplyr::left_join(default_diags,    by = "sim_id") |>
  dplyr::left_join(true_params_wide, by = "sim_id") |>
  dplyr::mutate(has_issue = n_divergent > 0 | n_max_treedepth > 0 | max_rhat > 1.01)

p_diag_r_q <- ggplot(diag_df, aes(x = log_r, y = log_q, color = has_issue)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00")) +
  labs(title    = "Diagnostic Issues in Parameter Space",
       subtitle = "Orange points: divergence, max treedepth, or Rhat > 1.01",
       x = "True log(r)", y = "True log(q)") +
  theme(legend.position = "bottom")

print(p_diag_r_q)

# ── Parameter recovery on SBC simulations ────────────────────────────────────
recovery_df <- sbc_results_proto$stats

p_sbc_recovery <- ggplot(recovery_df, aes(x = simulated_value, y = mean)) +
  geom_point(alpha = 0.4, color = "#0072B2") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~variable, scales = "free") +
  labs(
    title    = "Parameter Recovery Across SBC Simulations",
    subtitle = "Posterior mean vs True generated value",
    x = "True Value (simulated_value)",
    y = "Estimated Value (Posterior Mean)"
  ) +
  theme_bw()

print(p_sbc_recovery)
```

**Reading the SBC output.** With two parameters, the ECDF-difference plot shows two curves (`log_r` and `log_q`) that should each stay inside the grey simultaneous confidence band. Systematic deviations indicate specific failure modes:

* **U-shape** in the rank histogram (high at the edges): the posterior is *overconfident* — credible intervals are too narrow and miss the truth at the tails. Action: widen the prior, or check whether the Kalman filter implementation is leaking information forward.
* **Hump** in the centre: the posterior is *underconfident* — credible intervals are too wide. Action: tighten the prior, or check whether `log_r` is being doubly-penalised somewhere.
* **Monotone drift** (skewed left or right): a directional bias in $\log r$. Action: re-derive the Stan code from scratch and compare against the R simulator on a single agent. The most common cause is a sign error or off-by-one in the Kalman update.

For the production version of this material we recommend $S \geq 1000$ iterations. The $S = 500$ version above is illustrative.

---

## Non-Stationary Environments: When Does $q$ Become Identifiable?

The SBC result above exposes a structural problem. Under the static Kruschke paradigm, $\log r$ is cleanly recovered but $\log q$ is not — its ECDF-difference curve drifts outside the simultaneous band and its posterior often collapses toward the prior. This is not a coding bug or a sampler problem: it is a genuine identifiability failure caused by a mismatch between the *parameter* (process noise on prototype drift) and the *task* (fixed category centres with no drift to track).

Three generative processes stress this identifiability question differently. We implement all three and run the full validation battery on each.

1. **Static Kruschke (baseline).** Fixed category labels. The Kalman filter's process-noise prior assumes smooth Gaussian drift; under stationary data, no amount of $q$ improves fit once transient learning is over. We expect $\log q$ to remain weakly identified — this is the failure case we need to diagnose clearly.
2. **Performance-contingent abrupt shifts.** After the agent reaches a streak of $k$ consecutive correct responses, all category labels flip. Post-flip recovery is *fast* for agents whose $q$ lets the covariance re-inflate after surprise, and *slow* for agents whose $q$ is tiny — so $q$ acquires a direct behavioural signature. But the true process has abrupt jumps, which the filter's Gaussian-drift prior does not literally contain; this is a misspecified-but-informative scenario.
3. **Smooth drift.** Category centres in feature space follow an independent random walk; labels are assigned trial-by-trial by proximity to the current drifted centres. This is the **assumption-matched** scenario: the Kalman filter's generative story (prototypes drift as Gaussian random walks) is now literally true of the environment. We expect $\log q$ to be cleanly identified and SBC to pass.

The three scenarios span the spectrum of assumption match — **none → misspecified → exact** — and the SBC signatures should sort in that order.

### Assumption Match Across Scenarios

| Scenario | What drifts in the world | What the Kalman filter assumes | Match |
|---|---|---|---|
| Static | Nothing | Gaussian drift | **Misspecified** (but harmless: $q = 0$ limit works) |
| Contingent shift | Abrupt label reversals | Gaussian drift | **Misspecified** (filter smooths the reversal) |
| Smooth drift | Gaussian-like centre drift | Gaussian drift | **Exact** (this is the model's native story) |

This is the mirror of Chapter 11's GCM-with-decay analysis: the GCM's decay kernel fits contingent shifts as well as drift (because "forget old stuff" is a general-purpose mechanism), while the Kalman filter's smooth-drift prior fits smooth drift cleanly and reversals only approximately.

### Scenario-Aware Simulator

```{r ch13_scenario_simulator}
simulate_prototype_scenario <- function(r_value, q_value, schedule,
                                        scenario = c("static",
                                                     "contingent_shift",
                                                     "drift"),
                                        streak_target      = 8,
                                        drift_traj         = NULL,
                                        initial_mu_cat0    = c(2.5, 2.5),
                                        initial_mu_cat1    = c(2.5, 2.5),
                                        initial_sigma_diag = 10.0) {
  scenario <- match.arg(scenario)
  ntrials  <- nrow(schedule)
  obs_mat  <- as.matrix(schedule[, c("height", "position")])
  base_lab <- schedule$category_feedback
  nfeat    <- 2L

  mu0    <- initial_mu_cat0
  mu1    <- initial_mu_cat1
  sigma0 <- diag(initial_sigma_diag, nfeat)
  sigma1 <- diag(initial_sigma_diag, nfeat)
  R_mat  <- diag(r_value, nfeat)
  Q_mat  <- diag(q_value, nfeat)
  I_mat  <- diag(nfeat)

  label_flip <- FALSE
  streak     <- 0L

  prob_cat1         <- numeric(ntrials)
  sim_response      <- integer(ntrials)
  observed_feedback <- integer(ntrials)
  label_flip_trace  <- logical(ntrials)

  for (i in seq_len(ntrials)) {
    x <- as.numeric(obs_mat[i, ])

    # ── Prediction step ──────────────────────────────────────────────
    sigma0 <- sigma0 + Q_mat
    sigma1 <- sigma1 + Q_mat

    # ── Decision ─────────────────────────────────────────────────────
    cov0 <- sigma0 + R_mat
    cov1 <- sigma1 + R_mat
    lp0  <- tryCatch(mvtnorm::dmvnorm(x, mean = mu0, sigma = cov0, log = TRUE),
                     error = function(e) -Inf)
    lp1  <- tryCatch(mvtnorm::dmvnorm(x, mean = mu1, sigma = cov1, log = TRUE),
                     error = function(e) -Inf)
    m    <- max(lp0, lp1)
    p    <- exp(lp1 - (m + log(exp(lp0 - m) + exp(lp1 - m))))
    p    <- pmax(1e-9, pmin(1 - 1e-9, p))
    prob_cat1[i]    <- p
    sim_response[i] <- rbinom(1, 1, p)

    # ── Observed feedback depends on the scenario ────────────────────
    fb <- switch(scenario,
      static           = base_lab[i],
      contingent_shift = if (label_flip) 1L - base_lab[i] else base_lab[i],
      drift            = {
        d0 <- sum(abs(x - drift_traj$mu0[i, ]))
        d1 <- sum(abs(x - drift_traj$mu1[i, ]))
        as.integer(d1 < d0)
      }
    )
    observed_feedback[i] <- fb
    label_flip_trace[i]  <- label_flip

    # Update the streak/flip state for the contingent scenario
    if (scenario == "contingent_shift") {
      streak <- if (sim_response[i] == fb) streak + 1L else 0L
      if (streak >= streak_target) { label_flip <- !label_flip; streak <- 0L }
    }

    # ── Measurement update for the observed-feedback category ────────
    if (fb == 1) {
      innov  <- x - mu1
      S      <- sigma1 + R_mat
      K      <- sigma1 %*% solve(S)
      mu1    <- as.numeric(mu1 + K %*% innov)
      IK     <- I_mat - K
      sigma1 <- IK %*% sigma1 %*% t(IK) + K %*% R_mat %*% t(K)
      sigma1 <- (sigma1 + t(sigma1)) / 2
    } else {
      innov  <- x - mu0
      S      <- sigma0 + R_mat
      K      <- sigma0 %*% solve(S)
      mu0    <- as.numeric(mu0 + K %*% innov)
      IK     <- I_mat - K
      sigma0 <- IK %*% sigma0 %*% t(IK) + K %*% R_mat %*% t(K)
      sigma0 <- (sigma0 + t(sigma0)) / 2
    }
  }

  schedule |>
    mutate(
      prob_cat1         = prob_cat1,
      sim_response      = sim_response,
      observed_feedback = observed_feedback,
      label_flip        = label_flip_trace,
      correct           = as.integer(sim_response == observed_feedback)
    )
}
```

### Drift Trajectory Generator

The drift scenario pre-computes a trajectory for both category centres as independent bivariate Gaussian random walks.

```{r ch13_drift_trajectory}
make_drift_trajectory <- function(ntrials, drift_sigma = 0.05, seed = 1) {
  set.seed(seed)
  mu0_init <- c(1.75, 3.25)   # Kruschke cat-0 mean (height, position)
  mu1_init <- c(3.25, 1.75)   # Kruschke cat-1 mean
  d0 <- apply(matrix(rnorm(ntrials * 2, 0, drift_sigma), ncol = 2), 2, cumsum)
  d1 <- apply(matrix(rnorm(ntrials * 2, 0, drift_sigma), ncol = 2), 2, cumsum)
  list(
    mu0 = sweep(d0, 2, mu0_init, "+"),
    mu1 = sweep(d1, 2, mu1_init, "+")
  )
}
```

### Scenario Agent Wrapper

```{r ch13_scenario_agent_wrapper}
simulate_prototype_agent_scenario <- function(agent_id, scenario,
                                              r_value, q_value,
                                              subject_seed,
                                              streak_target = 8,
                                              drift_sigma   = 0.05) {
  schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  drift_tr <- if (scenario == "drift")
    make_drift_trajectory(nrow(schedule), drift_sigma, seed = subject_seed)
  else NULL

  out <- simulate_prototype_scenario(
    r_value       = r_value,
    q_value       = q_value,
    schedule      = schedule,
    scenario      = scenario,
    streak_target = streak_target,
    drift_traj    = drift_tr
  )

  out |>
    mutate(
      agent_id     = agent_id,
      scenario     = scenario,
      r_value_true = r_value,
      q_value_true = q_value,
      log_r_true   = log(r_value),
      log_q_true   = log(q_value)
    ) |>
    group_by(agent_id) |>
    mutate(cumulative_accuracy = cumsum(correct) / row_number()) |>
    ungroup()
}
```

### Visualising the Three Scenarios

We simulate 10 agents per cell across a 3 × 3 grid of scenarios × $q$ values (at fixed $r = 1.5$) and plot cumulative accuracy.

```{r ch13_scenario_viz}
scenario_viz_file <- here("simdata", "ch13_scenario_viz.csv")

if (regenerate_simulations || !file.exists(scenario_viz_file)) {
  viz_grid <- expand_grid(
    scenario    = c("static", "contingent_shift", "drift"),
    q_value     = c(0.02, 0.15, 0.6),
    agent_id    = 1:10
  ) |>
    mutate(subject_seed = 30000 + row_number())

  viz_sim <- pmap_dfr(
    list(agent_id     = viz_grid$agent_id,
         scenario     = viz_grid$scenario,
         q_value      = viz_grid$q_value,
         subject_seed = viz_grid$subject_seed),
    function(agent_id, scenario, q_value, subject_seed) {
      simulate_prototype_agent_scenario(
        agent_id     = agent_id,
        scenario     = scenario,
        r_value      = 1.5,
        q_value      = q_value,
        subject_seed = subject_seed
      )
    }
  )
  write_csv(viz_sim, scenario_viz_file)
} else {
  viz_sim <- read_csv(scenario_viz_file, show_col_types = FALSE)
}

ggplot(viz_sim,
       aes(x = trial_within_subject, y = cumulative_accuracy,
           color = factor(q_value_true))) +
  stat_summary(fun = mean, geom = "line",
               aes(group = factor(q_value_true)), linewidth = 1.1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.15,
               aes(fill = factor(q_value_true), group = factor(q_value_true)),
               color = NA) +
  facet_wrap(~scenario, ncol = 3,
             labeller = as_labeller(c(
               static           = "Static Kruschke",
               contingent_shift = "Performance-contingent flips",
               drift            = "Smooth drift"))) +
  scale_color_viridis_d(option = "plasma", name = "q") +
  scale_fill_viridis_d(option  = "plasma", name = "q") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Prototype Learning Curves Across Scenarios",
    subtitle = "Mean ± SE across 10 agents per cell; r = 1.5",
    x = "Trial", y = "Cumulative accuracy"
  ) +
  theme(legend.position = "bottom")
```

**Reading the panels.** The static panel looks much like the plots earlier in the chapter — accuracy climbs, asymptotes below ceiling for larger $q$ (because the covariance never fully shrinks). The contingent-shift panel shows characteristic sawtooth recovery: higher $q$ produces sharper but faster-rebounding dips. The drift panel shows all three $q$ curves plateauing below ceiling, with the plateau highest for the $q$ tuned to the environmental drift amplitude.

### Fitting One Agent per Scenario (Original Parameterisation)

We pick one representative agent per scenario with $r = 1.5$, $q = 0.15$, and fit the Kalman prototype model (original `(log_r, log_q)` parameterisation) to each.

```{r ch13_scenario_fit_per_scenario}
scenario_fit_file <- here("simmodels", "ch13_prototype_scenario_fits.rds")

scenario_single_fit_data <- list(
  static           = simulate_prototype_agent_scenario(
                        1, "static",           r_value = 1.5, q_value = 0.15,
                        subject_seed = 40001),
  contingent_shift = simulate_prototype_agent_scenario(
                        1, "contingent_shift", r_value = 1.5, q_value = 0.15,
                        subject_seed = 40002),
  drift            = simulate_prototype_agent_scenario(
                        1, "drift",            r_value = 1.5, q_value = 0.15,
                        subject_seed = 40003)
)

stan_data_from_scenario_agent <- function(agent_df) {
  list(
    ntrials            = nrow(agent_df),
    nfeatures          = 2L,
    cat_one            = as.integer(agent_df$observed_feedback),
    y                  = as.integer(agent_df$sim_response),
    obs                = as.matrix(agent_df[, c("height", "position")]),
    initial_mu_cat0    = c(2.5, 2.5),
    initial_mu_cat1    = c(2.5, 2.5),
    initial_sigma_diag = 10.0,
    prior_logr_mean    = LOG_R_PRIOR_MEAN,
    prior_logr_sd      = LOG_R_PRIOR_SD,
    prior_logq_mean    = LOG_Q_PRIOR_MEAN,
    prior_logq_sd      = LOG_Q_PRIOR_SD
  )
}

if (regenerate_simulations || !file.exists(scenario_fit_file)) {
  scenario_fits <- purrr::imap(scenario_single_fit_data, function(agent_df, scen) {
    dat <- stan_data_from_scenario_agent(agent_df)
    pf  <- tryCatch(
      mod_prototype_single$pathfinder(data = dat, num_paths = 4,
                                      refresh = 0, seed = 500),
      error = function(e) NULL
    )
    fit <- mod_prototype_single$sample(
      data            = dat,
      init            = if (!is.null(pf)) pf else 0.5,
      seed            = 500,
      chains          = 4,
      parallel_chains = min(4, availableCores()),
      iter_warmup     = 1000,
      iter_sampling   = 1500,
      refresh         = 0,
      adapt_delta     = 0.9
    )
    fit$save_object(here("simmodels",
                         paste0("ch13_prototype_scenario_", scen, ".rds")))
    fit
  })
  saveRDS(lapply(scenario_fits, function(f) f$output_files()),
          scenario_fit_file)
  cat("Scenario fits computed and saved.\n")
} else {
  scenario_fits <- setNames(
    lapply(names(scenario_single_fit_data), function(scen) {
      readRDS(here("simmodels",
                   paste0("ch13_prototype_scenario_", scen, ".rds")))
    }),
    names(scenario_single_fit_data)
  )
  cat("Loaded existing scenario fits.\n")
}
```

### MCMC Diagnostic Battery Across Scenarios

```{r ch13_scenario_diagnostics}
scenario_diag_tbls <- purrr::imap_dfr(scenario_fits, function(fit, scen) {
  tbl <- diagnostic_summary_table(fit, pars = c("log_r", "log_q",
                                                "r_value", "q_value"))
  tbl |> mutate(scenario = scen)
})
print(scenario_diag_tbls |>
        dplyr::select(scenario, metric, value, threshold, pass))

walk(names(scenario_fits), function(scen) {
  cat("=== Pair plot:", scen, "===\n")
  print(bayesplot::mcmc_pairs(
    scenario_fits[[scen]]$draws(c("log_r", "log_q")),
    diag_fun     = "dens",
    off_diag_fun = "hex",
    np           = nuts_params(scenario_fits[[scen]])
  ))
})
```

**What the pair plots reveal.** On static data the `(log_r, log_q)` posterior sits on a strongly negatively-correlated ridge — the canonical identifiability failure. On contingent-shift data the ridge is shorter and tilted: reversal dynamics load on $q$ specifically, so the degrees of freedom are no longer degenerate. On drift data the posterior is roughly circular in the two parameters; this is the visual signature of the assumption-matched case.

### Prior → Posterior Updates Across Scenarios

```{r ch13_scenario_prior_posterior}
scenario_pp_df <- purrr::imap_dfr(scenario_fits, function(fit, scen) {
  d <- as_draws_df(fit$draws(c("log_r", "log_q")))
  bind_rows(
    tibble(scenario = scen, parameter = "log_r", value = d$log_r, source = "posterior"),
    tibble(scenario = scen, parameter = "log_q", value = d$log_q, source = "posterior")
  )
})

set.seed(21)
n_prior <- 4000
prior_ref <- bind_rows(
  tibble(parameter = "log_r", value = rnorm(n_prior, LOG_R_PRIOR_MEAN, LOG_R_PRIOR_SD),
         source = "prior"),
  tibble(parameter = "log_q", value = rnorm(n_prior, LOG_Q_PRIOR_MEAN, LOG_Q_PRIOR_SD),
         source = "prior")
) |>
  tidyr::crossing(scenario = c("static", "contingent_shift", "drift"))

truth_lines <- tibble(
  parameter = c("log_r", "log_q"),
  value     = c(log(1.5), log(0.15))
)

ggplot(bind_rows(prior_ref, scenario_pp_df),
       aes(x = value, fill = source, color = source)) +
  geom_density(alpha = 0.4, linewidth = 0.5) +
  geom_vline(data = truth_lines, aes(xintercept = value),
             color = "red", linetype = "dashed") +
  facet_grid(scenario ~ parameter, scales = "free") +
  scale_fill_manual(values  = c(prior = "grey70", posterior = "#009E73")) +
  scale_color_manual(values = c(prior = "grey50", posterior = "#006D4E")) +
  labs(
    title    = "Prior → Posterior Update: Prototype Model Across Scenarios",
    subtitle = "Red dashed line = true generating value used for the single-agent fit",
    x = NULL, y = "Density"
  )
```

The density of `log_q` narrows visibly across scenarios: from prior-like on static data (information-starved), to sharply peaked around the truth on drift data (assumption-matched).

### One-Shot Parameter Recovery Across Scenarios

We sweep a 3 × 3 grid of $(r, q)$ truths with 5 iterations per cell per scenario (135 fits) and plot recovery on the log scale.

```{r ch13_scenario_recovery_sweep}
scenario_recovery_file <- here("simdata", "ch13_scenario_recovery_results.rds")

run_proto_scenario_recovery_iter <- function(scenario, r_true, q_true, iteration) {
  subject_seed <- iteration * 10000 +
                   round(r_true * 100) +
                   round(q_true * 1000) +
                   switch(scenario, static = 1, contingent_shift = 2, drift = 3)

  agent_df <- simulate_prototype_agent_scenario(
    agent_id     = iteration,
    scenario     = scenario,
    r_value      = r_true,
    q_value      = q_true,
    subject_seed = subject_seed
  )
  dat <- stan_data_from_scenario_agent(agent_df)
  pf  <- tryCatch(
    mod_prototype_single$pathfinder(data = dat, num_paths = 4,
                                    refresh = 0, seed = subject_seed),
    error = function(e) NULL
  )
  fit <- tryCatch(
    mod_prototype_single$sample(
      data            = dat,
      init            = if (!is.null(pf)) pf else 0.5,
      seed            = subject_seed,
      chains          = 2,
      parallel_chains = 2,
      iter_warmup     = 800,
      iter_sampling   = 1000,
      refresh         = 0,
      adapt_delta     = 0.9
    ),
    error = function(e) { message("Fit error: ", e$message); NULL }
  )
  if (is.null(fit)) return(NULL)

  s <- fit$summary(variables = c("log_r", "log_q"))
  diag <- fit$diagnostic_summary(quiet = TRUE)
  tibble(
    scenario       = scenario,
    r_true         = r_true,
    q_true         = q_true,
    iteration      = iteration,
    log_r_mean     = s$mean[s$variable == "log_r"],
    log_q_mean     = s$mean[s$variable == "log_q"],
    log_r_q5       = s$q5[s$variable == "log_r"],
    log_r_q95      = s$q95[s$variable == "log_r"],
    log_q_q5       = s$q5[s$variable == "log_q"],
    log_q_q95      = s$q95[s$variable == "log_q"],
    had_divergence = any(diag$num_divergent > 0)
  )
}

if (regenerate_simulations || !file.exists(scenario_recovery_file)) {
  recovery_plan <- expand_grid(
    scenario  = c("static", "contingent_shift", "drift"),
    r_true    = c(0.5, 1.5, 3.0),
    q_true    = c(0.02, 0.15, 0.6),
    iteration = 1:5
  )

  plan(multisession, workers = max(1, availableCores() - 1))
  scenario_recovery <- future_pmap_dfr(
    list(scenario  = recovery_plan$scenario,
         r_true    = recovery_plan$r_true,
         q_true    = recovery_plan$q_true,
         iteration = recovery_plan$iteration),
    run_proto_scenario_recovery_iter,
    .options = furrr_options(seed = TRUE, scheduling = 1)
  )
  plan(sequential)
  saveRDS(scenario_recovery, scenario_recovery_file)
  cat("Scenario recovery sweep complete.\n")
} else {
  scenario_recovery <- readRDS(scenario_recovery_file)
  cat("Loaded existing scenario recovery sweep.\n")
}

scenario_recovery_long <- scenario_recovery |>
  dplyr::select(scenario, r_true, q_true, iteration, had_divergence,
                log_r_mean, log_q_mean) |>
  pivot_longer(cols = c(log_r_mean, log_q_mean),
               names_to = "variable", values_to = "estimated") |>
  mutate(
    parameter = if_else(variable == "log_r_mean", "log_r", "log_q"),
    truth     = if_else(parameter == "log_r", log(r_true), log(q_true))
  )

ggplot(scenario_recovery_long, aes(x = truth, y = estimated,
                                    color = had_divergence)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.7, size = 2) +
  facet_grid(parameter ~ scenario, scales = "free") +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00"),
                     name = "Had divergence") +
  labs(
    title    = "One-Shot Parameter Recovery for the Kalman Prototype Model",
    subtitle = "Posterior mean vs true generating value; 5 iterations per (r, q) cell, 3 scenarios",
    x = "True value (log scale)", y = "Estimated value (posterior mean)"
  ) +
  theme(legend.position = "bottom")
```

**Interpretation.** The `log_r` row is near-diagonal across scenarios — $r$ is identified by within-category decision sharpness, which every scenario exercises. The `log_q` row is the diagnostic: diffuse scatter on static data (the parameter is under-constrained), tighter on drift, and comparably tight on contingent-shift. The contingent-shift result is noteworthy — the data-generating process has abrupt jumps that the filter's smooth-drift prior cannot exactly reproduce, yet recovery still works because post-shift recovery speed loads specifically on $q$.

### Simulation-Based Calibration Across Scenarios

SBC is the formal calibration test. Each scenario gets its own generator drawing $(\log r, \log q)$ from their priors.

```{r ch13_scenario_sbc_generator}
make_proto_sbc_generator <- function(scenario) {
  SBC_generator_function(
    function() {
      log_r_true <- rnorm(1, LOG_R_PRIOR_MEAN, LOG_R_PRIOR_SD)
      log_q_true <- rnorm(1, LOG_Q_PRIOR_MEAN, LOG_Q_PRIOR_SD)
      r_true     <- exp(log_r_true)
      q_true     <- exp(log_q_true)

      subject_seed <- sample.int(1e8, 1)
      agent_df <- simulate_prototype_agent_scenario(
        agent_id     = 1,
        scenario     = scenario,
        r_value      = r_true,
        q_value      = q_true,
        subject_seed = subject_seed
      )

      list(
        variables = list(log_r = log_r_true, log_q = log_q_true),
        generated = stan_data_from_scenario_agent(agent_df)
      )
    }
  )
}

run_proto_scenario_sbc <- function(scenario, n_sims = 200) {
  cache_file <- here("simdata", paste0("ch13_proto_sbc_", scenario, ".rds"))
  if (!regenerate_simulations && file.exists(cache_file)) {
    cat("Loaded cached prototype SBC for scenario:", scenario, "\n")
    return(readRDS(cache_file))
  }
  gen <- make_proto_sbc_generator(scenario)
  backend <- SBC_backend_cmdstan_sample(
    mod_prototype_single,
    iter_warmup   = 800,
    iter_sampling = 1000,
    chains        = 2,
    adapt_delta   = 0.9,
    refresh       = 0
  )
  datasets <- generate_datasets(gen, n_sims = n_sims)
  res <- compute_SBC(
    datasets,
    backend,
    cache_mode     = "results",
    cache_location = here("simdata", paste0("ch13_proto_sbc_cache_", scenario)),
    keep_fits      = FALSE
  )
  saveRDS(res, cache_file)
  res
}

proto_sbc_static <- run_proto_scenario_sbc("static",           n_sims = 200)
proto_sbc_shift  <- run_proto_scenario_sbc("contingent_shift", n_sims = 200)
proto_sbc_drift  <- run_proto_scenario_sbc("drift",            n_sims = 200)
```

```{r ch13_scenario_sbc_ecdf}
print(plot_ecdf_diff(proto_sbc_static) + ggtitle("SBC: static"))
print(plot_ecdf_diff(proto_sbc_shift)  + ggtitle("SBC: performance-contingent shifts"))
print(plot_ecdf_diff(proto_sbc_drift)  + ggtitle("SBC: smooth drift"))

print(plot_rank_hist(proto_sbc_static) + ggtitle("Rank histograms — static"))
print(plot_rank_hist(proto_sbc_shift)  + ggtitle("Rank histograms — contingent shifts"))
print(plot_rank_hist(proto_sbc_drift)  + ggtitle("Rank histograms — drift"))
```

```{r ch13_scenario_sbc_recovery_plots}
combine_proto_sbc_stats <- function(sbc_obj, scenario) {
  sbc_obj$stats |> mutate(scenario = scenario)
}

proto_sbc_stats <- bind_rows(
  combine_proto_sbc_stats(proto_sbc_static, "static"),
  combine_proto_sbc_stats(proto_sbc_shift,  "contingent_shift"),
  combine_proto_sbc_stats(proto_sbc_drift,  "drift")
) |>
  dplyr::filter(variable %in% c("log_r", "log_q"))

ggplot(proto_sbc_stats, aes(x = simulated_value, y = mean)) +
  geom_point(alpha = 0.35, color = "#0072B2") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_grid(variable ~ scenario, scales = "free") +
  labs(
    title    = "SBC-Level Parameter Recovery: Kalman Prototype",
    subtitle = "Posterior mean vs true generating value, pooled across SBC simulations",
    x = "True value", y = "Posterior mean"
  )

proto_sbc_stats |>
  group_by(scenario, variable) |>
  summarise(
    sbc_mean_bias = mean(mean - simulated_value, na.rm = TRUE),
    sbc_rmse      = sqrt(mean((mean - simulated_value)^2, na.rm = TRUE)),
    sbc_corr      = cor(mean, simulated_value,
                        use = "complete.obs", method = "pearson"),
    n_sims        = dplyr::n(),
    .groups       = "drop"
  ) |>
  print(n = Inf)
```

**Reading the three SBC results.** The static SBC for `log_q` shows the ECDF-difference curve drifting outside the simultaneous band and the rank histogram humping — posterior credible intervals do not contain the truth at the nominal rate. The contingent-shift SBC is noticeably better on `log_q` but may still show mild systematic deviation (a signature of the smooth-drift-vs-abrupt-jump misspecification). The drift SBC should show both parameters' curves squarely inside the band and rank histograms close to uniform. This is the formal version of the identifiability claim: calibration is achieved when task structure matches the model's generative story.

---

## Reparameterizing to Break the $q$/$r$ Ridge

The static-scenario pair plot showed a heavy negatively-correlated ridge on `(log_r, log_q)`. The ridge is *geometric* — a property of the posterior shape, not of the information content. It makes HMC sampling less efficient (ESS-per-draw falls as the posterior elongates into a thin diagonal), makes SBC harder to interpret (the failure mode could be either log_r or log_q but really it's their difference), and obscures what the data actually say.

### The SNR Parameterisation

Rewrite the two parameters as

$$\log r, \qquad \log \rho = \log q - \log r.$$

Here $\rho = q/r$ is the drift-to-observation-noise ratio — a dimensionless scale that directly controls the Kalman filter's steady-state gain. The original `log_q` is now a *derived* quantity: `log_q = log_r + log_rho`. If the data pin down the *sum* $\log r + \log \rho$ (i.e. $\log q$) cleanly but not the *difference* $\log r - \log \rho$, the SNR parameterisation represents this cleanly as a wide marginal on one of the two parameters — no diagonal ridge, no misleading pair plot.

We use Normal priors on both axes:

```{r ch13_snr_prior_constants}
# SNR parameterisation priors.
# log_r ~ Normal(0, 1)      — same as the original parameterisation.
# log_rho ~ Normal(-2, 1)   — centres rho at 0.135 (drift small vs. r).
# Marginal on log_q = log_r + log_rho is Normal(-2, sqrt(2)) ≈ Normal(-2, 1.41).
# Slightly wider than the original Normal(-2, 1), but the increase is modest and
# the shape of the implied prior predictive is comparable.
LOG_RHO_PRIOR_MEAN <- -2.0
LOG_RHO_PRIOR_SD   <- 1.0
```

### The SNR Stan Model

```{r ch13_snr_stan_compile}
stan_file_proto_snr <- here("stan", "ch12_prototype_single_snr.stan")
mod_prototype_snr   <- cmdstan_model(stan_file_proto_snr)
```

See `stan/ch12_prototype_single_snr.stan`. The only parameter-block change is `log_rho` replacing `log_q`; `log_q` is computed in transformed parameters.

### Fitting Both Parameterisations to the Same Static Data

```{r ch13_snr_fit_static}
snr_static_file <- here("simmodels", "ch13_proto_snr_static_fit.rds")

static_agent <- scenario_single_fit_data$static  # same dataset as above
snr_static_data <- list(
  ntrials            = nrow(static_agent),
  nfeatures          = 2L,
  cat_one            = as.integer(static_agent$observed_feedback),
  y                  = as.integer(static_agent$sim_response),
  obs                = as.matrix(static_agent[, c("height", "position")]),
  initial_mu_cat0    = c(2.5, 2.5),
  initial_mu_cat1    = c(2.5, 2.5),
  initial_sigma_diag = 10.0,
  prior_logr_mean    = LOG_R_PRIOR_MEAN,
  prior_logr_sd      = LOG_R_PRIOR_SD,
  prior_logrho_mean  = LOG_RHO_PRIOR_MEAN,
  prior_logrho_sd    = LOG_RHO_PRIOR_SD
)

if (regenerate_simulations || !file.exists(snr_static_file)) {
  pf_snr <- tryCatch(
    mod_prototype_snr$pathfinder(data = snr_static_data, num_paths = 4,
                                 refresh = 0, seed = 777),
    error = function(e) NULL
  )
  fit_proto_snr_static <- mod_prototype_snr$sample(
    data            = snr_static_data,
    init            = if (!is.null(pf_snr)) pf_snr else 0.5,
    seed            = 777,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1000,
    iter_sampling   = 1500,
    refresh         = 0,
    adapt_delta     = 0.9
  )
  fit_proto_snr_static$save_object(snr_static_file)
} else {
  fit_proto_snr_static <- readRDS(snr_static_file)
}

orig_static_fit <- scenario_fits$static

cat("=== Original (log_r, log_q) parameterisation ===\n")
print(orig_static_fit$summary(variables = c("log_r", "log_q", "r_value", "q_value")))
cat("\n=== SNR (log_r, log_rho) parameterisation ===\n")
print(fit_proto_snr_static$summary(variables = c("log_r", "log_rho", "log_q",
                                                  "r_value", "q_value")))
```

### Geometry Comparison

```{r ch13_snr_pair_compare}
p_orig <- bayesplot::mcmc_pairs(
  orig_static_fit$draws(c("log_r", "log_q")),
  diag_fun = "dens", off_diag_fun = "hex",
  np       = nuts_params(orig_static_fit)
)
p_snr  <- bayesplot::mcmc_pairs(
  fit_proto_snr_static$draws(c("log_r", "log_rho")),
  diag_fun = "dens", off_diag_fun = "hex",
  np       = nuts_params(fit_proto_snr_static)
)

print(p_orig)
print(p_snr)

# Side-by-side ESS comparison
ess_compare <- bind_rows(
  orig_static_fit$summary(c("log_r", "log_q"),
                          "ess_bulk", "ess_tail") |>
    mutate(parameterisation = "original"),
  fit_proto_snr_static$summary(c("log_r", "log_rho"),
                               "ess_bulk", "ess_tail") |>
    mutate(parameterisation = "SNR")
)
print(ess_compare)
```

**The expected pattern.** The original pair plot shows the heavy negative ridge; the SNR pair plot shows a wide-but-roughly-circular cloud on `(log_r, log_rho)`. ESS-per-iteration should improve for both parameters under SNR, because HMC no longer has to traverse a long thin diagonal. The posterior over `log_q` (a derived quantity under SNR) is *identical in substance* — the two models imply the same likelihood — but the Stan sampler reaches it more efficiently.

### SBC for the SNR Model on Drift Data

We rerun SBC on the drift scenario under the SNR parameterisation as the ultimate test: *assumption-matched data + decorrelated parameterisation = cleanest possible calibration.*

```{r ch13_snr_sbc_drift}
snr_sbc_drift_file <- here("simdata", "ch13_proto_snr_sbc_drift.rds")

make_snr_sbc_generator <- function(scenario) {
  SBC_generator_function(
    function() {
      log_r_true   <- rnorm(1, LOG_R_PRIOR_MEAN,   LOG_R_PRIOR_SD)
      log_rho_true <- rnorm(1, LOG_RHO_PRIOR_MEAN, LOG_RHO_PRIOR_SD)
      log_q_true   <- log_r_true + log_rho_true
      r_true       <- exp(log_r_true)
      q_true       <- exp(log_q_true)

      agent_df <- simulate_prototype_agent_scenario(
        agent_id     = 1,
        scenario     = scenario,
        r_value      = r_true,
        q_value      = q_true,
        subject_seed = sample.int(1e8, 1)
      )

      list(
        variables = list(log_r = log_r_true,
                         log_rho = log_rho_true,
                         log_q = log_q_true),
        generated = list(
          ntrials            = nrow(agent_df),
          nfeatures          = 2L,
          cat_one            = as.integer(agent_df$observed_feedback),
          y                  = as.integer(agent_df$sim_response),
          obs                = as.matrix(agent_df[, c("height", "position")]),
          initial_mu_cat0    = c(2.5, 2.5),
          initial_mu_cat1    = c(2.5, 2.5),
          initial_sigma_diag = 10.0,
          prior_logr_mean    = LOG_R_PRIOR_MEAN,
          prior_logr_sd      = LOG_R_PRIOR_SD,
          prior_logrho_mean  = LOG_RHO_PRIOR_MEAN,
          prior_logrho_sd    = LOG_RHO_PRIOR_SD
        )
      )
    }
  )
}

if (regenerate_simulations || !file.exists(snr_sbc_drift_file)) {
  gen <- make_snr_sbc_generator("drift")
  backend <- SBC_backend_cmdstan_sample(
    mod_prototype_snr,
    iter_warmup   = 800,
    iter_sampling = 1000,
    chains        = 2,
    adapt_delta   = 0.9,
    refresh       = 0
  )
  datasets <- generate_datasets(gen, n_sims = 200)
  snr_sbc_drift <- compute_SBC(
    datasets, backend,
    cache_mode     = "results",
    cache_location = here("simdata", "ch13_proto_snr_sbc_cache_drift"),
    keep_fits      = FALSE
  )
  saveRDS(snr_sbc_drift, snr_sbc_drift_file)
} else {
  snr_sbc_drift <- readRDS(snr_sbc_drift_file)
}

print(plot_ecdf_diff(snr_sbc_drift) + ggtitle("SNR-parameterised SBC on drift data"))
print(plot_rank_hist(snr_sbc_drift))
```

Under this combination we expect all three curves (`log_r`, `log_rho`, and the derived `log_q`) to stay inside the simultaneous band with flat rank histograms. Reparameterisation alone does not manufacture information: on static data, the SNR model will still show the drift-related direction as a wide marginal (just on `log_rho` now, not on a diagonal ridge). It is the *combination* of assumption-matched task and decorrelated parameterisation that closes the loop.

---

## Discussion: When Does $q$ Become Identifiable?

The six validation panels (three scenarios × two parameterisations) give the cleanest summary of the identifiability argument.

* **Static + original.** The posterior sits on a heavy negatively-correlated ridge. SBC for `log_q` fails; parameter recovery is diffuse. This is the problem the chapter opened with.
* **Static + SNR.** The ridge disappears. The poorly-identified direction is now cleanly visible as a wide marginal on `log_rho`. SBC still fails for the drift-related direction — reparameterisation does not manufacture information.
* **Contingent-shift + original.** Post-flip recovery dynamics load on $q$ directly, so the ridge shrinks and SBC for `log_q` passes (or nearly passes). Caveat: the filter's smooth-drift assumption is technically violated by abrupt jumps, so posterior predictive checks may reveal systematic underperformance right after each flip.
* **Contingent-shift + SNR.** Similar to contingent-shift + original, but with tighter posterior marginals and better MCMC geometry.
* **Smooth drift + original.** Assumption-matched generative story. The posterior is roughly circular and SBC passes for both parameters.
* **Smooth drift + SNR.** The best-case scenario: decorrelated parameterisation + assumption-matched task. Both SBC ECDFs inside the band, both rank histograms uniform, MCMC geometry clean.

**Paralleling Chapter 11.** The GCM-with-decay analysis in Chapter 11 showed that the decay parameter $\lambda$ becomes identifiable when the task exercises forgetting — whether through reversals or drift. The mechanism there is general-purpose ("downweight old stuff"), so reversals and drift work comparably well. The Kalman filter's $q$ is assumption-specific (it encodes a *Gaussian* drift prior), so the assumption-matched smooth-drift task gives the cleanest result; reversals work but with a mild misspecification signature. Both chapters make the same underlying point: *identifiability is a joint property of the model and the task, not of the model alone.*

**Implication for the multilevel analysis that follows.** The multilevel prototype model below uses the static paradigm by default to stay comparable with prior drafts of this chapter. Readers using this model for research should consider re-running the multilevel analysis on drift data when possible; the population-level `log_q` posterior that looked wide before is not a sign that multilevel inference failed — it was a sign that the task was not giving the model what it needed to pin $q$ down.

---

## Multilevel Prototype Model: Population Inference Across Subjects

Real experiments involve many participants, each bringing their own observation-noise tolerance and drift rate. Some learners trust individual examples heavily (low $r$); others discount them (high $r$). Some learners have more stable category representations (low $q$); others show faster adaptation (high $q$). The **multilevel prototype model** captures this variation by placing population distributions over both `log_r` and `log_q` and partially pooling information across subjects.

The structure is the minimal multilevel extension of the two-parameter single-subject model:

$$\text{pop\_log\_r\_mean} \sim \mathcal{N}(0, 1), \quad \text{pop\_log\_r\_sd} \sim \text{Exponential}(2)$$
$$z_{r,j} \sim \mathcal{N}(0, 1), \quad \log r_j = \text{pop\_log\_r\_mean} + z_{r,j} \cdot \text{pop\_log\_r\_sd}$$
$$\text{pop\_log\_q\_mean} \sim \mathcal{N}(-2, 1), \quad \text{pop\_log\_q\_sd} \sim \text{Exponential}(2)$$
$$z_{q,j} \sim \mathcal{N}(0, 1), \quad \log q_j = \text{pop\_log\_q\_mean} + z_{q,j} \cdot \text{pop\_log\_q\_sd}$$

The non-centred parameterisation (NCP) for the per-subject offsets avoids the funnel geometry that centred parameterisations produce when the population SD is small. With two per-subject parameters, the SBC rank histograms (below) show six curves: two population means, two population SDs, and two individual-level curves (`subj_log_r[1]`, `subj_log_q[1]`).

### Multilevel Forward Simulation

```{r ch13_multilevel_forward_sim}
# Simulate a full multilevel dataset from specified population parameters.
simulate_multilevel_prototype <- function(
    n_subjects,
    stimulus_info,
    n_blocks,
    base_seed,
    pop_log_r_mean,
    pop_log_r_sd,
    pop_log_q_mean = LOG_Q_PRIOR_MEAN,
    pop_log_q_sd   = 0.5) {

  set.seed(base_seed)
  z_log_r     <- rnorm(n_subjects, 0, 1)
  subj_log_r  <- pop_log_r_mean + pop_log_r_sd * z_log_r
  subj_r      <- exp(subj_log_r)

  z_log_q     <- rnorm(n_subjects, 0, 1)
  subj_log_q  <- pop_log_q_mean + pop_log_q_sd * z_log_q
  subj_q      <- exp(subj_log_q)

  true_params <- tibble(
    agent_id   = seq_len(n_subjects),
    log_r_true = subj_log_r,
    r_true     = subj_r,
    log_q_true = subj_log_q,
    q_true     = subj_q
  )

  sim_data <- map_dfr(seq_len(n_subjects), function(j) {
    sched <- make_subject_schedule(stimulus_info, n_blocks,
                                   seed = base_seed * 100 + j)
    obs   <- as.matrix(sched[, c("height", "position")])
    res   <- prototype_kalman(
      r_value            = subj_r[j],
      q_value            = subj_q[j],
      obs                = obs,
      cat_one            = sched$category_feedback,
      initial_mu         = list(rep(2.5, 2), rep(2.5, 2)),
      initial_sigma_diag = 10.0
    )
    sched |>
      mutate(
        agent_id     = j,
        sim_response = res$sim_response,
        correct      = as.integer(category_feedback == sim_response)
      ) |>
      group_by(agent_id) |>
      mutate(performance = cumsum(correct) / row_number()) |>
      ungroup()
  })

  list(data = sim_data, true_params = true_params)
}
```

```{r ch13_simulate_multilevel_data}
# Population parameters for multilevel simulation
pop_params_ml_proto <- list(
  pop_log_r_mean = log(1.0),   # prior median r = 1.0
  pop_log_r_sd   = 0.5,        # moderate between-subject heterogeneity
  n_subjects     = 20
)

sim_file_ml_proto    <- here("simdata", "ch13_proto_ml_sim_data.csv")
true_params_file_ml  <- here("simdata", "ch13_proto_ml_true_params.csv")

if (regenerate_simulations ||
    !file.exists(sim_file_ml_proto) ||
    !file.exists(true_params_file_ml)) {

  ml_sim_out <- simulate_multilevel_prototype(
    n_subjects     = pop_params_ml_proto$n_subjects,
    stimulus_info  = stimulus_info,
    n_blocks       = n_blocks,
    base_seed      = 42,
    pop_log_r_mean = pop_params_ml_proto$pop_log_r_mean,
    pop_log_r_sd   = pop_params_ml_proto$pop_log_r_sd
  )
  multilevel_proto_data    <- ml_sim_out$data
  multilevel_proto_params  <- ml_sim_out$true_params

  write_csv(multilevel_proto_data,   sim_file_ml_proto)
  write_csv(multilevel_proto_params, true_params_file_ml)
  cat("Multilevel prototype simulation complete. Results saved.\n")
} else {
  multilevel_proto_data   <- read_csv(sim_file_ml_proto,   show_col_types = FALSE)
  multilevel_proto_params <- read_csv(true_params_file_ml, show_col_types = FALSE)
  cat("Loaded existing multilevel prototype simulation results.\n")
}

ggplot(multilevel_proto_data,
       aes(x = trial_within_subject, y = performance, group = agent_id)) +
  geom_line(alpha = 0.25, color = "grey40") +
  stat_summary(fun = mean, geom = "line", aes(group = 1),
               color = "#009E73", linewidth = 1.5) +
  labs(
    title    = "Simulated Learning Curves: Multilevel Prototype Model",
    subtitle = "Grey lines: individual agents; green: group mean",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  ylim(0, 1)
```

### Implementing the Multilevel Prototype Model in Stan

The multilevel Stan model extends the single-subject architecture by:

1. Adding population hyperparameters `pop_log_r_mean`, `pop_log_r_sd`, `pop_log_q_mean`, and `pop_log_q_sd` to the `parameters` block.
2. Adding per-subject NCP offsets `z_log_r[j] ~ Normal(0, 1)` and `z_log_q[j] ~ Normal(0, 1)`.
3. Looping the Kalman filter (with prediction step) over subjects in `transformed parameters`, exactly as the GCM does in Chapter 11.

The resulting model has $4 + 2J$ parameters ($J$ is the number of subjects) — still relatively compact compared to exemplar models — which makes posterior geometry and SBC diagnostics particularly legible.

```{r ch13_multilevel_stan_model}
prototype_ml_stan <- "
// Kalman Filter Prototype Model — Multilevel (Non-Centred Parameterisation)
//
// New relative to the single-subject model:
//   1. Population hyperparameters for log_r AND log_q.
//   2. Per-subject NCP offsets z_log_r[j] and z_log_q[j] ~ Normal(0, 1).
//   3. Subject-indexed Kalman filter loop with prediction step in transformed parameters.
//   4. ALL prior hyperparameters passed through the data block.

data {
  int<lower=1> N_total;
  int<lower=1> N_subjects;
  int<lower=1> N_features;

  array[N_subjects] int<lower=1, upper=N_total> subj_start;
  array[N_subjects] int<lower=1, upper=N_total> subj_end;

  array[N_total] int<lower=0, upper=1> y;
  array[N_total] int<lower=0, upper=1> cat_one;
  array[N_total, N_features] real obs;

  // Fixed structural values (same as single-subject; not inferred)
  vector[N_features] initial_mu_cat0;
  vector[N_features] initial_mu_cat1;
  real<lower=0> initial_sigma_diag;

  // Prior hyperparameters for r
  real pop_log_r_mean_prior_mean;
  real<lower=0> pop_log_r_mean_prior_sd;
  real<lower=0> pop_log_r_sd_prior_rate;

  // Prior hyperparameters for q
  real pop_log_q_mean_prior_mean;
  real<lower=0> pop_log_q_mean_prior_sd;
  real<lower=0> pop_log_q_sd_prior_rate;
}

parameters {
  real pop_log_r_mean;
  real<lower=0> pop_log_r_sd;
  vector[N_subjects] z_log_r;

  real pop_log_q_mean;
  real<lower=0> pop_log_q_sd;
  vector[N_subjects] z_log_q;
}

transformed parameters {
  vector[N_subjects] subj_log_r;
  vector<lower=0>[N_subjects] subj_r;
  vector[N_subjects] subj_log_q;
  vector<lower=0>[N_subjects] subj_q;
  vector<lower=1e-9, upper=1-1e-9>[N_total] prob_cat1;

  for (j in 1:N_subjects) {
    subj_log_r[j] = pop_log_r_mean + z_log_r[j] * pop_log_r_sd;
    subj_r[j]     = exp(subj_log_r[j]);
    subj_log_q[j] = pop_log_q_mean + z_log_q[j] * pop_log_q_sd;
    subj_q[j]     = exp(subj_log_q[j]);
  }

  {
    matrix[N_features, N_features] I_mat =
      diag_matrix(rep_vector(1.0, N_features));

    for (j in 1:N_subjects) {
      vector[N_features] mu_cat0 = initial_mu_cat0;
      vector[N_features] mu_cat1 = initial_mu_cat1;
      matrix[N_features, N_features] sigma_cat0 =
        diag_matrix(rep_vector(initial_sigma_diag, N_features));
      matrix[N_features, N_features] sigma_cat1 =
        diag_matrix(rep_vector(initial_sigma_diag, N_features));
      matrix[N_features, N_features] r_matrix =
        diag_matrix(rep_vector(subj_r[j], N_features));
      matrix[N_features, N_features] q_matrix =
        diag_matrix(rep_vector(subj_q[j], N_features));

      for (i in subj_start[j]:subj_end[j]) {
        vector[N_features] x = to_vector(obs[i]);

        // ── Prediction step: add process noise to both categories ──────────
        sigma_cat0 = sigma_cat0 + q_matrix;
        sigma_cat1 = sigma_cat1 + q_matrix;

        // ── Decision ────────────────────────────────────────────────────────
        matrix[N_features, N_features] cov0 = sigma_cat0 + r_matrix;
        matrix[N_features, N_features] cov1 = sigma_cat1 + r_matrix;
        real log_p0 = multi_normal_lpdf(x | mu_cat0, cov0);
        real log_p1 = multi_normal_lpdf(x | mu_cat1, cov1);
        real p_i    = exp(log_p1 - log_sum_exp(log_p0, log_p1));
        prob_cat1[i] = fmax(1e-9, fmin(1 - 1e-9, p_i));

        // ── Update ───────────────────────────────────────────────────────────
        if (cat_one[i] == 1) {
          vector[N_features] innov = x - mu_cat1;
          matrix[N_features, N_features] S  = sigma_cat1 + r_matrix;
          matrix[N_features, N_features] K  = mdivide_right_spd(sigma_cat1, S);
          matrix[N_features, N_features] IK = I_mat - K;
          mu_cat1    = mu_cat1 + K * innov;
          sigma_cat1 = IK * sigma_cat1 * IK' + K * r_matrix * K';
          sigma_cat1 = 0.5 * (sigma_cat1 + sigma_cat1');
        } else {
          vector[N_features] innov = x - mu_cat0;
          matrix[N_features, N_features] S  = sigma_cat0 + r_matrix;
          matrix[N_features, N_features] K  = mdivide_right_spd(sigma_cat0, S);
          matrix[N_features, N_features] IK = I_mat - K;
          mu_cat0    = mu_cat0 + K * innov;
          sigma_cat0 = IK * sigma_cat0 * IK' + K * r_matrix * K';
          sigma_cat0 = 0.5 * (sigma_cat0 + sigma_cat0');
        }
      }
    }
  }
}

model {
  target += normal_lpdf(pop_log_r_mean | pop_log_r_mean_prior_mean,
                                         pop_log_r_mean_prior_sd);
  target += exponential_lpdf(pop_log_r_sd | pop_log_r_sd_prior_rate);
  target += std_normal_lpdf(z_log_r);

  target += normal_lpdf(pop_log_q_mean | pop_log_q_mean_prior_mean,
                                         pop_log_q_mean_prior_sd);
  target += exponential_lpdf(pop_log_q_sd | pop_log_q_sd_prior_rate);
  target += std_normal_lpdf(z_log_q);

  target += bernoulli_lpmf(y | prob_cat1);
}

generated quantities {
  vector[N_total] log_lik;
  real lprior;
  for (i in 1:N_total)
    log_lik[i] = bernoulli_lpmf(y[i] | prob_cat1[i]);
  lprior = normal_lpdf(pop_log_r_mean | pop_log_r_mean_prior_mean, pop_log_r_mean_prior_sd) +
           exponential_lpdf(pop_log_r_sd | pop_log_r_sd_prior_rate) +
           std_normal_lpdf(z_log_r) +
           normal_lpdf(pop_log_q_mean | pop_log_q_mean_prior_mean, pop_log_q_mean_prior_sd) +
           exponential_lpdf(pop_log_q_sd | pop_log_q_sd_prior_rate) +
           std_normal_lpdf(z_log_q);
}
"

stan_file_proto_ml <- here("stan", "ch13_prototype_ml.stan")
write_stan_file(prototype_ml_stan, dir = here("stan"),
                basename = "ch13_prototype_ml.stan")
mod_prototype_ml <- cmdstan_model(stan_file_proto_ml)
cat("Multilevel prototype Stan model compiled successfully.\n")
```

### Fitting the Multilevel Prototype Model

```{r ch13_fit_multilevel_prototype}
# ── Build the Stan data list ─────────────────────────────────────────────────
proto_ml_sorted <- multilevel_proto_data |>
  mutate(subj_id_stan = as.integer(factor(agent_id))) |>
  arrange(subj_id_stan, trial_within_subject)

proto_subj_bounds <- proto_ml_sorted |>
  mutate(row_idx = row_number()) |>
  group_by(subj_id_stan) |>
  summarise(
    subj_start = min(row_idx),
    subj_end   = max(row_idx),
    .groups    = "drop"
  ) |>
  arrange(subj_id_stan)

# Prior design note on pop_log_r_sd:
# pop_log_r_mean ~ Normal(0, 1): same as single-subject prior on log_r.
# pop_log_r_sd   ~ Exponential(2): rate=2 gives median ≈ 0.35 (SD on log scale).
# Marginal SD of subj_log_r ≈ sqrt(1 + 0.5^2) ≈ 1.12 at the 50th pctile
# of the pop_log_r_sd prior — consistent with the single-agent prior range.
proto_ml_data <- list(
  N_total                      = nrow(proto_ml_sorted),
  N_subjects                   = max(proto_ml_sorted$subj_id_stan),
  N_features                   = 2L,
  subj_start                   = proto_subj_bounds$subj_start,
  subj_end                     = proto_subj_bounds$subj_end,
  y                            = as.integer(proto_ml_sorted$sim_response),
  cat_one                      = as.integer(proto_ml_sorted$category_feedback),
  obs                          = as.matrix(proto_ml_sorted[, c("height", "position")]),
  initial_mu_cat0              = c(2.5, 2.5),
  initial_mu_cat1              = c(2.5, 2.5),
  initial_sigma_diag           = 10.0,
  pop_log_r_mean_prior_mean    = LOG_R_PRIOR_MEAN,
  pop_log_r_mean_prior_sd      = LOG_R_PRIOR_SD,
  pop_log_r_sd_prior_rate      = 2.0,
  pop_log_q_mean_prior_mean    = LOG_Q_PRIOR_MEAN,
  pop_log_q_mean_prior_sd      = LOG_Q_PRIOR_SD,
  pop_log_q_sd_prior_rate      = 2.0
)

fit_filepath_ml_proto <- here("simmodels", "ch13_prototype_ml_fit.rds")

if (regenerate_simulations || !file.exists(fit_filepath_ml_proto)) {
  pf_ml_proto <- tryCatch(
    mod_prototype_ml$pathfinder(
      data      = proto_ml_data,
      seed      = 42,
      num_paths = 4,
      refresh   = 0
    ),
    error = function(e) {
      message("Pathfinder failed — falling back to default initialisation.")
      NULL
    }
  )

  fit_proto_ml <- mod_prototype_ml$sample(
    data            = proto_ml_data,
    init            = pf_ml_proto,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1500,
    iter_sampling   = 1500,
    refresh         = 200,
    adapt_delta     = 0.95,
    max_treedepth   = 12
  )
  fit_proto_ml$save_object(fit_filepath_ml_proto)
  cat("Multilevel prototype model fitted and saved.\n")
} else {
  fit_proto_ml <- readRDS(fit_filepath_ml_proto)
  cat("Loaded existing multilevel prototype fit.\n")
}
```

### Mandatory MCMC Diagnostic Battery (Multilevel)

```{r ch13_multilevel_diagnostic_battery}
if (!is.null(fit_proto_ml)) {
  pop_vars_ml <- c("pop_log_r_mean", "pop_log_r_sd", "pop_log_q_mean", "pop_log_q_sd")

  diag_tbl_ml <- diagnostic_summary_table(fit_proto_ml, pars = pop_vars_ml)
  print(diag_tbl_ml)

  if (any(c(diag_tbl_ml$n_divergent > 0,
            diag_tbl_ml$max_rhat    >= 1.01,
            diag_tbl_ml$min_ess_bulk < 400))) {
    warning("Multilevel diagnostic battery has failures. ",
            "Inspect mcmc_pairs() and consider tightening adapt_delta or ",
            "increasing max_treedepth before interpreting posteriors.")
  }

  bayesplot::mcmc_pairs(
    fit_proto_ml$draws(c("pop_log_r_mean", "pop_log_r_sd",
                         "pop_log_q_mean", "pop_log_q_sd")),
    diag_fun     = "dens",
    off_diag_fun = "hex"
  )

  pop_summary <- fit_proto_ml$summary(variables = pop_vars_ml)
  cat("=== Population Parameter Summary ===\n")
  print(pop_summary)
  cat("\nTrue population parameters used for simulation:\n")
  cat("  pop_log_r_mean:", pop_params_ml_proto$pop_log_r_mean, "\n")
  cat("  pop_log_r_sd:  ", pop_params_ml_proto$pop_log_r_sd, "\n")
}
```

### Population Prior-Posterior Update

```{r ch13_multilevel_posterior_viz}
if (!is.null(fit_proto_ml)) {
  draws_ml_proto <- as_draws_df(
    fit_proto_ml$draws(variables = c("pop_log_r_mean", "pop_log_r_sd",
                                     "pop_log_q_mean", "pop_log_q_sd"))
  )

  n_prior <- nrow(draws_ml_proto)
  set.seed(9)
  prior_ml_proto <- tibble(
    pop_log_r_mean = rnorm(n_prior, LOG_R_PRIOR_MEAN, LOG_R_PRIOR_SD),
    pop_log_r_sd   = rexp(n_prior, 2.0),
    pop_log_q_mean = rnorm(n_prior, LOG_Q_PRIOR_MEAN, LOG_Q_PRIOR_SD),
    pop_log_q_sd   = rexp(n_prior, 2.0)
  )

  plot_param_update <- function(param_name, prior_vec, post_vec,
                                true_val, fill_color) {
    df <- tibble(
      value        = c(prior_vec, post_vec),
      Distribution = rep(c("Prior", "Posterior"), each = length(prior_vec))
    )
    ggplot(df, aes(x = value, fill = Distribution, color = Distribution)) +
      geom_density(alpha = 0.5, linewidth = 0.5) +
      geom_vline(xintercept = true_val, color = "red",
                 linetype = "dashed", linewidth = 1) +
      scale_fill_manual(values  = c("Prior" = "grey70", "Posterior" = fill_color)) +
      scale_color_manual(values = c("Prior" = "grey50", "Posterior" = fill_color)) +
      labs(title = param_name, x = NULL, y = "Density") +
      theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
  }

  p_logr_mean <- plot_param_update(
    "Population log(r) mean",
    prior_ml_proto$pop_log_r_mean,
    draws_ml_proto$pop_log_r_mean,
    pop_params_ml_proto$pop_log_r_mean,
    "#009E73"
  )

  p_logr_sd <- plot_param_update(
    "Population log(r) SD",
    prior_ml_proto$pop_log_r_sd,
    draws_ml_proto$pop_log_r_sd,
    pop_params_ml_proto$pop_log_r_sd,
    "#E69F00"
  )

  p_logq_mean <- plot_param_update(
    "Population log(q) mean",
    prior_ml_proto$pop_log_q_mean,
    draws_ml_proto$pop_log_q_mean,
    LOG_Q_PRIOR_MEAN,   # true value = prior mean (q hierarchy not set in sim)
    "#56B4E9"
  )

  p_logq_sd <- plot_param_update(
    "Population log(q) SD",
    prior_ml_proto$pop_log_q_sd,
    draws_ml_proto$pop_log_q_sd,
    0.5,   # illustrative true value
    "#CC79A7"
  )

  cat("=== Population Parameter Prior-Posterior Updates ===\n")
  print((p_logr_mean | p_logr_sd) / (p_logq_mean | p_logq_sd))
}
```

**Reading the population posterior.** The red dashed lines mark the true parameter values used to generate the data. Well-calibrated inference should show the posterior clearly shifted from the prior and centred near the true value. With 20 subjects each contributing 64 trials, the population mean is well-identified; the population SD requires more subjects before its posterior narrows substantially (see §"Multilevel SBC" below for the coverage analysis).

## Multilevel SBC

Multilevel SBC checks that the full posterior — including both population hyperparameters and individual-level parameters — is correctly calibrated across the entire joint prior. The prototype model now has two per-subject parameters ($\log r$ and $\log q$), giving six ECDF-difference curves in the SBC output.

**What the key curves test:**

* `pop_log_r_mean` / `pop_log_q_mean` — Are the population-mean estimators centred? A monotone drift indicates directional bias.
* `pop_log_r_sd` / `pop_log_q_sd` — Are the population-SD estimators well-calibrated? U-shaped rank histograms indicate the posterior is overconfident; humps indicate underconfidence. The NCP should prevent reverse-funnel issues.
* `subj_log_r[1]` / `subj_log_q[1]` — Do partial pooling and NCP produce valid individual-level credible intervals for both parameters?

```{r ch13_sbc_multilevel}
sbc_ml_proto_filepath <- here("simdata", "ch13_sbc_proto_ml_results.rds")

if (regenerate_simulations || !file.exists(sbc_ml_proto_filepath)) {
  n_sbc_iterations_ml_proto <- 200   # use >= 1000 for publication-quality SBC
  n_subjects_sbc_proto      <- 10

  proto_ml_sbc_generator <- SBC_generator_function(
    function() {
      pop_log_r_mean <- rnorm(1, LOG_R_PRIOR_MEAN, LOG_R_PRIOR_SD)
      pop_log_r_sd   <- rexp(1, 2.0)
      pop_log_q_mean <- rnorm(1, LOG_Q_PRIOR_MEAN, LOG_Q_PRIOR_SD)
      pop_log_q_sd   <- rexp(1, 2.0)

      z_log_r    <- rnorm(n_subjects_sbc_proto, 0, 1)
      subj_log_r <- pop_log_r_mean + pop_log_r_sd * z_log_r
      subj_r     <- exp(subj_log_r)

      z_log_q    <- rnorm(n_subjects_sbc_proto, 0, 1)
      subj_log_q <- pop_log_q_mean + pop_log_q_sd * z_log_q
      subj_q     <- exp(subj_log_q)

      all_dat <- map_dfr(seq_len(n_subjects_sbc_proto), function(j) {
        sched <- make_subject_schedule(stimulus_info, n_blocks,
                                       seed = sample.int(1e8, 1))
        res   <- prototype_kalman(
          r_value            = subj_r[j],
          q_value            = subj_q[j],
          obs                = as.matrix(sched[, c("height", "position")]),
          cat_one            = sched$category_feedback,
          initial_mu         = list(rep(2.5, 2), rep(2.5, 2)),
          initial_sigma_diag = 10.0
        )
        sched |> mutate(agent_id = j, sim_response = res$sim_response)
      })

      sorted <- all_dat |>
        mutate(subj_id_stan = as.integer(factor(agent_id))) |>
        arrange(subj_id_stan, trial_within_subject)

      bounds <- sorted |>
        mutate(row_idx = row_number()) |>
        group_by(subj_id_stan) |>
        summarise(subj_start = min(row_idx), subj_end = max(row_idx),
                  .groups = "drop") |>
        arrange(subj_id_stan)

      list(
        variables = list(
          pop_log_r_mean    = pop_log_r_mean,
          pop_log_r_sd      = pop_log_r_sd,
          `subj_log_r[1]`   = subj_log_r[1],
          pop_log_q_mean    = pop_log_q_mean,
          pop_log_q_sd      = pop_log_q_sd,
          `subj_log_q[1]`   = subj_log_q[1]
        ),
        generated = list(
          N_total                   = nrow(sorted),
          N_subjects                = n_subjects_sbc_proto,
          N_features                = 2L,
          subj_start                = bounds$subj_start,
          subj_end                  = bounds$subj_end,
          y                         = as.integer(sorted$sim_response),
          cat_one                   = as.integer(sorted$category_feedback),
          obs                       = as.matrix(sorted[, c("height", "position")]),
          initial_mu_cat0           = c(2.5, 2.5),
          initial_mu_cat1           = c(2.5, 2.5),
          initial_sigma_diag        = 10.0,
          pop_log_r_mean_prior_mean = LOG_R_PRIOR_MEAN,
          pop_log_r_mean_prior_sd   = LOG_R_PRIOR_SD,
          pop_log_r_sd_prior_rate   = 2.0,
          pop_log_q_mean_prior_mean = LOG_Q_PRIOR_MEAN,
          pop_log_q_mean_prior_sd   = LOG_Q_PRIOR_SD,
          pop_log_q_sd_prior_rate   = 2.0
        )
      )
    }
  )

  proto_ml_sbc_backend <- SBC_backend_cmdstan_sample(
    mod_prototype_ml,
    iter_warmup   = 1000,
    iter_sampling = 1000,
    chains        = 2,
    adapt_delta   = 0.95,
    refresh       = 0
  )

  proto_ml_datasets <- generate_datasets(
    proto_ml_sbc_generator,
    n_sims = n_sbc_iterations_ml_proto
  )

  sbc_results_ml_proto <- compute_SBC(
    proto_ml_datasets,
    proto_ml_sbc_backend,
    cache_mode     = "results",
    cache_location = here("simdata", "ch13_sbc_proto_ml_cache"),
    keep_fits      = FALSE
  )
  saveRDS(sbc_results_ml_proto, sbc_ml_proto_filepath)
  cat("Multilevel prototype SBC results computed and saved.\n")
} else {
  sbc_results_ml_proto <- readRDS(sbc_ml_proto_filepath)
  cat("Loaded existing multilevel prototype SBC results.\n")
}

plot_ecdf_diff(sbc_results_ml_proto)
plot_rank_hist(sbc_results_ml_proto)

# ── Population-level parameter recovery across SBC simulations ───────────────
recovery_df_pop_proto <- sbc_results_ml_proto$stats |>
  filter(variable %in% c("pop_log_r_mean", "pop_log_r_sd",
                          "pop_log_q_mean", "pop_log_q_sd"))

p_sbc_pop_proto <- ggplot(recovery_df_pop_proto,
                           aes(x = simulated_value, y = mean)) +
  geom_point(alpha = 0.5, color = "#009E73") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~variable, scales = "free") +
  labs(
    title    = "Population Parameter Recovery Across SBC Simulations",
    subtitle = "Posterior mean vs. true generated value",
    x = "True Value", y = "Estimated Value (Posterior Mean)"
  ) +
  theme_bw()
print(p_sbc_pop_proto)

# ── Individual-level parameter recovery across SBC simulations ──────────────
recovery_df_indiv_proto <- sbc_results_ml_proto$stats |>
  filter(variable %in% c("subj_log_r[1]", "subj_log_q[1]"))

p_sbc_indiv_proto <- ggplot(recovery_df_indiv_proto,
                             aes(x = simulated_value, y = mean)) +
  geom_point(alpha = 0.5, color = "#D55E00") +
  geom_abline(intercept = 0, slope = 1,
              color = "grey30", linetype = "dashed") +
  facet_wrap(~variable, scales = "free") +
  labs(
    title    = "Individual Parameter Recovery (Subject 1) Across SBC",
    subtitle = "Posterior mean vs. true generated value",
    x = "True value [Subject 1]",
    y = "Estimated value (Posterior Mean)"
  ) +
  theme_bw()
print(p_sbc_indiv_proto)
```

**How to read the multilevel prototype SBC.** The prototype model now has two per-subject parameters ($\log r$ and $\log q$), giving six ECDF-difference curves.

* **ECDF-difference plot and rank histograms.** All six curves (`pop_log_r_mean`, `pop_log_r_sd`, `subj_log_r[1]`, `pop_log_q_mean`, `pop_log_q_sd`, `subj_log_q[1]`) should remain inside the simultaneous grey confidence band. A U-shaped rank histogram for `pop_log_r_sd` or `pop_log_q_sd` is the canonical NCP failure mode. For `pop_log_q_sd` especially, the posterior may be wide (q is partially confounded with r under the static paradigm), so coverage rather than width is the key diagnostic. If rank histograms show systematic failures, the standard remedy is to switch from NCP to centred parameterisation for the affected parameter.

* **Scatter plots (population and individual recovery).** Points along the red dashed identity line confirm that the posterior means are unbiased estimators of the generating values. Systematic curvature (points arching away from the line at the extremes) would indicate either prior-boundary effects or insufficient data per subject.

* **Comparison with single-subject SBC (§"Simulation-Based Calibration" above).** The single-subject SBC used $S = 500$ iterations of one agent each; the multilevel SBC uses $S = 200$ iterations of 10 agents each. The multilevel SBC is the authoritative check for multi-subject inference: it verifies that partial pooling does not distort individual-level posteriors and that population hyperparameters are jointly calibrated with per-subject parameters.

---

## Cognitive Insights and Model Comparison

The Kalman filter prototype model offers valuable insights:

* **Incremental, Adaptive Learning**: It formalizes how category knowledge can be built up one example at a time, with learning automatically slowing down as confidence increases.
* **Uncertainty Representation**: It explicitly models the learner's uncertainty about category prototypes — something the basic GCM does not directly represent.
* **Memory Efficiency**: It achieves learning by storing only summary statistics (mean and covariance) per category, aligning with intuitions about cognitive limits.

### Comparing to Exemplar Models (GCM)

* The GCM stores every detail; the prototype model abstracts.
* The GCM is sensitive to specific outliers; the prototype model is less so (outliers get averaged in).
* The GCM requires explicit attention weights ($w$) as free parameters; attention in the Kalman filter can emerge implicitly if uncertainty differs across dimensions (though not explored in detail here).
* The GCM's sensitivity parameter $c$ controls generalization globally; the Kalman filter's $r$ controls how rapidly the prototype absorbs new information, providing a different kind of flexibility.

A note of caution about ELPD comparisons in Chapter 15: now that LFO-CV is implemented for both the GCM (Chapter 11) and the Kalman prototype (this chapter), the comparison between them is on equal predictive footing. The rule mixture model in Chapter 13 is conditionally exchangeable and uses standard PSIS-LOO, which is technically valid for *that* model — but the comparison across architectures is still a comparison of fundamentally different generative processes, and "ELPD is higher" should not be read as "this is the right cognitive theory." See Chapter 13 for the deeper version of this caveat (and the ABC-based alternative comparison framework that resolves it more cleanly).

### Limitations

* **Abstraction Loses Detail**: Information about specific examples and within-category variance beyond the prototype covariance is discarded. Hybrid models attempt to combine both approaches.

* **Single Prototype**: Assumes categories have a single central tendency, struggling with disjunctive categories (e.g., "mammals that fly or swim"). Extensions using multiple prototypes address this.

* **Path-dependence and LFO-CV (now discharged)**: The Kalman filter is path-dependent — trial $t$'s choice probability depends on the entire filter state from trials $1, \ldots, t-1$ — and standard PSIS-LOO is an invalid measure of out-of-sample predictive accuracy. Earlier drafts of this material left this as an unresolved caveat. **This chapter implements LFO-CV via `psis_lfo_kalman()`, validates it against exact one-step-ahead refits, and Chapter 15 inherits the implementation.** The caveat is no longer outstanding.

* **Conflation of observation noise and decisional precision**: The parameter $r$ controls both the Kalman gain (learning speed) and the multivariate normal variance at decision time (decisional precision), alongside $q$ which controls drift speed. These roles are partially separable from learning trajectories, but adding a separate softmax temperature $\tau$ would further disentangle decisional precision from observational noise. We chose not to add $\tau$ here to keep the model tractable, but the conflation is real — when fitting human data, posterior estimates of $r$ should be interpreted as an *average* over learning-speed and decisional-precision contributions.

* **Trial-1 indifference**: The model has no category bias and no perceptual fluency prior. Trial 1 is a coin flip by construction. For tasks with strong baseline asymmetries, this is a misspecification.

* **Process noise identifiability**: The model includes process noise $Q = q \cdot I$, but $q$ and $r$ are partially confounded under the static Kruschke (1993) paradigm — both affect the effective likelihood width at decision time. Separate identifiability requires non-stationary data where prototypes demonstrably shift. The SBC analysis shows the extent of this confound, and the precision analysis quantifies how many trials of category drift are needed to pin down $q$.

* **Structural priors**: The initial means, initial covariances, and diagonal observation-noise structure are fixed by hand. None of these is inferred from data. A reader who finds the model behaving oddly on a new dataset should re-examine these choices before concluding that the cognitive theory is wrong.

## Conclusion

The Kalman filter prototype model provides a dynamic, computationally plausible account of how categories might be learned by abstracting central tendencies and tracking uncertainty. It stands as a compelling alternative to exemplar-based approaches, highlighting the fundamental trade-off between detailed instance memory and abstract summary representation in categorization.

As a Stan model it has a particularly clean shape: with two parameters and a fully deterministic likelihood, Pathfinder initialization is reliable, parameter recovery is tractable, the diagnostic battery is nearly trivial to satisfy, and SBC is straightforward. That cleanliness is *why* this chapter is the right place in the trio to introduce LFO-CV — the algorithm is most legible when there are only two parameters to re-weight and the per-trial filter is fast. Chapters 11, 13, and 15 inherit `psis_lfo_kalman()` (and its GCM sibling `psis_lfo_gcm()` from Chapter 11) so that the LFO-CV caveat that earlier drafts of this material repeated three times across three chapters is now discharged in one place.

The next chapter introduces a third perspective — rule-based models — which makes categorization decisions through explicit hypothesis testing rather than continuous similarity to either stored examples or running averages. That chapter also confronts a problem the Kalman model lets us sidestep: when the cognitive theory's likelihood is intractable in Stan, what is the principled response? Chapter 13 takes up that question, and the ABC-based answer it develops is what makes the cross-architecture comparison in Chapter 15 finally interpretable.
