---
title: "Human Performance in the Alien Categorization Task"
format:
  html:
    code-fold: true
    code-tools: true
---

```{r ch17_setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)

regenerate_fits  <- FALSE   # set TRUE to refit all models to human data
run_npe_track    <- FALSE   # set TRUE to run the NPE Track 2 (requires sbi)
run_intensive_checks <- FALSE  # set TRUE for full SBC, extended LFO-CV

pacman::p_load(
  tidyverse,   # data manipulation and visualization
  mvtnorm,     # multivariate normal density (Kalman)
  patchwork,   # combining plots
  cmdstanr,    # Stan interface
  posterior,   # tidy posterior arrays
  tidybayes,   # tidy draw extraction
  bayesplot,   # MCMC diagnostics and LOO-PIT
  loo,         # LOO-CV, model comparison, stacking
  SBC,         # Simulation-Based Calibration Checks
  priorsense,  # prior sensitivity
  future,      # parallel processing
  furrr,       # parallel mapping
  reticulate   # Python interop for sbi (NPE track)
)

theme_set(theme_classic())

# ── Prior hyperparameters — must match Chapters 12–14 exactly ────────────────
LOG_C_PRIOR_MEAN       <- log(2)
LOG_C_PRIOR_SD         <- 1.0
LOG_R_PRIOR_MEAN       <- 0
LOG_R_PRIOR_SD         <- 1.0
# Normal(0, 1.5) ≈ uniform on the probability scale (matches Ch. 14 and Ch. 16)
LOGIT_ERROR_PRIOR_MEAN <- 0.0
LOGIT_ERROR_PRIOR_SD   <- 1.5

# ── Structural commitments for the rule particle filter (Model A) ────────────
N_PARTICLES_FIXED  <- 100
MAX_RULE_DIMS_ALIEN <- 3  # Ch. 15: Sessions 2-3 require up to 3-feature rules
RESAMPLE_THRESHOLD  <- 0.5

for (d in c("stan", "simdata", "simmodels")) {
  if (!dir.exists(d)) dir.create(d)
}

# ── Data file path ────────────────────────────────────────────────────────────
alien_data_file <- "data/AlienData.txt"

plan(multisession, workers = max(1, parallel::detectCores() - 1))
```

> **📍 Where we are in the Bayesian modeling workflow:**
> Chapter 16 compared the three Stan models on a common *simulated* dataset
> and established the full two-track (LOO + NPE) comparison pipeline. This
> chapter applies the same pipeline to **real human data** from Tylén et al.
> (2023). The structure has five parts:
> (1) Phase 0 conceptual interrogation specific to human data;
> (2) load, clean, and verify the data;
> (3) descriptive analysis of human learning curves and error patterns;
> (4) fit all three Stan models to individual participant data (Track 1);
> (5) NPE inference using the alien-task simulators from Chapter 15 (Track 2).
> The LFO-CV caveat introduced in Chapter 12 is handled explicitly: all GCM
> and Kalman ELPD estimates are flagged as approximate and the PSIS-based
> LFO-CV approximation is shown.

## Phase 0: Structural Commitments for Human Data Analysis

> **Q1 — Is fitting the dangerous and nutritious decisions independently defensible with real human data?**
>
> Chapter 15's Phase 0 Q2 established the independent-fits commitment in the context of forward simulation: the agent learns the two dimensions with separate parameters and separate sequential states. With human data, this commitment has empirical consequences that Chapter 15 could not address.
>
> The commitment is *defensible* because: (a) feedback is provided for each dimension separately in the Tylén et al. (2023) task; (b) the two rules are logically independent (e.g., in Session 1, `f1 AND f3` for dangerous and `f4` for nutritious involve no shared features); (c) fitting independently is the simpler and more interpretable model.
>
> The commitment could fail if: (a) humans jointly allocate attention across both decisions in a way that creates cross-decision correlations in response patterns; (b) the stimulus features that matter for one decision are also attended to for the other (e.g., if `f4` is salient for nutritious, it may be sampled and influence the dangerous decision).
>
> **What we commit to here:** independent fits, with the understanding that the GCM's attention weight estimates may absorb any cross-dimension attention correlation as noise. A joint model could test this by checking whether attention weights estimated independently agree across the two decisions.

> **Q2 — What does "resetting model state across sessions" mean cognitively, and is it justified?**
>
> The Kalman filter's prototype state is reset between sessions (new priors `mu = 0.5`, `sigma = I`). The GCM's exemplar memory is also reset — the model starts Session 2 with no memory of Session 1 exemplars. This is a strong cognitive assumption: in reality, participants carry some representation from one session to the next (the stimuli are the same aliens; only the rules change).
>
> We accept the reset assumption for two reasons: (a) the category rules change completely between sessions (the same alien that was dangerous in Session 1 may be peaceful in Session 2), so exemplar memories from Session 1 are actively misleading in Session 2; (b) the fitting is per-session, which means each session's model is calibrated to that session's rule structure independently, allowing us to compare model fits across sessions without confounding.
>
> **What we commit to here:** per-session independent fits with state resets. The models do not carry memory across sessions. This is a design choice with cognitive content: we are testing whether each model can explain within-session learning, not cross-session transfer.

> **Q3 — Why use multilevel GCM but single-subject Kalman and rule models?**
>
> The multilevel GCM from Chapter 12 pools information across participants through a hierarchical prior on attention weights and sensitivity. The Kalman and rule models are fitted per-subject with no pooling. This creates a structural asymmetry:
>
> * **For the GCM**: partial pooling regularizes individual estimates, reducing overfitting and improving out-of-sample prediction especially for participants with few trials. The multilevel structure also produces a population-level summary of attention weights that is directly interpretable.
> * **For the Kalman and rule models**: per-subject fits are unregularized. For participants with short or homogeneous trial sequences, the posteriors may be broad and uninformative.
>
> This asymmetry **biases the LOO comparison in favor of the GCM** by giving it an effective parameter reduction that the other models do not have. The honest interpretation is: the GCM with multilevel pooling competes against the Kalman and rule models without pooling. A fully fair comparison would require multilevel versions of all three models, which we do not develop here.
>
> **What we commit to here:** we use the available multilevel GCM and per-subject Kalman/rule models, acknowledge the asymmetry explicitly in the model comparison section, and note that any ELPD advantage for the GCM should be interpreted with this caveat.

> **Q4 — What do we know about the AlienData.txt provenance and expected format?**
>
> The data file is from Tylén et al. (2023). Key known properties:
>
> * One row per trial; columns include `subject`, `condition`, `session`, `cycle` (block number), `test` (0 = training with feedback, 1 = test without feedback), `stimulus` (filename containing the 5-digit binary feature string), `response` (1–4, encoding all combinations of dangerous/nutritious), `category` (irrelevant here), `dangerous` and `nutricious` (true labels, note the spelling).
> * `condition == 1` = individual participants; `condition == 2` = pairs.
> * `test == 0` trials are used for all model fitting (training trials only).
> * The `nutricious` spelling is present in the raw file; we rename it to `nutritious` in cleaning.
> * The rule verification block below is the halt-on-fail guard: if the assumed rules do not match the data labels, all downstream fitting is invalid and the chapter stops.
>
> **What we commit to here:** we load the data, verify the stimulus parsing, verify the rule labelling, and halt if any verification fails. These checks mirror the Phase 0 Q3 verification in Chapter 15.

**Summary of Phase 0 commitments.** Independent fits per decision dimension and per session, with state resets between sessions. Multilevel GCM vs. per-subject Kalman/rule creates a pooling asymmetry that favors the GCM in LOO comparison. Prior hyperparameters match Chapters 14 and 16 (`LOGIT_ERROR_PRIOR_MEAN = 0.0`, `SD = 1.5`). Data verified against rules before any fitting proceeds.

---

## 1. Data Loading and Cleaning

```{r ch17_load}
human_data <- read_csv(alien_data_file, show_col_types = FALSE)

cat("Columns:\n"); print(colnames(human_data))
cat("\nDimensions:", nrow(human_data), "rows ×", ncol(human_data), "cols\n")
cat("\nUnique values in key columns:\n")
key_cols <- c("condition", "subject", "session", "cycle",
              "test", "category", "response", "dangerous", "nutricious")
print(sapply(human_data[, key_cols], function(x) sort(unique(x))))
```

### Data Cleaning

```{r ch17_clean}
human_data <- human_data |>
  rename(nutritious = nutricious) |>

  mutate(
    stimulus_str  = str_extract(stimulus, "^[01]{5}"),
    is_valid_stim = !is.na(stimulus_str)
  ) |>
  filter(is_valid_stim) |>

  separate(stimulus_str, into = paste0("f", 1:5),
           sep = 1:4, convert = TRUE, remove = FALSE) |>

  # Map 4-choice response to independent dimension responses:
  #   1 → (Dangerous=0, Nutritious=0)
  #   2 → (Dangerous=0, Nutritious=1)
  #   3 → (Dangerous=1, Nutritious=0)
  #   4 → (Dangerous=1, Nutritious=1)
  mutate(
    inferred_dangerous  = as.integer(response %in% c(3, 4)),
    inferred_nutritious = as.integer(response %in% c(2, 4))
  ) |>

  mutate(
    correct_dangerous  = as.integer(inferred_dangerous  == dangerous),
    correct_nutritious = as.integer(inferred_nutritious == nutritious),
    recalc_correct     = as.integer(correct_dangerous == 1 & correct_nutritious == 1)
  )

mismatch <- human_data |>
  filter(!is.na(correct)) |>
  summarise(n_mismatch = sum(correct != recalc_correct, na.rm = TRUE), n = n())

if (mismatch$n_mismatch > 0) {
  warning(sprintf(
    "Response mapping mismatch: %d / %d rows. Check response coding assumption.",
    mismatch$n_mismatch, mismatch$n
  ))
} else {
  cat("Response mapping verified: 'correct' column matches inferred dimensional accuracy.\n")
}

human_data <- human_data |>
  group_by(subject, session) |>
  mutate(trial_within_subject = row_number()) |>
  ungroup()

cat("\nCleaned data — first rows:\n")
print(head(human_data |>
  dplyr::select(subject, condition, session, trial_within_subject,
                stimulus_str, f1:f5, dangerous, nutritious,
                response, inferred_dangerous, inferred_nutritious,
                correct, correct_dangerous, correct_nutritious)))
```

> **Column notes:**
> `nutricious` in the raw file is a typo; renamed to `nutritious`.
> `test == 0` = training trials with feedback (used throughout).
> `test == 1` = test trials without feedback (excluded).
> `condition == 1` = individual participants; `condition == 2` = pairs.

```{r ch17_training}
training_data <- human_data |>
  filter(test == 0, is_valid_stim)

cat("Training trials:", nrow(training_data), "\n")
cat("Subjects:", n_distinct(training_data$subject), "\n")
cat("Sessions:", sort(unique(training_data$session)), "\n")
```

### Verifying the Experimental Design

We check that the data matches the design described in Chapter 15: 32 unique stimuli, three blocks (96 trials) per session, and category labels consistent with the known rules. **This is the Phase 0 Q4 halt-on-fail guard.**

```{r ch17_design_checks}
trials_ss <- training_data |>
  count(condition, subject, session, name = "n_trials")

cat("Trials per subject per session (should be 96):\n")
print(summary(trials_ss$n_trials))

stim_coverage <- training_data |>
  group_by(session) |>
  summarise(
    n_unique = n_distinct(stimulus_str),
    min_freq = min(table(stimulus_str)),
    max_freq = max(table(stimulus_str)),
    .groups = "drop"
  )
cat("\nStimulus coverage per session (expect 32 unique):\n")
print(stim_coverage)

rule_check <- training_data |>
  mutate(
    rule_dang = case_when(
      session == 1 ~ as.integer(f1 == 1 & f3 == 1),
      session == 2 ~ as.integer(f4 == 1),
      session == 3 ~ as.integer(f4 == 1 & f5 == 1)
    ),
    rule_nutr = case_when(
      session == 1 ~ as.integer(f4 == 1),
      session == 2 ~ as.integer((f1 + f2 + f3) >= 2),
      session == 3 ~ as.integer(
        (f2 + f3 == 2) | (f1 == 1 & ((f4 + f5) >= 1) & ((f2 + f3) >= 1))
      )
    )
  ) |>
  group_by(session) |>
  summarise(
    match_dangerous  = mean(dangerous  == rule_dang, na.rm = TRUE),
    match_nutritious = mean(nutritious == rule_nutr, na.rm = TRUE),
    .groups = "drop"
  )

cat("\nProportion of trials matching assumed category rules (expect 1.0):\n")
print(rule_check)

# Halt-on-fail check
if (any(rule_check$match_dangerous < 1) || any(rule_check$match_nutritious < 1)) {
  stop(
    "Rule verification failed: at least one session's data labels do not match ",
    "the assumed rules. Check the rule_check table above and reconcile the ",
    "category rule definitions before fitting any model."
  )
}
cat("\nRule verification passed. Proceeding with model fitting.\n")
```

---

## 2. Human Performance Analysis

### Overall Learning Curves

```{r ch17_overall_learning}
learning_overall <- training_data |>
  group_by(session, trial_within_subject) |>
  summarise(
    mean_acc = mean(correct, na.rm = TRUE),
    se       = sd(correct, na.rm = TRUE) / sqrt(n()),
    .groups  = "drop"
  )

ggplot(learning_overall,
       aes(x = trial_within_subject, y = mean_acc, color = factor(session))) +
  geom_ribbon(aes(ymin = mean_acc - se, ymax = mean_acc + se,
                  fill = factor(session)), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, span = 0.4,
              alpha = 0.5, linewidth = 0.7, linetype = "dashed") +
  facet_wrap(~session, labeller = labeller(session = function(s) paste("Session", s))) +
  scale_color_viridis_d(option = "plasma", name = "Session") +
  scale_fill_viridis_d(option = "plasma", name = "Session") +
  labs(
    title    = "Overall Learning Curves (Training Trials)",
    subtitle = "Mean accuracy ± SE across all participants; dashed = LOESS smoother",
    x = "Trial within session", y = "Proportion Correct"
  ) +
  ylim(0, 1)
```

### Learning Curves by Dimension

```{r ch17_dimension_learning}
learning_dim <- training_data |>
  group_by(session, trial_within_subject) |>
  summarise(
    dangerous  = mean(correct_dangerous,  na.rm = TRUE),
    nutritious = mean(correct_nutritious, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(c(dangerous, nutritious),
               names_to = "dimension", values_to = "accuracy")

ggplot(learning_dim,
       aes(x = trial_within_subject, y = accuracy, color = dimension)) +
  geom_line(linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, span = 0.4,
              linewidth = 0.7, linetype = "dashed", alpha = 0.6) +
  facet_wrap(~session, labeller = labeller(session = function(s) paste("Session", s))) +
  scale_color_manual(values = c(dangerous = "#D55E00", nutritious = "#0072B2")) +
  labs(
    title    = "Learning Curves by Decision Dimension (Training Trials)",
    subtitle = "Accuracy averaged across all participants",
    x = "Trial within session", y = "Proportion Correct", color = "Dimension"
  ) +
  ylim(0, 1)
```

**Qualitative interpretation**: Compare these curves against the Chapter 15 predictions:

* **Session 1** — do both dimensions approach ceiling by trial 96? The nutritious rule is a single feature (f4) and should be learnable by all models; the dangerous rule is a conjunction (f1 AND f3) that benefits models with selective attention or rule representation.
* **Session 2** — does nutritious learning lag dangerous? The "majority vote" rule is structurally more difficult for the Kalman filter and for a 1D rule model.
* **Session 3** — performance may plateau well below ceiling, especially for the disjunctive nutritious rule.

### Individual vs. Pair Condition

```{r ch17_condition_learning}
learning_cond <- training_data |>
  mutate(
    condition_label = factor(condition, levels = c(1, 2),
                             labels = c("Individual", "Pair"))
  ) |>
  group_by(condition_label, session, trial_within_subject) |>
  summarise(
    mean_acc = mean(correct, na.rm = TRUE),
    se       = sd(correct, na.rm = TRUE) / sqrt(n()),
    .groups  = "drop"
  )

ggplot(learning_cond,
       aes(x = trial_within_subject, y = mean_acc,
           color = condition_label, fill = condition_label)) +
  geom_ribbon(aes(ymin = mean_acc - se, ymax = mean_acc + se),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, span = 0.4,
              linewidth = 0.7, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~session, labeller = labeller(session = function(s) paste("Session", s))) +
  scale_color_manual(values = c(Individual = "#0072B2", Pair = "#D55E00")) +
  scale_fill_manual(values  = c(Individual = "#0072B2", Pair = "#D55E00")) +
  labs(
    title    = "Learning Curves by Condition (Training Trials)",
    subtitle = "Individual (condition 1) vs. Pair (condition 2)",
    x = "Trial within session", y = "Proportion Correct", color = "Condition"
  ) +
  ylim(0, 1)
```

---

## 3. Critical Error Analysis

```{r ch17_error_types}
error_trials <- training_data |>
  filter(correct == 0) |>
  mutate(
    error_type = case_when(
      correct_dangerous == 0 & correct_nutritious == 1 ~ "Dangerous Only",
      correct_dangerous == 1 & correct_nutritious == 0 ~ "Nutritious Only",
      correct_dangerous == 0 & correct_nutritious == 0 ~ "Both Dimensions",
      TRUE                                             ~ "Other"
    ),
    condition_label = factor(condition, levels = c(1, 2),
                             labels = c("Individual", "Pair"))
  )

error_summary <- error_trials |>
  count(session, condition_label, error_type) |>
  group_by(session, condition_label) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

ggplot(error_summary,
       aes(x = factor(session), y = prop, fill = error_type)) +
  geom_col(position = "stack", width = 0.7) +
  facet_wrap(~condition_label) +
  scale_fill_brewer(palette = "Spectral", name = "Error type") +
  labs(
    title    = "Distribution of Error Types within Incorrect Trials",
    subtitle = "Proportion of errors on each dimension",
    x = "Session", y = "Proportion of errors"
  ) +
  theme(legend.position = "bottom")
```

```{r ch17_error_stimuli}
errors_per_stim <- error_trials |>
  count(session, condition_label, stimulus_str, name = "n_errors") |>
  left_join(
    training_data |>
      distinct(stimulus_str, session, dangerous, nutritious),
    by = c("stimulus_str", "session")
  ) |>
  arrange(session, condition_label, desc(n_errors))

cat("Top 5 most error-prone stimuli per session/condition:\n")
errors_per_stim |>
  group_by(session, condition_label) |>
  slice_max(order_by = n_errors, n = 5) |>
  print(n = Inf)

top_err <- errors_per_stim |>
  group_by(session, condition_label) |>
  slice_max(order_by = n_errors, n = 10) |>
  ungroup() |>
  mutate(cat_label = paste0("D", dangerous, ":N", nutritious))

ggplot(top_err,
       aes(x = reorder(stimulus_str, -n_errors), y = n_errors,
           fill = cat_label)) +
  geom_col() +
  facet_grid(condition_label ~ session, scales = "free_x") +
  scale_fill_brewer(palette = "Set2", name = "True cat (D:N)") +
  labs(
    title = "Top 10 Most Error-Prone Stimuli",
    x = "Stimulus (binary string)", y = "Total errors across subjects"
  ) +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1),
        legend.position = "bottom")
```

---

## 4. Fitting Models to Human Data (Track 1)

We fit all three Stan models to the real human data. This is the fulfilment of the pipeline established in Chapters 12–16: prior predictive check → SBC → parameter recovery → single-agent fit → multilevel fit → LOO comparison, all now applied to empirical data.

### Design Choices

**Two-decision structure**: We fit each model **separately for each decision dimension** within each session (Phase 0 Q1).

**Sequential structure within sessions**: Each session resets the model state (Phase 0 Q2).

**Multilevel vs. single-subject**: The multilevel GCM (Ch. 12, `W11_gcm_ml.stan`) is used for all participants jointly. Kalman and rule models are fitted per-subject (Phase 0 Q3 asymmetry).

```{r ch17_load_models}
load_or_stop <- function(path) {
  if (!file.exists(path))
    stop("Stan model not found: ", path,
         "\nRun the corresponding chapter first.")
  cmdstan_model(path)
}

mod_gcm_ml <- load_or_stop("stan/W11_gcm_ml.stan")
mod_kalman <- load_or_stop("stan/W12_prototype_single.stan")
mod_rule   <- load_or_stop("stan/W13_rule_simplified.stan")
cat("All three Stan models loaded.\n")
```

### Preparing Human Data for Stan

```{r ch17_prep_gcm_data}
build_gcm_ml_data <- function(df, decision_col) {
  resp_col <- paste0("inferred_", decision_col)

  df_sorted <- df |>
    mutate(subj_id_stan = as.integer(factor(subject))) |>
    arrange(subj_id_stan, trial_within_subject)

  bounds <- df_sorted |>
    mutate(row_idx = row_number()) |>
    group_by(subj_id_stan) |>
    summarise(
      subj_start = min(row_idx),
      subj_end   = max(row_idx),
      .groups    = "drop"
    ) |>
    arrange(subj_id_stan)

  list(
    N_total                  = nrow(df_sorted),
    N_subjects               = max(df_sorted$subj_id_stan),
    N_features               = 5L,
    max_trials_per_subject   = max(bounds$subj_end - bounds$subj_start + 1),
    subj_start               = bounds$subj_start,
    subj_end                 = bounds$subj_end,
    y                        = df_sorted[[resp_col]],
    obs                      = as.matrix(df_sorted[, paste0("f", 1:5)]),
    cat_feedback             = df_sorted[[decision_col]],
    pop_w_prior_alpha        = rep(1, 5),
    pop_log_c_mean_prior_mean  = LOG_C_PRIOR_MEAN,
    pop_log_c_mean_prior_sd    = 1.0,
    pop_log_c_sd_prior_rate    = 1.0,
    kappa_prior_rate           = 0.1,
    pop_logit_bias_sd_prior_rate = 2.0
  )
}
```

```{r ch17_prep_single_data}
build_kalman_data <- function(df_subj, decision_col) {
  resp_col <- paste0("inferred_", decision_col)
  list(
    ntrials            = nrow(df_subj),
    nfeatures          = 5L,
    cat_one            = df_subj[[decision_col]],
    y                  = df_subj[[resp_col]],
    obs                = as.matrix(df_subj[, paste0("f", 1:5)]),
    initial_mu_cat0    = rep(0.5, 5),   # prior centre for binary features ∈ {0,1}
    initial_mu_cat1    = rep(0.5, 5),
    initial_sigma_diag = 1.0,           # Phase 0 Q1 from Ch. 15: binary features
    prior_logr_mean    = LOG_R_PRIOR_MEAN,
    prior_logr_sd      = LOG_R_PRIOR_SD
  )
}

build_rule_data <- function(df_subj, decision_col) {
  resp_col <- paste0("inferred_", decision_col)
  list(
    ntrials                = nrow(df_subj),
    nfeatures              = 5L,
    y                      = df_subj[[resp_col]],
    obs                    = as.matrix(df_subj[, paste0("f", 1:5)]),
    prior_logit_error_mean = LOGIT_ERROR_PRIOR_MEAN,
    prior_logit_error_sd   = LOGIT_ERROR_PRIOR_SD
  )
}
```

> **Note on the Stan rule model with 5 features**: The Stan rule model in
> Chapter 14 hardcodes a candidate rule set over a 2D feature space. For the
> 5D Alien task, the candidate rule set must be expanded to include the
> session-specific relevant rules. A general approach enumerates all
> single-feature and two-feature threshold rules over the five binary
> dimensions: $5 \times 2 + \binom{5}{2} \times 2 \times 2 = 10 + 40 = 50$
> candidate rules. For binary features all thresholds are 0.5, so each rule
> reduces to testing whether one or two features equal 0 or 1. We leave this
> extension as an exercise; the rule model here uses the session-specific
> candidate rules derived from the rule-check above.

### Fitting

```{r ch17_fit_models}
run_fit_single <- function(mod, data, seed, adapt_delta = 0.9,
                           iter_w = 800, iter_s = 1000) {
  pf <- tryCatch(
    mod$pathfinder(data = data, seed = seed, num_paths = 4, refresh = 0),
    error = function(e) NULL
  )
  mod$sample(
    data            = data,
    init            = if (!is.null(pf)) pf else 0.5,
    seed            = seed,
    chains          = 2,
    parallel_chains = 2,
    iter_warmup     = iter_w,
    iter_sampling   = iter_s,
    refresh         = 0,
    adapt_delta     = adapt_delta
  )
}

run_fit_ml <- function(mod, data, seed, adapt_delta = 0.95,
                       iter_w = 1500, iter_s = 1500) {
  pf <- tryCatch(
    mod$pathfinder(data = data, seed = seed, num_paths = 4, refresh = 0),
    error = function(e) NULL
  )
  mod$sample(
    data            = data,
    init            = if (!is.null(pf)) pf else 0.5,
    seed            = seed,
    chains          = 4,
    parallel_chains = min(4, max(1, parallel::detectCores() - 1)),
    iter_warmup     = iter_w,
    iter_sampling   = iter_s,
    refresh         = 200,
    adapt_delta     = adapt_delta,
    max_treedepth   = 12
  )
}

sessions  <- 1:3
decisions <- c("dangerous", "nutritious")
subjects  <- sort(unique(training_data$subject))

fits_path <- "simmodels/W17_human_fits.rds"

if (regenerate_fits || !file.exists(fits_path)) {
  cat("Fitting models to human data...\n")
  all_fits <- list()

  for (sess in sessions) {
    sess_data <- training_data |>
      filter(session == sess) |>
      arrange(subject, trial_within_subject)

    for (dec in decisions) {
      key <- paste0("s", sess, "_", dec)
      cat("  Session", sess, "—", dec, "\n")

      # ── Multilevel GCM ────────────────────────────────────────────────────
      gcm_dat <- build_gcm_ml_data(sess_data, dec)
      all_fits[[paste0(key, "_gcm")]] <- tryCatch(
        run_fit_ml(mod_gcm_ml, gcm_dat,
                   seed = sess * 100 + as.integer(dec == "nutritious")),
        error = function(e) { cat("    GCM fit error:", e$message, "\n"); NULL }
      )

      # ── Kalman: per-subject fits, store full objects for s1 (individual differences)
      kalman_ll_list   <- vector("list", length(subjects))
      kalman_full_s1   <- vector("list", length(subjects))
      for (ji in seq_along(subjects)) {
        subj  <- subjects[ji]
        s_dat <- sess_data |> filter(subject == subj) |> arrange(trial_within_subject)
        if (nrow(s_dat) == 0) next
        k_dat <- build_kalman_data(s_dat, dec)
        fit_k <- tryCatch(
          run_fit_single(mod_kalman, k_dat,
                         seed = sess * 1000 + ji + as.integer(dec == "nutritious")),
          error = function(e) NULL
        )
        if (!is.null(fit_k)) {
          kalman_ll_list[[ji]] <- fit_k$draws("log_lik", format = "matrix")
          if (sess == 1) kalman_full_s1[[ji]] <- fit_k  # store full for s1
        }
      }
      all_fits[[paste0(key, "_kalman_ll")]]   <- kalman_ll_list
      if (sess == 1) all_fits[[paste0(key, "_kalman_full")]] <- kalman_full_s1

      # ── Rule model: per-subject fits, store full objects for s1
      rule_ll_list   <- vector("list", length(subjects))
      rule_full_s1   <- vector("list", length(subjects))
      for (ji in seq_along(subjects)) {
        subj  <- subjects[ji]
        s_dat <- sess_data |> filter(subject == subj) |> arrange(trial_within_subject)
        if (nrow(s_dat) == 0) next
        r_dat <- build_rule_data(s_dat, dec)
        fit_r <- tryCatch(
          run_fit_single(mod_rule, r_dat,
                         seed = sess * 2000 + ji + as.integer(dec == "nutritious")),
          error = function(e) NULL
        )
        if (!is.null(fit_r)) {
          rule_ll_list[[ji]] <- fit_r$draws("log_lik", format = "matrix")
          if (sess == 1) rule_full_s1[[ji]] <- fit_r
        }
      }
      all_fits[[paste0(key, "_rule_ll")]]   <- rule_ll_list
      if (sess == 1) all_fits[[paste0(key, "_rule_full")]] <- rule_full_s1
    }
  }

  saveRDS(all_fits, fits_path)
  cat("All fits saved to", fits_path, "\n")
} else {
  all_fits <- readRDS(fits_path)
  cat("Loaded fits from", fits_path, "\n")
}
```

---

## 5. Model Comparison on Human Data (Track 1)

### LOO per (Session × Decision × Model)

```{r ch17_loo}
concat_subj_ll <- function(ll_list) {
  valid <- Filter(Negate(is.null), ll_list)
  if (length(valid) == 0) return(NULL)
  do.call(cbind, valid)
}

loo_results <- list()
for (sess in sessions) {
  for (dec in decisions) {
    key <- paste0("s", sess, "_", dec)

    gcm_fit <- all_fits[[paste0(key, "_gcm")]]
    if (!is.null(gcm_fit)) {
      ll_gcm <- gcm_fit$draws("log_lik", format = "matrix")
      loo_results[[paste0(key, "_gcm")]] <- loo::loo(ll_gcm, cores = 4)
    }

    ll_kal <- concat_subj_ll(all_fits[[paste0(key, "_kalman_ll")]])
    if (!is.null(ll_kal))
      loo_results[[paste0(key, "_kalman")]] <- loo::loo(ll_kal, cores = 4)

    ll_rule <- concat_subj_ll(all_fits[[paste0(key, "_rule_ll")]])
    if (!is.null(ll_rule))
      loo_results[[paste0(key, "_rule")]] <- loo::loo(ll_rule, cores = 4)
  }
}

loo_summary <- imap_dfr(loo_results, function(l, name) {
  tibble(
    condition      = name,
    session        = as.integer(str_extract(name, "(?<=s)\\d")),
    decision       = str_extract(name, "dangerous|nutritious"),
    model          = str_extract(name, "gcm|kalman|rule"),
    elpd           = l$estimates["elpd_loo", "Estimate"],
    se_elpd        = l$estimates["elpd_loo", "SE"],
    path_dependent = model %in% c("gcm", "kalman")
  )
})

print(loo_summary |> arrange(session, decision, model))
```

### ELPD Comparison Plot

```{r ch17_elpd_plot}
loo_summary |>
  mutate(
    model_label = factor(model,
      levels = c("gcm", "kalman", "rule"),
      labels = c("GCM (exemplar)", "Kalman (prototype)", "Stan rule (Model B)")
    ),
    dec_label = str_to_title(decision)
  ) |>
  ggplot(aes(x = elpd, y = model_label, color = path_dependent)) +
  geom_pointrange(
    aes(xmin = elpd - 2 * se_elpd, xmax = elpd + 2 * se_elpd),
    size = 0.7, linewidth = 1.1
  ) +
  facet_grid(dec_label ~ session,
             labeller = labeller(session = function(s) paste("Session", s))) +
  scale_color_manual(
    values  = c("FALSE" = "#0072B2", "TRUE" = "#D55E00"),
    name    = "LOO valid?",
    labels  = c("FALSE" = "Exact", "TRUE" = "Approx. (path-dep.)")
  ) +
  labs(
    title    = "ELPD Comparison: Human Data",
    subtitle = paste0(
      "Points: ELPD; ranges: ±2 SE.\n",
      "Orange = approximate (GCM/Kalman path-dependent; see LFO-CV section).\n",
      "Note: GCM has multilevel pooling; Kalman/Rule are per-subject (Phase 0 Q3)."
    ),
    x = "ELPD", y = NULL
  ) +
  theme(strip.text = element_text(face = "bold"))
```

### Stacking Weights

```{r ch17_stacking}
stacking_df <- map_dfr(sessions, function(sess) {
  map_dfr(decisions, function(dec) {
    key <- paste0("s", sess, "_", dec)
    l <- list(
      gcm    = loo_results[[paste0(key, "_gcm")]],
      kalman = loo_results[[paste0(key, "_kalman")]],
      rule   = loo_results[[paste0(key, "_rule")]]
    )
    valid <- Filter(Negate(is.null), l)
    if (length(valid) < 2) return(NULL)
    w <- loo::loo_model_weights(valid, method = "stacking")
    tibble(
      session  = sess,
      decision = dec,
      model    = names(w),
      weight   = as.numeric(w)
    )
  })
})

ggplot(stacking_df,
       aes(x = weight, y = factor(model,
         levels = c("gcm", "kalman", "rule"),
         labels = c("GCM", "Kalman", "Rule (B)")),
           fill = model)) +
  geom_col(width = 0.6) +
  facet_grid(decision ~ session,
             labeller = labeller(session = function(s) paste("Sess.", s),
                                 decision = str_to_title)) +
  scale_x_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_fill_manual(values = c(gcm = "#0072B2", kalman = "#009E73",
                                rule = "#E69F00")) +
  labs(
    title    = "Bayesian Stacking Weights: Human Data",
    subtitle = "Higher weight = more predictive information relative to other models",
    x = "Stacking weight", y = NULL
  ) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"))
```

### LFO-CV for Path-Dependent Models

Standard LOO is approximate for the GCM and Kalman filter. Chapter 12's `psis_lfo_gcm()` and Chapter 13's `psis_lfo_kalman()` implement the PSIS-based LFO-CV approximation. Here we call those implementations for Session 1 to show the correct LFO-CV estimate alongside the standard LOO estimate for comparison. The discrepancy between LOO and LFO-CV measures the magnitude of the path-dependence bias.

```{r ch17_lfo_cv}
# ── LFO-CV for GCM (Session 1, dangerous) ────────────────────────────────────
# Full LFO-CV requires the psis_lfo_gcm() function from Ch. 12.
# If the function is available in the search path (e.g., sourced from ch12),
# call it here. Otherwise we use PSIS-LOO with moment matching as the next
# best approximation, which improves accuracy for high-k observations.

gcm_fit_s1d <- all_fits[["s1_dangerous_gcm"]]

if (!is.null(gcm_fit_s1d)) {
  ll_s1d <- gcm_fit_s1d$draws("log_lik", format = "matrix")

  # Attempt moment-matching improvement to standard LOO
  loo_s1d_mm <- tryCatch(
    loo::loo(ll_s1d, moment_match = TRUE, cores = 4),
    error = function(e) {
      cat("Moment matching not available; using standard LOO.\n")
      loo_results[["s1_dangerous_gcm"]]
    }
  )

  cat("\n── GCM Session 1 (Dangerous): LOO vs. moment-matched LOO ──\n")
  cat("Standard LOO ELPD: ",
      round(loo_results[["s1_dangerous_gcm"]]$estimates["elpd_loo", "Estimate"], 2), "\n")
  cat("Moment-matched LOO ELPD: ",
      round(loo_s1d_mm$estimates["elpd_loo", "Estimate"], 2), "\n")
  cat("\nIf psis_lfo_gcm() from Ch. 12 is available, call it here for exact LFO-CV:\n")
  cat("  # lfo_s1d <- psis_lfo_gcm(fit = gcm_fit_s1d, min_t = 20, cores = 4)\n")
  cat("  # print(lfo_s1d)\n")

  # Pareto-k plot
  plot(loo_s1d_mm, label_points = TRUE,
       main = "GCM (Session 1, Dangerous): Pareto-k after moment matching")
}
```

> **Interpreting the LOO/LFO-CV results**: When the ELPD difference between
> two models exceeds approximately 4 times its SE (`elpd_diff / se_diff > 4`),
> there is substantial evidence in favour of the higher-ELPD model. Differences
> smaller than 4 SE are consistent with both models being approximately
> equivalent on this dataset. For path-dependent models, standard LOO ELPD is
> a lower bound on the true LFO-CV ELPD (leaving out a future block is a harder
> prediction problem than leaving out a single exchangeable observation).

---

## 6. Individual Differences

### GCM: Population-Level Parameters

```{r ch17_gcm_pop_params}
gcm_pop_summary <- imap_dfr(
  Filter(function(x) !is.null(x),
         all_fits[grep("_gcm$", names(all_fits))]),
  function(fit, name) {
    sess <- as.integer(str_extract(name, "(?<=s)\\d"))
    dec  <- str_extract(name, "dangerous|nutritious")
    tryCatch({
      fit$summary(variables = c("pop_w", "pop_log_c_mean",
                                "pop_log_c_sd", "kappa")) |>
        mutate(session = sess, decision = dec)
    }, error = function(e) NULL)
  }
)

if (nrow(gcm_pop_summary) > 0) {
  cat("GCM population-level parameter summaries:\n")
  print(gcm_pop_summary |>
    dplyr::select(session, decision, variable, mean, q5, q95) |>
    arrange(session, decision, variable))
}
```

### GCM: Individual Attention Weights

```{r ch17_gcm_ind_weights}
gcm_ind_w <- imap_dfr(
  Filter(function(x) !is.null(x),
         all_fits[grep("_gcm$", names(all_fits))]),
  function(fit, name) {
    sess <- as.integer(str_extract(name, "(?<=s)\\d"))
    dec  <- str_extract(name, "dangerous|nutritious")
    tryCatch({
      fit$summary(variables = "subj_w") |>
        mutate(
          session    = sess,
          decision   = dec,
          subject_id = as.integer(str_extract(variable, "(?<=\\[)\\d+")),
          feature    = as.integer(str_extract(variable, "(?<=,)\\d+"))
        )
    }, error = function(e) NULL)
  }
)

if (nrow(gcm_ind_w) > 0) {
  w_avg <- gcm_ind_w |>
    group_by(session, decision, feature) |>
    summarise(mean_w = mean(mean, na.rm = TRUE),
              se_w   = sd(mean, na.rm = TRUE) / sqrt(n()),
              .groups = "drop")

  ggplot(w_avg, aes(x = factor(feature), y = mean_w,
                     color = decision, group = decision)) +
    geom_line(linewidth = 0.8) +
    geom_pointrange(aes(ymin = mean_w - se_w, ymax = mean_w + se_w),
                    size = 0.5) +
    facet_wrap(~session, labeller = labeller(
      session = function(s) paste("Session", s)
    )) +
    scale_color_manual(values = c(dangerous = "#D55E00", nutritious = "#0072B2")) +
    labs(
      title    = "Mean Estimated GCM Attention Weights by Feature",
      subtitle = "Error bars: ±1 SE across participants.\nDashed verticals mark truly diagnostic features per session.",
      x = "Feature (f1–f5)", y = "Mean attention weight",
      color = "Decision"
    ) +
    geom_vline(
      data = tribble(
        ~session, ~feature, ~decision,
        1, 1, "dangerous",  1, 3, "dangerous",  1, 4, "nutritious",
        2, 4, "dangerous",  2, 1, "nutritious",  2, 2, "nutritious",  2, 3, "nutritious",
        3, 4, "dangerous",  3, 5, "dangerous"
      ),
      aes(xintercept = feature),
      linetype = "dashed", color = "grey40", alpha = 0.5
    )
}
```

### Kalman: Individual Observation Noise

```{r ch17_kalman_ind_r}
# We stored full fit objects for Session 1 in the fitting loop (Phase 0 Q3 fix).
# Extract r_value posteriors from those stored fits.
kalman_r_s1 <- imap_dfr(
  c("s1_dangerous", "s1_nutritious"),
  function(key, idx) {
    full_fits <- all_fits[[paste0(key, "_kalman_full")]]
    if (is.null(full_fits)) return(NULL)
    imap_dfr(seq_along(subjects), function(ji, ...) {
      fit_k <- full_fits[[ji]]
      if (is.null(fit_k)) return(NULL)
      tryCatch({
        fit_k$summary("r_value") |>
          mutate(
            subject  = subjects[ji],
            session  = 1L,
            decision = str_extract(key, "dangerous|nutritious")
          )
      }, error = function(e) NULL)
    })
  }
)

if (nrow(kalman_r_s1) > 0) {
  ggplot(kalman_r_s1, aes(x = mean, fill = decision)) +
    geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
    facet_wrap(~decision, labeller = labeller(decision = str_to_title)) +
    scale_fill_manual(values = c(dangerous = "#D55E00", nutritious = "#0072B2")) +
    labs(
      title    = "Distribution of Individual r_value Estimates (Session 1)",
      subtitle = "Posterior mean r_value per participant.",
      x = "r_value (posterior mean)", y = "Count"
    ) +
    theme(legend.position = "none")
}
```

### Rule Model: Individual Error Tolerance

```{r ch17_rule_ind_eps}
# Same approach: full fit objects stored for Session 1.
rule_eps_s1 <- imap_dfr(
  c("s1_dangerous", "s1_nutritious"),
  function(key, idx) {
    full_fits <- all_fits[[paste0(key, "_rule_full")]]
    if (is.null(full_fits)) return(NULL)
    imap_dfr(seq_along(subjects), function(ji, ...) {
      fit_r <- full_fits[[ji]]
      if (is.null(fit_r)) return(NULL)
      tryCatch({
        fit_r$summary("error_prob") |>
          mutate(
            subject  = subjects[ji],
            session  = 1L,
            decision = str_extract(key, "dangerous|nutritious")
          )
      }, error = function(e) NULL)
    })
  }
)

if (nrow(rule_eps_s1) > 0) {
  ggplot(rule_eps_s1, aes(x = mean, fill = decision)) +
    geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
    facet_wrap(~decision, labeller = labeller(decision = str_to_title)) +
    scale_fill_manual(values = c(dangerous = "#D55E00", nutritious = "#0072B2")) +
    labs(
      title    = "Distribution of Individual error_prob Estimates (Session 1)",
      subtitle = "Posterior mean error_prob per participant.",
      x = "error_prob (posterior mean)", y = "Count"
    ) +
    theme(legend.position = "none")
}
```

---

## 7. Posterior Predictive Checks (Track 1)

```{r ch17_ppc}
gcm_fit_s1d <- all_fits[["s1_dangerous_gcm"]]

if (!is.null(gcm_fit_s1d)) {
  ll_mat <- gcm_fit_s1d$draws("log_lik", format = "matrix")
  loo_gcm_s1d <- loo::loo(ll_mat, cores = 4)
  plot(loo_gcm_s1d, label_points = TRUE,
       main = "GCM Pareto-k: Session 1, Dangerous")

  cat("\nPareto-k summary for GCM, Session 1, Dangerous:\n")
  print(summary(loo_gcm_s1d$diagnostics$pareto_k))
}
```

---

## 8. Track 2: NPE Inference on Human Data

> **📦 Python dependency.** Track 2 requires the `sbi` Python package.
> All code blocks are gated behind `run_npe_track <- FALSE`. The alien-task
> simulators use the functions defined in Chapter 15 (re-defined below for
> self-containment). Training 20,000 simulations per model × 6 conditions
> (3 sessions × 2 decisions) takes substantial compute; consider running on
> a server.

Chapter 16 trained NPE networks for the **Kruschke task** (2 continuous features). The Alien Game uses **5 binary features** — a different feature space. Track 2 for human data therefore requires new NPE networks trained on the alien-task simulators from Chapter 15. We train one network per (session × decision) condition, fitting all three models to each observed human response vector.

### Forward Simulators (Alien Task)

```{r ch17_alien_simulators, eval=run_npe_track}
# Simulators re-defined from Ch. 15 for self-containment.
# In a shared-code workflow these would be sourced from a common file.

n_features_alien <- 5L

# ── GCM simulator for alien task ─────────────────────────────────────────────
# theta = (logit_w1, ..., logit_w4, log_c) — 5 params (w5 = 1 - sum(w1:w4))
make_alien_gcm_simulator <- function(obs_matrix, cat_feedback) {
  function(theta) {
    # Softmax over 5 weights via log-sum-exp normalisation
    raw_w <- c(theta[1:4], 0)  # w5 is reference; use softmax
    w     <- exp(raw_w - log(sum(exp(raw_w))))
    c_val <- exp(theta[5])
    n <- nrow(obs_matrix)
    probs <- numeric(n)
    mem_obs <- matrix(NA_real_, 0, n_features_alien)
    mem_cat <- numeric(0)
    for (i in seq_len(n)) {
      stim  <- obs_matrix[i, ]
      n_mem <- nrow(mem_obs)
      if (n_mem == 0 || length(unique(mem_cat)) < 2) {
        p <- 0.5
      } else {
        sims <- vapply(seq_len(n_mem), function(e)
          exp(-c_val * sum(w * abs(stim - mem_obs[e, ]))), numeric(1))
        s1 <- sum(sims[mem_cat == 1]); s0 <- sum(sims[mem_cat == 0])
        den <- s1 + s0
        p   <- if (den > 1e-9) s1 / den else 0.5
      }
      probs[i] <- pmax(1e-9, pmin(1 - 1e-9, p))
      mem_obs  <- rbind(mem_obs, stim)
      mem_cat  <- c(mem_cat, cat_feedback[i])
    }
    rbinom(n, 1, probs)
  }
}

# ── Kalman simulator for alien task ──────────────────────────────────────────
# theta = (log_r) — 1 parameter
make_alien_kalman_simulator <- function(obs_matrix, cat_feedback) {
  function(theta) {
    r_val <- exp(theta[1])
    r_mat <- diag(r_val, n_features_alien)
    mu0   <- rep(0.5, n_features_alien); mu1 <- rep(0.5, n_features_alien)
    sig0  <- diag(1.0, n_features_alien); sig1 <- diag(1.0, n_features_alien)
    n     <- nrow(obs_matrix)
    probs <- numeric(n)
    for (i in seq_len(n)) {
      x <- as.numeric(obs_matrix[i, ])
      lp0 <- tryCatch(mvtnorm::dmvnorm(x, mu0, sig0 + r_mat, log = TRUE),
                      error = function(e) -Inf)
      lp1 <- tryCatch(mvtnorm::dmvnorm(x, mu1, sig1 + r_mat, log = TRUE),
                      error = function(e) -Inf)
      if (all(is.finite(c(lp0, lp1)))) {
        lse <- log(exp(lp0) + exp(lp1))
        probs[i] <- pmax(1e-9, pmin(1 - 1e-9, exp(lp1 - lse)))
      } else { probs[i] <- 0.5 }
      if (cat_feedback[i] == 1) {
        S <- sig1 + r_mat; K <- sig1 %*% solve(S)
        mu1  <- as.numeric(mu1 + K %*% (x - mu1))
        IK   <- diag(n_features_alien) - K
        sig1 <- (IK %*% sig1 %*% t(IK) + K %*% r_mat %*% t(K))
        sig1 <- (sig1 + t(sig1)) / 2
      } else {
        S <- sig0 + r_mat; K <- sig0 %*% solve(S)
        mu0  <- as.numeric(mu0 + K %*% (x - mu0))
        IK   <- diag(n_features_alien) - K
        sig0 <- (IK %*% sig0 %*% t(IK) + K %*% r_mat %*% t(K))
        sig0 <- (sig0 + t(sig0)) / 2
      }
    }
    rbinom(n, 1, probs)
  }
}

# ── Rule particle filter simulator for alien task ─────────────────────────────
# theta = (logit_eps) — 1 parameter (N_particles fixed at N_PARTICLES_FIXED)
make_alien_rule_simulator <- function(obs_matrix, cat_feedback) {
  n_feat <- ncol(obs_matrix)
  function(theta) {
    eps    <- plogis(theta[1])
    n_part <- N_PARTICLES_FIXED
    n      <- nrow(obs_matrix)

    gen_rule <- function() {
      nd <- sample(seq_len(min(MAX_RULE_DIMS_ALIEN, n_feat)), 1)
      dm <- sort(sample(seq_len(n_feat), nd, replace = FALSE))
      list(dims = dm, thresh = rep(0.5, nd),
           ops  = sample(c(">", "<"), nd, replace = TRUE),
           logic = if (nd > 1) sample(c("AND", "OR"), 1) else "AND",
           pred  = sample(0:1, 1))
    }
    eval_rule <- function(r, x) {
      tryCatch({
        ec <- function(d, t, o) switch(o, ">" = x[d] > t, "<" = x[d] < t, FALSE)
        if (length(r$dims) == 1) { cond <- ec(r$dims, r$thresh, r$ops)
        } else {
          cs <- mapply(ec, r$dims, r$thresh, r$ops)
          cond <- if (r$logic == "AND") all(cs) else any(cs)
        }
        ifelse(cond, r$pred, 1L - r$pred)
      }, error = function(e) sample(0:1, 1))
    }

    parts <- lapply(seq_len(n_part), function(i) gen_rule())
    wts   <- rep(1 / n_part, n_part)
    probs <- numeric(n)

    for (i in seq_len(n)) {
      x  <- as.numeric(obs_matrix[i, ])
      pd <- vapply(parts, eval_rule, x = x, numeric(1))
      probs[i] <- pmax(1e-9, pmin(1 - 1e-9,
        sum(wts * ifelse(pd == 1, 1 - eps, eps))))
      lk  <- vapply(parts, function(p) {
        pr <- eval_rule(p, x)
        if (pr == cat_feedback[i]) 1 - eps else eps
      }, numeric(1))
      nw  <- wts * lk; tot <- sum(nw)
      wts <- if (tot > 1e-9) nw / tot else rep(1 / n_part, n_part)
      ess <- 1 / sum(wts^2)
      if (!is.nan(ess) && ess < n_part * RESAMPLE_THRESHOLD) {
        idx   <- sample.int(n_part, n_part, replace = TRUE, prob = wts)
        parts <- parts[idx]; wts <- rep(1 / n_part, n_part)
      }
    }
    rbinom(n, 1, probs)
  }
}

cat("Alien-task forward simulators defined.\n")
```

### Training NPE and Inference on Human Data

We train one set of NPE networks per condition (session 1 shown as example; extend to all 6 conditions for full analysis). The trained networks are then applied to each participant's response vector.

```{r ch17_npe_train_human, eval=run_npe_track}
library(reticulate)
sbi   <- import("sbi")
torch <- import("torch")
utils <- import("sbi.utils")
infer_sbi <- import("sbi.inference")

n_train_alien <- 20000L

# ── Train NPE for Session 1, Dangerous ───────────────────────────────────────
sess_example <- 1; dec_example <- "dangerous"
sess_data_ex <- training_data |> filter(session == sess_example)

# Use the first subject's schedule as the reference schedule for training
ref_subj <- subjects[1]
obs_mat_ref <- sess_data_ex |>
  filter(subject == ref_subj) |>
  arrange(trial_within_subject) |>
  dplyr::select(f1:f5) |>
  as.matrix()
cat_fb_ref <- sess_data_ex |>
  filter(subject == ref_subj) |>
  arrange(trial_within_subject) |>
  pull(!!sym(dec_example))

n_trials_ex <- nrow(obs_mat_ref)

sim_gcm_ex    <- make_alien_gcm_simulator(obs_mat_ref, cat_fb_ref)
sim_kalman_ex <- make_alien_kalman_simulator(obs_mat_ref, cat_fb_ref)
sim_rule_ex   <- make_alien_rule_simulator(obs_mat_ref, cat_fb_ref)

train_alien_npe <- function(simulator_fn, prior_low, prior_high, n_params,
                             label, n_sims = n_train_alien) {
  cache_file <- sprintf("simdata/W17_npe_%s_%s_s%d.rds",
                        label, dec_example, sess_example)
  if (regenerate_fits || !file.exists(cache_file)) {
    cat("Generating", n_sims, "simulations for", label, "...\n")
    set.seed(42)
    thetas <- matrix(runif(n_sims * n_params, min = prior_low, max = prior_high),
                     ncol = n_params, byrow = TRUE)
    resps  <- future_map(seq_len(n_sims), function(i) simulator_fn(thetas[i, ]),
                         .options = furrr_options(seed = TRUE), .progress = TRUE)
    y_mat  <- do.call(rbind, resps)
    saveRDS(list(thetas = thetas, responses = y_mat), cache_file)
  } else {
    cached <- readRDS(cache_file); thetas <- cached$thetas; y_mat <- cached$responses
  }
  theta_t <- torch$tensor(thetas, dtype = torch$float32)
  x_t     <- torch$tensor(y_mat,  dtype = torch$float32)
  sbi_prior <- utils$BoxUniform(
    low  = torch$tensor(prior_low,  dtype = torch$float32),
    high = torch$tensor(prior_high, dtype = torch$float32)
  )
  inference_obj <- infer_sbi$SNPE(
    prior = sbi_prior,
    density_estimator = utils$posterior_nn(
      model = "nsf", hidden_features = 64L, num_transforms = 5L,
      embedding_net = torch$nn$Sequential(
        torch$nn$Linear(as.integer(n_trials_ex), 128L), torch$nn$ReLU(),
        torch$nn$Linear(128L, 64L), torch$nn$ReLU(),
        torch$nn$Linear(64L, 32L)
      )
    )
  )
  inference_obj$append_simulations(theta_t, x_t)
  dens <- inference_obj$train(training_batch_size = 256L, learning_rate = 5e-4,
                               show_train_summary = TRUE)
  list(posterior = inference_obj$build_posterior(dens),
       thetas = thetas, responses = y_mat)
}

npe_gcm_alien <- train_alien_npe(
  sim_gcm_ex,
  prior_low  = c(rep(-3, 4), -1),   # logit(w1:w4), log_c
  prior_high = c(rep(3,  4),  3),
  n_params = 5, label = "gcm"
)
npe_kalman_alien <- train_alien_npe(
  sim_kalman_ex,
  prior_low  = c(-2), prior_high = c(3),
  n_params = 1, label = "kalman"
)
npe_rule_alien <- train_alien_npe(
  sim_rule_ex,
  prior_low  = c(LOGIT_ERROR_PRIOR_MEAN - 3 * LOGIT_ERROR_PRIOR_SD),
  prior_high = c(LOGIT_ERROR_PRIOR_MEAN + 3 * LOGIT_ERROR_PRIOR_SD),
  n_params = 1, label = "rule"
)

cat("All three alien-task NPE networks trained for Session 1, Dangerous.\n")
```

### SBC Validation for Alien-Task NPE Networks

Full 1000-iteration Simulation-Based Calibration Checks (SBC) as promised in Chapter 16. We run 1000 SBC iterations for the GCM alien-task network when `run_intensive_checks = TRUE`; a 200-iteration spot-check otherwise.

```{r ch17_npe_sbc, eval=run_npe_track}
n_sbc <- if (run_intensive_checks) 1000L else 200L
cat("Running SBC with", n_sbc, "iterations for GCM alien-task NPE...\n")
set.seed(99)

sbc_thetas_alien <- matrix(
  runif(n_sbc * 5, min = c(rep(-3, 4), -1), max = c(rep(3, 4), 3)),
  ncol = 5, byrow = TRUE
)
sbc_ranks_alien <- matrix(0L, nrow = n_sbc, ncol = 5)

for (i in seq_len(n_sbc)) {
  y_i  <- sim_gcm_ex(sbc_thetas_alien[i, ])
  xi   <- torch$tensor(matrix(as.numeric(y_i), nrow = 1), dtype = torch$float32)
  post_i <- npe_gcm_alien$posterior$sample(c(1000L), x = xi)$numpy()
  for (p in 1:5) sbc_ranks_alien[i, p] <- sum(post_i[, p] < sbc_thetas_alien[i, p])
}

colnames(sbc_ranks_alien) <- c("logit_w1", "logit_w2", "logit_w3", "logit_w4", "log_c")
as_tibble(sbc_ranks_alien) |>
  pivot_longer(everything(), names_to = "parameter", values_to = "rank") |>
  ggplot(aes(x = rank)) +
  geom_histogram(bins = 20, fill = "#009E73", alpha = 0.7) +
  geom_hline(yintercept = n_sbc / 20, linetype = "dashed", color = "red") +
  facet_wrap(~parameter) +
  labs(
    title    = paste0("SBC (", n_sbc, " iterations): GCM NPE, Alien Task"),
    subtitle = "Should be approximately uniform.",
    x = "Rank", y = "Count"
  )
```

### Posterior Inference on Human Participants

```{r ch17_npe_human_inference, eval=run_npe_track}
n_post_alien <- 5000L

# Apply NPE to each participant for Session 1, Dangerous
human_npe_posteriors <- map_dfr(seq_along(subjects), function(ji) {
  subj  <- subjects[ji]
  s_dat <- training_data |>
    filter(session == sess_example, subject == subj) |>
    arrange(trial_within_subject)
  if (nrow(s_dat) == 0) return(NULL)

  y_subj <- s_dat |> pull(!!sym(dec_example))
  obs_t  <- torch$tensor(matrix(as.numeric(y_subj), nrow = 1),
                          dtype = torch$float32)

  # GCM posterior
  gcm_post  <- npe_gcm_alien$posterior$sample(c(n_post_alien), x = obs_t)$numpy()
  # Kalman posterior
  kal_post  <- npe_kalman_alien$posterior$sample(c(n_post_alien), x = obs_t)$numpy()
  # Rule posterior
  rule_post <- npe_rule_alien$posterior$sample(c(n_post_alien), x = obs_t)$numpy()

  bind_rows(
    tibble(subject = subj, model = "GCM",
           param = "c", value = mean(exp(gcm_post[, 5]))),
    tibble(subject = subj, model = "Kalman",
           param = "r", value = mean(exp(kal_post[, 1]))),
    tibble(subject = subj, model = "Rule",
           param = "eps", value = mean(plogis(rule_post[, 1])))
  )
})

if (nrow(human_npe_posteriors) > 0) {
  ggplot(human_npe_posteriors,
         aes(x = value, fill = model)) +
    geom_histogram(bins = 20, alpha = 0.7) +
    facet_wrap(~model, scales = "free_x") +
    scale_fill_viridis_d(option = "plasma") +
    labs(
      title    = "NPE Posterior Means by Participant (Session 1, Dangerous)",
      subtitle = "Distribution of posterior-mean parameters across participants.",
      x = "Posterior mean", y = "Count"
    ) +
    theme(legend.position = "none")
}
```

---

## 9. Synthesis and Discussion

### Which Model Best Accounts for Human Categorization?

The stacking weights from Section 5 provide a Track 1 summary: models with near-zero stacking weight add no predictive information beyond the other models for this session and decision. The Track 2 NPE posteriors (when `run_npe_track = TRUE`) provide the cognitive-theory complement: which model's forward simulator best accounts for individual response patterns.

**Key predictions from Chapter 15** to evaluate:

**GCM predictions**: Optimal attention weights should concentrate on the truly diagnostic features. Evaluate with the posterior attention weight plots in Section 6.

**Kalman filter predictions**: Model should struggle most with Session 2 nutritious and Session 3 nutritious rules where the category distribution is not unimodal.

**Rule model predictions**: Performance should be highest in Session 1 where rules are simple conjunctions. The model may show abrupt learning patterns.

### Individual Differences

The individual-level fits tell a richer story than the population average. Participants may genuinely differ in strategy: some showing GCM-like attention focusing (high estimated $c$, weights concentrated on diagnostic features), others showing Kalman-like smooth prototype updating (low $r$), and others showing rule-like learning (low $\varepsilon$, abrupt performance changes).

### The Condition Effect (Individual vs. Pair)

Do pairs show different model fit patterns from individuals? Possible hypotheses:

* **Cognitive division of labour**: Pairs may divide attention between the dangerous and nutritious dimensions.
* **Rule verbalization**: Pairs may more easily articulate and apply explicit rules, leading to higher stacking weight for the rule model.
* **Error correction**: Pairs may catch each other's errors, leading to lower estimated $\varepsilon$ or lower $r$.

---

## 10. Conclusion

This chapter applied the full Bayesian modeling pipeline — developed across Chapters 12–16 — to real human data from the Alien Game.

1. **Data quality**: The experimental design was verified against the stimulus rules; the `nutricious` typo was corrected; the response mapping was validated against the `correct` column. Data loading halts on rule-verification failure.

2. **Descriptive patterns**: Human learning shows the expected complexity gradient across sessions.

3. **Track 1 model comparison**: The ELPD comparison and stacking weights identify which model architecture best predicts held-out trial responses for each session and dimension. The multilevel-GCM vs. per-subject asymmetry is acknowledged (Phase 0 Q3).

4. **LFO-CV**: Moment-matched LOO is provided as an improved approximation for path-dependent models; the full `psis_lfo_gcm()` / `psis_lfo_kalman()` implementations from Chapters 12 and 13 can be called directly for exact LFO-CV.

5. **Individual differences**: The multilevel GCM provides population-level attention weight estimates. Individual Kalman and rule parameters are extracted from stored Session 1 fits.

6. **Track 2 (NPE)**: Alien-task NPE networks are trained on the Chapter 15 simulators, validated with Simulation-Based Calibration Checks, and applied to individual participant data when `run_npe_track = TRUE`.
