---
title: "12 — Prototype Models: Learning Category Averages"
output:
  html_document:
    mathjax: default
    toc: true
    toc_float: true
  pdf_document:
    toc: true
date: "2025-04-28"
---

```{r 12_setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)

# Set to TRUE to rerun simulations and Stan fits.
# FALSE loads pre-computed results from 'simdata/'.
regenerate_simulations <- FALSE

pacman::p_load(
  tidyverse,   # data manipulation and visualization
  mvtnorm,     # multivariate normal densities
  patchwork,   # combining plots
  cmdstanr,    # Stan interface
  posterior,   # tidy posterior arrays
  tidybayes,   # tidy extraction of draws
  bayesplot,   # MCMC diagnostic plots
  loo,         # LOO-CV
  SBC,         # Simulation-Based Calibration
  priorsense,  # prior sensitivity
  future,      # parallel processing
  furrr,       # parallel functional programming
  ellipse,     # uncertainty ellipses
  MASS         # ginv (pseudo-inverse fallback)
)

theme_set(theme_classic())

# Prior hyperparameter for log(r)
LOG_R_PRIOR_MEAN <- 0          # prior median r = 1
LOG_R_PRIOR_SD   <- 1.0        # covers r ≈ 0.13–7.4 at ±2 SD

for (d in c("stan", "simdata", "simmodels")) {
  if (!dir.exists(d)) dir.create(d)
}

# ── Kruschke (1993) stimulus set ─────────────────────────────────────────────
# Re-defined here for standalone use; identical to Ch. 11.
stimulus_info <- tibble(
  stimulus      = c(5, 3, 7, 1, 8, 2, 6, 4),
  height        = c(1, 1, 2, 2, 3, 3, 4, 4),
  position      = c(2, 3, 1, 4, 1, 4, 2, 3),
  category_true = c(0, 0, 1, 0, 1, 0, 1, 1)
)

n_blocks         <- 8
n_stim_per_block <- nrow(stimulus_info)
total_trials     <- n_blocks * n_stim_per_block

# ── Per-subject schedule helper ───────────────────────────────────────────────
# Identical to Ch. 11. Redefined here for standalone use.
make_subject_schedule <- function(stimulus_info, n_blocks, seed) {
  set.seed(seed)
  n_stim <- nrow(stimulus_info)
  sequence <- unlist(lapply(seq_len(n_blocks), function(b) {
    sample(stimulus_info$stimulus, n_stim, replace = FALSE)
  }))
  tibble(
    trial_within_subject = seq_along(sequence),
    block                = rep(seq_len(n_blocks), each = n_stim),
    stimulus_id          = sequence
  ) |>
    left_join(stimulus_info, by = c("stimulus_id" = "stimulus")) |>
    rename(category_feedback = category_true) |>
    dplyr::select(trial_within_subject, block, stimulus_id,
                  height, position, category_feedback)
}

plan(multisession, workers = max(1, availableCores() - 1))
```

> **📍 Where we are in the Bayesian modeling workflow:**
> Chapter 11 built the GCM — an exemplar model that stores every encountered
> stimulus and decides by summed similarity to memory. This chapter introduces
> a structurally different account: the **prototype model**, which abstracts
> each category to a single running estimate (its mean and uncertainty) rather
> than storing individual instances. We implement prototype learning as a
> multivariate Kalman filter and then, for the first time in this module, add a
> **Stan parameter-estimation model** for the prototype architecture. The full
> validation pipeline from Ch. 5 — prior predictive check, Pathfinder
> initialization, parameter recovery, SBC, LOO, and prior sensitivity — is
> applied to this new model. The same LFO-CV caveat from Ch. 11 applies: the
> Kalman filter is path-dependent, so standard LOO is an approximation only.

# Prototype-Based Models of Categorization

## Rethinking Category Representation: From Examples to Averages

In the previous chapter, we explored the Generalized Context Model (GCM), an exemplar model where every encountered example is stored in memory. This approach is powerful but raises questions about cognitive efficiency. Do we really store every single instance we encounter?

Consider learning about common categories like "birds" or "chairs". While specific examples matter, we also seem to develop a general sense of what constitutes a typical member. Prototype theory offers an alternative perspective: instead of storing individual exemplars, the mind abstracts a summary representation — a prototype — for each category. This prototype often represents the central tendency or "average" member.

## Core Ideas of Prototype Models

* **Abstraction, Not Storage**: The primary representation is a summary (the prototype), not a collection of instances.
* **Cognitive Economy**: Storing one prototype per category is potentially much more memory-efficient than storing numerous exemplars.
* **Similarity to Prototype**: New items are categorized based on their similarity to the stored prototypes of different categories.
* **Typicality**: Items closer to a category's prototype are judged as more "typical" members and are often categorized faster and more accurately. This contrasts with exemplar models, which emphasize detailed memory for specific instances.

| Feature | Prototype Models | Exemplar Models (GCM) |
|---|---|---|
| Representation | Abstract summary (e.g., average) | Collection of individual examples |
| Memory Cost | Low (one prototype per category) | High (potentially all examples) |
| Key Information | Central tendency & variance | Specific instances & their labels |
| Sensitivity | Less sensitive to specific examples | Highly sensitive to specific examples |

## Modeling Dynamic Prototypes: The Kalman Filter

Early prototype models often calculated a static average. However, human learning is dynamic; our understanding evolves with experience. How can we model a prototype that updates incrementally as new examples are encountered?

The Kalman filter provides a suitable mathematical framework. Originally developed for tracking physical systems amidst noise, it's well-suited for modeling prototype learning because it allows us to:

* **Track an Estimate**: The category prototype (its average feature values).
* **Represent Uncertainty**: Maintain not just the prototype's location but also our uncertainty about it.
* **Update Incrementally**: Refine the prototype with each new example.
* **Balance Old and New**: Optimally combine the current prototype (prior belief) with the new example (evidence).
* **Implement an Adaptive Learning Rate**: Automatically adjust how much influence a new example has (the "Kalman gain") based on current uncertainty. Learning is faster when uncertainty is high and slows as confidence grows.

### The Kalman Filter Logic: Updating an Estimate

Imagine tracking a prototype based on a single feature (e.g., average height). The Kalman filter maintains your Current State. At any moment, you have two key pieces of information:

* **Your Best Guess ($\mu$):** This is your current estimate of the prototype's average height.
* **Your Confidence ($\sigma^2$):** This isn't just *if* you're confident, but *how much*. It's the *variance* around your guess. A small variance means you're quite sure; a large variance means you're very unsure.

When a new category member (observation $x$) arrives, you measure its height.

* **The Catch:** Your measurement tool (or the example itself) isn't perfect. There's some inherent "noise" or variability in how examples relate to the true prototype. We represent the uncertainty of this *measurement process* with another variance, often called $R$.

**The Core Problem:** You have your guess (with $\sigma^2_{prev}$) and a new measurement (with $R$). How do you combine them to get a *new, better guess* and an *updated level of uncertainty*?

**Enter the Kalman Gain ($K$): The "Trust" Knob**

The Kalman filter calculates the Kalman Gain $K$, which tells you how much to trust the *new measurement* compared to your *current guess*:

$$K = \frac{\sigma^2_{prev}}{\sigma^2_{prev} + R}$$

If $\sigma^2_{prev}$ is large relative to $R$, then $K \to 1$: trust the measurement almost completely. If $\sigma^2_{prev}$ is small relative to $R$, then $K \to 0$: stick closer to your previous guess.

**Updating Your Guess:**

You nudge your old guess towards the new measurement. The size of the nudge is determined by the Kalman Gain $K$:

$$\mu_{new} = \mu_{prev} + K \cdot (x - \mu_{prev})$$

The term $(x - \mu_{prev})$ is the "prediction error" or "innovation" — how much the new data differs from what you expected.

**Updating Your Confidence (Uncertainty):**

After incorporating new information, uncertainty decreases:

$$\sigma^2_{new} = (1 - K) \cdot \sigma^2_{prev}$$

Since $K \in (0, 1)$, multiplying by $(1 - K)$ always shrinks uncertainty. The more you trusted the measurement (higher $K$), the more your uncertainty shrinks.

**Note**: This simple implementation assumes a static target (a true underlying prototype), so uncertainty monotonically decreases. In reality, the prototype may change over time, and more complex implementations introduce a **process noise term** $Q$. Before you get the next measurement, you predict how the system state might change over time. This prediction step increases uncertainty — because real-world processes aren't perfectly predictable — then the measurement update decreases it. If the state evolves as $x_k = F x_{k-1} + w_k$ (where $w_k$ is process noise with covariance $Q$), the predicted covariance becomes $P_{k|k-1} = F P_{k-1|k-1} F^T + Q$. The addition of $Q$ explicitly increases uncertainty between observations. We model categories as static targets here, which is appropriate for the blocked learning paradigm we use.

### Extending to Multiple Features: Multivariate Kalman Filter

Our stimuli have two features (height and position). We need a multivariate version:

* **Your Current State (Vectors and Matrices):**
    * **Your Best Guess ($\vec{\mu}$):** Now a *vector* containing your best guess for each feature.
    * **Your Confidence ($\Sigma$):** Now a *covariance matrix*. The diagonal elements give the variance for each feature individually; the off-diagonal elements give their covariance — for example, whether taller prototypes are also typically further to the right.

* **A New Hint Arrives:** Your new observation ($\vec{x}$) is also a vector.
    * **Measurement Noise ($R$):** Also a covariance matrix. Often simplified to a diagonal matrix, assuming measurement errors on different features are independent.

* **The Logic Stays the Same, the Math Uses Matrices:**
    * **Kalman Gain ($K$):** A *matrix*: $K = \Sigma_{prev} (\Sigma_{prev} + R)^{-1}$
    * **Updating Your Guess ($\vec{\mu}$):** $\vec{\mu}_{new} = \vec{\mu}_{prev} + K (\vec{x} - \vec{\mu}_{prev})$
    * **Updating Your Confidence ($\Sigma$):** We use the numerically stable Joseph form: $\Sigma_{new} = (I - K)\Sigma_{prev}(I - K)^T + KRK^T$. While it looks complex, the intuition is the same as the scalar case: uncertainty decreases after each update. The Joseph form is preferred because it guarantees the result remains symmetric and positive semi-definite even in the presence of floating-point rounding errors.

### In a Nutshell

The Kalman filter is a recursive algorithm that provides the mathematically optimal way to estimate the state of a system (like the features of a prototype) given a series of noisy measurements. It works by:

1. Maintaining an estimate ($\vec{\mu}$) and uncertainty ($\Sigma$).
2. Calculating a Kalman Gain $K$ that balances trust in the current estimate against trust in the new measurement.
3. Updating the estimate by moving it towards the new measurement, weighted by $K$.
4. Updating (reducing) the uncertainty based on the information gained.

The multivariate version simply uses vectors and matrices to do this for multiple, potentially correlated variables simultaneously.

### Implementing the Multivariate Update in R

```{r, 12 multivariate_kalman_update}
# One update step of a multivariate Kalman filter.
# Returns updated mean vector, covariance matrix, and Kalman gain.
multivariate_kalman_update <- function(mu_prev,      # previous mean vector
                                       sigma_prev,    # previous covariance matrix
                                       observation,   # observed feature vector
                                       r_matrix) {    # observation noise matrix
  mu_prev     <- as.numeric(mu_prev)
  observation <- as.numeric(observation)
  sigma_prev  <- as.matrix(sigma_prev)
  r_matrix    <- as.matrix(r_matrix)

  n_dim <- length(mu_prev)
  I     <- diag(n_dim)

  # Kalman gain: K = Sigma_prev * (Sigma_prev + R)^{-1}
  S     <- sigma_prev + r_matrix
  S_inv <- tryCatch(
    solve(S),
    error = function(e) {
      warning("Matrix inversion failed; using pseudo-inverse.", call. = FALSE)
      MASS::ginv(S)
    }
  )
  K <- sigma_prev %*% S_inv

  # Update mean
  innovation <- observation - mu_prev
  mu_new     <- as.numeric(mu_prev + K %*% innovation)

  # Update covariance — Joseph form for numerical stability
  IK        <- I - K
  sigma_new <- IK %*% sigma_prev %*% t(IK) + K %*% r_matrix %*% t(K)

  # Enforce symmetry (floating-point can introduce tiny asymmetry)
  sigma_new <- (sigma_new + t(sigma_new)) / 2

  list(mu = mu_new, sigma = sigma_new, k = K)
}
```

This function performs a single update for one category's prototype when it observes a new member.

---

## Building the Full Categorization Model

We now integrate this update mechanism into a full agent that learns prototypes for two categories and makes categorization decisions.

### Initialization

* We need separate prototypes ($\mu_0, \Sigma_0$ and $\mu_1, \Sigma_1$) for Category 0 and Category 1.
* We start with high uncertainty (large initial $\Sigma$) and uninformative means at the center of the feature space.
* We define the observation noise matrix $R = r_\text{val} \cdot I$. The parameter $r_\text{val}$ represents how much variability we assume exists within categories or in our perception of stimuli.

### Categorization Decision

How does the model decide the category for a new stimulus $\vec{x}$? It compares $\vec{x}$ to both prototypes, considering their current means ($\vec{\mu}$) and uncertainties ($\Sigma$). We use the probability density of the multivariate normal distribution:

$$P(\vec{x} \mid \text{Category } C) \propto \exp\left(-\frac{1}{2}(\vec{x} - \vec{\mu}_C)^T (\Sigma_C + R)^{-1} (\vec{x} - \vec{\mu}_C)\right)$$

The probability of choosing Category 1 is then calculated using the Luce choice rule (softmax):

$$P(\text{Choose Cat 1} \mid \vec{x}) = \frac{P(\vec{x} \mid \text{Category } 1)}{P(\vec{x} \mid \text{Category } 0) + P(\vec{x} \mid \text{Category } 1)}$$

We work with log-probabilities for numerical stability.

### The Learning Loop

The agent processes trials sequentially:

1. Observe stimulus $\vec{x}_i$.
2. Calculate $P(\text{Choose Cat 1} \mid \vec{x}_i)$ based on current $\mu_0, \Sigma_0, \mu_1, \Sigma_1$.
3. Generate a response.
4. Receive feedback (true category $y_i$).
5. Update the prototype of the correct category using the multivariate Kalman update.

### R Implementation of the Prototype Agent

The function now accepts `subject_seed` and calls `make_subject_schedule` internally — consistent with the path-dependence fix from Ch. 11. Each agent's learning trajectory depends on the order in which stimuli are encountered, and different randomizations produce genuinely different trajectories.

```{r, 12 prototype_model}
# Prototype model agent using Kalman filter for categorization.
# r_value   : observation noise variance (scalar > 0)
# obs       : matrix of observations (trials x features)
# cat_one   : vector of true category labels (0 or 1) for feedback
# Returns:  tibble with trial_within_subject, sim_response, prob_cat1,
#           correct, and performance columns alongside schedule columns.
prototype_kalman <- function(r_value,
                             obs,
                             cat_one,
                             initial_mu         = NULL,
                             initial_sigma_diag = 10.0,
                             quiet              = TRUE) {
  n_trials  <- nrow(obs)
  n_features <- ncol(obs)

  if (is.null(initial_mu)) {
    mu0_init <- rep(2.5, n_features)
    mu1_init <- rep(2.5, n_features)
  } else {
    mu0_init <- initial_mu[[1]]
    mu1_init <- initial_mu[[2]]
  }

  prototype_cat_0 <- list(mu = mu0_init, sigma = diag(initial_sigma_diag, n_features))
  prototype_cat_1 <- list(mu = mu1_init, sigma = diag(initial_sigma_diag, n_features))
  r_matrix        <- diag(r_value, n_features)

  response_probs <- numeric(n_trials)

  log_sum_exp <- function(v) {
    m <- max(v)
    m + log(sum(exp(v - m)))
  }

  for (i in seq_len(n_trials)) {
    if (!quiet && i %% 20 == 0) cat("Trial", i, "\n")

    current_obs <- as.numeric(obs[i, ])

    # ── Decision ────────────────────────────────────────────────────────────
    cov_cat_0 <- prototype_cat_0$sigma + r_matrix
    cov_cat_1 <- prototype_cat_1$sigma + r_matrix

    log_prob_0 <- tryCatch(
      mvtnorm::dmvnorm(current_obs, mean = prototype_cat_0$mu,
                       sigma = cov_cat_0, log = TRUE),
      error = function(e) -Inf
    )
    log_prob_1 <- tryCatch(
      mvtnorm::dmvnorm(current_obs, mean = prototype_cat_1$mu,
                       sigma = cov_cat_1, log = TRUE),
      error = function(e) -Inf
    )

    if (!is.finite(log_prob_0) && !is.finite(log_prob_1)) {
      prob_cat_1 <- 0.5
    } else if (!is.finite(log_prob_0)) {
      prob_cat_1 <- 1.0
    } else if (!is.finite(log_prob_1)) {
      prob_cat_1 <- 0.0
    } else {
      # No explicit bias parameter in the base model (b = 0.5)
      prob_cat_1 <- exp(log_prob_1 - log_sum_exp(c(log_prob_0, log_prob_1)))
    }
    response_probs[i] <- pmax(1e-9, pmin(1 - 1e-9, prob_cat_1))

    # ── Update ───────────────────────────────────────────────────────────────
    true_cat <- cat_one[i]
    if (true_cat == 1) {
      upd <- multivariate_kalman_update(prototype_cat_1$mu, prototype_cat_1$sigma,
                                        current_obs, r_matrix)
      prototype_cat_1$mu    <- upd$mu
      prototype_cat_1$sigma <- upd$sigma
    } else {
      upd <- multivariate_kalman_update(prototype_cat_0$mu, prototype_cat_0$sigma,
                                        current_obs, r_matrix)
      prototype_cat_0$mu    <- upd$mu
      prototype_cat_0$sigma <- upd$sigma
    }
  }

  tibble(
    prob_cat1    = response_probs,
    sim_response = rbinom(n_trials, 1, response_probs)
  )
}

# Wrapper: generates per-agent schedule then calls prototype_kalman
simulate_prototype_agent <- function(agent_id, r_value,
                                     stimulus_info, n_blocks, subject_seed) {
  schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  obs      <- as.matrix(schedule[, c("height", "position")])
  cat_one  <- schedule$category_feedback

  result <- prototype_kalman(
    r_value            = r_value,
    obs                = obs,
    cat_one            = cat_one,
    initial_mu         = list(rep(2.5, 2), rep(2.5, 2)),
    initial_sigma_diag = 10.0,
    quiet              = TRUE
  )

  schedule |>
    mutate(
      agent_id     = agent_id,
      r_value_true = r_value,
      log_r_true   = log(r_value),
      prob_cat1    = result$prob_cat1,
      sim_response = result$sim_response,
      correct      = as.integer(category_feedback == sim_response)
    ) |>
    group_by(agent_id) |>
    mutate(performance = cumsum(correct) / row_number()) |>
    ungroup()
}
```

### Simulating Categorization Behavior

Let's simulate behavior on the Kruschke (1993) task. The key parameter here is `r_value`, the observation noise. This parameter reflects the model's assumption about how much variability exists within categories or arises from perceptual noise.

* **Low `r_value`**: The model assumes observations are precise representations of the category; each new example exerts strong influence and prototypes update rapidly.
* **High `r_value`**: The model assumes observations are noisy; learning is slower as each example has less influence, but the model generalizes more broadly.

```{r, 12 simulate_prototype_responses}
param_df <- expand_grid(
  agent_id  = 1:10,
  r_value   = c(0.1, 0.5, 1.0, 2.0, 5.0)
) |>
  mutate(subject_seed = agent_id)    # unique, reproducible seed per agent

sim_file <- "simdata/W12_prototype_simulated_responses.csv"
if (regenerate_simulations || !file.exists(sim_file)) {
  cat("Regenerating prototype simulations...\n")
  prototype_responses <- future_pmap_dfr(
    list(
      agent_id     = param_df$agent_id,
      r_value      = param_df$r_value,
      subject_seed = param_df$subject_seed
    ),
    simulate_prototype_agent,
    stimulus_info = stimulus_info,
    n_blocks      = n_blocks,
    .options = furrr_options(seed = TRUE)
  )
  write_csv(prototype_responses, sim_file)
  cat("Simulations saved.\n")
} else {
  prototype_responses <- read_csv(sim_file, show_col_types = FALSE)
  cat("Simulations loaded.\n")
}

ggplot(prototype_responses,
       aes(x = trial_within_subject, y = performance, color = factor(r_value))) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.15,
               aes(fill = factor(r_value))) +
  scale_color_viridis_d(option = "plasma", name = "r_value") +
  scale_fill_viridis_d(option = "plasma", name = "r_value") +
  labs(
    title    = "Prototype Model Learning Performance",
    subtitle = "Effect of Observation Noise (r_value); mean ± SE across agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  ylim(0.4, 1.0)
```

**Interpretation**: Different values of `r_value` impact learning speed and asymptotic accuracy. Very low noise (`r_value = 0.1`) means each example exerts near-complete influence on the prototype, which can lead to rapid initial learning but sensitivity to early idiosyncrasies. Very high noise (`r_value = 5.0`) means the model trusts observations much less and learns more slowly. An intermediate value often achieves the best balance between responsiveness and robustness.

### Visualizing Prototype Evolution

Where do the prototypes end up? Let's track their movement and uncertainty over trials.

```{r, 12 track_prototypes}
# Track prototype means and covariance matrices over the course of learning.
# Returns a tibble with one row per (trial × category).
track_prototypes <- function(r_value, obs, cat_one,
                             initial_mu = NULL, initial_sigma_diag = 10.0) {
  n_trials  <- nrow(obs)
  n_features <- ncol(obs)

  if (is.null(initial_mu)) {
    mu0_init <- rep(2.5, n_features)
    mu1_init <- rep(2.5, n_features)
  } else {
    mu0_init <- initial_mu[[1]]
    mu1_init <- initial_mu[[2]]
  }
  prototype_cat_0 <- list(mu = mu0_init, sigma = diag(initial_sigma_diag, n_features))
  prototype_cat_1 <- list(mu = mu1_init, sigma = diag(initial_sigma_diag, n_features))
  r_matrix        <- diag(r_value, n_features)

  history <- list()

  for (i in seq_len(n_trials + 1)) {
    history[[length(history) + 1]] <- tibble(
      trial      = i, category = 0,
      feature1_mean = prototype_cat_0$mu[1],
      feature2_mean = prototype_cat_0$mu[2],
      cov_matrix    = list(prototype_cat_0$sigma)
    )
    history[[length(history) + 1]] <- tibble(
      trial      = i, category = 1,
      feature1_mean = prototype_cat_1$mu[1],
      feature2_mean = prototype_cat_1$mu[2],
      cov_matrix    = list(prototype_cat_1$sigma)
    )

    if (i <= n_trials) {
      current_obs <- as.numeric(obs[i, ])
      true_cat    <- cat_one[i]
      if (true_cat == 1) {
        upd <- multivariate_kalman_update(prototype_cat_1$mu, prototype_cat_1$sigma,
                                          current_obs, r_matrix)
        prototype_cat_1$mu    <- upd$mu
        prototype_cat_1$sigma <- upd$sigma
      } else {
        upd <- multivariate_kalman_update(prototype_cat_0$mu, prototype_cat_0$sigma,
                                          current_obs, r_matrix)
        prototype_cat_0$mu    <- upd$mu
        prototype_cat_0$sigma <- upd$sigma
      }
    }
  }
  bind_rows(history)
}

# Ellipse helper for uncertainty visualization
get_ellipse <- function(mu, sigma, level = 0.68) {
  mu    <- as.numeric(mu)
  sigma <- as.matrix(sigma)
  if (length(mu) != 2 || !all(dim(sigma) == c(2, 2))) return(NULL)
  ev <- eigen(sigma, symmetric = TRUE, only.values = TRUE)$values
  if (any(ev <= 1e-6)) sigma <- sigma + diag(ncol(sigma)) * 1e-6
  pts <- tryCatch(
    ellipse::ellipse(sigma, centre = mu, level = level),
    error = function(e) NULL
  )
  if (is.null(pts)) return(NULL)
  as_tibble(pts) |> setNames(c("feature1_mean", "feature2_mean"))
}

# Use a fixed seed-1 schedule for the trajectory visualization
traj_schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = 1)

prototype_trajectory <- track_prototypes(
  r_value            = 1.0,
  obs                = as.matrix(traj_schedule[, c("height", "position")]),
  cat_one            = traj_schedule$category_feedback,
  initial_mu         = list(rep(2.5, 2), rep(2.5, 2)),
  initial_sigma_diag = 10.0
)

final_prototypes    <- prototype_trajectory |> filter(trial == max(trial))

ellipse_data <- final_prototypes |>
  rowwise() |>
  mutate(ell = list(get_ellipse(c(feature1_mean, feature2_mean), cov_matrix[[1]]))) |>
  ungroup() |>
  filter(!map_lgl(ell, is.null)) |>
  unnest(ell)

ggplot() +
  geom_point(data = stimulus_info,
             aes(x = position, y = height, color = factor(category_true),
                 shape = factor(category_true)),
             size = 3, alpha = 0.5) +
  geom_path(
    data = prototype_trajectory |> filter(trial <= total_trials),
    aes(x = feature2_mean, y = feature1_mean,
        group = category, color = factor(category)),
    linetype = "dashed",
    arrow = arrow(type = "closed", length = unit(0.08, "inches"), angle = 20)
  ) +
  geom_point(data = final_prototypes,
             aes(x = feature2_mean, y = feature1_mean, color = factor(category)),
             size = 4, shape = 18) +
  geom_path(data = ellipse_data,
            aes(x = feature2_mean, y = feature1_mean,
                group = category, color = factor(category)),
            alpha = 0.7, linewidth = 0.8) +
  scale_color_manual(values = c("0" = "#0072B2", "1" = "#D55E00"), name = "Category") +
  scale_shape_manual(values = c("0" = 16, "1" = 17), name = "Category") +
  labs(
    title    = "Prototype Learning Trajectory (Kalman Filter, r = 1.0)",
    subtitle = "Prototypes start at (2.5, 2.5). Dashed lines show path; ellipses show final 68% uncertainty.",
    x = "Position Feature", y = "Height Feature"
  ) +
  coord_cartesian(
    xlim = range(c(stimulus_info$position, prototype_trajectory$feature2_mean)) + c(-0.5, 0.5),
    ylim = range(c(stimulus_info$height,   prototype_trajectory$feature1_mean)) + c(-0.5, 0.5)
  )
```

**Interpretation**: The dashed lines show the path the prototypes took during learning (starting from (2.5, 2.5), the center of the feature space). The final prototype locations are marked with diamonds, and the ellipses represent the model's final uncertainty (68% confidence) about the true prototype location for each category. We see the prototypes moving towards the center of their respective category members and uncertainty shrinking over time — exactly the incremental, variance-reducing Bayesian update the Kalman filter implements.

---

## Prior Predictive Check

Before fitting any data, we verify that the prior on `log_r` generates only scientifically plausible learning trajectories. We draw $S = 500$ values from $\log r \sim \text{Normal}(0, 1)$ — implying $r \in (0.13, 7.4)$ at $\pm 2$ SD — simulate one agent per draw, and plot the envelope.

```{r, 12 prior predictive check}
n_ppc_samples <- 500
set.seed(42)

ppc_log_r  <- rnorm(n_ppc_samples, LOG_R_PRIOR_MEAN, LOG_R_PRIOR_SD)
ppc_r      <- exp(ppc_log_r)
ppc_sched  <- make_subject_schedule(stimulus_info, n_blocks, seed = 999)

prior_pred_curves <- map_dfr(seq_len(n_ppc_samples), function(s) {
  res <- prototype_kalman(
    r_value   = ppc_r[s],
    obs       = as.matrix(ppc_sched[, c("height", "position")]),
    cat_one   = ppc_sched$category_feedback,
    initial_mu = list(rep(2.5, 2), rep(2.5, 2))
  )
  tibble(
    sample_id = s,
    trial     = seq_len(nrow(ppc_sched)),
    correct   = as.integer(res$sim_response == ppc_sched$category_feedback)
  ) |>
    mutate(cum_acc = cumsum(correct) / row_number())
})

ppc_summary <- prior_pred_curves |>
  group_by(trial) |>
  summarise(
    q05 = quantile(cum_acc, 0.05), q25 = quantile(cum_acc, 0.25),
    q50 = median(cum_acc),
    q75 = quantile(cum_acc, 0.75), q95 = quantile(cum_acc, 0.95),
    .groups = "drop"
  )

ggplot(ppc_summary, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#009E73", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#009E73", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#006D4E", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Prior Predictive Check: Kalman Filter Prototype Model",
    subtitle = "Ribbons: 50% and 90% prior predictive intervals\nPrior: log(r) ~ Normal(0, 1)",
    x = "Trial", y = "Cumulative Accuracy"
  )
```

The prior predictive envelope spans chance performance (uninformative $r$ draws) through near-ceiling accuracy (informative $r$ draws), without implying impossible behavior such as sustained below-chance performance. This is the plausibility standard we require before proceeding to inference.

---

## Estimating Parameters: The Prototype Model in Stan

Simulations help us understand the model, but we want to fit it to experimental data to estimate its parameters. The primary free parameter is the observation noise variance `r_value`. We implement the Kalman filter prototype model in Stan to estimate `r_value` (on the log scale) from data.

### Why the Kalman filter in `transformed parameters`?

The prototype model has a structural advantage over the GCM for Stan implementation: the Kalman filter state is entirely deterministic given `log_r` and the data. There are no latent variables beyond the single scalar `log_r`. The full trial-by-trial Kalman state — all prototype means, covariances, and choice probabilities — can be pre-computed in `transformed parameters` and then referenced directly in the `model` and `generated quantities` blocks. This means, crucially, that **`log_lik[i]` in `generated quantities` simply reads `p[i]` from `transformed parameters`** — there is no need to re-run the Kalman filter in the GQ block. This is a genuine architectural advantage over the GCM, where the memory state was a local variable that had to be re-computed in every block.

The same path-dependence caveat from Ch. 11 applies: trial $t$'s choice probability depends on all previous stimuli through the Kalman state, so standard LOO is an approximation and LFO-CV is the correct alternative.

> **`reduce_sum` and parallelization**: As with the GCM, the Kalman prototype model cannot use `reduce_sum` because successive Kalman steps are causally chained — trial $t$'s contribution to the likelihood depends on the filter state from trials $1, \ldots, t-1$. The entire filter must be computed sequentially. However, because the filter state is deterministic and cheap to compute (only two $2 \times 2$ matrix operations per trial), the model is already fast without parallelization.

### Stan Model

```{r, 12 stan_prototype_model}
prototype_single_stan <- "
// Kalman Filter Prototype Model — Single Subject
// Key design:
//   log_r is the only parameter (unconstrained); r_value = exp(log_r)
//   The full Kalman filter sequence (prototype states + choice probs)
//   is computed in transformed parameters — a deterministic function
//   of log_r and the data. This means log_lik in generated quantities
//   simply reads p[i] from transformed parameters without re-running
//   the filter. No duplication, no local-block memory management.

data {
  int<lower=1> ntrials;
  int<lower=1> nfeatures;
  array[ntrials] int<lower=0, upper=1> cat_one;   // true category (feedback)
  array[ntrials] int<lower=0, upper=1> y;          // observed responses
  array[ntrials, nfeatures] real obs;              // stimulus features

  // Fixed structural values (passed as data for transparency)
  vector[nfeatures] initial_mu_cat0;
  vector[nfeatures] initial_mu_cat1;
  real<lower=0> initial_sigma_diag;

  // Prior hyperparameters
  real prior_logr_mean;
  real<lower=0> prior_logr_sd;
}

parameters {
  real log_r;
}

transformed parameters {
  real<lower=0> r_value = exp(log_r);

  // Choice probability at each trial, computed via the Kalman filter.
  // Stored as a transformed parameter so generated quantities can reference
  // it directly without re-running the filter.
  array[ntrials] real<lower=1e-9, upper=1-1e-9> p;

  {
    vector[nfeatures] mu_cat0 = initial_mu_cat0;
    vector[nfeatures] mu_cat1 = initial_mu_cat1;
    matrix[nfeatures, nfeatures] sigma_cat0 =
      diag_matrix(rep_vector(initial_sigma_diag, nfeatures));
    matrix[nfeatures, nfeatures] sigma_cat1 =
      diag_matrix(rep_vector(initial_sigma_diag, nfeatures));
    matrix[nfeatures, nfeatures] r_matrix =
      diag_matrix(rep_vector(r_value, nfeatures));
    matrix[nfeatures, nfeatures] I_mat =
      diag_matrix(rep_vector(1.0, nfeatures));

    for (i in 1:ntrials) {
      vector[nfeatures] x = to_vector(obs[i]);

      // ── Decision (before this trial's update) ──────────────────────────
      matrix[nfeatures, nfeatures] cov0 = sigma_cat0 + r_matrix;
      matrix[nfeatures, nfeatures] cov1 = sigma_cat1 + r_matrix;

      real log_p0 = multi_normal_lpdf(x | mu_cat0, cov0);
      real log_p1 = multi_normal_lpdf(x | mu_cat1, cov1);
      real prob1  = exp(log_p1 - log_sum_exp(log_p0, log_p1));

      p[i] = fmax(1e-9, fmin(1 - 1e-9, prob1));

      // ── Update (on all trials, including the last) ──────────────────────
      if (cat_one[i] == 1) {
        vector[nfeatures] innov = x - mu_cat1;
        matrix[nfeatures, nfeatures] S  = sigma_cat1 + r_matrix;
        // K = Sigma * S^{-1}; mdivide_right_spd is numerically stable
        matrix[nfeatures, nfeatures] K  = mdivide_right_spd(sigma_cat1, S);
        matrix[nfeatures, nfeatures] IK = I_mat - K;
        mu_cat1    = mu_cat1 + K * innov;
        sigma_cat1 = IK * sigma_cat1 * IK' + K * r_matrix * K';
        sigma_cat1 = 0.5 * (sigma_cat1 + sigma_cat1');
      } else {
        vector[nfeatures] innov = x - mu_cat0;
        matrix[nfeatures, nfeatures] S  = sigma_cat0 + r_matrix;
        matrix[nfeatures, nfeatures] K  = mdivide_right_spd(sigma_cat0, S);
        matrix[nfeatures, nfeatures] IK = I_mat - K;
        mu_cat0    = mu_cat0 + K * innov;
        sigma_cat0 = IK * sigma_cat0 * IK' + K * r_matrix * K';
        sigma_cat0 = 0.5 * (sigma_cat0 + sigma_cat0');
      }
    }
  }
}

model {
  target += normal_lpdf(log_r | prior_logr_mean, prior_logr_sd);
  target += bernoulli_lpmf(y | p);
}

generated quantities {
  // p[i] is already available from transformed parameters — no re-computation needed.
  // This is the architectural advantage of the Kalman formulation vs. the GCM.
  vector[ntrials] log_lik;
  for (i in 1:ntrials)
    log_lik[i] = bernoulli_lpmf(y[i] | p[i]);
}
"

stan_file_proto_single <- "stan/W12_prototype_single.stan"
write_stan_file(prototype_single_stan, dir = "stan/",
                basename = "W12_prototype_single.stan")
mod_prototype_single <- cmdstan_model(stan_file_proto_single)
cat("Stan model compiled successfully.\n")
```

**Key aspects of the Stan implementation:**

* **Parameters block**: A single scalar `log_r`. The constraint `r_value > 0` is handled implicitly by `exp(log_r)`.
* **Transformed parameters**: Runs the full Kalman filter trial-by-trial, updating the two prototype distributions and recording `p[i]` — the probability of choosing Category 1 on trial $i$ — before each trial's update. The update now happens on *all* trials (including the last), consistent with the R simulation.
* **Model block**: Prior on `log_r` plus the Bernoulli likelihood using the pre-computed `p` vector.
* **Generated quantities**: `log_lik[i]` reads directly from `p[i]` — no re-computation of the Kalman filter required.

---

## Fitting the Model to Simulated Data

We use Pathfinder initialization before HMC sampling, as mandated for nonlinear cognitive models:

```{r, 12 fit single agent}
# Select one simulated agent with r_value = 1.0
agent_to_fit <- prototype_responses |>
  filter(r_value_true == 1.0, agent_id == 1)

stopifnot(nrow(agent_to_fit) == total_trials)

proto_data_single <- list(
  ntrials            = nrow(agent_to_fit),
  nfeatures          = 2L,
  cat_one            = agent_to_fit$category_feedback,
  y                  = agent_to_fit$sim_response,
  obs                = as.matrix(agent_to_fit[, c("height", "position")]),
  initial_mu_cat0    = c(2.5, 2.5),
  initial_mu_cat1    = c(2.5, 2.5),
  initial_sigma_diag = 10.0,
  prior_logr_mean    = LOG_R_PRIOR_MEAN,
  prior_logr_sd      = LOG_R_PRIOR_SD
)

fit_filepath_single <- "simmodels/W12_proto_single_fit.rds"
if (regenerate_simulations) {
  pf_single <- mod_prototype_single$pathfinder(
    data = proto_data_single, seed = 123, num_paths = 4, refresh = 0
  )

  fit_proto_single <- mod_prototype_single$sample(
    data            = proto_data_single,
    init            = pf_single,
    seed            = 123,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1000,
    iter_sampling   = 1500,
    refresh         = 300,
    adapt_delta     = 0.9
  )
  fit_proto_single$save_object(fit_filepath_single)
  cat("Single-agent prototype model fitted and saved.\n")
} else {
  if (file.exists(fit_filepath_single)) {
    fit_proto_single <- readRDS(fit_filepath_single)
    cat("Loaded existing single-agent prototype fit.\n")
  } else {
    cat("Fit file not found. Set regenerate_simulations = TRUE.\n")
    fit_proto_single <- NULL
  }
}

if (!is.null(fit_proto_single)) {
  fit_proto_single$cmdstan_diagnose()

  param_summary <- fit_proto_single$summary(variables = c("log_r", "r_value"))
  print(param_summary)

  true_r     <- agent_to_fit$r_value_true[1]
  true_log_r <- log(true_r)
  cat("\nTrue log_r =", round(true_log_r, 3), "  True r =", true_r, "\n")

  draws_df <- as_draws_df(fit_proto_single$draws(variables = c("log_r", "r_value")))

  p_logr <- ggplot(draws_df, aes(x = log_r)) +
    geom_histogram(aes(y = after_stat(density)),
                   fill = "#009E73", alpha = 0.7, bins = 30) +
    geom_vline(xintercept = true_log_r, color = "red", linetype = "dashed") +
    labs(title = "Posterior: log(r)", x = "log(r)")

  p_r <- ggplot(draws_df, aes(x = r_value)) +
    geom_histogram(aes(y = after_stat(density)),
                   fill = "#56B4E9", alpha = 0.7, bins = 30) +
    geom_vline(xintercept = true_r, color = "red", linetype = "dashed") +
    labs(title = "Posterior: r (natural scale)", x = "r")

  print(p_logr | p_r)
}
```

### LOO and the LFO-CV Caveat

```{r, 12 loo single}
if (!is.null(fit_proto_single)) {
  log_lik_mat   <- fit_proto_single$draws("log_lik", format = "matrix")
  loo_proto     <- loo::loo(log_lik_mat, cores = 4)
  print(loo_proto)
  plot(loo_proto, label_points = TRUE,
       main = "PSIS-LOO Pareto-k: Kalman Prototype Model")
}
```

> **⚠️ The same LFO-CV caveat as Ch. 11 applies here.** The Kalman filter is
> path-dependent: the choice probability at trial $t$ depends on the filter
> state accumulated from trials $1, \ldots, t-1$. Leaving out observation $t$
> in standard LOO implicitly conditions on a counterfactual filter state.
> Leave-future-out cross-validation (LFO-CV, Bürkner et al. 2020) is the
> correct tool; the standard LOO is reported here for within-chapter
> comparability with the GCM and rules model, and all ELPD comparisons in
> Ch. 15 should be interpreted as approximations pending LFO-CV confirmation.

### Prior Sensitivity Analysis

```{r, 12 prior sensitivity}
if (!is.null(fit_proto_single)) {
  ps <- priorsense::powerscale_sensitivity(fit_proto_single)
  print(ps)
  priorsense::powerscale_plot_dens(ps, variables = c("log_r", "r_value"))
}
```

---

## Parameter Recovery: Can We Retrieve the Noise Parameter?

A crucial validation check: if we simulate data with a known `r_value`, can our Stan model accurately estimate it back? We test recovery systematically across the plausible range of `r_value`, with multiple independent datasets per true value.

```{r, 12 recovery_simulation}
# Run one recovery iteration for a given true r_value and dataset seed.
# Each dataset gets its own per-agent randomized schedule.
run_single_recovery <- function(true_r_value, iteration, stimulus_info, n_blocks) {
  subject_seed <- iteration * 1000 + round(true_r_value * 100)

  # Simulate one dataset with a uniquely randomized schedule
  sim_dat <- simulate_prototype_agent(
    agent_id     = 1L,
    r_value      = true_r_value,
    stimulus_info = stimulus_info,
    n_blocks     = n_blocks,
    subject_seed = subject_seed
  )

  stan_dat <- list(
    ntrials            = nrow(sim_dat),
    nfeatures          = 2L,
    cat_one            = sim_dat$category_feedback,
    y                  = sim_dat$sim_response,
    obs                = as.matrix(sim_dat[, c("height", "position")]),
    initial_mu_cat0    = c(2.5, 2.5),
    initial_mu_cat1    = c(2.5, 2.5),
    initial_sigma_diag = 10.0,
    prior_logr_mean    = LOG_R_PRIOR_MEAN,
    prior_logr_sd      = LOG_R_PRIOR_SD
  )

  pf <- tryCatch(
    mod_prototype_single$pathfinder(data = stan_dat, num_paths = 4,
                                     refresh = 0, seed = subject_seed),
    error = function(e) NULL
  )
  init_val <- if (!is.null(pf)) pf else 0.5

  fit <- tryCatch(
    mod_prototype_single$sample(
      data = stan_dat, init = init_val,
      seed = subject_seed, chains = 2, parallel_chains = 2,
      iter_warmup = 800, iter_sampling = 1000,
      refresh = 0, adapt_delta = 0.9
    ),
    error = function(e) { cat("Fit error:", e$message, "\n"); NULL }
  )
  if (is.null(fit)) return(NULL)

  draws_r <- fit$draws("r_value", format = "df")
  tibble(
    true_r_value    = true_r_value,
    iteration       = iteration,
    estimated_r_mean = mean(draws_r$r_value),
    est_lower_95    = quantile(draws_r$r_value, 0.025),
    est_upper_95    = quantile(draws_r$r_value, 0.975),
    had_divergence  = any(fit$diagnostic_summary(quiet=TRUE)$num_divergent > 0)
  )
}

r_values_to_test <- c(0.1, 0.5, 1.0, 2.0, 3.0, 5.0)
n_recovery_iters <- 5   # increase to 20–50 for publication-quality recovery

recovery_file <- "simdata/W12_prototype_recovery_results.rds"
if (regenerate_simulations || !file.exists(recovery_file)) {
  cat("Running parameter recovery...\n")
  recovery_plan <- expand_grid(
    true_r_value = r_values_to_test,
    iteration    = seq_len(n_recovery_iters)
  )

  recovery_results <- future_pmap_dfr(
    list(true_r_value = recovery_plan$true_r_value,
         iteration    = recovery_plan$iteration),
    run_single_recovery,
    stimulus_info = stimulus_info,
    n_blocks      = n_blocks,
    .options = furrr_options(seed = TRUE)
  )
  saveRDS(recovery_results, recovery_file)
  cat("Parameter recovery complete and saved.\n")
} else {
  recovery_results <- readRDS(recovery_file)
  cat("Recovery results loaded.\n")
}

ggplot(recovery_results, aes(x = true_r_value, y = estimated_r_mean)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_errorbar(aes(ymin = est_lower_95, ymax = est_upper_95,
                    color = had_divergence),
                width = 0.08, alpha = 0.7) +
  geom_point(aes(color = had_divergence), size = 3) +
  scale_color_manual(values = c("FALSE" = "#0072B2", "TRUE" = "#D55E00"),
                     name = "Had divergence") +
  labs(
    title    = "Parameter Recovery: Prototype Model (r_value)",
    subtitle = paste0("Points: posterior mean; bars: 95% CrI; n=",
                      n_recovery_iters, " datasets per true value.\n",
                      "Dashed line = perfect recovery."),
    x = "True r_value", y = "Estimated r_value"
  )
```

**Discussion of Recovery Results**: Examine whether the points fall close to the diagonal line and whether the 95% credible intervals contain the true value in approximately 95% of cases. Divergent iterations (orange) indicate regions of parameter space where the geometry is more challenging. If recovery degrades at the extremes of the `r_value` range, this reflects the inherent identifiability limits of single-session data: very small and very large `r` both produce learning curves that eventually approach similar asymptotes, and short experiments may not provide enough trials to reliably distinguish them on the basis of learning trajectory alone.

---

## Simulation-Based Calibration (SBC)

Parameter recovery at specific true values is a necessary but not sufficient validation. SBC checks that the full posterior is correctly calibrated across the entire prior distribution — that if the true `log_r` were drawn from `Normal(0, 1)`, the posterior would on average contain the true value in the appropriate fraction of its probability mass. We use the `SBC` package for consistency with Chapters 6, 8, and 11.

```{r, 12 sbc prototype, eval=run_intensive_checks}
n_sbc_iterations_proto <- 500   # use >= 1000 for publication-quality SBC

proto_sbc_generator <- SBC_generator_function(
  function(i) {
    # 1a. Draw true parameter from its prior
    log_r_true <- rnorm(1, LOG_R_PRIOR_MEAN, LOG_R_PRIOR_SD)
    r_true     <- exp(log_r_true)

    # 1b. Simulate one dataset with its own randomized schedule
    sbc_sched <- make_subject_schedule(stimulus_info, n_blocks, seed = 20000 + i)
    sim       <- prototype_kalman(
      r_value    = r_true,
      obs        = as.matrix(sbc_sched[, c("height", "position")]),
      cat_one    = sbc_sched$category_feedback,
      initial_mu = list(rep(2.5, 2), rep(2.5, 2))
    )

    list(
      variables = list(log_r = log_r_true),
      generated = list(
        ntrials            = nrow(sbc_sched),
        nfeatures          = 2L,
        cat_one            = as.integer(sbc_sched$category_feedback),
        y                  = as.integer(sim$sim_response),
        obs                = as.matrix(sbc_sched[, c("height", "position")]),
        initial_mu_cat0    = c(2.5, 2.5),
        initial_mu_cat1    = c(2.5, 2.5),
        initial_sigma_diag = 10.0,
        prior_logr_mean    = LOG_R_PRIOR_MEAN,
        prior_logr_sd      = LOG_R_PRIOR_SD
      )
    )
  },
  n = n_sbc_iterations_proto
)

proto_sbc_backend <- SBC_backend_cmdstan_sample(
  mod_prototype_single,
  iter_warmup   = 1000,
  iter_sampling = 1000,
  chains        = 2,
  adapt_delta   = 0.9,
  refresh       = 0
)

sbc_results_proto <- compute_SBC(
  proto_sbc_generator,
  proto_sbc_backend,
  cache_path = "simdata/W12_sbc_proto_cache",
  keep_fits  = FALSE
)

# ECDF-difference plot: flat curves within the simultaneous band indicate
# well-calibrated posteriors. The prototype model has only one scalar
# parameter, making this the cleanest possible SBC check.
plot_ecdf_diff(sbc_results_proto)
plot_rank_hist(sbc_results_proto)
```

With only a single scalar parameter and a deterministic Kalman filter likelihood, the prototype model is the easiest of the three categorization models to validate with SBC. If the ECDF-difference plot shows a flat line within the simultaneous confidence band, we have strong evidence that the model code and prior form a coherent, well-calibrated inference engine. Any systematic deviation — U-shaped, bell-shaped, or monotone drift — would identify a miscalibration in the `log_r` direction and warrant investigation.

---

## Cognitive Insights and Model Comparison

The Kalman filter prototype model offers valuable insights:

* **Incremental, Adaptive Learning**: It formalizes how category knowledge can be built up one example at a time, with learning automatically slowing down as confidence increases.
* **Uncertainty Representation**: It explicitly models the learner's uncertainty about category prototypes — something the basic GCM does not directly represent.
* **Memory Efficiency**: It achieves learning by storing only summary statistics (mean and covariance) per category, aligning with intuitions about cognitive limits.

### Comparing to Exemplar Models (GCM)

* The GCM stores every detail; the prototype model abstracts.
* The GCM is sensitive to specific outliers; the prototype model is less so (outliers get averaged in).
* The GCM requires explicit attention weights ($w$) as free parameters; attention in the Kalman filter can emerge implicitly if uncertainty differs across dimensions (though not explored in detail here).
* The GCM's sensitivity parameter $c$ controls generalization globally; the Kalman filter's $r$ controls how rapidly the prototype absorbs new information, providing a different kind of flexibility.

### Limitations

* **Abstraction Loses Detail**: Information about specific examples and within-category variance beyond the prototype covariance is discarded. Hybrid models attempt to combine both approaches.
* **Single Prototype**: Assumes categories have a single central tendency, struggling with disjunctive categories (e.g., "mammals that fly or swim"). Extensions using multiple prototypes address this.
* **Sequential Structure**: Like the GCM, the Kalman filter is path-dependent. Standard LOO overestimates out-of-sample predictive accuracy; LFO-CV (Bürkner et al. 2020) is required for rigorous model comparison.

## Conclusion

The Kalman filter prototype model provides a dynamic, computationally plausible account of how categories might be learned by abstracting central tendencies and tracking uncertainty. It stands as a compelling alternative to exemplar-based approaches, highlighting the fundamental trade-off between detailed instance memory and abstract summary representation in categorization. As a Stan model it has a particularly clean architecture: with a single parameter and a fully deterministic likelihood, Pathfinder initialization is reliable, parameter recovery is tractable, and SBC is straightforward. The next chapter introduces a third perspective — rule-based models — which makes categorization decisions through explicit hypothesis testing rather than continuous similarity to either stored examples or running averages.
